### 2010-02-25 jar.tex
first commit

\documentclass[a4paper,10pt, runningheads]{llncs}
\usepackage{url}
\usepackage{bbm}
\usepackage{ucs}
% \usepackage[utf8x,utf8]{inputenc}
\newcommand{\N}{\ensuremath{\mathbbm{N}}}
\newcommand{\Z}{\ensuremath{\mathbbm{Z}}}
\newcommand{\Q}{\ensuremath{\mathbbm{Q}}}
%include coq2tex.tex
\arrayhs%i.e. no page breaks in code blocks
\divide\abovedisplayskip 4
\divide\belowdisplayskip 4
\begin{document}
\title{The algebraic hierarchy in type theory using type classes}
\author{Bas Spitters \and Eelis van der Weegen}
\institute{Radboud University Nijmegen}
\maketitle
\begin{abstract}
We propose generic design patterns to organize algebraic structures using type classes inside the
Coq system. We support multiple inheritance, sharing of notations and theories, and automated
structure inference. A library for many-sorted universal algebra abstracts from the specific
algebraic structures and aids the organization of the library. Category theory provides a further
useful abstraction. We provide abstract
interfaces for $\N$, $\Z$, $\Q$ as the initial semiring, the initial ring and a field of
fractions. Similarly, we provide an interface for polynomials over a ring $R$ using $R$-algebras.
We show that the standard implementations of these are in fact instances of these type classes.
We report on the efficiency penality of these abstractions.

Finally, we use type classes to conveniently implement a quote function using the prolog-like
abilities of type class unification. We integrate this with our development of universal algebra.
\end{abstract}

\section{Introduction}
Developing formalized libraries for mathematics requires a detailed analysis of the
architecture of mathematics~\cite{C-corn,DBLP:conf/types/HaftmannW08}, similar to that of a large
software project.
The development of an algebraic hierarchy is an important aspect. Challenges include: 1. multiple
inheritance: A ring is a type with two operations. How do we inherit from the two theories of
monoids? 2. Sharing of notations. 3. Combining syntactic and semantic methods, usually this is
addressed by reflection.

Several solutions have been proposed for the Coq theorem prover~\cite{Coq,BC04}: dependent
records~\cite{DBLP:journals/jsc/GeuversPWZ02} (a.k.a.\ telescopes), packed classes~\cite{Packed} and
occasionally modules. Although the latter option is quickly discarded%
\footnote{Modules are not first class, but currently they seem to be superior for extraction to
OCaml~\cite{letouzey02}. Modules may also be seen as parametrised sections. They allow management
of namespaces, notations, hints, etc. An |Import| command imports all the constants from a
module. Thus allowing the user to type |B| instead of |A.B|. Similar functionality would be useful
for type classes.}
algebraic theories as `interfaces', as opposed to `modules', has been an important
guiding principle~\cite{Packed}.

In this fashion we implement\footnote{The sources are available
at:~\url{http://www.eelis.net/research/math-classes/}}
 algebraic theories as type classes, not instances, thus addressing the
challenges above; see Section~\ref{classes}. Type classes have recently entered
Coq~\cite{DBLP:conf/tphol/SozeauO08}. In Coq,
type classes are \emph{first class} citizens based on dependent records. The presence of dependent types makes them even more
powerful than the already succesful second class implementations in Haskell and Isabelle. Moreover,
Coq's type classes are well integrated with setoid-rewriting~\cite{Setoid-rewrite}.

We abstract from concrete algebra and formalize many-sorted universal algebra, including a
construction of
free $\Sigma$-algebras; see Section~\ref{univ}. We are also concerned with efficiency.
In fact, our aim is to use the present development as a new basis for Bishop's program in type
theory~\cite{typesreal-article}.\marginpar{More}

So, we would like to replace data structures with more efficient ones. Concretely,
Coq provides unary, binary and machine numbers. We use type classes polymorphism to
develop a unified library for them; see Section~\ref{modul}.
The interfaces for $\N$, $\Z,\Q$ are the initial semiring, the initial ring and a field of
fractions; see Section~\ref{interfaces}. In future work we aim to base our development of the reals
on an abstract dense set, allowing us to use the efficient dyadic
rationals~\cite{boldo2009combining} as a base for exact real number computation in
Coq~\cite{Riemann,Oconnor:real}.

The paper is organised as follows: Section~\ref{classes} starts with a development of the concrete
algebraic hierarchy addressing the issues raised above. Section~\ref{univ} abstracts to universal
algebra, we develop the basic theory to the point of the first homomorphism theorem.
Section~\ref{interfaces} provides interfaces to usual datastructures by providing their universal
properties. To allow symbolic manipulation of semantic objects one needs a quote function. Usually,
this function is implemented in Coq's tactics language. A more convenient way is to use type class
unification; see section~\ref{quote}. In section~\ref{canonical}, we end with a short comparision
between type classes and canonical structures.

In this paper we focus on the Coq proof assistant. We conjecture that the methods can be transfered
to any type theory based proof assistant supporting type classes such as
Matita~\cite{asperti2007user}.

\section{Preliminaries}
\subsection{Coq type theory}
The Coq proof assistant is based on the calculus of inductive
constructions~\cite{CoquandHuet,CoquandPaulin}.\marginpar{More}
This type theory lacks quotient types, as it would make type checking undecidable. Instead it
supports setoids (aka Bishop sets), a pair of a type and an equivalence
relation~\cite{Bishop67,Hofmann,Capretta}. Recently, improved support has been added to the
system~\cite{Setoid-rewrite}. Functions respecting these equivalence relations are called |Proper|.
As we will see in Section~\ref{univ2} below this functionality can also be used to work with
quotient algebras.

\subsection{Type classes}
Implementation in Ltac, Notation etc
Coercions

\section{To bunde or not to bundle? (Or: What to bundle?)}

How should we define reflexive, transitive, and symmetric relations? In a first attempt, we
might define:
%format → = "\rightarrow"
%format Π = "\Pi"
\begin{code}
  Record ReflexiveRelation: Type :=
    { rr_carrier: Type
    ; rr_rel: rr_carrier → rr_carrier → Prop
    ; rr_proof: Π x, rr_rel x x }.
\end{code}
and similar for transitive and symmetric relations.

With this approach, we run into problems when we try to define |EquivalenceRelation| in terms of
|ReflexiveRelation| and its two siblings, because when posit three objects of type
|ReflexiveRelation|, |TransitiveRelation|, and |SymmetricRelation|, respectively, we have to
add equalities expressing that they talk about the same carrier and relation, and these equalities
are a nightmare to work with (because the three carriers and relations won't be convertible).
Technically, this leads to the research in manifest fields, the \emph{with} construction, to which
we return in Section~\ref{manifest}.

As a first step we would define |ReflexiveRelation| as:
\begin{code}
  Record ReflexiveRelation (carrier: Type): Type :=
    { rr_rel: rr_carrier → rr_carrier → Prop
    ; rr_proof: Π x, rr_rel x x }.
\end{code}

This would solve the problem for the carrier, but |EquivalenceRelation| would still need equalities
for the relation:
\begin{code}
  Record EquivalenceRelation: Type :=
    { er_carrier: Type
    ; er_refl: ReflexiveRelation er_carrier
    ; er_sym: SymmetricRelation er_carrier
    ; er_trans: TransitiveRelation er_trans
    ; same_rel: rr_rel er_refl = sr_rel er_sym
    ; same_rel': rr_rel er_refl = tr_rel er_trans }.
\end{code}

A natural solution is to pull the relation out of |ReflexiveRelation| (and its siblings) as well:
\begin{code}
  Record ReflexiveRelation (carrier: Type)
                           (rel: carrier → carrier → Prop): Prop :=
    { rr_proof: Π x, rr_rel x x }.
\end{code}
or equivalently:
\begin{code}
  Definition ReflexiveRelation (carrier: Type)
                               (rel: carrier → carrier → Prop): Prop :=
    Π x, rr_rel x x.
\end{code}
This is how reflexivity is defined in the standard library, and this is entirely sensible. The
change from |Type| to |Prop| reflects the fact that |ReflexiveRelation| is now strictly a property,
proofs of which can be nicely kept opaque;%
\footnote{Coq distinguishes between transparent and opague definitions. The former can be unfolded,
whereas this is impossible for the latter.}
we will return to this issue in Section~\ref{explicit}.
We can now define |EquivalenceRelation| without awkward equalities:
\begin{code}
  Record EquivalenceRelation: Type :=
    { er_carrier: Type
    ; er_rel: er_carrier → er_carrier → Prop
    ; er_refl: ReflexiveRelation er_carrier er_rel
    ; er_sym: SymmetricRelation er_carrier er_rel
    ; er_trans: TransitiveRelation er_trans er_rel }.
\end{code}

But looking at |EquivalenceRelation|, we see that it now essentially has the same form as the
original naive |ReflexiveRelation| record we started with. Clearly, equivalence relations
and reflexive relations are analogous concepts, and by the same argument we
conclude that |EquivalenceRelation| should really be defined as:

\begin{code}
Record EquivalenceRelation (er_carrier: Type)
                             (er_rel: er_carrier → er_carrier → Prop): Prop :=
  { er_refl: ReflexiveRelation er_carrier er_rel
  ; er_sym: SymmetricRelation er_carrier er_rel
  ; er_trans: TransitiveRelation er_trans er_rel }.
\end{code}

After all, when we define a type with
two different equivalence relations on it,
such as the quotient of a setoid,
we have the same considerations as for |ReflexiveRelation| above.

At the risk of overstressing the point, let us look at how the same problem and solution keep
appearing in composite structures. Suppose we use a definition of |Monoid| that bundles the
relation. Then if we proceed to define |SemiRing| as a double |Monoid|, we face a dilemma:
\begin{code}
  Record SemiRing A: Type :=
    { mult_monoid: Monoid A 
    ; plus_monoid: Monoid A 
    ; plus_comm: Commutative (mon_op plus_monoid)
    ; mult_comm: Commutative (mon_op mult_monoid)
    ; mult_0_l: Π x: A, ?
\end{code}
Which equality do we use in the expression ?, |mult_monoid|'s equality or |plus_monoid|'s?
Again, for this definition to be usable at all, we would need to add an equality expressing that the
two monoids contain the same equivalence relation. And again, the proper solution is to change
Monoid and unbundle the relation. But here, we have a choice as to how much we unbundle exactly.
(For the moment we ignore the question of whether the unit and binary operator should be unbundled
as well, focusing only on the equivalence relation and its equivalence proof.)

We \emph{could} become over-eager and decide to unbundle the equivalence proof term as well:
\begin{code}
Record Monoid (A: Type) (R: A → A → Prop)
  (e: Equivalence R) : Type := ...
\end{code}
However, this way if we happen to have two proofs |p| and |q| that a certain relation is an
equivalence, we end up with two non-convertible ways to express that the relation (together with
some unit and binary operation) form a Monoid: 'Monoid A R p ...' and 'Monoid A R q ...'. We must
either try to deal with this (by proving that Monoid proofs with different Equivalence proofs are
interchangable, and hope for automation, such as type class instance resolution, to use them
interchangably), or take great care to ensure that |Equivalence| proof terms are always shared,
essentially elevating them to `values', which goes against the principle of proof
irrelevance.\marginpar{What if we add proof irrelevance?}

Neither of these options are very appealing. We really should just make |Monoid| take as parameters
only the operations/relations. If we treat the monoid's unit and binary operation the same way, this
means that |Monoid| ends up in |Prop|, and that its instances can be left opaque, which is really
how things ought to be.

Another benefit of this approach is that it works very smoothly when these records are in fact type
classes. When relations and operations are bundled and do \emph{not} appear as parameters of the
structure record/class, expressing as a type class constraint that some collection of relations and
operations posess the structure seems to require manifest fields; see Section~\ref{manifest}. With
our unbundled approach, on the other hand, such type class constraints are easily expressed without
any extensions as simple type inhabitation queries
% (e.g. ``hey Coq, find me a proof of [Monoid some_rel some_op some_unit]'')
of types living in Prop, so that we don't even care what instances are found (thus embracing proof
irrelevance).

We remark that leaving the equivalence proofs bundled means that when one has a proof that something
is a semiring, one has two proofs (via the two monoids) that the equivalence relation is an
equivalence. Fortunately, having redundant proofs like this is entirely harmless when one's terms
never refer (directly or indirectly) to proofs, which is the case when the relations and operations
they are composed of are left unbundled. Here, too, we enjoy the warm embrace of proof irrelevance.

From the discussion so far, we derive some principles:
\begin{enumerate}
 \item structural concepts (like Reflexive, Setoid, Monoid) should have their structural components
(i.e. carriers, relations, and operations) as parameters, and nothing else;
 \item (as a consequence) these concepts should be defined as predicates (which includes records and
classes living in Prop), and their proofs may be kept opaque;
 \item (as a second consequence) algebraic expressions need/should never refer (directly or
indirectly) to proofs.
\end{enumerate}
% \setcounter{enumi}{3}

This list will grow in subsequent sections.

The principles listed so far are not particularly controversial, in fact this
approach ensures maximum flexibility~\cite{Hints}. However,
there is a common perception that unbundling carriers, relations, and operations this way invariably
leads either to unmanagably big algebraic expressions (involving endless projections and/or implicit
arguments), or to unmanageably long and tedious enumeration of carriers/relations/operations, or to
problems establishing canonical names for operations (since we no longer project them out of
structure instances all the time).

In the next sections, we show how systematic use of type classes (and their support infrastructure
in Coq), combined with the use of a special form of type class we call an `operational type class'
(as opposed to a `structural' type class we might use in place of the records shown thus far), lets
us avoid these problems.

\section{Managing unbundled structure with type classes}

So far, we have only looked at the definitions of the structures themselves, and have not yet
considered the perspective of someone trying to prove some theorems about these abstract structures.
We would like to approximate the common mathematical vernacular, where one
simply says:

for $x, y, z$ in a semigroup $G$, $x * y * z = z * y * x$, $\ldots$

Suppose we have defined |SemiGroup| in line with the aforementioned principles:
\begin{code}
Record SemiGroup (A: Type) (eq: A → A → Prop)
 (op: A → A → A): Prop := ...
\end{code}
Then our starting point is the obvious but awful:
\begin{code}
Lemma assoc1 A eq op (P: SemiGroup A eq op) x y z:
                   eq (op (op x y) z) (op (op z y) x).
\end{code}
As a first cleaning measure, we might hope to introduce some notations to let us use infix notation
for the semigroup operator (say, |&|) and for equality. Unfortunately, at this point the names |eq|
and |op| are just local names and there is no way to predeclare that they are to be associated with
any notation. What we really need are canonical names for these notions, so that we may fashion
notations for them.

This is a typical job for type classes, and they are used in this way in the |Clean| standard
library~\cite{Clean}. Similarly, Haskell uses type classes to overload notations.
We simply follow suit:
\begin{code}
  Class Equiv A := equiv: relation A.
  Class SemiGroupOp A := sg_op: A → A → A.

  Infix "=" := equiv: type_scope.
  Infix "&" := sg_op (at level 50, left associativity).
\end{code}
We choose to use |&| here and reserve the notation |*| for ring multiplication.

We use = for setoid equality, since Leibniz equality seems less important for the formalization of
mathematics. More precisely, setoid equality is the basic concept. In special case, one can prove
that setoid equality coincides with Leibniz equality.

We would have like to use a haskell style definition
\begin{code}
Class Equiv A := (=): relation A.
\end{code}
Unfortunately, currently this is not supported in Coq.

These single-field type classes are given special treatment in that they are not translated into
records the way classes with any other number of fields are, but are instead turned into ordinary
definitions. The effect of this is that |Equiv A| immediately reduces to |relation A|, and that the
equiv `projection' is the identity function whose argument is inferred using instance resolution.
Thus, for a very modest cost in indirection, we gain the ability to use canonical names, and with
it, notations:
\begin{code}
  Record SemiGroup (A: Type) (eq: Equiv A)
                             (op: SemiGroupOp A): Prop := ...
  Lemma assoc1 A eq op (P: SemiGroup A eq op) x y z:
                    x & y & z = z & y & x.
\end{code}
We call these type classes \emph{operational type classes}. We note that operational type classes
allow us to avoid Coq's notation |scope| mechanism.

It is important to point out that although the expression now \emph{looks} like it is bound to some
specific semigroup structure (which with traditional approaches would imply projections from bundles
and/or reference to proofs), it is really only referring near-directly to the actual operations
involved, with the semigroup proof existing only as opaque `knowledge about the operations' which we
may use in the proof. This lack of projections keeps our terms small and independent, and keeps
rewriting simple and sane.

On the other hand, it \emph{does} mean that all the carriers/relations/operations end up in the
context. The context for |assoc1| looks like:
\begin{code}
  A: Type
  eq: Equiv A
  op: SemiGroupOp A
  P: SemiGroup A eq op
\end{code}

While we are not particularly worried about overly large contexts (we feel that this will most
likely not be problematic for any but the most complex formalizations), having to \emph{declare}
these entities (like |eq| and |op| for |assoc1|) \emph{is} a bit of a chore. Fortunately, Coq
provides a feature called implicit generalization, which is exactly what we need.
\marginpar{describe how it works} With it, we can write |assoc1| as follows:
\begin{code}
  Lemma assoc1 `(SemiGroup A) x y z: x & y & z = z & y & x.
\end{code}
Thus, we have reached the mathematical ideal we aimed for.

Let us now stop and think to what extent this approach suffers from all the problems commonly
associated with it. In particular, let us imagine what happens to our terms and contexts when we
want to talk about nested structures such as polynomials over the ring of matrices over a
given ring; see Section~\ref{interfaces}.

\section{Subset types}
BEGIN SPECULATIVE

A record type is \marginpar{mostly?} syntactic sugar for a $\Sigma$-type. In the presence of a
squash operator, |[]:Type->Prop|, we may define a subset type as |\{A\mid B\}=Sig A. [B]|. This is
pursued by Barras~\cite{Barras:subset,Werner}. The program machinery fakes subset types in the
present Coq implementation. Proof irrelevanve in Coq would allow a faithful
implementation.

Squash is a monad and allows many constructions: subsets, unions, images, heterogeneous equality,
\ldots.

Question: Would the meta-theory simplify if we would just add a conversion rule $x\triangleright
\epsilon$ for $x:Prop$? We could add a term constructor $[a]$ which will have type $[A]$. There
seems to be no reason to connect this to |Prop| or impredicativity.

END SPECULATIVE

We add another principle:
\begin{itemize}\setcounter{enumi}{3}
 \item Working hypothesis: `every record can be reordered as a subset'. Consequently, we
rewrite this subset |\{A\mid B\}| as |Record r (a:A):={b:B}|.
\end{itemize}

\section{Algebraic hierarchy using type classes}\label{classes}
\paragraph{Hierarchy.}\label{hierarchy}
In our formalization of algebraic structures we use the structure as (implicit) parameters, but
pack the proofs. This can be nicely organized by using canonical names. For instance, we define:
\begin{code}
Class SemiRing A {e: Equiv A}{plus: RingPlus A}{mult: RingMult A}
                 {zero: RingZero A}{one: RingOne A}:Prop :=
  { semiring_mult_monoid:> Monoid A (op:=mult)(unit:=one)
  ; semiring_plus_monoid:> Monoid A (op:=plus)(unit:=zero)
  ; semiring_plus_comm:> Commutative plus
  ; semiring_mult_comm:> Commutative mult
  ; semiring_distr:> Distribute mult plus
  ; mult_0_l: forall x, 0 * x == 0 }.
\end{code}
where e.g.\ |RingPlus| is the operation class for the semigroup operator.
The notation |:>| makes |semiring_mult_monoid| a coercion: when asked for a monoid it suffices to
provide a ring.

\emph{Structure-as-parameters} helps setoid-rewriting: type class resolution
can find the equivalence relation in the context.
% A similar style should be possible for, say, the |Ring| tactic, instead of
% declaring the ring structure by a separate command, we would rely on type class resolution to find
% it.
We note that |op| does not depend on the proof that |e| is an equivalence. We use Coq's implicit
quantification (|`{}|) to avoid having to write all the parameters when \emph{stating} a theorem
and Coq's maximally inserted implicit arguments to find the parameters when \emph{applying} a
theorem. Both features are new in Coq and stem from the type class implementation.

We mention the trade-off between bigger contexts versus bigger terms. Our contexts are bigger than
those of telescopes or packed classes. In our experience, this has been relatively
harmless\footnote{Coq's data structure for contexts is not very efficient however, we believe this
may be a bottleneck for huge developments. It seems that the data structure chosen
in~\cite{asperti2009compact} will behave better.}%
: most
terms in the context are there to support canonical names. Bigger terms
\emph{do} cause problems: 1. when proofs are part of mathematical objects we need to share these
proofs to allow rewriting. Moreover, it prohibits Opaque proofs and `proof irrelevance'. 2. The
projection paths may not be canonical.

Coercion pullbacks~\cite{Hints} were introduced to address problems with multiple coercions paths,
as in the definition of a semiring: a type with two monoid structures on it. We avoid some
of these problems by explicitly specifying the fields. We emphasize that the semiring properties are automatically derived from the ring
properties, although the properties of a semiring are not structurally included in the ring
properties.

\subsection{Manifest fields}\label{manifest}

Manifest fields in records allow us to fix a field of a dependent record.
Luo~\cite{DBLP:conf/types/Luo08} has proposed coersive subtyping to implement manifest fields.
It was conjectured~\cite{Hints} that implementing the algebraic hierarchy with type classes would
be awkward due to problems with equality. We avoided these problems by shifting the fields
from the dependent record to its argument, thus mimicking manifest fields by using type classes.
Since |op| is \emph{instantiated} with |mult|, we need not worry about the
equality |op=mult|.

\section{Universal algebra}\label{univ}
After having implemented concrete algebraic structures, we generalized some of our constructions
by implementing many-sorted universal algebra. A formalization of universal algebra
already exists in Coq~\cite{DBLP:conf/tphol/Capretta99}. It has recently been extended with more
categorical results~\cite{dominguez2008formalizing}. We have conveniently redeveloped most of this
using type classes. We defined equational theories and their varieties:
the categories of their models. A quote function helps to translate a statement for an instance of
a type class to a statement in the equational logic.

\subsection{Category theory}\label{cats}
Using type classes, we provide a basic library for category theory: categories, functors,
adjunctions, monads,\ldots We follow the usual type theoretical definition of a\marginpar{Use
notation for comp} category~\cite{saibi1995constructive}
\begin{code}
 Class Category O `{!Arrows O} `{forall x y: O, Equiv (x --> y)}
   `{!CatId O} `{!CatComp O}: Prop :=
  { arrow_equiv:> forall x y: O, Setoid (x --> y)
  ; comp_proper:> forall x y z: O,
      Proper (equiv ==> equiv ==> equiv)%signature (@comp _ _ _ x y z)
  ; comp_assoc: forall (w x y z: O) (a: w --> x) (b: x --> y) (c: y --> z),
      comp c (comp b a) == comp (comp c b) a
  ; id_l: forall (x y: O) (a: y --> x), comp cat_id a == a
  ; id_r: forall (x y: O) (a: x --> y), comp a cat_id == a }.
\end{code}
This is based on the 2-categorical idea that we have equality on arrows, but not on objects.
Similarly, we have equality on natural transformations, but not on functors.

%format ◎ = "\circ"
For instance natural transformations are formalized as: |◎|
\begin{code}
  Context `{Category C} `{Category D} (F G: C -> D)
     `{!Functor F Fa} `{!Functor G Ga}.

  Class NaturalTransformation (η: forall c: C, F c --> G c): Prop :=
    natural: forall (x y: C) (f: x --> y), η y ◎ fmap F f == fmap G f ◎ η x.
\end{code}

The notation |◎| denotes function composition.
Here |fmap F| explicitly refers to the arrow part of the functor |F|.\\
In mathematics one would apply $F$ to both objects and arrows ($F x$, $F f$). However, this does
not seem to fit with type theory, hence we follow the customs in functional programming.

The proofs are short and follow the textbook easily.\marginpar{really?}

We provide the category of setoids, the category of categories, the dual of a category, products,
and hence co-products, the category of $\Sigma$-algebras.

We provide the construction of the term-algebra for a signature $\Sigma$ and show that it is
initial in the category of $\Sigma$-algebras. We proceed to prove that this provides a left-adjoint
to the forgetful functor. The resulting expression monad will be used below.

\subsection{Universal algebra continued}\label{univ2}

We have developed basic universal algebra~\cite{meinke1993universal} upto the first homomorphism
theorem.~\marginpar{Description of the work.}
%perhaps the soundness theorem for equational logic.

In order to prove the first homophism theorem, we define a congruence:
%theory/ua_congruence.v
\begin{code}
Class Congruence: Prop :=
    { congruence_proper:> Π s, Proper (equiv ==> equiv ==> iff) (e s)
    ; congruence_quotient:> @Algebra et v e _
    }.
\end{code}
We note the use of |Proper|.

We define the forgetful functor from the category of $\Sigma$-algebras to the category of sets.

We define the term algebra consisting of closed terms, terms with the empty type as set of
variables. % extensionally equal closed term???
We show that the term algebra is initial. One may expect that this could be useful to
\emph{define}, say, the natural numbers as the initial semiring. However, it seems difficult to
prove decidability of equality, as it seems to require a normalization procedure.

We define the product of two algebras and show that it is the categorical product.

Given a sub\emph{set} of an algebra, we define the sub\emph{algebra} generated by it.
An interesting fact to mention is the use of the following heterogeneous equality between
elements of the algebra and of the subalgebra, i.e.\ terms of different types may be equal.
\begin{code}
  Fixpoint heq {o}: op_type carrier o -> op_type v o -> Prop :=
    match o with
    | constant x => fun a b => `a == b
    | function x y => fun a b => forall u, heq (a u) (b (`u))
    end.
\end{code}

We connect the abstract theory to the concrete theory. Concretely, let $\Sigma$ be the theory
of monoids. Given a monoid, we can construct the corresponding $\Sigma$-algebra and an
object in the category of $\Sigma$-algebras. Conversely, given an object in the category, we
construct a monoid.
\begin{code}
 Variable o: Variety theory.
 Global Instance from_object: Monoid (o tt).
\end{code}
Similarly, for arrows.

%There is a slight problem when defining the forgetful functor from the category of
%$\Sigma$-algebras to Sets. This problem is caused by our use of the category of all
%categories\footnote{Perhaps we should have used higher categories for this.}.
%We conjecture that this problem disappears when Coq's universe polymorphism is extended to
%definitions.
\section{Interfaces using category theory}\label{interfaces}\label{modul}
The current work was stimulated by our desire~\cite{Riemann} to move between several
implementations in a flexible way. In haskell type class polymorphism is used for this purpose.
Hence, we do the same in type theory.

We have characterized the naturals as the initial object in the category of semirings and derived
many of their properties from this interface. Unary, binary and machine numbers are
instances of this interface, so we can directly apply these results to all these instances.
Similarly, for the integers are the initial ring. The rationals are the field of fractions of the
integers. 
Given a ring $R$, the $R$-algebra $R[X]$ of polynomials, is the free $R$-algebra of a set $X$.
We provide two implementations of polynomials: the
standard representation using lists of coefficients and the Bernstein representation. A Bernstein
basis polynomial is one of the form:
\[b_{\nu,n}(x) = {n \choose \nu} x^{\nu} \left( 1 - x \right)^{n - \nu}, \quad \nu = 0, \ldots, n.\]
A Bernstein polynomial is a linear combination of these basic polynomials. Bernstein polynomials
have been used for efficient computations inside the Coq system~\cite{ZumkellerPhD}.

We encourage more efficient implementations by assigning the default implementation a
low priority. For example, the distance function on the natural numbers, which is derived from its
semiring structure, is assigned priority 10.
\begin{code}
  Global Program Instance: NatDistance N | 10 := ...
\end{code}

The context grows as follows:\marginpar{expand}

\section{Type class quoting}\label{quote}
Unification hints~\cite{Hints} allow one to semi-automatically construct a quote
function\footnote{Gonthier provides similar functionality using canonical structures}.
Unification hints are absent in Coq. Fortunately, type classes provide similar functionality as
we will now show.

We define a term language.
\begin{code}
Inductive Expr (V: Type) := Mult (a b: Expr V) | Zero | Var (v: V).
\end{code}
This defines a monad on Type\marginpar{Do this!}.
The expression type is parameterized over the set of variable indices. Hence, we diverge
 from~\cite{Hints}, which uses |nat| for variables thereby introducing bounds problems and
 dummy variables.

\noindent An expression is only meaningful in the context of a variable assignment:\footnote{nat is
an Expr-algebra. eval is map vars composed with the algebra map.}
\begin{code}
Definition Value := nat.
Definition Vars V := V -> Value.

Fixpoint eval {V} (vs: Vars V) (e: Expr V): Value :=
  match e with
  | Zero => 0
  | Mult a b => eval vs a * eval vs b
  | Var v => vs v
  end.
\end{code}
%
Monads are trees with grafting. As explained by sigfpe~\cite{MonadsGrafting}, the tree monad
$TX:=1+T^2X$, may be seen as the prototypical monad. Indeed, the |Expr| monad, i.e.\ the free
$\Sigma$-algebra monad, behaves much like the tree monad. We now provide a direct definition of the
corresponding bind operation.

\begin{code}
Fixpoint bind {V W} (f: V -> Expr W) (e: Expr V): Expr W :=
  match e with
  | Zero => Zero
  | Mult a b => Mult (bind f a) (bind f b)
  | Var v => vs v
  end.
\end{code}

We have shown that |Expr|, i.e.\ |bind| together with |Var|, is a monad on the category of Types
with extensional functions between them~\marginpar{Really?}.

\noindent Some simple combinators for variable packs:
%
\begin{code}
Definition novars: Vars False := False_rect _.
Definition singlevar (x: Value): Vars unit := fun _ => x.
Definition merge {A B} (a: Vars A) (b: Vars B): Vars (A+B) :=
  fun i => if i then a j else b j.
\end{code}

\noindent These last two combinators are the `constructors' of an implicitly defined subset of
 Gallina terms, representing heaps, for which we implement syntactic lookup with type classes.
Given a heap and value, |Lookup| instances give the value's index in the heap:\marginpar{A
canonical structures approach would allow us to do this in Coq?}
\begin{code}
  Class Lookup {A} (x: Value) (f: Vars A) :=
    { lookup: A; lookup_correct: f lookup = x }.
\end{code}
% Context (x: Value) {A B} (va: Vars A) (vb: Vars B).

If the heap is a merge of two heaps and we can find the value's index in the left heap, we can
access it by indexing the merged heap (and vice versa)
\begin{code}
  Global Instance lookup_left `{!Lookup x va}: Lookup x (merge va vb)
    := { lookup := inl (lookup x va) }.
\end{code}
If the heap is just a singlevar, we can easily index it.
\begin{code}
  Global Program Instance: Lookup x (singlevar x) := { lookup := tt }.
\end{code}

One useful operation we need before we get to Quote relates to variables and expression
 evaluation. As its name suggests, |map_var| maps an expression's variable indices.\marginpar{monad}

\begin{code}
Definition map_var {V W: Type} (f: V -> W):
    Expr V -> Expr W :=
  fix F (e: Expr V): Expr W :=
    match e with
    | Mult a b => Mult (F a) (F b)
    | Zero => Zero
    | Var v => Var (f v)
    end.
\end{code}
In |Quote| below, the idea is that |V, l|, and |n| are all input variables, while |V'| and |r| are
output variables (in the sense that we will rely on unification to generate them). |V| and |l|
represent the current heap, |n| represents the value we want to quote, and |V'| and |r'| represent
the heap of newly encountered variables during the quotation.
  This explains the type of quote: it is an expression that refers either to variables from
the old heap, or to newly encountered variables. Finally, |eval_quote| is the usual correctness
property, which now merges the two heaps.

\begin{code}
  Class Quote {V} (l: Vars V) (n: Value) {V'} (r: Vars V') :=
    { quote: Expr (V + V')
    ; eval_quote: @eval (V+V') (merge l r) quote = n }.
\end{code}

Our first instance for |Zero| is easy. The `novars' in the result type reflects the fact that no
new variables are encountered.
\begin{code}
  Global Program Instance quote_zero V (v: Vars V):
    Quote v 0 novars :=
  { quote := Zero }.
\end{code}

The instance for multiplication is a bit more complex. The first line consists of
 variable declarations. The second line is important. `Quote x y z' must be read as
 `quoting y with existing heap x generates new heap z', so the second line basically just
shuffles heaps around.
 The third line contains some |map_var|'s because the heap shuffling must be
reflected in the variable indices, but apart from that it's just constructing a |Mult| term with
quoted subterms.

\begin{code}
Global Program Instance quote_mult V (v: Vars V)
  n V' (v': Vars V') m V'' (v'': Vars V'')
  `{!Quote v n v'} `{!Quote (merge v v') m v''}:
  Quote v (n * m) (merge v' v'') :=
  { quote := Mult (map_var bla (quote n)) (map_var sum_assoc (quote m)) }.
\end{code}
  Now follows the instance where we recognize values that are already in the heap. This
   is expressed by the Lookup requirement, which will only be fulfilled if the Lookup instances
   defined above can find the value in the heap. The novars in the |Quote v x novars| result
   reflects that this quotation does not generate new variables.

\begin{code}
Global Program Instance quote_old_var
  V (v: Vars V) x {i: Lookup x v}:
  Quote v x novars | 8 := { quote := Var (inl (lookup x v)) }.
\end{code}

\noindent Finally, the instance for new variables. We give this a lower priority so that it is only
used if Lookup fails. The |8| in the previous example is a random number less than |9| below. A
similar method is used in Coq's notation mechanism. We strongly recommend a change in
implementation supporting an agda style precedence relation which is only restricted to be a
directed acyclic graph~\cite{danielsson2009parsing}.

\begin{code}
Global Program Instance quote_new_var V (v: Vars V) x:
  Quote v x (singlevar x) | 9 := { quote := Var (inr tt) }.
\end{code}

\noindent As announced this provides a light-weight quoting function\footnote{|eval_quote'| is a
modification of |eval_quote| which works in the empty context.}:
\begin{code}
Goal forall x y (P: Value -> Prop), P ((x * y) * (x * 0)).
  intros. rewrite <- (eval_quote' _).
\end{code}
This turns the goal into | (eval variable_pack quote)|.

It would be interesting to use this technique to implement the |congruence| tactic, which implements
the congruence closure algorithm~\cite{corbineau2007deciding}. One would naturally obtain a tactic
which moreover works on setoid equalities.

\section{Canonical structures}\label{canonical}
Packed classes use canonical structures for the algebraic hierarchy. Both canonical structures and
type classes may be seen as instances of hints in unification~\cite{Hints}. Some uses of canonical
structures can be replaced by type class instances. The user manual (2.7.15) uses canonical
structures to derive the setoid equality on the natural
numbers in the following example |(S x)==(S y)|. We have conveniently reimplemented this with
type classes with superior proof terms. Canonical structures give
\begin{code}
@equiv(Build_Setoid nat (@eq nat)(@eq_equivalence nat))(S x)(S y)
\end{code}

which includes an explicit proof that |(@eq nat)| is an equivalence,
whereas we obtain |@equiv nat (@eq nat) (S x) (S y)|.

\subsection{Big operators}
Canonical structures have been used to provide a uniform treatment~\cite{bertot2008canonical} of big
operators (like $\Pi,\sum, \max$). These operators extend a pair of a binary and a 0-ary operation
to an $n$-ary operation for any $n$. Categorically, one considers the algebra maps from non-empty
lists, lists, inhabited finite sets and finite sets to the carrier of a semigroup, monoid,
commutative semigroup, commutative monoid. Hence we reuse the libraries for lists etc.

We use type classes to deduce the relevant monoid operation and write:
\begin{code}
Definition seq_sum
  `{Sequence A T} `{RingPlus A} `{z: RingZero A}: T -> A
  := @seq_to_monoid A T _ A ring_plus z id.
  Eval compute in seq_sum [3; 2].
\end{code}

\section{Constructive mathematics and proof irrelevance}\label{explicit}
As originally concieved~\cite{ITT,CMCP}, Propositions are Types and we can extract information from
them. In practice, it is convenient to assume that all proofs of a proposition are equal, the
principle of proof irrelevance. These two views clearly contradict eachother. In Coq, a middle way
was choosen, we are forbidden to extract information from proofs, but proofs are not identified.
Here we need to pay, but do not seem to get anything in return.%
\footnote{To be precise, what we get in return is the possibility to extract precisely the program
we want.}
There are various proposals to add the principle of proof irrelevance to Coq, perhaps a
promissing one is the Calculus of implicit
constructions~\cite{miquel2001implicit,barras2008implicit}. Such a change would fit naturally with
the Hoare-style separation between programs and proofs which is used by the |Program| machinery.

On the other hand, we found instances where we needed proof \emph{relevance}. In this case, it lead
to a duplication of code which may have been avoided by universe polymorphism for |Fixpoint|s. In
practice, manual dead-code analysis seems useful. Making Harrop formulas proof irrelevant seems to
be a good first approximation~\cite{lcf:spi:03}.

\section{Conclusions}
Telescopes have been criticized~\cite{Packed} for the lack of multiple inheritance and
the efficiency penalty of a long chain of coercion projections. Packed classes~\cite{Packed} provide
a solution to these problems. We provide an alternative solution. We have already
discussed multiple inheritance in Section~\ref{hierarchy}. Moreover, by exposing the
carrier we avoid the chains of projections.

We conclude that unification hints, canonical structures and type classes all appear to have
similar expressive powers. Canonical structures, being tied to the unifier, are more robust, but
seems to require more inguinity. Type classes are easier to use. Unification hints, may be a
good generalization, but are absent from Coq. We encourage their inclusion to the
Coq system.

An obvious topic for future research is the extension from equational logic to partial Horn
logic~\cite{palmgren2007partial}. Another topic would be to fully, but practically, embrase the
categorical approach to universal algebra~\cite{pitts2001categorical}.

According to |coqwc|, our development consists of 3103 lines of specifications and 707 lines of
proofs.\marginpar{update,targz}

% Alt-Ergo congruence closure parametrized by an equational theory.
% As usual either implement in Coq (verify the algorithm), or Check traces, or Check Alt-Ergo (Work
% in progress).

% http://lara.epfl.ch/dokuwiki/sav09:congruence_closure_algorithm_and_correctness
% I understand that this is the algorithm underlying congruence and perhaps first-order.
% 
% Type classes vs soft typing (Mizar types).
% Type classes and types in homalg?

\paragraph{Acknowledgements.}
This research was stimulated by the new possibilities provided by the introduction of type classes
in Coq. In fact, this seems to be the first substantial development that tries to exploid all their
possibilities. As a consequence, we found many small bugs and unintended behavior in the type
class implementation. All these issues were quickly solved by Matthieu Sozeau. Discussions with
Georges Gonthier and Claudio Sacerdoti Coen have helped to
sharpen our understanding of the relation with Canonical Structures and with Unification Hints.
%Jeremy, Thierry, Claudio, Georges?, Wouter
%\enlargethispage{5mm}
\bibliographystyle{plain}
\bibliography{alg}
\end{document}

### 2010-02-25 processed.tex
first commit

--- 2010-02-25 jar.tex
@@
 \newcommand{\Q}{\ensuremath{\mathbbm{Q}}}
-%include coq2tex.tex
+%% ODER: format ==         = "\mathrel{==}"
+%% ODER: format /=         = "\neq "
+%
+%
+\makeatletter
+\@ifundefined{lhs2tex.lhs2tex.sty.read}%
+  {\@namedef{lhs2tex.lhs2tex.sty.read}{}%
+   \newcommand\SkipToFmtEnd{}%
+   \newcommand\EndFmtInput{}%
+   \long\def\SkipToFmtEnd#1\EndFmtInput{}%
+  }\SkipToFmtEnd
+
+\newcommand\ReadOnlyOnce[1]{\@ifundefined{#1}{\@namedef{#1}{}}\SkipToFmtEnd}
+\usepackage{amstext}
+\usepackage{amssymb}
+\usepackage{stmaryrd}
+\DeclareFontFamily{OT1}{cmtex}{}
+\DeclareFontShape{OT1}{cmtex}{m}{n}
+  {<5><6><7><8>cmtex8
+   <9>cmtex9
+   <10><10.95><12><14.4><17.28><20.74><24.88>cmtex10}{}
+\DeclareFontShape{OT1}{cmtex}{m}{it}
+  {<-> ssub * cmtt/m/it}{}
+\newcommand{\texfamily}{\fontfamily{cmtex}\selectfont}
+\DeclareFontShape{OT1}{cmtt}{bx}{n}
+  {<5><6><7><8>cmtt8
+   <9>cmbtt9
+   <10><10.95><12><14.4><17.28><20.74><24.88>cmbtt10}{}
+\DeclareFontShape{OT1}{cmtex}{bx}{n}
+  {<-> ssub * cmtt/bx/n}{}
+\newcommand{\tex}[1]{\text{\texfamily#1}}	% NEU
+
+\newcommand{\Sp}{\hskip.33334em\relax}
+
+\newcommand{\Conid}[1]{\mathit{#1}}
+\newcommand{\Varid}[1]{\mathit{#1}}
+\newcommand{\anonymous}{\kern0.06em \vbox{\hrule\@width.5em}}
+\newcommand{\plus}{\mathbin{+\!\!\!+}}
+\newcommand{\bind}{\mathbin{>\!\!\!>\mkern-6.7mu=}}
+\newcommand{\rbind}{\mathbin{=\mkern-6.7mu<\!\!\!<}}% suggested by Neil Mitchell
+\newcommand{\sequ}{\mathbin{>\!\!\!>}}
+\renewcommand{\leq}{\leqslant}
+\renewcommand{\geq}{\geqslant}
+\usepackage{polytable}
+
+%mathindent has to be defined
+\@ifundefined{mathindent}%
+  {\newdimen\mathindent\mathindent\leftmargini}%
+  {}%
+
+\def\resethooks{%
+  \global\let\SaveRestoreHook\empty
+  \global\let\ColumnHook\empty}
+\newcommand*{\savecolumns}[1][default]%
+  {\g@addto@macro\SaveRestoreHook{\savecolumns[#1]}}
+\newcommand*{\restorecolumns}[1][default]%
+  {\g@addto@macro\SaveRestoreHook{\restorecolumns[#1]}}
+\newcommand*{\aligncolumn}[2]%
+  {\g@addto@macro\ColumnHook{\column{#1}{#2}}}
+
+\resethooks
+
+\newcommand{\onelinecommentchars}{\quad-{}- }
+\newcommand{\commentbeginchars}{\enskip\{-}
+\newcommand{\commentendchars}{-\}\enskip}
+
+\newcommand{\visiblecomments}{%
+  \let\onelinecomment=\onelinecommentchars
+  \let\commentbegin=\commentbeginchars
+  \let\commentend=\commentendchars}
+
+\newcommand{\invisiblecomments}{%
+  \let\onelinecomment=\empty
+  \let\commentbegin=\empty
+  \let\commentend=\empty}
+
+\visiblecomments
+
+\newlength{\blanklineskip}
+\setlength{\blanklineskip}{0.66084ex}
+
+\newcommand{\hsindent}[1]{\quad}% default is fixed indentation
+\let\hspre\empty
+\let\hspost\empty
+\newcommand{\NB}{\textbf{NB}}
+\newcommand{\Todo}[1]{$\langle$\textbf{To do:}~#1$\rangle$}
+
+\EndFmtInput
+\makeatother
+%
+%
+%
+%
+%
+%
+% This package provides two environments suitable to take the place
+% of hscode, called "plainhscode" and "arrayhscode". 
+%
+% The plain environment surrounds each code block by vertical space,
+% and it uses \abovedisplayskip and \belowdisplayskip to get spacing
+% similar to formulas. Note that if these dimensions are changed,
+% the spacing around displayed math formulas changes as well.
+% All code is indented using \leftskip.
+%
+% Changed 19.08.2004 to reflect changes in colorcode. Should work with
+% CodeGroup.sty.
+%
+\ReadOnlyOnce{polycode.fmt}%
+\makeatletter
+
+\newcommand{\hsnewpar}[1]%
+  {{\parskip=0pt\parindent=0pt\par\vskip #1\noindent}}
+
+% can be used, for instance, to redefine the code size, by setting the
+% command to \small or something alike
+\newcommand{\hscodestyle}{}
+
+% The command \sethscode can be used to switch the code formatting
+% behaviour by mapping the hscode environment in the subst directive
+% to a new LaTeX environment.
+
+\newcommand{\sethscode}[1]%
+  {\expandafter\let\expandafter\hscode\csname #1\endcsname
+   \expandafter\let\expandafter\endhscode\csname end#1\endcsname}
+
+% "compatibility" mode restores the non-polycode.fmt layout.
+
+\newenvironment{compathscode}%
+  {\par\noindent
+   \advance\leftskip\mathindent
+   \hscodestyle
+   \let\\=\@normalcr
+   \(\pboxed}%
+  {\endpboxed\)%
+   \par\noindent
+   \ignorespacesafterend}
+
+\newcommand{\compaths}{\sethscode{compathscode}}
+
+% "plain" mode is the proposed default.
+% It should now work with \centering.
+% This required some changes. The old version
+% is still available for reference as oldplainhscode.
+
+\newenvironment{plainhscode}%
+  {\hsnewpar\abovedisplayskip
+   \advance\leftskip\mathindent
+   \hscodestyle
+   \let\hspre\(\let\hspost\)%
+   \pboxed}%
+  {\endpboxed%
+   \hsnewpar\belowdisplayskip
+   \ignorespacesafterend}
+
+\newenvironment{oldplainhscode}%
+  {\hsnewpar\abovedisplayskip
+   \advance\leftskip\mathindent
+   \hscodestyle
+   \let\\=\@normalcr
+   \(\pboxed}%
+  {\endpboxed\)%
+   \hsnewpar\belowdisplayskip
+   \ignorespacesafterend}
+
+% Here, we make plainhscode the default environment.
+
+\newcommand{\plainhs}{\sethscode{plainhscode}}
+\newcommand{\oldplainhs}{\sethscode{oldplainhscode}}
+\plainhs
+
+% The arrayhscode is like plain, but makes use of polytable's
+% parray environment which disallows page breaks in code blocks.
+
+\newenvironment{arrayhscode}%
+  {\hsnewpar\abovedisplayskip
+   \advance\leftskip\mathindent
+   \hscodestyle
+   \let\\=\@normalcr
+   \(\parray}%
+  {\endparray\)%
+   \hsnewpar\belowdisplayskip
+   \ignorespacesafterend}
+
+\newcommand{\arrayhs}{\sethscode{arrayhscode}}
+
+% The mathhscode environment also makes use of polytable's parray 
+% environment. It is supposed to be used only inside math mode 
+% (I used it to typeset the type rules in my thesis).
+
+\newenvironment{mathhscode}%
+  {\parray}{\endparray}
+
+\newcommand{\mathhs}{\sethscode{mathhscode}}
+
+% texths is similar to mathhs, but works in text mode.
+
+\newenvironment{texthscode}%
+  {\(\parray}{\endparray\)}
+
+\newcommand{\texths}{\sethscode{texthscode}}
+
+% The framed environment places code in a framed box.
+
+\def\codeframewidth{\arrayrulewidth}
+\RequirePackage{calc}
+
+\newenvironment{framedhscode}%
+  {\parskip=\abovedisplayskip\par\noindent
+   \hscodestyle
+   \arrayrulewidth=\codeframewidth
+   \tabular{@{}|p{\linewidth-2\arraycolsep-2\arrayrulewidth-2pt}|@{}}%
+   \hline\framedhslinecorrect\\{-1.5ex}%
+   \let\endoflinesave=\\
+   \let\\=\@normalcr
+   \(\pboxed}%
+  {\endpboxed\)%
+   \framedhslinecorrect\endoflinesave{.5ex}\hline
+   \endtabular
+   \parskip=\belowdisplayskip\par\noindent
+   \ignorespacesafterend}
+
+\newcommand{\framedhslinecorrect}[2]%
+  {#1[#2]}
+
+\newcommand{\framedhs}{\sethscode{framedhscode}}
+
+% The inlinehscode environment is an experimental environment
+% that can be used to typeset displayed code inline.
+
+\newenvironment{inlinehscode}%
+  {\(\def\column##1##2{}%
+   \let\>\undefined\let\<\undefined\let\\\undefined
+   \newcommand\>[1][]{}\newcommand\<[1][]{}\newcommand\\[1][]{}%
+   \def\fromto##1##2##3{##3}%
+   \def\nextline{}}{\) }%
+
+\newcommand{\inlinehs}{\sethscode{inlinehscode}}
+
+% The joincode environment is a separate environment that
+% can be used to surround and thereby connect multiple code
+% blocks.
+
+\newenvironment{joincode}%
+  {\let\orighscode=\hscode
+   \let\origendhscode=\endhscode
+   \def\endhscode{\def\hscode{\endgroup\def\@currenvir{hscode}\\}\begingroup}
+   %\let\SaveRestoreHook=\empty
+   %\let\ColumnHook=\empty
+   %\let\resethooks=\empty
+   \orighscode\def\hscode{\endgroup\def\@currenvir{hscode}}}%
+  {\origendhscode
+   \global\let\hscode=\orighscode
+   \global\let\endhscode=\origendhscode}%
+
+\makeatother
+\EndFmtInput
+%
+
+  % Without these, lhs2tex uses three-line equal signs.
+
 \arrayhs%i.e. no page breaks in code blocks
@@
 OCaml~\cite{letouzey02}. Modules may also be seen as parametrised sections. They allow management
-of namespaces, notations, hints, etc. An |Import| command imports all the constants from a
-module. Thus allowing the user to type |B| instead of |A.B|. Similar functionality would be useful
+of namespaces, notations, hints, etc. An \ensuremath{\Conid{Import}} command imports all the constants from a
+module. Thus allowing the user to type \ensuremath{\Conid{B}} instead of \ensuremath{\Conid{A.B}}. Similar functionality would be useful
 for type classes.}
@@
 relation~\cite{Bishop67,Hofmann,Capretta}. Recently, improved support has been added to the
-system~\cite{Setoid-rewrite}. Functions respecting these equivalence relations are called |Proper|.
+system~\cite{Setoid-rewrite}. Functions respecting these equivalence relations are called \ensuremath{\Conid{Proper}}.
 As we will see in Section~\ref{univ2} below this functionality can also be used to work with
@@
 might define:
-%format → = "\rightarrow"
-%format Π = "\Pi"
-\begin{code}
-  Record ReflexiveRelation: Type :=
-    { rr_carrier: Type
-    ; rr_rel: rr_carrier → rr_carrier → Prop
-    ; rr_proof: Π x, rr_rel x x }.
-\end{code}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{20}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[3]{}\mathbf{Record}\;\Conid{ReflexiveRelation:}\;\Conid{Type}\;\Conid{:=}\;{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\hspace*{0mm}\{\Varid{rr\char95 carrier:}\;\Conid{Type}{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{};\Varid{rr\char95 rel:}\;\Varid{rr\char95 carrier}\;\rightarrow\;\Varid{rr\char95 carrier}\;\rightarrow\;\Conid{Prop}{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{};\Varid{rr\char95 proof:}\;\Pi\;{}\<[20]%
+\>[20]{}\Varid{x},\Varid{rr\char95 rel}\;\Varid{x}\;\Varid{x}\mskip1.5mu\}\;.{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
 and similar for transitive and symmetric relations.
 
-With this approach, we run into problems when we try to define |EquivalenceRelation| in terms of
-|ReflexiveRelation| and its two siblings, because when posit three objects of type
-|ReflexiveRelation|, |TransitiveRelation|, and |SymmetricRelation|, respectively, we have to
+With this approach, we run into problems when we try to define \ensuremath{\Conid{EquivalenceRelation}} in terms of
+\ensuremath{\Conid{ReflexiveRelation}} and its two siblings, because when posit three objects of type
+\ensuremath{\Conid{ReflexiveRelation}}, \ensuremath{\Conid{TransitiveRelation}}, and \ensuremath{\Conid{SymmetricRelation}}, respectively, we have to
 add equalities expressing that they talk about the same carrier and relation, and these equalities
@@
 
-As a first step we would define |ReflexiveRelation| as:
-\begin{code}
-  Record ReflexiveRelation (carrier: Type): Type :=
-    { rr_rel: rr_carrier → rr_carrier → Prop
-    ; rr_proof: Π x, rr_rel x x }.
-\end{code}
-
-This would solve the problem for the carrier, but |EquivalenceRelation| would still need equalities
+As a first step we would define \ensuremath{\Conid{ReflexiveRelation}} as:
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{20}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[3]{}\mathbf{Record}\;\Conid{ReflexiveRelation}\;(\Varid{carrier:}\;\Conid{Type})\;\mathbin{:}\;\Conid{Type}\;\Conid{:=}\;{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\hspace*{0mm}\{\Varid{rr\char95 rel:}\;\Varid{rr\char95 carrier}\;\rightarrow\;\Varid{rr\char95 carrier}\;\rightarrow\;\Conid{Prop}{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{};\Varid{rr\char95 proof:}\;\Pi\;{}\<[20]%
+\>[20]{}\Varid{x},\Varid{rr\char95 rel}\;\Varid{x}\;\Varid{x}\mskip1.5mu\}\;.{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+
+This would solve the problem for the carrier, but \ensuremath{\Conid{EquivalenceRelation}} would still need equalities
 for the relation:
-\begin{code}
-  Record EquivalenceRelation: Type :=
-    { er_carrier: Type
-    ; er_refl: ReflexiveRelation er_carrier
-    ; er_sym: SymmetricRelation er_carrier
-    ; er_trans: TransitiveRelation er_trans
-    ; same_rel: rr_rel er_refl = sr_rel er_sym
-    ; same_rel': rr_rel er_refl = tr_rel er_trans }.
-\end{code}
-
-A natural solution is to pull the relation out of |ReflexiveRelation| (and its siblings) as well:
-\begin{code}
-  Record ReflexiveRelation (carrier: Type)
-                           (rel: carrier → carrier → Prop): Prop :=
-    { rr_proof: Π x, rr_rel x x }.
-\end{code}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[3]{}\mathbf{Record}\;\Conid{EquivalenceRelation:}\;\Conid{Type}\;\Conid{:=}\;{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\hspace*{0mm}\{\Varid{er\char95 carrier:}\;\Conid{Type}{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{};\Varid{er\char95 refl:}\;\Conid{ReflexiveRelation}\;\Varid{er\char95 carrier}{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{};\Varid{er\char95 sym:}\;\Conid{SymmetricRelation}\;\Varid{er\char95 carrier}{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{};\Varid{er\char95 trans:}\;\Conid{TransitiveRelation}\;\Varid{er\char95 trans}{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{};\Varid{same\char95 rel:}\;\Varid{rr\char95 rel}\;\Varid{er\char95 refl}\;\mathrel{=}\;\Varid{sr\char95 rel}\;\Varid{er\char95 sym}{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{};\Varid{same\char95 rel':}\;\Varid{rr\char95 rel}\;\Varid{er\char95 refl}\;\mathrel{=}\;\Varid{tr\char95 rel}\;\Varid{er\char95 trans}\mskip1.5mu\}\;.{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+
+A natural solution is to pull the relation out of \ensuremath{\Conid{ReflexiveRelation}} (and its siblings) as well:
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{20}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{28}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[3]{}\mathbf{Record}\;\Conid{ReflexiveRelation}\;(\Varid{carrier:}\;\Conid{Type})\;{}\<[E]%
+\\
+\>[3]{}\hsindent{25}{}\<[28]%
+\>[28]{}(\Varid{rel:}\;\Varid{carrier}\;\rightarrow\;\Varid{carrier}\;\rightarrow\;\Conid{Prop})\;\mathbin{:}\;\Conid{Prop}\;\Conid{:=}\;{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\hspace*{0mm}\{\Varid{rr\char95 proof:}\;\Pi\;{}\<[20]%
+\>[20]{}\Varid{x},\Varid{rr\char95 rel}\;\Varid{x}\;\Varid{x}\mskip1.5mu\}\;.{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
 or equivalently:
-\begin{code}
-  Definition ReflexiveRelation (carrier: Type)
-                               (rel: carrier → carrier → Prop): Prop :=
-    Π x, rr_rel x x.
-\end{code}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{8}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{32}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[3]{}\mathbf{Definition}\;\Conid{ReflexiveRelation}\;(\Varid{carrier:}\;\Conid{Type})\;{}\<[E]%
+\\
+\>[3]{}\hsindent{29}{}\<[32]%
+\>[32]{}(\Varid{rel:}\;\Varid{carrier}\;\rightarrow\;\Varid{carrier}\;\rightarrow\;\Conid{Prop})\;\mathbin{:}\;\Conid{Prop}\;\Conid{:=}\;{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Pi\;{}\<[8]%
+\>[8]{}\Varid{x},\Varid{rr\char95 rel}\;\Varid{x}\;\Varid{x.}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
 This is how reflexivity is defined in the standard library, and this is entirely sensible. The
-change from |Type| to |Prop| reflects the fact that |ReflexiveRelation| is now strictly a property,
+change from \ensuremath{\Conid{Type}} to \ensuremath{\Conid{Prop}} reflects the fact that \ensuremath{\Conid{ReflexiveRelation}} is now strictly a property,
 proofs of which can be nicely kept opaque;%
@@
 we will return to this issue in Section~\ref{explicit}.
-We can now define |EquivalenceRelation| without awkward equalities:
-\begin{code}
-  Record EquivalenceRelation: Type :=
-    { er_carrier: Type
-    ; er_rel: er_carrier → er_carrier → Prop
-    ; er_refl: ReflexiveRelation er_carrier er_rel
-    ; er_sym: SymmetricRelation er_carrier er_rel
-    ; er_trans: TransitiveRelation er_trans er_rel }.
-\end{code}
-
-But looking at |EquivalenceRelation|, we see that it now essentially has the same form as the
-original naive |ReflexiveRelation| record we started with. Clearly, equivalence relations
+We can now define \ensuremath{\Conid{EquivalenceRelation}} without awkward equalities:
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[3]{}\mathbf{Record}\;\Conid{EquivalenceRelation:}\;\Conid{Type}\;\Conid{:=}\;{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\hspace*{0mm}\{\Varid{er\char95 carrier:}\;\Conid{Type}{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{};\Varid{er\char95 rel:}\;\Varid{er\char95 carrier}\;\rightarrow\;\Varid{er\char95 carrier}\;\rightarrow\;\Conid{Prop}{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{};\Varid{er\char95 refl:}\;\Conid{ReflexiveRelation}\;\Varid{er\char95 carrier}\;\Varid{er\char95 rel}{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{};\Varid{er\char95 sym:}\;\Conid{SymmetricRelation}\;\Varid{er\char95 carrier}\;\Varid{er\char95 rel}{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{};\Varid{er\char95 trans:}\;\Conid{TransitiveRelation}\;\Varid{er\char95 trans}\;\Varid{er\char95 rel}\mskip1.5mu\}\;.{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+
+But looking at \ensuremath{\Conid{EquivalenceRelation}}, we see that it now essentially has the same form as the
+original naive \ensuremath{\Conid{ReflexiveRelation}} record we started with. Clearly, equivalence relations
 and reflexive relations are analogous concepts, and by the same argument we
-conclude that |EquivalenceRelation| should really be defined as:
-
-\begin{code}
-Record EquivalenceRelation (er_carrier: Type)
-                             (er_rel: er_carrier → er_carrier → Prop): Prop :=
-  { er_refl: ReflexiveRelation er_carrier er_rel
-  ; er_sym: SymmetricRelation er_carrier er_rel
-  ; er_trans: TransitiveRelation er_trans er_rel }.
-\end{code}
+conclude that \ensuremath{\Conid{EquivalenceRelation}} should really be defined as:
+
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{30}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathbf{Record}\;\Conid{EquivalenceRelation}\;(\Varid{er\char95 carrier:}\;\Conid{Type})\;{}\<[E]%
+\\
+\>[B]{}\hsindent{30}{}\<[30]%
+\>[30]{}(\Varid{er\char95 rel:}\;\Varid{er\char95 carrier}\;\rightarrow\;\Varid{er\char95 carrier}\;\rightarrow\;\Conid{Prop})\;\mathbin{:}\;\Conid{Prop}\;\Conid{:=}\;{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\hspace*{0mm}\{\Varid{er\char95 refl:}\;\Conid{ReflexiveRelation}\;\Varid{er\char95 carrier}\;\Varid{er\char95 rel}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{};\Varid{er\char95 sym:}\;\Conid{SymmetricRelation}\;\Varid{er\char95 carrier}\;\Varid{er\char95 rel}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{};\Varid{er\char95 trans:}\;\Conid{TransitiveRelation}\;\Varid{er\char95 trans}\;\Varid{er\char95 rel}\mskip1.5mu\}\;.{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
 
@@
 such as the quotient of a setoid,
-we have the same considerations as for |ReflexiveRelation| above.
+we have the same considerations as for \ensuremath{\Conid{ReflexiveRelation}} above.
 
 At the risk of overstressing the point, let us look at how the same problem and solution keep
-appearing in composite structures. Suppose we use a definition of |Monoid| that bundles the
-relation. Then if we proceed to define |SemiRing| as a double |Monoid|, we face a dilemma:
-\begin{code}
-  Record SemiRing A: Type :=
-    { mult_monoid: Monoid A 
-    ; plus_monoid: Monoid A 
-    ; plus_comm: Commutative (mon_op plus_monoid)
-    ; mult_comm: Commutative (mon_op mult_monoid)
-    ; mult_0_l: Π x: A, ?
-\end{code}
-Which equality do we use in the expression ?, |mult_monoid|'s equality or |plus_monoid|'s?
+appearing in composite structures. Suppose we use a definition of \ensuremath{\Conid{Monoid}} that bundles the
+relation. Then if we proceed to define \ensuremath{\Conid{SemiRing}} as a double \ensuremath{\Conid{Monoid}}, we face a dilemma:
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{20}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[3]{}\mathbf{Record}\;\Conid{SemiRing}\;\Conid{A:}\;\Conid{Type}\;\Conid{:=}\;{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\hspace*{0mm}\{\Varid{mult\char95 monoid:}\;\Conid{Monoid}\;\Conid{A}{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{};\Varid{plus\char95 monoid:}\;\Conid{Monoid}\;\Conid{A}{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{};\Varid{plus\char95 comm:}\;\Conid{Commutative}\;(\Varid{mon\char95 op}\;\Varid{plus\char95 monoid}){}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{};\Varid{mult\char95 comm:}\;\Conid{Commutative}\;(\Varid{mon\char95 op}\;\Varid{mult\char95 monoid}){}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{};\Varid{mult\char95 0\char95 l:}\;\Pi\;{}\<[20]%
+\>[20]{}\Varid{x:}\;\Conid{A},\Varid{?}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+Which equality do we use in the expression ?, \ensuremath{\Varid{mult\char95 monoid}}'s equality or \ensuremath{\Varid{plus\char95 monoid}}'s?
 Again, for this definition to be usable at all, we would need to add an equality expressing that the
@@
 We \emph{could} become over-eager and decide to unbundle the equivalence proof term as well:
-\begin{code}
-Record Monoid (A: Type) (R: A → A → Prop)
-  (e: Equivalence R) : Type := ...
-\end{code}
-However, this way if we happen to have two proofs |p| and |q| that a certain relation is an
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathbf{Record}\;\Conid{Monoid}\;(\Conid{A:}\;\Conid{Type})\;(\Conid{R:}\;\Conid{A}\;\rightarrow\;\Conid{A}\;\rightarrow\;\Conid{Prop})\;{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}(\Varid{e:}\;\Conid{Equivalence}\;\mathbb{R})\;\mathbin{:}\;\Conid{Type}\;\Conid{:=}\;\Varid{...}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+However, this way if we happen to have two proofs \ensuremath{\Varid{p}} and \ensuremath{\Varid{q}} that a certain relation is an
 equivalence, we end up with two non-convertible ways to express that the relation (together with
@@
 interchangable, and hope for automation, such as type class instance resolution, to use them
-interchangably), or take great care to ensure that |Equivalence| proof terms are always shared,
+interchangably), or take great care to ensure that \ensuremath{\Conid{Equivalence}} proof terms are always shared,
 essentially elevating them to `values', which goes against the principle of proof
@@
 
-Neither of these options are very appealing. We really should just make |Monoid| take as parameters
+Neither of these options are very appealing. We really should just make \ensuremath{\Conid{Monoid}} take as parameters
 only the operations/relations. If we treat the monoid's unit and binary operation the same way, this
-means that |Monoid| ends up in |Prop|, and that its instances can be left opaque, which is really
+means that \ensuremath{\Conid{Monoid}} ends up in \ensuremath{\Conid{Prop}}, and that its instances can be left opaque, which is really
 how things ought to be.
@@
 
-Suppose we have defined |SemiGroup| in line with the aforementioned principles:
-\begin{code}
-Record SemiGroup (A: Type) (eq: A → A → Prop)
- (op: A → A → A): Prop := ...
-\end{code}
+Suppose we have defined \ensuremath{\Conid{SemiGroup}} in line with the aforementioned principles:
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[3]{}\mathbf{Record}\;\Conid{SemiGroup}\;(\Conid{A:}\;\Conid{Type})\;(\Varid{eq:}\;\Conid{A}\;\rightarrow\;\Conid{A}\;\rightarrow\;\Conid{Prop})\;(\Varid{op:}\;\Conid{A}\;\rightarrow\;\Conid{A}\;\rightarrow\;\Conid{A})\;\mathbin{:}\;\Conid{Prop}\;\Conid{:=}\;\Varid{...}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
 Then our starting point is the obvious but awful:
-\begin{code}
-Lemma assoc1 A eq op (P: SemiGroup A eq op) x y z:
-                   eq (op (op x y) z) (op (op z y) x).
-\end{code}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{20}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[3]{}\mathbf{Lemma}\;\Varid{assoc1}\;\Conid{A}\;\Varid{eq}\;\Varid{op}\;(\Conid{P:}\;\Conid{SemiGroup}\;\Conid{A}\;\Varid{eq}\;\Varid{op})\;\Varid{x}\;\Varid{y}\;\Varid{z:}\;{}\<[E]%
+\\
+\>[3]{}\hsindent{17}{}\<[20]%
+\>[20]{}\Varid{eq}\;(\Varid{op}\;(\Varid{op}\;\Varid{x}\;\Varid{y})\;\Varid{z})\;(\Varid{op}\;(\Varid{op}\;\Varid{z}\;\Varid{y})\;\Varid{x})\;.{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
 As a first cleaning measure, we might hope to introduce some notations to let us use infix notation
-for the semigroup operator (say, |&|) and for equality. Unfortunately, at this point the names |eq|
-and |op| are just local names and there is no way to predeclare that they are to be associated with
+for the semigroup operator (say, \ensuremath{\Varid{\&}}) and for equality. Unfortunately, at this point the names \ensuremath{\Varid{eq}}
+and \ensuremath{\Varid{op}} are just local names and there is no way to predeclare that they are to be associated with
 any notation. What we really need are canonical names for these notions, so that we may fashion
@@
 
-This is a typical job for type classes, and they are used in this way in the |Clean| standard
+This is a typical job for type classes, and they are used in this way in the \ensuremath{\Conid{Clean}} standard
 library~\cite{Clean}. Similarly, Haskell uses type classes to overload notations.
 We simply follow suit:
-\begin{code}
-  Class Equiv A := equiv: relation A.
-  Class SemiGroupOp A := sg_op: A → A → A.
-
-  Infix "=" := equiv: type_scope.
-  Infix "&" := sg_op (at level 50, left associativity).
-\end{code}
-We choose to use |&| here and reserve the notation |*| for ring multiplication.
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[3]{}\mathbf{Class}\;\Conid{Equiv}\;\Conid{A}\;\Conid{:=}\;\Varid{equiv:}\;\Varid{relation}\;\Conid{A.}{}\<[E]%
+\\
+\>[3]{}\mathbf{Class}\;\Conid{SemiGroupOp}\;\Conid{A}\;\Conid{:=}\;\Varid{sg\char95 op:}\;\Conid{A}\;\rightarrow\;\Conid{A}\;\rightarrow\;\Conid{A.}{}\<[E]%
+\\[\blanklineskip]%
+\>[3]{}\mathbf{Infix}\;\text{\tt \char34 =\char34}\;\Conid{:=}\;\Varid{equiv:}\;\Varid{type\char95 scope.}{}\<[E]%
+\\
+\>[3]{}\mathbf{Infix}\;\text{\tt \char34 \&\char34}\;\Conid{:=}\;\Varid{sg\char95 op}\;(\Varid{at}\;\Varid{level}\;\Varid{50},\Varid{left}\;\Varid{associativity})\;.{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+We choose to use \ensuremath{\Varid{\&}} here and reserve the notation \ensuremath{\Varid{*}} for ring multiplication.
 
@@
 We would have like to use a haskell style definition
-\begin{code}
-Class Equiv A := (=): relation A.
-\end{code}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathbf{Class}\;\Conid{Equiv}\;\Conid{A}\;\Conid{:=}\;(\mathrel{=})\;\mathbin{:}\;\Varid{relation}\;\Conid{A.}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
 Unfortunately, currently this is not supported in Coq.
@@
 records the way classes with any other number of fields are, but are instead turned into ordinary
-definitions. The effect of this is that |Equiv A| immediately reduces to |relation A|, and that the
+definitions. The effect of this is that \ensuremath{\Conid{Equiv}\;\Conid{A}} immediately reduces to \ensuremath{\Varid{relation}\;\Conid{A}}, and that the
 equiv `projection' is the identity function whose argument is inferred using instance resolution.
@@
 it, notations:
-\begin{code}
-  Record SemiGroup (A: Type) (eq: Equiv A)
-                             (op: SemiGroupOp A): Prop := ...
-  Lemma assoc1 A eq op (P: SemiGroup A eq op) x y z:
-                    x & y & z = z & y & x.
-\end{code}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{21}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{30}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[3]{}\mathbf{Record}\;\Conid{SemiGroup}\;(\Conid{A:}\;\Conid{Type})\;(\Varid{eq:}\;\Conid{Equiv}\;\Conid{A})\;{}\<[E]%
+\\
+\>[3]{}\hsindent{27}{}\<[30]%
+\>[30]{}(\Varid{op:}\;\Conid{SemiGroupOp}\;\Conid{A})\;\mathbin{:}\;\Conid{Prop}\;\Conid{:=}\;\Varid{...}{}\<[E]%
+\\
+\>[3]{}\mathbf{Lemma}\;\Varid{assoc1}\;\Conid{A}\;\Varid{eq}\;\Varid{op}\;(\Conid{P:}\;\Conid{SemiGroup}\;\Conid{A}\;\Varid{eq}\;\Varid{op})\;\Varid{x}\;\Varid{y}\;\Varid{z:}\;{}\<[E]%
+\\
+\>[3]{}\hsindent{18}{}\<[21]%
+\>[21]{}\Varid{x}\;\Varid{\&}\;\Varid{y}\;\Varid{\&}\;\Varid{z}\;\mathrel{=}\;\Varid{z}\;\Varid{\&}\;\Varid{y}\;\Varid{\&}\;\Varid{x.}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
 We call these type classes \emph{operational type classes}. We note that operational type classes
-allow us to avoid Coq's notation |scope| mechanism.
+allow us to avoid Coq's notation \ensuremath{\mathbf{scope}} mechanism.
 
@@
 On the other hand, it \emph{does} mean that all the carriers/relations/operations end up in the
-context. The context for |assoc1| looks like:
-\begin{code}
-  A: Type
-  eq: Equiv A
-  op: SemiGroupOp A
-  P: SemiGroup A eq op
-\end{code}
+context. The context for \ensuremath{\Varid{assoc1}} looks like:
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[3]{}\Conid{A:}\;\Conid{Type}{}\<[E]%
+\\
+\>[3]{}\Varid{eq:}\;\Conid{Equiv}\;\Conid{A}{}\<[E]%
+\\
+\>[3]{}\Varid{op:}\;\Conid{SemiGroupOp}\;\Conid{A}{}\<[E]%
+\\
+\>[3]{}\Conid{P:}\;\Conid{SemiGroup}\;\Conid{A}\;\Varid{eq}\;\Varid{op}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
 
@@
 likely not be problematic for any but the most complex formalizations), having to \emph{declare}
-these entities (like |eq| and |op| for |assoc1|) \emph{is} a bit of a chore. Fortunately, Coq
+these entities (like \ensuremath{\Varid{eq}} and \ensuremath{\Varid{op}} for \ensuremath{\Varid{assoc1}}) \emph{is} a bit of a chore. Fortunately, Coq
 provides a feature called implicit generalization, which is exactly what we need.
-\marginpar{describe how it works} With it, we can write |assoc1| as follows:
-\begin{code}
-  Lemma assoc1 `(SemiGroup A) x y z: x & y & z = z & y & x.
-\end{code}
+\marginpar{describe how it works} With it, we can write \ensuremath{\Varid{assoc1}} as follows:
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[3]{}\mathbf{Lemma}\;\Varid{assoc1}\hspace*{0mm}`(\Conid{SemiGroup}\;\Conid{A})\;\Varid{x}\;\Varid{y}\;\Varid{z:}\;\Varid{x}\;\Varid{\&}\;\Varid{y}\;\Varid{\&}\;\Varid{z}\;\mathrel{=}\;\Varid{z}\;\Varid{\&}\;\Varid{y}\;\Varid{\&}\;\Varid{x.}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
 Thus, we have reached the mathematical ideal we aimed for.
@@
 A record type is \marginpar{mostly?} syntactic sugar for a $\Sigma$-type. In the presence of a
-squash operator, |[]:Type->Prop|, we may define a subset type as |\{A\mid B\}=Sig A. [B]|. This is
+squash operator, \ensuremath{[\mskip1.5mu \mskip1.5mu]\;\Conid{:Type->Prop}}, we may define a subset type as \ensuremath{\lambda \;\hspace*{0mm}\{\Conid{A\char92 mid}\;\Conid{B\char92 }\mskip1.5mu\}\;\Varid{=Sig}\;\Conid{A.}\;[\mskip1.5mu \Conid{B}\mskip1.5mu]}.
+This is
 pursued by Barras~\cite{Barras:subset,Werner}. The program machinery fakes subset types in the
@@
 \epsilon$ for $x:Prop$? We could add a term constructor $[a]$ which will have type $[A]$. There
-seems to be no reason to connect this to |Prop| or impredicativity.
+seems to be no reason to connect this to \ensuremath{\Conid{Prop}} or impredicativity.
 
@@
  \item Working hypothesis: `every record can be reordered as a subset'. Consequently, we
-rewrite this subset |\{A\mid B\}| as |Record r (a:A):={b:B}|.
+rewrite this subset \ensuremath{\lambda \;\hspace*{0mm}\{\Conid{A\char92 mid}\;\Conid{B\char92 }\mskip1.5mu\}} as \ensuremath{\mathbf{Record}\;\Varid{r}\;(\Varid{a:A})\;\Conid{:=}\;\hspace*{0mm}\{\Varid{b:B}\mskip1.5mu\}}.
 \end{itemize}
@@
 pack the proofs. This can be nicely organized by using canonical names. For instance, we define:
-\begin{code}
-Class SemiRing A {e: Equiv A}{plus: RingPlus A}{mult: RingMult A}
-                 {zero: RingZero A}{one: RingOne A}:Prop :=
-  { semiring_mult_monoid:> Monoid A (op:=mult)(unit:=one)
-  ; semiring_plus_monoid:> Monoid A (op:=plus)(unit:=zero)
-  ; semiring_plus_comm:> Commutative plus
-  ; semiring_mult_comm:> Commutative mult
-  ; semiring_distr:> Distribute mult plus
-  ; mult_0_l: forall x, 0 * x == 0 }.
-\end{code}
-where e.g.\ |RingPlus| is the operation class for the semigroup operator.
-The notation |:>| makes |semiring_mult_monoid| a coercion: when asked for a monoid it suffices to
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{18}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathbf{Class}\;\Conid{SemiRing}\;\Conid{A}\;\hspace*{0mm}\{\Varid{e:}\;\Conid{Equiv}\;\Conid{A}\mskip1.5mu\}\;\hspace*{0mm}\{\Varid{plus:}\;\Conid{RingPlus}\;\Conid{A}\mskip1.5mu\}\;\hspace*{0mm}\{\Varid{mult:}\;\Conid{RingMult}\;\Conid{A}\mskip1.5mu\}\;{}\<[E]%
+\\
+\>[B]{}\hsindent{18}{}\<[18]%
+\>[18]{}\hspace*{0mm}\{\Varid{zero:}\;\Conid{RingZero}\;\Conid{A}\mskip1.5mu\}\;\hspace*{0mm}\{\Varid{one:}\;\Conid{RingOne}\;\Conid{A}\mskip1.5mu\}\;\Conid{:Prop}\;\Conid{:=}\;{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\hspace*{0mm}\{\Varid{semiring\char95 mult\char95 monoid:>}\;\Conid{Monoid}\;\Conid{A}\;(\Varid{op:=mult})\;(\Varid{unit:=one}){}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{};\Varid{semiring\char95 plus\char95 monoid:>}\;\Conid{Monoid}\;\Conid{A}\;(\Varid{op:=plus})\;(\Varid{unit:=zero}){}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{};\Varid{semiring\char95 plus\char95 comm:>}\;\Conid{Commutative}\;\Varid{plus}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{};\Varid{semiring\char95 mult\char95 comm:>}\;\Conid{Commutative}\;\Varid{mult}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{};\Varid{semiring\char95 distr:>}\;\Conid{Distribute}\;\Varid{mult}\;\Varid{plus}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{};\Varid{mult\char95 0\char95 l:}\;\forall\;\Varid{x},\Varid{0}\;\Varid{*}\;\Varid{x}\;= \;\Varid{0}\mskip1.5mu\}\;.{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+where e.g.\ \ensuremath{\Conid{RingPlus}} is the operation class for the semigroup operator.
+The notation \ensuremath{\Conid{:>}} makes \ensuremath{\Varid{semiring\char95 mult\char95 monoid}} a coercion: when asked for a monoid it suffices to
 provide a ring.
@@
 % it.
-We note that |op| does not depend on the proof that |e| is an equivalence. We use Coq's implicit
-quantification (|`{}|) to avoid having to write all the parameters when \emph{stating} a theorem
+We note that \ensuremath{\Varid{op}} does not depend on the proof that \ensuremath{\Varid{e}} is an equivalence. We use Coq's implicit
+quantification (\ensuremath{\hspace*{0mm}`\hspace*{0mm}\{\mskip1.5mu\}}) to avoid having to write all the parameters when \emph{stating} a theorem
 and Coq's maximally inserted implicit arguments to find the parameters when \emph{applying} a
@@
 from the dependent record to its argument, thus mimicking manifest fields by using type classes.
-Since |op| is \emph{instantiated} with |mult|, we need not worry about the
-equality |op=mult|.
+Since \ensuremath{\Varid{op}} is \emph{instantiated} with \ensuremath{\Varid{mult}}, we need not worry about the
+equality \ensuremath{\Varid{op=mult}}.
 
@@
 notation for comp} category~\cite{saibi1995constructive}
-\begin{code}
- Class Category O `{!Arrows O} `{forall x y: O, Equiv (x --> y)}
-   `{!CatId O} `{!CatComp O}: Prop :=
-  { arrow_equiv:> forall x y: O, Setoid (x --> y)
-  ; comp_proper:> forall x y z: O,
-      Proper (equiv ==> equiv ==> equiv)%signature (@comp _ _ _ x y z)
-  ; comp_assoc: forall (w x y z: O) (a: w --> x) (b: x --> y) (c: y --> z),
-      comp c (comp b a) == comp (comp c b) a
-  ; id_l: forall (x y: O) (a: y --> x), comp cat_id a == a
-  ; id_r: forall (x y: O) (a: x --> y), comp a cat_id == a }.
-\end{code}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{4}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{7}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathbf{Class}\;\Conid{Category}\;\Conid{O}\hspace*{0mm}`\hspace*{0mm}\{\Varid{!Arrows}\;\Conid{O}\mskip1.5mu\}\hspace*{0mm}`\hspace*{0mm}\{\forall\;\Varid{x}\;\Varid{y:}\;\Conid{O},\Conid{Equiv}\;(\Varid{x}\;\longrightarrow\;\Varid{y})\mskip1.5mu\}{}\<[E]%
+\\
+\>[B]{}\hsindent{4}{}\<[4]%
+\>[4]{}\hspace*{0mm}`\hspace*{0mm}\{\Varid{!CatId}\;\Conid{O}\mskip1.5mu\}\hspace*{0mm}`\hspace*{0mm}\{\Varid{!CatComp}\;\Conid{O}\mskip1.5mu\}\;\mathbin{:}\;\Conid{Prop}\;\Conid{:=}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\hspace*{0mm}\{\Varid{arrow\char95 equiv:>}\;\forall\;\Varid{x}\;\Varid{y:}\;\Conid{O},\Conid{Setoid}\;(\Varid{x}\;\longrightarrow\;\Varid{y}){}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{};\Varid{comp\char95 proper:>}\;\forall\;\Varid{x}\;\Varid{y}\;\Varid{z:}\;\Conid{O},{}\<[E]%
+\\
+\>[3]{}\hsindent{4}{}\<[7]%
+\>[7]{}\Conid{Proper}\;(\Varid{equiv}\;\Rightarrow \;\Varid{equiv}\;\Rightarrow \;\Varid{equiv})\;\Varid{\%signature}\;(\Varid{@comp}\;\anonymous \;\anonymous \;\anonymous \;\Varid{x}\;\Varid{y}\;\Varid{z}){}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{};\Varid{comp\char95 assoc:}\;\forall\;(\Varid{w}\;\Varid{x}\;\Varid{y}\;\Varid{z:}\;\Conid{O})\;(\Varid{a:}\;\Varid{w}\;\longrightarrow\;\Varid{x})\;(\Varid{b:}\;\Varid{x}\;\longrightarrow\;\Varid{y})\;(\Varid{c:}\;\Varid{y}\;\longrightarrow\;\Varid{z}),{}\<[E]%
+\\
+\>[3]{}\hsindent{4}{}\<[7]%
+\>[7]{}\Varid{comp}\;\Varid{c}\;(\Varid{comp}\;\Varid{b}\;\Varid{a})\;= \;\Varid{comp}\;(\Varid{comp}\;\Varid{c}\;\Varid{b})\;\Varid{a}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{};\Varid{id\char95 l:}\;\forall\;(\Varid{x}\;\Varid{y:}\;\Conid{O})\;(\Varid{a:}\;\Varid{y}\;\longrightarrow\;\Varid{x}),\Varid{comp}\;\Varid{cat\char95 id}\;\Varid{a}\;= \;\Varid{a}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{};\Varid{id\char95 r:}\;\forall\;(\Varid{x}\;\Varid{y:}\;\Conid{O})\;(\Varid{a:}\;\Varid{x}\;\longrightarrow\;\Varid{y}),\Varid{comp}\;\Varid{a}\;\Varid{cat\char95 id}\;= \;\Varid{a}\mskip1.5mu\}\;.{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
 This is based on the 2-categorical idea that we have equality on arrows, but not on objects.
@@
 
-%format ◎ = "\circ"
-For instance natural transformations are formalized as: |◎|
-\begin{code}
-  Context `{Category C} `{Category D} (F G: C -> D)
-     `{!Functor F Fa} `{!Functor G Ga}.
-
-  Class NaturalTransformation (η: forall c: C, F c --> G c): Prop :=
-    natural: forall (x y: C) (f: x --> y), η y ◎ fmap F f == fmap G f ◎ η x.
-\end{code}
-
-The notation |◎| denotes function composition.
-Here |fmap F| explicitly refers to the arrow part of the functor |F|.\\
+For instance natural transformations are formalized as: \ensuremath{\circ}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{6}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[3]{}\Conid{Context}\hspace*{0mm}`\hspace*{0mm}\{\Conid{Category}\;\Conid{C}\mskip1.5mu\}\hspace*{0mm}`\hspace*{0mm}\{\Conid{Category}\;\Conid{D}\mskip1.5mu\}\;(\Conid{F}\;\Conid{G:}\;\Conid{C}\;\to \;\Conid{D}){}\<[E]%
+\\
+\>[3]{}\hsindent{3}{}\<[6]%
+\>[6]{}\hspace*{0mm}`\hspace*{0mm}\{\Varid{!Functor}\;\Conid{F}\;\Conid{Fa}\mskip1.5mu\}\hspace*{0mm}`\hspace*{0mm}\{\Varid{!Functor}\;\Conid{G}\;\Conid{Ga}\mskip1.5mu\}\;.{}\<[E]%
+\\[\blanklineskip]%
+\>[3]{}\mathbf{Class}\;\Conid{NaturalTransformation}\;(\Conid{η:}\;\forall\;\Varid{c:}\;\Conid{C},\Conid{F}\;\Varid{c}\;\longrightarrow\;\Conid{G}\;\Varid{c})\;\mathbin{:}\;\Conid{Prop}\;\Conid{:=}\;{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Varid{natural:}\;\forall\;(\Varid{x}\;\Varid{y:}\;\Conid{C})\;(\Varid{f:}\;\Varid{x}\;\longrightarrow\;\Varid{y}),\Conid{η}\;\Varid{y}\;\circ\;\Varid{fmap}\;\Conid{F}\;\Varid{f}\;= \;\Varid{fmap}\;\Conid{G}\;\Varid{f}\;\circ\;\Conid{η}\;\Varid{x.}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+
+The notation \ensuremath{\circ} denotes function composition.
+Here \ensuremath{\Varid{fmap}\;\Conid{F}} explicitly refers to the arrow part of the functor \ensuremath{\Conid{F}}.\\
 In mathematics one would apply $F$ to both objects and arrows ($F x$, $F f$). However, this does
@@
 %theory/ua_congruence.v
-\begin{code}
-Class Congruence: Prop :=
-    { congruence_proper:> Π s, Proper (equiv ==> equiv ==> iff) (e s)
-    ; congruence_quotient:> @Algebra et v e _
-    }.
-\end{code}
-We note the use of |Proper|.
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{30}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathbf{Class}\;\Conid{Congruence:}\;\Conid{Prop}\;\Conid{:=}\;{}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}\hspace*{0mm}\{\Varid{congruence\char95 proper:>}\;\Pi\;{}\<[30]%
+\>[30]{}\Varid{s},\Conid{Proper}\;(\Varid{equiv}\;\Rightarrow \;\Varid{equiv}\;\Rightarrow \;\Varid{iff})\;(\Varid{e}\;\Varid{s}){}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{};\Varid{congruence\char95 quotient:>}\;\Varid{@Algebra}\;\Varid{et}\;\Varid{v}\;\Varid{e}\;\anonymous {}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}\mskip1.5mu\}\;.{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+We note the use of \ensuremath{\Conid{Proper}}.
 
@@
 elements of the algebra and of the subalgebra, i.e.\ terms of different types may be equal.
-\begin{code}
-  Fixpoint heq {o}: op_type carrier o -> op_type v o -> Prop :=
-    match o with
-    | constant x => fun a b => `a == b
-    | function x y => fun a b => forall u, heq (a u) (b (`u))
-    end.
-\end{code}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[3]{}\mathbf{Fixpoint}\;\Varid{heq}\;\hspace*{0mm}\{\Varid{o}\mskip1.5mu\}\;\mathbin{:}\;\Varid{op\char95 type}\;\Varid{carrier}\;\Varid{o}\;\to \;\Varid{op\char95 type}\;\Varid{v}\;\Varid{o}\;\to \;\Conid{Prop}\;\Conid{:=}\;{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\mathbf{match}\;\Varid{o}\;\mathbf{with}{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\mid \;\Varid{constant}\;\Varid{x}\;\Rightarrow \;\lambda\hspace{-1mm} \;\Varid{a}\;\Varid{b}\;\Rightarrow \hspace*{0mm}`\Varid{a}\;= \;\Varid{b}{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\mid \;\Varid{function}\;\Varid{x}\;\Varid{y}\;\Rightarrow \;\lambda\hspace{-1mm} \;\Varid{a}\;\Varid{b}\;\Rightarrow \;\forall\;\Varid{u},\Varid{heq}\;(\Varid{a}\;\Varid{u})\;(\Varid{b}\;(\hspace*{0mm}`\Varid{u})){}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Varid{end.}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
 
@@
 construct a monoid.
-\begin{code}
- Variable o: Variety theory.
- Global Instance from_object: Monoid (o tt).
-\end{code}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathbf{Variable}\;\Varid{o:}\;\Conid{Variety}\;\Varid{theory.}{}\<[E]%
+\\
+\>[B]{}\mathbf{Global}\;\mathbf{Instance}\;\Varid{from\char95 object:}\;\Conid{Monoid}\;(\Varid{o}\;\Varid{tt})\;.{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
 Similarly, for arrows.
@@
 semiring structure, is assigned priority 10.
-\begin{code}
-  Global Program Instance: NatDistance N | 10 := ...
-\end{code}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[3]{}\mathbf{Global}\;\mathbf{Program}\;\Conid{Instance:}\;\Conid{NatDistance}\;\Conid{N}\;\mid \;\Varid{10}\;\Conid{:=}\;\Varid{...}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
 
@@
 We define a term language.
-\begin{code}
-Inductive Expr (V: Type) := Mult (a b: Expr V) | Zero | Var (v: V).
-\end{code}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathbf{Inductive}\;\Conid{Expr}\;(\Conid{V:}\;\Conid{Type})\;\Conid{:=}\;\Conid{Mult}\;(\Varid{a}\;\Varid{b:}\;\Conid{Expr}\;\Conid{V})\;\mid \;\Conid{Zero}\;\mid \;\Conid{Var}\;(\Varid{v:}\;\Conid{V})\;.{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
 This defines a monad on Type\marginpar{Do this!}.
 The expression type is parameterized over the set of variable indices. Hence, we diverge
- from~\cite{Hints}, which uses |nat| for variables thereby introducing bounds problems and
+ from~\cite{Hints}, which uses \ensuremath{\mathbb{N}} for variables thereby introducing bounds problems and
  dummy variables.
@@
 an Expr-algebra. eval is map vars composed with the algebra map.}
-\begin{code}
-Definition Value := nat.
-Definition Vars V := V -> Value.
-
-Fixpoint eval {V} (vs: Vars V) (e: Expr V): Value :=
-  match e with
-  | Zero => 0
-  | Mult a b => eval vs a * eval vs b
-  | Var v => vs v
-  end.
-\end{code}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathbf{Definition}\;\Conid{Value}\;\Conid{:=}\;\Varid{nat.}{}\<[E]%
+\\
+\>[B]{}\mathbf{Definition}\;\Conid{Vars}\;\Conid{V}\;\Conid{:=}\;\Conid{V}\;\to \;\Conid{Value.}{}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\mathbf{Fixpoint}\;\Varid{eval}\;\hspace*{0mm}\{\Conid{V}\mskip1.5mu\}\;(\Varid{vs:}\;\Conid{Vars}\;\Conid{V})\;(\Varid{e:}\;\Conid{Expr}\;\Conid{V})\;\mathbin{:}\;\Conid{Value}\;\Conid{:=}\;{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\mathbf{match}\;\Varid{e}\;\mathbf{with}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\mid \;\Conid{Zero}\;\Rightarrow \;\Varid{0}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\mid \;\Conid{Mult}\;\Varid{a}\;\Varid{b}\;\Rightarrow \;\Varid{eval}\;\Varid{vs}\;\Varid{a}\;\Varid{*}\;\Varid{eval}\;\Varid{vs}\;\Varid{b}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\mid \;\Conid{Var}\;\Varid{v}\;\Rightarrow \;\Varid{vs}\;\Varid{v}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{end.}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
 %
 Monads are trees with grafting. As explained by sigfpe~\cite{MonadsGrafting}, the tree monad
-$TX:=1+T^2X$, may be seen as the prototypical monad. Indeed, the |Expr| monad, i.e.\ the free
+$TX:=1+T^2X$, may be seen as the prototypical monad. Indeed, the \ensuremath{\Conid{Expr}} monad, i.e.\ the free
 $\Sigma$-algebra monad, behaves much like the tree monad. We now provide a direct definition of the
@@
 
-\begin{code}
-Fixpoint bind {V W} (f: V -> Expr W) (e: Expr V): Expr W :=
-  match e with
-  | Zero => Zero
-  | Mult a b => Mult (bind f a) (bind f b)
-  | Var v => vs v
-  end.
-\end{code}
-
-We have shown that |Expr|, i.e.\ |bind| together with |Var|, is a monad on the category of Types
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathbf{Fixpoint}\;\Varid{bind}\;\hspace*{0mm}\{\Conid{V}\;\Conid{W}\mskip1.5mu\}\;(\Varid{f:}\;\Conid{V}\;\to \;\Conid{Expr}\;\Conid{W})\;(\Varid{e:}\;\Conid{Expr}\;\Conid{V})\;\mathbin{:}\;\Conid{Expr}\;\Conid{W}\;\Conid{:=}\;{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\mathbf{match}\;\Varid{e}\;\mathbf{with}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\mid \;\Conid{Zero}\;\Rightarrow \;\Conid{Zero}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\mid \;\Conid{Mult}\;\Varid{a}\;\Varid{b}\;\Rightarrow \;\Conid{Mult}\;(\Varid{bind}\;\Varid{f}\;\Varid{a})\;(\Varid{bind}\;\Varid{f}\;\Varid{b}){}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\mid \;\Conid{Var}\;\Varid{v}\;\Rightarrow \;\Varid{vs}\;\Varid{v}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{end.}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+
+We have shown that \ensuremath{\Conid{Expr}}, i.e.\ \ensuremath{\Varid{bind}} together with \ensuremath{\Conid{Var}}, is a monad on the category of Types
 with extensional functions between them~\marginpar{Really?}.
@@
 %
-\begin{code}
-Definition novars: Vars False := False_rect _.
-Definition singlevar (x: Value): Vars unit := fun _ => x.
-Definition merge {A B} (a: Vars A) (b: Vars B): Vars (A+B) :=
-  fun i => if i then a j else b j.
-\end{code}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathbf{Definition}\;\Varid{novars:}\;\Conid{Vars}\;\Conid{False}\;\Conid{:=}\;\Conid{False\char95 rect}\;\Varid{\char95 .}{}\<[E]%
+\\
+\>[B]{}\mathbf{Definition}\;\Varid{singlevar}\;(\Varid{x:}\;\Conid{Value})\;\mathbin{:}\;\Conid{Vars}\;\Varid{unit}\;\Conid{:=}\;\lambda\hspace{-1mm} \;\anonymous \;\Rightarrow \;\Varid{x.}{}\<[E]%
+\\
+\>[B]{}\mathbf{Definition}\;\Varid{merge}\;\hspace*{0mm}\{\Conid{A}\;\Conid{B}\mskip1.5mu\}\;(\Varid{a:}\;\Conid{Vars}\;\Conid{A})\;(\Varid{b:}\;\Conid{Vars}\;\Conid{B})\;\mathbin{:}\;\Conid{Vars}\;(\Conid{A+B})\;\Conid{:=}\;{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\lambda\hspace{-1mm} \;\Varid{i}\;\Rightarrow \;\Varid{if}\;\Varid{i}\;\Varid{then}\;\Varid{a}\;\Varid{j}\;\Varid{else}\;\Varid{b}\;\Varid{j.}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
 
@@
  Gallina terms, representing heaps, for which we implement syntactic lookup with type classes.
-Given a heap and value, |Lookup| instances give the value's index in the heap:\marginpar{A
+Given a heap and value, \ensuremath{\Conid{Lookup}} instances give the value's index in the heap:\marginpar{A
 canonical structures approach would allow us to do this in Coq?}
-\begin{code}
-  Class Lookup {A} (x: Value) (f: Vars A) :=
-    { lookup: A; lookup_correct: f lookup = x }.
-\end{code}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[3]{}\mathbf{Class}\;\Conid{Lookup}\;\hspace*{0mm}\{\Conid{A}\mskip1.5mu\}\;(\Varid{x:}\;\Conid{Value})\;(\Varid{f:}\;\Conid{Vars}\;\Conid{A})\;\Conid{:=}\;{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\hspace*{0mm}\{\Varid{lookup:}\;\Conid{A};\Varid{lookup\char95 correct:}\;\Varid{f}\;\Varid{lookup}\;\mathrel{=}\;\Varid{x}\mskip1.5mu\}\;.{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
 % Context (x: Value) {A B} (va: Vars A) (vb: Vars B).
@@
 access it by indexing the merged heap (and vice versa)
-\begin{code}
-  Global Instance lookup_left `{!Lookup x va}: Lookup x (merge va vb)
-    := { lookup := inl (lookup x va) }.
-\end{code}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[3]{}\mathbf{Global}\;\mathbf{Instance}\;\Varid{lookup\char95 left}\hspace*{0mm}`\hspace*{0mm}\{\Varid{!Lookup}\;\Varid{x}\;\Varid{va}\mskip1.5mu\}\;\mathbin{:}\;\Conid{Lookup}\;\Varid{x}\;(\Varid{merge}\;\Varid{va}\;\Varid{vb}){}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Conid{:=}\;\hspace*{0mm}\{\Varid{lookup}\;\Conid{:=}\;\Varid{inl}\;(\Varid{lookup}\;\Varid{x}\;\Varid{va})\mskip1.5mu\}\;.{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
 If the heap is just a singlevar, we can easily index it.
-\begin{code}
-  Global Program Instance: Lookup x (singlevar x) := { lookup := tt }.
-\end{code}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[3]{}\mathbf{Global}\;\mathbf{Program}\;\Conid{Instance:}\;\Conid{Lookup}\;\Varid{x}\;(\Varid{singlevar}\;\Varid{x})\;\Conid{:=}\;\hspace*{0mm}\{\Varid{lookup}\;\Conid{:=}\;\Varid{tt}\mskip1.5mu\}\;.{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
 
 One useful operation we need before we get to Quote relates to variables and expression
- evaluation. As its name suggests, |map_var| maps an expression's variable indices.\marginpar{monad}
-
-\begin{code}
-Definition map_var {V W: Type} (f: V -> W):
-    Expr V -> Expr W :=
-  fix F (e: Expr V): Expr W :=
-    match e with
-    | Mult a b => Mult (F a) (F b)
-    | Zero => Zero
-    | Var v => Var (f v)
-    end.
-\end{code}
-In |Quote| below, the idea is that |V, l|, and |n| are all input variables, while |V'| and |r| are
-output variables (in the sense that we will rely on unification to generate them). |V| and |l|
-represent the current heap, |n| represents the value we want to quote, and |V'| and |r'| represent
+ evaluation. As its name suggests, \ensuremath{\Varid{map\char95 var}} maps an expression's variable indices.\marginpar{monad}
+
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathbf{Definition}\;\Varid{map\char95 var}\;\hspace*{0mm}\{\Conid{V}\;\Conid{W:}\;\Conid{Type}\mskip1.5mu\}\;(\Varid{f:}\;\Conid{V}\;\to \;\Conid{W})\;\mathbin{:}\;{}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}\Conid{Expr}\;\Conid{V}\;\to \;\Conid{Expr}\;\Conid{W}\;\Conid{:=}\;{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{fix}\;\Conid{F}\;(\Varid{e:}\;\Conid{Expr}\;\Conid{V})\;\mathbin{:}\;\Conid{Expr}\;\Conid{W}\;\Conid{:=}\;{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\mathbf{match}\;\Varid{e}\;\mathbf{with}{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\mid \;\Conid{Mult}\;\Varid{a}\;\Varid{b}\;\Rightarrow \;\Conid{Mult}\;(\Conid{F}\;\Varid{a})\;(\Conid{F}\;\Varid{b}){}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\mid \;\Conid{Zero}\;\Rightarrow \;\Conid{Zero}{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\mid \;\Conid{Var}\;\Varid{v}\;\Rightarrow \;\Conid{Var}\;(\Varid{f}\;\Varid{v}){}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Varid{end.}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+In \ensuremath{\Conid{Quote}} below, the idea is that \ensuremath{\Conid{V},\Varid{l}}, and \ensuremath{\Varid{n}} are all input variables, while \ensuremath{\Conid{V'}} and \ensuremath{\Varid{r}} are
+output variables (in the sense that we will rely on unification to generate them). \ensuremath{\Conid{V}} and \ensuremath{\Varid{l}}
+represent the current heap, \ensuremath{\Varid{n}} represents the value we want to quote, and \ensuremath{\Conid{V'}} and \ensuremath{\Varid{r'}} represent
 the heap of newly encountered variables during the quotation.
   This explains the type of quote: it is an expression that refers either to variables from
-the old heap, or to newly encountered variables. Finally, |eval_quote| is the usual correctness
+the old heap, or to newly encountered variables. Finally, \ensuremath{\Varid{eval\char95 quote}} is the usual correctness
 property, which now merges the two heaps.
 
-\begin{code}
-  Class Quote {V} (l: Vars V) (n: Value) {V'} (r: Vars V') :=
-    { quote: Expr (V + V')
-    ; eval_quote: @eval (V+V') (merge l r) quote = n }.
-\end{code}
-
-Our first instance for |Zero| is easy. The `novars' in the result type reflects the fact that no
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[3]{}\mathbf{Class}\;\Conid{Quote}\;\hspace*{0mm}\{\Conid{V}\mskip1.5mu\}\;(\Varid{l:}\;\Conid{Vars}\;\Conid{V})\;(\Varid{n:}\;\Conid{Value})\;\hspace*{0mm}\{\Conid{V'}\mskip1.5mu\}\;(\Varid{r:}\;\Conid{Vars}\;\Conid{V'})\;\Conid{:=}\;{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\hspace*{0mm}\{\Varid{quote:}\;\Conid{Expr}\;(\Conid{V}\;\Varid{+}\;\Conid{V'}){}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{};\Varid{eval\char95 quote:}\;\Varid{@eval}\;(\Conid{V+V'})\;(\Varid{merge}\;\Varid{l}\;\Varid{r})\;\Varid{quote}\;\mathrel{=}\;\Varid{n}\mskip1.5mu\}\;.{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+
+Our first instance for \ensuremath{\Conid{Zero}} is easy. The `novars' in the result type reflects the fact that no
 new variables are encountered.
-\begin{code}
-  Global Program Instance quote_zero V (v: Vars V):
-    Quote v 0 novars :=
-  { quote := Zero }.
-\end{code}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[3]{}\mathbf{Global}\;\mathbf{Program}\;\mathbf{Instance}\;\Varid{quote\char95 zero}\;\Conid{V}\;(\Varid{v:}\;\Conid{Vars}\;\Conid{V})\;\mathbin{:}\;{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Conid{Quote}\;\Varid{v}\;\Varid{0}\;\Varid{novars}\;\Conid{:=}{}\<[E]%
+\\
+\>[3]{}\hspace*{0mm}\{\Varid{quote}\;\Conid{:=}\;\Conid{Zero}\mskip1.5mu\}\;.{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
 
@@
 shuffles heaps around.
- The third line contains some |map_var|'s because the heap shuffling must be
-reflected in the variable indices, but apart from that it's just constructing a |Mult| term with
+ The third line contains some \ensuremath{\Varid{map\char95 var}}'s because the heap shuffling must be
+reflected in the variable indices, but apart from that it's just constructing a \ensuremath{\Conid{Mult}} term with
 quoted subterms.
 
-\begin{code}
-Global Program Instance quote_mult V (v: Vars V)
-  n V' (v': Vars V') m V'' (v'': Vars V'')
-  `{!Quote v n v'} `{!Quote (merge v v') m v''}:
-  Quote v (n * m) (merge v' v'') :=
-  { quote := Mult (map_var bla (quote n)) (map_var sum_assoc (quote m)) }.
-\end{code}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathbf{Global}\;\mathbf{Program}\;\mathbf{Instance}\;\Varid{quote\char95 mult}\;\Conid{V}\;(\Varid{v:}\;\Conid{Vars}\;\Conid{V})\;{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{n}\;\Conid{V'}\;(\Varid{v':}\;\Conid{Vars}\;\Conid{V'})\;\Varid{m}\;\Conid{V''}\;(\Varid{v'':}\;\Conid{Vars}\;\Conid{V''}){}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\hspace*{0mm}`\hspace*{0mm}\{\Varid{!Quote}\;\Varid{v}\;\Varid{n}\;\Varid{v'}\mskip1.5mu\}\hspace*{0mm}`\hspace*{0mm}\{\Varid{!Quote}\;(\Varid{merge}\;\Varid{v}\;\Varid{v'})\;\Varid{m}\;\Varid{v''}\mskip1.5mu\}\;\mathbin{:}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Conid{Quote}\;\Varid{v}\;(\Varid{n}\;\Varid{*}\;\Varid{m})\;(\Varid{merge}\;\Varid{v'}\;\Varid{v''})\;\Conid{:=}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\hspace*{0mm}\{\Varid{quote}\;\Conid{:=}\;\Conid{Mult}\;(\Varid{map\char95 var}\;\Varid{bla}\;(\Varid{quote}\;\Varid{n}))\;(\Varid{map\char95 var}\;\Varid{sum\char95 assoc}\;(\Varid{quote}\;\Varid{m}))\mskip1.5mu\}\;.{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
   Now follows the instance where we recognize values that are already in the heap. This
    is expressed by the Lookup requirement, which will only be fulfilled if the Lookup instances
-   defined above can find the value in the heap. The novars in the |Quote v x novars| result
+   defined above can find the value in the heap. The novars in the \ensuremath{\Conid{Quote}\;\Varid{v}\;\Varid{x}\;\Varid{novars}} result
    reflects that this quotation does not generate new variables.
 
-\begin{code}
-Global Program Instance quote_old_var
-  V (v: Vars V) x {i: Lookup x v}:
-  Quote v x novars | 8 := { quote := Var (inl (lookup x v)) }.
-\end{code}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathbf{Global}\;\mathbf{Program}\;\mathbf{Instance}\;\Varid{quote\char95 old\char95 var}\;{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Conid{V}\;(\Varid{v:}\;\Conid{Vars}\;\Conid{V})\;\Varid{x}\;\hspace*{0mm}\{\Varid{i:}\;\Conid{Lookup}\;\Varid{x}\;\Varid{v}\mskip1.5mu\}\;\mathbin{:}\;{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Conid{Quote}\;\Varid{v}\;\Varid{x}\;\Varid{novars}\;\mid \;\Varid{8}\;\Conid{:=}\;\hspace*{0mm}\{\Varid{quote}\;\Conid{:=}\;\Conid{Var}\;(\Varid{inl}\;(\Varid{lookup}\;\Varid{x}\;\Varid{v}))\mskip1.5mu\}\;.{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
 
 \noindent Finally, the instance for new variables. We give this a lower priority so that it is only
-used if Lookup fails. The |8| in the previous example is a random number less than |9| below. A
+used if Lookup fails. The \ensuremath{\Varid{8}} in the previous example is a random number less than \ensuremath{\Varid{9}} below. A
 similar method is used in Coq's notation mechanism. We strongly recommend a change in
@@
 
-\begin{code}
-Global Program Instance quote_new_var V (v: Vars V) x:
-  Quote v x (singlevar x) | 9 := { quote := Var (inr tt) }.
-\end{code}
-
-\noindent As announced this provides a light-weight quoting function\footnote{|eval_quote'| is a
-modification of |eval_quote| which works in the empty context.}:
-\begin{code}
-Goal forall x y (P: Value -> Prop), P ((x * y) * (x * 0)).
-  intros. rewrite <- (eval_quote' _).
-\end{code}
-This turns the goal into | (eval variable_pack quote)|.
-
-It would be interesting to use this technique to implement the |congruence| tactic, which implements
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathbf{Global}\;\mathbf{Program}\;\mathbf{Instance}\;\Varid{quote\char95 new\char95 var}\;\Conid{V}\;(\Varid{v:}\;\Conid{Vars}\;\Conid{V})\;\Varid{x:}\;{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Conid{Quote}\;\Varid{v}\;\Varid{x}\;(\Varid{singlevar}\;\Varid{x})\;\mid \;\Varid{9}\;\Conid{:=}\;\hspace*{0mm}\{\Varid{quote}\;\Conid{:=}\;\Conid{Var}\;(\Varid{inr}\;\Varid{tt})\mskip1.5mu\}\;.{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+
+\noindent As announced this provides a light-weight quoting function\footnote{\ensuremath{\Varid{eval\char95 quote'}} is a
+modification of \ensuremath{\Varid{eval\char95 quote}} which works in the empty context.}:
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Conid{Goal}\;\forall\;\Varid{x}\;\Varid{y}\;(\Conid{P:}\;\Conid{Value}\;\to \;\Conid{Prop}),\Conid{P}\;((\Varid{x}\;\Varid{*}\;\Varid{y})\;\Varid{*}\;(\Varid{x}\;\Varid{*}\;\Varid{0}))\;.{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{intros.}\;\Varid{rewrite}\;\leftarrow \;(\Varid{eval\char95 quote'}\;\anonymous )\;.{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+This turns the goal into \ensuremath{(\Varid{eval}\;\Varid{variable\char95 pack}\;\Varid{quote})}.
+
+It would be interesting to use this technique to implement the \ensuremath{\Varid{congruence}} tactic, which implements
 the congruence closure algorithm~\cite{corbineau2007deciding}. One would naturally obtain a tactic
@@
 structures to derive the setoid equality on the natural
-numbers in the following example |(S x)==(S y)|. We have conveniently reimplemented this with
+numbers in the following example \ensuremath{(\Conid{S}\;\Varid{x})\;= \;(\Conid{S}\;\Varid{y})}. We have conveniently reimplemented this with
 type classes with superior proof terms. Canonical structures give
-\begin{code}
-@equiv(Build_Setoid nat (@eq nat)(@eq_equivalence nat))(S x)(S y)
-\end{code}
-
-which includes an explicit proof that |(@eq nat)| is an equivalence,
-whereas we obtain |@equiv nat (@eq nat) (S x) (S y)|.
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{@equiv}\;(\Conid{Build\char95 Setoid}\;\mathbb{N}\;(\Varid{@eq}\;\mathbb{N})\;(\Varid{@eq\char95 equivalence}\;\mathbb{N}))\;(\Conid{S}\;\Varid{x})\;(\Conid{S}\;\Varid{y}){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+
+which includes an explicit proof that \ensuremath{(\Varid{@eq}\;\mathbb{N})} is an equivalence,
+whereas we obtain \ensuremath{\Varid{@equiv}\;\mathbb{N}\;(\Varid{@eq}\;\mathbb{N})\;(\Conid{S}\;\Varid{x})\;(\Conid{S}\;\Varid{y})}.
 
@@
 We use type classes to deduce the relevant monoid operation and write:
-\begin{code}
-Definition seq_sum
-  `{Sequence A T} `{RingPlus A} `{z: RingZero A}: T -> A
-  := @seq_to_monoid A T _ A ring_plus z id.
-  Eval compute in seq_sum [3; 2].
-\end{code}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathbf{Definition}\;\Varid{seq\char95 sum}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\hspace*{0mm}`\hspace*{0mm}\{\Conid{Sequence}\;\Conid{A}\;\Conid{T}\mskip1.5mu\}\hspace*{0mm}`\hspace*{0mm}\{\Conid{RingPlus}\;\Conid{A}\mskip1.5mu\}\hspace*{0mm}`\hspace*{0mm}\{\Varid{z:}\;\Conid{RingZero}\;\Conid{A}\mskip1.5mu\}\;\mathbin{:}\;\Conid{T}\;\to \;\Conid{A}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Conid{:=}\;\Varid{@seq\char95 to\char95 monoid}\;\Conid{A}\;\Conid{T}\;\anonymous \;\Conid{A}\;\Varid{ring\char95 plus}\;\Varid{z}\;\Varid{id.}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\mathbf{Eval}\;\Varid{compute}\;\mathbf{in}\;\Varid{seq\char95 sum}\;[\mskip1.5mu \Varid{3};\Varid{2}\mskip1.5mu]\;.{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
 
@@
 constructions~\cite{miquel2001implicit,barras2008implicit}. Such a change would fit naturally with
-the Hoare-style separation between programs and proofs which is used by the |Program| machinery.
+the Hoare-style separation between programs and proofs which is used by the \ensuremath{\mathbf{Program}} machinery.
 
 On the other hand, we found instances where we needed proof \emph{relevance}. In this case, it lead
-to a duplication of code which may have been avoided by universe polymorphism for |Fixpoint|s. In
+to a duplication of code which may have been avoided by universe polymorphism for \ensuremath{\mathbf{Fixpoint}}s. In
 practice, manual dead-code analysis seems useful. Making Harrop formulas proof irrelevant seems to
@@
 
-According to |coqwc|, our development consists of 3103 lines of specifications and 707 lines of
+According to \ensuremath{\Varid{coqwc}}, our development consists of 3103 lines of specifications and 707 lines of
 proofs.\marginpar{update,targz}

### 2010-02-25 roughdiamond.tex
first commit

--- 2010-02-25 jar.tex
@@
-\documentclass[a4paper,10pt, runningheads]{llncs}
+\documentclass[a4paper,10pt]{llncs}
 \usepackage{url}
 \usepackage{bbm}
-\usepackage{ucs}
-% \usepackage[utf8x,utf8]{inputenc}
+\usepackage[utf8x]{inputenc}
 \newcommand{\N}{\ensuremath{\mathbbm{N}}}
@@
 %include coq2tex.tex
-\arrayhs%i.e. no page breaks in code blocks
+\arrayhs %i.e. no page breaks in code blocks
 \divide\abovedisplayskip 4
@@
 \begin{document}
-\title{The algebraic hierarchy in type theory using type classes}
+\title{Developing the algebraic hierarchy with type classes in Coq}
 \author{Bas Spitters \and Eelis van der Weegen}
@@
 structure inference. A library for many-sorted universal algebra abstracts from the specific
-algebraic structures and aids the organization of the library. Category theory provides a further
-useful abstraction. We provide abstract
+algebraic structures and aids the organization of the library. Finally, we provide abstract
 interfaces for $\N$, $\Z$, $\Q$ as the initial semiring, the initial ring and a field of
-fractions. Similarly, we provide an interface for polynomials over a ring $R$ using $R$-algebras.
-We show that the standard implementations of these are in fact instances of these type classes.
-We report on the efficiency penality of these abstractions.
-
-Finally, we use type classes to conveniently implement a quote function using the prolog-like
-abilities of type class unification. We integrate this with our development of universal algebra.
+fractions.
 \end{abstract}
@@
 \section{Introduction}
-Developing formalized libraries for mathematics requires a detailed analysis of the
-architecture of mathematics~\cite{C-corn,DBLP:conf/types/HaftmannW08}, similar to that of a large
-software project.
+Developing formalized libraries for mathematics requires a detailed analysis of the architecture~\cite{C-corn,DBLP:conf/types/HaftmannW08}, similar to that of a large software project. 
 The development of an algebraic hierarchy is an important aspect. Challenges include: 1. multiple
 inheritance: A ring is a type with two operations. How do we inherit from the two theories of
-monoids? 2. Sharing of notations. 3. Combining syntactic and semantic methods, usually this is
-addressed by reflection.
+monoids? 2. Sharing of notations.
 
-Several solutions have been proposed for the Coq theorem prover~\cite{Coq,BC04}: dependent
+Several solutions have been proposed for the Coq theorem prover%~\cite{Coq,BC04}
+: dependent
 records~\cite{DBLP:journals/jsc/GeuversPWZ02} (a.k.a.\ telescopes), packed classes~\cite{Packed} and
 occasionally modules. Although the latter option is quickly discarded%
-\footnote{Modules are not first class, but currently they seem to be superior for extraction to
-OCaml~\cite{letouzey02}. Modules may also be seen as parametrised sections. They allow management
-of namespaces, notations, hints, etc. An |Import| command imports all the constants from a
-module. Thus allowing the user to type |B| instead of |A.B|. Similar functionality would be useful
-for type classes.}
+\footnote{Modules are not first class, but currently they seem to be superior for extraction
+purposes%~\cite{letouzey02}%
+.}, considering
 algebraic theories as `interfaces', as opposed to `modules', has been an important
@@
 
-In this fashion we implement\footnote{The sources are available
-at:~\url{http://www.eelis.net/research/math-classes/}}
+In this fashion we implement\footnote{The sources are available at:~\url{http://www.eelis.net/research/math-classes/}}%
  algebraic theories as type classes, not instances, thus addressing the
-challenges above; see Section~\ref{classes}. Type classes have recently entered
-Coq~\cite{DBLP:conf/tphol/SozeauO08}. In Coq,
+challenges above; see Section~\ref{classes}. Type classes have recently entered Coq~\cite{DBLP:conf/tphol/SozeauO08}. In Coq
 type classes are \emph{first class} citizens based on dependent records. The presence of dependent types makes them even more
@@
 
-We abstract from concrete algebra and formalize many-sorted universal algebra, including a
-construction of
-free $\Sigma$-algebras; see Section~\ref{univ}. We are also concerned with efficiency.
-In fact, our aim is to use the present development as a new basis for Bishop's program in type
-theory~\cite{typesreal-article}.\marginpar{More}
+% In
+% Coq, type classes have advantages over modules\footnote{
+% Modules are sections on steroids: namespace management, notations, hints, etc.
+% The [Import] command imports all the constants from a module. Thus allowing the user to type B
+% instead of A.B
+% }:
+% they are first class and extract~\cite{letouzey02} to Haskell\footnote{
+% Would module like extraction be possible?}.
 
-So, we would like to replace data structures with more efficient ones. Concretely,
-Coq provides unary, binary and machine numbers. We use type classes polymorphism to
-develop a unified library for them; see Section~\ref{modul}.
-The interfaces for $\N$, $\Z,\Q$ are the initial semiring, the initial ring and a field of
-fractions; see Section~\ref{interfaces}. In future work we aim to base our development of the reals
-on an abstract dense set, allowing us to use the efficient dyadic
-rationals~\cite{boldo2009combining} as a base for exact real number computation in
-Coq~\cite{Riemann,Oconnor:real}.
+We abstract from concrete algebra and formalize many-sorted universal algebra, including a construction of
+free $\Sigma$-algebras; Section~\ref{univ}. Finally, we are concerned with efficiency. So, we would like to replace data structures with more efficient ones. For instance, 
+Coq provides unary, binary and machine numbers. Hence, we use ad-hoc polymorphism with type classes to
+develop a unified library for them; see Section~\ref{modul}. The interfaces for $\N$, $\Z,\Q$ are the initial semiring, the initial ring and a field of fractions; see Section~\ref{interfaces}.
 
-The paper is organised as follows: Section~\ref{classes} starts with a development of the concrete
-algebraic hierarchy addressing the issues raised above. Section~\ref{univ} abstracts to universal
-algebra, we develop the basic theory to the point of the first homomorphism theorem.
-Section~\ref{interfaces} provides interfaces to usual datastructures by providing their universal
-properties. To allow symbolic manipulation of semantic objects one needs a quote function. Usually,
-this function is implemented in Coq's tactics language. A more convenient way is to use type class
-unification; see section~\ref{quote}. In section~\ref{canonical}, we end with a short comparision
-between type classes and canonical structures.
+%Our contributions are:\footnote{The sources are available at:~\url{http://www.eelis.net/research/math-classes/}}%
+%\begin{itemize}\itemsep=0pt
+% \item The implementation of an algebraic hierarchy based on type classes, addressing the
+%issues raised above; see Section~\ref{classes}.
+% \item A formalization of many-sorted universal algebra in Coq, including a construction of
+%free $\Sigma$-algebras; Section~\ref{univ}.
+%\end{itemize}
 
-In this paper we focus on the Coq proof assistant. We conjecture that the methods can be transfered
-to any type theory based proof assistant supporting type classes such as
-Matita~\cite{asperti2007user}.
-
-\section{Preliminaries}
-\subsection{Coq type theory}
-The Coq proof assistant is based on the calculus of inductive
-constructions~\cite{CoquandHuet,CoquandPaulin}.\marginpar{More}
-This type theory lacks quotient types, as it would make type checking undecidable. Instead it
-supports setoids (aka Bishop sets), a pair of a type and an equivalence
-relation~\cite{Bishop67,Hofmann,Capretta}. Recently, improved support has been added to the
-system~\cite{Setoid-rewrite}. Functions respecting these equivalence relations are called |Proper|.
-As we will see in Section~\ref{univ2} below this functionality can also be used to work with
-quotient algebras.
-
-\subsection{Type classes}
-Implementation in Ltac, Notation etc
-Coercions
-
-\section{To bunde or not to bundle? (Or: What to bundle?)}
-
-How should we define reflexive, transitive, and symmetric relations? In a first attempt, we
-might define:
-%format → = "\rightarrow"
-%format Π = "\Pi"
-\begin{code}
-  Record ReflexiveRelation: Type :=
-    { rr_carrier: Type
-    ; rr_rel: rr_carrier → rr_carrier → Prop
-    ; rr_proof: Π x, rr_rel x x }.
-\end{code}
-and similar for transitive and symmetric relations.
-
-With this approach, we run into problems when we try to define |EquivalenceRelation| in terms of
-|ReflexiveRelation| and its two siblings, because when posit three objects of type
-|ReflexiveRelation|, |TransitiveRelation|, and |SymmetricRelation|, respectively, we have to
-add equalities expressing that they talk about the same carrier and relation, and these equalities
-are a nightmare to work with (because the three carriers and relations won't be convertible).
-Technically, this leads to the research in manifest fields, the \emph{with} construction, to which
-we return in Section~\ref{manifest}.
-
-As a first step we would define |ReflexiveRelation| as:
-\begin{code}
-  Record ReflexiveRelation (carrier: Type): Type :=
-    { rr_rel: rr_carrier → rr_carrier → Prop
-    ; rr_proof: Π x, rr_rel x x }.
-\end{code}
-
-This would solve the problem for the carrier, but |EquivalenceRelation| would still need equalities
-for the relation:
-\begin{code}
-  Record EquivalenceRelation: Type :=
-    { er_carrier: Type
-    ; er_refl: ReflexiveRelation er_carrier
-    ; er_sym: SymmetricRelation er_carrier
-    ; er_trans: TransitiveRelation er_trans
-    ; same_rel: rr_rel er_refl = sr_rel er_sym
-    ; same_rel': rr_rel er_refl = tr_rel er_trans }.
-\end{code}
-
-A natural solution is to pull the relation out of |ReflexiveRelation| (and its siblings) as well:
-\begin{code}
-  Record ReflexiveRelation (carrier: Type)
-                           (rel: carrier → carrier → Prop): Prop :=
-    { rr_proof: Π x, rr_rel x x }.
-\end{code}
-or equivalently:
-\begin{code}
-  Definition ReflexiveRelation (carrier: Type)
-                               (rel: carrier → carrier → Prop): Prop :=
-    Π x, rr_rel x x.
-\end{code}
-This is how reflexivity is defined in the standard library, and this is entirely sensible. The
-change from |Type| to |Prop| reflects the fact that |ReflexiveRelation| is now strictly a property,
-proofs of which can be nicely kept opaque;%
-\footnote{Coq distinguishes between transparent and opague definitions. The former can be unfolded,
-whereas this is impossible for the latter.}
-we will return to this issue in Section~\ref{explicit}.
-We can now define |EquivalenceRelation| without awkward equalities:
-\begin{code}
-  Record EquivalenceRelation: Type :=
-    { er_carrier: Type
-    ; er_rel: er_carrier → er_carrier → Prop
-    ; er_refl: ReflexiveRelation er_carrier er_rel
-    ; er_sym: SymmetricRelation er_carrier er_rel
-    ; er_trans: TransitiveRelation er_trans er_rel }.
-\end{code}
-
-But looking at |EquivalenceRelation|, we see that it now essentially has the same form as the
-original naive |ReflexiveRelation| record we started with. Clearly, equivalence relations
-and reflexive relations are analogous concepts, and by the same argument we
-conclude that |EquivalenceRelation| should really be defined as:
-
-\begin{code}
-Record EquivalenceRelation (er_carrier: Type)
-                             (er_rel: er_carrier → er_carrier → Prop): Prop :=
-  { er_refl: ReflexiveRelation er_carrier er_rel
-  ; er_sym: SymmetricRelation er_carrier er_rel
-  ; er_trans: TransitiveRelation er_trans er_rel }.
-\end{code}
-
-After all, when we define a type with
-two different equivalence relations on it,
-such as the quotient of a setoid,
-we have the same considerations as for |ReflexiveRelation| above.
-
-At the risk of overstressing the point, let us look at how the same problem and solution keep
-appearing in composite structures. Suppose we use a definition of |Monoid| that bundles the
-relation. Then if we proceed to define |SemiRing| as a double |Monoid|, we face a dilemma:
-\begin{code}
-  Record SemiRing A: Type :=
-    { mult_monoid: Monoid A 
-    ; plus_monoid: Monoid A 
-    ; plus_comm: Commutative (mon_op plus_monoid)
-    ; mult_comm: Commutative (mon_op mult_monoid)
-    ; mult_0_l: Π x: A, ?
-\end{code}
-Which equality do we use in the expression ?, |mult_monoid|'s equality or |plus_monoid|'s?
-Again, for this definition to be usable at all, we would need to add an equality expressing that the
-two monoids contain the same equivalence relation. And again, the proper solution is to change
-Monoid and unbundle the relation. But here, we have a choice as to how much we unbundle exactly.
-(For the moment we ignore the question of whether the unit and binary operator should be unbundled
-as well, focusing only on the equivalence relation and its equivalence proof.)
-
-We \emph{could} become over-eager and decide to unbundle the equivalence proof term as well:
-\begin{code}
-Record Monoid (A: Type) (R: A → A → Prop)
-  (e: Equivalence R) : Type := ...
-\end{code}
-However, this way if we happen to have two proofs |p| and |q| that a certain relation is an
-equivalence, we end up with two non-convertible ways to express that the relation (together with
-some unit and binary operation) form a Monoid: 'Monoid A R p ...' and 'Monoid A R q ...'. We must
-either try to deal with this (by proving that Monoid proofs with different Equivalence proofs are
-interchangable, and hope for automation, such as type class instance resolution, to use them
-interchangably), or take great care to ensure that |Equivalence| proof terms are always shared,
-essentially elevating them to `values', which goes against the principle of proof
-irrelevance.\marginpar{What if we add proof irrelevance?}
-
-Neither of these options are very appealing. We really should just make |Monoid| take as parameters
-only the operations/relations. If we treat the monoid's unit and binary operation the same way, this
-means that |Monoid| ends up in |Prop|, and that its instances can be left opaque, which is really
-how things ought to be.
-
-Another benefit of this approach is that it works very smoothly when these records are in fact type
-classes. When relations and operations are bundled and do \emph{not} appear as parameters of the
-structure record/class, expressing as a type class constraint that some collection of relations and
-operations posess the structure seems to require manifest fields; see Section~\ref{manifest}. With
-our unbundled approach, on the other hand, such type class constraints are easily expressed without
-any extensions as simple type inhabitation queries
-% (e.g. ``hey Coq, find me a proof of [Monoid some_rel some_op some_unit]'')
-of types living in Prop, so that we don't even care what instances are found (thus embracing proof
-irrelevance).
-
-We remark that leaving the equivalence proofs bundled means that when one has a proof that something
-is a semiring, one has two proofs (via the two monoids) that the equivalence relation is an
-equivalence. Fortunately, having redundant proofs like this is entirely harmless when one's terms
-never refer (directly or indirectly) to proofs, which is the case when the relations and operations
-they are composed of are left unbundled. Here, too, we enjoy the warm embrace of proof irrelevance.
-
-From the discussion so far, we derive some principles:
-\begin{enumerate}
- \item structural concepts (like Reflexive, Setoid, Monoid) should have their structural components
-(i.e. carriers, relations, and operations) as parameters, and nothing else;
- \item (as a consequence) these concepts should be defined as predicates (which includes records and
-classes living in Prop), and their proofs may be kept opaque;
- \item (as a second consequence) algebraic expressions need/should never refer (directly or
-indirectly) to proofs.
-\end{enumerate}
-% \setcounter{enumi}{3}
-
-This list will grow in subsequent sections.
-
-The principles listed so far are not particularly controversial, in fact this
-approach ensures maximum flexibility~\cite{Hints}. However,
-there is a common perception that unbundling carriers, relations, and operations this way invariably
-leads either to unmanagably big algebraic expressions (involving endless projections and/or implicit
-arguments), or to unmanageably long and tedious enumeration of carriers/relations/operations, or to
-problems establishing canonical names for operations (since we no longer project them out of
-structure instances all the time).
-
-In the next sections, we show how systematic use of type classes (and their support infrastructure
-in Coq), combined with the use of a special form of type class we call an `operational type class'
-(as opposed to a `structural' type class we might use in place of the records shown thus far), lets
-us avoid these problems.
-
-\section{Managing unbundled structure with type classes}
-
-So far, we have only looked at the definitions of the structures themselves, and have not yet
-considered the perspective of someone trying to prove some theorems about these abstract structures.
-We would like to approximate the common mathematical vernacular, where one
-simply says:
-
-for $x, y, z$ in a semigroup $G$, $x * y * z = z * y * x$, $\ldots$
-
-Suppose we have defined |SemiGroup| in line with the aforementioned principles:
-\begin{code}
-Record SemiGroup (A: Type) (eq: A → A → Prop)
- (op: A → A → A): Prop := ...
-\end{code}
-Then our starting point is the obvious but awful:
-\begin{code}
-Lemma assoc1 A eq op (P: SemiGroup A eq op) x y z:
-                   eq (op (op x y) z) (op (op z y) x).
-\end{code}
-As a first cleaning measure, we might hope to introduce some notations to let us use infix notation
-for the semigroup operator (say, |&|) and for equality. Unfortunately, at this point the names |eq|
-and |op| are just local names and there is no way to predeclare that they are to be associated with
-any notation. What we really need are canonical names for these notions, so that we may fashion
-notations for them.
-
-This is a typical job for type classes, and they are used in this way in the |Clean| standard
-library~\cite{Clean}. Similarly, Haskell uses type classes to overload notations.
-We simply follow suit:
+\section{Algebraic hierarchy using type classes}\label{classes}
+\vspace*{-2ex}
+\paragraph{Canonical names.}\label{names}
+Sharing of notations is an important aspect of the design of a mathematical library. Haskell uses
+type classes to overload notations. In Coq we assign a class to every operation, thus avoiding Coq's notation |scope| mechanism.
+ Each such operation class has only \emph{one} constructor, hence no record is constructed.
 \begin{code}
  Class Equiv A := equiv: relation A.
-  Class SemiGroupOp A := sg_op: A → A → A.
-
-  Infix "=" := equiv: type_scope.
-  Infix "&" := sg_op (at level 50, left associativity).
+ Infix "==" := equiv (at level 70, no associativity).
 \end{code}
-We choose to use |&| here and reserve the notation |*| for ring multiplication.
-
-We use = for setoid equality, since Leibniz equality seems less important for the formalization of
-mathematics. More precisely, setoid equality is the basic concept. In special case, one can prove
-that setoid equality coincides with Leibniz equality.
-
-We would have like to use a haskell style definition
-\begin{code}
-Class Equiv A := (=): relation A.
-\end{code}
-Unfortunately, currently this is not supported in Coq.
-
-These single-field type classes are given special treatment in that they are not translated into
-records the way classes with any other number of fields are, but are instead turned into ordinary
-definitions. The effect of this is that |Equiv A| immediately reduces to |relation A|, and that the
-equiv `projection' is the identity function whose argument is inferred using instance resolution.
-Thus, for a very modest cost in indirection, we gain the ability to use canonical names, and with
-it, notations:
-\begin{code}
-  Record SemiGroup (A: Type) (eq: Equiv A)
-                             (op: SemiGroupOp A): Prop := ...
-  Lemma assoc1 A eq op (P: SemiGroup A eq op) x y z:
-                    x & y & z = z & y & x.
-\end{code}
-We call these type classes \emph{operational type classes}. We note that operational type classes
-allow us to avoid Coq's notation |scope| mechanism.
-
-It is important to point out that although the expression now \emph{looks} like it is bound to some
-specific semigroup structure (which with traditional approaches would imply projections from bundles
-and/or reference to proofs), it is really only referring near-directly to the actual operations
-involved, with the semigroup proof existing only as opaque `knowledge about the operations' which we
-may use in the proof. This lack of projections keeps our terms small and independent, and keeps
-rewriting simple and sane.
-
-On the other hand, it \emph{does} mean that all the carriers/relations/operations end up in the
-context. The context for |assoc1| looks like:
-\begin{code}
-  A: Type
-  eq: Equiv A
-  op: SemiGroupOp A
-  P: SemiGroup A eq op
-\end{code}
-
-While we are not particularly worried about overly large contexts (we feel that this will most
-likely not be problematic for any but the most complex formalizations), having to \emph{declare}
-these entities (like |eq| and |op| for |assoc1|) \emph{is} a bit of a chore. Fortunately, Coq
-provides a feature called implicit generalization, which is exactly what we need.
-\marginpar{describe how it works} With it, we can write |assoc1| as follows:
-\begin{code}
-  Lemma assoc1 `(SemiGroup A) x y z: x & y & z = z & y & x.
-\end{code}
-Thus, we have reached the mathematical ideal we aimed for.
-
-Let us now stop and think to what extent this approach suffers from all the problems commonly
-associated with it. In particular, let us imagine what happens to our terms and contexts when we
-want to talk about nested structures such as polynomials over the ring of matrices over a
-given ring; see Section~\ref{interfaces}.
-
-\section{Subset types}
-BEGIN SPECULATIVE
-
-A record type is \marginpar{mostly?} syntactic sugar for a $\Sigma$-type. In the presence of a
-squash operator, |[]:Type->Prop|, we may define a subset type as |\{A\mid B\}=Sig A. [B]|. This is
-pursued by Barras~\cite{Barras:subset,Werner}. The program machinery fakes subset types in the
-present Coq implementation. Proof irrelevanve in Coq would allow a faithful
-implementation.
-
-Squash is a monad and allows many constructions: subsets, unions, images, heterogeneous equality,
-\ldots.
-
-Question: Would the meta-theory simplify if we would just add a conversion rule $x\triangleright
-\epsilon$ for $x:Prop$? We could add a term constructor $[a]$ which will have type $[A]$. There
-seems to be no reason to connect this to |Prop| or impredicativity.
-
-END SPECULATIVE
-
-We add another principle:
-\begin{itemize}\setcounter{enumi}{3}
- \item Working hypothesis: `every record can be reordered as a subset'. Consequently, we
-rewrite this subset |\{A\mid B\}| as |Record r (a:A):={b:B}|.
-\end{itemize}
-
-\section{Algebraic hierarchy using type classes}\label{classes}
+% Instance: Params (@equiv) 2.
+\vspace*{-5ex}
 \paragraph{Hierarchy.}\label{hierarchy}
@@
 where e.g.\ |RingPlus| is the operation class for the semigroup operator.
-The notation |:>| makes |semiring_mult_monoid| a coercion: when asked for a monoid it suffices to
-provide a ring.
 
@@
 We mention the trade-off between bigger contexts versus bigger terms. Our contexts are bigger than
-those of telescopes or packed classes. In our experience, this has been relatively
-harmless\footnote{Coq's data structure for contexts is not very efficient however, we believe this
-may be a bottleneck for huge developments. It seems that the data structure chosen
-in~\cite{asperti2009compact} will behave better.}%
-: most
+those of telescopes or packed classes. In our experience, this has been relatively harmless: most
 terms in the context are there to support canonical names. Bigger terms
@@
 properties.
-
-\subsection{Manifest fields}\label{manifest}
 
@@
 by implementing many-sorted universal algebra. A formalization of universal algebra
-already exists in Coq~\cite{DBLP:conf/tphol/Capretta99}. It has recently been extended with more
-categorical results~\cite{dominguez2008formalizing}. We have conveniently redeveloped most of this
-using type classes. We defined equational theories and their varieties:
+already exists in Coq~\cite{DBLP:conf/tphol/Capretta99}. We have conveniently redeveloped most of it
+using type classes, moreover we have added a categorical framework. For instance, we show that the
+term-algebra for a signature is initial. We defined equational theories and their varieties:
 the categories of their models. A quote function helps to translate a statement for an instance of
-a type class to a statement in the equational logic.
+a type class to a statement in the equational logic. We are currently pursuing quoting with
+unification~\cite{Hints} using type classes.
 
-\subsection{Category theory}\label{cats}
-Using type classes, we provide a basic library for category theory: categories, functors,
-adjunctions, monads,\ldots We follow the usual type theoretical definition of a\marginpar{Use
-notation for comp} category~\cite{saibi1995constructive}
-\begin{code}
- Class Category O `{!Arrows O} `{forall x y: O, Equiv (x --> y)}
-   `{!CatId O} `{!CatComp O}: Prop :=
-  { arrow_equiv:> forall x y: O, Setoid (x --> y)
-  ; comp_proper:> forall x y z: O,
-      Proper (equiv ==> equiv ==> equiv)%signature (@comp _ _ _ x y z)
-  ; comp_assoc: forall (w x y z: O) (a: w --> x) (b: x --> y) (c: y --> z),
-      comp c (comp b a) == comp (comp c b) a
-  ; id_l: forall (x y: O) (a: y --> x), comp cat_id a == a
-  ; id_r: forall (x y: O) (a: x --> y), comp a cat_id == a }.
-\end{code}
-This is based on the 2-categorical idea that we have equality on arrows, but not on objects.
-Similarly, we have equality on natural transformations, but not on functors.
-
-%format ◎ = "\circ"
-For instance natural transformations are formalized as: |◎|
-\begin{code}
-  Context `{Category C} `{Category D} (F G: C -> D)
-     `{!Functor F Fa} `{!Functor G Ga}.
-
-  Class NaturalTransformation (η: forall c: C, F c --> G c): Prop :=
-    natural: forall (x y: C) (f: x --> y), η y ◎ fmap F f == fmap G f ◎ η x.
-\end{code}
-
-The notation |◎| denotes function composition.
-Here |fmap F| explicitly refers to the arrow part of the functor |F|.\\
-In mathematics one would apply $F$ to both objects and arrows ($F x$, $F f$). However, this does
-not seem to fit with type theory, hence we follow the customs in functional programming.
-
-The proofs are short and follow the textbook easily.\marginpar{really?}
-
-We provide the category of setoids, the category of categories, the dual of a category, products,
-and hence co-products, the category of $\Sigma$-algebras.
-
-We provide the construction of the term-algebra for a signature $\Sigma$ and show that it is
-initial in the category of $\Sigma$-algebras. We proceed to prove that this provides a left-adjoint
-to the forgetful functor. The resulting expression monad will be used below.
-
-\subsection{Universal algebra continued}\label{univ2}
-
-We have developed basic universal algebra~\cite{meinke1993universal} upto the first homomorphism
-theorem.~\marginpar{Description of the work.}
-%perhaps the soundness theorem for equational logic.
-
-In order to prove the first homophism theorem, we define a congruence:
-%theory/ua_congruence.v
-\begin{code}
-Class Congruence: Prop :=
-    { congruence_proper:> Π s, Proper (equiv ==> equiv ==> iff) (e s)
-    ; congruence_quotient:> @Algebra et v e _
-    }.
-\end{code}
-We note the use of |Proper|.
-
-We define the forgetful functor from the category of $\Sigma$-algebras to the category of sets.
-
-We define the term algebra consisting of closed terms, terms with the empty type as set of
-variables. % extensionally equal closed term???
-We show that the term algebra is initial. One may expect that this could be useful to
-\emph{define}, say, the natural numbers as the initial semiring. However, it seems difficult to
-prove decidability of equality, as it seems to require a normalization procedure.
-
-We define the product of two algebras and show that it is the categorical product.
-
-Given a sub\emph{set} of an algebra, we define the sub\emph{algebra} generated by it.
-An interesting fact to mention is the use of the following heterogeneous equality between
-elements of the algebra and of the subalgebra, i.e.\ terms of different types may be equal.
-\begin{code}
-  Fixpoint heq {o}: op_type carrier o -> op_type v o -> Prop :=
-    match o with
-    | constant x => fun a b => `a == b
-    | function x y => fun a b => forall u, heq (a u) (b (`u))
-    end.
-\end{code}
-
-We connect the abstract theory to the concrete theory. Concretely, let $\Sigma$ be the theory
-of monoids. Given a monoid, we can construct the corresponding $\Sigma$-algebra and an
-object in the category of $\Sigma$-algebras. Conversely, given an object in the category, we
-construct a monoid.
-\begin{code}
- Variable o: Variety theory.
- Global Instance from_object: Monoid (o tt).
-\end{code}
-Similarly, for arrows.
+We plan to include the first homomorphism theorem and perhaps the soundness theorem for equational
+logic
 
@@
 %definitions.
+
 \section{Interfaces using category theory}\label{interfaces}\label{modul}
 The current work was stimulated by our desire~\cite{Riemann} to move between several
-implementations in a flexible way. In haskell type class polymorphism is used for this purpose.
-Hence, we do the same in type theory.
-
-We have characterized the naturals as the initial object in the category of semirings and derived
-many of their properties from this interface. Unary, binary and machine numbers are
-instances of this interface, so we can directly apply these results to all these instances.
-Similarly, for the integers are the initial ring. The rationals are the field of fractions of the
-integers. 
-Given a ring $R$, the $R$-algebra $R[X]$ of polynomials, is the free $R$-algebra of a set $X$.
-We provide two implementations of polynomials: the
-standard representation using lists of coefficients and the Bernstein representation. A Bernstein
-basis polynomial is one of the form:
-\[b_{\nu,n}(x) = {n \choose \nu} x^{\nu} \left( 1 - x \right)^{n - \nu}, \quad \nu = 0, \ldots, n.\]
-A Bernstein polynomial is a linear combination of these basic polynomials. Bernstein polynomials
-have been used for efficient computations inside the Coq system~\cite{ZumkellerPhD}.
+implementations in a flexible way. We use ad-hoc polymorphism to capture the natural
+numbers without referring to their actual implementation, we have
+characterized them as the initial object in the category of semirings and derived many of its
+properties from this interface. Unary, binary and machine numbers are
+instances of this interface, so we can directly apply these results to them.
 
@@
 \end{code}
+\vspace*{-5ex}
+\section{Conclusions}
+Telescopes have been criticized~\cite{Packed} for the lack of multiple inheritance and
+the efficiency penalty of a long chain of coercion projections. Packed classes~\cite{Packed} provide
+a solution to these problems. We provide an alternative solution. We have already
+discussed multiple inheritance in Section~\ref{hierarchy}. Moreover, by exposing the
+carrier we avoid the chains of projections.
 
-The context grows as follows:\marginpar{expand}
-
-\section{Type class quoting}\label{quote}
-Unification hints~\cite{Hints} allow one to semi-automatically construct a quote
-function\footnote{Gonthier provides similar functionality using canonical structures}.
-Unification hints are absent in Coq. Fortunately, type classes provide similar functionality as
-we will now show.
-
-We define a term language.
-\begin{code}
-Inductive Expr (V: Type) := Mult (a b: Expr V) | Zero | Var (v: V).
-\end{code}
-This defines a monad on Type\marginpar{Do this!}.
-The expression type is parameterized over the set of variable indices. Hence, we diverge
- from~\cite{Hints}, which uses |nat| for variables thereby introducing bounds problems and
- dummy variables.
-
-\noindent An expression is only meaningful in the context of a variable assignment:\footnote{nat is
-an Expr-algebra. eval is map vars composed with the algebra map.}
-\begin{code}
-Definition Value := nat.
-Definition Vars V := V -> Value.
-
-Fixpoint eval {V} (vs: Vars V) (e: Expr V): Value :=
-  match e with
-  | Zero => 0
-  | Mult a b => eval vs a * eval vs b
-  | Var v => vs v
-  end.
-\end{code}
-%
-Monads are trees with grafting. As explained by sigfpe~\cite{MonadsGrafting}, the tree monad
-$TX:=1+T^2X$, may be seen as the prototypical monad. Indeed, the |Expr| monad, i.e.\ the free
-$\Sigma$-algebra monad, behaves much like the tree monad. We now provide a direct definition of the
-corresponding bind operation.
-
-\begin{code}
-Fixpoint bind {V W} (f: V -> Expr W) (e: Expr V): Expr W :=
-  match e with
-  | Zero => Zero
-  | Mult a b => Mult (bind f a) (bind f b)
-  | Var v => vs v
-  end.
-\end{code}
-
-We have shown that |Expr|, i.e.\ |bind| together with |Var|, is a monad on the category of Types
-with extensional functions between them~\marginpar{Really?}.
-
-\noindent Some simple combinators for variable packs:
-%
-\begin{code}
-Definition novars: Vars False := False_rect _.
-Definition singlevar (x: Value): Vars unit := fun _ => x.
-Definition merge {A B} (a: Vars A) (b: Vars B): Vars (A+B) :=
-  fun i => if i then a j else b j.
-\end{code}
-
-\noindent These last two combinators are the `constructors' of an implicitly defined subset of
- Gallina terms, representing heaps, for which we implement syntactic lookup with type classes.
-Given a heap and value, |Lookup| instances give the value's index in the heap:\marginpar{A
-canonical structures approach would allow us to do this in Coq?}
-\begin{code}
-  Class Lookup {A} (x: Value) (f: Vars A) :=
-    { lookup: A; lookup_correct: f lookup = x }.
-\end{code}
-% Context (x: Value) {A B} (va: Vars A) (vb: Vars B).
-
-If the heap is a merge of two heaps and we can find the value's index in the left heap, we can
-access it by indexing the merged heap (and vice versa)
-\begin{code}
-  Global Instance lookup_left `{!Lookup x va}: Lookup x (merge va vb)
-    := { lookup := inl (lookup x va) }.
-\end{code}
-If the heap is just a singlevar, we can easily index it.
-\begin{code}
-  Global Program Instance: Lookup x (singlevar x) := { lookup := tt }.
-\end{code}
-
-One useful operation we need before we get to Quote relates to variables and expression
- evaluation. As its name suggests, |map_var| maps an expression's variable indices.\marginpar{monad}
-
-\begin{code}
-Definition map_var {V W: Type} (f: V -> W):
-    Expr V -> Expr W :=
-  fix F (e: Expr V): Expr W :=
-    match e with
-    | Mult a b => Mult (F a) (F b)
-    | Zero => Zero
-    | Var v => Var (f v)
-    end.
-\end{code}
-In |Quote| below, the idea is that |V, l|, and |n| are all input variables, while |V'| and |r| are
-output variables (in the sense that we will rely on unification to generate them). |V| and |l|
-represent the current heap, |n| represents the value we want to quote, and |V'| and |r'| represent
-the heap of newly encountered variables during the quotation.
-  This explains the type of quote: it is an expression that refers either to variables from
-the old heap, or to newly encountered variables. Finally, |eval_quote| is the usual correctness
-property, which now merges the two heaps.
-
-\begin{code}
-  Class Quote {V} (l: Vars V) (n: Value) {V'} (r: Vars V') :=
-    { quote: Expr (V + V')
-    ; eval_quote: @eval (V+V') (merge l r) quote = n }.
-\end{code}
-
-Our first instance for |Zero| is easy. The `novars' in the result type reflects the fact that no
-new variables are encountered.
-\begin{code}
-  Global Program Instance quote_zero V (v: Vars V):
-    Quote v 0 novars :=
-  { quote := Zero }.
-\end{code}
-
-The instance for multiplication is a bit more complex. The first line consists of
- variable declarations. The second line is important. `Quote x y z' must be read as
- `quoting y with existing heap x generates new heap z', so the second line basically just
-shuffles heaps around.
- The third line contains some |map_var|'s because the heap shuffling must be
-reflected in the variable indices, but apart from that it's just constructing a |Mult| term with
-quoted subterms.
-
-\begin{code}
-Global Program Instance quote_mult V (v: Vars V)
-  n V' (v': Vars V') m V'' (v'': Vars V'')
-  `{!Quote v n v'} `{!Quote (merge v v') m v''}:
-  Quote v (n * m) (merge v' v'') :=
-  { quote := Mult (map_var bla (quote n)) (map_var sum_assoc (quote m)) }.
-\end{code}
-  Now follows the instance where we recognize values that are already in the heap. This
-   is expressed by the Lookup requirement, which will only be fulfilled if the Lookup instances
-   defined above can find the value in the heap. The novars in the |Quote v x novars| result
-   reflects that this quotation does not generate new variables.
-
-\begin{code}
-Global Program Instance quote_old_var
-  V (v: Vars V) x {i: Lookup x v}:
-  Quote v x novars | 8 := { quote := Var (inl (lookup x v)) }.
-\end{code}
-
-\noindent Finally, the instance for new variables. We give this a lower priority so that it is only
-used if Lookup fails. The |8| in the previous example is a random number less than |9| below. A
-similar method is used in Coq's notation mechanism. We strongly recommend a change in
-implementation supporting an agda style precedence relation which is only restricted to be a
-directed acyclic graph~\cite{danielsson2009parsing}.
-
-\begin{code}
-Global Program Instance quote_new_var V (v: Vars V) x:
-  Quote v x (singlevar x) | 9 := { quote := Var (inr tt) }.
-\end{code}
-
-\noindent As announced this provides a light-weight quoting function\footnote{|eval_quote'| is a
-modification of |eval_quote| which works in the empty context.}:
-\begin{code}
-Goal forall x y (P: Value -> Prop), P ((x * y) * (x * 0)).
-  intros. rewrite <- (eval_quote' _).
-\end{code}
-This turns the goal into | (eval variable_pack quote)|.
-
-It would be interesting to use this technique to implement the |congruence| tactic, which implements
-the congruence closure algorithm~\cite{corbineau2007deciding}. One would naturally obtain a tactic
-which moreover works on setoid equalities.
-
-\section{Canonical structures}\label{canonical}
 Packed classes use canonical structures for the algebraic hierarchy. Both canonical structures and
@@
 
-\subsection{Big operators}
-Canonical structures have been used to provide a uniform treatment~\cite{bertot2008canonical} of big
-operators (like $\Pi,\sum, \max$). These operators extend a pair of a binary and a 0-ary operation
-to an $n$-ary operation for any $n$. Categorically, one considers the algebra maps from non-empty
-lists, lists, inhabited finite sets and finite sets to the carrier of a semigroup, monoid,
-commutative semigroup, commutative monoid. Hence we reuse the libraries for lists etc.
+According to |coqwc|, our development consists of 3103 lines of specifications and 707 lines of
+proofs. %targz
 
-We use type classes to deduce the relevant monoid operation and write:
-\begin{code}
-Definition seq_sum
-  `{Sequence A T} `{RingPlus A} `{z: RingZero A}: T -> A
-  := @seq_to_monoid A T _ A ring_plus z id.
-  Eval compute in seq_sum [3; 2].
-\end{code}
-
-\section{Constructive mathematics and proof irrelevance}\label{explicit}
-As originally concieved~\cite{ITT,CMCP}, Propositions are Types and we can extract information from
-them. In practice, it is convenient to assume that all proofs of a proposition are equal, the
-principle of proof irrelevance. These two views clearly contradict eachother. In Coq, a middle way
-was choosen, we are forbidden to extract information from proofs, but proofs are not identified.
-Here we need to pay, but do not seem to get anything in return.%
-\footnote{To be precise, what we get in return is the possibility to extract precisely the program
-we want.}
-There are various proposals to add the principle of proof irrelevance to Coq, perhaps a
-promissing one is the Calculus of implicit
-constructions~\cite{miquel2001implicit,barras2008implicit}. Such a change would fit naturally with
-the Hoare-style separation between programs and proofs which is used by the |Program| machinery.
-
-On the other hand, we found instances where we needed proof \emph{relevance}. In this case, it lead
-to a duplication of code which may have been avoided by universe polymorphism for |Fixpoint|s. In
-practice, manual dead-code analysis seems useful. Making Harrop formulas proof irrelevant seems to
-be a good first approximation~\cite{lcf:spi:03}.
-
-\section{Conclusions}
-Telescopes have been criticized~\cite{Packed} for the lack of multiple inheritance and
-the efficiency penalty of a long chain of coercion projections. Packed classes~\cite{Packed} provide
-a solution to these problems. We provide an alternative solution. We have already
-discussed multiple inheritance in Section~\ref{hierarchy}. Moreover, by exposing the
-carrier we avoid the chains of projections.
-
-We conclude that unification hints, canonical structures and type classes all appear to have
-similar expressive powers. Canonical structures, being tied to the unifier, are more robust, but
-seems to require more inguinity. Type classes are easier to use. Unification hints, may be a
-good generalization, but are absent from Coq. We encourage their inclusion to the
-Coq system.
-
-An obvious topic for future research is the extension from equational logic to partial Horn
-logic~\cite{palmgren2007partial}. Another topic would be to fully, but practically, embrase the
-categorical approach to universal algebra~\cite{pitts2001categorical}.
-
-According to |coqwc|, our development consists of 3103 lines of specifications and 707 lines of
-proofs.\marginpar{update,targz}
-
-% Alt-Ergo congruence closure parametrized by an equational theory.
-% As usual either implement in Coq (verify the algorithm), or Check traces, or Check Alt-Ergo (Work
-% in progress).
-
-% http://lara.epfl.ch/dokuwiki/sav09:congruence_closure_algorithm_and_correctness
-% I understand that this is the algorithm underlying congruence and perhaps first-order.
-% 
-% Type classes vs soft typing (Mizar types).
-% Type classes and types in homalg?
-
+% \section{Other systems}
+% We will provide a fuller comparison with other systems at a later stage.
+% We specifically mention recent work on unification hints in Matita~\cite{Hints}. They however, do
+% not address the challenge of proofs in mathematical structures.
 \paragraph{Acknowledgements.}
-This research was stimulated by the new possibilities provided by the introduction of type classes
-in Coq. In fact, this seems to be the first substantial development that tries to exploid all their
-possibilities. As a consequence, we found many small bugs and unintended behavior in the type
-class implementation. All these issues were quickly solved by Matthieu Sozeau. Discussions with
-Georges Gonthier and Claudio Sacerdoti Coen have helped to
-sharpen our understanding of the relation with Canonical Structures and with Unification Hints.
-%Jeremy, Thierry, Claudio, Georges?, Wouter
+We would like to thank Matthieu Sozeau for discussions and quickly solving numerous small bugs and
+feature requests. %Jeremy,Thierry, Claudio, Georges?, Wouter
 %\enlargethispage{5mm}

### 2010-02-25 compile.sh

@@ -1,0 +1,13 @@
+#!/bin/sh
+
+lhs2TeX --agda< $1.tex > processed.tex
+cp processed.tex tmp.tex
+
+set -e
+xelatex tmp.tex
+bibtex tmp
+xelatex tmp.tex
+bibtex tmp
+xelatex tmp.tex
+cat tmp.pdf > $1.pdf
+rm tmp.*

### 2010-02-26 jar.tex
Major rewrite.

@@
 \usepackage{ucs}
+\usepackage{amsmath}
 % \usepackage[utf8x,utf8]{inputenc}
@@
 \begin{document}
-\title{The algebraic hierarchy in type theory using type classes}
+\title{DRAFT: The algebraic hierarchy in type theory using type classes}
 \author{Bas Spitters \and Eelis van der Weegen}
@@
 construction of
-free $\Sigma$-algebras; see Section~\ref{univ}. We are also concerned with efficiency.
+free $\tau$-algebras; see Section~\ref{univ}. We are also concerned with efficiency.
 In fact, our aim is to use the present development as a new basis for Bishop's program in type
@@
 
+Future versions of Coq are planned to support proof irrelevance: every |p,q:P:Prop| are
+convertible; see section~\cite{explit}.
+
 \subsection{Type classes}
-Implementation in Ltac, Notation etc
-Coercions
+Implementation in Ltac, Notation etc\\
+Coercions\\
+Unification hints
 
@@
 
-We \emph{could} become over-eager and decide to unbundle the equivalence proof term as well:
-\begin{code}
-Record Monoid (A: Type) (R: A → A → Prop)
-  (e: Equivalence R) : Type := ...
-\end{code}
-However, this way if we happen to have two proofs |p| and |q| that a certain relation is an
-equivalence, we end up with two non-convertible ways to express that the relation (together with
-some unit and binary operation) form a Monoid: 'Monoid A R p ...' and 'Monoid A R q ...'. We must
-either try to deal with this (by proving that Monoid proofs with different Equivalence proofs are
-interchangable, and hope for automation, such as type class instance resolution, to use them
-interchangably), or take great care to ensure that |Equivalence| proof terms are always shared,
-essentially elevating them to `values', which goes against the principle of proof
-irrelevance.\marginpar{What if we add proof irrelevance?}
+% We \emph{could} become over-eager and decide to unbundle the equivalence proof term as well:
+% \begin{code}
+% Record Monoid (A: Type) (R: A → A → Prop)
+%   (e: Equivalence R) : Type := ...
+% \end{code}
+% However, this way if we happen to have two proofs |p| and |q| that a certain relation is an
+% equivalence, we end up with two non-convertible ways to express that the relation (together with
+% some unit and binary operation) form a Monoid: 'Monoid A R p ...' and 'Monoid A R q ...'. We must
+% either try to deal with this (by proving that Monoid proofs with different Equivalence proofs are
+% interchangable, and hope for automation, such as type class instance resolution, to use them
+% interchangably), or take great care to ensure that |Equivalence| proof terms are always shared,
+% essentially elevating them to `values', which goes against the principle of proof
+% irrelevance.\marginpar{What if we add proof irrelevance?}
+% BOTTOMLINE: Bundle as much as we can!
+% 
+% Neither of these options are very appealing. We really should just make |Monoid| take as
+% parameters
+% only the operations/relations. If we treat the monoid's unit and binary operation the same way,
+% this
+% means that |Monoid| ends up in |Prop|, and that its instances can be left opaque, which is really
+% how things ought to be.
 
-Neither of these options are very appealing. We really should just make |Monoid| take as parameters
-only the operations/relations. If we treat the monoid's unit and binary operation the same way, this
-means that |Monoid| ends up in |Prop|, and that its instances can be left opaque, which is really
-how things ought to be.
-
-Another benefit of this approach is that it works very smoothly when these records are in fact type
+Another benefit of our approach is that it works very smoothly when these records are in fact type
 classes. When relations and operations are bundled and do \emph{not} appear as parameters of the
-structure record/class, expressing as a type class constraint that some collection of relations and
-operations posess the structure seems to require manifest fields; see Section~\ref{manifest}. With
+structure record/class, to express that some collection of relations and
+operations posess the structure as a type class constraint seems to require manifest fields; see
+Section~\ref{manifest}. With
 our unbundled approach, on the other hand, such type class constraints are easily expressed without
-any extensions as simple type inhabitation queries
+any extensions as type inhabitation queries
 % (e.g. ``hey Coq, find me a proof of [Monoid some_rel some_op some_unit]'')
@@
 
-We remark that leaving the equivalence proofs bundled means that when one has a proof that something
+Leaving the equivalence proofs bundled means that when one has a proof that a structure
 is a semiring, one has two proofs (via the two monoids) that the equivalence relation is an
-equivalence. Fortunately, having redundant proofs like this is entirely harmless when one's terms
+equivalence. Fortunately, this redundancy is entirely harmless since our terms
 never refer (directly or indirectly) to proofs, which is the case when the relations and operations
-they are composed of are left unbundled. Here, too, we enjoy the warm embrace of proof irrelevance.
+they are composed from are left unbundled. Here, too, we enjoy the warm embrace of proof
+irrelevance.
 
@@
  \item (as a second consequence) algebraic expressions need/should never refer (directly or
-indirectly) to proofs.
+indirectly) to proofs about the algebraic structure.\footnote{The reciprocal in a field
+famously does refer to a proof that the element is non-zero, but not to a proof that the operations
+indeed form a field.}
 \end{enumerate}
@@
 \begin{code}
-Lemma assoc1 A eq op (P: SemiGroup A eq op) x y z:
+Lemma perm A eq op (P: SemiGroup A eq op) x y z:
                    eq (op (op x y) z) (op (op z y) x).
@@
                  (op: SemiGroupOp A): Prop := ...
-  Lemma assoc1 A eq op (P: SemiGroup A eq op) x y z:
+Lemma perm A eq op (P: SemiGroup A eq op) x y z:
                  x & y & z = z & y & x.
@@
 may use in the proof. This lack of projections keeps our terms small and independent, and keeps
-rewriting simple and sane.
+rewriting simple and sane; see Section~\ref{canonical}.
 
 On the other hand, it \emph{does} mean that all the carriers/relations/operations end up in the
-context. The context for |assoc1| looks like:
+context. The context for |perm| looks like:
 \begin{code}
@@
 likely not be problematic for any but the most complex formalizations), having to \emph{declare}
-these entities (like |eq| and |op| for |assoc1|) \emph{is} a bit of a chore. Fortunately, Coq
+these entities (like |eq| and |op| for |perm|) \emph{is} a bit of a chore. Fortunately, Coq
 provides a feature called implicit generalization, which is exactly what we need.
-\marginpar{describe how it works} With it, we can write |assoc1| as follows:
+With it, we can write |perm| as follows:
 \begin{code}
-  Lemma assoc1 `(SemiGroup A) x y z: x & y & z = z & y & x.
+  Lemma perm `(SemiGroup A) x y z: x & y & z = z & y & x.
 \end{code}
@@
 
-\section{Subset types}
-BEGIN SPECULATIVE
-
-A record type is \marginpar{mostly?} syntactic sugar for a $\Sigma$-type. In the presence of a
-squash operator, |[]:Type->Prop|, we may define a subset type as |\{A\mid B\}=Sig A. [B]|. This is
-pursued by Barras~\cite{Barras:subset,Werner}. The program machinery fakes subset types in the
-present Coq implementation. Proof irrelevanve in Coq would allow a faithful
-implementation.
-
-Squash is a monad and allows many constructions: subsets, unions, images, heterogeneous equality,
-\ldots.
-
-Question: Would the meta-theory simplify if we would just add a conversion rule $x\triangleright
-\epsilon$ for $x:Prop$? We could add a term constructor $[a]$ which will have type $[A]$. There
-seems to be no reason to connect this to |Prop| or impredicativity.
-
-END SPECULATIVE
-
-We add another principle:
-\begin{itemize}\setcounter{enumi}{3}
- \item Working hypothesis: `every record can be reordered as a subset'. Consequently, we
-rewrite this subset |\{A\mid B\}| as |Record r (a:A):={b:B}|.
-\end{itemize}
-
-\section{Algebraic hierarchy using type classes}\label{classes}
-\paragraph{Hierarchy.}\label{hierarchy}
-In our formalization of algebraic structures we use the structure as (implicit) parameters, but
+\section{Putting it all together}\label{classes}
+Combining the ideas in the previous sections we end up with a formalization of algebraic
+structures where we use the structure as (implicit) parameters, but
 pack the proofs. This can be nicely organized by using canonical names. For instance, we define:
@@
 % it.
-We note that |op| does not depend on the proof that |e| is an equivalence. We use Coq's implicit
-quantification (|`{}|) to avoid having to write all the parameters when \emph{stating} a theorem
-and Coq's maximally inserted implicit arguments to find the parameters when \emph{applying} a
-theorem. Both features are new in Coq and stem from the type class implementation.
+We note that |op| does not depend on the proof that |e| is an equivalence. As explained above we
+use Coq's implicit quantification (|`{}|) to avoid having to write all the parameters when
+\emph{stating} a theorem and Coq's maximally inserted implicit arguments to find the parameters when
+\emph{applying} a theorem. Both features are new in Coq and stem from the type class implementation.
 
@@
 those of telescopes or packed classes. In our experience, this has been relatively
-harmless\footnote{Coq's data structure for contexts is not very efficient however, we believe this
+harmless\footnote{Coq's data structure for contexts is not very efficient. Gonthier fears that this
 may be a bottleneck for huge developments. It seems that the data structure chosen
 in~\cite{asperti2009compact} will behave better.}%
-: most
-terms in the context are there to support canonical names. Bigger terms
+: most terms in the context are there to support canonical names. Bigger terms
 \emph{do} cause problems: 1. when proofs are part of mathematical objects we need to share these
@@
 
-Manifest fields in records allow us to fix a field of a dependent record.
+Manifest fields in records allow us to fix a field of a dependent record. For example, as in the
+example above, |Monoid with op=mult|. 
 Luo~\cite{DBLP:conf/types/Luo08} has proposed coersive subtyping to implement manifest fields.
@@
 Since |op| is \emph{instantiated} with |mult|, we need not worry about the
-equality |op=mult|.
+equality |op=mult|. This seems to be a general phenomenon. For instance, instead of searching
+for a |Monoid with op=mult| we would use type class resolution to find an instance of |Monoid mult|.
+We have thus addressed the problem of multiple inheritance.
 
 \section{Universal algebra}\label{univ}
-After having implemented concrete algebraic structures, we generalized some of our constructions
+After having implemented concrete algebraic structures, we generalize some of our constructions
 by implementing many-sorted universal algebra. A formalization of universal algebra
-already exists in Coq~\cite{DBLP:conf/tphol/Capretta99}. It has recently been extended with more
-categorical results~\cite{dominguez2008formalizing}. We have conveniently redeveloped most of this
-using type classes. We defined equational theories and their varieties:
-the categories of their models. A quote function helps to translate a statement for an instance of
-a type class to a statement in the equational logic.
+already exists in Coq~\cite{DBLP:conf/tphol/Capretta99} and has recently been extended with
+more categorical results~\cite{dominguez2008formalizing}. We have conveniently redeveloped most of
+this using type classes. We define equational theories and their varieties of algebras, i.e.\ their
+finitary algebraic category of models. A quote function helps to translate a statement for an
+instance of
+a type class to a statement in the equational logic. We will come back to this in
+Section~\ref{quote}.
 
-\subsection{Category theory}\label{cats}
-Using type classes, we provide a basic library for category theory: categories, functors,
-adjunctions, monads,\ldots We follow the usual type theoretical definition of a\marginpar{Use
-notation for comp} category~\cite{saibi1995constructive}
-\begin{code}
- Class Category O `{!Arrows O} `{forall x y: O, Equiv (x --> y)}
-   `{!CatId O} `{!CatComp O}: Prop :=
-  { arrow_equiv:> forall x y: O, Setoid (x --> y)
-  ; comp_proper:> forall x y z: O,
-      Proper (equiv ==> equiv ==> equiv)%signature (@comp _ _ _ x y z)
-  ; comp_assoc: forall (w x y z: O) (a: w --> x) (b: x --> y) (c: y --> z),
-      comp c (comp b a) == comp (comp c b) a
-  ; id_l: forall (x y: O) (a: y --> x), comp cat_id a == a
-  ; id_r: forall (x y: O) (a: x --> y), comp a cat_id == a }.
-\end{code}
-This is based on the 2-categorical idea that we have equality on arrows, but not on objects.
-Similarly, we have equality on natural transformations, but not on functors.
-
-%format ◎ = "\circ"
-For instance natural transformations are formalized as: |◎|
-\begin{code}
-  Context `{Category C} `{Category D} (F G: C -> D)
-     `{!Functor F Fa} `{!Functor G Ga}.
-
-  Class NaturalTransformation (η: forall c: C, F c --> G c): Prop :=
-    natural: forall (x y: C) (f: x --> y), η y ◎ fmap F f == fmap G f ◎ η x.
-\end{code}
-
-The notation |◎| denotes function composition.
-Here |fmap F| explicitly refers to the arrow part of the functor |F|.\\
-In mathematics one would apply $F$ to both objects and arrows ($F x$, $F f$). However, this does
-not seem to fit with type theory, hence we follow the customs in functional programming.
-
-The proofs are short and follow the textbook easily.\marginpar{really?}
-
-We provide the category of setoids, the category of categories, the dual of a category, products,
-and hence co-products, the category of $\Sigma$-algebras.
-
-We provide the construction of the term-algebra for a signature $\Sigma$ and show that it is
-initial in the category of $\Sigma$-algebras. We proceed to prove that this provides a left-adjoint
-to the forgetful functor. The resulting expression monad will be used below.
-
-\subsection{Universal algebra continued}\label{univ2}
+% \subsection{Universal algebra continued}\label{univ2}
 
@@
 theorem.~\marginpar{Description of the work.}
-%perhaps the soundness theorem for equational logic.
+Discuss: the duplication of code which may have been avoided by universe
+polymorphism for |Fixpoint|s.
+%perhaps the soundness theorem for equational logic?
 
-In order to prove the first homophism theorem, we define a congruence:
+In order to prove the first homophism theorem, we define a congruence.
+The compatibility of the relation |e| can be stated in two provably  equivalent ways.
+Directly, |eAlgebra := @Algebra et v e _| or  indirectly, by stating that the relation
+as a set of pairs is a subalgebra in the product algebra.
 %theory/ua_congruence.v
+% \begin{code}
+%  eSub := @ua_subalgebraT.ClosedSubset et vv _ _ Q.
+% \end{code}
+Subsequently, we choose for former as a definition:
 \begin{code}
@@
 \end{code}
-We note the use of |Proper|.
+A congruence relation is coarser than the setoid relation, as indicated by the use of |Proper|
+above. Since our theory is founded on a congruence, the initial setoid equality, the development of 
+congruences and quotients benefits from this.
 
-We define the forgetful functor from the category of $\Sigma$-algebras to the category of sets.
+We define the forgetful functor from the category of $\tau$-algebras to the category of sets.
+We define the term algebra consisting of closed terms, terms with the empty type as set of
+variables. The term algebra is the free $\tau$-algebra and hence\marginpar{?}
+initial. Categorically: we have defined a left-adjoint to the forgetful functor. A formal proof
+connects the categorical and the algebraic definition.
+One may expect that this term algebra could be useful to
+\emph{define}, say, the natural numbers as the term algebra for the theory of semiring. However, it
+seems difficult to prove decidability of equality, which requires a normalization procedure.
 
-We define the term algebra consisting of closed terms, terms with the empty type as set of
-variables. % extensionally equal closed term???
-We show that the term algebra is initial. One may expect that this could be useful to
-\emph{define}, say, the natural numbers as the initial semiring. However, it seems difficult to
-prove decidability of equality, as it seems to require a normalization procedure.
-
-We define the product of two algebras and show that it is the categorical product.
+The product of two $\tau$-algebras is their categorical product.
 
@@
 
-We connect the abstract theory to the concrete theory. Concretely, let $\Sigma$ be the theory
-of monoids. Given a monoid, we can construct the corresponding $\Sigma$-algebra and an
-object in the category of $\Sigma$-algebras. Conversely, given an object in the category, we
+We connect the abstract theory to the concrete theory. Concretely, let $\tau_m$ be the theory
+of monoids. Given a monoid, we can construct the corresponding $\tau_m$-algebra and an
+object in the category of $\\tau$-algebras. Conversely, given an object in the category/variety, we
 construct a monoid.
+%varieties/monoid.v
 \begin{code}
- Variable o: Variety theory.
+Variable o: variety.Object theory.
+Global Instance: SemiGroupOp (o tt) := algebra_op theory mult.
+Global Instance: MonoidUnit (o tt) := algebra_op theory one.
 Global Instance from_object: Monoid (o tt).
 \end{code}
-Similarly, for arrows.
+In the last line, instance resolution `automatically' finds the operation and the unit specified in
+the lines before.\marginpar{Why this way?}
 
-%There is a slight problem when defining the forgetful functor from the category of
-%$\Sigma$-algebras to Sets. This problem is caused by our use of the category of all
-%categories\footnote{Perhaps we should have used higher categories for this.}.
-%We conjecture that this problem disappears when Coq's universe polymorphism is extended to
-%definitions.
+\subsection{Category theory}\label{cats}
+Using type classes, we provide a basic library for category theory: categories, functors,
+adjunctions, monads,\ldots We follow the usual type theoretical definition of a
+category~\cite{saibi1995constructive}
+
+%format ◎ = "\circ"
+%interfaces/abstract_algebra.v
+\begin{code}
+Class Category O `{!Arrows O} `{Π x y: O, Equiv (x ⟶ y)}
+  `{!CatId O} `{!CatComp O}: Prop :=
+  { arrow_equiv:> Π x y: O, Setoid (x ⟶ y)
+  ; comp_proper:> Π x y z,
+      Proper (equiv ==> equiv ==> equiv)%signature (@comp _ _ _ x y z)
+  ; comp_assoc (w x y z: O) (a: w ⟶ x) (b: x ⟶ y) (c: y ⟶ z):
+      c ◎ (b ◎ a) = (c ◎ b) ◎ a
+  ; id_l `(a: x ⟶ y): cat_id ◎ a = a
+  ; id_r `(a: x ⟶ y): a ◎ cat_id = a }.
+\end{code}
+Where |%signature|
+tells Coq how to interpret |equiv|.\marginpar{Where is this defined?}
+This is based on the 2-categorical idea that we have equality on arrows, but not on objects.
+Similarly, we have equality on natural transformations, but not on functors.
+
+%theory/categories.v
+Natural transformations are formalized as:
+\begin{code}
+  Context `{Category C} `{Category D}
+  `{!Functor (F: C → D) Fa} `{!Functor (G: C → D) Ga}.
+
+  Class NaturalTransformation (η: Π c, F c ⟶ G c): Prop :=
+    natural: Π (x y: C) (f: x ⟶ y), η y ◎ fmap F f = fmap G f ◎ η x.
+\end{code}
+
+Here |fmap F| explicitly refers to the arrow part of the functor |F|.
+In mathematics one would apply $F$ to both objects and arrows ($F x$, $F f$). However, this does
+not seem to fit with type theory.\marginpar{Explain why.}
+
+The proofs are short and follow the textbook easily.\marginpar{really?}
+
+We have defined the category of setoids, the category of categories, the dual of a category,
+products, and hence co-products, the category of $\tau$-algebras.
+
+We provide the construction of the term-algebra for a type $\tau$ and show that it is
+initial in the category of $\tau$-algebras. We proceed to prove that this provides a left-adjoint
+to the forgetful functor. The resulting expression monad will be used in Section~\ref{quote} below.
+
 \section{Interfaces using category theory}\label{interfaces}\label{modul}
@@
 instances of this interface, so we can directly apply these results to all these instances.
-Similarly, for the integers are the initial ring. The rationals are the field of fractions of the
+Similarly, the integers are the initial ring. The rationals are the field of fractions of the
 integers. 
-Given a ring $R$, the $R$-algebra $R[X]$ of polynomials, is the free $R$-algebra of a set $X$.
+Given a ring $R$, the $R$-algebra $R[X]$ of polynomials, is the free $R$-algebra on a set $X$.
 We provide two implementations of polynomials: the
@@
 
-The context grows as follows:\marginpar{expand}
+%There is a slight problem when defining the forgetful functor from the category of
+%$\Sigma$-algebras to Sets. This problem is caused by our use of the category of all
+%categories\footnote{Perhaps we should have used higher categories for this.}.
+%We conjecture that this problem disappears when Coq's universe polymorphism is extended to
+%definitions.
 
@@
 
-We define a term language.
+We define a term language for monoids:
 \begin{code}
@@
 \end{code}
-This defines a monad on Type\marginpar{Do this!}.
+This defines a monad on the category of |Type|s with extensional functions.\marginpar{Proof this
+formally}.
 The expression type is parameterized over the set of variable indices. Hence, we diverge
@@
 
-\noindent An expression is only meaningful in the context of a variable assignment:\footnote{nat is
-an Expr-algebra. eval is map vars composed with the algebra map.}
+\noindent An expression is only meaningful in the context of a variable
+assignment:\footnote{Categorically: |nat| is
+an |Expr|-algebra and |eval| is |map vars| composed with the algebra map.}
 \begin{code}
@@
 %
-Monads are trees with grafting. As explained by sigfpe~\cite{MonadsGrafting}, the tree monad
+Monads are trees with grafting~\cite{MonadsGrafting}: the tree monad
 $TX:=1+T^2X$, may be seen as the prototypical monad. Indeed, the |Expr| monad, i.e.\ the free
-$\Sigma$-algebra monad, behaves much like the tree monad. We now provide a direct definition of the
-corresponding bind operation.
+|Expr|-algebra monad, behaves much like the tree monad. We now provide a direct definition of the
+corresponding bind operation, the arrows in the Kleisli category.
 
@@
 
-We have shown that |Expr|, i.e.\ |bind| together with |Var|, is a monad on the category of Types
-with extensional functions between them~\marginpar{Really?}.
+% We have shown that |Expr|, i.e.\ |bind| together with |Var|, is a monad on the category of Types
+% with extensional functions between them~\marginpar{Really?}.
 
-\noindent Some simple combinators for variable packs:
+\noindent We define some simple combinators for variable packs:
 %
@@
  Gallina terms, representing heaps, for which we implement syntactic lookup with type classes.
-Given a heap and value, |Lookup| instances give the value's index in the heap:\marginpar{A
-canonical structures approach would allow us to do this in Coq?}
+The heap can also be defined explitly, with no essential change in the code.
+Given a heap and value, |Lookup| instances give the value's index in the heap:
+% \marginpar{A
+% canonical structures approach would allow us to do this in Coq?}
 \begin{code}
@@
 One useful operation we need before we get to Quote relates to variables and expression
- evaluation. As its name suggests, |map_var| maps an expression's variable indices.\marginpar{monad}
+ evaluation. As its name suggests, |map_var| maps an expression's variable indices.
+This reindexing function is the the |map| of the |Expr|-monad.
 
@@
 used if Lookup fails. The |8| in the previous example is a random number less than |9| below. A
-similar method is used in Coq's notation mechanism. We strongly recommend a change in
-implementation supporting an agda style precedence relation which is only restricted to be a
-directed acyclic graph~\cite{danielsson2009parsing}.
+similar method is used in Coq's notation mechanism.
+% We recommend a change in
+% implementation supporting an agda style precedence relation which is only restricted to be a
+% directed acyclic graph~\cite{danielsson2009parsing}.
 
@@
 structures to derive the setoid equality on the natural
-numbers in the following example |(S x)==(S y)|. We have conveniently reimplemented this with
-type classes with superior proof terms. Canonical structures give
+numbers in the following example |(S x)==(S y)|. In this case type classes provide superior proof
+terms. Canonical structures give
 \begin{code}
@@
 \end{code}
-
 which includes an explicit proof that |(@eq nat)| is an equivalence,
@@
 lists, lists, inhabited finite sets and finite sets to the carrier of a semigroup, monoid,
-commutative semigroup, commutative monoid. Hence we reuse the libraries for lists etc.
+commutative semigroup, commutative monoid. Hence we want to reuse the libraries for lists etc.
 
@@
 \end{code}
+\marginpar{distributivity?}
 
-\section{Constructive mathematics and proof irrelevance}\label{explicit}
+\section{Subset types and proof irrelevance}
+We have mentioned proof irrelevance a number of times before. It is an important theme in our
+approach. The |Program| machinery allows one to conveniently write programs with Hoare-style
+pre-
+and post-conditions. I.e.\ functions $f: \{ A \mid P \} \to \{ B \mid Q \}$. Both side conditions
+are intended to be proof irrelevant, they are in |Prop|. Presently, Coq does not support such subset
+types, thus forcing the user to manually prove that |f| does not depend on the proof of |P|.
+The addition of proof irrelevance and subset types to Coq is pursued by
+Barras~\cite{Barras:subset,Werner}.
+
+% Squash is a monad and allows many constructions: subsets, unions, images, heterogeneous equality,
+% \ldots.
+
+We add another principle:
+\begin{enumerate}\setcounter{enumi}{3}
+ \item Working hypothesis: every record can be reordered in such a way that all the non-informative
+fields come last.
+\end{enumerate}
+
+Pollack~\cite{pollack2000dependently} introduces record types with telescopes. 
+Given a telescope\[
+T=[x_1:A_1][x_2:A_2(x_1)]\cdots[x_k:A_k(x_1,\ldots,x_{k_1})]
+\]
+There is an inductively defined type $\Sigma T$, with a single constructor, given by the following 
+formation and introduction rule in pseudo-code for $k=2$:
+\begin{align*}
+ A_1 :& Type\\
+ A_2 :& \Pi A_1,Type\\
+\cline{1-2}
+\Sigma T :& Type
+\end{align*}
+
+\begin{align*}
+ \Sigma T:Type\quad a_1 &:A_1 \quad a_2 : A_2(a_1)\\
+\cline{1-3}
+(a_1,a_2) &:\Sigma T
+\end{align*}
+
+For our methodology to work we have to be able to transform the occuring record into a telescopic
+subset $\{ A \mid  B\}$. We moreover conjecture that the terms in |B| only depend on the
+variables in |A|, but not in |B| itself. However, we do not need this.
+
+\subsection{Propositions as [ ]-Types}\label{explicit}
 As originally concieved~\cite{ITT,CMCP}, Propositions are Types and we can extract information from
-them. In practice, it is convenient to assume that all proofs of a proposition are equal, the
+them. In practice, it is convenient to assume that all proofs of a proposition are equal: the
 principle of proof irrelevance. These two views clearly contradict eachother. In Coq, a middle way
 was choosen, we are forbidden to extract information from proofs, but proofs are not identified.
-Here we need to pay, but do not seem to get anything in return.%
-\footnote{To be precise, what we get in return is the possibility to extract precisely the program
-we want.}
-There are various proposals to add the principle of proof irrelevance to Coq, perhaps a
-promissing one is the Calculus of implicit
-constructions~\cite{miquel2001implicit,barras2008implicit}. Such a change would fit naturally with
-the Hoare-style separation between programs and proofs which is used by the |Program| machinery.
+This allows us to extract the program we want, but we want more:
+We would like to use Coq as a dependently typed programming language. To do so,
+one would like to add the principle of proof irrelevance to Coq. A promissing proposal is the
+Calculus of implicit constructions with implicit
+$\Sigma$-types~\cite{miquel2001implicit,barras2008implicit,barrasSigma}. Such a change would fit
+naturally with the Hoare-style separation between programs and proofs which is used by the |Program|
+machinery.
 
-On the other hand, we found instances where we needed proof \emph{relevance}. In this case, it lead
-to a duplication of code which may have been avoided by universe polymorphism for |Fixpoint|s. In
-practice, manual dead-code analysis seems useful. Making Harrop formulas proof irrelevant seems to
-be a good first approximation~\cite{lcf:spi:03}.
+One may be tempted to propose Propositions as [ ]-types~\cite{awodey2004propositions}. Here the
+construction $[P]:=\{\top\mid p:P\}$ squashes all the information about how we proved $P$. However,
+there are places where we needed proof \emph{relevance}; e.g.\ in the proof of the first isomorphism
+theorem above.In practice, manual dead-code analysis seems useful. Making Harrop formulas proof
+irrelevant seems to be a good first approximation~\cite{lcf:spi:03}. However, a further refinement
+of Propositions as types seems necessary.
 
@@
 a solution to these problems. We provide an alternative solution. We have already
-discussed multiple inheritance in Section~\ref{hierarchy}. Moreover, by exposing the
+discussed multiple inheritance in Section~\ref{classes}. Moreover, by exposing the
 carrier we avoid the chains of projections.
@@
 similar expressive powers. Canonical structures, being tied to the unifier, are more robust, but
-seems to require more inguinity. Type classes are easier to use. Unification hints, may be a
+seem to require more inguinity. Type classes are easier to use. Unification hints, may be a
 good generalization, but are absent from Coq. We encourage their inclusion to the
@@
 %Jeremy, Thierry, Claudio, Georges?, Wouter
-%\enlargethispage{5mm}
 \bibliographystyle{plain}

### 2010-02-26 processed.tex
Major rewrite.

@@
 
+%%%format R = "\mathbb{R}"
+
 \arrayhs%i.e. no page breaks in code blocks
@@
 
+Future versions of Coq are planned to support proof irrelevance: every \ensuremath{\Varid{p},\Varid{q:P:Prop}} are
+convertible; see section~\cite{explit}.
+
 \subsection{Type classes}
-Implementation in Ltac, Notation etc
-Coercions
+Implementation in Ltac, Notation etc\\
+Coercions\\
+Unification hints
 
@@
 
-We \emph{could} become over-eager and decide to unbundle the equivalence proof term as well:
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\mathbf{Record}\;\Conid{Monoid}\;(\Conid{A:}\;\Conid{Type})\;(\Conid{R:}\;\Conid{A}\;\rightarrow\;\Conid{A}\;\rightarrow\;\Conid{Prop})\;{}\<[E]%
-\\
-\>[B]{}\hsindent{3}{}\<[3]%
-\>[3]{}(\Varid{e:}\;\Conid{Equivalence}\;\mathbb{R})\;\mathbin{:}\;\Conid{Type}\;\Conid{:=}\;\Varid{...}{}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-However, this way if we happen to have two proofs \ensuremath{\Varid{p}} and \ensuremath{\Varid{q}} that a certain relation is an
-equivalence, we end up with two non-convertible ways to express that the relation (together with
-some unit and binary operation) form a Monoid: 'Monoid A R p ...' and 'Monoid A R q ...'. We must
-either try to deal with this (by proving that Monoid proofs with different Equivalence proofs are
-interchangable, and hope for automation, such as type class instance resolution, to use them
-interchangably), or take great care to ensure that \ensuremath{\Conid{Equivalence}} proof terms are always shared,
-essentially elevating them to `values', which goes against the principle of proof
-irrelevance.\marginpar{What if we add proof irrelevance?}
-
-Neither of these options are very appealing. We really should just make \ensuremath{\Conid{Monoid}} take as parameters
-only the operations/relations. If we treat the monoid's unit and binary operation the same way, this
-means that \ensuremath{\Conid{Monoid}} ends up in \ensuremath{\Conid{Prop}}, and that its instances can be left opaque, which is really
-how things ought to be.
-
-Another benefit of this approach is that it works very smoothly when these records are in fact type
+% We \emph{could} become over-eager and decide to unbundle the equivalence proof term as well:
+% \begin{code}
+% Record Monoid (A: Type) (R: A → A → Prop)
+%   (e: Equivalence R) : Type := ...
+% \end{code}
+% However, this way if we happen to have two proofs |p| and |q| that a certain relation is an
+% equivalence, we end up with two non-convertible ways to express that the relation (together with
+% some unit and binary operation) form a Monoid: 'Monoid A R p ...' and 'Monoid A R q ...'. We must
+% either try to deal with this (by proving that Monoid proofs with different Equivalence proofs are
+% interchangable, and hope for automation, such as type class instance resolution, to use them
+% interchangably), or take great care to ensure that |Equivalence| proof terms are always shared,
+% essentially elevating them to `values', which goes against the principle of proof
+% irrelevance.\marginpar{What if we add proof irrelevance?}
+% BOTTOMLINE: Bundle as much as we can!
+% 
+% Neither of these options are very appealing. We really should just make |Monoid| take as
+% parameters
+% only the operations/relations. If we treat the monoid's unit and binary operation the same way,
+% this
+% means that |Monoid| ends up in |Prop|, and that its instances can be left opaque, which is really
+% how things ought to be.
+
+Another benefit of our approach is that it works very smoothly when these records are in fact type
 classes. When relations and operations are bundled and do \emph{not} appear as parameters of the
-structure record/class, expressing as a type class constraint that some collection of relations and
-operations posess the structure seems to require manifest fields; see Section~\ref{manifest}. With
+structure record/class, to express that some collection of relations and
+operations posess the structure as a type class constraint seems to require manifest fields; see
+Section~\ref{manifest}. With
 our unbundled approach, on the other hand, such type class constraints are easily expressed without
-any extensions as simple type inhabitation queries
+any extensions as type inhabitation queries
 % (e.g. ``hey Coq, find me a proof of [Monoid some_rel some_op some_unit]'')
@@
 \column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[3]{}\mathbf{Record}\;\Conid{SemiGroup}\;(\Conid{A:}\;\Conid{Type})\;(\Varid{eq:}\;\Conid{A}\;\rightarrow\;\Conid{A}\;\rightarrow\;\Conid{Prop})\;(\Varid{op:}\;\Conid{A}\;\rightarrow\;\Conid{A}\;\rightarrow\;\Conid{A})\;\mathbin{:}\;\Conid{Prop}\;\Conid{:=}\;\Varid{...}{}\<[E]%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathbf{Record}\;\Conid{SemiGroup}\;(\Conid{A:}\;\Conid{Type})\;(\Varid{eq:}\;\Conid{A}\;\rightarrow\;\Conid{A}\;\rightarrow\;\Conid{Prop})\;{}\<[E]%
+\\
+\>[B]{}(\Varid{op:}\;\Conid{A}\;\rightarrow\;\Conid{A}\;\rightarrow\;\Conid{A})\;\mathbin{:}\;\Conid{Prop}\;\Conid{:=}\;\Varid{...}{}\<[E]%
 \ColumnHook
@@
 \column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
 \column{20}{@{}>{\hspre}l<{\hspost}@{}}%
 \column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[3]{}\mathbf{Lemma}\;\Varid{assoc1}\;\Conid{A}\;\Varid{eq}\;\Varid{op}\;(\Conid{P:}\;\Conid{SemiGroup}\;\Conid{A}\;\Varid{eq}\;\Varid{op})\;\Varid{x}\;\Varid{y}\;\Varid{z:}\;{}\<[E]%
-\\
-\>[3]{}\hsindent{17}{}\<[20]%
+\>[B]{}\mathbf{Lemma}\;\Varid{perm}\;\Conid{A}\;\Varid{eq}\;\Varid{op}\;(\Conid{P:}\;\Conid{SemiGroup}\;\Conid{A}\;\Varid{eq}\;\Varid{op})\;\Varid{x}\;\Varid{y}\;\Varid{z:}\;{}\<[E]%
+\\
+\>[B]{}\hsindent{20}{}\<[20]%
 \>[20]{}\Varid{eq}\;(\Varid{op}\;(\Varid{op}\;\Varid{x}\;\Varid{y})\;\Varid{z})\;(\Varid{op}\;(\Varid{op}\;\Varid{z}\;\Varid{y})\;\Varid{x})\;.{}\<[E]%
@@
 \column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{21}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{30}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[3]{}\mathbf{Record}\;\Conid{SemiGroup}\;(\Conid{A:}\;\Conid{Type})\;(\Varid{eq:}\;\Conid{Equiv}\;\Conid{A})\;{}\<[E]%
-\\
-\>[3]{}\hsindent{27}{}\<[30]%
-\>[30]{}(\Varid{op:}\;\Conid{SemiGroupOp}\;\Conid{A})\;\mathbin{:}\;\Conid{Prop}\;\Conid{:=}\;\Varid{...}{}\<[E]%
-\\
-\>[3]{}\mathbf{Lemma}\;\Varid{assoc1}\;\Conid{A}\;\Varid{eq}\;\Varid{op}\;(\Conid{P:}\;\Conid{SemiGroup}\;\Conid{A}\;\Varid{eq}\;\Varid{op})\;\Varid{x}\;\Varid{y}\;\Varid{z:}\;{}\<[E]%
-\\
-\>[3]{}\hsindent{18}{}\<[21]%
-\>[21]{}\Varid{x}\;\Varid{\&}\;\Varid{y}\;\Varid{\&}\;\Varid{z}\;\mathrel{=}\;\Varid{z}\;\Varid{\&}\;\Varid{y}\;\Varid{\&}\;\Varid{x.}{}\<[E]%
+\column{18}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathbf{Record}\;\Conid{SemiGroup}\;(\Conid{A:}\;\Conid{Type})\;(\Varid{eq:}\;\Conid{Equiv}\;\Conid{A})\;{}\<[E]%
+\\
+\>[B]{}\hsindent{18}{}\<[18]%
+\>[18]{}(\Varid{op:}\;\Conid{SemiGroupOp}\;\Conid{A})\;\mathbin{:}\;\Conid{Prop}\;\Conid{:=}\;\Varid{...}{}\<[E]%
+\\
+\>[B]{}\mathbf{Lemma}\;\Varid{perm}\;\Conid{A}\;\Varid{eq}\;\Varid{op}\;(\Conid{P:}\;\Conid{SemiGroup}\;\Conid{A}\;\Varid{eq}\;\Varid{op})\;\Varid{x}\;\Varid{y}\;\Varid{z:}\;{}\<[E]%
+\\
+\>[B]{}\hsindent{18}{}\<[18]%
+\>[18]{}\Varid{x}\;\Varid{\&}\;\Varid{y}\;\Varid{\&}\;\Varid{z}\;\mathrel{=}\;\Varid{z}\;\Varid{\&}\;\Varid{y}\;\Varid{\&}\;\Varid{x.}{}\<[E]%
 \ColumnHook
@@
 may use in the proof. This lack of projections keeps our terms small and independent, and keeps
-rewriting simple and sane.
+rewriting simple and sane; see Section~\ref{canonical}.
 
 On the other hand, it \emph{does} mean that all the carriers/relations/operations end up in the
-context. The context for \ensuremath{\Varid{assoc1}} looks like:
+context. The context for \ensuremath{\Varid{perm}} looks like:
 \begin{hscode}\SaveRestoreHook
@@
 likely not be problematic for any but the most complex formalizations), having to \emph{declare}
-these entities (like \ensuremath{\Varid{eq}} and \ensuremath{\Varid{op}} for \ensuremath{\Varid{assoc1}}) \emph{is} a bit of a chore. Fortunately, Coq
+these entities (like \ensuremath{\Varid{eq}} and \ensuremath{\Varid{op}} for \ensuremath{\Varid{perm}}) \emph{is} a bit of a chore. Fortunately, Coq
 provides a feature called implicit generalization, which is exactly what we need.
-\marginpar{describe how it works} With it, we can write \ensuremath{\Varid{assoc1}} as follows:
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[3]{}\mathbf{Lemma}\;\Varid{assoc1}\hspace*{0mm}`(\Conid{SemiGroup}\;\Conid{A})\;\Varid{x}\;\Varid{y}\;\Varid{z:}\;\Varid{x}\;\Varid{\&}\;\Varid{y}\;\Varid{\&}\;\Varid{z}\;\mathrel{=}\;\Varid{z}\;\Varid{\&}\;\Varid{y}\;\Varid{\&}\;\Varid{x.}{}\<[E]%
+With it, we can write \ensuremath{\Varid{perm}} as follows:
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[3]{}\mathbf{Lemma}\;\Varid{perm}\hspace*{0mm}`(\Conid{SemiGroup}\;\Conid{A})\;\Varid{x}\;\Varid{y}\;\Varid{z:}\;\Varid{x}\;\Varid{\&}\;\Varid{y}\;\Varid{\&}\;\Varid{z}\;\mathrel{=}\;\Varid{z}\;\Varid{\&}\;\Varid{y}\;\Varid{\&}\;\Varid{x.}{}\<[E]%
 \ColumnHook
@@
 
-\section{Subset types}
-BEGIN SPECULATIVE
-
-A record type is \marginpar{mostly?} syntactic sugar for a $\Sigma$-type. In the presence of a
-squash operator, \ensuremath{[\mskip1.5mu \mskip1.5mu]\;\Conid{:Type->Prop}}, we may define a subset type as \ensuremath{\lambda \;\hspace*{0mm}\{\Conid{A\char92 mid}\;\Conid{B\char92 }\mskip1.5mu\}\;\Varid{=Sig}\;\Conid{A.}\;[\mskip1.5mu \Conid{B}\mskip1.5mu]}.
-This is
-pursued by Barras~\cite{Barras:subset,Werner}. The program machinery fakes subset types in the
-present Coq implementation. Proof irrelevanve in Coq would allow a faithful
-implementation.
-
-Squash is a monad and allows many constructions: subsets, unions, images, heterogeneous equality,
-\ldots.
-
-Question: Would the meta-theory simplify if we would just add a conversion rule $x\triangleright
-\epsilon$ for $x:Prop$? We could add a term constructor $[a]$ which will have type $[A]$. There
-seems to be no reason to connect this to \ensuremath{\Conid{Prop}} or impredicativity.
-
-END SPECULATIVE
-
-We add another principle:
-\begin{itemize}\setcounter{enumi}{3}
- \item Working hypothesis: `every record can be reordered as a subset'. Consequently, we
-rewrite this subset \ensuremath{\lambda \;\hspace*{0mm}\{\Conid{A\char92 mid}\;\Conid{B\char92 }\mskip1.5mu\}} as \ensuremath{\mathbf{Record}\;\Varid{r}\;(\Varid{a:A})\;\Conid{:=}\;\hspace*{0mm}\{\Varid{b:B}\mskip1.5mu\}}.
-\end{itemize}
-
-\section{Algebraic hierarchy using type classes}\label{classes}
-\paragraph{Hierarchy.}\label{hierarchy}
-In our formalization of algebraic structures we use the structure as (implicit) parameters, but
+\section{Putting it all together}\label{classes}
+Combining the ideas in the previous sections we end up with a formalization of algebraic
+structures where we use the structure as (implicit) parameters, but
 pack the proofs. This can be nicely organized by using canonical names. For instance, we define:
@@
 % it.
-We note that \ensuremath{\Varid{op}} does not depend on the proof that \ensuremath{\Varid{e}} is an equivalence. We use Coq's implicit
-quantification (\ensuremath{\hspace*{0mm}`\hspace*{0mm}\{\mskip1.5mu\}}) to avoid having to write all the parameters when \emph{stating} a theorem
-and Coq's maximally inserted implicit arguments to find the parameters when \emph{applying} a
-theorem. Both features are new in Coq and stem from the type class implementation.
+We note that \ensuremath{\Varid{op}} does not depend on the proof that \ensuremath{\Varid{e}} is an equivalence. As explained above we
+use Coq's implicit quantification (\ensuremath{\hspace*{0mm}`\hspace*{0mm}\{\mskip1.5mu\}}) to avoid having to write all the parameters when
+\emph{stating} a theorem and Coq's maximally inserted implicit arguments to find the parameters when
+\emph{applying} a theorem. Both features are new in Coq and stem from the type class implementation.
 
@@
 
-Manifest fields in records allow us to fix a field of a dependent record.
+Manifest fields in records allow us to fix a field of a dependent record. For example, as in the
+example above, \ensuremath{\Conid{Monoid}\;\mathbf{with}\;\Varid{op=mult}}. 
 Luo~\cite{DBLP:conf/types/Luo08} has proposed coersive subtyping to implement manifest fields.
@@
 Since \ensuremath{\Varid{op}} is \emph{instantiated} with \ensuremath{\Varid{mult}}, we need not worry about the
-equality \ensuremath{\Varid{op=mult}}.
+equality \ensuremath{\Varid{op=mult}}. This seems to be a general phenomenon. For instance, instead of searching
+for a \ensuremath{\Conid{Monoid}\;\mathbf{with}\;\Varid{op=mult}} we would use type class resolution to find an instance of \ensuremath{\Conid{Monoid}\;\Varid{mult}}.
+We have thus addressed the problem of multiple inheritance.
 
 \section{Universal algebra}\label{univ}
-After having implemented concrete algebraic structures, we generalized some of our constructions
+After having implemented concrete algebraic structures, we generalize some of our constructions
 by implementing many-sorted universal algebra. A formalization of universal algebra
-already exists in Coq~\cite{DBLP:conf/tphol/Capretta99}. It has recently been extended with more
-categorical results~\cite{dominguez2008formalizing}. We have conveniently redeveloped most of this
-using type classes. We defined equational theories and their varieties:
-the categories of their models. A quote function helps to translate a statement for an instance of
-a type class to a statement in the equational logic.
-
-\subsection{Category theory}\label{cats}
-Using type classes, we provide a basic library for category theory: categories, functors,
-adjunctions, monads,\ldots We follow the usual type theoretical definition of a\marginpar{Use
-notation for comp} category~\cite{saibi1995constructive}
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{4}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{7}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\mathbf{Class}\;\Conid{Category}\;\Conid{O}\hspace*{0mm}`\hspace*{0mm}\{\Varid{!Arrows}\;\Conid{O}\mskip1.5mu\}\hspace*{0mm}`\hspace*{0mm}\{\forall\;\Varid{x}\;\Varid{y:}\;\Conid{O},\Conid{Equiv}\;(\Varid{x}\;\longrightarrow\;\Varid{y})\mskip1.5mu\}{}\<[E]%
-\\
-\>[B]{}\hsindent{4}{}\<[4]%
-\>[4]{}\hspace*{0mm}`\hspace*{0mm}\{\Varid{!CatId}\;\Conid{O}\mskip1.5mu\}\hspace*{0mm}`\hspace*{0mm}\{\Varid{!CatComp}\;\Conid{O}\mskip1.5mu\}\;\mathbin{:}\;\Conid{Prop}\;\Conid{:=}{}\<[E]%
-\\
-\>[B]{}\hsindent{3}{}\<[3]%
-\>[3]{}\hspace*{0mm}\{\Varid{arrow\char95 equiv:>}\;\forall\;\Varid{x}\;\Varid{y:}\;\Conid{O},\Conid{Setoid}\;(\Varid{x}\;\longrightarrow\;\Varid{y}){}\<[E]%
-\\
-\>[B]{}\hsindent{3}{}\<[3]%
-\>[3]{};\Varid{comp\char95 proper:>}\;\forall\;\Varid{x}\;\Varid{y}\;\Varid{z:}\;\Conid{O},{}\<[E]%
-\\
-\>[3]{}\hsindent{4}{}\<[7]%
-\>[7]{}\Conid{Proper}\;(\Varid{equiv}\;\Rightarrow \;\Varid{equiv}\;\Rightarrow \;\Varid{equiv})\;\Varid{\%signature}\;(\Varid{@comp}\;\anonymous \;\anonymous \;\anonymous \;\Varid{x}\;\Varid{y}\;\Varid{z}){}\<[E]%
-\\
-\>[B]{}\hsindent{3}{}\<[3]%
-\>[3]{};\Varid{comp\char95 assoc:}\;\forall\;(\Varid{w}\;\Varid{x}\;\Varid{y}\;\Varid{z:}\;\Conid{O})\;(\Varid{a:}\;\Varid{w}\;\longrightarrow\;\Varid{x})\;(\Varid{b:}\;\Varid{x}\;\longrightarrow\;\Varid{y})\;(\Varid{c:}\;\Varid{y}\;\longrightarrow\;\Varid{z}),{}\<[E]%
-\\
-\>[3]{}\hsindent{4}{}\<[7]%
-\>[7]{}\Varid{comp}\;\Varid{c}\;(\Varid{comp}\;\Varid{b}\;\Varid{a})\;= \;\Varid{comp}\;(\Varid{comp}\;\Varid{c}\;\Varid{b})\;\Varid{a}{}\<[E]%
-\\
-\>[B]{}\hsindent{3}{}\<[3]%
-\>[3]{};\Varid{id\char95 l:}\;\forall\;(\Varid{x}\;\Varid{y:}\;\Conid{O})\;(\Varid{a:}\;\Varid{y}\;\longrightarrow\;\Varid{x}),\Varid{comp}\;\Varid{cat\char95 id}\;\Varid{a}\;= \;\Varid{a}{}\<[E]%
-\\
-\>[B]{}\hsindent{3}{}\<[3]%
-\>[3]{};\Varid{id\char95 r:}\;\forall\;(\Varid{x}\;\Varid{y:}\;\Conid{O})\;(\Varid{a:}\;\Varid{x}\;\longrightarrow\;\Varid{y}),\Varid{comp}\;\Varid{a}\;\Varid{cat\char95 id}\;= \;\Varid{a}\mskip1.5mu\}\;.{}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-This is based on the 2-categorical idea that we have equality on arrows, but not on objects.
-Similarly, we have equality on natural transformations, but not on functors.
-
-For instance natural transformations are formalized as: \ensuremath{\circ}
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{6}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[3]{}\Conid{Context}\hspace*{0mm}`\hspace*{0mm}\{\Conid{Category}\;\Conid{C}\mskip1.5mu\}\hspace*{0mm}`\hspace*{0mm}\{\Conid{Category}\;\Conid{D}\mskip1.5mu\}\;(\Conid{F}\;\Conid{G:}\;\Conid{C}\;\to \;\Conid{D}){}\<[E]%
-\\
-\>[3]{}\hsindent{3}{}\<[6]%
-\>[6]{}\hspace*{0mm}`\hspace*{0mm}\{\Varid{!Functor}\;\Conid{F}\;\Conid{Fa}\mskip1.5mu\}\hspace*{0mm}`\hspace*{0mm}\{\Varid{!Functor}\;\Conid{G}\;\Conid{Ga}\mskip1.5mu\}\;.{}\<[E]%
-\\[\blanklineskip]%
-\>[3]{}\mathbf{Class}\;\Conid{NaturalTransformation}\;(\Conid{η:}\;\forall\;\Varid{c:}\;\Conid{C},\Conid{F}\;\Varid{c}\;\longrightarrow\;\Conid{G}\;\Varid{c})\;\mathbin{:}\;\Conid{Prop}\;\Conid{:=}\;{}\<[E]%
-\\
-\>[3]{}\hsindent{2}{}\<[5]%
-\>[5]{}\Varid{natural:}\;\forall\;(\Varid{x}\;\Varid{y:}\;\Conid{C})\;(\Varid{f:}\;\Varid{x}\;\longrightarrow\;\Varid{y}),\Conid{η}\;\Varid{y}\;\circ\;\Varid{fmap}\;\Conid{F}\;\Varid{f}\;= \;\Varid{fmap}\;\Conid{G}\;\Varid{f}\;\circ\;\Conid{η}\;\Varid{x.}{}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-
-The notation \ensuremath{\circ} denotes function composition.
-Here \ensuremath{\Varid{fmap}\;\Conid{F}} explicitly refers to the arrow part of the functor \ensuremath{\Conid{F}}.\\
-In mathematics one would apply $F$ to both objects and arrows ($F x$, $F f$). However, this does
-not seem to fit with type theory, hence we follow the customs in functional programming.
-
-The proofs are short and follow the textbook easily.\marginpar{really?}
-
-We provide the category of setoids, the category of categories, the dual of a category, products,
-and hence co-products, the category of $\Sigma$-algebras.
-
-We provide the construction of the term-algebra for a signature $\Sigma$ and show that it is
-initial in the category of $\Sigma$-algebras. We proceed to prove that this provides a left-adjoint
-to the forgetful functor. The resulting expression monad will be used below.
-
-\subsection{Universal algebra continued}\label{univ2}
+already exists in Coq~\cite{DBLP:conf/tphol/Capretta99} and has recently been extended with
+more categorical results~\cite{dominguez2008formalizing}. We have conveniently redeveloped most of
+this using type classes. We define equational theories and their varieties of algebras, i.e.\ their
+finitary algebraic category of models. A quote function helps to translate a statement for an
+instance of
+a type class to a statement in the equational logic. We will come back to this in
+Section~\ref{quote}.
+
+% \subsection{Universal algebra continued}\label{univ2}
 
@@
 theorem.~\marginpar{Description of the work.}
-%perhaps the soundness theorem for equational logic.
-
-In order to prove the first homophism theorem, we define a congruence:
+Discuss: the duplication of code which may have been avoided by universe
+polymorphism for \ensuremath{\mathbf{Fixpoint}}s.
+%perhaps the soundness theorem for equational logic?
+
+In order to prove the first homophism theorem, we define a congruence.
+The compatibility of the relation \ensuremath{\Varid{e}} can be stated in two provably  equivalent ways.
+Directly, \ensuremath{\Varid{eAlgebra}\;\Conid{:=}\;\Varid{@Algebra}\;\Varid{et}\;\Varid{v}\;\Varid{e}\;\anonymous } or  indirectly, by stating that the relation
+as a set of pairs is a subalgebra in the product algebra.
 %theory/ua_congruence.v
+% \begin{code}
+%  eSub := @ua_subalgebraT.ClosedSubset et vv _ _ Q.
+% \end{code}
+Subsequently, we choose for former as a definition:
 \begin{hscode}\SaveRestoreHook
@@
 \end{hscode}\resethooks
-We note the use of \ensuremath{\Conid{Proper}}.
+A congruence relation is coarser than the setoid relation, as indicated by the use of \ensuremath{\Conid{Proper}}
+above. Since our theory is founded on a congruence, the initial setoid equality, the development of 
+congruences and quotients benefits from this.
 
 We define the forgetful functor from the category of $\Sigma$-algebras to the category of sets.
-
 We define the term algebra consisting of closed terms, terms with the empty type as set of
-variables. % extensionally equal closed term???
-We show that the term algebra is initial. One may expect that this could be useful to
-\emph{define}, say, the natural numbers as the initial semiring. However, it seems difficult to
-prove decidability of equality, as it seems to require a normalization procedure.
-
-We define the product of two algebras and show that it is the categorical product.
+variables. The term algebra is initial. One may expect that this could be useful to
+\emph{define}, say, the natural numbers as the term algebra for the theory of semiring. However, it
+seems difficult to prove decidability of equality, which requires a normalization procedure.
+
+The product of two $\Sigma$-algebras is their categorical product.
 
@@
 
-We connect the abstract theory to the concrete theory. Concretely, let $\Sigma$ be the theory
-of monoids. Given a monoid, we can construct the corresponding $\Sigma$-algebra and an
-object in the category of $\Sigma$-algebras. Conversely, given an object in the category, we
+We connect the abstract theory to the concrete theory. Concretely, let $\tau_m$ be the theory
+of monoids. Given a monoid, we can construct the corresponding $\tau_m$-algebra and an
+object in the category of $\\tau$-algebras. Conversely, given an object in the category/variety, we
 construct a monoid.
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\mathbf{Variable}\;\Varid{o:}\;\Conid{Variety}\;\Varid{theory.}{}\<[E]%
+%varieties/monoid.v
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathbf{Variable}\;\Varid{o:}\;\Varid{variety.Object}\;\Varid{theory.}{}\<[E]%
+\\
+\>[B]{}\mathbf{Global}\;\Conid{Instance:}\;\Conid{SemiGroupOp}\;(\Varid{o}\;\Varid{tt})\;\Conid{:=}\;\Varid{algebra\char95 op}\;\Varid{theory}\;\Varid{mult.}{}\<[E]%
+\\
+\>[B]{}\mathbf{Global}\;\Conid{Instance:}\;\Conid{MonoidUnit}\;(\Varid{o}\;\Varid{tt})\;\Conid{:=}\;\Varid{algebra\char95 op}\;\Varid{theory}\;\Varid{one.}{}\<[E]%
 \\
@@
 \end{hscode}\resethooks
-Similarly, for arrows.
-
-%There is a slight problem when defining the forgetful functor from the category of
-%$\Sigma$-algebras to Sets. This problem is caused by our use of the category of all
-%categories\footnote{Perhaps we should have used higher categories for this.}.
-%We conjecture that this problem disappears when Coq's universe polymorphism is extended to
-%definitions.
+In the last line, instance resolution `automatically' finds the operation and the unit specified in
+the lines before.\marginpar{Why this way?}
+
+\subsection{Category theory}\label{cats}
+Using type classes, we provide a basic library for category theory: categories, functors,
+adjunctions, monads,\ldots We follow the usual type theoretical definition of a
+category~\cite{saibi1995constructive}
+
+%interfaces/abstract_algebra.v
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{7}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{22}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{36}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathbf{Class}\;\Conid{Category}\;\Conid{O}\hspace*{0mm}`\hspace*{0mm}\{\Varid{!Arrows}\;\Conid{O}\mskip1.5mu\}\hspace*{0mm}`\hspace*{0mm}\{\Pi\;{}\<[36]%
+\>[36]{}\Varid{x}\;\Varid{y:}\;\Conid{O},\Conid{Equiv}\;(\Varid{x}\;\Varid{⟶}\;\Varid{y})\mskip1.5mu\}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\hspace*{0mm}`\hspace*{0mm}\{\Varid{!CatId}\;\Conid{O}\mskip1.5mu\}\hspace*{0mm}`\hspace*{0mm}\{\Varid{!CatComp}\;\Conid{O}\mskip1.5mu\}\;\mathbin{:}\;\Conid{Prop}\;\Conid{:=}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\hspace*{0mm}\{\Varid{arrow\char95 equiv:>}\;\Pi\;{}\<[22]%
+\>[22]{}\Varid{x}\;\Varid{y:}\;\Conid{O},\Conid{Setoid}\;(\Varid{x}\;\Varid{⟶}\;\Varid{y}){}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{};\Varid{comp\char95 proper:>}\;\Pi\;{}\<[22]%
+\>[22]{}\Varid{x}\;\Varid{y}\;\Varid{z},{}\<[E]%
+\\
+\>[3]{}\hsindent{4}{}\<[7]%
+\>[7]{}\Conid{Proper}\;(\Varid{equiv}\;\Rightarrow \;\Varid{equiv}\;\Rightarrow \;\Varid{equiv})\;\Varid{\%signature}\;(\Varid{@comp}\;\anonymous \;\anonymous \;\anonymous \;\Varid{x}\;\Varid{y}\;\Varid{z}){}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{};\Varid{comp\char95 assoc}\;(\Varid{w}\;\Varid{x}\;\Varid{y}\;\Varid{z:}\;\Conid{O})\;(\Varid{a:}\;\Varid{w}\;\Varid{⟶}\;\Varid{x})\;(\Varid{b:}\;\Varid{x}\;\Varid{⟶}\;\Varid{y})\;(\Varid{c:}\;\Varid{y}\;\Varid{⟶}\;\Varid{z})\;\mathbin{:}\;{}\<[E]%
+\\
+\>[3]{}\hsindent{4}{}\<[7]%
+\>[7]{}\Varid{c}\;\circ\;(\Varid{b}\;\circ\;\Varid{a})\;\mathrel{=}\;(\Varid{c}\;\circ\;\Varid{b})\;\circ\;\Varid{a}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{};\Varid{id\char95 l}\hspace*{0mm}`(\Varid{a:}\;\Varid{x}\;\Varid{⟶}\;\Varid{y})\;\mathbin{:}\;\Varid{cat\char95 id}\;\circ\;\Varid{a}\;\mathrel{=}\;\Varid{a}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{};\Varid{id\char95 r}\hspace*{0mm}`(\Varid{a:}\;\Varid{x}\;\Varid{⟶}\;\Varid{y})\;\mathbin{:}\;\Varid{a}\;\circ\;\Varid{cat\char95 id}\;\mathrel{=}\;\Varid{a}\mskip1.5mu\}\;.{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+Where \ensuremath{\Varid{\%signature}}
+tells Coq how to interpret \ensuremath{\Varid{equiv}}.\marginpar{Where is this defined?}
+This is based on the 2-categorical idea that we have equality on arrows, but not on objects.
+Similarly, we have equality on natural transformations, but not on functors.
+
+%theory/categories.v
+Natural transformations are formalized as:
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{17}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{39}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[3]{}\Conid{Context}\hspace*{0mm}`\hspace*{0mm}\{\Conid{Category}\;\Conid{C}\mskip1.5mu\}\hspace*{0mm}`\hspace*{0mm}\{\Conid{Category}\;\Conid{D}\mskip1.5mu\}{}\<[E]%
+\\
+\>[3]{}\hspace*{0mm}`\hspace*{0mm}\{\Varid{!Functor}\;(\Conid{F:}\;\Conid{C}\;\rightarrow\;\Conid{D})\;\Conid{Fa}\mskip1.5mu\}\hspace*{0mm}`\hspace*{0mm}\{\Varid{!Functor}\;(\Conid{G:}\;\Conid{C}\;\rightarrow\;\Conid{D})\;\Conid{Ga}\mskip1.5mu\}\;.{}\<[E]%
+\\[\blanklineskip]%
+\>[3]{}\mathbf{Class}\;\Conid{NaturalTransformation}\;(\Conid{η:}\;\Pi\;{}\<[39]%
+\>[39]{}\Varid{c},\Conid{F}\;\Varid{c}\;\Varid{⟶}\;\Conid{G}\;\Varid{c})\;\mathbin{:}\;\Conid{Prop}\;\Conid{:=}\;{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Varid{natural:}\;\Pi\;{}\<[17]%
+\>[17]{}(\Varid{x}\;\Varid{y:}\;\Conid{C})\;(\Varid{f:}\;\Varid{x}\;\Varid{⟶}\;\Varid{y}),\Conid{η}\;\Varid{y}\;\circ\;\Varid{fmap}\;\Conid{F}\;\Varid{f}\;\mathrel{=}\;\Varid{fmap}\;\Conid{G}\;\Varid{f}\;\circ\;\Conid{η}\;\Varid{x.}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+
+Here \ensuremath{\Varid{fmap}\;\Conid{F}} explicitly refers to the arrow part of the functor \ensuremath{\Conid{F}}.
+In mathematics one would apply $F$ to both objects and arrows ($F x$, $F f$). However, this does
+not seem to fit with type theory.\marginpar{Explain why.}
+
+The proofs are short and follow the textbook easily.\marginpar{really?}
+
+We have defined the category of setoids, the category of categories, the dual of a category,
+products, and hence co-products, the category of $\tau$-algebras.
+
+We provide the construction of the term-algebra for a type $\tau$ and show that it is
+initial in the category of $\tau$-algebras. We proceed to prove that this provides a left-adjoint
+to the forgetful functor. The resulting expression monad will be used in Section~\ref{quote} below.
+
 \section{Interfaces using category theory}\label{interfaces}\label{modul}
@@
 
-We define a term language.
+We define a term language for monoids:
 \begin{hscode}\SaveRestoreHook
@@
 \end{hscode}\resethooks
-This defines a monad on Type\marginpar{Do this!}.
+This defines a monad on the category of \ensuremath{\Conid{Type}}s with extensional functions.\marginpar{Proof this
+formally}.
 The expression type is parameterized over the set of variable indices. Hence, we diverge
@@
 
-\noindent An expression is only meaningful in the context of a variable assignment:\footnote{nat is
-an Expr-algebra. eval is map vars composed with the algebra map.}
+\noindent An expression is only meaningful in the context of a variable
+assignment:\footnote{Categorically: \ensuremath{\mathbb{N}} is
+an \ensuremath{\Conid{Expr}}-algebra and \ensuremath{\Varid{eval}} is \ensuremath{\Varid{map}\;\Varid{vars}} composed with the algebra map.}
 \begin{hscode}\SaveRestoreHook
@@
 %
-Monads are trees with grafting. As explained by sigfpe~\cite{MonadsGrafting}, the tree monad
+Monads are trees with grafting~\cite{MonadsGrafting}: the tree monad
 $TX:=1+T^2X$, may be seen as the prototypical monad. Indeed, the \ensuremath{\Conid{Expr}} monad, i.e.\ the free
-$\Sigma$-algebra monad, behaves much like the tree monad. We now provide a direct definition of the
-corresponding bind operation.
+\ensuremath{\Conid{Expr}}-algebra monad, behaves much like the tree monad. We now provide a direct definition of the
+corresponding bind operation, the arrows in the Kleisli category.
 
@@
 
-We have shown that \ensuremath{\Conid{Expr}}, i.e.\ \ensuremath{\Varid{bind}} together with \ensuremath{\Conid{Var}}, is a monad on the category of Types
-with extensional functions between them~\marginpar{Really?}.
-
-\noindent Some simple combinators for variable packs:
+% We have shown that |Expr|, i.e.\ |bind| together with |Var|, is a monad on the category of Types
+% with extensional functions between them~\marginpar{Really?}.
+
+\noindent We define some simple combinators for variable packs:
 %
@@
  Gallina terms, representing heaps, for which we implement syntactic lookup with type classes.
-Given a heap and value, \ensuremath{\Conid{Lookup}} instances give the value's index in the heap:\marginpar{A
-canonical structures approach would allow us to do this in Coq?}
+The heap can also be defined explitly, with no essential change in the code.
+Given a heap and value, \ensuremath{\Conid{Lookup}} instances give the value's index in the heap:
+% \marginpar{A
+% canonical structures approach would allow us to do this in Coq?}
 \begin{hscode}\SaveRestoreHook
@@
 One useful operation we need before we get to Quote relates to variables and expression
- evaluation. As its name suggests, \ensuremath{\Varid{map\char95 var}} maps an expression's variable indices.\marginpar{monad}
+ evaluation. As its name suggests, \ensuremath{\Varid{map\char95 var}} maps an expression's variable indices.
+This reindexing function is the the \ensuremath{\Varid{map}} of the \ensuremath{\Conid{Expr}}-monad.
 
@@
 used if Lookup fails. The \ensuremath{\Varid{8}} in the previous example is a random number less than \ensuremath{\Varid{9}} below. A
-similar method is used in Coq's notation mechanism. We strongly recommend a change in
-implementation supporting an agda style precedence relation which is only restricted to be a
-directed acyclic graph~\cite{danielsson2009parsing}.
+similar method is used in Coq's notation mechanism.
+% We recommend a change in
+% implementation supporting an agda style precedence relation which is only restricted to be a
+% directed acyclic graph~\cite{danielsson2009parsing}.
 
@@
 structures to derive the setoid equality on the natural
-numbers in the following example \ensuremath{(\Conid{S}\;\Varid{x})\;= \;(\Conid{S}\;\Varid{y})}. We have conveniently reimplemented this with
-type classes with superior proof terms. Canonical structures give
+numbers in the following example \ensuremath{(\Conid{S}\;\Varid{x})\;= \;(\Conid{S}\;\Varid{y})}. In this case type classes provide superior proof
+terms. Canonical structures give
 \begin{hscode}\SaveRestoreHook
@@
 \end{hscode}\resethooks
-
 which includes an explicit proof that \ensuremath{(\Varid{@eq}\;\mathbb{N})} is an equivalence,
@@
 \end{hscode}\resethooks
-
-\section{Constructive mathematics and proof irrelevance}\label{explicit}
+\marginpar{distributivity?}
+
+\section{Subset types and proof irrelevance}
+We have mentioned proof irrelevance a number of times before. It is an important theme in our
+approach. The \ensuremath{\mathbf{Program}} machinery allows one to conveniently write programs with Hoare-style
+pre-
+and post-conditions. I.e.\ functions $f: \{ A \mid P \} \to \{ B \mid Q \}$. Both side conditions
+are intended to be proof irrelevant, they are in \ensuremath{\Conid{Prop}}. Presently, Coq does not support such subset
+types, thus forcing the user to manually prove that \ensuremath{\Varid{f}} does not depend on the proof of \ensuremath{\Conid{P}}.
+The addition of proof irrelevance and subset types to Coq is pursued by
+Barras~\cite{Barras:subset,Werner}.
+
+% Squash is a monad and allows many constructions: subsets, unions, images, heterogeneous equality,
+% \ldots.
+
+We add another principle:
+\begin{enumerate}\setcounter{enumi}{3}
+ \item Working hypothesis: every record can be reordered in such a way that all the non-informative
+fields come last.
+\end{enumerate}
+
+Pollack~\cite{pollack2000dependently} introduces record types with telescopes. 
+Given a telescope\[
+T=[x_1:A_1][x_2:A_2(x_1)]\cdots[x_k:A_k(x_1,\ldots,x_{k_1})]
+\]
+There is an inductively defined type $\Sigma T$, with a single constructor, given by the following 
+formation and introduction rule in pseudo-code for $k=2$:
+\begin{align*}
+ A_1 :& Type\\
+ A_2 :& \Pi A_1,Type\\
+\cline{1-2}
+\Sigma T :& Type
+\end{align*}
+
+\begin{align*}
+ \Sigma T:Type\quad a_1 &:A_1 \quad a_2 : A_2(a_1)\\
+\cline{1-3}
+(a_1,a_2) &:\Sigma T
+\end{align*}
+
+For our methodology to work we have to be able to transform the occuring record into a telescopic
+subset $\{ A \mid  B\}$. We moreover conjecture that the terms in \ensuremath{\Conid{B}} only depend on the
+variables in \ensuremath{\Conid{A}}, but not in \ensuremath{\Conid{B}} itself. However, we do not need this.
+
+\subsection{Propositions as [ ]-Types}\label{explicit}
 As originally concieved~\cite{ITT,CMCP}, Propositions are Types and we can extract information from
-them. In practice, it is convenient to assume that all proofs of a proposition are equal, the
+them. In practice, it is convenient to assume that all proofs of a proposition are equal: the
 principle of proof irrelevance. These two views clearly contradict eachother. In Coq, a middle way
 was choosen, we are forbidden to extract information from proofs, but proofs are not identified.
-Here we need to pay, but do not seem to get anything in return.%
-\footnote{To be precise, what we get in return is the possibility to extract precisely the program
-we want.}
-There are various proposals to add the principle of proof irrelevance to Coq, perhaps a
-promissing one is the Calculus of implicit
-constructions~\cite{miquel2001implicit,barras2008implicit}. Such a change would fit naturally with
-the Hoare-style separation between programs and proofs which is used by the \ensuremath{\mathbf{Program}} machinery.
-
-On the other hand, we found instances where we needed proof \emph{relevance}. In this case, it lead
-to a duplication of code which may have been avoided by universe polymorphism for \ensuremath{\mathbf{Fixpoint}}s. In
-practice, manual dead-code analysis seems useful. Making Harrop formulas proof irrelevant seems to
-be a good first approximation~\cite{lcf:spi:03}.
+This allows us to extract the program we want, but we want more:
+We would like to use Coq as a dependently typed programming language. To do so,
+one would like to add the principle of proof irrelevance to Coq. A promissing proposal is the
+Calculus of implicit constructions with implicit
+$\Sigma$-types~\cite{miquel2001implicit,barras2008implicit,barrasSigma}. Such a change would fit
+naturally with the Hoare-style separation between programs and proofs which is used by the \ensuremath{\mathbf{Program}}
+machinery.
+
+One may be tempted to propose Propositions as [ ]-types~\cite{awodey2004propositions}. Here the
+construction $[P]:=\{\top\mid p:P\}$ squashes all the information about how we proved $P$. However,
+there are places where we needed proof \emph{relevance}; e.g.\ in the proof of the first isomorphism
+theorem above.In practice, manual dead-code analysis seems useful. Making Harrop formulas proof
+irrelevant seems to be a good first approximation~\cite{lcf:spi:03}. However, a further refinement
+of Propositions as types seems necessary.
 

### 2010-02-26 jar.tex
Small changes

@@
 We abstract from concrete algebra and formalize many-sorted universal algebra, including a
-construction of
-free $\tau$-algebras; see Section~\ref{univ}. We are also concerned with efficiency.
-In fact, our aim is to use the present development as a new basis for Bishop's program in type
-theory~\cite{typesreal-article}.\marginpar{More}
+construction of free $\tau$-algebras; see Section~\ref{univ}.
+
+% We are also concerned with efficiency.
+% In fact, our aim is to use the present development as a new basis for Bishop's program in type
+% theory~\cite{typesreal-article}.\marginpar{More}
 
@@
 system~\cite{Setoid-rewrite}. Functions respecting these equivalence relations are called |Proper|.
-As we will see in Section~\ref{univ2} below this functionality can also be used to work with
+As we will see in Section~\ref{univ} below this functionality can also be used to work with
 quotient algebras.
 
-Future versions of Coq are planned to support proof irrelevance: every |p,q:P:Prop| are
-convertible; see section~\cite{explit}.
+Coq has two kinds of sorts |Prop| for propostions and |Type| for data types. Coq
+is scheduled to support proof irrelevance: every |p,q:P:Prop| are
+convertible in the future; see section~\ref{PI}.
 
-\subsection{Type classes}
-Implementation in Ltac, Notation etc\\
-Coercions\\
-Unification hints
+% \subsection{Type classes}
+% Implementation in Ltac, Notation etc\\
+% Coercions\\
+% Unification hints
 
@@
 whereas this is impossible for the latter.}
-we will return to this issue in Section~\ref{explicit}.
+we will return to this issue in Section~\ref{PI}.
 We can now define |EquivalenceRelation| without awkward equalities:
@@
 % \setcounter{enumi}{3}
-
-This list will grow in subsequent sections.
 
@@
 more categorical results~\cite{dominguez2008formalizing}. We have conveniently redeveloped most of
-this using type classes. We define equational theories and their varieties of algebras, i.e.\ their
-finitary algebraic category of models. A quote function helps to translate a statement for an
-instance of
-a type class to a statement in the equational logic. We will come back to this in
-Section~\ref{quote}.
+this using type classes. We define equational theories and their category of algebras. A quote
+function helps to translate a statement for an instance of a type class to a statement in the
+equational logic. We will come back to this in Section~\ref{quote}.
 
@@
 Discuss: the duplication of code which may have been avoided by universe
-polymorphism for |Fixpoint|s.
+polymorphism for |Fixpoint|s.\marginpar{universe polymorphism}
 %perhaps the soundness theorem for equational logic?
@@
 In order to prove the first homophism theorem, we define a congruence.
-The compatibility of the relation |e| can be stated in two provably  equivalent ways.
-Directly, |eAlgebra := @Algebra et v e _| or  indirectly, by stating that the relation
-as a set of pairs is a subalgebra in the product algebra.
-%theory/ua_congruence.v
-% \begin{code}
-%  eSub := @ua_subalgebraT.ClosedSubset et vv _ _ Q.
-% \end{code}
-Subsequently, we choose for former as a definition:
 \begin{code}
@@
 \end{code}
+Alternatively, we could have state the compatibility of the relation |e| by stating that the
+relation as a set of pairs is a subalgebra in the product algebra. We prove that both approaches
+are equivalent.
+
 A congruence relation is coarser than the setoid relation, as indicated by the use of |Proper|
 above. Since our theory is founded on a congruence, the initial setoid equality, the development of 
-congruences and quotients benefits from this.
+congruences and quotients benefits from this.\marginpar{how}
 
@@
 of monoids. Given a monoid, we can construct the corresponding $\tau_m$-algebra and an
-object in the category of $\\tau$-algebras. Conversely, given an object in the category/variety, we
-construct a monoid.
+object in the category of $\tau$-algebras. Conversely, given an object in the category/variety, we
+construct an instance of the |Monoid| type class.\marginpar{Update code}
 %varieties/monoid.v
@@
 adjunctions, monads,\ldots We follow the usual type theoretical definition of a
-category~\cite{saibi1995constructive}
+category~\cite{saibi1995constructive}.\marginpar{Update code}
 
@@
 \end{code}
-Where |%signature|
-tells Coq how to interpret |equiv|.\marginpar{Where is this defined?}
 This is based on the 2-categorical idea that we have equality on arrows, but not on objects.
@@
 We have defined the category of setoids, the category of categories, the dual of a category,
-products, and hence co-products, the category of $\tau$-algebras.
+products, and hence co-products and the category of $\tau$-algebras, for a given type $\tau$.
 
@@
 Unification hints~\cite{Hints} allow one to semi-automatically construct a quote
-function\footnote{Gonthier provides similar functionality using canonical structures}.
-Unification hints are absent in Coq. Fortunately, type classes provide similar functionality as
+function\footnote{Gonthier provides similar functionality using canonical structures}. However,
+this feature is absent from Coq. Fortunately, type classes provide similar functionality as
 we will now show.
@@
 This defines a monad on the category of |Type|s with extensional functions.\marginpar{Proof this
-formally}.
+formally?}
 The expression type is parameterized over the set of variable indices. Hence, we diverge
@@
 
-\section{Subset types and proof irrelevance}
+\section{Subset types and proof irrelevance}\label{PI}
 We have mentioned proof irrelevance a number of times before. It is an important theme in our
 approach. The |Program| machinery allows one to conveniently write programs with Hoare-style
-pre-
-and post-conditions. I.e.\ functions $f: \{ A \mid P \} \to \{ B \mid Q \}$. Both side conditions
+pre- and post-conditions. I.e.\ functions $f: \{ A \mid P \} \to \{ B \mid Q \}$. Both side
+conditions
 are intended to be proof irrelevant, they are in |Prop|. Presently, Coq does not support such subset
@@
 Barras~\cite{Barras:subset,Werner}. 
-
 % Squash is a monad and allows many constructions: subsets, unions, images, heterogeneous equality,
@@
 
-We add another principle:
-\begin{enumerate}\setcounter{enumi}{3}
- \item Working hypothesis: every record can be reordered in such a way that all the non-informative
-fields come last.
-\end{enumerate}
-
 Pollack~\cite{pollack2000dependently} introduces record types with telescopes. 
 Given a telescope\[
-T=[x_1:A_1][x_2:A_2(x_1)]\cdots[x_k:A_k(x_1,\ldots,x_{k_1})]
+T=[x_1:A_1][x_2:A_2(x_1)]\cdots[x_k:A_k(x_1,\ldots,x_k)].
 \]
@@
 
-\subsection{Propositions as [ ]-Types}\label{explicit}
+\subsection{Propositions as [ ]-Types}
 As originally concieved~\cite{ITT,CMCP}, Propositions are Types and we can extract information from
@@
 One may be tempted to propose Propositions as [ ]-types~\cite{awodey2004propositions}. Here the
-construction $[P]:=\{\top\mid p:P\}$ squashes all the information about how we proved $P$. However,
+construction $[P]:=\{\top\mid p:P\}$ squashes all the information about how we proved $P$. In Coq,
+the [ ]-types are isomorphic to the type of sort |Prop|. However,
 there are places where we needed proof \emph{relevance}; e.g.\ in the proof of the first isomorphism

### 2010-03-19 roughdiamond.tex
Referee revisions

@@
-\documentclass[a4paper,10pt]{llncs}
+\documentclass{llncs}
 \usepackage{url}
@@
 records~\cite{DBLP:journals/jsc/GeuversPWZ02} (a.k.a.\ telescopes), packed classes~\cite{Packed} and
-occasionally modules. Although the latter option is quickly discarded%
+occasionally modules. Although the latter option is quickly discarded
 \footnote{Modules are not first class, but currently they seem to be superior for extraction
@@
 
-In this fashion we implement\footnote{The sources are available at:~\url{http://www.eelis.net/research/math-classes/}}%
+In this fashion we implement\footnote{The sources are available
+at:~\url{http://www.eelis.net/research/math-classes/}}
  algebraic theories as type classes, not instances, thus addressing the
@@
 We abstract from concrete algebra and formalize many-sorted universal algebra, including a construction of
-free $\Sigma$-algebras; Section~\ref{univ}. Finally, we are concerned with efficiency. So, we would like to replace data structures with more efficient ones. For instance, 
-Coq provides unary, binary and machine numbers. Hence, we use ad-hoc polymorphism with type classes to
-develop a unified library for them; see Section~\ref{modul}. The interfaces for $\N$, $\Z,\Q$ are the initial semiring, the initial ring and a field of fractions; see Section~\ref{interfaces}.
+free $\Sigma$-algebras; Section~\ref{univ}. We intend to use the current work as a basis for
+a library for computational analysis.  Thus we want to be able to replace data structures
+with more efficient ones. For instance, Coq provides unary, binary and machine numbers. We
+use ad-hoc polymorphism with type classes to develop a unified library for them. The interfaces for
+$\N$, $\Z,\Q$ are the initial semiring, the initial ring and a field of fractions; see
+Section~\ref{interfaces}.
 
@@
 using type classes, moreover we have added a categorical framework. For instance, we show that the
-term-algebra for a signature is initial. We defined equational theories and their varieties:
-the categories of their models. A quote function helps to translate a statement for an instance of
-a type class to a statement in the equational logic. We are currently pursuing quoting with
-unification~\cite{Hints} using type classes.
+term-algebra for a signature is initial. We defined equational theories and their categories of
+their models. A quote function helps to translate a statement for an instance of
+a type class to a statement in the equational logic. We have a prototype automatic quoting mechanism
+using type classes similar to that using unification hints~\cite{Hints} .
 
@@
 \end{code}
-\vspace*{-5ex}
+
 \section{Conclusions}
@@
 
-Packed classes use canonical structures for the algebraic hierarchy. Both canonical structures and
+Packed classes use canonical structures for the algebraic hierarchy and a uniform notation. Both
+canonical structures and
 type classes may be seen as instances of hints in unification~\cite{Hints}. Some uses of canonical
@@
 structures to derive the setoid equality on the natural
-numbers in the following example |(S x)==(S y)|. We have conveniently reimplemented this with
-type classes with superior proof terms. Canonical structures give 
+numbers in the following example |(S x)==(S y)|. We have reimplemented this with
+type classes with similar proof terms. Canonical structures give
 \begin{code}
@@
 
-According to |coqwc|, our development consists of 3103 lines of specifications and 707 lines of
-proofs. %targz
+According to |coqwc|, our development consists of 4479 lines of specifications and 722 lines of
+proofs.
 
-% \section{Other systems}
-% We will provide a fuller comparison with other systems at a later stage.
-% We specifically mention recent work on unification hints in Matita~\cite{Hints}. They however, do
-% not address the challenge of proofs in mathematical structures.
 \paragraph{Acknowledgements.}
 We would like to thank Matthieu Sozeau for discussions and quickly solving numerous small bugs and
-feature requests. %Jeremy,Thierry, Claudio, Georges?, Wouter
-%\enlargethispage{5mm}
+feature requests. 
 \bibliographystyle{plain}

### 2010-03-19 compile.sh

@@ -4,10 +4,10 @@ lhs2TeX --agda< $1.tex > processed.tex
 cp processed.tex tmp.tex
 
 set -e
-xelatex tmp.tex
+pdflatex tmp.tex
 bibtex tmp
-xelatex tmp.tex
+pdflatex tmp.tex
 bibtex tmp
-xelatex tmp.tex
+pdflatex tmp.tex
 cat tmp.pdf > $1.pdf
 rm tmp.*

### 2010-03-19 jar.tex
Fix

@@
-\documentclass[a4paper,10pt, runningheads]{llncs}
+\documentclass[a4paper,10pt,runningheads]{llncs}
 \usepackage{url}
 \usepackage{bbm}
-\usepackage{ucs}
 \usepackage{amsmath}
-% \usepackage[utf8x,utf8]{inputenc}
+\usepackage[utf8x]{inputenx}
 \newcommand{\N}{\ensuremath{\mathbbm{N}}}
@@
 \newcommand{\Q}{\ensuremath{\mathbbm{Q}}}
+%include polycode.fmt
 %include coq2tex.tex
@@
 \divide\belowdisplayskip 4
+\newcommand{\textPi}{\ensuremath\Pi}
+\newcommand{\textSigma}{\ensuremath\Sigma}
+\newcommand{\texttau}{\ensuremath\tau}
+% \newcommand{\textforall}{\ensuremath\forall}
 \begin{document}
@@
 \begin{abstract}
-We propose generic design patterns to organize algebraic structures using type classes inside the
+We\footnote{\today} propose generic design patterns to organize algebraic structures using
+type classes inside the
 Coq system. We support multiple inheritance, sharing of notations and theories, and automated
@@
 The Coq proof assistant is based on the calculus of inductive
-constructions~\cite{CoquandHuet,CoquandPaulin}.\marginpar{More}
-This type theory lacks quotient types, as it would make type checking undecidable. Instead it
+constructions~\cite{CoquandHuet,CoquandPaulin}, a dependent type theory with (co)inductive types.
+We will not elaborate on this, but instead refer to~\cite{Coq,BC04}.
+
+The Coq type theory lacks quotient
+types, as it would make type checking undecidable. Instead it
 supports setoids (aka Bishop sets), a pair of a type and an equivalence
-relation~\cite{Bishop67,Hofmann,Capretta}. Recently, improved support has been added to the
-system~\cite{Setoid-rewrite}. Functions respecting these equivalence relations are called |Proper|.
+relation~\cite{Bishop67,Hofmann,Capretta}. Recently, improved support for setoid has been added to
+the system~\cite{Setoid-rewrite}. Functions respecting these equivalence relations are called
+|Proper|.
 As we will see in Section~\ref{univ} below this functionality can also be used to work with
@@
 Coq has two kinds of sorts |Prop| for propostions and |Type| for data types. Coq
-is scheduled to support proof irrelevance: every |p,q:P:Prop| are
-convertible in the future; see section~\ref{PI}.
+is scheduled to support proof irrelevance, every |p,q:P:Prop| are
+convertible, in the future; see section~\ref{PI}.
 
@@
 \section{To bunde or not to bundle? (Or: What to bundle?)}
+We provide a rationale for our approach. We begin gently.
 
@@
     { rr_carrier: Type
-    ; rr_rel: rr_carrier → rr_carrier → Prop
+    ; rr_rel: relation rr_carrier
     ; rr_proof: Π x, rr_rel x x }.
@@
 
-With this approach, we run into problems when we try to define |EquivalenceRelation| in terms of
+With this approach, we run into problems when defining |EquivalenceRelation| in terms of
 |ReflexiveRelation| and its two siblings, because when posit three objects of type
 |ReflexiveRelation|, |TransitiveRelation|, and |SymmetricRelation|, respectively, we have to
-add equalities expressing that they talk about the same carrier and relation, and these equalities
-are a nightmare to work with (because the three carriers and relations won't be convertible).
-Technically, this leads to the research in manifest fields, the \emph{with} construction, to which
+add equalities expressing that they talk about the same carrier and relation. These equalities
+cause difficulties, because the three carriers and relations won't be convertible.
+Technically, this leads to the research in manifest fields, i.e.\ the \emph{with} construction, to
+which
 we return in Section~\ref{manifest}.
@@
   Record ReflexiveRelation (carrier: Type): Type :=
-    { rr_rel: rr_carrier → rr_carrier → Prop
+    { rr_rel: relation rr_carrier
     ; rr_proof: Π x, rr_rel x x }.
@@
   Record ReflexiveRelation (carrier: Type)
-                           (rel: carrier → carrier → Prop): Prop :=
+                           (rel: relation carrier): Prop :=
     { rr_proof: Π x, rr_rel x x }.
@@
   Definition ReflexiveRelation (carrier: Type)
-                               (rel: carrier → carrier → Prop): Prop :=
+                               (rel: relation carrier): Prop :=
     Π x, rr_rel x x.
@@
 \footnote{Coq distinguishes between transparent and opague definitions. The former can be unfolded,
-whereas this is impossible for the latter.}
+whereas this is impossible for the latter. Since we never need to distinguish between two proofs of
+a propositions, Propositions can be Opague.}
 we will return to this issue in Section~\ref{PI}.
@@
     { er_carrier: Type
-    ; er_rel: er_carrier → er_carrier → Prop
+    ; er_rel: relation er_carrier
     ; er_refl: ReflexiveRelation er_carrier er_rel
@@
 Record EquivalenceRelation (er_carrier: Type)
-                             (er_rel: er_carrier → er_carrier → Prop): Prop :=
+                             (er_rel: relation er_carrier): Prop :=
   { er_refl: ReflexiveRelation er_carrier er_rel
@@
     ; mult_comm: Commutative (mon_op mult_monoid)
-    ; mult_0_l: Π x: A, ?
+    ; mult_0_l: Π x: A, H
 \end{code}
-Which equality do we use in the expression ?, |mult_monoid|'s equality or |plus_monoid|'s?
+Do we use |mult_monoid|'s equality or |plus_monoid|'s equality in |H|?
 Again, for this definition to be usable at all, we would need to add an equality expressing that the
@@
 
-Let us now stop and think to what extent this approach suffers from all the problems commonly
-associated with it. In particular, let us imagine what happens to our terms and contexts when we
-want to talk about nested structures such as polynomials over the ring of matrices over a
-given ring; see Section~\ref{interfaces}.
-
 \section{Putting it all together}\label{classes}
@@
 as in the definition of a semiring: a type with two monoid structures on it. We avoid some
-of these problems by explicitly specifying the fields. We emphasize that the semiring properties are automatically derived from the ring
-properties, although the properties of a semiring are not structurally included in the ring
-properties.
+of these problems by explicitly specifying the fields. We emphasize that the semiring properties are
+automatically derived from the ring properties, although the properties of a semiring are not
+structurally included in the ring properties.
+
+Let us now stop and think to what extent this approach suffers from all the problems commonly
+associated with it. In particular, let us imagine what happens to our terms and contexts when we
+want to talk about nested structures such as polynomials over the ring. For a concrete
+representation of the polynomials the context will just contain the context for an abstract ring.
+For abstract reasoning about polynomials the context will grow with abstract operations on
+polynomials. Consequently, the context will grow linearly, as opposed to exponentially.
 
@@
 
-We have developed basic universal algebra~\cite{meinke1993universal} upto the first homomorphism
-theorem.~\marginpar{Description of the work.}
-Discuss: the duplication of code which may have been avoided by universe
-polymorphism for |Fixpoint|s.\marginpar{universe polymorphism}
+We have developed basic multi-sorted universal algebra~\cite{meinke1993universal} upto the first
+homomorphism theorem. More precisely, we provide the following:
+\begin{itemize}
+ \item Definition of a signature Σ. Definition of a type (Σ,E). Σ-algebras and
+(Σ,$E$)-algebras
+ \item A forgetful functor from (Σ,$E$)-algebras via Σ-algebras to Sets
+ \item Products and subalgebras
+ \item The term algebra for a type $\tau$. This is an initial object in the category of
+$\tau$-algebras.
+\end{itemize}
 %perhaps the soundness theorem for equational logic?
+
+Coq provides different sorts. To simplify the discussion, we assume that there are only two sorts:
+|Prop| and |Type|. It occassionally happens that we want to prove a similar theorem for both sorts.
+To avoid code duplication, Coq provides universe polymorphism for inductive types.
+Unfortunately,
+definitions and fixpoints are not (yet?) polymorphic. To avoid code duplication we
+change some Fixpoints into
+Inductives. Concretely, the image $Im f:=\{b:B \mid \exists a. b=(f
+a)\}$ is informative (in |Type|) since we need the witness in the proof of the first homomorphism
+theorem.\marginpar{Does |relation| as a definition cause universe inconsistencies?}
+% theory/ua_subalgebraT.v
+% find . -iname "*.v" |xargs grep "niverse"
 
@@
 \end{code}
-Alternatively, we could have state the compatibility of the relation |e| by stating that the
+Alternatively, we could have stated the compatibility of the relation |e| by stating that the
 relation as a set of pairs is a subalgebra in the product algebra. We prove that both approaches
-are equivalent.
+are equivalent. The former definition seems more natural in type theory. In set theory, a quotient
+is defined as a collection of equivalence classes. In type theory, a quotient is more naturally
+defined by changing the equality, while leaving the carrier unchanged.%
+\footnote{In Bishop's words~\cite[p.12]{Bishop/Bridges:1985}: The axiom of choice is used to extract
+elements from equivalence classes where they should never have been put in the first place.}
+ In this way we profit from
+the infrastructure for setoids: A congruence relation is coarser than the setoid relation, as
+indicated by the use of |Proper| above.
 
-A congruence relation is coarser than the setoid relation, as indicated by the use of |Proper|
-above. Since our theory is founded on a congruence, the initial setoid equality, the development of 
-congruences and quotients benefits from this.\marginpar{how}
-
+%varieties/closed_terms.v
 We define the forgetful functor from the category of $\tau$-algebras to the category of sets.
-We define the term algebra consisting of closed terms, terms with the empty type as set of
-variables. The term algebra is the free $\tau$-algebra and hence\marginpar{?}
-initial. Categorically: we have defined a left-adjoint to the forgetful functor. A formal proof
-connects the categorical and the algebraic definition.
+We define the term algebra consisting of closed terms, i.e.\ terms with the empty type as set of
+variables. The term algebra is the initial $\tau$-algebra. Categorically: we have defined a
+left-adjoint to the forgetful functor.\marginpar{Did we proof this?}
+A formal proof connects the categorical and the algebraic definition.
 One may expect that this term algebra could be useful to
@@
 
-%format ◎ = "\circ"
+\newcommand{\rightarrowtext}{\ensuremath{\rightarrow}}
+\newcommand{\textocirc}{\ensuremath\circ}
 %interfaces/abstract_algebra.v
-\begin{code}
-Class Category O `{!Arrows O} `{Π x y: O, Equiv (x ⟶ y)}
-  `{!CatId O} `{!CatComp O}: Prop :=
-  { arrow_equiv:> Π x y: O, Setoid (x ⟶ y)
-  ; comp_proper:> Π x y z,
-      Proper (equiv ==> equiv ==> equiv)%signature (@comp _ _ _ x y z)
-  ; comp_assoc (w x y z: O) (a: w ⟶ x) (b: x ⟶ y) (c: y ⟶ z):
-      c ◎ (b ◎ a) = (c ◎ b) ◎ a
-  ; id_l `(a: x ⟶ y): cat_id ◎ a = a
-  ; id_r `(a: x ⟶ y): a ◎ cat_id = a }.
-\end{code}
+
+CODE MISSING.
+% \begin{code}
+% Class Category O `{!Arrows O} `{Π x y: O, Equiv (x ⟶ y)}
+%   `{!CatId O} `{!CatComp O}: Prop :=
+%   { arrow_equiv:> Π x y: O, Setoid (x ⟶ y)
+%   ; comp_proper:> Π x y z,
+%       Proper (equiv ==> equiv ==> equiv)%signature (@comp _ _ _ x y z)
+%   ; comp_assoc (w x y z: O) (a: w ⟶ x) (b: x ⟶ y) (c: y ⟶ z):
+%       c ◎ (b ◎ a) = (c ◎ b) ◎ a
+%   ; id_l `(a: x ⟶ y): cat_id ◎ a = a
+%   ; id_r `(a: x ⟶ y): a ◎ cat_id = a }.
+% \end{code}
 This is based on the 2-categorical idea that we have equality on arrows, but not on objects.
@@
 Natural transformations are formalized as:
-\begin{code}
-  Context `{Category C} `{Category D}
-  `{!Functor (F: C → D) Fa} `{!Functor (G: C → D) Ga}.
-
-  Class NaturalTransformation (η: Π c, F c ⟶ G c): Prop :=
-    natural: Π (x y: C) (f: x ⟶ y), η y ◎ fmap F f = fmap G f ◎ η x.
-\end{code}
+CODE MISSING.
+% \begin{code}
+%   Context `{Category C} `{Category D}
+%   `{!Functor (F: C → D) Fa} `{!Functor (G: C → D) Ga}.
+% 
+%   Class NaturalTransformation (η: Π c, F c ⟶ G c): Prop :=
+%     natural: Π (x y: C) (f: x ⟶ y), η y ◎ fmap F f = fmap G f ◎ η x.
+% \end{code}
 
@@
 products, and hence co-products and the category of $\tau$-algebras, for a given type $\tau$.
+
+Universe polymorphism allows us to define the category of categories. However, we need to avoid
+making |Relation| a |Definition|, since |Definitions| are not universe polymorphic.
 
@@
 
+Similarly, we use type classes to flexibly combine two representations of a dense set of the
+reals.
+Let |Q| be the rationals and |D| be the dyadic rationals, i.e.\ rationals where the denominator is
+a power of 2. Let |inj:D>->Q| be a coercion. Then we want to define heterogeneous multiplication
+as follows:
+
 %There is a slight problem when defining the forgetful functor from the category of
@@
 Unification hints~\cite{Hints} allow one to semi-automatically construct a quote
-function\footnote{Gonthier provides similar functionality using canonical structures}. However,
+function.\footnote{Gonthier provides similar functionality using canonical structures.} However,
 this feature is absent from Coq. Fortunately, type classes provide similar functionality as
@@
 subset $\{ A \mid  B\}$. We moreover conjecture that the terms in |B| only depend on the
-variables in |A|, but not in |B| itself. However, we do not need this.
+variables in |A|, but not in |B| itself. However, we do not need this.\\
+
+We observed that we were naturally lead to the following methodology: Let φ be a statement in type
+theory, i.e.\ an alternation of Πs and Σs. By the axiom of choice, we first Skolemize to $Σ h. Πa.P$
+and then transfer the variable $h$ to the parameters. For example, consider surjectivity. This is
+skolemized to a surjection with a right inverse, i.e.\ a split epi. We then move the right inverse
+to the parameters.
+
+It is interesting to note that classical logic is conservative for purely universal statements.
+
+The addition of implicit Sigma types to Coq would change many things. Perhaps it would even allow us
+to pack the proofs together with the operations, but it is too early to tell.
 
@@
 good generalization, but are absent from Coq. We encourage their inclusion to the
-Coq system.
+Coq system.\marginpar{for instance what about control in
+proof/instance search?\\ or what about its efficiency?}
 

### 2010-03-19 processed.tex
Fix

--- 2010-02-25 jar.tex
@@
-\documentclass[a4paper,10pt, runningheads]{llncs}
+\documentclass{llncs}
 \usepackage{url}
 \usepackage{bbm}
-\usepackage{ucs}
-% \usepackage[utf8x,utf8]{inputenc}
+\usepackage[utf8x]{inputenc}
 \newcommand{\N}{\ensuremath{\mathbbm{N}}}
@@
 \newcommand{\Q}{\ensuremath{\mathbbm{Q}}}
-%include coq2tex.tex
-\arrayhs%i.e. no page breaks in code blocks
+%% ODER: format ==         = "\mathrel{==}"
+%% ODER: format /=         = "\neq "
+%
+%
+\makeatletter
+\@ifundefined{lhs2tex.lhs2tex.sty.read}%
+  {\@namedef{lhs2tex.lhs2tex.sty.read}{}%
+   \newcommand\SkipToFmtEnd{}%
+   \newcommand\EndFmtInput{}%
+   \long\def\SkipToFmtEnd#1\EndFmtInput{}%
+  }\SkipToFmtEnd
+
+\newcommand\ReadOnlyOnce[1]{\@ifundefined{#1}{\@namedef{#1}{}}\SkipToFmtEnd}
+\usepackage{amstext}
+\usepackage{amssymb}
+\usepackage{stmaryrd}
+\DeclareFontFamily{OT1}{cmtex}{}
+\DeclareFontShape{OT1}{cmtex}{m}{n}
+  {<5><6><7><8>cmtex8
+   <9>cmtex9
+   <10><10.95><12><14.4><17.28><20.74><24.88>cmtex10}{}
+\DeclareFontShape{OT1}{cmtex}{m}{it}
+  {<-> ssub * cmtt/m/it}{}
+\newcommand{\texfamily}{\fontfamily{cmtex}\selectfont}
+\DeclareFontShape{OT1}{cmtt}{bx}{n}
+  {<5><6><7><8>cmtt8
+   <9>cmbtt9
+   <10><10.95><12><14.4><17.28><20.74><24.88>cmbtt10}{}
+\DeclareFontShape{OT1}{cmtex}{bx}{n}
+  {<-> ssub * cmtt/bx/n}{}
+\newcommand{\tex}[1]{\text{\texfamily#1}}	% NEU
+
+\newcommand{\Sp}{\hskip.33334em\relax}
+
+\newcommand{\Conid}[1]{\mathit{#1}}
+\newcommand{\Varid}[1]{\mathit{#1}}
+\newcommand{\anonymous}{\kern0.06em \vbox{\hrule\@width.5em}}
+\newcommand{\plus}{\mathbin{+\!\!\!+}}
+\newcommand{\bind}{\mathbin{>\!\!\!>\mkern-6.7mu=}}
+\newcommand{\rbind}{\mathbin{=\mkern-6.7mu<\!\!\!<}}% suggested by Neil Mitchell
+\newcommand{\sequ}{\mathbin{>\!\!\!>}}
+\renewcommand{\leq}{\leqslant}
+\renewcommand{\geq}{\geqslant}
+\usepackage{polytable}
+
+%mathindent has to be defined
+\@ifundefined{mathindent}%
+  {\newdimen\mathindent\mathindent\leftmargini}%
+  {}%
+
+\def\resethooks{%
+  \global\let\SaveRestoreHook\empty
+  \global\let\ColumnHook\empty}
+\newcommand*{\savecolumns}[1][default]%
+  {\g@addto@macro\SaveRestoreHook{\savecolumns[#1]}}
+\newcommand*{\restorecolumns}[1][default]%
+  {\g@addto@macro\SaveRestoreHook{\restorecolumns[#1]}}
+\newcommand*{\aligncolumn}[2]%
+  {\g@addto@macro\ColumnHook{\column{#1}{#2}}}
+
+\resethooks
+
+\newcommand{\onelinecommentchars}{\quad-{}- }
+\newcommand{\commentbeginchars}{\enskip\{-}
+\newcommand{\commentendchars}{-\}\enskip}
+
+\newcommand{\visiblecomments}{%
+  \let\onelinecomment=\onelinecommentchars
+  \let\commentbegin=\commentbeginchars
+  \let\commentend=\commentendchars}
+
+\newcommand{\invisiblecomments}{%
+  \let\onelinecomment=\empty
+  \let\commentbegin=\empty
+  \let\commentend=\empty}
+
+\visiblecomments
+
+\newlength{\blanklineskip}
+\setlength{\blanklineskip}{0.66084ex}
+
+\newcommand{\hsindent}[1]{\quad}% default is fixed indentation
+\let\hspre\empty
+\let\hspost\empty
+\newcommand{\NB}{\textbf{NB}}
+\newcommand{\Todo}[1]{$\langle$\textbf{To do:}~#1$\rangle$}
+
+\EndFmtInput
+\makeatother
+%
+%
+%
+%
+%
+%
+% This package provides two environments suitable to take the place
+% of hscode, called "plainhscode" and "arrayhscode". 
+%
+% The plain environment surrounds each code block by vertical space,
+% and it uses \abovedisplayskip and \belowdisplayskip to get spacing
+% similar to formulas. Note that if these dimensions are changed,
+% the spacing around displayed math formulas changes as well.
+% All code is indented using \leftskip.
+%
+% Changed 19.08.2004 to reflect changes in colorcode. Should work with
+% CodeGroup.sty.
+%
+\ReadOnlyOnce{polycode.fmt}%
+\makeatletter
+
+\newcommand{\hsnewpar}[1]%
+  {{\parskip=0pt\parindent=0pt\par\vskip #1\noindent}}
+
+% can be used, for instance, to redefine the code size, by setting the
+% command to \small or something alike
+\newcommand{\hscodestyle}{}
+
+% The command \sethscode can be used to switch the code formatting
+% behaviour by mapping the hscode environment in the subst directive
+% to a new LaTeX environment.
+
+\newcommand{\sethscode}[1]%
+  {\expandafter\let\expandafter\hscode\csname #1\endcsname
+   \expandafter\let\expandafter\endhscode\csname end#1\endcsname}
+
+% "compatibility" mode restores the non-polycode.fmt layout.
+
+\newenvironment{compathscode}%
+  {\par\noindent
+   \advance\leftskip\mathindent
+   \hscodestyle
+   \let\\=\@normalcr
+   \(\pboxed}%
+  {\endpboxed\)%
+   \par\noindent
+   \ignorespacesafterend}
+
+\newcommand{\compaths}{\sethscode{compathscode}}
+
+% "plain" mode is the proposed default.
+% It should now work with \centering.
+% This required some changes. The old version
+% is still available for reference as oldplainhscode.
+
+\newenvironment{plainhscode}%
+  {\hsnewpar\abovedisplayskip
+   \advance\leftskip\mathindent
+   \hscodestyle
+   \let\hspre\(\let\hspost\)%
+   \pboxed}%
+  {\endpboxed%
+   \hsnewpar\belowdisplayskip
+   \ignorespacesafterend}
+
+\newenvironment{oldplainhscode}%
+  {\hsnewpar\abovedisplayskip
+   \advance\leftskip\mathindent
+   \hscodestyle
+   \let\\=\@normalcr
+   \(\pboxed}%
+  {\endpboxed\)%
+   \hsnewpar\belowdisplayskip
+   \ignorespacesafterend}
+
+% Here, we make plainhscode the default environment.
+
+\newcommand{\plainhs}{\sethscode{plainhscode}}
+\newcommand{\oldplainhs}{\sethscode{oldplainhscode}}
+\plainhs
+
+% The arrayhscode is like plain, but makes use of polytable's
+% parray environment which disallows page breaks in code blocks.
+
+\newenvironment{arrayhscode}%
+  {\hsnewpar\abovedisplayskip
+   \advance\leftskip\mathindent
+   \hscodestyle
+   \let\\=\@normalcr
+   \(\parray}%
+  {\endparray\)%
+   \hsnewpar\belowdisplayskip
+   \ignorespacesafterend}
+
+\newcommand{\arrayhs}{\sethscode{arrayhscode}}
+
+% The mathhscode environment also makes use of polytable's parray 
+% environment. It is supposed to be used only inside math mode 
+% (I used it to typeset the type rules in my thesis).
+
+\newenvironment{mathhscode}%
+  {\parray}{\endparray}
+
+\newcommand{\mathhs}{\sethscode{mathhscode}}
+
+% texths is similar to mathhs, but works in text mode.
+
+\newenvironment{texthscode}%
+  {\(\parray}{\endparray\)}
+
+\newcommand{\texths}{\sethscode{texthscode}}
+
+% The framed environment places code in a framed box.
+
+\def\codeframewidth{\arrayrulewidth}
+\RequirePackage{calc}
+
+\newenvironment{framedhscode}%
+  {\parskip=\abovedisplayskip\par\noindent
+   \hscodestyle
+   \arrayrulewidth=\codeframewidth
+   \tabular{@{}|p{\linewidth-2\arraycolsep-2\arrayrulewidth-2pt}|@{}}%
+   \hline\framedhslinecorrect\\{-1.5ex}%
+   \let\endoflinesave=\\
+   \let\\=\@normalcr
+   \(\pboxed}%
+  {\endpboxed\)%
+   \framedhslinecorrect\endoflinesave{.5ex}\hline
+   \endtabular
+   \parskip=\belowdisplayskip\par\noindent
+   \ignorespacesafterend}
+
+\newcommand{\framedhslinecorrect}[2]%
+  {#1[#2]}
+
+\newcommand{\framedhs}{\sethscode{framedhscode}}
+
+% The inlinehscode environment is an experimental environment
+% that can be used to typeset displayed code inline.
+
+\newenvironment{inlinehscode}%
+  {\(\def\column##1##2{}%
+   \let\>\undefined\let\<\undefined\let\\\undefined
+   \newcommand\>[1][]{}\newcommand\<[1][]{}\newcommand\\[1][]{}%
+   \def\fromto##1##2##3{##3}%
+   \def\nextline{}}{\) }%
+
+\newcommand{\inlinehs}{\sethscode{inlinehscode}}
+
+% The joincode environment is a separate environment that
+% can be used to surround and thereby connect multiple code
+% blocks.
+
+\newenvironment{joincode}%
+  {\let\orighscode=\hscode
+   \let\origendhscode=\endhscode
+   \def\endhscode{\def\hscode{\endgroup\def\@currenvir{hscode}\\}\begingroup}
+   %\let\SaveRestoreHook=\empty
+   %\let\ColumnHook=\empty
+   %\let\resethooks=\empty
+   \orighscode\def\hscode{\endgroup\def\@currenvir{hscode}}}%
+  {\origendhscode
+   \global\let\hscode=\orighscode
+   \global\let\endhscode=\origendhscode}%
+
+\makeatother
+\EndFmtInput
+%
+
+  % Without these, lhs2tex uses three-line equal signs.
+
+%%%format R = "\mathbb{R}"
+
+\arrayhs %i.e. no page breaks in code blocks
 \divide\abovedisplayskip 4
@@
 \section{Introduction}
-Developing formalized libraries for mathematics requires a detailed analysis of the
-architecture of mathematics~\cite{C-corn,DBLP:conf/types/HaftmannW08}, similar to that of a large
-software project.
+Developing formalized libraries for mathematics requires a detailed analysis of the architecture~\cite{C-corn,DBLP:conf/types/HaftmannW08}, similar to that of a large software project. 
 The development of an algebraic hierarchy is an important aspect. Challenges include: 1. multiple
 inheritance: A ring is a type with two operations. How do we inherit from the two theories of
-monoids? 2. Sharing of notations. 3. Combining syntactic and semantic methods, usually this is
-addressed by reflection.
+monoids? 2. Sharing of notations.
 
-Several solutions have been proposed for the Coq theorem prover~\cite{Coq,BC04}: dependent
+Several solutions have been proposed for the Coq theorem prover%~\cite{Coq,BC04}
+: dependent
 records~\cite{DBLP:journals/jsc/GeuversPWZ02} (a.k.a.\ telescopes), packed classes~\cite{Packed} and
-occasionally modules. Although the latter option is quickly discarded%
-\footnote{Modules are not first class, but currently they seem to be superior for extraction to
-OCaml~\cite{letouzey02}. Modules may also be seen as parametrised sections. They allow management
-of namespaces, notations, hints, etc. An |Import| command imports all the constants from a
-module. Thus allowing the user to type |B| instead of |A.B|. Similar functionality would be useful
-for type classes.}
+occasionally modules. Although the latter option is quickly discarded
+\footnote{Modules are not first class, but currently they seem to be superior for extraction
+purposes%~\cite{letouzey02}%
+.}, considering
 algebraic theories as `interfaces', as opposed to `modules', has been an important
@@
  algebraic theories as type classes, not instances, thus addressing the
-challenges above; see Section~\ref{classes}. Type classes have recently entered
-Coq~\cite{DBLP:conf/tphol/SozeauO08}. In Coq,
+challenges above; see Section~\ref{classes}. Type classes have recently entered Coq~\cite{DBLP:conf/tphol/SozeauO08}. In Coq
 type classes are \emph{first class} citizens based on dependent records. The presence of dependent types makes them even more
@@
 
-We abstract from concrete algebra and formalize many-sorted universal algebra, including a
-construction of
-free $\Sigma$-algebras; see Section~\ref{univ}. We are also concerned with efficiency.
-In fact, our aim is to use the present development as a new basis for Bishop's program in type
-theory~\cite{typesreal-article}.\marginpar{More}
+% In
+% Coq, type classes have advantages over modules\footnote{
+% Modules are sections on steroids: namespace management, notations, hints, etc.
+% The [Import] command imports all the constants from a module. Thus allowing the user to type B
+% instead of A.B
+% }:
+% they are first class and extract~\cite{letouzey02} to Haskell\footnote{
+% Would module like extraction be possible?}.
 
-So, we would like to replace data structures with more efficient ones. Concretely,
-Coq provides unary, binary and machine numbers. We use type classes polymorphism to
-develop a unified library for them; see Section~\ref{modul}.
-The interfaces for $\N$, $\Z,\Q$ are the initial semiring, the initial ring and a field of
-fractions; see Section~\ref{interfaces}. In future work we aim to base our development of the reals
-on an abstract dense set, allowing us to use the efficient dyadic
-rationals~\cite{boldo2009combining} as a base for exact real number computation in
-Coq~\cite{Riemann,Oconnor:real}.
+We abstract from concrete algebra and formalize many-sorted universal algebra, including a construction of
+free $\Sigma$-algebras; Section~\ref{univ}. We intend to use the current work as a basis for
+a library for computational analysis.  Thus we want to be able to replace data structures
+with more efficient ones. For instance, Coq provides unary, binary and machine numbers. We
+use ad-hoc polymorphism with type classes to develop a unified library for them. The interfaces for
+$\N$, $\Z,\Q$ are the initial semiring, the initial ring and a field of fractions; see
+Section~\ref{interfaces}.
 
-The paper is organised as follows: Section~\ref{classes} starts with a development of the concrete
-algebraic hierarchy addressing the issues raised above. Section~\ref{univ} abstracts to universal
-algebra, we develop the basic theory to the point of the first homomorphism theorem.
-Section~\ref{interfaces} provides interfaces to usual datastructures by providing their universal
-properties. To allow symbolic manipulation of semantic objects one needs a quote function. Usually,
-this function is implemented in Coq's tactics language. A more convenient way is to use type class
-unification; see section~\ref{quote}. In section~\ref{canonical}, we end with a short comparision
-between type classes and canonical structures.
-
-In this paper we focus on the Coq proof assistant. We conjecture that the methods can be transfered
-to any type theory based proof assistant supporting type classes such as
-Matita~\cite{asperti2007user}.
-
-\section{Preliminaries}
-\subsection{Coq type theory}
-The Coq proof assistant is based on the calculus of inductive
-constructions~\cite{CoquandHuet,CoquandPaulin}.\marginpar{More}
-This type theory lacks quotient types, as it would make type checking undecidable. Instead it
-supports setoids (aka Bishop sets), a pair of a type and an equivalence
-relation~\cite{Bishop67,Hofmann,Capretta}. Recently, improved support has been added to the
-system~\cite{Setoid-rewrite}. Functions respecting these equivalence relations are called |Proper|.
-As we will see in Section~\ref{univ2} below this functionality can also be used to work with
-quotient algebras.
-
-\subsection{Type classes}
-Implementation in Ltac, Notation etc
-Coercions
-
-\section{To bunde or not to bundle? (Or: What to bundle?)}
-
-How should we define reflexive, transitive, and symmetric relations? In a first attempt, we
-might define:
-%format → = "\rightarrow"
-%format Π = "\Pi"
-\begin{code}
-  Record ReflexiveRelation: Type :=
-    { rr_carrier: Type
-    ; rr_rel: rr_carrier → rr_carrier → Prop
-    ; rr_proof: Π x, rr_rel x x }.
-\end{code}
-and similar for transitive and symmetric relations.
-
-With this approach, we run into problems when we try to define |EquivalenceRelation| in terms of
-|ReflexiveRelation| and its two siblings, because when posit three objects of type
-|ReflexiveRelation|, |TransitiveRelation|, and |SymmetricRelation|, respectively, we have to
-add equalities expressing that they talk about the same carrier and relation, and these equalities
-are a nightmare to work with (because the three carriers and relations won't be convertible).
-Technically, this leads to the research in manifest fields, the \emph{with} construction, to which
-we return in Section~\ref{manifest}.
-
-As a first step we would define |ReflexiveRelation| as:
-\begin{code}
-  Record ReflexiveRelation (carrier: Type): Type :=
-    { rr_rel: rr_carrier → rr_carrier → Prop
-    ; rr_proof: Π x, rr_rel x x }.
-\end{code}
-
-This would solve the problem for the carrier, but |EquivalenceRelation| would still need equalities
-for the relation:
-\begin{code}
-  Record EquivalenceRelation: Type :=
-    { er_carrier: Type
-    ; er_refl: ReflexiveRelation er_carrier
-    ; er_sym: SymmetricRelation er_carrier
-    ; er_trans: TransitiveRelation er_trans
-    ; same_rel: rr_rel er_refl = sr_rel er_sym
-    ; same_rel': rr_rel er_refl = tr_rel er_trans }.
-\end{code}
-
-A natural solution is to pull the relation out of |ReflexiveRelation| (and its siblings) as well:
-\begin{code}
-  Record ReflexiveRelation (carrier: Type)
-                           (rel: carrier → carrier → Prop): Prop :=
-    { rr_proof: Π x, rr_rel x x }.
-\end{code}
-or equivalently:
-\begin{code}
-  Definition ReflexiveRelation (carrier: Type)
-                               (rel: carrier → carrier → Prop): Prop :=
-    Π x, rr_rel x x.
-\end{code}
-This is how reflexivity is defined in the standard library, and this is entirely sensible. The
-change from |Type| to |Prop| reflects the fact that |ReflexiveRelation| is now strictly a property,
-proofs of which can be nicely kept opaque;%
-\footnote{Coq distinguishes between transparent and opague definitions. The former can be unfolded,
-whereas this is impossible for the latter.}
-we will return to this issue in Section~\ref{explicit}.
-We can now define |EquivalenceRelation| without awkward equalities:
-\begin{code}
-  Record EquivalenceRelation: Type :=
-    { er_carrier: Type
-    ; er_rel: er_carrier → er_carrier → Prop
-    ; er_refl: ReflexiveRelation er_carrier er_rel
-    ; er_sym: SymmetricRelation er_carrier er_rel
-    ; er_trans: TransitiveRelation er_trans er_rel }.
-\end{code}
-
-But looking at |EquivalenceRelation|, we see that it now essentially has the same form as the
-original naive |ReflexiveRelation| record we started with. Clearly, equivalence relations
-and reflexive relations are analogous concepts, and by the same argument we
-conclude that |EquivalenceRelation| should really be defined as:
-
-\begin{code}
-Record EquivalenceRelation (er_carrier: Type)
-                             (er_rel: er_carrier → er_carrier → Prop): Prop :=
-  { er_refl: ReflexiveRelation er_carrier er_rel
-  ; er_sym: SymmetricRelation er_carrier er_rel
-  ; er_trans: TransitiveRelation er_trans er_rel }.
-\end{code}
-
-After all, when we define a type with
-two different equivalence relations on it,
-such as the quotient of a setoid,
-we have the same considerations as for |ReflexiveRelation| above.
-
-At the risk of overstressing the point, let us look at how the same problem and solution keep
-appearing in composite structures. Suppose we use a definition of |Monoid| that bundles the
-relation. Then if we proceed to define |SemiRing| as a double |Monoid|, we face a dilemma:
-\begin{code}
-  Record SemiRing A: Type :=
-    { mult_monoid: Monoid A 
-    ; plus_monoid: Monoid A 
-    ; plus_comm: Commutative (mon_op plus_monoid)
-    ; mult_comm: Commutative (mon_op mult_monoid)
-    ; mult_0_l: Π x: A, ?
-\end{code}
-Which equality do we use in the expression ?, |mult_monoid|'s equality or |plus_monoid|'s?
-Again, for this definition to be usable at all, we would need to add an equality expressing that the
-two monoids contain the same equivalence relation. And again, the proper solution is to change
-Monoid and unbundle the relation. But here, we have a choice as to how much we unbundle exactly.
-(For the moment we ignore the question of whether the unit and binary operator should be unbundled
-as well, focusing only on the equivalence relation and its equivalence proof.)
-
-We \emph{could} become over-eager and decide to unbundle the equivalence proof term as well:
-\begin{code}
-Record Monoid (A: Type) (R: A → A → Prop)
-  (e: Equivalence R) : Type := ...
-\end{code}
-However, this way if we happen to have two proofs |p| and |q| that a certain relation is an
-equivalence, we end up with two non-convertible ways to express that the relation (together with
-some unit and binary operation) form a Monoid: 'Monoid A R p ...' and 'Monoid A R q ...'. We must
-either try to deal with this (by proving that Monoid proofs with different Equivalence proofs are
-interchangable, and hope for automation, such as type class instance resolution, to use them
-interchangably), or take great care to ensure that |Equivalence| proof terms are always shared,
-essentially elevating them to `values', which goes against the principle of proof
-irrelevance.\marginpar{What if we add proof irrelevance?}
-
-Neither of these options are very appealing. We really should just make |Monoid| take as parameters
-only the operations/relations. If we treat the monoid's unit and binary operation the same way, this
-means that |Monoid| ends up in |Prop|, and that its instances can be left opaque, which is really
-how things ought to be.
-
-Another benefit of this approach is that it works very smoothly when these records are in fact type
-classes. When relations and operations are bundled and do \emph{not} appear as parameters of the
-structure record/class, expressing as a type class constraint that some collection of relations and
-operations posess the structure seems to require manifest fields; see Section~\ref{manifest}. With
-our unbundled approach, on the other hand, such type class constraints are easily expressed without
-any extensions as simple type inhabitation queries
-% (e.g. ``hey Coq, find me a proof of [Monoid some_rel some_op some_unit]'')
-of types living in Prop, so that we don't even care what instances are found (thus embracing proof
-irrelevance).
-
-We remark that leaving the equivalence proofs bundled means that when one has a proof that something
-is a semiring, one has two proofs (via the two monoids) that the equivalence relation is an
-equivalence. Fortunately, having redundant proofs like this is entirely harmless when one's terms
-never refer (directly or indirectly) to proofs, which is the case when the relations and operations
-they are composed of are left unbundled. Here, too, we enjoy the warm embrace of proof irrelevance.
-
-From the discussion so far, we derive some principles:
-\begin{enumerate}
- \item structural concepts (like Reflexive, Setoid, Monoid) should have their structural components
-(i.e. carriers, relations, and operations) as parameters, and nothing else;
- \item (as a consequence) these concepts should be defined as predicates (which includes records and
-classes living in Prop), and their proofs may be kept opaque;
- \item (as a second consequence) algebraic expressions need/should never refer (directly or
-indirectly) to proofs.
-\end{enumerate}
-% \setcounter{enumi}{3}
-
-This list will grow in subsequent sections.
-
-The principles listed so far are not particularly controversial, in fact this
-approach ensures maximum flexibility~\cite{Hints}. However,
-there is a common perception that unbundling carriers, relations, and operations this way invariably
-leads either to unmanagably big algebraic expressions (involving endless projections and/or implicit
-arguments), or to unmanageably long and tedious enumeration of carriers/relations/operations, or to
-problems establishing canonical names for operations (since we no longer project them out of
-structure instances all the time).
-
-In the next sections, we show how systematic use of type classes (and their support infrastructure
-in Coq), combined with the use of a special form of type class we call an `operational type class'
-(as opposed to a `structural' type class we might use in place of the records shown thus far), lets
-us avoid these problems.
-
-\section{Managing unbundled structure with type classes}
-
-So far, we have only looked at the definitions of the structures themselves, and have not yet
-considered the perspective of someone trying to prove some theorems about these abstract structures.
-We would like to approximate the common mathematical vernacular, where one
-simply says:
-
-for $x, y, z$ in a semigroup $G$, $x * y * z = z * y * x$, $\ldots$
-
-Suppose we have defined |SemiGroup| in line with the aforementioned principles:
-\begin{code}
-Record SemiGroup (A: Type) (eq: A → A → Prop)
- (op: A → A → A): Prop := ...
-\end{code}
-Then our starting point is the obvious but awful:
-\begin{code}
-Lemma assoc1 A eq op (P: SemiGroup A eq op) x y z:
-                   eq (op (op x y) z) (op (op z y) x).
-\end{code}
-As a first cleaning measure, we might hope to introduce some notations to let us use infix notation
-for the semigroup operator (say, |&|) and for equality. Unfortunately, at this point the names |eq|
-and |op| are just local names and there is no way to predeclare that they are to be associated with
-any notation. What we really need are canonical names for these notions, so that we may fashion
-notations for them.
-
-This is a typical job for type classes, and they are used in this way in the |Clean| standard
-library~\cite{Clean}. Similarly, Haskell uses type classes to overload notations.
-We simply follow suit:
-\begin{code}
-  Class Equiv A := equiv: relation A.
-  Class SemiGroupOp A := sg_op: A → A → A.
-
-  Infix "=" := equiv: type_scope.
-  Infix "&" := sg_op (at level 50, left associativity).
-\end{code}
-We choose to use |&| here and reserve the notation |*| for ring multiplication.
-
-We use = for setoid equality, since Leibniz equality seems less important for the formalization of
-mathematics. More precisely, setoid equality is the basic concept. In special case, one can prove
-that setoid equality coincides with Leibniz equality.
-
-We would have like to use a haskell style definition
-\begin{code}
-Class Equiv A := (=): relation A.
-\end{code}
-Unfortunately, currently this is not supported in Coq.
-
-These single-field type classes are given special treatment in that they are not translated into
-records the way classes with any other number of fields are, but are instead turned into ordinary
-definitions. The effect of this is that |Equiv A| immediately reduces to |relation A|, and that the
-equiv `projection' is the identity function whose argument is inferred using instance resolution.
-Thus, for a very modest cost in indirection, we gain the ability to use canonical names, and with
-it, notations:
-\begin{code}
-  Record SemiGroup (A: Type) (eq: Equiv A)
-                             (op: SemiGroupOp A): Prop := ...
-  Lemma assoc1 A eq op (P: SemiGroup A eq op) x y z:
-                    x & y & z = z & y & x.
-\end{code}
-We call these type classes \emph{operational type classes}. We note that operational type classes
-allow us to avoid Coq's notation |scope| mechanism.
-
-It is important to point out that although the expression now \emph{looks} like it is bound to some
-specific semigroup structure (which with traditional approaches would imply projections from bundles
-and/or reference to proofs), it is really only referring near-directly to the actual operations
-involved, with the semigroup proof existing only as opaque `knowledge about the operations' which we
-may use in the proof. This lack of projections keeps our terms small and independent, and keeps
-rewriting simple and sane.
-
-On the other hand, it \emph{does} mean that all the carriers/relations/operations end up in the
-context. The context for |assoc1| looks like:
-\begin{code}
-  A: Type
-  eq: Equiv A
-  op: SemiGroupOp A
-  P: SemiGroup A eq op
-\end{code}
-
-While we are not particularly worried about overly large contexts (we feel that this will most
-likely not be problematic for any but the most complex formalizations), having to \emph{declare}
-these entities (like |eq| and |op| for |assoc1|) \emph{is} a bit of a chore. Fortunately, Coq
-provides a feature called implicit generalization, which is exactly what we need.
-\marginpar{describe how it works} With it, we can write |assoc1| as follows:
-\begin{code}
-  Lemma assoc1 `(SemiGroup A) x y z: x & y & z = z & y & x.
-\end{code}
-Thus, we have reached the mathematical ideal we aimed for.
-
-Let us now stop and think to what extent this approach suffers from all the problems commonly
-associated with it. In particular, let us imagine what happens to our terms and contexts when we
-want to talk about nested structures such as polynomials over the ring of matrices over a
-given ring; see Section~\ref{interfaces}.
-
-\section{Subset types}
-BEGIN SPECULATIVE
-
-A record type is \marginpar{mostly?} syntactic sugar for a $\Sigma$-type. In the presence of a
-squash operator, |[]:Type->Prop|, we may define a subset type as |\{A\mid B\}=Sig A. [B]|. This is
-pursued by Barras~\cite{Barras:subset,Werner}. The program machinery fakes subset types in the
-present Coq implementation. Proof irrelevanve in Coq would allow a faithful
-implementation.
-
-Squash is a monad and allows many constructions: subsets, unions, images, heterogeneous equality,
-\ldots.
-
-Question: Would the meta-theory simplify if we would just add a conversion rule $x\triangleright
-\epsilon$ for $x:Prop$? We could add a term constructor $[a]$ which will have type $[A]$. There
-seems to be no reason to connect this to |Prop| or impredicativity.
-
-END SPECULATIVE
-
-We add another principle:
-\begin{itemize}\setcounter{enumi}{3}
- \item Working hypothesis: `every record can be reordered as a subset'. Consequently, we
-rewrite this subset |\{A\mid B\}| as |Record r (a:A):={b:B}|.
-\end{itemize}
+%Our contributions are:\footnote{The sources are available at:~\url{http://www.eelis.net/research/math-classes/}}%
+%\begin{itemize}\itemsep=0pt
+% \item The implementation of an algebraic hierarchy based on type classes, addressing the
+%issues raised above; see Section~\ref{classes}.
+% \item A formalization of many-sorted universal algebra in Coq, including a construction of
+%free $\Sigma$-algebras; Section~\ref{univ}.
+%\end{itemize}
 
 \section{Algebraic hierarchy using type classes}\label{classes}
+\vspace*{-2ex}
+\paragraph{Canonical names.}\label{names}
+Sharing of notations is an important aspect of the design of a mathematical library. Haskell uses
+type classes to overload notations. In Coq we assign a class to every operation, thus avoiding Coq's notation \ensuremath{\mathbf{scope}} mechanism.
+ Each such operation class has only \emph{one} constructor, hence no record is constructed.
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathbf{Class}\;\Conid{Equiv}\;\Conid{A}\;\Conid{:=}\;\Varid{equiv:}\;\Varid{relation}\;\Conid{A.}{}\<[E]%
+\\
+\>[B]{}\mathbf{Infix}\;\text{\tt \char34 ==\char34}\;\Conid{:=}\;\Varid{equiv}\;(\Varid{at}\;\Varid{level}\;\Varid{70},\Varid{no}\;\Varid{associativity})\;.{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+% Instance: Params (@equiv) 2.
+\vspace*{-5ex}
 \paragraph{Hierarchy.}\label{hierarchy}
@@
 pack the proofs. This can be nicely organized by using canonical names. For instance, we define:
-\begin{code}
-Class SemiRing A {e: Equiv A}{plus: RingPlus A}{mult: RingMult A}
-                 {zero: RingZero A}{one: RingOne A}:Prop :=
-  { semiring_mult_monoid:> Monoid A (op:=mult)(unit:=one)
-  ; semiring_plus_monoid:> Monoid A (op:=plus)(unit:=zero)
-  ; semiring_plus_comm:> Commutative plus
-  ; semiring_mult_comm:> Commutative mult
-  ; semiring_distr:> Distribute mult plus
-  ; mult_0_l: forall x, 0 * x == 0 }.
-\end{code}
-where e.g.\ |RingPlus| is the operation class for the semigroup operator.
-The notation |:>| makes |semiring_mult_monoid| a coercion: when asked for a monoid it suffices to
-provide a ring.
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{18}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathbf{Class}\;\Conid{SemiRing}\;\Conid{A}\;\hspace*{0mm}\{\Varid{e:}\;\Conid{Equiv}\;\Conid{A}\mskip1.5mu\}\;\hspace*{0mm}\{\Varid{plus:}\;\Conid{RingPlus}\;\Conid{A}\mskip1.5mu\}\;\hspace*{0mm}\{\Varid{mult:}\;\Conid{RingMult}\;\Conid{A}\mskip1.5mu\}\;{}\<[E]%
+\\
+\>[B]{}\hsindent{18}{}\<[18]%
+\>[18]{}\hspace*{0mm}\{\Varid{zero:}\;\Conid{RingZero}\;\Conid{A}\mskip1.5mu\}\;\hspace*{0mm}\{\Varid{one:}\;\Conid{RingOne}\;\Conid{A}\mskip1.5mu\}\;\Conid{:Prop}\;\Conid{:=}\;{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\hspace*{0mm}\{\Varid{semiring\char95 mult\char95 monoid:>}\;\Conid{Monoid}\;\Conid{A}\;(\Varid{op:=mult})\;(\Varid{unit:=one}){}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{};\Varid{semiring\char95 plus\char95 monoid:>}\;\Conid{Monoid}\;\Conid{A}\;(\Varid{op:=plus})\;(\Varid{unit:=zero}){}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{};\Varid{semiring\char95 plus\char95 comm:>}\;\Conid{Commutative}\;\Varid{plus}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{};\Varid{semiring\char95 mult\char95 comm:>}\;\Conid{Commutative}\;\Varid{mult}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{};\Varid{semiring\char95 distr:>}\;\Conid{Distribute}\;\Varid{mult}\;\Varid{plus}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{};\Varid{mult\char95 0\char95 l:}\;\forall\;\Varid{x},\Varid{0}\;\Varid{*}\;\Varid{x}\;= \;\Varid{0}\mskip1.5mu\}\;.{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+where e.g.\ \ensuremath{\Conid{RingPlus}} is the operation class for the semigroup operator.
 
@@
 
-\subsection{Manifest fields}\label{manifest}
-
 Manifest fields in records allow us to fix a field of a dependent record.
@@
 by implementing many-sorted universal algebra. A formalization of universal algebra
-already exists in Coq~\cite{DBLP:conf/tphol/Capretta99}. It has recently been extended with more
-categorical results~\cite{dominguez2008formalizing}. We have conveniently redeveloped most of this
-using type classes. We defined equational theories and their varieties:
-the categories of their models. A quote function helps to translate a statement for an instance of
-a type class to a statement in the equational logic.
+already exists in Coq~\cite{DBLP:conf/tphol/Capretta99}. We have conveniently redeveloped most of it
+using type classes, moreover we have added a categorical framework. For instance, we show that the
+term-algebra for a signature is initial. We defined equational theories and their categories of
+their models. A quote function helps to translate a statement for an instance of
+a type class to a statement in the equational logic. We have a prototype automatic quoting mechanism
+using type classes similar to that using unification hints~\cite{Hints} .
 
-\subsection{Category theory}\label{cats}
-Using type classes, we provide a basic library for category theory: categories, functors,
-adjunctions, monads,\ldots We follow the usual type theoretical definition of a\marginpar{Use
-notation for comp} category~\cite{saibi1995constructive}
-\begin{code}
- Class Category O `{!Arrows O} `{forall x y: O, Equiv (x --> y)}
-   `{!CatId O} `{!CatComp O}: Prop :=
-  { arrow_equiv:> forall x y: O, Setoid (x --> y)
-  ; comp_proper:> forall x y z: O,
-      Proper (equiv ==> equiv ==> equiv)%signature (@comp _ _ _ x y z)
-  ; comp_assoc: forall (w x y z: O) (a: w --> x) (b: x --> y) (c: y --> z),
-      comp c (comp b a) == comp (comp c b) a
-  ; id_l: forall (x y: O) (a: y --> x), comp cat_id a == a
-  ; id_r: forall (x y: O) (a: x --> y), comp a cat_id == a }.
-\end{code}
-This is based on the 2-categorical idea that we have equality on arrows, but not on objects.
-Similarly, we have equality on natural transformations, but not on functors.
-
-%format ◎ = "\circ"
-For instance natural transformations are formalized as: |◎|
-\begin{code}
-  Context `{Category C} `{Category D} (F G: C -> D)
-     `{!Functor F Fa} `{!Functor G Ga}.
-
-  Class NaturalTransformation (η: forall c: C, F c --> G c): Prop :=
-    natural: forall (x y: C) (f: x --> y), η y ◎ fmap F f == fmap G f ◎ η x.
-\end{code}
-
-The notation |◎| denotes function composition.
-Here |fmap F| explicitly refers to the arrow part of the functor |F|.\\
-In mathematics one would apply $F$ to both objects and arrows ($F x$, $F f$). However, this does
-not seem to fit with type theory, hence we follow the customs in functional programming.
-
-The proofs are short and follow the textbook easily.\marginpar{really?}
-
-We provide the category of setoids, the category of categories, the dual of a category, products,
-and hence co-products, the category of $\Sigma$-algebras.
-
-We provide the construction of the term-algebra for a signature $\Sigma$ and show that it is
-initial in the category of $\Sigma$-algebras. We proceed to prove that this provides a left-adjoint
-to the forgetful functor. The resulting expression monad will be used below.
-
-\subsection{Universal algebra continued}\label{univ2}
-
-We have developed basic universal algebra~\cite{meinke1993universal} upto the first homomorphism
-theorem.~\marginpar{Description of the work.}
-%perhaps the soundness theorem for equational logic.
-
-In order to prove the first homophism theorem, we define a congruence:
-%theory/ua_congruence.v
-\begin{code}
-Class Congruence: Prop :=
-    { congruence_proper:> Π s, Proper (equiv ==> equiv ==> iff) (e s)
-    ; congruence_quotient:> @Algebra et v e _
-    }.
-\end{code}
-We note the use of |Proper|.
-
-We define the forgetful functor from the category of $\Sigma$-algebras to the category of sets.
-
-We define the term algebra consisting of closed terms, terms with the empty type as set of
-variables. % extensionally equal closed term???
-We show that the term algebra is initial. One may expect that this could be useful to
-\emph{define}, say, the natural numbers as the initial semiring. However, it seems difficult to
-prove decidability of equality, as it seems to require a normalization procedure.
-
-We define the product of two algebras and show that it is the categorical product.
-
-Given a sub\emph{set} of an algebra, we define the sub\emph{algebra} generated by it.
-An interesting fact to mention is the use of the following heterogeneous equality between
-elements of the algebra and of the subalgebra, i.e.\ terms of different types may be equal.
-\begin{code}
-  Fixpoint heq {o}: op_type carrier o -> op_type v o -> Prop :=
-    match o with
-    | constant x => fun a b => `a == b
-    | function x y => fun a b => forall u, heq (a u) (b (`u))
-    end.
-\end{code}
-
-We connect the abstract theory to the concrete theory. Concretely, let $\Sigma$ be the theory
-of monoids. Given a monoid, we can construct the corresponding $\Sigma$-algebra and an
-object in the category of $\Sigma$-algebras. Conversely, given an object in the category, we
-construct a monoid.
-\begin{code}
- Variable o: Variety theory.
- Global Instance from_object: Monoid (o tt).
-\end{code}
-Similarly, for arrows.
+We plan to include the first homomorphism theorem and perhaps the soundness theorem for equational
+logic
 
@@
 semiring structure, is assigned priority 10.
-\begin{code}
-  Global Program Instance: NatDistance N | 10 := ...
-\end{code}
-
-The context grows as follows:\marginpar{expand}
-
-\section{Type class quoting}\label{quote}
-Unification hints~\cite{Hints} allow one to semi-automatically construct a quote
-function\footnote{Gonthier provides similar functionality using canonical structures}.
-Unification hints are absent in Coq. Fortunately, type classes provide similar functionality as
-we will now show.
-
-We define a term language.
-\begin{code}
-Inductive Expr (V: Type) := Mult (a b: Expr V) | Zero | Var (v: V).
-\end{code}
-This defines a monad on Type\marginpar{Do this!}.
-The expression type is parameterized over the set of variable indices. Hence, we diverge
- from~\cite{Hints}, which uses |nat| for variables thereby introducing bounds problems and
- dummy variables.
-
-\noindent An expression is only meaningful in the context of a variable assignment:\footnote{nat is
-an Expr-algebra. eval is map vars composed with the algebra map.}
-\begin{code}
-Definition Value := nat.
-Definition Vars V := V -> Value.
-
-Fixpoint eval {V} (vs: Vars V) (e: Expr V): Value :=
-  match e with
-  | Zero => 0
-  | Mult a b => eval vs a * eval vs b
-  | Var v => vs v
-  end.
-\end{code}
-%
-Monads are trees with grafting. As explained by sigfpe~\cite{MonadsGrafting}, the tree monad
-$TX:=1+T^2X$, may be seen as the prototypical monad. Indeed, the |Expr| monad, i.e.\ the free
-$\Sigma$-algebra monad, behaves much like the tree monad. We now provide a direct definition of the
-corresponding bind operation.
-
-\begin{code}
-Fixpoint bind {V W} (f: V -> Expr W) (e: Expr V): Expr W :=
-  match e with
-  | Zero => Zero
-  | Mult a b => Mult (bind f a) (bind f b)
-  | Var v => vs v
-  end.
-\end{code}
-
-We have shown that |Expr|, i.e.\ |bind| together with |Var|, is a monad on the category of Types
-with extensional functions between them~\marginpar{Really?}.
-
-\noindent Some simple combinators for variable packs:
-%
-\begin{code}
-Definition novars: Vars False := False_rect _.
-Definition singlevar (x: Value): Vars unit := fun _ => x.
-Definition merge {A B} (a: Vars A) (b: Vars B): Vars (A+B) :=
-  fun i => if i then a j else b j.
-\end{code}
-
-\noindent These last two combinators are the `constructors' of an implicitly defined subset of
- Gallina terms, representing heaps, for which we implement syntactic lookup with type classes.
-Given a heap and value, |Lookup| instances give the value's index in the heap:\marginpar{A
-canonical structures approach would allow us to do this in Coq?}
-\begin{code}
-  Class Lookup {A} (x: Value) (f: Vars A) :=
-    { lookup: A; lookup_correct: f lookup = x }.
-\end{code}
-% Context (x: Value) {A B} (va: Vars A) (vb: Vars B).
-
-If the heap is a merge of two heaps and we can find the value's index in the left heap, we can
-access it by indexing the merged heap (and vice versa)
-\begin{code}
-  Global Instance lookup_left `{!Lookup x va}: Lookup x (merge va vb)
-    := { lookup := inl (lookup x va) }.
-\end{code}
-If the heap is just a singlevar, we can easily index it.
-\begin{code}
-  Global Program Instance: Lookup x (singlevar x) := { lookup := tt }.
-\end{code}
-
-One useful operation we need before we get to Quote relates to variables and expression
- evaluation. As its name suggests, |map_var| maps an expression's variable indices.\marginpar{monad}
-
-\begin{code}
-Definition map_var {V W: Type} (f: V -> W):
-    Expr V -> Expr W :=
-  fix F (e: Expr V): Expr W :=
-    match e with
-    | Mult a b => Mult (F a) (F b)
-    | Zero => Zero
-    | Var v => Var (f v)
-    end.
-\end{code}
-In |Quote| below, the idea is that |V, l|, and |n| are all input variables, while |V'| and |r| are
-output variables (in the sense that we will rely on unification to generate them). |V| and |l|
-represent the current heap, |n| represents the value we want to quote, and |V'| and |r'| represent
-the heap of newly encountered variables during the quotation.
-  This explains the type of quote: it is an expression that refers either to variables from
-the old heap, or to newly encountered variables. Finally, |eval_quote| is the usual correctness
-property, which now merges the two heaps.
-
-\begin{code}
-  Class Quote {V} (l: Vars V) (n: Value) {V'} (r: Vars V') :=
-    { quote: Expr (V + V')
-    ; eval_quote: @eval (V+V') (merge l r) quote = n }.
-\end{code}
-
-Our first instance for |Zero| is easy. The `novars' in the result type reflects the fact that no
-new variables are encountered.
-\begin{code}
-  Global Program Instance quote_zero V (v: Vars V):
-    Quote v 0 novars :=
-  { quote := Zero }.
-\end{code}
-
-The instance for multiplication is a bit more complex. The first line consists of
- variable declarations. The second line is important. `Quote x y z' must be read as
- `quoting y with existing heap x generates new heap z', so the second line basically just
-shuffles heaps around.
- The third line contains some |map_var|'s because the heap shuffling must be
-reflected in the variable indices, but apart from that it's just constructing a |Mult| term with
-quoted subterms.
-
-\begin{code}
-Global Program Instance quote_mult V (v: Vars V)
-  n V' (v': Vars V') m V'' (v'': Vars V'')
-  `{!Quote v n v'} `{!Quote (merge v v') m v''}:
-  Quote v (n * m) (merge v' v'') :=
-  { quote := Mult (map_var bla (quote n)) (map_var sum_assoc (quote m)) }.
-\end{code}
-  Now follows the instance where we recognize values that are already in the heap. This
-   is expressed by the Lookup requirement, which will only be fulfilled if the Lookup instances
-   defined above can find the value in the heap. The novars in the |Quote v x novars| result
-   reflects that this quotation does not generate new variables.
-
-\begin{code}
-Global Program Instance quote_old_var
-  V (v: Vars V) x {i: Lookup x v}:
-  Quote v x novars | 8 := { quote := Var (inl (lookup x v)) }.
-\end{code}
-
-\noindent Finally, the instance for new variables. We give this a lower priority so that it is only
-used if Lookup fails. The |8| in the previous example is a random number less than |9| below. A
-similar method is used in Coq's notation mechanism. We strongly recommend a change in
-implementation supporting an agda style precedence relation which is only restricted to be a
-directed acyclic graph~\cite{danielsson2009parsing}.
-
-\begin{code}
-Global Program Instance quote_new_var V (v: Vars V) x:
-  Quote v x (singlevar x) | 9 := { quote := Var (inr tt) }.
-\end{code}
-
-\noindent As announced this provides a light-weight quoting function\footnote{|eval_quote'| is a
-modification of |eval_quote| which works in the empty context.}:
-\begin{code}
-Goal forall x y (P: Value -> Prop), P ((x * y) * (x * 0)).
-  intros. rewrite <- (eval_quote' _).
-\end{code}
-This turns the goal into | (eval variable_pack quote)|.
-
-It would be interesting to use this technique to implement the |congruence| tactic, which implements
-the congruence closure algorithm~\cite{corbineau2007deciding}. One would naturally obtain a tactic
-which moreover works on setoid equalities.
-
-\section{Canonical structures}\label{canonical}
-Packed classes use canonical structures for the algebraic hierarchy. Both canonical structures and
-type classes may be seen as instances of hints in unification~\cite{Hints}. Some uses of canonical
-structures can be replaced by type class instances. The user manual (2.7.15) uses canonical
-structures to derive the setoid equality on the natural
-numbers in the following example |(S x)==(S y)|. We have conveniently reimplemented this with
-type classes with superior proof terms. Canonical structures give
-\begin{code}
-@equiv(Build_Setoid nat (@eq nat)(@eq_equivalence nat))(S x)(S y)
-\end{code}
-
-which includes an explicit proof that |(@eq nat)| is an equivalence,
-whereas we obtain |@equiv nat (@eq nat) (S x) (S y)|.
-
-\subsection{Big operators}
-Canonical structures have been used to provide a uniform treatment~\cite{bertot2008canonical} of big
-operators (like $\Pi,\sum, \max$). These operators extend a pair of a binary and a 0-ary operation
-to an $n$-ary operation for any $n$. Categorically, one considers the algebra maps from non-empty
-lists, lists, inhabited finite sets and finite sets to the carrier of a semigroup, monoid,
-commutative semigroup, commutative monoid. Hence we reuse the libraries for lists etc.
-
-We use type classes to deduce the relevant monoid operation and write:
-\begin{code}
-Definition seq_sum
-  `{Sequence A T} `{RingPlus A} `{z: RingZero A}: T -> A
-  := @seq_to_monoid A T _ A ring_plus z id.
-  Eval compute in seq_sum [3; 2].
-\end{code}
-
-\section{Constructive mathematics and proof irrelevance}\label{explicit}
-As originally concieved~\cite{ITT,CMCP}, Propositions are Types and we can extract information from
-them. In practice, it is convenient to assume that all proofs of a proposition are equal, the
-principle of proof irrelevance. These two views clearly contradict eachother. In Coq, a middle way
-was choosen, we are forbidden to extract information from proofs, but proofs are not identified.
-Here we need to pay, but do not seem to get anything in return.%
-\footnote{To be precise, what we get in return is the possibility to extract precisely the program
-we want.}
-There are various proposals to add the principle of proof irrelevance to Coq, perhaps a
-promissing one is the Calculus of implicit
-constructions~\cite{miquel2001implicit,barras2008implicit}. Such a change would fit naturally with
-the Hoare-style separation between programs and proofs which is used by the |Program| machinery.
-
-On the other hand, we found instances where we needed proof \emph{relevance}. In this case, it lead
-to a duplication of code which may have been avoided by universe polymorphism for |Fixpoint|s. In
-practice, manual dead-code analysis seems useful. Making Harrop formulas proof irrelevant seems to
-be a good first approximation~\cite{lcf:spi:03}.
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[3]{}\mathbf{Global}\;\mathbf{Program}\;\Conid{Instance:}\;\Conid{NatDistance}\;\Conid{N}\;\mid \;\Varid{10}\;\Conid{:=}\;\Varid{...}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
 
@@
 
-We conclude that unification hints, canonical structures and type classes all appear to have
-similar expressive powers. Canonical structures, being tied to the unifier, are more robust, but
-seems to require more inguinity. Type classes are easier to use. Unification hints, may be a
-good generalization, but are absent from Coq. We encourage their inclusion to the
-Coq system.
+Packed classes use canonical structures for the algebraic hierarchy and a uniform notation. Both
+canonical structures and
+type classes may be seen as instances of hints in unification~\cite{Hints}. Some uses of canonical
+structures can be replaced by type class instances. The user manual (2.7.15) uses canonical
+structures to derive the setoid equality on the natural
+numbers in the following example \ensuremath{(\Conid{S}\;\Varid{x})\;= \;(\Conid{S}\;\Varid{y})}. We have reimplemented this with
+type classes with similar proof terms. Canonical structures give
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{@equiv}\;(\Conid{Build\char95 Setoid}\;\mathbb{N}\;(\Varid{@eq}\;\mathbb{N})\;(\Varid{@eq\char95 equivalence}\;\mathbb{N}))\;(\Conid{S}\;\Varid{x})\;(\Conid{S}\;\Varid{y}){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
 
-An obvious topic for future research is the extension from equational logic to partial Horn
-logic~\cite{palmgren2007partial}. Another topic would be to fully, but practically, embrase the
-categorical approach to universal algebra~\cite{pitts2001categorical}.
+which includes an explicit proof that \ensuremath{(\Varid{@eq}\;\mathbb{N})} is an equivalence,
+whereas we obtain \ensuremath{\Varid{@equiv}\;\mathbb{N}\;(\Varid{@eq}\;\mathbb{N})\;(\Conid{S}\;\Varid{x})\;(\Conid{S}\;\Varid{y})}.
 
-According to |coqwc|, our development consists of 3103 lines of specifications and 707 lines of
-proofs.\marginpar{update,targz}
-
-% Alt-Ergo congruence closure parametrized by an equational theory.
-% As usual either implement in Coq (verify the algorithm), or Check traces, or Check Alt-Ergo (Work
-% in progress).
-
-% http://lara.epfl.ch/dokuwiki/sav09:congruence_closure_algorithm_and_correctness
-% I understand that this is the algorithm underlying congruence and perhaps first-order.
-% 
-% Type classes vs soft typing (Mizar types).
-% Type classes and types in homalg?
+According to \ensuremath{\Varid{coqwc}}, our development consists of 4479 lines of specifications and 722 lines of
+proofs.
 
 \paragraph{Acknowledgements.}
-This research was stimulated by the new possibilities provided by the introduction of type classes
-in Coq. In fact, this seems to be the first substantial development that tries to exploid all their
-possibilities. As a consequence, we found many small bugs and unintended behavior in the type
-class implementation. All these issues were quickly solved by Matthieu Sozeau. Discussions with
-Georges Gonthier and Claudio Sacerdoti Coen have helped to
-sharpen our understanding of the relation with Canonical Structures and with Unification Hints.
-%Jeremy, Thierry, Claudio, Georges?, Wouter
-%\enlargethispage{5mm}
+We would like to thank Matthieu Sozeau for discussions and quickly solving numerous small bugs and
+feature requests. 
 \bibliographystyle{plain}

### 2010-03-22 roughdiamond.tex
Rewrite first two pages in Eelis-style.

@@
 \begin{abstract}
-We propose generic design patterns to organize algebraic structures using type classes inside the
-Coq system. We support multiple inheritance, sharing of notations and theories, and automated
-structure inference. A library for many-sorted universal algebra abstracts from the specific
-algebraic structures and aids the organization of the library. Finally, we provide abstract
-interfaces for $\N$, $\Z$, $\Q$ as the initial semiring, the initial ring and a field of
-fractions.
+We present a new formalization of the algebraic hierarchy in Coq, exploiting its newly obtained type class implementation to make practical a solution formerly thought infeasible.
+Our approach addresses both traditional challenges as well as new ones resulting from our ambition to build upon this development a library of constructive analysis in which abstraction penalties affecting efficient computation are reduced to a bare minimum.
 \end{abstract}
@@
 \section{Introduction}
-Developing formalized libraries for mathematics requires a detailed analysis of the architecture~\cite{C-corn,DBLP:conf/types/HaftmannW08}, similar to that of a large software project. 
-The development of an algebraic hierarchy is an important aspect. Challenges include: 1. multiple
-inheritance: A ring is a type with two operations. How do we inherit from the two theories of
-monoids? 2. Sharing of notations.
+The development of libraries for formalized mathematics presents many software engineering challenges ~\cite{C-corn,DBLP:conf/types/HaftmannW08} arising from the fact that it is far from obvious how the clean, idealized concepts from everyday mathematics should be represented using the facilitities provided by concrete theorem provers and their formalisms, in a way that is both mathematically faithful and convenient to work with.
 
-Several solutions have been proposed for the Coq theorem prover%~\cite{Coq,BC04}
-: dependent
-records~\cite{DBLP:journals/jsc/GeuversPWZ02} (a.k.a.\ telescopes), packed classes~\cite{Packed} and
-occasionally modules. Although the latter option is quickly discarded
-\footnote{Modules are not first class, but currently they seem to be superior for extraction
-purposes%~\cite{letouzey02}%
-.}, considering
-algebraic theories as `interfaces', as opposed to `modules', has been an important
-guiding principle~\cite{Packed}.
+For the algebraic hierarchy---a critical component in any library of formalized mathematics---these challenges include structure inference, handling of multiple inheritance, idiomatic use of notations, and convenient algebraic manipulation.
 
-In this fashion we implement\footnote{The sources are available
-at:~\url{http://www.eelis.net/research/math-classes/}}
- algebraic theories as type classes, not instances, thus addressing the
-challenges above; see Section~\ref{classes}. Type classes have recently entered Coq~\cite{DBLP:conf/tphol/SozeauO08}. In Coq
-type classes are \emph{first class} citizens based on dependent records. The presence of dependent types makes them even more
-powerful than the already succesful second class implementations in Haskell and Isabelle. Moreover,
-Coq's type classes are well integrated with setoid-rewriting~\cite{Setoid-rewrite}.
+Several solutions have been proposed for the Coq theorem prover: dependent records~\cite{DBLP:journals/jsc/GeuversPWZ02} (a.k.a.
+telescopes), packed classes~\cite{Packed}, and occasionally modules.
+We present a new approach based entirely on Coq's recently obtained type classes, and show how their features together with a key design pattern let us effectively address the challenges mentioned above.
+
+Since we intend to use this development as a basis for constructive analysis with practical certified exact real arithmetic, an additional objective and motivation in our design is to facilitate \emph{efficient} computation.
+In particular, we want to be able to effortlessly swap implementations of number representations.
+Doing this requires that we have clean abstract interfaces, and mathematics tells us what these should look like: we represent $\N$, $\Z$, and $\Q$ as \emph{interfaces} specifying an initial semiring, an initial ring, and a field of integral fractions, respectively.
+
+To express these elegantly and without duplication, our development\footnote{The sources are available
+at:~\url{http://www.eelis.net/research/math-classes/}} includes an integrated formalization of the basics of category theory and multi-sorted universal algebra, all expressed using type classes where appropriate.
 
@@
 % Would module like extraction be possible?}.
-
-We abstract from concrete algebra and formalize many-sorted universal algebra, including a construction of
-free $\Sigma$-algebras; Section~\ref{univ}. We intend to use the current work as a basis for
-a library for computational analysis.  Thus we want to be able to replace data structures
-with more efficient ones. For instance, Coq provides unary, binary and machine numbers. We
-use ad-hoc polymorphism with type classes to develop a unified library for them. The interfaces for
-$\N$, $\Z,\Q$ are the initial semiring, the initial ring and a field of fractions; see
-Section~\ref{interfaces}.
 
@@
 \vspace*{-2ex}
-\paragraph{Canonical names.}\label{names}
-Sharing of notations is an important aspect of the design of a mathematical library. Haskell uses
-type classes to overload notations. In Coq we assign a class to every operation, thus avoiding Coq's notation |scope| mechanism.
- Each such operation class has only \emph{one} constructor, hence no record is constructed.
+We represent each structure in the algebraic hierarchy as a type class.
+In Coq, type classes and their instances are realized as record types (``dictionaries'') and registered constants of these types.
+This immediately leads to the familiar question of which components of the structure are to become parameters, and which are to become record(/class) fields.
+By far the most important design choice in our development is the decision to turn all \emph{structural} components (i.e.
+carriers, relations, and operations) into parameters, keeping only \emph{properties} as fields.
+Type classes defined this way are essentially predicates with automatically resolved proofs.
+
+Conventional wisdom warns that while this approach is theoretically very flexible (as we will see momentarily), one risks extreme inconvenience both in having to declare and pass around all these structural components all the time, as well as in losing notations (because we no longer project named operations out of records).
+
+These are legitimate concerns that we avoid by exploiting the way Coq type classes and their support infrastructure work, using \emph{operational type classes}: classes with a single field representing a single relation or operation in isolation.
+Such classes are treated specially by Coq in being translated to mere definitions rather than records, with the field projection becoming the identity function.
 \begin{code}
  Class Equiv A := equiv: relation A.
- Infix "==" := equiv (at level 70, no associativity).
+ Infix "=" := equiv (at level 70, no associativity).
 \end{code}
-% Instance: Params (@equiv) 2.
-\vspace*{-5ex}
-\paragraph{Hierarchy.}\label{hierarchy}
-In our formalization of algebraic structures we use the structure as (implicit) parameters, but
-pack the proofs. This can be nicely organized by using canonical names. For instance, we define:
+These operational type classes serve to establish \emph{canonical names}, which not only let us bind notations, but also allow us to make their declaration and passing implicit in most contexts.
+For instance, using the following definition of semirings, all structural parameters (represented by operational classes declared with curly brackets) will be implicitly resolved by the type class mechanism rather than listed explicitly whenever we talk about semirings.
 \begin{code}
@@
 \end{code}
-where e.g.\ |RingPlus| is the operation class for the semigroup operator.
+The concrete Coq infrastructure (added specifically to support type classes) that makes this work consists of implicit quantification (when declaring a semiring), and maximally inserted implicit arguments (when stating that something is a semiring, and when referring to operations and relations).
 
-\emph{Structure-as-parameters} helps setoid-rewriting: type class resolution
-can find the equivalence relation in the context.
-% A similar style should be possible for, say, the |Ring| tactic, instead of
-% declaring the ring structure by a separate command, we would rely on type class resolution to find
-% it.
-We note that |op| does not depend on the proof that |e| is an equivalence. We use Coq's implicit
-quantification (|`{}|) to avoid having to write all the parameters when \emph{stating} a theorem
-and Coq's maximally inserted implicit arguments to find the parameters when \emph{applying} a
-theorem. Both features are new in Coq and stem from the type class implementation.
+Having argued that the \emph{all-structure-as-parameters} approach \emph{can} be made practical, we enumerate some of the benefits that make it worthwhile.
 
-We mention the trade-off between bigger contexts versus bigger terms. Our contexts are bigger than
-those of telescopes or packed classes. In our experience, this has been relatively harmless: most
-terms in the context are there to support canonical names. Bigger terms
-\emph{do} cause problems: 1. when proofs are part of mathematical objects we need to share these
-proofs to allow rewriting. Moreover, it prohibits Opaque proofs and `proof irrelevance'. 2. The
-projection paths may not be canonical.
+First, multiple inheritance becomes trivial: |SemiRing| inherits two |Monoid| structures on the same carrier and setoid relation, using ordinary named arguments (rather than dedicated extensions \cite{DBLP:conf/types/Luo08}) to achieve ``manifest fields''.
 
-Coercion pullbacks~\cite{Hints} were introduced to address problems with multiple coercions paths,
-as in the definition of a semiring: a type with two monoid structures on it. We avoid some
-of these problems by explicitly specifying the fields. We emphasize that the semiring properties are automatically derived from the ring
-properties, although the properties of a semiring are not structurally included in the ring
-properties.
+Second, because our terms are small and independent and never refer to proofs, we are invulnerable to concerns about efficiency and ambiguity of projection paths (obviating the need for things like the proposed coercion pullbacks~\cite{Hints}).
 
-Manifest fields in records allow us to fix a field of a dependent record.
-Luo~\cite{DBLP:conf/types/Luo08} has proposed coersive subtyping to implement manifest fields.
-It was conjectured~\cite{Hints} that implementing the algebraic hierarchy with type classes would
-be awkward due to problems with equality. We avoided these problems by shifting the fields
-from the dependent record to its argument, thus mimicking manifest fields by using type classes.
-Since |op| is \emph{instantiated} with |mult|, we need not worry about the
-equality |op=mult|.
+Third, since our structural type classes are mere (proof-irrelevant) predicates, overlap between their instances is a non-issue.
+Together with the previous point, this gives us tremendous freedom to posit multiple structures on the same operations and relations, including ones derived implicitly via subclasses.
+
+% Coercion pullbacks~\cite{Hints} were introduced to address problems with multiple coercions paths,
+% as in the definition of a semiring: a type with two monoid structures on it. We avoid some
+% of these problems by explicitly specifying the fields. We emphasize that the semiring properties are automatically derived from the ring
+% properties, although the properties of a semiring are not structurally included in the ring
+% properties.
 

### 2010-03-22 roughdiamond.tex
Use "penalties inhibiting efficient computation" rather than "... affecting ...".

@@
 We present a new formalization of the algebraic hierarchy in Coq, exploiting its newly obtained type class implementation to make practical a solution formerly thought infeasible.
-Our approach addresses both traditional challenges as well as new ones resulting from our ambition to build upon this development a library of constructive analysis in which abstraction penalties affecting efficient computation are reduced to a bare minimum.
+Our approach addresses both traditional challenges as well as new ones resulting from our ambition to build upon this development a library of constructive analysis in which abstraction penalties inhibiting efficient computation are reduced to a bare minimum.
 \end{abstract}

### 2010-03-25 roughdiamond.tex
Revise last two pages in diamond.

@@
 \begin{document}
-\title{Developing the algebraic hierarchy with type classes in Coq}
+\title{Developing the Algebraic Hierarchy with Type Classes in Coq}
 \author{Bas Spitters \and Eelis van der Weegen}
@@
 \begin{abstract}
-We present a new formalization of the algebraic hierarchy in Coq, exploiting its newly obtained type class implementation to make practical a solution formerly thought infeasible.
+We present a new formalization of the algebraic hierarchy in Coq, exploiting its new type class mechanism to make practical a solution formerly thought infeasible.
 Our approach addresses both traditional challenges as well as new ones resulting from our ambition to build upon this development a library of constructive analysis in which abstraction penalties inhibiting efficient computation are reduced to a bare minimum.
+To support mathematically sound abstract interfaces for $\N$, $\Z$, and $\Q$, our formalization includes portions of category theory and multisorted universal algebra.
 \end{abstract}
@@
 \section{Introduction}
-The development of libraries for formalized mathematics presents many software engineering challenges ~\cite{C-corn,DBLP:conf/types/HaftmannW08} arising from the fact that it is far from obvious how the clean, idealized concepts from everyday mathematics should be represented using the facilitities provided by concrete theorem provers and their formalisms, in a way that is both mathematically faithful and convenient to work with.
+The development of libraries for formalized mathematics presents many software engineering challenges~\cite{C-corn,DBLP:conf/types/HaftmannW08}, because it is far from obvious how the clean, idealized concepts from everyday mathematics should be represented using the facilitities provided by concrete theorem provers and their formalisms, in a way that is both mathematically faithful and convenient to work with.
 
@@
 telescopes), packed classes~\cite{Packed}, and occasionally modules.
-We present a new approach based entirely on Coq's recently obtained type classes, and show how their features together with a key design pattern let us effectively address the challenges mentioned above.
+We present a new approach based entirely on Coq's new type class mechanism, and show how its features together with a key design pattern let us effectively address the challenges mentioned above.
 
@@
 To express these elegantly and without duplication, our development\footnote{The sources are available
-at:~\url{http://www.eelis.net/research/math-classes/}} includes an integrated formalization of the basics of category theory and multi-sorted universal algebra, all expressed using type classes where appropriate.
+at:~\url{http://www.eelis.net/research/math-classes/}} includes an integrated formalization of parts of category theory and multi-sorted universal algebra, all expressed using type classes for optimum effect.
 
@@
 
-%Our contributions are:\footnote{The sources are available at:~\url{http://www.eelis.net/research/math-classes/}}%
-%\begin{itemize}\itemsep=0pt
-% \item The implementation of an algebraic hierarchy based on type classes, addressing the
-%issues raised above; see Section~\ref{classes}.
-% \item A formalization of many-sorted universal algebra in Coq, including a construction of
-%free $\Sigma$-algebras; Section~\ref{univ}.
-%\end{itemize}
+\section{The Type-Classified Algebraic Hierarchy}\label{classes}
 
-\section{Algebraic hierarchy using type classes}\label{classes}
-\vspace*{-2ex}
+Unlike Haskell's and Isabelle's second class type classes, Coq's type classes are first class: classes and their instances are realized as ordinary record types (``dictionaries'') and registered constants of these types.
+
 We represent each structure in the algebraic hierarchy as a type class.
-In Coq, type classes and their instances are realized as record types (``dictionaries'') and registered constants of these types.
-This immediately leads to the familiar question of which components of the structure are to become parameters, and which are to become record(/class) fields.
+This immediately leads to the familiar question of which components of the structure are to become parameters of the class, and which are to become fields.
 By far the most important design choice in our development is the decision to turn all \emph{structural} components (i.e.
@@
 
-First, multiple inheritance becomes trivial: |SemiRing| inherits two |Monoid| structures on the same carrier and setoid relation, using ordinary named arguments (rather than dedicated extensions \cite{DBLP:conf/types/Luo08}) to achieve ``manifest fields''.
+First, multiple inheritance (a major headache/complication in current solutions) becomes trivial: |SemiRing| inherits two |Monoid| structures on the same carrier and setoid relation, using ordinary named arguments (rather than dedicated extensions~\cite{DBLP:conf/types/Luo08}) to achieve ``manifest fields''.
 
-Second, because our terms are small and independent and never refer to proofs, we are invulnerable to concerns about efficiency and ambiguity of projection paths (obviating the need for things like the proposed coercion pullbacks~\cite{Hints}).
+Second, because our terms are small and independent and never refer to proofs, we are invulnerable to concerns about efficiency and ambiguity of projection paths that plague existing solutions, obviating the need for extensions like the proposed coercion pullbacks~\cite{Hints}.
 
-Third, since our structural type classes are mere (proof-irrelevant) predicates, overlap between their instances is a non-issue.
-Together with the previous point, this gives us tremendous freedom to posit multiple structures on the same operations and relations, including ones derived implicitly via subclasses.
+Third, since our structural type classes are mere predicates, overlap between their instances---another worry with existing solutions---is a non-issue.
+Together with the previous point, this gives us tremendous freedom to posit multiple structures on the same operations and relations, including ones derived implicitly via subclasses: by simply declaring a |SemiRing| class instance showing that a ring is a semiring, results about semirings immediately apply implicitly to any known ring, without us having to explicitly encode this relation in the hierarchy definition itself, and without needing any projection or translation of carriers or operations.
 
-% Coercion pullbacks~\cite{Hints} were introduced to address problems with multiple coercions paths,
-% as in the definition of a semiring: a type with two monoid structures on it. We avoid some
-% of these problems by explicitly specifying the fields. We emphasize that the semiring properties are automatically derived from the ring
-% properties, although the properties of a semiring are not structurally included in the ring
-% properties.
+\section{Category Theory and Universal Algebra}\label{interfaces}\label{modul}
 
-\section{Universal algebra}\label{univ}
-After having implemented concrete algebraic structures, we generalized some of our constructions
-by implementing many-sorted universal algebra. A formalization of universal algebra
-already exists in Coq~\cite{DBLP:conf/tphol/Capretta99}. We have conveniently redeveloped most of it
-using type classes, moreover we have added a categorical framework. For instance, we show that the
-term-algebra for a signature is initial. We defined equational theories and their categories of
-their models. A quote function helps to translate a statement for an instance of
-a type class to a statement in the equational logic. We have a prototype automatic quoting mechanism
-using type classes similar to that using unification hints~\cite{Hints} .
+Motivated originally by our desire to cleanly express interfaces for basic numeric data types such as $\N$ and $\Z$ in terms of their categorical characterization as initial objects in the categories of semirings and rings, respectively, we initially introduced only the very basics of category theory into our development, again using type classes where possible to achieve the same benefits mentioned before.
 
-We plan to include the first homomorphism theorem and perhaps the soundness theorem for equational
-logic
+Realizing that much code duplication for the various algebraic structures in the hierarchy could be avoided by employing universal algebra constructions, we then proceeded to formalize some of the theory of multisorted universal algebra and equational theories, using it to automatically construct algebra/variety categories.
+We purposely did not try to use Capretta's existing Coq formalization ~\cite{DBLP:conf/tphol/Capretta99} of universal algebra for this, because an important goal for us is to find out exactly what level of elegance, convenience, and integration can be achieved using the state of the art technology offered by the Coq system (of which type classes are the most important instance).
+
+At the time of writing, our development includes a fully integrated formalization of a nontrivial portion of category theory and multisorted universal algebra, including various categories (e.g.
+the category |Cat| of categories, and generic algebra/variety categories which we instantiate to obtain the categories of monoids, semirings, and rings), functors (including automatically generated forgetful functors), natural transformations, adjunctions, initial models of varieties constructed from term algebras, transference of proofs between isomorphic models of varieties, subalgebras/subvarieties, universal congruences and quotients and products, the first homomorphism theorem, et cetera.
+
+There is an interesting interplay in our development between concrete algebraic structure type classes and their expressions on the one hand, and models of universal algebras and varieties instantiated with equational theories on the other.
+While occasionally a source of tension in that translation in either direction is not (yet) fully automatic, this duality also opens the door to the possibility of fully internalized implementations of generic tactics for algebraic manipulation, no longer requiring plugins.
+One missing piece in this puzzle is automatic quotation of concrete expressions into universal algebra expressions.
+We have already implemented a proof of concept showing that like unification hints~\cite{Hints}, type classes can be used to implement Ltac/OCaml-free quotation.
 
@@
 
-\section{Interfaces using category theory}\label{interfaces}\label{modul}
-The current work was stimulated by our desire~\cite{Riemann} to move between several
-implementations in a flexible way. We use ad-hoc polymorphism to capture the natural
-numbers without referring to their actual implementation, we have
-characterized them as the initial object in the category of semirings and derived many of its
-properties from this interface. Unary, binary and machine numbers are
-instances of this interface, so we can directly apply these results to them.
-
-We encourage more efficient implementations by assigning the default implementation a
-low priority. For example, the distance function on the natural numbers, which is derived from its
-semiring structure, is assigned priority 10.
-\begin{code}
-  Global Program Instance: NatDistance N | 10 := ...
-\end{code}
+% We encourage more efficient implementations by assigning the default implementation a
+% low priority. For example, the distance function on the natural numbers, which is derived from its
+% semiring structure, is assigned priority 10.
+% \begin{code}
+%   Global Program Instance: NatDistance N | 10 := ...
+% \end{code}
+%
 
 \section{Conclusions}
-Telescopes have been criticized~\cite{Packed} for the lack of multiple inheritance and
-the efficiency penalty of a long chain of coercion projections. Packed classes~\cite{Packed} provide
-a solution to these problems. We provide an alternative solution. We have already
-discussed multiple inheritance in Section~\ref{hierarchy}. Moreover, by exposing the
-carrier we avoid the chains of projections.
 
-Packed classes use canonical structures for the algebraic hierarchy and a uniform notation. Both
-canonical structures and
-type classes may be seen as instances of hints in unification~\cite{Hints}. Some uses of canonical
-structures can be replaced by type class instances. The user manual (2.7.15) uses canonical
-structures to derive the setoid equality on the natural
-numbers in the following example |(S x)==(S y)|. We have reimplemented this with
-type classes with similar proof terms. Canonical structures give
-\begin{code}
-@equiv(Build_Setoid nat (@eq nat)(@eq_equivalence nat))(S x)(S y)
-\end{code}
+Our development (which according to \texttt{coqwc} consists of about 5K lines of specifications and 1K lines of proofs) shows that the first class type class implementation in Coq is already an extremely powerful piece of technology which enables new practical and elegant solutions to old problems.
 
-which includes an explicit proof that |(@eq nat)| is an equivalence,
-whereas we obtain |@equiv nat (@eq nat) (S x) (S y)|.
-
-According to |coqwc|, our development consists of 4479 lines of specifications and 722 lines of
-proofs.
+In our work we push the type class implementation and the new generalized rewriting infrastructure~\cite{Setoid-rewrite} to their limits, revealing both innocent bugs as well as more serious issues (concerning both efficiency and functionality) that the Coq development team is already working on (for instance with the soon to be revealed new proof engine).
 

### 2010-03-26 roughdiamond.tex
Small improvements suggested today by Bas.

@@
 We represent each structure in the algebraic hierarchy as a type class.
-This immediately leads to the familiar question of which components of the structure are to become parameters of the class, and which are to become fields.
+This immediately leads to the familiar question of which components of the structure should become parameters of the class, and which should become fields.
 By far the most important design choice in our development is the decision to turn all \emph{structural} components (i.e.
 carriers, relations, and operations) into parameters, keeping only \emph{properties} as fields.
-Type classes defined this way are essentially predicates with automatically resolved proofs.
+Type classes defined in this way are essentially predicates with automatically resolved proofs.
 
@@
 \end{code}
-These operational type classes serve to establish \emph{canonical names}, which not only let us bind notations, but also allow us to make their declaration and passing implicit in most contexts.
+These operational type classes serve to establish \emph{canonical names}, which not only lets us bind notations to them, but also makes their declaration and use implicit in most contexts.
 For instance, using the following definition of semirings, all structural parameters (represented by operational classes declared with curly brackets) will be implicitly resolved by the type class mechanism rather than listed explicitly whenever we talk about semirings.
@@
 \end{code}
-The concrete Coq infrastructure (added specifically to support type classes) that makes this work consists of implicit quantification (when declaring a semiring), and maximally inserted implicit arguments (when stating that something is a semiring, and when referring to operations and relations).
+The two key Coq features that make this work are implicit quantification (when declaring a semiring), and maximally inserted implicit arguments (when stating that something is a semiring, and when referencing operations and relations).
+Both were added specifically to support type classes.
 
@@
 
-First, multiple inheritance (a major headache/complication in current solutions) becomes trivial: |SemiRing| inherits two |Monoid| structures on the same carrier and setoid relation, using ordinary named arguments (rather than dedicated extensions~\cite{DBLP:conf/types/Luo08}) to achieve ``manifest fields''.
+First, multiple inheritance (a major complication in current solutions) becomes trivial: |SemiRing| inherits two |Monoid| structures on the same carrier and setoid relation, using ordinary named arguments (rather than dedicated extensions~\cite{DBLP:conf/types/Luo08}) to achieve ``manifest fields''.
 
@@
 
-Realizing that much code duplication for the various algebraic structures in the hierarchy could be avoided by employing universal algebra constructions, we then proceeded to formalize some of the theory of multisorted universal algebra and equational theories, using it to automatically construct algebra/variety categories.
-We purposely did not try to use Capretta's existing Coq formalization ~\cite{DBLP:conf/tphol/Capretta99} of universal algebra for this, because an important goal for us is to find out exactly what level of elegance, convenience, and integration can be achieved using the state of the art technology offered by the Coq system (of which type classes are the most important instance).
+Realizing that much code duplication for the various algebraic structures in the hierarchy could be avoided by employing universal algebra constructions, we then proceeded to formalize some of the theory of multisorted universal algebra and equational theories, using it to automatically construct varieties of algebras.
+We purposely did not try to use existing Coq formalizations~\cite{DBLP:conf/tphol/Capretta99,dominguez2008formalizing} of universal algebra for this, because an important goal for us is to find out exactly what level of elegance, convenience, and integration can be achieved using the state of the art technology offered by the Coq system (of which type classes are the most important instance).
 
 At the time of writing, our development includes a fully integrated formalization of a nontrivial portion of category theory and multisorted universal algebra, including various categories (e.g.
-the category |Cat| of categories, and generic algebra/variety categories which we instantiate to obtain the categories of monoids, semirings, and rings), functors (including automatically generated forgetful functors), natural transformations, adjunctions, initial models of varieties constructed from term algebras, transference of proofs between isomorphic models of varieties, subalgebras/subvarieties, universal congruences and quotients and products, the first homomorphism theorem, et cetera.
+the category |Cat| of categories, and generic variety categories which we instantiate to obtain the categories of monoids, semirings, and rings), functors (including automatically generated forgetful functors), natural transformations, adjunctions, initial models of equational theories constructed from term algebras, transference of proofs between isomorphic models of equational theories, subalgebras, congruences, quotients, products, and the first homomorphism theorem.
 

### 2010-05-07 jar.tex
Merging changes from rough diammond.
Many changes.

@@
 \documentclass[a4paper,10pt,runningheads]{llncs}
+\usepackage{listings}
+% The package listingsutf8 supports utf8 for \lstinputlisting
+% However, we do not want to put all our code snippets in separate files.
 \usepackage{url}
@@
 \usepackage{amsmath}
+\usepackage[mathletters]{ucs}
 \usepackage[utf8x]{inputenx}
+\usepackage{comment}
 \newcommand{\N}{\ensuremath{\mathbbm{N}}}
@@
 \newcommand{\Q}{\ensuremath{\mathbbm{Q}}}
-%include polycode.fmt
-%include coq2tex.tex
-\arrayhs%i.e. no page breaks in code blocks
-\divide\abovedisplayskip 4
-\divide\belowdisplayskip 4
-\newcommand{\textPi}{\ensuremath\Pi}
-\newcommand{\textSigma}{\ensuremath\Sigma}
-\newcommand{\texttau}{\ensuremath\tau}
-% \newcommand{\textforall}{\ensuremath\forall}
+\usepackage{color}
+\definecolor{dkblue}{rgb}{0,0.1,0.5}
+\definecolor{lightblue}{rgb}{0,0.5,0.5}
+\definecolor{dkgreen}{rgb}{0,0.4,0}
+\definecolor{dk2green}{rgb}{0.4,0,0}
+\definecolor{dkviolet}{rgb}{0.6,0,0.8}
+\def\lstlanguagefiles{defManSSR.tex}
+\lstset{language=SSR}
 \begin{document}
-\title{DRAFT: The algebraic hierarchy in type theory using type classes}
+\title{The algebraic hierarchy in type theory using type classes}
 \author{Bas Spitters \and Eelis van der Weegen}
 \institute{Radboud University Nijmegen}
+\date{today}
 \maketitle
 \begin{abstract}
-We\footnote{\today} propose generic design patterns to organize algebraic structures using
-type classes inside the
-Coq system. We support multiple inheritance, sharing of notations and theories, and automated
-structure inference. A library for many-sorted universal algebra abstracts from the specific
-algebraic structures and aids the organization of the library. Category theory provides a further
-useful abstraction. We provide abstract
-interfaces for $\N$, $\Z$, $\Q$ as the initial semiring, the initial ring and a field of
-fractions. Similarly, we provide an interface for polynomials over a ring $R$ using $R$-algebras.
-We show that the standard implementations of these are in fact instances of these type classes.
-We report on the efficiency penality of these abstractions.
-
-Finally, we use type classes to conveniently implement a quote function using the prolog-like
-abilities of type class unification. We integrate this with our development of universal algebra.
+We present a new formalization of the algebraic hierarchy in Coq, exploiting its new type class mechanism to make practical a solution formerly thought infeasible.
+Our approach addresses both traditional challenges as well as new ones resulting from our ambition to build upon this development a library of constructive analysis in which abstraction penalties inhibiting efficient computation are reduced to a bare minimum.
+To support mathematically sound abstract interfaces for $\N$, $\Z$, and $\Q$, our formalization includes portions of category theory and multisorted universal algebra.
+Algebra flourishes by the interplay between syntax and semantics. The Prolog-like
+abilities of type class unification allow us to conveniently define a quote function thus facilitating the use of reflective techniques.
 \end{abstract}
@@
 \section{Introduction}
-Developing formalized libraries for mathematics requires a detailed analysis of the
-architecture of mathematics~\cite{C-corn,DBLP:conf/types/HaftmannW08}, similar to that of a large
-software project.
-The development of an algebraic hierarchy is an important aspect. Challenges include: 1. multiple
-inheritance: A ring is a type with two operations. How do we inherit from the two theories of
-monoids? 2. Sharing of notations. 3. Combining syntactic and semantic methods, usually this is
-addressed by reflection.
+The development of libraries for formalized mathematics presents many software engineering challenges~\cite{C-corn,DBLP:conf/types/HaftmannW08}, because it is far from obvious how the clean, idealized concepts from everyday mathematics should be represented using the facilities provided by concrete theorem provers and their formalisms, in a way that is both mathematically faithful and convenient to work with.
 
-Several solutions have been proposed for the Coq theorem prover~\cite{Coq,BC04}: dependent
-records~\cite{DBLP:journals/jsc/GeuversPWZ02} (a.k.a.\ telescopes), packed classes~\cite{Packed} and
-occasionally modules. Although the latter option is quickly discarded%
-\footnote{Modules are not first class, but currently they seem to be superior for extraction to
-OCaml~\cite{letouzey02}. Modules may also be seen as parametrised sections. They allow management
-of namespaces, notations, hints, etc. An |Import| command imports all the constants from a
-module. Thus allowing the user to type |B| instead of |A.B|. Similar functionality would be useful
-for type classes.}
-algebraic theories as `interfaces', as opposed to `modules', has been an important
-guiding principle~\cite{Packed}.
+For the algebraic hierarchy---a critical component in any library of formalized mathematics---these challenges include structure inference, handling of multiple inheritance, idiomatic use of notations, and convenient algebraic manipulation.
 
-In this fashion we implement\footnote{The sources are available
-at:~\url{http://www.eelis.net/research/math-classes/}}
- algebraic theories as type classes, not instances, thus addressing the
-challenges above; see Section~\ref{classes}. Type classes have recently entered
-Coq~\cite{DBLP:conf/tphol/SozeauO08}. In Coq,
-type classes are \emph{first class} citizens based on dependent records. The presence of dependent types makes them even more
-powerful than the already succesful second class implementations in Haskell and Isabelle. Moreover,
-Coq's type classes are well integrated with setoid-rewriting~\cite{Setoid-rewrite}.
+Several solutions have been proposed for the Coq theorem prover: dependent records~\cite{DBLP:journals/jsc/GeuversPWZ02} (a.k.a.
+telescopes), packed classes~\cite{Packed}, and occasionally modules.
+We present a new approach based entirely on Coq's new type class mechanism, and show how its features together with a key design pattern let us effectively address the challenges mentioned above.
 
-We abstract from concrete algebra and formalize many-sorted universal algebra, including a
-construction of free $\tau$-algebras; see Section~\ref{univ}.
+Since we intend to use this development as a basis for constructive analysis with practical certified exact real arithmetic, an additional objective and motivation in our design is to facilitate \emph{efficient} computation.
+In particular, we want to be able to effortlessly swap implementations of number representations.
+Doing this requires that we have clean abstract interfaces, and mathematics tells us what these should look like: we represent $\N$, $\Z$, and $\Q$ as \emph{interfaces} specifying an initial semiring, an initial ring, and a field of integral fractions, respectively.
 
-% We are also concerned with efficiency.
-% In fact, our aim is to use the present development as a new basis for Bishop's program in type
-% theory~\cite{typesreal-article}.\marginpar{More}
+To express these elegantly and without duplication, our development\footnote{The sources are available
+at:~\url{http://www.eelis.net/research/math-classes/}} includes an integrated formalization of parts of category theory and multi-sorted universal algebra, all expressed using type classes for optimum effect.
 
-So, we would like to replace data structures with more efficient ones. Concretely,
-Coq provides unary, binary and machine numbers. We use type classes polymorphism to
-develop a unified library for them; see Section~\ref{modul}.
-The interfaces for $\N$, $\Z,\Q$ are the initial semiring, the initial ring and a field of
-fractions; see Section~\ref{interfaces}. In future work we aim to base our development of the reals
-on an abstract dense set, allowing us to use the efficient dyadic
-rationals~\cite{boldo2009combining} as a base for exact real number computation in
-Coq~\cite{Riemann,Oconnor:real}.
+\section{The Type-Classified Algebraic Hierarchy}\label{classes}
+Unlike Haskell's and Isabelle's second class type classes, Coq's type classes are first class~\cite{DBLP:conf/tphol/SozeauO08}: classes and their instances are realized as ordinary record types (``dictionaries'') and registered constants of these types.
 
-The paper is organised as follows: Section~\ref{classes} starts with a development of the concrete
-algebraic hierarchy addressing the issues raised above. Section~\ref{univ} abstracts to universal
+We represent each structure in the algebraic hierarchy as a type class.
+This immediately leads to the familiar question of which components of the structure should become parameters of the class, and which should become fields.
+By far the most important design choice in our development is the decision to turn all \emph{structural} components (i.e.
+carriers, relations, and operations) into parameters, keeping only \emph{properties} as fields.
+Type classes defined in this way are essentially predicates with automatically resolved proofs.
+
+Conventional wisdom warns that while this approach is theoretically very flexible, one risks extreme inconvenience both in having to declare and pass around all these structural components all the time, as well as in losing notations (because we no longer project named operations out of records).
+These legitimate concerns will be addressed in Sections~\ref{bundle}-\ref{classes}.
+Section~\ref{univ} abstracts to universal
 algebra, we develop the basic theory to the point of the first homomorphism theorem.
-Section~\ref{interfaces} provides interfaces to usual datastructures by providing their universal
+Section~\ref{interfaces} provides interfaces to usual data structures by providing their universal
 properties. To allow symbolic manipulation of semantic objects one needs a quote function. Usually,
 this function is implemented in Coq's tactics language. A more convenient way is to use type class
-unification; see section~\ref{quote}. In section~\ref{canonical}, we end with a short comparision
+unification; see section~\ref{quote}. In section~\ref{canonical}, we end with a short comparison
 between type classes and canonical structures.
 
-In this paper we focus on the Coq proof assistant. We conjecture that the methods can be transfered
+In this paper we focus on the Coq proof assistant. We conjecture that the methods can be transferred
 to any type theory based proof assistant supporting type classes such as
@@
 The Coq proof assistant is based on the calculus of inductive
-constructions~\cite{CoquandHuet,CoquandPaulin}, a dependent type theory with (co)inductive types.
-We will not elaborate on this, but instead refer to~\cite{Coq,BC04}.
+constructions~\cite{CoquandHuet,CoquandPaulin}, a dependent type theory with (co)inductive types; see~\cite{Coq,BC04}.
 
 The Coq type theory lacks quotient
-types, as it would make type checking undecidable. Instead it
-supports setoids (aka Bishop sets), a pair of a type and an equivalence
-relation~\cite{Bishop67,Hofmann,Capretta}. Recently, improved support for setoid has been added to
-the system~\cite{Setoid-rewrite}. Functions respecting these equivalence relations are called
-|Proper|.
+types, as it would make type checking undecidable.
+Instead it supports setoids (aka Bishop sets), a pair of a type and an equivalence relation~\cite{Bishop67,Hofmann,Capretta}.
+Recently, improved support for setoids has been added to the system~\cite{Setoid-rewrite}.
+Functions respecting these equivalence relations are called \lstinline|Proper|.
 As we will see in Section~\ref{univ} below this functionality can also be used to work with
@@
 
-Coq has two kinds of sorts |Prop| for propostions and |Type| for data types. Coq
-is scheduled to support proof irrelevance, every |p,q:P:Prop| are
+Coq has two kinds of sorts \lstinline|Prop| for propositions and \lstinline|Type| for data types. Coq
+is scheduled to support proof irrelevance, every \lstinline|p,q:P:Prop| are
 convertible, in the future; see section~\ref{PI}.
@@
 
-\section{To bunde or not to bundle? (Or: What to bundle?)}
+\section{To bundle or not to bundle? (Or: What to bundle?)}\label{bundle}
 We provide a rationale for our approach. We begin gently.
@@
 might define:
-%format → = "\rightarrow"
-%format Π = "\Pi"
-\begin{code}
+% \lstinputlisting[language=SSR,inputencoding=utf8/latin1]{ReflexiveRelation.v}
+
+\begin{lstlisting}
   Record ReflexiveRelation: Type :=
@@
     ; rr_rel: relation rr_carrier
-    ; rr_proof: Π x, rr_rel x x }.
-\end{code}
+    ; rr_proof: $\Pi$ x, rr_rel x x }.
+\end{lstlisting}
+
 and similar for transitive and symmetric relations.
 
-With this approach, we run into problems when defining |EquivalenceRelation| in terms of
-|ReflexiveRelation| and its two siblings, because when posit three objects of type
-|ReflexiveRelation|, |TransitiveRelation|, and |SymmetricRelation|, respectively, we have to
+With this approach, we run into problems when defining \lstinline|EquivalenceRelation| in terms of
+\lstinline|ReflexiveRelation| and its two siblings, because when posit three objects of type
+\lstinline|ReflexiveRelation|, \lstinline|TransitiveRelation|, and \lstinline|SymmetricRelation|, respectively, we have to
 add equalities expressing that they talk about the same carrier and relation. These equalities
@@
 Technically, this leads to the research in manifest fields, i.e.\ the \emph{with} construction, to
-which
-we return in Section~\ref{manifest}.
+which we return in Section~\ref{manifest}.
 
-As a first step we would define |ReflexiveRelation| as:
-\begin{code}
+As a first step we would define \lstinline|ReflexiveRelation| as:
+\begin{lstlisting}
   Record ReflexiveRelation (carrier: Type): Type :=
     { rr_rel: relation rr_carrier
-    ; rr_proof: Π x, rr_rel x x }.
-\end{code}
+    ; rr_proof: $\Pi$ x, rr_rel x x }.
+\end{lstlisting}
 
-This would solve the problem for the carrier, but |EquivalenceRelation| would still need equalities
-for the relation:
-\begin{code}
+This would solve the problem for the carrier, but \lstinline|EquivalenceRelation| would still need equalities for the relation:
+\begin{lstlisting}
   Record EquivalenceRelation: Type :=
@@
     ; same_rel': rr_rel er_refl = tr_rel er_trans }.
-\end{code}
+\end{lstlisting}
 
-A natural solution is to pull the relation out of |ReflexiveRelation| (and its siblings) as well:
-\begin{code}
+A natural solution is to pull the relation out of \lstinline|ReflexiveRelation| (and its siblings) as well:
+\begin{lstlisting}
   Record ReflexiveRelation (carrier: Type)
                            (rel: relation carrier): Prop :=
-    { rr_proof: Π x, rr_rel x x }.
-\end{code}
+    { rr_proof: $\Pi$ x, rr_rel x x }.
+\end{lstlisting}
+
 or equivalently:
-\begin{code}
+\begin{lstlisting}
   Definition ReflexiveRelation (carrier: Type)
                                (rel: relation carrier): Prop :=
-    Π x, rr_rel x x.
-\end{code}
+    $\Pi$ x, rr_rel x x.
+\end{lstlisting}
+
 This is how reflexivity is defined in the standard library, and this is entirely sensible. The
-change from |Type| to |Prop| reflects the fact that |ReflexiveRelation| is now strictly a property,
+change from \lstinline|Type| to \lstinline|Prop| reflects the fact that \lstinline|ReflexiveRelation| is now strictly a property,
 proofs of which can be nicely kept opaque;%
-\footnote{Coq distinguishes between transparent and opague definitions. The former can be unfolded,
+\footnote{Coq distinguishes between transparent and opaque definitions. The former can be unfolded,
 whereas this is impossible for the latter. Since we never need to distinguish between two proofs of
-a propositions, Propositions can be Opague.}
+a propositions, Propositions can be Opaque.}
 we will return to this issue in Section~\ref{PI}.
-We can now define |EquivalenceRelation| without awkward equalities:
-\begin{code}
+We can now define \lstinline|EquivalenceRelation| without awkward equalities:
+\begin{lstlisting}
   Record EquivalenceRelation: Type :=
@@
     ; er_trans: TransitiveRelation er_trans er_rel }.
-\end{code}
+\end{lstlisting}
 
-But looking at |EquivalenceRelation|, we see that it now essentially has the same form as the
-original naive |ReflexiveRelation| record we started with. Clearly, equivalence relations
+But looking at \lstinline|EquivalenceRelation|, we see that it now essentially has the same form as the
+original naive \lstinline|ReflexiveRelation| record we started with. Clearly, equivalence relations
 and reflexive relations are analogous concepts, and by the same argument we
-conclude that |EquivalenceRelation| should really be defined as:
+conclude that \lstinline|EquivalenceRelation| should really be defined as:
 
-\begin{code}
+\begin{lstlisting}
 Record EquivalenceRelation (er_carrier: Type)
@@
   ; er_trans: TransitiveRelation er_trans er_rel }.
-\end{code}
+\end{lstlisting}
 
@@
 such as the quotient of a setoid,
-we have the same considerations as for |ReflexiveRelation| above.
+we have the same considerations as for \lstinline|ReflexiveRelation| above.
 
 At the risk of overstressing the point, let us look at how the same problem and solution keep
-appearing in composite structures. Suppose we use a definition of |Monoid| that bundles the
-relation. Then if we proceed to define |SemiRing| as a double |Monoid|, we face a dilemma:
-\begin{code}
+appearing in composite structures. Suppose we use a definition of \lstinline|Monoid| that bundles the
+relation. Then if we proceed to define \lstinline|SemiRing| as a double \lstinline|Monoid|, we face a dilemma:
+\begin{lstlisting}
   Record SemiRing A: Type :=
@@
     ; mult_comm: Commutative (mon_op mult_monoid)
-    ; mult_0_l: Π x: A, H
-\end{code}
-Do we use |mult_monoid|'s equality or |plus_monoid|'s equality in |H|?
+    ; mult_0_l: $\Pi$ x: A, H
+\end{lstlisting}
+
+Do we use \lstinline|mult_monoid|'s equality or \lstinline|plus_monoid|'s equality in \lstinline|H|?
 Again, for this definition to be usable at all, we would need to add an equality expressing that the
@@
 % We \emph{could} become over-eager and decide to unbundle the equivalence proof term as well:
-% \begin{code}
+% \begin{lstlisting}
+
 % Record Monoid (A: Type) (R: A → A → Prop)
 %   (e: Equivalence R) : Type := ...
-% \end{code}
+% \end{lstlisting}
+
 % However, this way if we happen to have two proofs |p| and |q| that a certain relation is an
@@
 structure record/class, to express that some collection of relations and
-operations posess the structure as a type class constraint seems to require manifest fields; see
+operations possess the structure as a type class constraint seems to require manifest fields; see
 Section~\ref{manifest}. With
@@
 there is a common perception that unbundling carriers, relations, and operations this way invariably
-leads either to unmanagably big algebraic expressions (involving endless projections and/or implicit
+leads either to unmanageably big algebraic expressions (involving endless projections and/or implicit
 arguments), or to unmanageably long and tedious enumeration of carriers/relations/operations, or to
@@
 
-Suppose we have defined |SemiGroup| in line with the aforementioned principles:
-\begin{code}
-Record SemiGroup (A: Type) (eq: A → A → Prop)
- (op: A → A → A): Prop := ...
-\end{code}
+Suppose we have defined \lstinline|SemiGroup| in line with the aforementioned principles:
+\begin{lstlisting}
+Record SemiGroup (A: Type) (eq: A $\to$ A $\to$ Prop)
+ (op: A $\to$ A $\to$ A): Prop := ...
+\end{lstlisting}
+
 Then our starting point is the obvious but awful:
-\begin{code}
+\begin{lstlisting}
 Lemma perm A eq op (P: SemiGroup A eq op) x y z:
                    eq (op (op x y) z) (op (op z y) x).
-\end{code}
+\end{lstlisting}
+
 As a first cleaning measure, we might hope to introduce some notations to let us use infix notation
-for the semigroup operator (say, |&|) and for equality. Unfortunately, at this point the names |eq|
-and |op| are just local names and there is no way to predeclare that they are to be associated with
-any notation. What we really need are canonical names for these notions, so that we may fashion
-notations for them.
+for the semigroup operator (say, \lstinline|&|) and for equality.
+Unfortunately, at this point the names \lstinline|eq| and \lstinline|op| are just local names and there is no way to predeclare that they are to be associated with any notation.
+What we really need are canonical names for these notions, so that we may fashion notations for them.
 
-This is a typical job for type classes, and they are used in this way in the |Clean| standard
-library~\cite{Clean}. Similarly, Haskell uses type classes to overload notations.
+This is a typical job for type classes, and they are used in this way in the \lstinline|Clean| standard
+library. Similarly, Haskell uses type classes to overload notations.
 We simply follow suit:
-\begin{code}
+\begin{lstlisting}
   Class Equiv A := equiv: relation A.
-  Class SemiGroupOp A := sg_op: A → A → A.
+  Class SemiGroupOp A := sg_op: A $\to$ A $\to$ A.
 
@@
   Infix "&" := sg_op (at level 50, left associativity).
-\end{code}
-We choose to use |&| here and reserve the notation |*| for ring multiplication.
+\end{lstlisting}
+
+We choose to use \lstinline|&| here and reserve the notation \lstinline|*| for ring multiplication.
 
@@
 We would have like to use a haskell style definition
-\begin{code}
+\begin{lstlisting}
 Class Equiv A := (=): relation A.
-\end{code}
+\end{lstlisting}
+
 Unfortunately, currently this is not supported in Coq.
@@
 records the way classes with any other number of fields are, but are instead turned into ordinary
-definitions. The effect of this is that |Equiv A| immediately reduces to |relation A|, and that the
-equiv `projection' is the identity function whose argument is inferred using instance resolution.
-Thus, for a very modest cost in indirection, we gain the ability to use canonical names, and with
-it, notations:
-\begin{code}
+definitions.
+The effect of this is that \lstinline|Equiv A| immediately reduces to \lstinline|relation A|, and that the equiv `projection' is the identity function whose argument is inferred using instance resolution.
+Thus, for a very modest cost in indirection, we gain the ability to use canonical names, and with it, notations:
+\begin{lstlisting}
 Record SemiGroup (A: Type) (eq: Equiv A)
@@
                  x & y & z = z & y & x.
-\end{code}
+\end{lstlisting}
+
 We call these type classes \emph{operational type classes}. We note that operational type classes
-allow us to avoid Coq's notation |scope| mechanism.
+allow us to avoid Coq's notation \lstinline|scope| mechanism.
 
@@
 On the other hand, it \emph{does} mean that all the carriers/relations/operations end up in the
-context. The context for |perm| looks like:
-\begin{code}
+context. The context for \lstinline|perm| looks like:
+\begin{lstlisting}
   A: Type
@@
   P: SemiGroup A eq op
-\end{code}
+\end{lstlisting}
 
@@
 likely not be problematic for any but the most complex formalizations), having to \emph{declare}
-these entities (like |eq| and |op| for |perm|) \emph{is} a bit of a chore. Fortunately, Coq
-provides a feature called implicit generalization, which is exactly what we need.
-With it, we can write |perm| as follows:
-\begin{code}
+these entities (like \lstinline|eq| and \lstinline|op| for \lstinline|perm|) \emph{is} a bit of a chore.
+Fortunately, Coq provides a feature called implicit generalization, which is exactly what we need.
+With it, we can write \lstinline|perm| as follows:
+\begin{lstlisting}
   Lemma perm `(SemiGroup A) x y z: x & y & z = z & y & x.
-\end{code}
+\end{lstlisting}
+
 Thus, we have reached the mathematical ideal we aimed for.
@@
 pack the proofs. This can be nicely organized by using canonical names. For instance, we define:
-\begin{code}
+\begin{lstlisting}
 Class SemiRing A {e: Equiv A}{plus: RingPlus A}{mult: RingMult A}
@@
   ; mult_0_l: forall x, 0 * x == 0 }.
-\end{code}
-where e.g.\ |RingPlus| is the operation class for the semigroup operator.
-The notation |:>| makes |semiring_mult_monoid| a coercion: when asked for a monoid it suffices to
+\end{lstlisting}
+
+where e.g.\ \lstinline|RingPlus| is the operation class for the semigroup operator.
+The notation \lstinline|:>| makes \lstinline|semiring_mult_monoid| a coercion: when asked for a monoid it suffices to
 provide a ring.
 
-\emph{Structure-as-parameters} helps setoid-rewriting: type class resolution
-can find the equivalence relation in the context.
-% A similar style should be possible for, say, the |Ring| tactic, instead of
-% declaring the ring structure by a separate command, we would rely on type class resolution to find
-% it.
-We note that |op| does not depend on the proof that |e| is an equivalence. As explained above we
-use Coq's implicit quantification (|`{}|) to avoid having to write all the parameters when
-\emph{stating} a theorem and Coq's maximally inserted implicit arguments to find the parameters when
-\emph{applying} a theorem. Both features are new in Coq and stem from the type class implementation.
+Having argued that the \emph{all-structure-as-parameters} approach \emph{can} be made practical, we enumerate some of the benefits that make it worthwhile.
+
+First, multiple inheritance becomes trivial: |SemiRing| inherits two |Monoid| structures on the same carrier and setoid relation, using ordinary named arguments to achieve ``manifest fields''; see section~\ref{manifest}.
+
+Second, because our terms are small and independent and never refer to proofs, we are invulnerable to concerns about efficiency and ambiguity of projection paths that plague existing solutions, obviating the need for extensions like the proposed coercion pullbacks~\cite{Hints}.
+
+Third, since our structural type classes are mere predicates, overlap between their instances is a non-issue.
+Together with the previous point, this gives us tremendous freedom to posit multiple structures on the same operations and relations, including ones derived implicitly via subclasses: by simply declaring a |SemiRing| class instance showing that a ring is a semiring, results about semirings immediately apply implicitly to any known ring, without us having to explicitly encode this relation in the hierarchy definition itself, and without needing any projection or translation of carriers or operations.
+
+Fourth, \emph{Structure-as-parameters} helps setoid-rewriting: type class resolution
+can find the equivalence relation in the context.\footnote{
+A similar style should be possible for, say, the |Ring| tactic, instead of
+declaring the ring structure by a separate command, we would rely on type class resolution to find it.}
+We note that \lstinline|op| does not depend on the proof that \lstinline|e| is an equivalence.
+As explained above we use Coq's implicit quantification (\lstinline|`{}|) to avoid having to write all the parameters when \emph{stating} a theorem and Coq's maximally inserted implicit arguments to find the parameters when \emph{applying} a theorem.
+Both features are new in Coq and stem from the type class implementation.
 
@@
 Manifest fields in records allow us to fix a field of a dependent record. For example, as in the
-example above, |Monoid with op=mult|. 
+example above, \lstinline|Monoid with op=mult|. 
 Luo~\cite{DBLP:conf/types/Luo08} has proposed coersive subtyping to implement manifest fields.
@@
 from the dependent record to its argument, thus mimicking manifest fields by using type classes.
-Since |op| is \emph{instantiated} with |mult|, we need not worry about the
-equality |op=mult|. This seems to be a general phenomenon. For instance, instead of searching
-for a |Monoid with op=mult| we would use type class resolution to find an instance of |Monoid mult|.
+Since \lstinline|op| is \emph{instantiated} with \lstinline|mult|, we need not worry about the
+equality \lstinline|op=mult|. This seems to be a general phenomenon. For instance, instead of searching
+for a \lstinline|Monoid with op=mult| we would use type class resolution to find an instance of \lstinline|Monoid mult|.
 We have thus addressed the problem of multiple inheritance.
@@
 \section{Universal algebra}\label{univ}
-After having implemented concrete algebraic structures, we generalize some of our constructions
-by implementing many-sorted universal algebra. A formalization of universal algebra
-already exists in Coq~\cite{DBLP:conf/tphol/Capretta99} and has recently been extended with
-more categorical results~\cite{dominguez2008formalizing}. We have conveniently redeveloped most of
-this using type classes. We define equational theories and their category of algebras. A quote
-function helps to translate a statement for an instance of a type class to a statement in the
-equational logic. We will come back to this in Section~\ref{quote}.
+Motivated originally by our desire to cleanly express interfaces for basic numeric data types such as $\N$ and $\Z$ in terms of their categorical characterization as initial objects in the categories of semirings and rings, respectively, we initially introduced only the very basics of category theory into our development, again using type classes where possible to achieve the same benefits mentioned before.
 
-% \subsection{Universal algebra continued}\label{univ2}
+Realizing that much code duplication for the various algebraic structures in the hierarchy could be avoided by employing universal algebra constructions, we then proceeded to formalize some of the theory of multisorted universal algebra and equational theories, using it to automatically construct varieties of algebras.
+We avoided existing formalizations~\cite{DBLP:conf/tphol/Capretta99,dominguez2008formalizing} of universal algebra, because we aimed to find out what level of elegance, convenience, and integration can be achieved using the state of the art technology (of which type classes are the most important instance).
+
+At the time of writing, our development includes a fully integrated formalization of a nontrivial portion of category theory and multisorted universal algebra, including various categories (e.g.\ the category |Cat| of categories, and generic variety categories which we instantiate to obtain the categories of monoids, semirings, and rings), functors (including automatically generated forgetful functors), natural transformations, adjunctions, initial models of equational theories constructed from term algebras, transference of proofs between isomorphic models of equational theories, subalgebras, congruences, quotients, products, and the first homomorphism theorem.
+
+There is an interesting interplay in our development between concrete algebraic structure type classes and their expressions on the one hand, and models of universal algebras and varieties instantiated with equational theories on the other.
+While occasionally a source of tension in that translation in either direction is not (yet) fully automatic, this duality also opens the door to the possibility of fully internalized implementations of generic tactics for algebraic manipulation, no longer requiring plugins.
+One missing piece in this puzzle is automatic quotation of concrete expressions into universal algebra expressions; see Section~\ref{quote}.
 
@@
 \begin{itemize}
- \item Definition of a signature Σ. Definition of a type (Σ,E). Σ-algebras and
-(Σ,$E$)-algebras
- \item A forgetful functor from (Σ,$E$)-algebras via Σ-algebras to Sets
+ \item Definition of a signature $\Sigma$. Definition of a type ($\Sigma$,E). $\Sigma$-algebras and
+($\Sigma$,$E$)-algebras
+ \item A forgetful functor from ($\Sigma$,$E$)-algebras via $\Sigma$-algebras to Sets
  \item Products and subalgebras
@@
 Coq provides different sorts. To simplify the discussion, we assume that there are only two sorts:
-|Prop| and |Type|. It occassionally happens that we want to prove a similar theorem for both sorts.
+\lstinline|Prop| and \lstinline|Type|. It occasionally happens that we want to prove a similar theorem for both sorts.
 To avoid code duplication, Coq provides universe polymorphism for inductive types.
-Unfortunately,
-definitions and fixpoints are not (yet?) polymorphic. To avoid code duplication we
-change some Fixpoints into
-Inductives. Concretely, the image $Im f:=\{b:B \mid \exists a. b=(f
-a)\}$ is informative (in |Type|) since we need the witness in the proof of the first homomorphism
-theorem.\marginpar{Does |relation| as a definition cause universe inconsistencies?}
+Unfortunately, definitions and fixpoints are not (yet?) polymorphic. To avoid code duplication we
+change some Fixpoints into Inductives. 
+Concretely, the image $Im f:=\{b:B \mid \exists a.
+b=(f a)\}$ is informative (in \lstinline|Type|) since we need the witness in the proof of the first homomorphism theorem.
+A similar problem raised by Coquand and Spiwack~\cite{coquand-towards}.
 % theory/ua_subalgebraT.v
@@
 
-In order to prove the first homophism theorem, we define a congruence.
-\begin{code}
+In order to prove the first homomophism theorem, we define a congruence.
+\begin{lstlisting}
 Class Congruence: Prop :=
-    { congruence_proper:> Π s, Proper (equiv ==> equiv ==> iff) (e s)
+    { congruence_proper:> $\Pi$ s, Proper (equiv ==> equiv ==> iff) (e s)
     ; congruence_quotient:> @Algebra et v e _
     }.
-\end{code}
-Alternatively, we could have stated the compatibility of the relation |e| by stating that the
+\end{lstlisting}
+
+Alternatively, we could have stated the compatibility of the relation \lstinline|e| by stating that the
 relation as a set of pairs is a subalgebra in the product algebra. We prove that both approaches
@@
 elements from equivalence classes where they should never have been put in the first place.}
- In this way we profit from
-the infrastructure for setoids: A congruence relation is coarser than the setoid relation, as
-indicated by the use of |Proper| above.
+ In this way we profit from the infrastructure for setoids: A congruence relation is coarser than the setoid relation, as indicated by the use of \lstinline|Proper| above.
 
@@
 variables. The term algebra is the initial $\tau$-algebra. Categorically: we have defined a
-left-adjoint to the forgetful functor.\marginpar{Did we proof this?}
+left-adjoint to the forgetful functor.
 A formal proof connects the categorical and the algebraic definition.
@@
 elements of the algebra and of the subalgebra, i.e.\ terms of different types may be equal.
-\begin{code}
+\begin{lstlisting}
   Fixpoint heq {o}: op_type carrier o -> op_type v o -> Prop :=
@@
     end.
-\end{code}
+\end{lstlisting}
 
 We connect the abstract theory to the concrete theory. Concretely, let $\tau_m$ be the theory
-of monoids. Given a monoid, we can construct the corresponding $\tau_m$-algebra and an
-object in the category of $\tau$-algebras. Conversely, given an object in the category/variety, we
-construct an instance of the |Monoid| type class.\marginpar{Update code}
+of monoids. Given a monoid, we can construct the corresponding $\tau_m$-algebra. 
+Conversely, given a $\tau_m$-algebra, we construct an instance of the \lstinline|Monoid| type class.
 %varieties/monoid.v
-\begin{code}
+\begin{lstlisting}
 Variable o: variety.Object theory.
@@
 Global Instance from_object: Monoid (o tt).
-\end{code}
+\end{lstlisting}
+
 In the last line, instance resolution `automatically' finds the operation and the unit specified in
-the lines before.\marginpar{Why this way?}
+the lines before.
 
@@
 adjunctions, monads,\ldots We follow the usual type theoretical definition of a
-category~\cite{saibi1995constructive}.\marginpar{Update code}
+category~\cite{saibi1995constructive}.
 
-\newcommand{\rightarrowtext}{\ensuremath{\rightarrow}}
-\newcommand{\textocirc}{\ensuremath\circ}
+% \newcommand{\rightarrowtext}{\ensuremath{\rightarrow}}
+% \newcommand{\textocirc}{\ensuremath\circ}
 %interfaces/abstract_algebra.v
 
-CODE MISSING.
-% \begin{code}
-% Class Category O `{!Arrows O} `{Π x y: O, Equiv (x ⟶ y)}
-%   `{!CatId O} `{!CatComp O}: Prop :=
-%   { arrow_equiv:> Π x y: O, Setoid (x ⟶ y)
-%   ; comp_proper:> Π x y z,
-%       Proper (equiv ==> equiv ==> equiv)%signature (@comp _ _ _ x y z)
-%   ; comp_assoc (w x y z: O) (a: w ⟶ x) (b: x ⟶ y) (c: y ⟶ z):
-%       c ◎ (b ◎ a) = (c ◎ b) ◎ a
-%   ; id_l `(a: x ⟶ y): cat_id ◎ a = a
-%   ; id_r `(a: x ⟶ y): a ◎ cat_id = a }.
-% \end{code}
+\begin{lstlisting}
+Class Category O`{!Arrows O}`{$\Pi$ x y:O, Equiv(x $\to$ y)}
+                `{!CatId O}`{!CatComp O}:Prop :=
+  { arrow_equiv:> $\Pi$ x y, Setoid (x $\to$ y)
+  ; comp_proper:> $\Pi$ x y z,
+    Proper (equiv $\Rightarrow$ equiv $\Rightarrow$ equiv)%signature
+    (comp: (y $\to$ z) $\to$ (x$\to$ y) $\to$ x $\to$ z)
+  ; comp_assoc w x y z (a: w $\to$ x) (b: x $\to$ y) (c: y $\to$ z):
+      c $\circ$ (b $\circ$ a) = (c $\circ$ b) $\circ$ a
+  ; id_l `(a: x $\to$ y): cat_id $\circ$ a = a
+  ; id_r `(a: x $\to$ y): a $\circ$ cat_id = a }.
+\end{lstlisting}
+
 This is based on the 2-categorical idea that we have equality on arrows, but not on objects.
@@
 Natural transformations are formalized as:
-CODE MISSING.
-% \begin{code}
-%   Context `{Category C} `{Category D}
-%   `{!Functor (F: C → D) Fa} `{!Functor (G: C → D) Ga}.
-% 
-%   Class NaturalTransformation (η: Π c, F c ⟶ G c): Prop :=
-%     natural: Π (x y: C) (f: x ⟶ y), η y ◎ fmap F f = fmap G f ◎ η x.
-% \end{code}
+\begin{lstlisting}
+   Context `{Category C} `{Category D}
+   `{!Functor (F: C $\to$ D) Fa} `{!Functor (G: C $\to$ D) Ga}.
+   Class NaturalTransformation ($\eta$: $\Pi$ c, F c $\to$ G c): Prop :=
+     natural: $\Pi$ (x y: C) (f: x $\to$ y), $\eta$ y $\circ$ fmap F f = fmap G f $\circ \eta$ x.
+\end{lstlisting}
 
-Here |fmap F| explicitly refers to the arrow part of the functor |F|.
+Here \lstinline|fmap F| explicitly refers to the arrow part of the functor \lstinline|F|.
 In mathematics one would apply $F$ to both objects and arrows ($F x$, $F f$). However, this does
-not seem to fit with type theory.\marginpar{Explain why.}
-
-The proofs are short and follow the textbook easily.\marginpar{really?}
+not seem to fit with type theory as the type of \lstinline|fmap F| depends on the object part of
+\lstinline|F|.
 
@@
 products, and hence co-products and the category of $\tau$-algebras, for a given type $\tau$.
+We prove that two definitions of adjunction are equivalent. The proofs are concise and follow the
+textbooks.
 
 Universe polymorphism allows us to define the category of categories. However, we need to avoid
-making |Relation| a |Definition|, since |Definitions| are not universe polymorphic.
+making \lstinline|Relation| a \lstinline|Definition|, since \lstinline|Definitions| are not universe polymorphic.
 
@@
 \section{Interfaces using category theory}\label{interfaces}\label{modul}
-The current work was stimulated by our desire~\cite{Riemann} to move between several
-implementations in a flexible way. In haskell type class polymorphism is used for this purpose.
-Hence, we do the same in type theory.
-
 We have characterized the naturals as the initial object in the category of semirings and derived
@@
 semiring structure, is assigned priority 10.
-\begin{code}
+\begin{lstlisting}
   Global Program Instance: NatDistance N | 10 := ...
-\end{code}
+\end{lstlisting}
 
-Similarly, we use type classes to flexibly combine two representations of a dense set of the
-reals.
-Let |Q| be the rationals and |D| be the dyadic rationals, i.e.\ rationals where the denominator is
-a power of 2. Let |inj:D>->Q| be a coercion. Then we want to define heterogeneous multiplication
-as follows:
-
-%There is a slight problem when defining the forgetful functor from the category of
-%$\Sigma$-algebras to Sets. This problem is caused by our use of the category of all
-%categories\footnote{Perhaps we should have used higher categories for this.}.
-%We conjecture that this problem disappears when Coq's universe polymorphism is extended to
-%definitions.
+In future work we aim to base our development of the reals
+on an abstract dense set, allowing us to use the efficient dyadic
+rationals~\cite{boldo2009combining} as a base for exact real number computation in
+Coq~\cite{Riemann,Oconnor:real}. The use of category theory has been important in these developments.
 
@@
 We define a term language for monoids:
-\begin{code}
+\begin{lstlisting}
 Inductive Expr (V: Type) := Mult (a b: Expr V) | Zero | Var (v: V).
-\end{code}
-This defines a monad on the category of |Type|s with extensional functions.\marginpar{Proof this
-formally?}
-The expression type is parameterized over the set of variable indices. Hence, we diverge
- from~\cite{Hints}, which uses |nat| for variables thereby introducing bounds problems and
- dummy variables.
+\end{lstlisting}
+
+This defines a monad on the category of \lstinline|Type|s with extensional functions.\marginpar{Proof in classquote}
+The expression type is parametrized over the set of variable indices. Hence, we diverge from~\cite{Hints}, 
+which uses \lstinline|nat| for variables thereby introducing bounds problems and dummy variables.
 
 \noindent An expression is only meaningful in the context of a variable
-assignment:\footnote{Categorically: |nat| is
-an |Expr|-algebra and |eval| is |map vars| composed with the algebra map.}
-\begin{code}
+assignment:\footnote{Categorically: \lstinline|nat| is
+an \lstinline|Expr|-algebra and \lstinline|eval| is \lstinline|map vars| composed with the algebra map.}
+\begin{lstlisting}
 Definition Value := nat.
@@
   end.
-\end{code}
+\end{lstlisting}
+
 %
 Monads are trees with grafting~\cite{MonadsGrafting}: the tree monad
-$TX:=1+T^2X$, may be seen as the prototypical monad. Indeed, the |Expr| monad, i.e.\ the free
-|Expr|-algebra monad, behaves much like the tree monad. We now provide a direct definition of the
+$TX:=1+T^2X$, may be seen as the prototypical monad. Indeed, the \lstinline|Expr| monad, i.e.\ the free
+\lstinline|Expr|-algebra monad, behaves much like the tree monad. We now provide a direct definition of the
 corresponding bind operation, the arrows in the Kleisli category.
 
-\begin{code}
+\begin{lstlisting}
 Fixpoint bind {V W} (f: V -> Expr W) (e: Expr V): Expr W :=
@@
   end.
-\end{code}
+\end{lstlisting}
 
@@
 %
-\begin{code}
+\begin{lstlisting}
 Definition novars: Vars False := False_rect _.
@@
   fun i => if i then a j else b j.
-\end{code}
+\end{lstlisting}
 
 \noindent These last two combinators are the `constructors' of an implicitly defined subset of
- Gallina terms, representing heaps, for which we implement syntactic lookup with type classes.
-The heap can also be defined explitly, with no essential change in the code.
-Given a heap and value, |Lookup| instances give the value's index in the heap:
+ Galina terms, representing heaps, for which we implement syntactic lookup with type classes.
+The heap can also be defined explicitly, with no essential change in the code.
+Given a heap and value, \lstinline|Lookup| instances give the value's index in the heap:
 % \marginpar{A
 % canonical structures approach would allow us to do this in Coq?}
-\begin{code}
+\begin{lstlisting}
   Class Lookup {A} (x: Value) (f: Vars A) :=
     { lookup: A; lookup_correct: f lookup = x }.
-\end{code}
+\end{lstlisting}
+
 % Context (x: Value) {A B} (va: Vars A) (vb: Vars B).
@@
 access it by indexing the merged heap (and vice versa)
-\begin{code}
+\begin{lstlisting}
   Global Instance lookup_left `{!Lookup x va}: Lookup x (merge va vb)
     := { lookup := inl (lookup x va) }.
-\end{code}
+\end{lstlisting}
+
 If the heap is just a singlevar, we can easily index it.
-\begin{code}
+\begin{lstlisting}
   Global Program Instance: Lookup x (singlevar x) := { lookup := tt }.
-\end{code}
+\end{lstlisting}
 
 One useful operation we need before we get to Quote relates to variables and expression
- evaluation. As its name suggests, |map_var| maps an expression's variable indices.
-This reindexing function is the the |map| of the |Expr|-monad.
+ evaluation. As its name suggests, \lstinline|map_var| maps an expression's variable indices.
+This reindexing function is the the \lstinline|map| of the \lstinline|Expr|-monad.
 
-\begin{code}
+\begin{lstlisting}
 Definition map_var {V W: Type} (f: V -> W):
@@
     end.
-\end{code}
-In |Quote| below, the idea is that |V, l|, and |n| are all input variables, while |V'| and |r| are
-output variables (in the sense that we will rely on unification to generate them). |V| and |l|
-represent the current heap, |n| represents the value we want to quote, and |V'| and |r'| represent
+\end{lstlisting}
+
+In \lstinline|Quote| below, the idea is that \lstinline|V, l|, and \lstinline|n| are all input variables, while \lstinline|V'| and \lstinline|r| are
+output variables (in the sense that we will rely on unification to generate them). \lstinline|V| and \lstinline|l|
+represent the current heap, \lstinline|n| represents the value we want to quote, and \lstinline|V'| and \lstinline|r'| represent
 the heap of newly encountered variables during the quotation.
   This explains the type of quote: it is an expression that refers either to variables from
-the old heap, or to newly encountered variables. Finally, |eval_quote| is the usual correctness
-property, which now merges the two heaps.
+the old heap, or to newly encountered variables. Finally, \lstinline|eval_quote| is the usual correctness property, which now merges the two heaps.
 
-\begin{code}
+\begin{lstlisting}
   Class Quote {V} (l: Vars V) (n: Value) {V'} (r: Vars V') :=
@@
     ; eval_quote: @eval (V+V') (merge l r) quote = n }.
-\end{code}
+\end{lstlisting}
 
-Our first instance for |Zero| is easy. The `novars' in the result type reflects the fact that no
+Our first instance for \lstinline|Zero| is easy. The `novars' in the result type reflects the fact that no
 new variables are encountered.
-\begin{code}
+\begin{lstlisting}
   Global Program Instance quote_zero V (v: Vars V):
@@
   { quote := Zero }.
-\end{code}
+\end{lstlisting}
 
@@
 shuffles heaps around.
- The third line contains some |map_var|'s because the heap shuffling must be
-reflected in the variable indices, but apart from that it's just constructing a |Mult| term with
+ The third line contains some \lstinline|map_var|'s because the heap shuffling must be
+reflected in the variable indices, but apart from that it's just constructing a \lstinline|Mult| term with
 quoted subterms.
 
-\begin{code}
+\begin{lstlisting}
 Global Program Instance quote_mult V (v: Vars V)
@@
   { quote := Mult (map_var bla (quote n)) (map_var sum_assoc (quote m)) }.
-\end{code}
+\end{lstlisting}
+
   Now follows the instance where we recognize values that are already in the heap. This
    is expressed by the Lookup requirement, which will only be fulfilled if the Lookup instances
-   defined above can find the value in the heap. The novars in the |Quote v x novars| result
+   defined above can find the value in the heap. The novars in the \lstinline|Quote v x novars| result
    reflects that this quotation does not generate new variables.
 
-\begin{code}
+\begin{lstlisting}
 Global Program Instance quote_old_var
@@
   Quote v x novars | 8 := { quote := Var (inl (lookup x v)) }.
-\end{code}
+\end{lstlisting}
 
 \noindent Finally, the instance for new variables. We give this a lower priority so that it is only
-used if Lookup fails. The |8| in the previous example is a random number less than |9| below. A
-similar method is used in Coq's notation mechanism.
+used if Lookup fails. The \lstinline|8| in the previous example is a random number less than \lstinline|9| below. A similar method is used in Coq's notation mechanism.
 % We recommend a change in
@@
 
-\begin{code}
+\begin{lstlisting}
 Global Program Instance quote_new_var V (v: Vars V) x:
   Quote v x (singlevar x) | 9 := { quote := Var (inr tt) }.
-\end{code}
+\end{lstlisting}
 
-\noindent As announced this provides a light-weight quoting function\footnote{|eval_quote'| is a
-modification of |eval_quote| which works in the empty context.}:
-\begin{code}
-Goal forall x y (P: Value -> Prop), P ((x * y) * (x * 0)).
+\noindent As announced this provides a light-weight quoting
+function\footnote{\lstinline|eval_quote'| is a
+modification of \lstinline|eval_quote| which works in the empty context.}:
+\begin{lstlisting}
+Goal forall x y (P: Value $\to$ Prop), P ((x * y) * (x * 0)).
   intros. rewrite <- (eval_quote' _).
-\end{code}
-This turns the goal into | (eval variable_pack quote)|.
+\end{lstlisting}
 
-It would be interesting to use this technique to implement the |congruence| tactic, which implements
+This turns the goal into \lstinline|(eval variable_pack quote)|.
+
+It would be interesting to use this technique to implement the \lstinline|congruence| tactic, which implements
 the congruence closure algorithm~\cite{corbineau2007deciding}. One would naturally obtain a tactic
@@
 structures to derive the setoid equality on the natural
-numbers in the following example |(S x)==(S y)|. In this case type classes provide superior proof
+numbers in the following example \lstinline|(S x)==(S y)|. In this case type classes provide superior proof
 terms. Canonical structures give
-\begin{code}
+\begin{lstlisting}
 @equiv(Build_Setoid nat (@eq nat)(@eq_equivalence nat))(S x)(S y)
-\end{code}
-which includes an explicit proof that |(@eq nat)| is an equivalence,
-whereas we obtain |@equiv nat (@eq nat) (S x) (S y)|.
+\end{lstlisting}
+
+which includes an explicit proof that \lstinline|(@eq nat)| is an equivalence,
+whereas we obtain \lstinline|@equiv nat (@eq nat) (S x) (S y)|.
 
@@
 We use type classes to deduce the relevant monoid operation and write:
-\begin{code}
+\begin{lstlisting}
 Definition seq_sum
-  `{Sequence A T} `{RingPlus A} `{z: RingZero A}: T -> A
+  `{Sequence A T} `{RingPlus A} `{z: RingZero A}: T $\to$ A
   := @seq_to_monoid A T _ A ring_plus z id.
   Eval compute in seq_sum [3; 2].
-\end{code}
-\marginpar{distributivity?}
+\end{lstlisting}
+\marginpar{Prove distributivity?}
 
@@
 We have mentioned proof irrelevance a number of times before. It is an important theme in our
-approach. The |Program| machinery allows one to conveniently write programs with Hoare-style
+approach. The \lstinline|Program| machinery allows one to conveniently write programs with Hoare-style
 pre- and post-conditions. I.e.\ functions $f: \{ A \mid P \} \to \{ B \mid Q \}$. Both side
 conditions
-are intended to be proof irrelevant, they are in |Prop|. Presently, Coq does not support such subset
-types, thus forcing the user to manually prove that |f| does not depend on the proof of |P|.
+are intended to be proof irrelevant, they are in \lstinline|Prop|. Presently, Coq does not support such subset
+types, thus forcing the user to manually prove that \lstinline|f| does not depend on the proof of \lstinline|P|.
 The addition of proof irrelevance and subset types to Coq is pursued by
-Barras~\cite{Barras:subset,Werner}. 
+Barras~\cite{Barras:subset,Werner08}. 
 % Squash is a monad and allows many constructions: subsets, unions, images, heterogeneous equality,
@@
 \cline{1-3}
-(a_1,a_2) &:\Sigma T
+(a_1,a_2) &: \Sigma T
 \end{align*}
 
-For our methodology to work we have to be able to transform the occuring record into a telescopic
-subset $\{ A \mid  B\}$. We moreover conjecture that the terms in |B| only depend on the
-variables in |A|, but not in |B| itself. However, we do not need this.\\
+For our methodology to work we have to be able to transform the occurring record into a telescopic
+subset $\{ A \mid  B\}$. We moreover conjecture that the terms in \lstinline|B| only depend on the
+variables in \lstinline|A|, but not in \lstinline|B| itself. However, we do not need this.\\
 
-We observed that we were naturally lead to the following methodology: Let φ be a statement in type
-theory, i.e.\ an alternation of Πs and Σs. By the axiom of choice, we first Skolemize to $Σ h. Πa.P$
-and then transfer the variable $h$ to the parameters. For example, consider surjectivity. This is
-skolemized to a surjection with a right inverse, i.e.\ a split epi. We then move the right inverse
+We observed that we were naturally lead to the following methodology: Let $\phi$ be a statement in type
+theory, i.e.\ an alternation of $\Pi$s and $\Sigma$s.
+By the axiom of choice, we first Skolemize to $\Sigma h.
+\Pi a.P$ and then transfer the variable $h$ to the parameters.
+For example, consider surjectivity.
+This is skolemized to a surjection with a right inverse, i.e.\ a split epi.
+We then move the right inverse
 to the parameters.
@@
 \subsection{Propositions as [ ]-Types}
-As originally concieved~\cite{ITT,CMCP}, Propositions are Types and we can extract information from
+As originally conceived~\cite{ITT,CMCP}, Propositions are Types and we can extract information from
 them. In practice, it is convenient to assume that all proofs of a proposition are equal: the
-principle of proof irrelevance. These two views clearly contradict eachother. In Coq, a middle way
-was choosen, we are forbidden to extract information from proofs, but proofs are not identified.
+principle of proof irrelevance. These two views clearly contradict each other. In Coq, a middle way
+was chosen, we are forbidden to extract information from proofs, but proofs are not identified.
 This allows us to extract the program we want, but we want more:
 We would like to use Coq as a dependently typed programming language. To do so,
-one would like to add the principle of proof irrelevance to Coq. A promissing proposal is the
+one would like to add the principle of proof irrelevance to Coq. A promising proposal is the
 Calculus of implicit constructions with implicit
-$\Sigma$-types~\cite{miquel2001implicit,barras2008implicit,barrasSigma}. Such a change would fit
-naturally with the Hoare-style separation between programs and proofs which is used by the |Program|
-machinery.
+$\Sigma$-types~\cite{miquel2001implicit,barras2008implicit,Bernardo}. Such a change would fit
+naturally with the Hoare-style separation between programs and proofs which is used by the \lstinline|Program| machinery.
 
@@
 construction $[P]:=\{\top\mid p:P\}$ squashes all the information about how we proved $P$. In Coq,
-the [ ]-types are isomorphic to the type of sort |Prop|. However,
+the [ ]-types are isomorphic to the type of sort \lstinline|Prop|. However,
 there are places where we needed proof \emph{relevance}; e.g.\ in the proof of the first isomorphism
@@
 similar expressive powers. Canonical structures, being tied to the unifier, are more robust, but
-seem to require more inguinity. Type classes are easier to use. Unification hints, may be a
+seem to require more ingenuity. Type classes are easier to use. Unification hints, may be a
 good generalization, but are absent from Coq. We encourage their inclusion to the
-Coq system.\marginpar{for instance what about control in
-proof/instance search?\\ or what about its efficiency?}
+Coq system.
 
 An obvious topic for future research is the extension from equational logic to partial Horn
-logic~\cite{palmgren2007partial}. Another topic would be to fully, but practically, embrase the
+logic~\cite{palmgren2007partial}. Another topic would be to fully, but practically, embrace the
 categorical approach to universal algebra~\cite{pitts2001categorical}.
 
-According to |coqwc|, our development consists of 3103 lines of specifications and 707 lines of
-proofs.\marginpar{update,targz}
+According to \lstinline|coqwc|, our development consists of 4895 lines of
+specifications and 769 lines of proofs.\marginpar{update,targz}
 
@@
 This research was stimulated by the new possibilities provided by the introduction of type classes
-in Coq. In fact, this seems to be the first substantial development that tries to exploid all their
+in Coq. In fact, this seems to be the first substantial development that tries to exploit all their
 possibilities. As a consequence, we found many small bugs and unintended behavior in the type

### 2010-05-07 defManSSR.tex
Adding defManSSR.tex

%=======================================================================
% Définit le style ssr Gallina pour les listings (Assia Mahboubi 2007)

\lstdefinelanguage{SSR} {

% Anything betweeen $ becomes LaTeX math mode
mathescape=true,						
% Comments may or not include Latex commands
texcl=false,

% Vernacular commands
morekeywords=[1]{
Section, Module, End, Require, Import, Export,
Variable, Variables, Parameter, Parameters, Axiom, Hypothesis, Hypotheses,
Notation, Local, Tactic, Reserved, Scope, Open, Close, Bind, Delimit,
Definition, Let, Ltac, Fixpoint, CoFixpoint, Add, Morphism, Relation,
Implicit, Arguments, Set, Unset, Contextual, Strict, Prenex, Implicits,
Inductive, CoInductive, Record, Structure, Canonical, Coercion,
Theorem, Lemma, Corollary, Proposition, Fact, Remark, Example,
Proof, Goal, Save, Qed, Defined, Hint, Resolve, Rewrite, View,
Search, Show, Print, Printing, All, Graph, Projections, inside, outside},

% Gallina
morekeywords=[2]{forall, exists, exists2, fun, fix, cofix, struct,
      match, with, end, as, in, return, let, if, is, then, else,
      for, of, nosimpl},

% Sorts
morekeywords=[3]{Type, Prop},

% Various tactics, some are std Coq subsumed by ssr, for the manual purpose
morekeywords=[4]{
         pose, set, move, case, elim, apply, clear,
            hnf, intro, intros, generalize, rename, pattern, after,
	    destruct, induction, using, refine, inversion, injection,
         rewrite, congr, unlock, compute, ring, field, fourier, 
            replace, fold, unfold, change, cutrewrite, simpl,
         have, suff, wlog, suffices, without, loss, nat_norm,
            assert, cut, trivial, revert, bool_congr, nat_congr,
	 symmetry, transitivity, auto, split, left, right, autorewrite},

% Terminators
morekeywords=[5]{
         by, done, exact, reflexivity, tauto, romega, omega,
         assumption, solve, contradiction, discriminate},

% Control
morekeywords=[6]{do, last, first, try, idtac, repeat},

% Various symbols
% For the ssr manual we turn off the prettyprint of formulas
% literate=	
%	{->}{{$\rightarrow\,$}}2		
% 	{->}{{\tt ->}}3
%	{<-}{{$\leftarrow\,$}}2	
% 	{<-}{{\tt <-}}2			
% 	{>->}{{$\mapsto$}}3		
% 	{<=}{{$\leq$}}1			
% 	{>=}{{$\geq$}}1		
% 	{<>}{{$\neq$}}1
% 	{/\\}{{$\wedge$}}2	
% 	{\\/}{{$\vee$}}2 		
% 	{<->}{{$\leftrightarrow\;$}}3
% 	{<=>}{{$\Leftrightarrow\;$}}3
% 	{:nat}{{$~\in\mathbb{N}$}}3	
%	{fforall\ }{{$\forall_f\,$}}1		
%	{forall\ }{{$\forall\,$}}1		
%	{exists\ }{{$\exists\,$}}1
% 	{negb}{{$\neg$}}1
% 	{spp}{{:*:\,}}1
% 	{~}{{$\sim$}}1
% 	{\\in}{{$\in\;$}}1
% 	{/\\}{$\land\,$}1
% 	{:*:}{{$*$}}2
%	{=>}{{$\,\Rightarrow\ $}}1
% 	{=>}{{\tt =>}}2
% 	{:=}{{{\tt:=}\,\,}}2
% 	{==}{{$\equiv$}\,}2
% 	{!=}{{$\neq$}\,}2
% 	{^-1}{{$^{-1}$}}1
% 	{elt'}{elt'}1
%       {=}{{\tt=}\,\,}2
%       {+}{{\tt+}\,\,}2,

% Comments delimiters, we do turn this off for the manual
comment=[s]{(*}{*)},

% Spaces are not displayed as a special character
showstringspaces=false,

% String delimiters
morestring=[b]",
morestring=[d],

% Size of tabulations
tabsize=3,

% Enables ASCII chars 128 to 255
extendedchars=true,

% Case sensitivity
sensitive=true,

% Automatic breaking of long lines
breaklines=true,

% Default style fors listings
basicstyle=\sffamily,

% Position of captions is bottom
captionpos=b,

% Full flexible columns 
columns=[l]fullflexible,

% Style for (listings') identifiers
identifierstyle={\sffamily\color{black}},
% Note : highlighting of Coq identifiers is done through a new
% delimiter definition through an lstset at the begining of the
% document. Don't know how to do better.

% Style for declaration keywords
keywordstyle=[1]{\sffamily\color{dkviolet}},

% Style for gallina keywords
keywordstyle=[2]{\sffamily\color{dkgreen}},

% Style for sorts keywords
keywordstyle=[3]{\sffamily\color{lightblue}},

% Style for tactics keywords
keywordstyle=[4]{\sffamily\color{dkblue}},

% Style for terminators keywords
keywordstyle=[5]{\sffamily\color{red}},

%Style for iterators
%keywordstyle=[6]{\ttfamily\color{dkpink}},

% Style for strings
stringstyle=\sffamily,

% Style for comments
commentstyle={\ttfamily\footnotesize},

}

### 2010-05-07 jar.tex
Small changes

@@
 \begin{lstlisting}
-  Record ReflexiveRelation (carrier: Type)
-                           (rel: relation carrier): Prop :=
+  Record ReflexiveRelation (carrier: Type) (rel: relation carrier): Prop :=
     { rr_proof: $\Pi$ x, rr_rel x x }.
@@
 \begin{lstlisting}
-  Definition ReflexiveRelation (carrier: Type)
-                               (rel: relation carrier): Prop :=
+  Definition ReflexiveRelation (carrier: Type) (rel: relation carrier): Prop :=
     $\Pi$ x, rr_rel x x.
@@
 whereas this is impossible for the latter. Since we never need to distinguish between two proofs of
-a propositions, Propositions can be Opaque.}
+a propositions, propositions can be opaque.}
 we will return to this issue in Section~\ref{PI}.
@@
 So far, we have only looked at the definitions of the structures themselves, and have not yet
-considered the perspective of someone trying to prove some theorems about these abstract structures.
+considered their use when proving theorems about these structures.
 We would like to approximate the common mathematical vernacular, where one
@@
 
-We choose to use \lstinline|&| here and reserve the notation \lstinline|*| for ring multiplication.
-
-We use = for setoid equality, since Leibniz equality seems less important for the formalization of
-mathematics. More precisely, setoid equality is the basic concept. In special case, one can prove
+We choose to use \lstinline|&| here and reserve the notation \lstinline|*| for ring multiplication. We use = for setoid equality, since Leibniz equality seems less important for the formalization of
+mathematics. More precisely, setoid equality is the basic concept. In some special cases, one can prove
 that setoid equality coincides with Leibniz equality.
-
+%Listings don't seem to work will with footnotes.
 We would have like to use a haskell style definition
@@
 \end{lstlisting}
-
 Unfortunately, currently this is not supported in Coq.
@@
 
-It is important to point out that although the expression now \emph{looks} like it is bound to some
+Although the expression now \emph{looks} like it is bound to some
 specific semigroup structure (which with traditional approaches would imply projections from bundles
@@
 where e.g.\ \lstinline|RingPlus| is the operation class for the semigroup operator.
-The notation \lstinline|:>| makes \lstinline|semiring_mult_monoid| a coercion: when asked for a monoid it suffices to
+The notation \lstinline|:>| makes \lstinline|semiring_mult_monoid| a subclass: when asked for a monoid it suffices to
 provide a ring.
@@
 
-First, multiple inheritance becomes trivial: |SemiRing| inherits two |Monoid| structures on the same carrier and setoid relation, using ordinary named arguments to achieve ``manifest fields''; see section~\ref{manifest}.
+First, multiple inheritance becomes trivial: \lstinline|SemiRing| inherits two \lstinline|Monoid| structures on the same carrier and setoid relation, using ordinary named arguments to achieve ``manifest fields''; see section~\ref{manifest}.
 
@@
 Third, since our structural type classes are mere predicates, overlap between their instances is a non-issue.
-Together with the previous point, this gives us tremendous freedom to posit multiple structures on the same operations and relations, including ones derived implicitly via subclasses: by simply declaring a |SemiRing| class instance showing that a ring is a semiring, results about semirings immediately apply implicitly to any known ring, without us having to explicitly encode this relation in the hierarchy definition itself, and without needing any projection or translation of carriers or operations.
+Together with the previous point, this gives us tremendous freedom to posit multiple structures on the same operations and relations, including ones derived implicitly via subclasses: by simply declaring a \lstinline|SemiRing| class instance showing that a ring is a semiring, results about semirings immediately apply implicitly to any known ring, without us having to explicitly encode this relation in the hierarchy definition itself, and without needing any projection or translation of carriers or operations.
 
@@
 can find the equivalence relation in the context.\footnote{
-A similar style should be possible for, say, the |Ring| tactic, instead of
+A similar style should be possible for, say, the \lstinline|Ring| tactic, instead of
 declaring the ring structure by a separate command, we would rely on type class resolution to find it.}

### 2010-05-19 jar.tex
Now reads better.

@@
 those of telescopes or packed classes. In our experience, this has been relatively
-harmless\footnote{Coq's data structure for contexts is not very efficient. Gonthier fears that this
+harmless\footnote{However, Coq's data structure for contexts is not very efficient. Gonthier fears that this
 may be a bottleneck for huge developments. It seems that the data structure chosen
@@
 
-Manifest fields in records allow us to fix a field of a dependent record. For example, as in the
-example above, \lstinline|Monoid with op=mult|. 
-Luo~\cite{DBLP:conf/types/Luo08} has proposed coersive subtyping to implement manifest fields.
-It was conjectured~\cite{Hints} that implementing the algebraic hierarchy with type classes would
+Manifest fields in records allow us to fix a field of a dependent record. For example, in the
+example above one would write \lstinline|Monoid with op=mult|. 
+Luo~\cite{DBLP:conf/types/Luo08} has proposed coercive subtyping to implement manifest fields.
+It was conjectured~\cite{Hints} that implementing the algebraic hierarchy with type classes in this way would
 be awkward due to problems with equality. We avoided these problems by shifting the fields
@@
 
-At the time of writing, our development includes a fully integrated formalization of a nontrivial portion of category theory and multisorted universal algebra, including various categories (e.g.\ the category |Cat| of categories, and generic variety categories which we instantiate to obtain the categories of monoids, semirings, and rings), functors (including automatically generated forgetful functors), natural transformations, adjunctions, initial models of equational theories constructed from term algebras, transference of proofs between isomorphic models of equational theories, subalgebras, congruences, quotients, products, and the first homomorphism theorem.
+At the time of writing, our development includes a fully integrated formalization of a nontrivial portion of category theory and multisorted universal algebra, including various categories (e.g.\ the category \lstinline|Cat| of categories, and finitary algebraic categories  defined by a theory which we instantiate to obtain the categories of monoids, semirings, and rings), functors (including automatically generated forgetful functors), natural transformations, adjunctions, initial models of equational theories constructed from term algebras, transference of proofs between isomorphic models of equational theories, subalgebras, congruences, quotients, products, and the first homomorphism theorem~\cite{meinke1993universal}.
 
-There is an interesting interplay in our development between concrete algebraic structure type classes and their expressions on the one hand, and models of universal algebras and varieties instantiated with equational theories on the other.
-While occasionally a source of tension in that translation in either direction is not (yet) fully automatic, this duality also opens the door to the possibility of fully internalized implementations of generic tactics for algebraic manipulation, no longer requiring plugins.
-One missing piece in this puzzle is automatic quotation of concrete expressions into universal algebra expressions; see Section~\ref{quote}.
+We consider the first homomorphism theorem in more detail:
+\begin{theorem}[First homomorphism theorem]
+If $A$ and $B$ are algebras, and $f$ is a homomorphism from $A$ to $B$, then the equivalence relation defined by $a\sim b$ if and only if $f(a)=f(b)$ is a congruence on $A$, and the algebra $A/\sim$ is isomorphic to the image of $f$, which is a subalgebra of B.
+\end{theorem}
 
-We have developed basic multi-sorted universal algebra~\cite{meinke1993universal} upto the first
-homomorphism theorem. More precisely, we provide the following:
-\begin{itemize}
- \item Definition of a signature $\Sigma$. Definition of a type ($\Sigma$,E). $\Sigma$-algebras and
-($\Sigma$,$E$)-algebras
- \item A forgetful functor from ($\Sigma$,$E$)-algebras via $\Sigma$-algebras to Sets
- \item Products and subalgebras
- \item The term algebra for a type $\tau$. This is an initial object in the category of
-$\tau$-algebras.
-\end{itemize}
-%perhaps the soundness theorem for equational logic?
-
-Coq provides different sorts. To simplify the discussion, we assume that there are only two sorts:
-\lstinline|Prop| and \lstinline|Type|. It occasionally happens that we want to prove a similar theorem for both sorts.
-To avoid code duplication, Coq provides universe polymorphism for inductive types.
-Unfortunately, definitions and fixpoints are not (yet?) polymorphic. To avoid code duplication we
-change some Fixpoints into Inductives. 
-Concretely, the image $Im f:=\{b:B \mid \exists a.
-b=(f a)\}$ is informative (in \lstinline|Type|) since we need the witness in the proof of the first homomorphism theorem.
-A similar problem raised by Coquand and Spiwack~\cite{coquand-towards}.
 % theory/ua_subalgebraT.v
 % find . -iname "*.v" |xargs grep "niverse"
-
-In order to prove the first homomophism theorem, we define a congruence.
+We define congruence:
 \begin{lstlisting}
@@
 
-Alternatively, we could have stated the compatibility of the relation \lstinline|e| by stating that the
-relation as a set of pairs is a subalgebra in the product algebra. We prove that both approaches
-are equivalent. The former definition seems more natural in type theory. In set theory, a quotient
-is defined as a collection of equivalence classes. In type theory, a quotient is more naturally
+In set theory, a quotient is defined as a collection of equivalence classes. In type theory, a quotient is more naturally
 defined by changing the equality, while leaving the carrier unchanged.%
@@
 
-%varieties/closed_terms.v
-We define the forgetful functor from the category of $\tau$-algebras to the category of sets.
-We define the term algebra consisting of closed terms, i.e.\ terms with the empty type as set of
-variables. The term algebra is the initial $\tau$-algebra. Categorically: we have defined a
-left-adjoint to the forgetful functor.
-A formal proof connects the categorical and the algebraic definition.
-One may expect that this term algebra could be useful to
-\emph{define}, say, the natural numbers as the term algebra for the theory of semiring. However, it
-seems difficult to prove decidability of equality, which requires a normalization procedure.
+Alternatively, we could have stated the compatibility of the relation \lstinline|e| by stating that the
+relation as a set of pairs is a subalgebra in the product algebra. We have proved that both approaches
+are equivalent. As said, the present definition seems natural in type theory.
 
-The product of two $\tau$-algebras is their categorical product.
+Coq provides different sorts. To simplify the discussion, we assume that there are only two sorts:
+\lstinline|Prop|, for propositions, and \lstinline|Type|, for data types.
+A subalgebra is defined by a predicate on the algebra, that is a map \lstinline|A->Prop|.
+However, the image $Im f:=\{b:B \mid \exists a.
+b=(f a)\}$ is informative (in \lstinline|Type|) as it includes a witness for the existential.
+This witness is needed to map $b\in Im f$ to one of its pre-images in $A$.
+We thus need two flavors of subalgebras, one \lstinline|Prop|-valued an one \lstinline|Type|-valued; see also~\cite{coquand-towards}.
+Coq provides universe polymorphism for inductive types to avoid code duplication in such cases.
+Unfortunately, definitions and fixpoints are not (yet?) polymorphic.
+Hence we had to resort to changing some Fixpoints into Inductives.
 
-Given a sub\emph{set} of an algebra, we define the sub\emph{algebra} generated by it.
-An interesting fact to mention is the use of the following heterogeneous equality between
-elements of the algebra and of the subalgebra, i.e.\ terms of different types may be equal.
-\begin{lstlisting}
-  Fixpoint heq {o}: op_type carrier o -> op_type v o -> Prop :=
-    match o with
-    | constant x => fun a b => `a == b
-    | function x y => fun a b => forall u, heq (a u) (b (`u))
-    end.
-\end{lstlisting}
-
-We connect the abstract theory to the concrete theory. Concretely, let $\tau_m$ be the theory
-of monoids. Given a monoid, we can construct the corresponding $\tau_m$-algebra. 
-Conversely, given a $\tau_m$-algebra, we construct an instance of the \lstinline|Monoid| type class.
-%varieties/monoid.v
-\begin{lstlisting}
-Variable o: variety.Object theory.
-Global Instance: SemiGroupOp (o tt) := algebra_op theory mult.
-Global Instance: MonoidUnit (o tt) := algebra_op theory one.
-Global Instance from_object: Monoid (o tt).
-\end{lstlisting}
-
-In the last line, instance resolution `automatically' finds the operation and the unit specified in
-the lines before.
+Having discussed the first homomorphism theorem we now turn to category theory which helps to organize our development of universal algebra.
 
@@
 
-We have defined the category of setoids, the category of categories, the dual of a category,
-products, and hence co-products and the category of $\tau$-algebras, for a given type $\tau$.
-We prove that two definitions of adjunction are equivalent. The proofs are concise and follow the
-textbooks.
+We have defined the category of setoids, the dual of a category, products, and hence co-products. 
+We prove that two definitions of adjunction are equivalent. The proofs are concise and follow the textbooks.
 
-Universe polymorphism allows us to define the category of categories. However, we need to avoid
-making \lstinline|Relation| a \lstinline|Definition|, since \lstinline|Definitions| are not universe polymorphic.
+The usual size problems in the definition of the category of categories can be avoided by using universe polymorphism. 
+However, we need to avoid making \lstinline|Relation| a \lstinline|Definition|, since \lstinline|Definitions| are not (yet?) universe polymorphic.
 
-We provide the construction of the term-algebra for a type $\tau$ and show that it is
-initial in the category of $\tau$-algebras. We proceed to prove that this provides a left-adjoint
-to the forgetful functor. The resulting expression monad will be used in Section~\ref{quote} below.
+\subsection{Categorical universal algebra}
+Having discussed category theory, we now turn to its application to universal algebra.
+
+%varieties/closed_terms.v
+We define the forgetful functor $U$ from the category of τ-algebras to the category of sets and its left-adjoint $F$: the term algebra consisting of closed terms, i.e.\ terms with the empty type as set of variables.
+This term algebra is the initial τ-algebra.
+Every adjunction defines a monad by composition of the functors $U$ and $F$.
+We will use the resulting expression monad in Section~\ref{quote} below.
+
+One may hope that this term algebra could be useful to \emph{define} inductive data types, say, to define the natural numbers 
+as the term algebra for the theory of semirings. However, it seems difficult to prove decidability of equality, 
+as this requires a normalization procedure.
+
+%The product of two $\tau$-algebras is their categorical product.
+
+\begin{comment}
+Given a sub\emph{set} of an algebra, we define the sub\emph{algebra} generated by it.
+An interesting fact to mention is the use of the following heterogeneous equality between
+elements of the algebra and of the subalgebra, i.e.\ terms of different types may be equal.
+\begin{lstlisting}
+  Fixpoint heq {o}: op_type carrier o -> op_type v o -> Prop :=
+    match o with
+    | constant x => fun a b => `a == b
+    | function x y => fun a b => forall u, heq (a u) (b (`u))
+    end.
+\end{lstlisting}
+\end{comment}
+
+We connect the abstract theory to the concrete theory. Concretely, let $\tau_m$ be the theory
+of monoids. Given a (concrete) monoid, we can construct the corresponding $\tau_m$-algebra. 
+Conversely, given a $\tau_m$-algebra, we construct an instance of the \lstinline|Monoid| type class.
+%varieties/monoid.v
+\begin{lstlisting}
+Variable o: variety.Object theory.
+Global Instance: SemiGroupOp (o tt) := algebra_op theory mult.
+Global Instance: MonoidUnit (o tt) := algebra_op theory one.
+Global Instance from_object: Monoid (o tt).
+\end{lstlisting}
+
+In the last line, instance resolution `automatically' finds the operation and the unit specified in
+the lines before.
+
+This interplay between concrete algebraic structure and their expressions on the one hand, and models of equational theories on the other is occasionally a source of tension (in that translation in either direction is not yet fully automatic).
+However, it opens the door to the possibility of fully internalized implementations of generic tactics for algebraic manipulation, no longer requiring plugins.
+We come back to this when we describe automatic quotation of concrete expressions into universal algebra expressions; see Section~\ref{quote}.
 
@@
 Unification hints~\cite{Hints} allow one to semi-automatically construct a quote
-function.\footnote{Gonthier provides similar functionality using canonical structures.} However,
-this feature is absent from Coq. Fortunately, type classes provide similar functionality as
-we will now show.
+function.\footnote{Gonthier provides similar functionality using canonical structures.} This feature is absent from Coq. Fortunately, type classes provide similar functionality as we will now show.
 
-We define a term language for monoids:
+We define a term language for monoids
 \begin{lstlisting}
@@
 
-This defines a monad on the category of \lstinline|Type|s with extensional functions.\marginpar{Proof in classquote}
 The expression type is parametrized over the set of variable indices. Hence, we diverge from~\cite{Hints}, 
-which uses \lstinline|nat| for variables thereby introducing bounds problems and dummy variables.
+which uses \lstinline|nat| for variables thereby introducing bound problems and dummy variables.
 
-\noindent An expression is only meaningful in the context of a variable
+ An expression is only meaningful in the context of a variable
 assignment:\footnote{Categorically: \lstinline|nat| is
@@
 
-%
-Monads are trees with grafting~\cite{MonadsGrafting}: the tree monad
-$TX:=1+T^2X$, may be seen as the prototypical monad. Indeed, the \lstinline|Expr| monad, i.e.\ the free
-\lstinline|Expr|-algebra monad, behaves much like the tree monad. We now provide a direct definition of the
-corresponding bind operation, the arrows in the Kleisli category.
-
+%Monads are trees with grafting~\cite{MonadsGrafting}: the tree monad
+%$TX:=1+T^2X$, may be seen as the prototypical monad. 
+The \lstinline|Expr| monad, i.e.\ the free \lstinline|Expr|-algebra monad\footnote{on the category of \lstinline|Type|s with extensional functions.%\marginpar{Proof in classquote}
+}, provides a bind operation which describes the arrows in the Kleisli category. Concretely,
 \begin{lstlisting}
@@
 
-One useful operation we need before we get to Quote relates to variables and expression
- evaluation. As its name suggests, \lstinline|map_var| maps an expression's variable indices.
+%One useful operation we need before we get to Quote relates to variables and expression evaluation. 
+ As its name suggests, \lstinline|map_var| maps an expression's variable indices.
 This reindexing function is the the \lstinline|map| of the \lstinline|Expr|-monad.
@@
   Global Program Instance quote_zero V (v: Vars V):
-    Quote v 0 novars :=
-  { quote := Zero }.
+    Quote v 0 novars := { quote := Zero }.
 \end{lstlisting}
@@
  The third line contains some \lstinline|map_var|'s because the heap shuffling must be
-reflected in the variable indices, but apart from that it's just constructing a \lstinline|Mult| term with
+reflected in the variable indices, but apart from that it is just constructing a \lstinline|Mult| term with
 quoted subterms.
@@
 
-  Now follows the instance where we recognize values that are already in the heap. This
-   is expressed by the Lookup requirement, which will only be fulfilled if the Lookup instances
-   defined above can find the value in the heap. The novars in the \lstinline|Quote v x novars| result
+The instance where we recognize values that are already in the heap is expressed by the Lookup requirement, which will only be fulfilled if the Lookup instances defined above can find the value in the heap.
+The novars in the \lstinline|Quote v x novars| result
    reflects that this quotation does not generate new variables.
@@
 \noindent Finally, the instance for new variables. We give this a lower priority so that it is only
-used if Lookup fails. The \lstinline|8| in the previous example is a random number less than \lstinline|9| below. A similar method is used in Coq's notation mechanism.
+used if Lookup fails. The \lstinline|8| in the previous example is a random number less than \lstinline|9| below. 
+%A similar method is used in Coq's notation mechanism.
 % We recommend a change in
@@
 
-\noindent As announced this provides a light-weight quoting
-function\footnote{\lstinline|eval_quote'| is a
-modification of \lstinline|eval_quote| which works in the empty context.}:
+We have defined a light-weight quoting function. The following code turns the goal into \lstinline|(eval variable_pack quote)|.
+Here \lstinline|eval_quote'| is a modification of \lstinline|eval_quote| which works in the empty context:
 \begin{lstlisting}
@@
 
-This turns the goal into \lstinline|(eval variable_pack quote)|.
-
-It would be interesting to use this technique to implement the \lstinline|congruence| tactic, which implements
-the congruence closure algorithm~\cite{corbineau2007deciding}. One would naturally obtain a tactic
-which moreover works on setoid equalities.
+We have used this technique to implement a tactic to normalize monoid expressions.
+%It would be interesting to use this technique to implement the \lstinline|congruence| tactic, which implements
+%the congruence closure algorithm~\cite{corbineau2007deciding}. One would naturally obtain a tactic
+%which moreover works on setoid equalities.
 
@@
 structures to derive the setoid equality on the natural
-numbers in the following example \lstinline|(S x)==(S y)|. In this case type classes provide superior proof
+numbers in the following example \lstinline|(S x)==(S y)|. In this case type classes provide similar proof
 terms. Canonical structures give
@@
 lists, lists, inhabited finite sets and finite sets to the carrier of a semigroup, monoid,
-commutative semigroup, commutative monoid. Hence we want to reuse the libraries for lists etc.
+commutative semigroup, commutative monoid. Hence we want to reuse the libraries for lists etc.\ as much as possible.
 
-We use type classes to deduce the relevant monoid operation and write:
+Again, we can use type classes, instead of canonical structures, to deduce the relevant monoid operation:
 \begin{lstlisting}
@@
 
-Pollack~\cite{pollack2000dependently} introduces record types with telescopes. 
-Given a telescope\[
+Our approach may be intuitively explained by `telescopic' subset types. We recall the use of telescopes in record types~\cite{pollack2000dependently}. Given a telescope\[
 T=[x_1:A_1][x_2:A_2(x_1)]\cdots[x_k:A_k(x_1,\ldots,x_k)].
@@
 subset $\{ A \mid  B\}$. We moreover conjecture that the terms in \lstinline|B| only depend on the
-variables in \lstinline|A|, but not in \lstinline|B| itself. However, we do not need this.\\
+variables in \lstinline|A|, but not in \lstinline|B| itself. However, we do not need this.
 
-We observed that we were naturally lead to the following methodology: Let $\phi$ be a statement in type
+We are naturally lead to the following methodology: Let $\phi$ be a statement in type
 theory, i.e.\ an alternation of $\Pi$s and $\Sigma$s.
-By the axiom of choice, we first Skolemize to $\Sigma h.
-\Pi a.P$ and then transfer the variable $h$ to the parameters.
-For example, consider surjectivity.
-This is skolemized to a surjection with a right inverse, i.e.\ a split epi.
-We then move the right inverse
-to the parameters.
+By the type theoretic axiom of choice, we first Skolemize to $\Sigma \overrightarrow{h}.
+\Pi \overrightarrow{a}.
+P$ and then transfer the variable $\overrightarrow{h}$ to the parameters.
+Concretely, the statement of surjectivity is skolemized to a surjection with an explicit right inverse, i.e.\ a split epi.
+We move the right inverse to the parameters.
+As a side effect, the distinction between classical and constructive mathematics virtually disappears for such explicit statements.
 
-It is interesting to note that classical logic is conservative for purely universal statements.
-
-The addition of implicit Sigma types to Coq would change many things. Perhaps it would even allow us
+The addition of implicit Σ-types to Coq would change many things. Perhaps it would even allow us
 to pack the proofs together with the operations, but it is too early to tell.
 
-\subsection{Propositions as [ ]-Types}
+\subsection{Propositions as Types}
 As originally conceived~\cite{ITT,CMCP}, Propositions are Types and we can extract information from
-them. In practice, it is convenient to assume that all proofs of a proposition are equal: the
-principle of proof irrelevance. These two views clearly contradict each other. In Coq, a middle way
-was chosen, we are forbidden to extract information from proofs, but proofs are not identified.
-This allows us to extract the program we want, but we want more:
-We would like to use Coq as a dependently typed programming language. To do so,
-one would like to add the principle of proof irrelevance to Coq. A promising proposal is the
-Calculus of implicit constructions with implicit
-$\Sigma$-types~\cite{miquel2001implicit,barras2008implicit,Bernardo}. Such a change would fit
-naturally with the Hoare-style separation between programs and proofs which is used by the \lstinline|Program| machinery.
+them.
+Often we do \emph{not} want functions to depend on proofs.
+In Coq, this is the paradigmatic way of working.
+We are forbidden to extract information from proofs of \lstinline|Prop|s.
+Especially with the aid of the Program machinery, this allows us to flexibly work with simply typed program and their Hoare style correctness proofs.
+However, we would also like to use Coq as a dependently typed programming language.
+This requires a hybrid approach, were we put informative propositions in \lstinline|Type| and non-informative ones in \lstinline|Prop|.
+Making Harrop formulas proof irrelevant seems to be a good first approximation~\cite{lcf:spi:03}.
+An orthogonal and further reaching proposal is to replace the Coq type theory with the Calculus of implicit constructions with implicit
+Σ-types~\cite{miquel2001implicit,barras2008implicit,Bernardo}.
 
-One may be tempted to propose Propositions as [ ]-types~\cite{awodey2004propositions}. Here the
-construction $[P]:=\{\top\mid p:P\}$ squashes all the information about how we proved $P$. In Coq,
-the [ ]-types are isomorphic to the type of sort \lstinline|Prop|. However,
-there are places where we needed proof \emph{relevance}; e.g.\ in the proof of the first isomorphism
-theorem above.In practice, manual dead-code analysis seems useful. Making Harrop formulas proof
-irrelevant seems to be a good first approximation~\cite{lcf:spi:03}. However, a further refinement
-of Propositions as types seems necessary.
+%One may be tempted to propose Propositions as [ ]-types~\cite{awodey2004propositions}. Here the
+%construction $[P]:=\{\top\mid p:P\}$ squashes all the information about how we proved $P$. In Coq,
+%the [ ]-types are isomorphic to the type of sort \lstinline|Prop|. However,
+%there are places where we needed proof \emph{relevance}; e.g.\ in the proof of the first isomorphism
+%theorem above.In practice, manual dead-code analysis seems useful. Making Harrop formulas proof
+%irrelevant seems to be a good first approximation~\cite{lcf:spi:03}. However, a further refinement
+%of Propositions as types seems necessary.
 
@@
 the efficiency penalty of a long chain of coercion projections. Packed classes~\cite{Packed} provide
-a solution to these problems. We provide an alternative solution. We have already
-discussed multiple inheritance in Section~\ref{classes}. Moreover, by exposing the
-carrier we avoid the chains of projections.
+a solution to these problems. We have provided an alternative solution based on type classes.
 
 We conclude that unification hints, canonical structures and type classes all appear to have
-similar expressive powers. Canonical structures, being tied to the unifier, are more robust, but
+similar expressive power. Canonical structures, being tied to the unifier, are more robust, but
 seem to require more ingenuity. Type classes are easier to use. Unification hints, may be a
@@
 sharpen our understanding of the relation with Canonical Structures and with Unification Hints.
-%Jeremy, Thierry, Claudio, Georges?, Wouter
+%Jeremy, Thierry
 \bibliographystyle{plain}

### 2010-05-28 defManSSR.tex
Use Unicode in listings.

@@
 Inductive, CoInductive, Record, Structure, Canonical, Coercion,
+Context, Class, Global, Instance, Program, Infix,
 Theorem, Lemma, Corollary, Proposition, Fact, Remark, Example,
@@
 % Enables ASCII chars 128 to 255
-extendedchars=true,
+extendedchars=false,
 

### 2010-05-28 jar.tex
Use Unicode in listings.

@@
 \documentclass[a4paper,10pt,runningheads]{llncs}
-\usepackage{listings}
-% The package listingsutf8 supports utf8 for \lstinputlisting
-% However, we do not want to put all our code snippets in separate files.
 \usepackage{url}
@@
 \usepackage{amsmath}
+\usepackage{upgreek}
 \usepackage[mathletters]{ucs}
@@
 \usepackage{color}
+
 \definecolor{dkblue}{rgb}{0,0.1,0.5}
@@
 \definecolor{dkviolet}{rgb}{0.6,0,0.8}
+
+% listings:
+
+\usepackage{listings}
+  % The package listingsutf8 supports utf8 for \lstinputlisting
+  % However, we do not want to put all our code snippets in separate files.
+
 \def\lstlanguagefiles{defManSSR.tex}
 \lstset{language=SSR}
+\lstset{literate=
+  % symbols that actually occur as unicode in our source:
+  {λ}{{$\uplambda\ $}}1
+  {∃}{{$\exists$}}1
+  {→}{{$\to\ $}}1
+  {≠}{{$\ne\ $}}1
+  {¬}{{$\neg\ $}}1
+  {⟶}{{$\longrightarrow\ $}}1
+  {∧}{{$\land$}}1
+  {∀}{{$\forall$}}1
+  {Π}{{$\Uppi\ $}}1
+  {η}{{$\upeta$}}1
+  {⊓}{{$\sqcap$}}1
+  {∘}{{$\circ\ $}}1
+  {◎}{{$\odot\ $}}1
+  % things we can't make pretty in the actual source, but can make pretty here!:
+  {=>}{{$\,\Rightarrow\ $}}1
+  {==>}{{$\Rightarrow\ $}}1
+  {<-}{{$\leftarrow\ $}}1
+}
+
 \begin{document}
@@
 
-How should we define reflexive, transitive, and symmetric relations? In a first attempt, we
-might define:
-% \lstinputlisting[language=SSR,inputencoding=utf8/latin1]{ReflexiveRelation.v}
-
+How should we define reflexive, transitive, and symmetric relations? In a first attempt, we might define:
 \begin{lstlisting}
@@
     ; rr_rel: relation rr_carrier
-    ; rr_proof: $\Pi$ x, rr_rel x x }.
+    ; rr_proof: Π x, rr_rel x x }.
 \end{lstlisting}
@@
     { rr_rel: relation rr_carrier
-    ; rr_proof: $\Pi$ x, rr_rel x x }.
+    ; rr_proof: Π x, rr_rel x x }.
 \end{lstlisting}
@@
   Record ReflexiveRelation (carrier: Type) (rel: relation carrier): Prop :=
-    { rr_proof: $\Pi$ x, rr_rel x x }.
+    { rr_proof: Π x, rr_rel x x }.
 \end{lstlisting}
@@
   Definition ReflexiveRelation (carrier: Type) (rel: relation carrier): Prop :=
-    $\Pi$ x, rr_rel x x.
+    Π x, rr_rel x x.
 \end{lstlisting}
@@
     ; mult_comm: Commutative (mon_op mult_monoid)
-    ; mult_0_l: $\Pi$ x: A, H
+    ; mult_0_l: Π x: A, H
 \end{lstlisting}
@@
   ; semiring_distr:> Distribute mult plus
-  ; mult_0_l: forall x, 0 * x == 0 }.
+  ; mult_0_l: Π x, 0 * x = 0 }.
 \end{lstlisting}
@@
 Class Congruence: Prop :=
-    { congruence_proper:> $\Pi$ s, Proper (equiv ==> equiv ==> iff) (e s)
+    { congruence_proper:> Π s, Proper (equiv ==> equiv ==> iff) (e s)
     ; congruence_quotient:> @Algebra et v e _
@@
 \begin{lstlisting}
-Class Category O`{!Arrows O}`{$\Pi$ x y:O, Equiv(x $\to$ y)}
-                `{!CatId O}`{!CatComp O}:Prop :=
-  { arrow_equiv:> $\Pi$ x y, Setoid (x $\to$ y)
-  ; comp_proper:> $\Pi$ x y z,
-    Proper (equiv $\Rightarrow$ equiv $\Rightarrow$ equiv)%signature
-    (comp: (y $\to$ z) $\to$ (x$\to$ y) $\to$ x $\to$ z)
-  ; comp_assoc w x y z (a: w $\to$ x) (b: x $\to$ y) (c: y $\to$ z):
-      c $\circ$ (b $\circ$ a) = (c $\circ$ b) $\circ$ a
-  ; id_l `(a: x $\to$ y): cat_id $\circ$ a = a
-  ; id_r `(a: x $\to$ y): a $\circ$ cat_id = a }.
+Class Category O `{!Arrows O} `{Π x y: O, Equiv (x ⟶ y)}
+    `{!CatId O} `{!CatComp O}: Prop :=
+  { arrow_equiv:> Π x y, Setoid (x ⟶ y)
+  ; comp_proper:> Π x y z,
+    Proper (equiv ==> equiv ==> equiv) comp
+  ; comp_assoc w x y z (a: w ⟶ x) (b: x ⟶ y) (c: y ⟶ z):
+      c ◎ (b ◎ a) = (c ◎ b) ◎ a
+  ; id_l `(a: x ⟶ y): cat_id ◎ a = a
+  ; id_r `(a: x ⟶ y): a ◎ cat_id = a }.
 \end{lstlisting}
@@
    Context `{Category C} `{Category D}
-   `{!Functor (F: C $\to$ D) Fa} `{!Functor (G: C $\to$ D) Ga}.
-   Class NaturalTransformation ($\eta$: $\Pi$ c, F c $\to$ G c): Prop :=
-     natural: $\Pi$ (x y: C) (f: x $\to$ y), $\eta$ y $\circ$ fmap F f = fmap G f $\circ \eta$ x.
+   `{!Functor (F: C ⟶ D) Fa} `{!Functor (G: C ⟶ D) Ga}.
+   Class NaturalTransformation (η: Π c, F c ⟶ G c): Prop :=
+     natural: Π (x y: C) (f: x ⟶ y), η$$ y ∘ fmap F f = fmap G f ∘ η$$ x.
 \end{lstlisting}
+  % Todo: Having to add "$$" to get whitespace is awful. we can't add "\ " in η's literate-replacement though, because then we also get space before the colon in "η: t".
 
@@
 Definition Value := nat.
-Definition Vars V := V -> Value.
+Definition Vars V := V → Value.
 
@@
 \begin{lstlisting}
-Fixpoint bind {V W} (f: V -> Expr W) (e: Expr V): Expr W :=
+Fixpoint bind {V W} (f: V → Expr W) (e: Expr V): Expr W :=
   match e with
@@
 Definition novars: Vars False := False_rect _.
-Definition singlevar (x: Value): Vars unit := fun _ => x.
+Definition singlevar (x: Value): Vars unit := λ _ => x.
 Definition merge {A B} (a: Vars A) (b: Vars B): Vars (A+B) :=
-  fun i => if i then a j else b j.
+  λ i => if i then a j else b j.
 \end{lstlisting}
@@
 \begin{lstlisting}
-Definition map_var {V W: Type} (f: V -> W):
-    Expr V -> Expr W :=
+Definition map_var {V W: Type} (f: V → W):
+    Expr V → Expr W :=
   fix F (e: Expr V): Expr W :=
@@
 \begin{lstlisting}
-Goal forall x y (P: Value $\to$ Prop), P ((x * y) * (x * 0)).
+Goal Π x y (P: Value $\to$ Prop), P ((x * y) * (x * 0)).
   intros. rewrite <- (eval_quote' _).

### 2010-06-04 jar.tex
Snapshot of introduction/preliminaries/bundling sections rewrite (*heavy* WIP).

@@
   {◎}{{$\odot\ $}}1
+  { ≡ }{{$\equiv\ $}}1
   % things we can't make pretty in the actual source, but can make pretty here!:
@@
 
-For the algebraic hierarchy---a critical component in any library of formalized mathematics---these challenges include structure inference, handling of multiple inheritance, idiomatic use of notations, and convenient algebraic manipulation.
+For the algebraic hierarchy---a critical component in any library of formalized mathematics---these challenges include structure inference, handling of multiple inheritance, idiomatic use of notations, and convenient algebraic manipulation (e.g.
+rewriting).
 
@@
 
-\section{The Type-Classified Algebraic Hierarchy}\label{classes}
-Unlike Haskell's and Isabelle's second class type classes, Coq's type classes are first class~\cite{DBLP:conf/tphol/SozeauO08}: classes and their instances are realized as ordinary record types (``dictionaries'') and registered constants of these types.
+% ideally should be more or less readable by Haskell person. that would be neat. so that's what the preliminaries should prepare for.
 
-We represent each structure in the algebraic hierarchy as a type class.
-This immediately leads to the familiar question of which components of the structure should become parameters of the class, and which should become fields.
-By far the most important design choice in our development is the decision to turn all \emph{structural} components (i.e.
-carriers, relations, and operations) into parameters, keeping only \emph{properties} as fields.
-Type classes defined in this way are essentially predicates with automatically resolved proofs.
+\paragraph{Outline}
+In section \ref{preliminaries} we briefly describe the Coq system and some of its facilities that are of particular interest to us (most notably its implementation of type classes).
+Then, in section \ref{bundling}, we give a very concrete introduction to the issue of \emph{bundling}, as this is arguably the biggest design dimension when building interfaces for abstract structures.
+Having derived some principles concerning bundling, we then proceed to show (in section \ref{managing}) how type classes may be employed to adopt these principles without compromising ease of use.
+Next, in section \ref{hierarchy}, we introduce a number of additional conventions that let us complete a consistent idiom for interfaces to algebraic structures.
+With this in place, we continue with a discussion of
 
-Conventional wisdom warns that while this approach is theoretically very flexible, one risks extreme inconvenience both in having to declare and pass around all these structural components all the time, as well as in losing notations (because we no longer project named operations out of records).
-These legitimate concerns will be addressed in Sections~\ref{bundle}-\ref{classes}.
-Section~\ref{univ} abstracts to universal
-algebra, we develop the basic theory to the point of the first homomorphism theorem.
-Section~\ref{interfaces} provides interfaces to usual data structures by providing their universal
-properties. To allow symbolic manipulation of semantic objects one needs a quote function. Usually,
-this function is implemented in Coq's tactics language. A more convenient way is to use type class
-unification; see section~\ref{quote}. In section~\ref{canonical}, we end with a short comparison
-between type classes and canonical structures.
+to the
+
+% outline:
+% 
+% - preliminaries
+%     -> coq (couple of sentences about programming in a combined framework with logic)
+%     -> type classes
+% - on bundling
+%     -> introduce problem
+%     -> naive solution
+%     -> etc
+%     -> "we first begin by showing how unbundling is in some sense the "right" way to do things"
+%     in doing this, however, we will find that doing things this way is somewhat painful
+%     -> derived principles
+% - Managing unbundled structure with type classes
+%     -> painful if flexible, which is why not done before
+%     -> with type classes, managable.
+% 
+% - toward the algebraic hierarchy
+% 
+%   leaving things unbundled is one thing, but we need more for the alg. hierarchy. suppose we define a class for semigroups based on yada.
+%   -> operational type classes (parameterized over a single operation = constant), structural type classes (parameterized over all the components that make up the structure).
+% 
+% - advantages
+% 
+% - numeric classes
+% 
+% - cat.theory
+% 
+% - univ. alg.
+% 
+% - quoting
+
+%todo: have a word somewhere about "super classes"
 
@@
 \section{Preliminaries}
-\subsection{Coq type theory}
+\label{preliminaries}
+
+% ideally, this should make the paper readable for Haskell people
+
 The Coq proof assistant is based on the calculus of inductive
 constructions~\cite{CoquandHuet,CoquandPaulin}, a dependent type theory with (co)inductive types; see~\cite{Coq,BC04}.
+In true Curry-Howard fashion, it can be thought of either as an excessively pure (if somewhat pedantic) functional programming language with an extremely expressive type system, or as a language for mathematical statements and proofs.
+We highlight some aspect of Coq that are of particular relevance to our development.
 
-The Coq type theory lacks quotient
-types, as it would make type checking undecidable.
-Instead it supports setoids (aka Bishop sets), a pair of a type and an equivalence relation~\cite{Bishop67,Hofmann,Capretta}.
+\paragraph{Types and propositions}
+
+Propositions in Coq are types, which themselves have types called \emph{sorts}.
+Coq features a distinguished sort called \lstinline|Prop| that one may voluntarily choose to use as the sort of types representing propositions.
+The distinguishing feature of the \lstinline|Prop| sort is that Coq will not permit terms of non-\lstinline|Prop| types to depend on the values of inhabitants of \lstinline|Prop| types (i.e.
+proof terms).
+
+Adopting this regime of discrimination establishes a weak form of proof irrelevance, in that changing a proof can never affect the result of value computations.
+On a very practical level, this lets the extraction mechanism safely erase all \lstinline|Prop| components when extracting certified programs to OCaml or Haskell.
+
+A stronger notion of proof irrelevance in which any two inhabitants of a \lstinline|Prop| type are made \emph{convertible} is planned for a future Coq version.
+
+\paragraph{Equality, setoids, and rewriting}
+
+The ``native'' notion of equality in Coq is that of terms being convertible, which in this context constitutes Leibniz equality, naturally propositionalized by an inductive type family \lstinline|eq| with a single constructor \lstinline|eq_refl| of type \lstinline|Π (T: Type) (x: T), x ≡ x|, where ``\lstinline|a ≡ b|'' is notation for \lstinline|eq T a b|~\footnote{We diverge from Coq traditional and reserve a single equals sign for setoid equality.}.
+Importantly, since convertibility is a congruence, a proof of \lstinline|a ≡ b| lets us substitute \lstinline|b| for \lstinline|a| anywhere inside a term without further conditions.
+
+We mention this explicitly only because rewriting \emph{does} give rise to conditions when we depart from Leibniz equality and introduce equivalence relations representing abstractions over common representations of the same conceptual object (for instance when identifying formal fractions of integers representing the same ratio).
+Rewriting a subterm using such an equality is permitted only if the subterm is an argument of a function that has been proven to \emph{respect} the equality.
+Such a function is called \emph{proper}, and propriety must be proved for each function in whose arguments we wish to enable rewriting.
+
+Effectively keeping track of, resolving, and combining proofs of equivalence-ness and propriety when the user attempts to substitute a given (sub)term using a given equality, requires nontrivial infrastructure and support from the system.
+The Coq implementation of these mechanisms was largely rewritten in 200?
+by Matthieu Sozeau \ref{rewriting}.
+(todo: say that it's now much better than before)
+
+\paragraph{Type classes}
+
+Type classes have been a great success story in the Haskell functional programming language, as a means for organizing interfaces of abstract structures.
+The type class implementation that is now part of Coq provides a superset of their functionality, but implemented in a different way.
+
+In Haskell and Isabelle, type classes and their instances are second class, in that they are handled as specialized syntactic constructs whose semantics are given specifically by the type class apparatus.
+By contrast, the expressivity of dependent types and inductive families as supported in Coq, combined with the use of pre-existing technology in the system (namely proof search and implicit arguments) enable a \emph{first class} type class implementation~\cite{DBLP:conf/tphol/SozeauO08}: classes are ordinary record types (``dictionaries''), instances are ordinary constants of these record types (registered as ``hints'' with the proof search machinery), class constraints are (fairly) ordinary implicit parameters, and instance resolution is achieved by augmenting the unification algorithm to invoke ordinary proof search for implicit arguments of class type.
+
+Thus, type classes in Coq are realized by relatively minor syntactic aides that bring together existing facilities of the theory and the system into a coherent idiom, rather than by introduction of a new category of qualitatively different definitions with their own dedicated semantics.
+
+% Often, we will want to use a \emph{different} equivalence relation when representing algebraic structures, for instance when representing integers by the hypothetical subtraction of a pair of natural numbers.
+A type and an equivalence relation over that type together form a \emph{setoid}.
+% 
+% An extremely important
+% 
+% 
+% when doing algebra, however, one wants to use custom equivalence relations, for instance when
+% 
+% , for the purposes of doing algebra.
+% 
+% where term reduction is
+% 
+% the native
+% 
+% Leibniz equality 
+% 
+% (which in the Coq system amounts to
+% 
+% the ``native'' equality is term equality modulo reduction.
+% 
+% which amounts to 
+% 
+% normal forms
+% 
+% A trivial example being the method of representing integers by hypothetical subtractions of one natural from another.
+% 
+% 
+% The Coq type theory lacks quotient
+% types, as it would make type checking undecidable. % todo: reference
+% Instead it supports setoids (aka Bishop sets), a pair of a type and an equivalence relation~\cite{Bishop67,Hofmann,Capretta}.
 Recently, improved support for setoids has been added to the system~\cite{Setoid-rewrite}.
 Functions respecting these equivalence relations are called \lstinline|Proper|.
-As we will see in Section~\ref{univ} below this functionality can also be used to work with
-quotient algebras.
+% As we will see in Section~\ref{univ} below this functionality can also be used to work with
+% quotient algebras.
 
-Coq has two kinds of sorts \lstinline|Prop| for propositions and \lstinline|Type| for data types. Coq
-is scheduled to support proof irrelevance, every \lstinline|p,q:P:Prop| are
-convertible, in the future; see section~\ref{PI}.
+% \section{The Type-Classified Algebraic Hierarchy}\label{classes}
+% We represent each structure in the algebraic hierarchy as a type class.
+This immediately leads to the familiar question of which components of the structure should become parameters of the class, and which should become fields.
+By far the most important design choice in our development is the decision to turn all \emph{structural} components (i.e.
+carriers, relations, and operations) into parameters, keeping only \emph{properties} as fields.
+Type classes defined in this way are essentially predicates with automatically resolved proofs.
+% 
+% Conventional wisdom warns that while this approach is theoretically very flexible, one risks extreme inconvenience both in having to declare and pass around all these structural components all the time, as well as in losing notations (because we no longer project named operations out of records).
+These legitimate concerns will be addressed in Sections~\ref{bundle}-\ref{classes}.
+Section~\ref{univ} abstracts to universal
+% algebra, we develop the basic theory to the point of the first homomorphism theorem.
+% Section~\ref{interfaces} provides interfaces to usual data structures by providing their universal
+% properties. To allow symbolic manipulation of semantic objects one needs a quote function. Usually,
+% this function is implemented in Coq's tactics language. A more convenient way is to use type class
+% unification; see section~\ref{quote}. In section~\ref{canonical}, we end with a short comparison
+% between type classes and canonical structures.
 
@@
 
-\section{To bundle or not to bundle? (Or: What to bundle?)}\label{bundle}
-We provide a rationale for our approach. We begin gently.
+\section{Bundling is bad}\label{bundling}
 
-How should we define reflexive, transitive, and symmetric relations? In a first attempt, we might define:
+Algebraic structures are expressed in terms of a number of carriers, a number of operations and relations on these carriers, and a number of laws that the operations and relations satisfy.
+In a system like Coq, we have different options when it comes to representing the grouping of these components.
+On one end of the spectrum, we can simply define the (conjunction of) laws as a $n$-ary predicate over $n$ components, basically forgoing grouping altogether.
+For instance, for the mundane example of a reflexive relation, we could use:
+
+\begin{lstlisting}
+Definition reflexive {A} (R: relation A): Prop := Π a, R a a.
+\end{lstlisting}
+(The curly brackets used for \lstinline|A| mark it as an implicit argument.)
+
+Higher structures, too, can all be expressed as predicates (expressing laws) over a number of carriers, relations, and operations.
+While optimally flexible in theory (todo: ref), in practice \emph{naive} adoption of this approach in Coq (i.e.
+without using type classes) leads to substantial inconveniences in actual use: when \emph{stating} theorems about abstract instances of such structures, one must enumerate all components along with the structure (predicate) of interest.
+And when \emph{applying} such theorems, one must either enumerate any non-inferrable components, or resort to awkward existentials.
+Importantly, this also hinders proof search for proofs of the structure predicates, making any nontrivial use of theorems a laborious experience.
+Finally, the lack of \emph{canonical names} for particular components of abstract structures makes it impossible to associate convenient notations with them.
+
+% hm, i used to have paragraphs here about how it would be unclear how this approach could support structure inference and structure inheritance, but those are actually pretty straightforward using ordinary proof search.
+
+% hm, say something about opacity
+
+In the absence of type classes, these are all very real problems, and for this reason the two biggest formalizations of abstract algebraic structures in Coq today (namely CoRN\ref{corn} and Ssreflect\ref{ssreflect}), both use \emph{bundled} representations: records with one or more of the components as fields.
+For reflexive relations, the following is a fully bundled representation---the other end of the spectrum:
 \begin{lstlisting}
@@
 \end{lstlisting}
+Superficially, this instantly solves the problems described above: reflexive relations can now be declared and passed as self-contained packages, and the \lstinline|rr_rel| projection now constitutes a ``canonical name'' for relations that are known to be reflexive, and we could bind a notation to it\footnote{While there is no conventional notation for reflexive relations, the situation is the same in the context of, say, a semiring, where we would bind ``+'' and ``*'' notations to the record field projections for the addition and multiplication operations, respectively.}.
 
-and similar for transitive and symmetric relations.
+% While not observable in this example, when combined with other specialized facilities in Coq (namely coercions and/or canonical structures), this bundled representation also lets one address the problems of structure inference inheritance to some extent.
 
-With this approach, we run into problems when defining \lstinline|EquivalenceRelation| in terms of
-\lstinline|ReflexiveRelation| and its two siblings, because when posit three objects of type
-\lstinline|ReflexiveRelation|, \lstinline|TransitiveRelation|, and \lstinline|SymmetricRelation|, respectively, we have to
-add equalities expressing that they talk about the same carrier and relation. These equalities
-cause difficulties, because the three carriers and relations won't be convertible.
-Technically, this leads to the research in manifest fields, i.e.\ the \emph{with} construction, to
-which we return in Section~\ref{manifest}.
+Unfortunately, despite its virtues, the bundled representation introduces serious problems of its own, the most immediate and prominent one being a lack of support for \emph{sharing} components between structures.
 
-As a first step we would define \lstinline|ReflexiveRelation| as:
-\begin{lstlisting}
-  Record ReflexiveRelation (carrier: Type): Type :=
-    { rr_rel: relation rr_carrier
-    ; rr_proof: Π x, rr_rel x x }.
-\end{lstlisting}
+In our example, lack of sharing support rears its head as soon as we try to define \lstinline|EquivalenceRelation| in terms of \lstinline|ReflexiveRelation| and its hypothetical siblings bundling symmetric and transitive relations.
+There, we would need some way to make sure that when we bundle the three constituents of types \lstinline|ReflexiveRelation|, \lstinline|SymmetricRelation|, and \lstinline|TransitiveRelation|, respectively together in a record, they all refer to the same carrier and relation.
+Positing additional equalities between the three bundled carriers and relations is neither easily accomplished (because one would need to work with heterogenous equality) nor would it permit natural use of the resulting structure (because one would constantly have to rewrite things back and forth).
 
-This would solve the problem for the carrier, but \lstinline|EquivalenceRelation| would still need equalities for the relation:
-\begin{lstlisting}
-  Record EquivalenceRelation: Type :=
-    { er_carrier: Type
-    ; er_refl: ReflexiveRelation er_carrier
-    ; er_sym: SymmetricRelation er_carrier
-    ; er_trans: TransitiveRelation er_trans
-    ; same_rel: rr_rel er_refl = sr_rel er_sym
-    ; same_rel': rr_rel er_refl = tr_rel er_trans }.
-\end{lstlisting}
+Manifest records \ref{manifestrecords} have been proposed and implemented in the Matita system to address exactly this problem.
+We hope to convince the reader that type system extensions of this nature, designed to mitigate particular symptoms of the bundled approach, are less elegant than a solution that avoids the problem altogether by employing type classes instead of bundled records.
 
-A natural solution is to pull the relation out of \lstinline|ReflexiveRelation| (and its siblings) as well:
-\begin{lstlisting}
-  Record ReflexiveRelation (carrier: Type) (rel: relation carrier): Prop :=
-    { rr_proof: Π x, rr_rel x x }.
-\end{lstlisting}
-
-or equivalently:
-\begin{lstlisting}
-  Definition ReflexiveRelation (carrier: Type) (rel: relation carrier): Prop :=
-    Π x, rr_rel x x.
-\end{lstlisting}
-
-This is how reflexivity is defined in the standard library, and this is entirely sensible. The
-change from \lstinline|Type| to \lstinline|Prop| reflects the fact that \lstinline|ReflexiveRelation| is now strictly a property,
-proofs of which can be nicely kept opaque;%
-\footnote{Coq distinguishes between transparent and opaque definitions. The former can be unfolded,
-whereas this is impossible for the latter. Since we never need to distinguish between two proofs of
-a propositions, propositions can be opaque.}
-we will return to this issue in Section~\ref{PI}.
-We can now define \lstinline|EquivalenceRelation| without awkward equalities:
+If we revert back to the predicate formulation of reflexive (et.al) relations, we \emph{could} define \lstinline|EquivalenceRelation| in a bundled fashion without the need for equalities:
 \begin{lstlisting}
@@
 \end{lstlisting}
+However, the exact same argument inevitably leads to the conclusion that \lstinline|EquivalenceRelation|, too, should be a predicate.
+Indeed, it would be rather strange for the interface of equivalence relations to differ qualitatively from the interface of reflexive relations.
 
-But looking at \lstinline|EquivalenceRelation|, we see that it now essentially has the same form as the
-original naive \lstinline|ReflexiveRelation| record we started with. Clearly, equivalence relations
-and reflexive relations are analogous concepts, and by the same argument we
-conclude that \lstinline|EquivalenceRelation| should really be defined as:
+Another attempt to regain some grouping might be to bundle the carrier with the relation into a record.
+Unfortunately, this leads to the same problem of sharing, in different clothes: as soon as we try to define an algebraic structure with two reflexive relations on the same carrier, we need awkward hacks to establish equality between the carrier projections of two different (carrier, relation) bundles.
+To give a slightly less silly example of this phenomenon: bundling a monoid's constant and binary operation into a record means that when defining a hypothetical algebraic structure with two binary relations and a constant such that both binary relations form a monoid with the constant, one needs awkard hacks to establish equality of the two operation bundles' constants.
 
-\begin{lstlisting}
-Record EquivalenceRelation (er_carrier: Type)
-                             (er_rel: relation er_carrier): Prop :=
-  { er_refl: ReflexiveRelation er_carrier er_rel
-  ; er_sym: SymmetricRelation er_carrier er_rel
-  ; er_trans: TransitiveRelation er_trans er_rel }.
-\end{lstlisting}
+A second problem with bundling components is that as the bundled records are stacked to represent higher and higher structures, the projection paths for their components grow longer and longer, resulting in ever more unwieldy terms (although admittedly coercions and notations can make this less painful).
 
-After all, when we define a type with
-two different equivalence relations on it,
-such as the quotient of a setoid,
-we have the same considerations as for \lstinline|ReflexiveRelation| above.
+% hm, what to say about terms referring on proofs.
 
-At the risk of overstressing the point, let us look at how the same problem and solution keep
-appearing in composite structures. Suppose we use a definition of \lstinline|Monoid| that bundles the
-relation. Then if we proceed to define \lstinline|SemiRing| as a double \lstinline|Monoid|, we face a dilemma:
-\begin{lstlisting}
-  Record SemiRing A: Type :=
-    { mult_monoid: Monoid A 
-    ; plus_monoid: Monoid A 
-    ; plus_comm: Commutative (mon_op plus_monoid)
-    ; mult_comm: Commutative (mon_op mult_monoid)
-    ; mult_0_l: Π x: A, H
-\end{lstlisting}
+Thus, bundled representations come at a substantial cost in flexibility.
+Historically, this cost has been acceptable because (1) the unbundled alternative was such a pain, and (2) the standard algebraic hierarchy (up to, say, fields and modules) is not all that wild.
 
-Do we use \lstinline|mult_monoid|'s equality or \lstinline|plus_monoid|'s equality in \lstinline|H|?
-Again, for this definition to be usable at all, we would need to add an equality expressing that the
-two monoids contain the same equivalence relation. And again, the proper solution is to change
-Monoid and unbundle the relation. But here, we have a choice as to how much we unbundle exactly.
-(For the moment we ignore the question of whether the unit and binary operator should be unbundled
-as well, focusing only on the equivalence relation and its equivalence proof.)
+In the next section, we show that type-classification of structure predicates and their component parameters have the potential to remedy the problems traditionally associated with their use, giving us the best of both worlds.
+
+For completeness, we mention an alternative record-based approach known as ``packed classes'', as used in the ssreflect library.
+In the packed classes idiom, each algebraic structure is represented by not one but a trinity of interdependent records along with packing/unpacking functions.
+These are then stacked in an intricate manner designed to reach a particular hybrid packing balance that, combined with the use of canonical structures and coercions, accomplishes some of the goals above.
+
+We have found it hard to assess the general merits of this approach, partly because of its complexity, and partly because the only actual usage example (the hierarchy in the ssreflect library) is not only wedded to a very specialized view on equality and decidability that does not support setoid relations from the outset, but also diverges significantly from standard mathematical dichotomy, making it hard to recognize familiar structures and presentation.
+The reasons for this (as we understand them) are pragmatic and twofold.
+
+First, the ssreflect library was originally developed strictly to support very practical work in a specialized domain of mathematics involving decidable and mostly finite structures.
+Second, at the time of ssreflect's initial development, support for setoid rewriting in Coq was not nearly as stable and refined as it is today, which made ``Leibniz equality at all costs'' a more reasonable strategy at the time than it might be considered today.
+
+Work to extend the ssreflect framework with techniques to allow representation of a less restricted class of mathematical structures has only recently begun and is still ongoing, though there are currently no plans to support undecidable structures.
+
+Because in our development we will be working with computable reals, equality on which can never be either Leibniz or decidable, we conclude that the ssreflect library is not usable for us.
+
+% todo: mention proof irrelevance here somewhere, and terms that depend on proofs.
 
@@
 
-Another benefit of our approach is that it works very smoothly when these records are in fact type
-classes. When relations and operations are bundled and do \emph{not} appear as parameters of the
-structure record/class, to express that some collection of relations and
-operations possess the structure as a type class constraint seems to require manifest fields; see
-Section~\ref{manifest}. With
-our unbundled approach, on the other hand, such type class constraints are easily expressed without
-any extensions as type inhabitation queries
-% (e.g. ``hey Coq, find me a proof of [Monoid some_rel some_op some_unit]'')
-of types living in Prop, so that we don't even care what instances are found (thus embracing proof
-irrelevance).
+% Another benefit of our approach is that it works very smoothly when these records are in fact type
+% classes. When relations and operations are bundled and do \emph{not} appear as parameters of the
+% structure record/class, to express that some collection of relations and
+% operations possess the structure as a type class constraint seems to require manifest fields; see
+% Section~\ref{manifest}. With
+% our unbundled approach, on the other hand, such type class constraints are easily expressed without
+% any extensions as type inhabitation queries
+% % (e.g. ``hey Coq, find me a proof of [Monoid some_rel some_op some_unit]'')
+% of types living in Prop, so that we don't even care what instances are found (thus embracing proof
+% irrelevance).
 
-Leaving the equivalence proofs bundled means that when one has a proof that a structure
-is a semiring, one has two proofs (via the two monoids) that the equivalence relation is an
-equivalence. Fortunately, this redundancy is entirely harmless since our terms
-never refer (directly or indirectly) to proofs, which is the case when the relations and operations
-they are composed from are left unbundled. Here, too, we enjoy the warm embrace of proof
-irrelevance.
-
-From the discussion so far, we derive some principles:
-\begin{enumerate}
- \item structural concepts (like Reflexive, Setoid, Monoid) should have their structural components
-(i.e. carriers, relations, and operations) as parameters, and nothing else;
- \item (as a consequence) these concepts should be defined as predicates (which includes records and
-classes living in Prop), and their proofs may be kept opaque;
- \item (as a second consequence) algebraic expressions need/should never refer (directly or
-indirectly) to proofs about the algebraic structure.\footnote{The reciprocal in a field
-famously does refer to a proof that the element is non-zero, but not to a proof that the operations
-indeed form a field.}
-\end{enumerate}
-% \setcounter{enumi}{3}
-
-The principles listed so far are not particularly controversial, in fact this
-approach ensures maximum flexibility~\cite{Hints}. However,
-there is a common perception that unbundling carriers, relations, and operations this way invariably
-leads either to unmanageably big algebraic expressions (involving endless projections and/or implicit
-arguments), or to unmanageably long and tedious enumeration of carriers/relations/operations, or to
-problems establishing canonical names for operations (since we no longer project them out of
-structure instances all the time).
-
-In the next sections, we show how systematic use of type classes (and their support infrastructure
-in Coq), combined with the use of a special form of type class we call an `operational type class'
-(as opposed to a `structural' type class we might use in place of the records shown thus far), lets
-us avoid these problems.
+% Leaving the equivalence proofs bundled means that when one has a proof that a structure
+% is a semiring, one has two proofs (via the two monoids) that the equivalence relation is an
+% equivalence. Fortunately, this redundancy is entirely harmless since our terms
+% never refer (directly or indirectly) to proofs, which is the case when the relations and operations
+% they are composed from are left unbundled. Here, too, we enjoy the warm embrace of proof
+% irrelevance.
+% 
+% From the discussion so far, we derive some principles:
+% \begin{enumerate}
+%  \item structural concepts (like Reflexive, Setoid, Monoid) should have their structural components
+% (i.e. carriers, relations, and operations) as parameters, and nothing else;
+%  \item (as a consequence) these concepts should be defined as predicates (which includes records and
+% classes living in Prop), and their proofs may be kept opaque;
+%  \item (as a second consequence) algebraic expressions need/should never refer (directly or
+% indirectly) to proofs about the algebraic structure.\footnote{The reciprocal in a field
+% famously does refer to a proof that the element is non-zero, but not to a proof that the operations
+% indeed form a field.}
+% \end{enumerate}
+% % \setcounter{enumi}{3}
+% 
+% The principles listed so far are not particularly controversial, in fact this
+% approach ensures maximum flexibility~\cite{Hints}. However,
+% there is a common perception that unbundling carriers, relations, and operations this way invariably
+% leads either to unmanageably big algebraic expressions (involving endless projections and/or implicit
+% arguments), or to unmanageably long and tedious enumeration of carriers/relations/operations, or to
+% problems establishing canonical names for operations (since we no longer project them out of
+% structure instances all the time).
+% 
+% In the next sections, we show how systematic use of type classes (and their support infrastructure
+% in Coq), combined with the use of a special form of type class we call an `operational type class'
+% (as opposed to a `structural' type class we might use in place of the records shown thus far), lets
+% us avoid these problems.
 

### 2010-06-14 jar.tex
Another snapshot. First 12 pages now more or less say what I want them to say.

@@
 
+\usepackage{graphicx}
+
 \definecolor{dkblue}{rgb}{0,0.1,0.5}
@@
 \definecolor{dkviolet}{rgb}{0.6,0,0.8}
+
+% preliminary choices:
+%   don't discuss difference between maximally-inserted and non-maximally-inserted implicit arguments.
+
+% todo:
+  % what to say about terms referring on proofs?
+  % cite that other type class based development in the context of the alternative bundling schemes.
+  % maybe have a chapter called "interfaces are primary, not implementations" (also involve polynomials)
+  % say this somewhere: \footnote{For an interesting (if fairly restrictive) alternative design in which unicity of term representations \emph{is} required for all objects, see the algebraic hierarchy in the ssreflect\cite{ssreflect} standard library.}
 
@@
 abilities of type class unification allow us to conveniently define a quote function thus facilitating the use of reflective techniques.
+  % todo: handle.
 \end{abstract}
@@
 telescopes), packed classes~\cite{Packed}, and occasionally modules.
-We present a new approach based entirely on Coq's new type class mechanism, and show how its features together with a key design pattern let us effectively address the challenges mentioned above.
+We present a new approach based entirely on Coq's new type class mechanism, and show how it can be used to make fully ``unbundled'' predicate-representations of algebraic structures practical to work with.
 
@@
 In section \ref{preliminaries} we briefly describe the Coq system and some of its facilities that are of particular interest to us (most notably its implementation of type classes).
-Then, in section \ref{bundling}, we give a very concrete introduction to the issue of \emph{bundling}, as this is arguably the biggest design dimension when building interfaces for abstract structures.
-Having derived some principles concerning bundling, we then proceed to show (in section \ref{managing}) how type classes may be employed to adopt these principles without compromising ease of use.
+Then, in section \ref{bundling}, we give a very concrete introduction to the issue of \emph{bundling}, arguably the biggest design dimension when building
+interfaces for abstract structures.
+In section \ref{predicateclasses} we show how type classes can make the use of ``unbundled'' purely predicate based interfaces for abstract structures practical.
+Next, in section \ref{hierarchy}, we discuss our algebraic hierarchy implemented using such predicate classes.
+
+ %Having derived some principles concerning bundling, we then proceed to show (in section \ref{managing}) how type classes may be employed to adopt these principles without compromising ease of use.
 Next, in section \ref{hierarchy}, we introduce a number of additional conventions that let us complete a consistent idiom for interfaces to algebraic structures.
@@
 
-to the
+%to the
 
@@
 
-The ``native'' notion of equality in Coq is that of terms being convertible, which in this context constitutes Leibniz equality, naturally propositionalized by an inductive type family \lstinline|eq| with a single constructor \lstinline|eq_refl| of type \lstinline|Π (T: Type) (x: T), x ≡ x|, where ``\lstinline|a ≡ b|'' is notation for \lstinline|eq T a b|~\footnote{We diverge from Coq traditional and reserve a single equals sign for setoid equality.}.
+The ``native'' notion of equality in Coq is that of terms being convertible, which in this context constitutes Leibniz equality, naturally propositionalized by an inductive type family \lstinline|eq| with a single constructor \lstinline|eq_refl| of type \lstinline|Π (T: Type) (x: T), x ≡ x|, where ``\lstinline|a ≡ b|'' is notation for \lstinline|eq T a b|~\footnote{We diverge from Coq traditional and reserve ``\lstinline|=|'' for setoid equality, as this is the equality we will be working with most of the time.}.
 Importantly, since convertibility is a congruence, a proof of \lstinline|a ≡ b| lets us substitute \lstinline|b| for \lstinline|a| anywhere inside a term without further conditions.
@@
 In Haskell and Isabelle, type classes and their instances are second class, in that they are handled as specialized syntactic constructs whose semantics are given specifically by the type class apparatus.
-By contrast, the expressivity of dependent types and inductive families as supported in Coq, combined with the use of pre-existing technology in the system (namely proof search and implicit arguments) enable a \emph{first class} type class implementation~\cite{DBLP:conf/tphol/SozeauO08}: classes are ordinary record types (``dictionaries''), instances are ordinary constants of these record types (registered as ``hints'' with the proof search machinery), class constraints are (fairly) ordinary implicit parameters, and instance resolution is achieved by augmenting the unification algorithm to invoke ordinary proof search for implicit arguments of class type.
-
-Thus, type classes in Coq are realized by relatively minor syntactic aides that bring together existing facilities of the theory and the system into a coherent idiom, rather than by introduction of a new category of qualitatively different definitions with their own dedicated semantics.
-
-% Often, we will want to use a \emph{different} equivalence relation when representing algebraic structures, for instance when representing integers by the hypothetical subtraction of a pair of natural numbers.
-A type and an equivalence relation over that type together form a \emph{setoid}.
-% 
-% An extremely important
-% 
-% 
-% when doing algebra, however, one wants to use custom equivalence relations, for instance when
-% 
-% , for the purposes of doing algebra.
-% 
-% where term reduction is
-% 
-% the native
-% 
-% Leibniz equality 
-% 
-% (which in the Coq system amounts to
-% 
-% the ``native'' equality is term equality modulo reduction.
-% 
-% which amounts to 
-% 
-% normal forms
-% 
-% A trivial example being the method of representing integers by hypothetical subtractions of one natural from another.
-% 
-% 
+By contrast, the expressivity of dependent types and inductive families as supported in Coq, combined with the use of pre-existing technology in the system (namely proof search and implicit arguments) enable a \emph{first class} type class implementation~\cite{DBLP:conf/tphol/SozeauO08}: classes are ordinary record types (``dictionaries''), instances are ordinary constants of these record types (registered as ``hints'' with the proof search machinery), class constraints are ordinary implicit parameters, and instance resolution is achieved by augmenting the unification algorithm to invoke ordinary proof search for implicit arguments of class type.
+
+Thus, type classes in Coq are realized by relatively minor syntactic aids that bring together existing facilities of the theory and the system into a coherent idiom, rather than by introduction of a new category of qualitatively different definitions with their own dedicated semantics.
+
 % The Coq type theory lacks quotient
@@
 In a system like Coq, we have different options when it comes to representing the grouping of these components.
-On one end of the spectrum, we can simply define the (conjunction of) laws as a $n$-ary predicate over $n$ components, basically forgoing grouping altogether.
+On one end of the spectrum, we can simply define the (conjunction of) laws as an $n$-ary predicate over $n$ components, forgoing explicit grouping altogether.
 For instance, for the mundane example of a reflexive relation, we could use:
@@
 \begin{lstlisting}
-Definition reflexive {A} (R: relation A): Prop := Π a, R a a.
+  Definition reflexive {A: Type} (R: relation A): Prop := Π a, R a a.
 \end{lstlisting}
@@
 Higher structures, too, can all be expressed as predicates (expressing laws) over a number of carriers, relations, and operations.
-While optimally flexible in theory (todo: ref), in practice \emph{naive} adoption of this approach in Coq (i.e.
+While optimally flexible in principle, in practice \emph{naive} adoption of this approach (i.e.
 without using type classes) leads to substantial inconveniences in actual use: when \emph{stating} theorems about abstract instances of such structures, one must enumerate all components along with the structure (predicate) of interest.
-And when \emph{applying} such theorems, one must either enumerate any non-inferrable components, or resort to awkward existentials.
+And when \emph{applying} such theorems, one must either enumerate any non-inferrable components, or let the system spawn awkward metavariables to be resolved at a later time.
 Importantly, this also hinders proof search for proofs of the structure predicates, making any nontrivial use of theorems a laborious experience.
-Finally, the lack of \emph{canonical names} for particular components of abstract structures makes it impossible to associate convenient notations with them.
+Finally, the lack of \emph{canonical names} for particular components of abstract structures makes it impossible to associate idiomatic notations with them.
 
@@
 
-In the absence of type classes, these are all very real problems, and for this reason the two biggest formalizations of abstract algebraic structures in Coq today (namely CoRN\ref{corn} and Ssreflect\ref{ssreflect}), both use \emph{bundled} representations: records with one or more of the components as fields.
+In the absence of type classes, these are all very real problems, and for this reason the two biggest formalizations of abstract algebraic structures in Coq today (namely CoRN\ref{corn} and Ssreflect\ref{ssreflect}), both use \emph{bundled} representation schemes, using records with one or more of the components as fields instead of parameters.
 For reflexive relations, the following is a fully bundled representation---the other end of the spectrum:
@@
 \end{lstlisting}
-Superficially, this instantly solves the problems described above: reflexive relations can now be declared and passed as self-contained packages, and the \lstinline|rr_rel| projection now constitutes a ``canonical name'' for relations that are known to be reflexive, and we could bind a notation to it\footnote{While there is no conventional notation for reflexive relations, the situation is the same in the context of, say, a semiring, where we would bind ``+'' and ``*'' notations to the record field projections for the addition and multiplication operations, respectively.}.
+Superficially, this instantly solves the problems described above: reflexive relations can now be declared and passed as self-contained packages, and the \lstinline|rr_rel| projection now constitutes a ``canonical name'' for relations that are known to be reflexive, and we could bind a notation to it\footnote{While there is no conventional notation for reflexive relations, the situation is the same in the context of, say, a semiring, where we would bind ``$+$'' and ``$*$'' notations to the record field projections for the addition and multiplication operations, respectively.}.
 
@@
 
-Unfortunately, despite its virtues, the bundled representation introduces serious problems of its own, the most immediate and prominent one being a lack of support for \emph{sharing} components between structures.
+Unfortunately, despite its apparent virtues, the bundled representation introduces serious problems of its own, the most immediate and prominent one being a lack of support for \emph{sharing} components between structures, which is needed to cope with overlapping multiple inheritance.
 
 In our example, lack of sharing support rears its head as soon as we try to define \lstinline|EquivalenceRelation| in terms of \lstinline|ReflexiveRelation| and its hypothetical siblings bundling symmetric and transitive relations.
-There, we would need some way to make sure that when we bundle the three constituents of types \lstinline|ReflexiveRelation|, \lstinline|SymmetricRelation|, and \lstinline|TransitiveRelation|, respectively together in a record, they all refer to the same carrier and relation.
-Positing additional equalities between the three bundled carriers and relations is neither easily accomplished (because one would need to work with heterogenous equality) nor would it permit natural use of the resulting structure (because one would constantly have to rewrite things back and forth).
-
-Manifest records \ref{manifestrecords} have been proposed and implemented in the Matita system to address exactly this problem.
-We hope to convince the reader that type system extensions of this nature, designed to mitigate particular symptoms of the bundled approach, are less elegant than a solution that avoids the problem altogether by employing type classes instead of bundled records.
-
-If we revert back to the predicate formulation of reflexive (et.al) relations, we \emph{could} define \lstinline|EquivalenceRelation| in a bundled fashion without the need for equalities:
+There, we would need some way to make sure that when we ``inherit'' \lstinline|ReflexiveRelation|, \lstinline|SymmetricRelation|, and \lstinline|TransitiveRelation| by adding them as fields in our bundled record, they all refer to the same carrier and relation.
+Adding additional fields stating equalities between the three bundled carriers and relations is neither easily accomplished (because one would need to work with heterogenous equality) nor would it permit natural use of the resulting structure (because one would constantly have to rewrite things back and forth).
+
+Manifest fields \ref{manifestrecords} have been proposed and implemented in the Matita system to address exactly this problem.
+We hope to convince the reader that type system extensions of this nature, designed to mitigate particular symptoms of the bundled approach, are less elegant than a solution that avoids the problem altogether by using predicate-like type classes in place of bundled records.
+
+If we revert back to the predicate formulation of reflexive (et. al) relations, we \emph{could} still define \lstinline|EquivalenceRelation| in a bundled fashion without the need for equalities:
 \begin{lstlisting}
@@
 \end{lstlisting}
-However, the exact same argument inevitably leads to the conclusion that \lstinline|EquivalenceRelation|, too, should be a predicate.
+However, the same argument as before leads to the conclusion that \lstinline|EquivalenceRelation|, too, should be a predicate.
 Indeed, it would be rather strange for the interface of equivalence relations to differ qualitatively from the interface of reflexive relations.
 
-Another attempt to regain some grouping might be to bundle the carrier with the relation into a record.
-Unfortunately, this leads to the same problem of sharing, in different clothes: as soon as we try to define an algebraic structure with two reflexive relations on the same carrier, we need awkward hacks to establish equality between the carrier projections of two different (carrier, relation) bundles.
-To give a slightly less silly example of this phenomenon: bundling a monoid's constant and binary operation into a record means that when defining a hypothetical algebraic structure with two binary relations and a constant such that both binary relations form a monoid with the constant, one needs awkard hacks to establish equality of the two operation bundles' constants.
-
-A second problem with bundling components is that as the bundled records are stacked to represent higher and higher structures, the projection paths for their components grow longer and longer, resulting in ever more unwieldy terms (although admittedly coercions and notations can make this less painful).
-
-% hm, what to say about terms referring on proofs.
+Another attempt to recover some grouping might be to bundle the carrier with the relation into a (lawless) record, but this too hinders sharing: as soon as we try to define an algebraic structure with two reflexive relations on the same carrier, we need awkward hacks to establish equality between the carrier projections of two different (carrier, relation) bundles.
+
+Even bundling just the operations of an algebraic structure together in a record (with the carrier as a parameter) does not work, as it again leads to the same problem when, for example, one attempts to define a hypothetical algebraic structure with two binary relations and a constant such that both binary relations form a monoid with the constant.
+
+A second problem with bundling is that as the bundled records are stacked to represent higher and higher structures, the projection paths for their components grow longer and longer, resulting in ever more unwieldy terms (coercions and notations can make this less painful).
+Further, if one tries to implement some semblance of sharing in a bundled representation, these projection paths additionally become non-canonical, and still more extensions have been proposed to address this symptom (e.g.
+coercion pullbacks~\cite{Hints}).
 
 Thus, bundled representations come at a substantial cost in flexibility.
-Historically, this cost has been acceptable because (1) the unbundled alternative was such a pain, and (2) the standard algebraic hierarchy (up to, say, fields and modules) is not all that wild.
-
-In the next section, we show that type-classification of structure predicates and their component parameters have the potential to remedy the problems traditionally associated with their use, giving us the best of both worlds.
-
-For completeness, we mention an alternative record-based approach known as ``packed classes'', as used in the ssreflect library.
+Historically, using bundled representations has nevertheless been an acceptable trade-off, because (1) the unbundled alternative was such a pain, and (2) the standard algebraic hierarchy (up to, say, fields and modules) is not all that wild.
+
+In the next section, we show that type-classification of structure predicates and their component parameters has the potential to remedy the problems associated with the naive unbundled predicate approach, giving us the best of both worlds.
+
+%For completeness, we mention an alternative record-based approach known as ``packed classes'', as used in the ssreflect library.
 In the packed classes idiom, each algebraic structure is represented by not one but a trinity of interdependent records along with packing/unpacking functions.
@@
 
-We have found it hard to assess the general merits of this approach, partly because of its complexity, and partly because the only actual usage example (the hierarchy in the ssreflect library) is not only wedded to a very specialized view on equality and decidability that does not support setoid relations from the outset, but also diverges significantly from standard mathematical dichotomy, making it hard to recognize familiar structures and presentation.
+%We have found it hard to assess the general merits of this approach, partly because of its complexity, and partly because the only actual usage example (the hierarchy in the ssreflect library) is not only wedded to a very specialized view on equality and decidability that does not support setoid relations from the outset, but also diverges significantly from standard mathematical dichotomy, making it hard to recognize familiar structures and presentation.
 The reasons for this (as we understand them) are pragmatic and twofold.
 
-First, the ssreflect library was originally developed strictly to support very practical work in a specialized domain of mathematics involving decidable and mostly finite structures.
+%First, the ssreflect library was originally developed strictly to support very practical work in a specialized domain of mathematics involving decidable and mostly finite structures.
 Second, at the time of ssreflect's initial development, support for setoid rewriting in Coq was not nearly as stable and refined as it is today, which made ``Leibniz equality at all costs'' a more reasonable strategy at the time than it might be considered today.
 
-Work to extend the ssreflect framework with techniques to allow representation of a less restricted class of mathematical structures has only recently begun and is still ongoing, though there are currently no plans to support undecidable structures.
-
-Because in our development we will be working with computable reals, equality on which can never be either Leibniz or decidable, we conclude that the ssreflect library is not usable for us.
+%Work to extend the ssreflect framework with techniques to allow representation of a less restricted class of mathematical structures has only recently begun and is still ongoing, though there are currently no plans to support undecidable structures.
+
+%Because in our development we will be working with computable reals, equality on which can never be either Leibniz or decidable, we conclude that the ssreflect library is not usable for us.
 
@@
 
-% We \emph{could} become over-eager and decide to unbundle the equivalence proof term as well:
-% \begin{lstlisting}
-
-% Record Monoid (A: Type) (R: A → A → Prop)
-%   (e: Equivalence R) : Type := ...
-% \end{lstlisting}
-
-% However, this way if we happen to have two proofs |p| and |q| that a certain relation is an
-% equivalence, we end up with two non-convertible ways to express that the relation (together with
-% some unit and binary operation) form a Monoid: 'Monoid A R p ...' and 'Monoid A R q ...'. We must
-% either try to deal with this (by proving that Monoid proofs with different Equivalence proofs are
-% interchangable, and hope for automation, such as type class instance resolution, to use them
-% interchangably), or take great care to ensure that |Equivalence| proof terms are always shared,
-% essentially elevating them to `values', which goes against the principle of proof
-% irrelevance.\marginpar{What if we add proof irrelevance?}
-% BOTTOMLINE: Bundle as much as we can!
-% 
-% Neither of these options are very appealing. We really should just make |Monoid| take as
-% parameters
-% only the operations/relations. If we treat the monoid's unit and binary operation the same way,
-% this
-% means that |Monoid| ends up in |Prop|, and that its instances can be left opaque, which is really
-% how things ought to be.
+The observant reader may wonder whether it might be beneficial to go one step further and unbundle proof of laws and inherited substructures as well.
+This is not the case, because there is no point in sharing them.
+After all, by (weak) proof irrelevance, the ``value'' of such proofs can be of no consequence anyway.
+Indeed, parameterizing on proofs would be actively harmful because instantiations differing only in the proof argument would express the same thing yet be nonconvertible, requiring awkward conversions and hindering automation.
 
@@
 
-\section{Managing unbundled structure with type classes}
-
-So far, we have only looked at the definitions of the structures themselves, and have not yet
-considered their use when proving theorems about these structures.
-We would like to approximate the common mathematical vernacular, where one
-simply says:
-
-for $x, y, z$ in a semigroup $G$, $x * y * z = z * y * x$, $\ldots$
-
-Suppose we have defined \lstinline|SemiGroup| in line with the aforementioned principles:
-\begin{lstlisting}
-Record SemiGroup (A: Type) (eq: A $\to$ A $\to$ Prop)
- (op: A $\to$ A $\to$ A): Prop := ...
-\end{lstlisting}
-
-Then our starting point is the obvious but awful:
-\begin{lstlisting}
-Lemma perm A eq op (P: SemiGroup A eq op) x y z:
-                   eq (op (op x y) z) (op (op z y) x).
-\end{lstlisting}
-
-As a first cleaning measure, we might hope to introduce some notations to let us use infix notation
-for the semigroup operator (say, \lstinline|&|) and for equality.
-Unfortunately, at this point the names \lstinline|eq| and \lstinline|op| are just local names and there is no way to predeclare that they are to be associated with any notation.
-What we really need are canonical names for these notions, so that we may fashion notations for them.
-
-This is a typical job for type classes, and they are used in this way in the \lstinline|Clean| standard
-library. Similarly, Haskell uses type classes to overload notations.
-We simply follow suit:
+\section{Predicate classes}
+
+We will tackle the problems associated with the structure predicates approach one by one, starting with those encountered during theorem \emph{application}.
+Suppose we have defined \lstinline|SemiGroup| as a structure predicate as follows \footnote{Note that defining \lstinline|SemiGroup| as a record instead of a straight conjunction does not make it any less of a predicate.
+The record form is simply more convenient in that it immediately gives us named projections for laws and substructures.}:
+\begin{lstlisting}
+  Record SemiGroup (G: Type) (e: relation G) (op: G → G → G): Prop :=
+    { sg_setoid: Equivalence e
+    ; sg_ass: Associative op
+    ; sg_proper: Proper (e ==> e ==> e) op }.
+\end{lstlisting}
+Then by (1) making \lstinline|SemiGroup| a \emph{class} (by replacing the \lstinline|Record| keyword with the \lstinline|Class| keyword), (2) marking its proofs as \emph{instances} (again by replacing a single keyword), and (3) marking the \lstinline|SemiGroup| parameter of semigroup theorems as implicit (by using curly instead of round brackets), we no longer have to pass \lstinline|SemiGroup| proofs around manually ourselves, letting instance resolution do it for us instead.
+Because instance resolution is part of the unifier, this also works when the statement of the theorem we wish to apply only mentions some of the components (which admittedly doesn't make much sense for semigroups).
+
+Next, we turn to problems concerning theorem \emph{declaration}. As a reference point, our ideal will be the common mathematical vernacular, where one simply says:
+\begin{quote}
+Theorem L: For $x, y, z$ in a semigroup $G$, $x * y * z = z * y * x$.
+\end{quote}
+Without further support from the system, this would have to be written as
+\begin{lstlisting}
+  Theorem L G e op {P: SemiGroup G e op}:
+    Π x y z, e (op (op x y) z) (op (op z y) x).
+\end{lstlisting}
+(The curly brackets mark \lstinline|P| as an implicit argument.) Because \lstinline|e| and \lstinline{op} are freshly introduced local names, we cannot bind notations to them prior to this theorem.
+Hence, if we want notations, what we really need are canonical names for these components.
+This is easily accomplished with single-field type classes containing one component each, which we will call \emph{operational type classes}\footnote{These single-field type classes are used in the same way in the \lstinline|Clean| standard library \cite{something}.}:
 \begin{lstlisting}
@@
 \end{lstlisting}
-
-We choose to use \lstinline|&| here and reserve the notation \lstinline|*| for ring multiplication. We use = for setoid equality, since Leibniz equality seems less important for the formalization of
-mathematics. More precisely, setoid equality is the basic concept. In some special cases, one can prove
-that setoid equality coincides with Leibniz equality.
-%Listings don't seem to work will with footnotes.
-We would have like to use a haskell style definition
-\begin{lstlisting}
-Class Equiv A := (=): relation A.
-\end{lstlisting}
-Unfortunately, currently this is not supported in Coq.
-
-These single-field type classes are given special treatment in that they are not translated into
-records the way classes with any other number of fields are, but are instead turned into ordinary
-definitions.
-The effect of this is that \lstinline|Equiv A| immediately reduces to \lstinline|relation A|, and that the equiv `projection' is the identity function whose argument is inferred using instance resolution.
-Thus, for a very modest cost in indirection, we gain the ability to use canonical names, and with it, notations:
-\begin{lstlisting}
-Record SemiGroup (A: Type) (eq: Equiv A)
-                 (op: SemiGroupOp A): Prop := ...
-Lemma perm A eq op (P: SemiGroup A eq op) x y z:
-                 x & y & z = z & y & x.
-\end{lstlisting}
-
-We call these type classes \emph{operational type classes}. We note that operational type classes
-allow us to avoid Coq's notation \lstinline|scope| mechanism.
-
-Although the expression now \emph{looks} like it is bound to some
-specific semigroup structure (which with traditional approaches would imply projections from bundles
-and/or reference to proofs), it is really only referring near-directly to the actual operations
-involved, with the semigroup proof existing only as opaque `knowledge about the operations' which we
-may use in the proof. This lack of projections keeps our terms small and independent, and keeps
-rewriting simple and sane; see Section~\ref{canonical}.
-
-On the other hand, it \emph{does} mean that all the carriers/relations/operations end up in the
-context. The context for \lstinline|perm| looks like:
-\begin{lstlisting}
-  A: Type
-  eq: Equiv A
-  op: SemiGroupOp A
-  P: SemiGroup A eq op
-\end{lstlisting}
-
-While we are not particularly worried about overly large contexts (we feel that this will most
-likely not be problematic for any but the most complex formalizations), having to \emph{declare}
-these entities (like \lstinline|eq| and \lstinline|op| for \lstinline|perm|) \emph{is} a bit of a chore.
-Fortunately, Coq provides a feature called implicit generalization, which is exactly what we need.
-With it, we can write \lstinline|perm| as follows:
-\begin{lstlisting}
-  Lemma perm `(SemiGroup A) x y z: x & y & z = z & y & x.
-\end{lstlisting}
-
+We use \lstinline|&| here and reserve the notation \lstinline|*| for (semi)ring multiplication.
+
+As an aside, we note that the distinction between the class field name and the infix operator notation bound to it is really just a mildly awkward Coq artifact.
+In Haskell, where operators can themselves be used as names, there would be no need to have the \lstinline|equiv| and \lstinline|sg_op| names in addition to the operator ``names''.
+
+If we now retype \lstinline|SemiGroup| as:
+\begin{lstlisting}
+  Π (G: Type) (e: Equiv G) (op: SemiGroupOp G), Prop
+\end{lstlisting}
+then we can declare the theorem with:
+\begin{lstlisting}
+  Theorem L G e op {P: SemiGroup G e op}: Π x y z, x & y & z = z & y & x.
+\end{lstlisting}
+This works because instance resolution, invoked by the use of \lstinline|=| and \lstinline|&|, will find \lstinline|e| and \lstinline|op|, respectively. Hence, the above is really
+
+\begin{lstlisting}
+  Theorem L G e op {P: SemiGroup G e op}: Π x y z,
+    equiv e (sg_op op (sg_op op x y) z) (sg_op op (sg_op op x y) z).
+\end{lstlisting}
+where e and the op's are filled in by instance resolution.
+
+At this point, a legitimate worry might be that the \lstinline|Equiv|/\lstinline|SemiGroup| classes and their \lstinline|equiv|/\lstinline|sg_op| projections imply constant construction and deconstruction of records, harming the simplicity and flexibility of the predicate approach that we are trying so hard to preserve.
+No such construction and destruction takes place, however, because type classes with only a single field are not desugared into an actual record with field projections the way classes with any other number of fields are.
+Instead, both class itself and its field projection are defined as the identity function with a fancy type.
+Thus, the introduction of these canonical names is essentially free; the structure predicate's new type reduces straight back to what it was before.
+
+A remaining eyesore in the theorem declaration is the enumeration of \lstinline|e| and \lstinline|op|.
+To remove these, we use a new parameter declaration feature called \emph{implicit generalization}, introduced in Coq specifically to support type classes.
+Using implicit generalization, we can write:
+
+\begin{lstlisting}
+  Lemma L `{SemiGroup G}: Π x y z: G, x & y & z = z & y & x.
+\end{lstlisting}
+% TODO: describe implicit generalization in two words
+% TODO: at some point down the line in our coq-inadequacies section, describe the implicit generalization improvements we need.
 Thus, we have reached the mathematical ideal we aimed for.
 
-\section{Putting it all together}\label{classes}
-Combining the ideas in the previous sections we end up with a formalization of algebraic
-structures where we use the structure as (implicit) parameters, but
-pack the proofs. This can be nicely organized by using canonical names. For instance, we define:
-\begin{lstlisting}
-Class SemiRing A {e: Equiv A}{plus: RingPlus A}{mult: RingMult A}
-                 {zero: RingZero A}{one: RingOne A}:Prop :=
-  { semiring_mult_monoid:> Monoid A (op:=mult)(unit:=one)
-  ; semiring_plus_monoid:> Monoid A (op:=plus)(unit:=zero)
-  ; semiring_plus_comm:> Commutative plus
-  ; semiring_mult_comm:> Commutative mult
+% We note that operational type classes allow us to avoid Coq's notation \lstinline|scope| mechanism.
+% hm, i don't think this is true.
+we don't need scopes for different representations' use of operators because we properly recognize their commonality, but that can be done without operational type classes as well.
+
+% TODO: get this in:
+% Although the expression now \emph{looks} like it is bound to some
+% specific semigroup structure (which with traditional approaches would imply projections from bundles
+% and/or reference to proofs), it is really only referring near-directly to the actual operations
+% involved, with the semigroup proof existing only as opaque `knowledge about the operations' which we
+% may use in the proof. This lack of projections keeps our terms small and independent, and keeps
+% rewriting simple and sane; see Section~\ref{canonical}.
+
+One aspect of the predicate approach we have not mentioned thus far is that in proofs parameterized by abstract structures, all components become hypotheses in the context.
+For the example theorem above, the context looks like:
+\begin{lstlisting}
+  G: Type
+  e: Equiv G
+  op: SemiGroupOp G
+  P: SemiGroup G e op
+\end{lstlisting}
+We are not particularly worried about overly large contexts, especially because most of the ``extra'' hypotheses we have compared to bundled approaches are declarations of relations, operators, and constants, which are all in some sense ``inert'' with respect to proof search.
+Hence, we do not foresee problems with large contexts for any but the most complex formalizations.
+
+\section{The algebraic hierarchy}\label{classes}
+
+We have developed an algebraic hierarchy composed entirely out of predicate classes and operational classes as described in the previous section. For instance, our semiring interface looks as follows:
+\begin{lstlisting}
+  Class SemiRing A {e: Equiv A}
+      {plus: RingPlus A} {mult: RingMult A}
+      {zero: RingZero A} {one: RingOne A}: Prop :=
+    { semiring_mult_monoid:> CommutativeMonoid A (op:=mult)(unit:=one)
+    ; semiring_plus_monoid:> CommutativeMonoid A (op:=plus)(unit:=zero)
     ; semiring_distr:> Distribute mult plus
-  ; mult_0_l: Π x, 0 * x = 0 }.
-\end{lstlisting}
-
-where e.g.\ \lstinline|RingPlus| is the operation class for the semigroup operator.
-The notation \lstinline|:>| makes \lstinline|semiring_mult_monoid| a subclass: when asked for a monoid it suffices to
-provide a ring.
-
-Having argued that the \emph{all-structure-as-parameters} approach \emph{can} be made practical, we enumerate some of the benefits that make it worthwhile.
-
-First, multiple inheritance becomes trivial: \lstinline|SemiRing| inherits two \lstinline|Monoid| structures on the same carrier and setoid relation, using ordinary named arguments to achieve ``manifest fields''; see section~\ref{manifest}.
-
-Second, because our terms are small and independent and never refer to proofs, we are invulnerable to concerns about efficiency and ambiguity of projection paths that plague existing solutions, obviating the need for extensions like the proposed coercion pullbacks~\cite{Hints}.
-
-Third, since our structural type classes are mere predicates, overlap between their instances is a non-issue.
+    ; semiring_left_absorb:> LeftAbsorb mult zero }.
+\end{lstlisting}
+All of \lstinline|Equiv|, \lstinline|RingPlus|, \lstinline|RingMult|, \lstinline|RingZero|, and \lstinline|RingOne| are operational (single-field) classes, with bound notations \lstinline|=|, \lstinline|+|, \lstinline|*|, \lstinline|0|, and \lstinline|1|, respectively.
+
+Let us briefly highlight some additional aspects of this style of structure definition in more detail.
+
+Fields declared with \lstinline|:>| are registered as hints for instance resolution, so that in any context where \lstinline|(A, =, +, 0, *, 1)| is known to be a \lstinline|SemiRing|, \lstinline|(A, =, +, 0)| and \lstinline|(A, =, *, 1)| are automatically known to be \lstinline|CommutativeMonoid|s (and so on, transitively, because instance resolution is recursive).
+In our hierarchy, these fields by themselves establish the inheritance diagram in Figure \ref{inheritancediagram}.
+% todo: for some reason the figure number reference is wrong here..
+
+\begin{figure}
+\label{inheritancediagram}
+\centering
+\includegraphics{hierarchy.pdf}
+\caption{Inheritance relations established by \lstinline|:>| fields.}
+\end{figure}
+
+However, we can easily add additional inheritance relations by declaring corresponding class instances.
+For instance, while our \lstinline|Ring| class does not have a \lstinline|SemiRing| field, the following instance declaration has the exact same effect for the purposes of instance resolution (at least once proved, which is trivial):
+\begin{lstlisting}
+  Instance ring_as_semiring `{Ring R}: SemiRing R.
+\end{lstlisting}
+
+Thus, axiomatic structural properties and inheritance have precisely the same status as separately proved structural properties and inheritance, reflecting natural mathematical ideology.
+Again, contrast this with bundled approaches, where axiomatic inheritance relations determine projection paths, and where additional inheritance relations lead to additional and ambiguous projection paths for the same operations.
+
+% todo: problem with ambiguous projection paths: quoting will not notice convertability.
+
+The declarations of the two inherited \lstinline|CommutativeMonoid| structures nicely illustrate how predicate classes naturally support not just multiple inheritance, but \emph{overlapping} multiple inheritance, where the inherited structures may share components (in this case carrier and equivalence relation).
+The carrier \lstinline|A|, being an explicit argument, is specified as normal.
+The equivalence relation, being an implicit argument of class type, is resolved automatically to \lstinline|e| and passed.
+The binary operation and constant would normally be automatically resolved and passed as well, but we override the inference mechanism locally using Coq's existing named argument facility (which is only syntactic sugar of the most superficial kind) in order to specify the correct pairings.
+Again, contrast this with type system extensions such as Matita's manifest records, which are required to make this work when the records bundle components such as \lstinline|op| and \lstinline{unit} as \emph{fields} instead of parameters.
+
+Since \lstinline|CommutativeMonoid| indirectly includes a \lstinline|SemiGroup| field, which in turn includes a \lstinline|Equivalence| field, having a \lstinline|SemiRing| proof means having two distinct proofs that the equality relation is an equivalence.
+This kind of redundant knowledge (which arises naturally in mathematics) is never a problem in our setup, because the use of operational type classes ensures that terms composed of algebraic operations and relations never refer to structure proofs.
+We find that this is a tremendous relief compared to approaches that \emph{do} intermix the two and where one must be careful to ensure that such terms refer to the ``right'' proofs of properties.
+There, even \emph{strong} proof irrelevance (which would make terms convertible that differ only in what proofs they refer) would not make these difficulties go away entirely, because high-level tactics that rely on quotation of terms require syntactic identity (rather than ``mere'' convertibility) to recognize identical subterms.
+
+% rather strict separation of operations and proofs of properties about them (resolved by instance resolution) provides a very pleasing environment to work in, without any need for hacks.
+% terms composed of algebraic operations and relations 
+% separation of , this kind of redundance is never a problem for us. 
+% it is /never/ a problem to know a fact twice.
+% multiple proofs of equality, and even of semigroup-ness (since both Monoid), but that's simply not a problem, since none of our terms ever refers to such proofs.
+We have found that this rather strict separation of operations and proofs of properties about them (resolved by instance resolution) provides a very pleasing environment to work in, without any need for hacks.
+
+The reader will note that we use names for properties like distributivity and absorption, and in fact these are type classes as well (which is why we declare them with \lstinline|:>|).
+It has been our experience that almost any generic property worth naming is worth representing as a type class, so that its proofs are be resolved as instances behind the scenes whenever possible.
+Doing this consistently minimizes administrative noise in the code, bringing us closer to ordinary mathematical vernacular.
+Indeed, we believe that type classes are an elegant and apt formalization of the seemingly casual manner in which ordinary mathematical presentation assumes implicit administration and use of a ``database'' of structural properties previously proved, much more so than existing solutions using, for instance, canonical structures.
+
+% Unfortunately, aggressive . More on taht in section bla
+
+% Having argued that the \emph{all-structure-as-parameters} approach \emph{can} be made practical, we enumerate some of the benefits that make it worthwhile.
+% 
+% First, multiple inheritance becomes trivial: \lstinline|SemiRing| inherits two \lstinline|Monoid| structures on the same carrier and setoid relation, using ordinary named arguments to achieve ``manifest fields''; see section~\ref{manifest}.
+% 
+% Second, because our terms are small and independent and never refer to proofs, we are invulnerable to concerns about efficiency and ambiguity of projection paths that plague existing solutions, obviating the need for extensions like the proposed coercion pullbacks~\cite{Hints}.
+% 
+% Third, since our structural type classes are mere predicates, overlap between their instances is a non-issue.
 Together with the previous point, this gives us tremendous freedom to posit multiple structures on the same operations and relations, including ones derived implicitly via subclasses: by simply declaring a \lstinline|SemiRing| class instance showing that a ring is a semiring, results about semirings immediately apply implicitly to any known ring, without us having to explicitly encode this relation in the hierarchy definition itself, and without needing any projection or translation of carriers or operations.
-
-Fourth, \emph{Structure-as-parameters} helps setoid-rewriting: type class resolution
-can find the equivalence relation in the context.\footnote{
-A similar style should be possible for, say, the \lstinline|Ring| tactic, instead of
-declaring the ring structure by a separate command, we would rely on type class resolution to find it.}
-We note that \lstinline|op| does not depend on the proof that \lstinline|e| is an equivalence.
+% 
+% Fourth, \emph{Structure-as-parameters} helps setoid-rewriting: type class resolution
+% can find the equivalence relation in the context.\footnote{
+% A similar style should be possible for, say, the \lstinline|Ring| tactic, instead of
+% declaring the ring structure by a separate command, we would rely on type class resolution to find it.}
+% We note that \lstinline|op| does not depend on the proof that \lstinline|e| is an equivalence.
 As explained above we use Coq's implicit quantification (\lstinline|`{}|) to avoid having to write all the parameters when \emph{stating} a theorem and Coq's maximally inserted implicit arguments to find the parameters when \emph{applying} a theorem.
 Both features are new in Coq and stem from the type class implementation.
-
-We mention the trade-off between bigger contexts versus bigger terms. Our contexts are bigger than
-those of telescopes or packed classes. In our experience, this has been relatively
-harmless\footnote{However, Coq's data structure for contexts is not very efficient. Gonthier fears that this
-may be a bottleneck for huge developments. It seems that the data structure chosen
-in~\cite{asperti2009compact} will behave better.}%
-: most terms in the context are there to support canonical names. Bigger terms
-\emph{do} cause problems: 1. when proofs are part of mathematical objects we need to share these
-proofs to allow rewriting. Moreover, it prohibits Opaque proofs and `proof irrelevance'. 2. The
-projection paths may not be canonical.
-
-Coercion pullbacks~\cite{Hints} were introduced to address problems with multiple coercions paths,
-as in the definition of a semiring: a type with two monoid structures on it. We avoid some
-of these problems by explicitly specifying the fields. We emphasize that the semiring properties are
-automatically derived from the ring properties, although the properties of a semiring are not
-structurally included in the ring properties.
-
-Let us now stop and think to what extent this approach suffers from all the problems commonly
-associated with it. In particular, let us imagine what happens to our terms and contexts when we
-want to talk about nested structures such as polynomials over the ring. For a concrete
-representation of the polynomials the context will just contain the context for an abstract ring.
-For abstract reasoning about polynomials the context will grow with abstract operations on
-polynomials. Consequently, the context will grow linearly, as opposed to exponentially.
-
-\subsection{Manifest fields}\label{manifest}
-
-Manifest fields in records allow us to fix a field of a dependent record. For example, in the
-example above one would write \lstinline|Monoid with op=mult|. 
-Luo~\cite{DBLP:conf/types/Luo08} has proposed coercive subtyping to implement manifest fields.
-It was conjectured~\cite{Hints} that implementing the algebraic hierarchy with type classes in this way would
-be awkward due to problems with equality. We avoided these problems by shifting the fields
-from the dependent record to its argument, thus mimicking manifest fields by using type classes.
-Since \lstinline|op| is \emph{instantiated} with \lstinline|mult|, we need not worry about the
-equality \lstinline|op=mult|. This seems to be a general phenomenon. For instance, instead of searching
-for a \lstinline|Monoid with op=mult| we would use type class resolution to find an instance of \lstinline|Monoid mult|.
-We have thus addressed the problem of multiple inheritance.
+% 
+% We mention the trade-off between bigger contexts versus bigger terms. Our contexts are bigger than
+% those of telescopes or packed classes. In our experience, this has been relatively
+% harmless\footnote{However, Coq's data structure for contexts is not very efficient. Gonthier fears that this
+% may be a bottleneck for huge developments. It seems that the data structure chosen
+% in~\cite{asperti2009compact} will behave better.}%
+% : most terms in the context are there to support canonical names. Bigger terms
+% \emph{do} cause problems: 1. when proofs are part of mathematical objects we need to share these
+% proofs to allow rewriting. Moreover, it prohibits Opaque proofs and `proof irrelevance'. 2. The
+% projection paths may not be canonical.
+% 
+% Coercion pullbacks~\cite{Hints} were introduced to address problems with multiple coercions paths,
+% as in the definition of a semiring: a type with two monoid structures on it. We avoid some
+% of these problems by explicitly specifying the fields. We emphasize that the semiring properties are
+% automatically derived from the ring properties, although the properties of a semiring are not
+% structurally included in the ring properties.
+% 
+% Let us now stop and think to what extent this approach suffers from all the problems commonly
+% associated with it. In particular, let us imagine what happens to our terms and contexts when we
+% want to talk about nested structures such as polynomials over the ring. For a concrete
+% representation of the polynomials the context will just contain the context for an abstract ring.
+% For abstract reasoning about polynomials the context will grow with abstract operations on
+% polynomials. Consequently, the context will grow linearly, as opposed to exponentially.
+
+Having described the basic principles of our approach, in the remainder of this paper we tour other parts of our development, further illustrating what a state of the art formal development of foundational mathematical structures can look like with a modern proof assistant.
+
+These parts were originally motivated by our desire to cleanly express interfaces for basic numeric data types such as $\N$ and $\Z$ in terms of their categorical characterization as initial objects in the categories of semirings and rings, respectively.
+Let us start, therefore, with basic category theory.
+
+\section{Category theory}\label{cat}
+
+Following our idiom, we introduce operational type classes for the \emph{components} of a category:
+
+\begin{lstlisting}
+Class Arrows (O: Type): Type := Arrow: O → O → Type.
+Class CatId O `{Arrows O} := cat_id: `(x ⟶ x).
+Class CatComp O `{Arrows O} :=
+  comp: Π {x y z}, (y ⟶ z) → (x ⟶ y) → (x ⟶ z).
+
+Infix "⟶" := Arrow (at level 90, right associativity).
+Infix "◎" := comp (at level 40, left associativity).
+\end{lstlisting}
+(The categorical arrow is distinguished from the primitive function space arrow by its length.)
+
+With these in place, our type class for categories follows the usual type-theoretical definition of a
+category~\cite{saibi1995constructive}:
+
+\begin{lstlisting}
+Class Category (O: Type) `{!Arrows O} `{Π x y: O, Equiv (x ⟶ y)}
+    `{!CatId O} `{!CatComp O}: Prop :=
+  { arrow_equiv:> Π x y, Setoid (x ⟶ y)
+  ; comp_proper:> Π x y z, Proper (equiv ==> equiv ==> equiv) comp
+  ; comp_assoc w x y z (a: w ⟶ x) (b: x ⟶ y) (c: y ⟶ z):
+      c ◎ (b ◎ a) = (c ◎ b) ◎ a
+  ; id_l `(a: x ⟶ y): cat_id ◎ a = a
+  ; id_r `(a: x ⟶ y): a ◎ cat_id = a }.
+\end{lstlisting}
+This definition is based on the 2-categorical idea of having equality only on arrows, not on objects.
+Similarly, we will have equality on natural transformations, but not on functors. % todo: ORLY?
+
+Initiality, too, is defined by a combination of an operational and a predicate class:
+
+\begin{lstlisting}
+  Context `{Category X}.
+  Class InitialArrow (x: X): Type := initial_arrow: Π y, x ⟶ y.
+  Class Initial (x: X) `{InitialArrow x}: Prop :=
+    initial_arrow_unique: Π y (a: x ⟶ y), a = initial_arrow y.
+\end{lstlisting}
+
+Having \lstinline|InitialArrow| as a type class means that we can always simply say \lstinline|initial_arrow y| whenever \lstinline|y| is known to be an object in a category known to have an initial object (where ``known to'' refers to the instances available to instance resolution).
+
+Strictly speaking the above is all we need in order to continue with the story line leading up to the numerical interfaces, but just to give a further taste of what category theory with this setup looks like in practice, we briefly mention a few more definitions and theorems.
+
+\subsection{Functors}
+
+In our definition of functors we see the by now familiar refrain once more:
+\begin{lstlisting}
+  Context `{Category C} `{Category D} (map_obj: C → D).
+
+  Class Fmap: Type :=
+    fmap: Π {v w: C}, (v ⟶ w) → (map_obj v ⟶ map_obj w).
+
+  Class Functor `(Fmap): Prop :=
+    { functor_from: Category C
+    ; functor_to: Category D
+    ; functor_morphism:> Π a b: C, Setoid_Morphism (@fmap _ a b)
+    ; preserves_id: `(fmap (cat_id: a ⟶ a) = cat_id)
+    ; preserves_comp `(f: y ⟶ z) `(g: x ⟶ y): fmap (f ◎ g) = fmap f ◎ fmap g }.
+\end{lstlisting}
+Some clarification is in order to explain the role of the \lstinline|Context| declaration of the two categories.
+While it may look we are parameterizing both \lstinline|Fmap| and \lstinline|Functor| on proofs (which would be a gross violation of our idiom), in fact they depend and are parameterized only on the \emph{components} declared through implicit generalization of the \lstinline|Category| declarations.
+Indeed, we use the latter strictly as a convenient way to declare the former.
+
+Notice that \lstinline|map_obj| is \emph{not} made into an operational type class.
+The reason for this is that \lstinline|map_obj| is analogous to the carrier type in earlier structure predicate class definitions, in that it serves as the primary identification for the structure, and should therefore not be inferred.
+
+The \lstinline|functor_to| and \lstinline|functor_from| fields in \lstinline|Functor| are not an absolute necessity, but eliminate the need for theorems to declare \lstinline|Category| parameters when they already declare \lstinline|Functor|s between them.
+This is an instance where we can freely posit structural properties without worrying about potential problems when such information turns out to be redundant in contexts where the source and target of the functor are already known to be \lstinline|Category|'s.
+
+Unfortunately, there is actually one annoying wrinkle here, which will also explain why, as the observant reader will have noticed, we do not register them as instance resolution hints (by declaring them with \lstinline|:>|).
+
+What we really want these fields to express is ``\emph{if} in a certain context we know something to be a functor, \emph{then} realize that the source and target are categories''.
+However, the current instance resolution implementation does not support this style of exclusively-forward implications, and only supports backward implications: had we registered \lstinline|functor_to| and \lstinline|functor_from| as instance resolution hints, we would in fact be saying ``\emph{if} trying to establish that something is a category, \emph{then} you might try finding a functor to or from it'', which quickly degenerates into a wild goose chase.
+
+We will say more about this unfortunate limitation of the current type class implementation in section \ref{wishlist}.
+
+Finally, we ought to say a few words about our use of \lstinline|fmap|.
+The usual mathematical notational convention for functor application is to use the name of the functor to refer to both its object map and its arrow map, relying on additional conventions regarding object/arrow names for disambiguation (i.e.
+``\lstinline|F x|'' and ``\lstinline|F f|'' map an object and an arrow, respectively, because ``\lstinline|x|'' and ``\lstinline|f|'' are conventional names for objects and arrows, respectively.
+
+In Coq, for a term \lstinline|F| to function as though it had two different types simultaneously (namely the object map and the arrow map), there must either (1) be coercions from the type of F to either function, or (2) F must be (coercible to) a single function that is able to consume both object and arrow arguments.
+In addition to not being supported by Coq, option (1) would violate our policy of leaving components unbundled.
+For (2), if it could be made to work at all (which is not clear at all), F would need a pretty egregious type considering that arrow types are indexed by objects, and that the type of the arrow map (namely ``\lstinline|Π x y, (x ⟶ y) → (F x ⟶ F y)|'') must refer to the object map.
+
+We feel that these issues are not limitations of the Coq system, but merely reflect the fact that notationally identifying these two different and interdependent maps is a typical example of an ``abus de notation'' that by its very nature is ill-suited to a formal development where software engineering concerns apply.
+Hence, we do not adopt this practice, and use ``\lstinline|fmap F|'' (name taken from the Haskell standard library) to refer to the arrow map of a functor \lstinline|F|.
+
+\subsection{Natural transformations}
+blabla, routine.
+
+%theory/categories.v
+Natural transformations are formalized as:
+\begin{lstlisting}
+   Context `{Category C} `{Category D}
+   `{!Functor (F: C ⟶ D) Fa} `{!Functor (G: C ⟶ D) Ga}.
+   Class NaturalTransformation (η: Π c, F c ⟶ G c): Prop :=
+     natural: Π (x y: C) (f: x ⟶ y), η$$ y ∘ fmap F f = fmap G f ∘ η$$ x.
+\end{lstlisting}
+  % Todo: Having to add "$$" to get whitespace is awful. we can't add "\ " in η's literate-replacement though, because then we also get space before the colon in "η: t".
+
+Here \lstinline|fmap F| explicitly refers to the arrow part of the functor \lstinline|F|.
+In mathematics one would apply $F$ to both objects and arrows ($F x$, $F f$). However, this does
+not seem to fit with type theory as the type of \lstinline|fmap F| depends on the object part of
+\lstinline|F|.
+
+We have defined the category of setoids, the dual of a category, products, and hence co-products. 
+We prove that two definitions of adjunction are equivalent. The proofs are concise and follow the textbooks.
+
+The usual size problems in the definition of the category of categories can be avoided by using universe polymorphism. 
+However, we need to avoid making \lstinline|Relation| a \lstinline|Definition|, since \lstinline|Definitions| are not (yet?) universe polymorphic.
+
+\subsection{Adjunctions}
+blabla, example of nontrivial theorem. nice stress-test, worked out mostly fine.
 
 \section{Universal algebra}\label{univ}
+
 Motivated originally by our desire to cleanly express interfaces for basic numeric data types such as $\N$ and $\Z$ in terms of their categorical characterization as initial objects in the categories of semirings and rings, respectively, we initially introduced only the very basics of category theory into our development, again using type classes where possible to achieve the same benefits mentioned before.
@@
 Having discussed the first homomorphism theorem we now turn to category theory which helps to organize our development of universal algebra.
-
-\subsection{Category theory}\label{cats}
-Using type classes, we provide a basic library for category theory: categories, functors,
-adjunctions, monads,\ldots We follow the usual type theoretical definition of a
-category~\cite{saibi1995constructive}.
-
-% \newcommand{\rightarrowtext}{\ensuremath{\rightarrow}}
-% \newcommand{\textocirc}{\ensuremath\circ}
-%interfaces/abstract_algebra.v
-
-\begin{lstlisting}
-Class Category O `{!Arrows O} `{Π x y: O, Equiv (x ⟶ y)}
-    `{!CatId O} `{!CatComp O}: Prop :=
-  { arrow_equiv:> Π x y, Setoid (x ⟶ y)
-  ; comp_proper:> Π x y z,
-    Proper (equiv ==> equiv ==> equiv) comp
-  ; comp_assoc w x y z (a: w ⟶ x) (b: x ⟶ y) (c: y ⟶ z):
-      c ◎ (b ◎ a) = (c ◎ b) ◎ a
-  ; id_l `(a: x ⟶ y): cat_id ◎ a = a
-  ; id_r `(a: x ⟶ y): a ◎ cat_id = a }.
-\end{lstlisting}
-
-This is based on the 2-categorical idea that we have equality on arrows, but not on objects.
-Similarly, we have equality on natural transformations, but not on functors.
-
-%theory/categories.v
-Natural transformations are formalized as:
-\begin{lstlisting}
-   Context `{Category C} `{Category D}
-   `{!Functor (F: C ⟶ D) Fa} `{!Functor (G: C ⟶ D) Ga}.
-   Class NaturalTransformation (η: Π c, F c ⟶ G c): Prop :=
-     natural: Π (x y: C) (f: x ⟶ y), η$$ y ∘ fmap F f = fmap G f ∘ η$$ x.
-\end{lstlisting}
-  % Todo: Having to add "$$" to get whitespace is awful. we can't add "\ " in η's literate-replacement though, because then we also get space before the colon in "η: t".
-
-Here \lstinline|fmap F| explicitly refers to the arrow part of the functor \lstinline|F|.
-In mathematics one would apply $F$ to both objects and arrows ($F x$, $F f$). However, this does
-not seem to fit with type theory as the type of \lstinline|fmap F| depends on the object part of
-\lstinline|F|.
-
-We have defined the category of setoids, the dual of a category, products, and hence co-products. 
-We prove that two definitions of adjunction are equivalent. The proofs are concise and follow the textbooks.
-
-The usual size problems in the definition of the category of categories can be avoided by using universe polymorphism. 
-However, we need to avoid making \lstinline|Relation| a \lstinline|Definition|, since \lstinline|Definitions| are not (yet?) universe polymorphic.
 
@@
 
+\section{Coq wishlist}
+% todo: maybe merge with conclusions and add "and future work"
+
+While we have already found the predicate-class-based style of formalization to work very well for our development thus far, and believe it has the potential to be \emph{the} modern solution to this problem, in our work we have encountered a number of limitations in the Coq implementation of key features of the tool, that keep us from making an unconditional recommendation for universal adoption of these techniques.
+
+\paragraph{Universe-polymorphic definitions}
+Coq has supported universe-polymorphism for parameterized inductive definitions (different instantiations of which are automatically of the lowest possible sort taking into account the universe level constraints on the arguments) for some time.
+However, the same functionality has not been available for \emph{definitions}, primarily because of efficiency concerns.
+In our development, bla bla misery and drama when doing this and that.
+In response to our inquiries about this issue, the Coq development team has agreed to implement ``opt-in'' universe polymorphism, where only expressly annotated definitions are made universe-polymorphic.
+This should solve our problems.
+
+\paragraph{Implicit generalization}
+
+The current syntax for implicit generalization 
+%current syntax very subtle.
+various combinations of `, class-ness, {} vs (), !
+or no !, make _ mean different things.
+Worse, one strategy of ``infer if possible, generalize otherwise'' is not supported while it seems like a very commonly needed thing.
+The Coq development team recognizes this as a legitimate problem, but fears the implementation might not be straightforward, and as of this writing we are not aware of plans to address this.
+That said, a new proof engine has recently, whcih might make this more attainable.
+Fortunately, not that big a problem.
+Just a remaining source of the occasional uglyness in declarations.
+
+\paragraph{Instance resolution efficiency} ...
+
+The current implementation is in some sense only a prototype.
+
+Backward
+
+Forward
+
+It seems likely that known techniques from established resolution engines can be reused to significantly increase performance.
+
+In reply to our inquiries about these issues the Coq development team has agreed that these are legitimate concerns and wishes, and indicated that they are committed to addressing them.
+
 \section{Conclusions}

### 2010-06-14 compile.sh

@@ -1,7 +1,7 @@
 #!/bin/sh
 
-lhs2TeX --agda< $1.tex > processed.tex
-cp processed.tex tmp.tex
+cp jar.tex tmp.tex
+dot -Tpdf hierarchy.dot > hierarchy.pdf
 
 set -e
 pdflatex tmp.tex
@@ -9,5 +9,5 @@ bibtex tmp
 pdflatex tmp.tex
 bibtex tmp
 pdflatex tmp.tex
-cat tmp.pdf > $1.pdf
+cat tmp.pdf > jar.pdf
 rm tmp.*

### 2010-06-15 jar.tex
spelling

@@
 We present a new formalization of the algebraic hierarchy in Coq, exploiting its new type class mechanism to make practical a solution formerly thought infeasible.
-Our approach addresses both traditional challenges as well as new ones resulting from our ambition to build upon this development a library of constructive analysis in which abstraction penalties inhibiting efficient computation are reduced to a bare minimum.
+Our approach addresses both traditional challenges as well as new ones resulting from our ambition to build upon this development a library of constructive analysis in which abstraction penalties inhibiting efficient computation are reduced to a minimum.
 To support mathematically sound abstract interfaces for $\N$, $\Z$, and $\Q$, our formalization includes portions of category theory and multisorted universal algebra.
@@
 Adopting this regime of discrimination establishes a weak form of proof irrelevance, in that changing a proof can never affect the result of value computations.
-On a very practical level, this lets the extraction mechanism safely erase all \lstinline|Prop| components when extracting certified programs to OCaml or Haskell.
+On a very practical level, this lets Coq safely erase all \lstinline|Prop| components when extracting certified programs to OCaml or Haskell.
 
@@
 
-Manifest fields \ref{manifestrecords} have been proposed and implemented in the Matita system to address exactly this problem.
+Manifest fields~\ref{manifestrecords} have been proposed and implemented in the Matita system to address exactly this problem.
 We hope to convince the reader that type system extensions of this nature, designed to mitigate particular symptoms of the bundled approach, are less elegant than a solution that avoids the problem altogether by using predicate-like type classes in place of bundled records.
@@
 %$TX:=1+T^2X$, may be seen as the prototypical monad. 
-The \lstinline|Expr| monad, i.e.\ the free \lstinline|Expr|-algebra monad\footnote{on the category of \lstinline|Type|s with extensional functions.%\marginpar{Proof in classquote}
-}, provides a bind operation which describes the arrows in the Kleisli category. Concretely,
+The \lstinline|Expr| monad, i.e.\ the free \lstinline|Expr|-algebra monad on the category of \lstinline|Type|s with extensional functions%\marginpar{Proof in classquote}
+, provides a bind operation which describes the arrows in the Kleisli category. Concretely,
 \begin{lstlisting}

### 2010-06-16 jar.tex
spelling

@@
 
-The ``native'' notion of equality in Coq is that of terms being convertible, which in this context constitutes Leibniz equality, naturally propositionalized by an inductive type family \lstinline|eq| with a single constructor \lstinline|eq_refl| of type \lstinline|Π (T: Type) (x: T), x ≡ x|, where ``\lstinline|a ≡ b|'' is notation for \lstinline|eq T a b|~\footnote{We diverge from Coq traditional and reserve ``\lstinline|=|'' for setoid equality, as this is the equality we will be working with most of the time.}.
+The ``native'' notion of equality in Coq is that of terms being convertible, which in this context constitutes Leibniz equality, naturally propositionalized by an inductive type family \lstinline|eq| with a single constructor \lstinline|eq_refl| of type \lstinline|Π (T: Type) (x: T), x ≡ x|, where ``\lstinline|a ≡ b|'' is notation for \lstinline|eq T a b|.
+Here  we diverge from Coq tradition and reserve ``\lstinline|=|'' for setoid equality, as this is the equality we will be working with most of the time.
 Importantly, since convertibility is a congruence, a proof of \lstinline|a ≡ b| lets us substitute \lstinline|b| for \lstinline|a| anywhere inside a term without further conditions.
@@
 \end{lstlisting}
-(The curly brackets used for \lstinline|A| mark it as an implicit argument.)
+The curly brackets used for \lstinline|A| mark it as an implicit argument.
 
@@
 \end{lstlisting}
-Superficially, this instantly solves the problems described above: reflexive relations can now be declared and passed as self-contained packages, and the \lstinline|rr_rel| projection now constitutes a ``canonical name'' for relations that are known to be reflexive, and we could bind a notation to it\footnote{While there is no conventional notation for reflexive relations, the situation is the same in the context of, say, a semiring, where we would bind ``$+$'' and ``$*$'' notations to the record field projections for the addition and multiplication operations, respectively.}.
+Superficially, this instantly solves the problems described above: reflexive relations can now be declared and passed as self-contained packages, and the \lstinline|rr_rel| projection now constitutes a ``canonical name'' for relations that are known to be reflexive, and we could bind a notation to it.
+While there is no conventional notation for reflexive relations, the situation is the same in the context of, say, a semiring, where we would bind ``$+$'' and ``$*$'' notations to the record field projections for the addition and multiplication operations, respectively.
 
@@
 
-If we revert back to the predicate formulation of reflexive (et. al) relations, we \emph{could} still define \lstinline|EquivalenceRelation| in a bundled fashion without the need for equalities:
+If we revert back to the predicate formulation of relations, we \emph{could} still define \lstinline|EquivalenceRelation| in a bundled fashion without the need for equalities:
 \begin{lstlisting}
@@
 \end{lstlisting}
-However, the same argument as before leads to the conclusion that \lstinline|EquivalenceRelation|, too, should be a predicate.
+However, the same argument as before leads to the conclusion that \mbox{\lstinline|EquivalenceRelation|,} too, should be a predicate.
 Indeed, it would be rather strange for the interface of equivalence relations to differ qualitatively from the interface of reflexive relations.
@@
 We find that this is a tremendous relief compared to approaches that \emph{do} intermix the two and where one must be careful to ensure that such terms refer to the ``right'' proofs of properties.
-There, even \emph{strong} proof irrelevance (which would make terms convertible that differ only in what proofs they refer) would not make these difficulties go away entirely, because high-level tactics that rely on quotation of terms require syntactic identity (rather than ``mere'' convertibility) to recognize identical subterms.
+There, even \emph{strong} proof irrelevance (which would make terms convertible that differ only in what proofs they refer to) would not make these difficulties go away entirely, because high-level tactics that rely on quotation of terms require syntactic identity (rather than ``mere'' convertibility) to recognize identical subterms.
 
@@
 
-The reader will note that we use names for properties like distributivity and absorption, and in fact these are type classes as well (which is why we declare them with \lstinline|:>|).
-It has been our experience that almost any generic property worth naming is worth representing as a type class, so that its proofs are be resolved as instances behind the scenes whenever possible.
+The reader will note that we use names for properties like distributivity and absorption.
+Indeed, these are type classes as well (which is why we declare them with \lstinline|:>|).
+It has been our experience that almost any generic property worth naming is worth representing as a type class, so that its proofs will be resolved as instances behind the scenes whenever possible.
 Doing this consistently minimizes administrative noise in the code, bringing us closer to ordinary mathematical vernacular.
@@
 
-We feel that these issues are not limitations of the Coq system, but merely reflect the fact that notationally identifying these two different and interdependent maps is a typical example of an ``abus de notation'' that by its very nature is ill-suited to a formal development where software engineering concerns apply.
+We feel that these issues are not limitations of the Coq system, but merely reflect the fact that notationally identifying these two different and interdependent maps is an abuse of notation that by its very nature is ill-suited to a formal development where software engineering concerns apply.
 Hence, we do not adopt this practice, and use ``\lstinline|fmap F|'' (name taken from the Haskell standard library) to refer to the arrow map of a functor \lstinline|F|.

### 2010-06-18 jar.tex
Notes on what to put in
Adopt some chktex suggestions and mention 'substructure' term.

@@
 \paragraph{Outline}
-In section \ref{preliminaries} we briefly describe the Coq system and some of its facilities that are of particular interest to us (most notably its implementation of type classes).
-Then, in section \ref{bundling}, we give a very concrete introduction to the issue of \emph{bundling}, arguably the biggest design dimension when building
-interfaces for abstract structures.
-In section \ref{predicateclasses} we show how type classes can make the use of ``unbundled'' purely predicate based interfaces for abstract structures practical.
-Next, in section \ref{hierarchy}, we discuss our algebraic hierarchy implemented using such predicate classes.
+In section~\ref{preliminaries} we briefly describe the Coq system and some of its facilities that are of particular interest to us (most notably its implementation of type classes).
+Then, in section~\ref{bundling}, we give a very concrete introduction to the issue of \emph{bundling}, arguably the biggest design dimension when building interfaces for abstract structures.
+In section~\ref{predicateclasses} we show how type classes can make the use of ``unbundled'' purely predicate based interfaces for abstract structures practical.
+Next, in section~\ref{hierarchy}, we discuss our algebraic hierarchy implemented using such predicate classes.
 
@@
 With this in place, we continue with a discussion of
-
-%to the
 
@@
 The Coq implementation of these mechanisms was largely rewritten in 200?
-by Matthieu Sozeau \ref{rewriting}.
+by Matthieu Sozeau~\ref{rewriting}.
 (todo: say that it's now much better than before)
@@
 
+% BS: put back
 % The Coq type theory lacks quotient
@@
 We will tackle the problems associated with the structure predicates approach one by one, starting with those encountered during theorem \emph{application}.
-Suppose we have defined \lstinline|SemiGroup| as a structure predicate as follows \footnote{Note that defining \lstinline|SemiGroup| as a record instead of a straight conjunction does not make it any less of a predicate.
+Suppose we have defined \lstinline|SemiGroup| as a structure predicate as follows\footnote{Note that defining \lstinline|SemiGroup| as a record instead of a straight conjunction does not make it any less of a predicate.
 The record form is simply more convenient in that it immediately gives us named projections for laws and substructures.}:
@@
 Fields declared with \lstinline|:>| are registered as hints for instance resolution, so that in any context where \lstinline|(A, =, +, 0, *, 1)| is known to be a \lstinline|SemiRing|, \lstinline|(A, =, +, 0)| and \lstinline|(A, =, *, 1)| are automatically known to be \lstinline|CommutativeMonoid|s (and so on, transitively, because instance resolution is recursive).
-In our hierarchy, these fields by themselves establish the inheritance diagram in Figure \ref{inheritancediagram}.
+In our hierarchy, these ``substructures'' by themselves establish the inheritance diagram in Figure~\ref{inheritancediagram}.
 % todo: for some reason the figure number reference is wrong here..
@@
 
-We will say more about this unfortunate limitation of the current type class implementation in section \ref{wishlist}.
+We will say more about this unfortunate limitation of the current type class implementation in section~\ref{wishlist}.
 

### 2010-06-22 jar.tex
Another snapshot.

@@
   {λ}{{$\uplambda\ $}}1
+  {σ}{{$\upsigma\ $}}1
   {∃}{{$\exists$}}1
@@
   {<-}{{$\leftarrow\ $}}1
+  {~}{{$\sim\ $}}1
 }
@@
 Indeed, these are type classes as well (which is why we declare them with \lstinline|:>|).
-It has been our experience that almost any generic property worth naming is worth representing as a type class, so that its proofs will be resolved as instances behind the scenes whenever possible.
+It has been our experience that almost any generic predicate worth naming is worth representing as a predicate type class, so that its proofs will be resolved as instances behind the scenes whenever possible.
 Doing this consistently minimizes administrative noise in the code, bringing us closer to ordinary mathematical vernacular.
-Indeed, we believe that type classes are an elegant and apt formalization of the seemingly casual manner in which ordinary mathematical presentation assumes implicit administration and use of a ``database'' of structural properties previously proved, much more so than existing solutions using, for instance, canonical structures.
-
-% Unfortunately, aggressive . More on taht in section bla
+Indeed, we believe that type classes are an elegant and apt formalization of the seemingly casual manner in which ordinary mathematical presentation assumes implicit administration and use of a ``database'' of properties previously proved.
+Much more so than the existing canonical structures facility, because canonical structures can only be used with bundled representations, which we argued against in section~\ref{bundling}.
+
+The hierarchy of predicate classes for the abstract structures themselves is mirrored by a hierarchy of predicate classes for morphisms. For instance:
+\begin{lstlisting}
+  Context `{Monoid A} `{Monoid B}.
+
+  Class Monoid_Morphism (f: A → B) :=
+    { monmor_from: Monoid A
+    ; monmor_to: Monoid B
+    ; monmor_sgmor:> SemiGroup_Morphism f
+    ; preserves_mon_unit: f mon_unit = mon_unit }.
+\end{lstlisting}
+
+Some clarification is in order to explain the role of the \lstinline|Context| declaration of the two monoids.
+While it may look we are parameterizing \lstinline|Monoid_Morphism| on monoid-ness proofs (which would be a gross violation of our idiom), in fact it is only parameterized on the monoid \emph{components} declared through implicit generalization of the \lstinline|Monoid| declarations, because it only refers to those.
+Here, we use declarations of predicate class parameters merely as convenient shorthands to declare their components.
+
+Notice that \lstinline|f| is \emph{not} made into an operational type class.
+The reason for this is that the role of \lstinline|f| is analogous to the carrier type in the previous predicate class definitions, in that it serves as the primary identification for the structure, and should therefore not be inferred.
+
+The \lstinline|monmor_to| and \lstinline|monmor_from| fields are not an absolute necessity, but eliminate the need for theorems to declare \lstinline|Monoid| parameters when they already declare \lstinline|Monoid_Morphism| parameters.
+This is an instance where we can freely posit structural properties without worrying about potential problems when such information turns out to be redundant in contexts where the source and target of the morphism are already known to be \lstinline|Monoid|s.
+
+Unfortunately, there is actually one annoying wrinkle here, which will also explain why, as the observant reader will have noticed, we do not register these two fields as instance resolution hints (by declaring them with \lstinline|:>|).
+What we really want these fields to express is ``\emph{if} in a certain context we know something to be a \lstinline|Monoid_Morphism|, \emph{then} realize that the source and target are \lstinline|Monoid|s''.
+However, the current instance resolution implementation does not support this style of \emph{forward} reasoning, and only supports \emph{backward} reasoning: had we registered \lstinline|monmor_to| and \lstinline|monmor_from| as instance resolution hints, we would in fact be saying ``\emph{if} trying to establish that something is a \lstinline|Monoid|, \emph{then} try finding a \lstinline|Monoid_Morphism| to or from it'', which quickly degenerates into a wild goose chase.
+We will return to this point in section~\ref{conclusion}.
+% todo: do this
+
+% : with predicate classes we can essentially pose queries of the form ``do I already know that proposition P holds?'', whereas canonical structures
+% ``do I already have a composite structure X of which Y is a specific component?
+% 
+% arbitrary predicates, based on associating particular
 
@@
 This definition is based on the 2-categorical idea of having equality only on arrows, not on objects.
-Similarly, we will have equality on natural transformations, but not on functors. % todo: ORLY?
+%Similarly, we will have equality on natural transformations, but not on functors. % todo: ORLY?
 
 Initiality, too, is defined by a combination of an operational and a predicate class:
-
 \begin{lstlisting}
   Context `{Category X}.
-  Class InitialArrow (x: X): Type := initial_arrow: Π y, x ⟶ y.
-  Class Initial (x: X) `{InitialArrow x}: Prop :=
+  Class InitialArrows (x: X): Type := initial_arrow: Π y, x ⟶ y.
+  Class Initial (x: X) `{InitialArrows x}: Prop :=
     initial_arrow_unique: Π y (a: x ⟶ y), a = initial_arrow y.
 \end{lstlisting}
-
-Having \lstinline|InitialArrow| as a type class means that we can always simply say \lstinline|initial_arrow y| whenever \lstinline|y| is known to be an object in a category known to have an initial object (where ``known to'' refers to the instances available to instance resolution).
+The operational class \lstinline|InitialArrows| designates the arrows that originate from an initial object \lstinline|x| by virtue of it being initial.
+The \lstinline|Initial| class itself further requires these ``initial arrows'' to be unique.
+Having \lstinline|InitialArrows| as a type class means that we can always simply say \lstinline|initial_arrow y| whenever \lstinline|y| is known to be an object in a category known to have an initial object (where ``known to'' should be read as ``can be determined by instance resolution'').
 
@@
     ; preserves_id: `(fmap (cat_id: a ⟶ a) = cat_id)
-    ; preserves_comp `(f: y ⟶ z) `(g: x ⟶ y): fmap (f ◎ g) = fmap f ◎ fmap g }.
-\end{lstlisting}
-Some clarification is in order to explain the role of the \lstinline|Context| declaration of the two categories.
-While it may look we are parameterizing both \lstinline|Fmap| and \lstinline|Functor| on proofs (which would be a gross violation of our idiom), in fact they depend and are parameterized only on the \emph{components} declared through implicit generalization of the \lstinline|Category| declarations.
-Indeed, we use the latter strictly as a convenient way to declare the former.
-
-Notice that \lstinline|map_obj| is \emph{not} made into an operational type class.
-The reason for this is that \lstinline|map_obj| is analogous to the carrier type in earlier structure predicate class definitions, in that it serves as the primary identification for the structure, and should therefore not be inferred.
-
-The \lstinline|functor_to| and \lstinline|functor_from| fields in \lstinline|Functor| are not an absolute necessity, but eliminate the need for theorems to declare \lstinline|Category| parameters when they already declare \lstinline|Functor|s between them.
-This is an instance where we can freely posit structural properties without worrying about potential problems when such information turns out to be redundant in contexts where the source and target of the functor are already known to be \lstinline|Category|'s.
-
-Unfortunately, there is actually one annoying wrinkle here, which will also explain why, as the observant reader will have noticed, we do not register them as instance resolution hints (by declaring them with \lstinline|:>|).
-
-What we really want these fields to express is ``\emph{if} in a certain context we know something to be a functor, \emph{then} realize that the source and target are categories''.
-However, the current instance resolution implementation does not support this style of exclusively-forward implications, and only supports backward implications: had we registered \lstinline|functor_to| and \lstinline|functor_from| as instance resolution hints, we would in fact be saying ``\emph{if} trying to establish that something is a category, \emph{then} you might try finding a functor to or from it'', which quickly degenerates into a wild goose chase.
-
-We will say more about this unfortunate limitation of the current type class implementation in section~\ref{wishlist}.
-
-Finally, we ought to say a few words about our use of \lstinline|fmap|.
+    ; preserves_comp `(f: y ⟶ z) `(g: x ⟶ y):
+      fmap (f ◎ g) = fmap f ◎ fmap g }.
+\end{lstlisting}
+We ought to say a few words about our use of \lstinline|fmap|.
 The usual mathematical notational convention for functor application is to use the name of the functor to refer to both its object map and its arrow map, relying on additional conventions regarding object/arrow names for disambiguation (i.e.
@@
 
-We feel that these issues are not limitations of the Coq system, but merely reflect the fact that notationally identifying these two different and interdependent maps is an abuse of notation that by its very nature is ill-suited to a formal development where software engineering concerns apply.
+We feel that these issues are not limitations of the Coq system, but merely reflect the fact that notationally identifying these two distinct and interdependent maps is an abuse of notation of sufficient severety to make it ill-suited to a formal development where software engineering concerns apply.
 Hence, we do not adopt this practice, and use ``\lstinline|fmap F|'' (name taken from the Haskell standard library) to refer to the arrow map of a functor \lstinline|F|.
@@
 \subsection{Natural transformations}
-blabla, routine.
-
-%theory/categories.v
 Natural transformations are formalized as:
@@
    Context `{Category C} `{Category D}
-   `{!Functor (F: C ⟶ D) Fa} `{!Functor (G: C ⟶ D) Ga}.
+   `{!Functor (F: C → D) Fa} `{!Functor (G: C → D) Ga}.
    Class NaturalTransformation (η: Π c, F c ⟶ G c): Prop :=
@@
 
-Here \lstinline|fmap F| explicitly refers to the arrow part of the functor \lstinline|F|.
-In mathematics one would apply $F$ to both objects and arrows ($F x$, $F f$). However, this does
-not seem to fit with type theory as the type of \lstinline|fmap F| depends on the object part of
-\lstinline|F|.
-
 We have defined the category of setoids, the dual of a category, products, and hence co-products. 
@@
 \subsection{Adjunctions}
-blabla, example of nontrivial theorem. nice stress-test, worked out mostly fine.
+
+Adjunctions can be defined in different ways. A nice symmetric definition is the following:
+
+\begin{lstlisting}
+  Context `{Category A} `{Category X}
+    F `{!Functor (F: X → A) F'}
+    G `{!Functor (G: A → X) G'}
+    (φ: Π {x a}, (F x ⟶ a) → (x ⟶ G a))
+    `{Π x a, Inverse (φ x a)}
+    `{Π x a, Bijective (φ x a)}.
+
+  Class Adjunction: Prop :=
+   { adjunction_left_functor: Functor F _
+   ; adjunction_right_functor: Functor G _
+   ; natural_left `(k: a ⟶ a'): `((fmap G k ◎) ∘ φ = @φ x _ ∘ (k ◎))
+   ; natural_right `(h: x' ⟶ x):`((◎ h) ∘ @φ _ a = φ ∘ (◎ fmap F h)) }.
+\end{lstlisting}
+
+blabla, alternative definition, equivalence theorem nice stress test, preliminary results promising except for rewriting/instance-resolution efficiency.
 
@@
 
-Motivated originally by our desire to cleanly express interfaces for basic numeric data types such as $\N$ and $\Z$ in terms of their categorical characterization as initial objects in the categories of semirings and rings, respectively, we initially introduced only the very basics of category theory into our development, again using type classes where possible to achieve the same benefits mentioned before.
-
-Realizing that much code duplication for the various algebraic structures in the hierarchy could be avoided by employing universal algebra constructions, we then proceeded to formalize some of the theory of multisorted universal algebra and equational theories, using it to automatically construct varieties of algebras.
-We avoided existing formalizations~\cite{DBLP:conf/tphol/Capretta99,dominguez2008formalizing} of universal algebra, because we aimed to find out what level of elegance, convenience, and integration can be achieved using the state of the art technology (of which type classes are the most important instance).
-
-At the time of writing, our development includes a fully integrated formalization of a nontrivial portion of category theory and multisorted universal algebra, including various categories (e.g.\ the category \lstinline|Cat| of categories, and finitary algebraic categories  defined by a theory which we instantiate to obtain the categories of monoids, semirings, and rings), functors (including automatically generated forgetful functors), natural transformations, adjunctions, initial models of equational theories constructed from term algebras, transference of proofs between isomorphic models of equational theories, subalgebras, congruences, quotients, products, and the first homomorphism theorem~\cite{meinke1993universal}.
-
-We consider the first homomorphism theorem in more detail:
+To specify the natural numbers and the integers as initial objects in the categories of semirings and rings, respectively, definitions of these categories are needed.
+While one could define both of them ``manually'', greater economy can be achieved by recognizing that both semirings and rings can be defined by equational theories, \emph{varieties} (that is, categories consisting of models for a fixed theory, with homomorphisms between them) of which can be defined generically.
+
+To this end, we have formalized some of the theory of multisorted universal algebra and equational theories.
+We chose not to revive existing old (and somewhat stale) formalizations ~\cite{DBLP:conf/tphol/Capretta99,dominguez2008formalizing} of universal algebra, because an important aim for us has been to find out what level of elegance, convenience, and integration can be achieved by leveraging the state of the art in Coq facilities (of which type classes are the most important example).
+
+\subsection{Signatures and algebras}
+
+A multisorted signature enumerates sorts, operations, and specifies the ``types'' of the operations as non-empty lists of sorts, where the final element denotes the result type:
+\begin{lstlisting}
+  Inductive Signature: Type :=
+    { sorts: Set
+    ; operation:> Set
+    ; operation_type:> operation → ne_list sorts }.
+\end{lstlisting}
+Given an interpretation of the sorts (mapping each symbolic sort to a carrier), an interpretation for the operations is easily written as an operational type class:
+\begin{lstlisting}
+  Variables (σ: Signature) (carriers: sorts σ → Type).
+
+  Definition op_type: ne_list (sorts σ) → Type := fold (→) ∘ map carrier.
+
+  Class AlgebraOps (σ: Signature) (carriers: sorts σ → Type)
+    := algebra_op: Π o: operation σ, op_type carriers (σ o).
+\end{lstlisting}
+Because our carriers will normally be equipped with a setoid equality, we further define the predicate class \lstinline|Algebra|, stating that each of the operations respects the setoid equality on the carriers:
+\begin{lstlisting}
+  Class Algebra `{Π a, Equiv (carriers a)} `{AlgebraOps σ carriers}: Prop :=
+      { algebra_setoids:> Π a, Setoid (carriers a)
+      ; algebra_propers:> Π o: σ, Proper (=) (algebra_op o) }.
+\end{lstlisting}
+The \lstinline|(=)| referred to in \lstinline|algebra_propers| is an |Equiv| instance expressing setoid-respecting extensionality for the function types generated by \lstinline|op_type|.
+
+We do not unbundle \lstinline|Signature| because it represents a triple that will always be specifically constructed for subsequent use with the universal algebra facilities.
+We have no ambition to recognize signature triples ``in the wild'', nor will we ever talk about multiple signatures sharing sort- or operation enumerations.
+
+\subsection{Equational theories and varieties}
+\label{varieties}
+
+Algebras are insufficient to characterize structures such as semirings and rings.
+For that, we need equational theories, consisting of a signature together with laws (represented by a predicate over equality entailments):
+\begin{lstlisting}
+Record EquationalTheory :=
+  { et_sig:> Signature
+  ; et_laws:> EqEntailment et_sig → Prop }.
+\end{lstlisting}
+An \lstinline|EqEntailment| consists of premises and a conclusion represented by an inductively defined statement grammar, which in turn uses an inductively defined term grammar.
+A detailed discussion of these definitions and the theory developed for them is beyond the scope of this paper,
+
+We now introduce a predicate class designating algebras that satisfy the laws of an equational theory:
+\begin{lstlisting}
+Class InVariety
+  (et: EquationalTheory) (carriers: sorts et → Type)
+  {e: Π a, Equiv (carriers a)} `{!AlgebraOps et carriers}: Prop :=
+  { variety_algebra:> Algebra et carriers
+  ; variety_laws: Π s, et_laws et s → (Π vars, eval_stmt et vars s) }.
+\end{lstlisting}
+
+What remains is to show that carriers+\lstinline|Equivs|+\lstinline|AlgebraOps| satisfying \lstinline|InVariety| for a given \lstinline|EquationalTheory| do indeed form a \lstinline|Category| (the ``variety'').
+Since we need a type for the objects in the |Category|, at this point we have no choice but to bundle components and proof together in a record:
+\begin{lstlisting}
+  Variable et: EquationalTheory.
+
+  Record ObjectInVariety: Type := object_in_variety
+    { variety_carriers:> sorts et → Type
+    ; variety_equiv: Π a, Equiv (variety_carriers a)
+    ; variety_op: AlgebraOps et variety_carriers
+    ; variety_proof: InVariety et variety_carriers }.
+\end{lstlisting}
+The arrows will be homomorphisms, which are also defined generically for any equational theory:
+
+\begin{lstlisting}
+  Instance: Arrows Object := λ X Y: Object => sig (HomoMorphism et X Y).
+\end{lstlisting}
+
+The instance definitions for identity arrows, arrow composition, arrow setoid equality, and composition propriety, are all trivial, as is the final \lstinline|Category| instance:
+\begin{lstlisting}
+  Instance: Category ObjectInVariety.
+\end{lstlisting}
+
+In addition to this ``variety'' category, we also have categories of lawless algebras, as well as forgetful functors from the former to the latter, and from the latter to the category of setoids.
+
+\subsection{The first homomorphism theorem}
+
+To give a further taste of what universal algebra in our development looks like, we consider the definitions involved in the first homomorphism theorem in more detail:
 \begin{theorem}[First homomorphism theorem]
@@
 
-% theory/ua_subalgebraT.v
-% find . -iname "*.v" |xargs grep "niverse"
-We define congruence:
-\begin{lstlisting}
-Class Congruence: Prop :=
+A set of relations \lstinline|e| (one for each sort) is a congruence for an existing algebra if (1) \lstinline|e| respects that algebra's existing setoid equality, and (2) the operations with \lstinline|e| again form an algebra (namely the quotient algebra):
+\begin{lstlisting}
+Context `{Algebra σ A}.
+Class Congruence (e: Π s: sorts σ, relation (v s)): Prop :=
     { congruence_proper:> Π s, Proper (equiv ==> equiv ==> iff) (e s)
-    ; congruence_quotient:> @Algebra et v e _
-    }.
-\end{lstlisting}
-
-In set theory, a quotient is defined as a collection of equivalence classes. In type theory, a quotient is more naturally
-defined by changing the equality, while leaving the carrier unchanged.%
-\footnote{In Bishop's words~\cite[p.12]{Bishop/Bridges:1985}: The axiom of choice is used to extract
-elements from equivalence classes where they should never have been put in the first place.}
- In this way we profit from the infrastructure for setoids: A congruence relation is coarser than the setoid relation, as indicated by the use of \lstinline|Proper| above.
-
-Alternatively, we could have stated the compatibility of the relation \lstinline|e| by stating that the
-relation as a set of pairs is a subalgebra in the product algebra. We have proved that both approaches
-are equivalent. As said, the present definition seems natural in type theory.
-
-Coq provides different sorts. To simplify the discussion, we assume that there are only two sorts:
-\lstinline|Prop|, for propositions, and \lstinline|Type|, for data types.
-A subalgebra is defined by a predicate on the algebra, that is a map \lstinline|A->Prop|.
-However, the image $Im f:=\{b:B \mid \exists a.
-b=(f a)\}$ is informative (in \lstinline|Type|) as it includes a witness for the existential.
-This witness is needed to map $b\in Im f$ to one of its pre-images in $A$.
-We thus need two flavors of subalgebras, one \lstinline|Prop|-valued an one \lstinline|Type|-valued; see also~\cite{coquand-towards}.
-Coq provides universe polymorphism for inductive types to avoid code duplication in such cases.
-Unfortunately, definitions and fixpoints are not (yet?) polymorphic.
-Hence we had to resort to changing some Fixpoints into Inductives.
-
-Having discussed the first homomorphism theorem we now turn to category theory which helps to organize our development of universal algebra.
+    ; congruence_quotient:> @Algebra et v e _ }.
+\end{lstlisting}
+We have proved that this natural and economical type-theoretic formulation that leverages our systematic integration of setoid equality is equivalent to the traditional definition of congruences as relations that, represented as sets of pairs, form a subalgebra of the product algebra.
+
+For the homomorphism theorem, we begin by declaring our dramatis personae:
+\begin{lstlisting}
+  Context `{Algebra σ A} `{Algebra σ B} `{!HomoMorphism σ A B f}.
+\end{lstlisting}
+With \lstinline|~| defined as described, the first part of the proof is simply the definition of the following instance:
+\begin{lstlisting}
+   Instance co: Congruence σ (~).
+\end{lstlisting}
+
+For the second part, we describe the image of \lstinline|f| as a predicate over \lstinline|B|, and show that it is closed under the operations of the algebra: 
+\begin{lstlisting}
+  Definition image s (b: B s): Type := sigT (λ a => f s a = b).
+
+  Instance: ClosedSubset image.
+\end{lstlisting}
+\lstinline|sigT| represents a \lstinline|Type|-sorted existential. \lstinline|ClosedSubset| is defined elsewhere as:
+\begin{lstlisting}
+  Context `{Algebra σ A} (P: Π s, A s → Type).
+  Class ClosedSubset: Type :=
+    { subset_proper: Π s x x', x = x' → iffT (P s x) (P s x')
+    ; subset_closed: Π o, op_closed (algebra_op o) }.
+\end{lstlisting}
+\lstinline|op_closed| is defined by recursion over the symbolic operation types.
+
+The reason we define \lstinline|image| and \lstinline|ClosedSubset| in \lstinline|Type| rather than \lstinline|Prop| is that since the final goal of the proof is to establish an isomorphism in the category of σ-algebras (where arrows are algebra homomorphisms), we will eventually need to map elements in the subalgebra defined by \lstinline|image| back to their pre-image in \lstinline|A|.
+
+However, there are contexts (in other proofs) where \lstinline|Prop|-sorted construction of subalgebras really \emph{is} appropriate.
+Unfortunately, Coq's universe polymorphism is not yet up to the task of letting us use a single set of definitions to handle both cases.
+In particular, there is no universe polymorphism for definitions (as opposed to inductive definitions) yet.
+We will return to this point later.
+In our development, we have two sets of definitions, one for \lstinline|Prop| and one for \lstinline|Type|, resulting in about a hundred lines of duplication.
+
+For the main theorem, we now bundle the quotient algebra and the subalgebra into records akin to \lstinline|ObjectInVariety| from section~\ref{varieties}:
+\begin{lstlisting}
+  Definition quot_obj: algebra.Object σ := algebra.object σ A (algebra_equiv:=(~)).
+  Definition subobject: algebra.Object σ := algebra.object σ (ua_subalgebraT.carrier image).
+\end{lstlisting}
+Here, \lstinline|algebra| is the module defining the bundled algebra record \lstinline|Object| with constructor \lstinline|object|.
+\lstinline|ua_subalgebraT| is the module for constructing subalgebras.
+
+Finally, we define a pair of arrows between the two and show that these arrows form an isomorphism:
+\begin{lstlisting}
+   Program Definition back: subobject ⟶ quot_obj
+    := λ _ X => projT1 (projT2 X).
+
+  Program Definition forth: quot_obj ⟶ subobject
+    := λ a X => existT _ (f a X) (existT _ X (reflexivity _)).
+
+  Theorem first_iso: iso_arrows back forth.
+\end{lstlisting}
+The \lstinline|Program| command generates proof obligations (not shown) expressing that these two arrows are indeed homomorphisms. The theorem itself is trivial.
+
+% \footnote{In Bishop's words~\cite[p.12]{Bishop/Bridges:1985}: The axiom of choice is used to extract
+%elements from equivalence classes where they should never have been put in the first place.}
+
+\section{Numerical interfaces}
+
+\lstinline|EquationalTheory|'s for semirings and rings are easy to define, and so from section~\ref{varieties} we get corresponding categories in which we can postulate initial objects:
+\begin{lstlisting}
+  Class Naturals (A: ObjectInVariety semiring_theory) `{!InitialArrow A}: Prop :=
+    { naturals_initial:> Initial A }.
+\end{lstlisting}
+While succinct, this definition is not a satisfactory abstraction because the use of \lstinline|ObjectInVariety| for the type of the \lstinline|A| component ``leaks'' the fact that we used this one particular universal algebraic construction of the category, which is just an implementation choice.
+Furthermore, this definition needs an additional layer of class instances to relate it to the \lstinline|SemiRing| class from our algebraic hierarchy.
+
+What we \emph{really} want to say is that an implementation of the natural numbers ought to be an a-priori \lstinline|SemiRing| that, when \emph{bundled} into an \lstinline|ObjectInVariety semiring_theory|, is initial in said category.
+This is a typical example where conversion functions between concrete classes such as \lstinline|SemiRing| and instantiations of more abstract classes such as \lstinline|InVariety| and \lstinline|Category| are required in our development in order to leverage and apply concepts and theory defined for the latter to the former.
+While sometimes a source of some tension in that these conversions are not yet applied completely transparently whenever needed, the ability to move between ``down to earth'' and ``high in the sky'' perspectives on the same abstract structures has proved invaluable in our development, and we will give more examples of this in a moment.
+
+Taking these conversion functions for granted, we will also need a ``down-to-earth'' representation of the initiality arrows if we are to give a \lstinline|SemiRing|-based definition of the interface for natural numbers.
+Once again, we introduce an operational type class to represent this particular component:
+\begin{lstlisting}
+  Class NaturalsToSemiRing (A: Type) :=
+    naturals_to_semiring: Π B `{RingMult B} `{RingPlus B} `{RingOne B} `{RingZero B}, A → B.
+\end{lstlisting}
+In order to use this with \lstinline|Initial|, we define an additional conversion instance that takes a \lstinline|NaturalsToSemiRing| along with a proof showing that it yields \lstinline|SemiRing_Morphism|s, and builds an \lstinline|InitialArrow| instance out of it invoking another conversion function that translates concrete \lstinline|SemiRing_Morphism| proofs into univeral algebra \lstinline|Homomorphism|s instantiated with the semiring signature, which make up the arrows in the category.
+
+With these instances in place, we can now define the improved natural numbers specification:
+\begin{lstlisting}
+  Class Naturals A {e plus mult zero one} `{!NaturalsToSemiRing A} :=
+    { naturals_ring:> @SemiRing A e plus mult zero one
+    ; naturals_to_semiring_mor:> Π `{SemiRing B}, SemiRing_Morphism (naturals_to_semiring A B)
+    ; naturals_initial:> Initial (bundle_semiring A) }.
+\end{lstlisting}
+Basing theory and programs on this abstract interface instead of on specific implementation (such as the ubiquitous Peano naturals \lstinline|nat| in the Coq standard library) is not only cleaner mathematically, but also facilitates easy swapping between implementations.
+And this benefit is far from theoretical, as diverse and non-contrived representations of the natural numbers are abound (think unary, binary, factor multisets,
+arrays of native machine words, etc.).
+
+Some theory and programs about naturals are more easily (or can only be) proved (resp.
+defined) for concrete implementations and then ``lifted'' to the abstract interface.
+For relatively simple things like decidability of equality, all that is needed to make this lifting easy is a handful of simple lemmas translating the initiality property back into statements saying that \lstinline|naturals_to_semiring| is extensionally equal to other semiring morphisms and involutive between \lstinline|Naturals| implementations.
+For the example of decidability, these let us define decidability for an arbitrary \lstinline|Naturals| implementation simply by mapping to the \lstinline|nat| implementation using \lstinline|naturals_to_semiring|, deciding equality there, and reflecting the equality or inequality proof back to the abstract implementation using involutivity of \lstinline|naturals_to_semiring|.
+
+For more complex properties such as injectivity of partially applied addition and multiplication, we take a longer detour.
+As part of our universal algebra theory, we have proved that proofs of statements in the language of an equational theory can be transferred between isomorphic implementations.
+Combined with the trivially proved property that any two initial objects in a category are isomorphic, this lets us tranfer proofs of such statements between implementations of \lstinline|Naturals|, requiring only that we reflect the concrete statement (expressed in terms of the operational type classes) to a symbolic representation in the equational theory of semirings.
+We intend to eventually make this reflection completely automatic using type class based quotation techniques along the lines of those described in~\ref{quoting}.
+
+Thanks to our close integration of universal algebra we can actually obtain a \lstinline|Naturals| implementation completely automatically, by invoking a generic construction of initial models built from the closed term algebra for the signature along with a setoid equality expressing the congruence closure of the identities in the equational theory.
+However, this implementation is very useful neither in terms of efficiency, nor as a canonical implementation (to be used as the basis for theory and programs that are then subsequently lifted), because for example deciding equality for the aforementioned setoid equality is far harder than for, say, \lstinline|nat|, because it requires a normalization procedure.
+
+\subsection{Specialization}
+
+The generic \lstinline|Decision| instance for \lstinline|Naturals| equality implemented by mapping to \lstinline|nat| will typically be far less efficient than a specialized implementation for a particular representation of the natural numbers.
+Fortunately, with Coq's type classes it is no problem for instances overlapping in this way to co-exist, and we can even deprioritize the generic instance so that instance resolution will always pick the specialization when the representation is known.
+
+To permit a \emph{generic} function operating on naturals to take advantage of specialized operations, we simply introduce an additional instance parameter:
+\begin{lstlisting}
+Definition calculate_things `{Naturals N} `{!forall n m: N, Decision (n = m)}
+  (a b: nat): ... := ... decide (a = b) ... .
+\end{lstlisting}
+Without the \lstinline|Decision| parameter \lstinline|calculate_things| would be equally correct, but could be less efficient.
+Thus, by this scheme one can start by writing correct-but-possibly-inefficient programs that make use of generic operation instances, and then selectively improve efficiency of key algorithms simply by adding additional operational type class instance parameters where profiling shows it to make a significant difference, without changing their definition body.
+
+Other operations on natural numbers that are sensible choices for specialization are subtraction, distance, division/multiplication by 2, etc.
+
+\subsection{Integers and rationals}
+
+% The concrete implementation we use are just the Peano naturals \lstinline|nat|.
+Showing that \lstinline|nat| with its arithmetic operations and Leibniz equality form a \lstinline|SemiRing| is entirely routine, as is defining a map to an arbitrary other semiring and showing that this map is unique as a \lstinline|SemiRing_Morphism|, so that we get initiality.
+
+%integers all the same:
+
+% At the time of writing, our development includes a fully integrated formalization of a nontrivial portion of category theory and multisorted universal algebra, including various categories (e.g.\ the category \lstinline|Cat| of categories, and finitary algebraic categories  defined by a theory which we instantiate to obtain the categories of monoids, semirings, and rings), functors (including automatically generated forgetful functors), natural transformations, adjunctions, initial models of equational theories constructed from term algebras, transference of proofs between isomorphic models of equational theories, subalgebras, congruences, quotients, products, and the first homomorphism theorem~\cite{meinke1993universal}.
+
+EELIS UP TO HERE
 
@@
 We will use the resulting expression monad in Section~\ref{quote} below.
-
-One may hope that this term algebra could be useful to \emph{define} inductive data types, say, to define the natural numbers 
-as the term algebra for the theory of semirings. However, it seems difficult to prove decidability of equality, 
-as this requires a normalization procedure.
 

### 2010-06-22 jar.tex
small changes

@@
 To this end, we have formalized some of the theory of multisorted universal algebra and equational theories.
-We chose not to revive existing old (and somewhat stale) formalizations ~\cite{DBLP:conf/tphol/Capretta99,dominguez2008formalizing} of universal algebra, because an important aim for us has been to find out what level of elegance, convenience, and integration can be achieved by leveraging the state of the art in Coq facilities (of which type classes are the most important example).
+We chose not to revive existing old (and somewhat stale) formalizations~\cite{DBLP:conf/tphol/Capretta99,dominguez2008formalizing} of universal algebra, because an important aim for us has been to find out what level of elegance, convenience, and integration can be achieved by leveraging the state of the art in Coq facilities (of which type classes are the most important example).
 
@@
 \end{lstlisting}
-The \lstinline|(=)| referred to in \lstinline|algebra_propers| is an |Equiv| instance expressing setoid-respecting extensionality for the function types generated by \lstinline|op_type|.
+The \lstinline|(=)| referred to in \lstinline|algebra_propers| is an \lstinline|Equiv| instance expressing setoid-respecting extensionality for the function types generated by \lstinline|op_type|.
 
@@
 
-Algebras are insufficient to characterize structures such as semirings and rings.
-For that, we need equational theories, consisting of a signature together with laws (represented by a predicate over equality entailments):
+To characterize structures such as semirings and rings, we need equational theories, consisting of a signature together with laws (represented by a predicate over equality entailments):
 \begin{lstlisting}
@@
 What remains is to show that carriers+\lstinline|Equivs|+\lstinline|AlgebraOps| satisfying \lstinline|InVariety| for a given \lstinline|EquationalTheory| do indeed form a \lstinline|Category| (the ``variety'').
-Since we need a type for the objects in the |Category|, at this point we have no choice but to bundle components and proof together in a record:
+Since we need a type for the objects in the \lstinline|Category|, at this point we have no choice but to bundle components and proof together in a record:
 \begin{lstlisting}
@@
 \begin{lstlisting}
-  Definition quot_obj: algebra.Object σ := algebra.object σ A (algebra_equiv:=(~)).
-  Definition subobject: algebra.Object σ := algebra.object σ (ua_subalgebraT.carrier image).
-\end{lstlisting}
-Here, \lstinline|algebra| is the module defining the bundled algebra record \lstinline|Object| with constructor \lstinline|object|.
-\lstinline|ua_subalgebraT| is the module for constructing subalgebras.
+  Definition quot_obj: algebra.Object σ :=
+    algebra.object σ A (algebra_equiv:=(~)).
+  Definition subobject: algebra.Object σ :=
+    algebra.object σ (ua_subalgebraT.carrier image).
+\end{lstlisting}
+Here, \lstinline|algebra| is the module defining the bundled algebra record \lstinline|Object| with constructor \lstinline|object|. The module \lstinline|ua_subalgebraT| constructs subalgebras.
 
@@
   Class NaturalsToSemiRing (A: Type) :=
-    naturals_to_semiring: Π B `{RingMult B} `{RingPlus B} `{RingOne B} `{RingZero B}, A → B.
+    naturals_to_semiring: Π B `{RingMult B} `{RingPlus B} `{RingOne B}
+      `{RingZero B}, A → B.
 \end{lstlisting}
@@
 Basing theory and programs on this abstract interface instead of on specific implementation (such as the ubiquitous Peano naturals \lstinline|nat| in the Coq standard library) is not only cleaner mathematically, but also facilitates easy swapping between implementations.
-And this benefit is far from theoretical, as diverse and non-contrived representations of the natural numbers are abound (think unary, binary, factor multisets,
-arrays of native machine words, etc.).
+And this benefit is far from theoretical, as diverse and non-contrived representations of the natural numbers are abound ( unary, binary, factor multisets, arrays of native machine words, etc.).
 

### 2010-06-22 jar.tex
Add subsection about implicit syntax-directed algorithm composition (to establish the Decision class).

--- 2010-06-22 jar.tex
@@
 
-\section{Predicate classes}
+\section{Predicate classes and operational classes}
 
@@
 Hence, we do not foresee problems with large contexts for any but the most complex formalizations.
+
+\subsection{Implicit syntax-directed algorithm composition}
+
+Before we proceed with a discussion of the algebraic hierarchy based on type classes, we briefly highlight one specific operational type class because we will use it later and because it is a particularly nice illustration of another neat application of operational type classes.
+The operation in question is that of deciding a proposition:
+\begin{lstlisting}
+  Class Decision (P: Prop): Type := decide: sumbool P (¬ P).
+\end{lstlisting}
+Here, \lstinline|sumbool| is just the (informative) sum of proofs.
+
+\lstinline|Decision| is a very general purpose type class that also works for predicates.
+For instance, to declare a parameter expressing decidability of, say, (setoid) equality on a type \lstinline|X|, we write: \lstinline|`{Π a b: X, Decision (a = b)}|.
+To then use this (unnamed) decider to decide a particular equality, we simply say \lstinline|decide (x = y)|, and instance resolution will resolve the decider we declared.
+
+With \lstinline|Decision| as a type class, we can very easily define composite deciders for things like conjunctions and quantifications over (finite) domains:
+\begin{lstlisting}
+  Instance decide_conj `{Decision P} `{Decision Q}: Decision (P ∧ Q).
+\end{lstlisting}
+With these in place, we can just say \lstinline|decide (x = y ∧ p = q)| and let instance resolution automatically compose a decision procedure that can decide the specified proposition.
+This style of syntax-directed implicit composition of algorithms is very convenient and highly expressive.
 
@@
 We will return to this point later.
-In our development, we have two sets of definitions, one for \lstinline|Prop| and one for \lstinline|Type|, resulting in about a hundred lines of duplication.
+In our development, we have two sets of definitions, one for \lstinline|Prop| and one for \lstinline|Type|, resulting in duplication of about a hundred lines of code.
 
@@
 \end{lstlisting}
-The \lstinline|Program| command generates proof obligations (not shown) expressing that these two arrows are indeed homomorphisms. The theorem itself is trivial.
+The \lstinline|Program| command generates proof obligations (not shown) expressing that these two arrows are indeed homomorphisms. The proof of the theorem itself is trivial.
 
@@
 Basing theory and programs on this abstract interface instead of on specific implementation (such as the ubiquitous Peano naturals \lstinline|nat| in the Coq standard library) is not only cleaner mathematically, but also facilitates easy swapping between implementations.
-And this benefit is far from theoretical, as diverse and non-contrived representations of the natural numbers are abound (think unary, binary, factor multisets,
-arrays of native machine words, etc.).
+And this benefit is far from theoretical, as diverse and non-contrived representations of the natural numbers are abound (think unary, binary, factor multisets, arrays of native machine words, etc.).
 
@@
 As part of our universal algebra theory, we have proved that proofs of statements in the language of an equational theory can be transferred between isomorphic implementations.
-Combined with the trivially proved property that any two initial objects in a category are isomorphic, this lets us tranfer proofs of such statements between implementations of \lstinline|Naturals|, requiring only that we reflect the concrete statement (expressed in terms of the operational type classes) to a symbolic representation in the equational theory of semirings.
+Combined with the trivially proved property that any two initial objects in a category are isomorphic, this lets us tranfer proofs of such statements between implementations of \lstinline|Naturals|, requiring only that we reflect the concrete statement (expressed in terms of the operational type classes) to a symbolic statement in the language of semirings.
 We intend to eventually make this reflection completely automatic using type class based quotation techniques along the lines of those described in~\ref{quoting}.
@@
 Thanks to our close integration of universal algebra we can actually obtain a \lstinline|Naturals| implementation completely automatically, by invoking a generic construction of initial models built from the closed term algebra for the signature along with a setoid equality expressing the congruence closure of the identities in the equational theory.
-However, this implementation is very useful neither in terms of efficiency, nor as a canonical implementation (to be used as the basis for theory and programs that are then subsequently lifted), because for example deciding equality for the aforementioned setoid equality is far harder than for, say, \lstinline|nat|, because it requires a normalization procedure.
+However, this implementation is very useful neither in terms of efficiency, nor as a canonical implementation (to be used as the basis for theory and programs that are then subsequently lifted), because for example deciding equality for the aforementioned setoid equality is far harder than for, say, \lstinline|nat| (because it requires a normalization procedure).
 
@@
 
+% todo: decision
 The generic \lstinline|Decision| instance for \lstinline|Naturals| equality implemented by mapping to \lstinline|nat| will typically be far less efficient than a specialized implementation for a particular representation of the natural numbers.
@@
 \begin{lstlisting}
-Definition calculate_things `{Naturals N} `{!forall n m: N, Decision (n = m)}
+Definition calculate_things `{Naturals N} `{!Π n m: N, Decision (n = m)}
   (a b: nat): ... := ... decide (a = b) ... .

### 2010-06-22 jar.tex

@@
 Basing theory and programs on this abstract interface instead of on specific implementation (such as the ubiquitous Peano naturals \lstinline|nat| in the Coq standard library) is not only cleaner mathematically, but also facilitates easy swapping between implementations.
-And this benefit is far from theoretical, as diverse and non-contrived representations of the natural numbers are abound (think unary, binary, factor multisets, arrays of native machine words, etc.).
+And this benefit is far from theoretical, as diverse and non-contrived representations of the natural numbers are abound (unary, binary, factor multisets, arrays of native machine words, etc.).
 

### 2010-06-22 jar.tex
Added formath
Add blurb about integers, rationals, and polynomials.

@@
 
-\subsection{Integers and rationals}
-
-% The concrete implementation we use are just the Peano naturals \lstinline|nat|.
-Showing that \lstinline|nat| with its arithmetic operations and Leibniz equality form a \lstinline|SemiRing| is entirely routine, as is defining a map to an arbitrary other semiring and showing that this map is unique as a \lstinline|SemiRing_Morphism|, so that we get initiality.
+\subsection{Integers, rationals, and polynomials}
+
+The abstract interface for integers is completely analogous to the one for natural numbers:
+\begin{lstlisting}
+  Class Integers A {e plus mult zero one opp} `{IntegersToRing A} :=
+    { integers_ring:> @Ring A e plus mult zero one opp
+    ; integers_to_ring_mor:> Π `{Ring B}, Ring_Morphism (integers_to_ring A B)
+    ; integers_initial:> Initial (ring.object A) }.
+\end{lstlisting}
+
+The rationals are characterized as a decidable field with an injective ring morphism from a canonical implementation of the integers and for which division of such integers is surjective:
+\begin{lstlisting}
+  Class Rationals A {e plus mult opp zero one mult_inv } `{Π x y: A, Decision (x = y)} {inj_inv}: Prop :=
+    { rationals_field:> @Field A e plus mult opp zero one mult_inv
+    ; rationals_frac: Surjective (λ p => integers_to_ring (Z nat) A (fst p) * / integers_to_ring (Z nat) A (snd p)) (inv:=inj_inv)
+    ; rationals_embed_ints: Injective (integers_to_ring (Z nat) A) }.
+\end{lstlisting}
+Here, \lstinline|Z| is an \lstinling|Integers| implementation paramerized by a \lstinline|Naturals| implementation, for which we just take \lstinline|nat|.
+The choice of \lstinline|Z nat| here is immaterial; we could have picked any other (or even an explicitly arbitrary) implementation of \lstinilne|Integers|, but doing so would provide no benefit.
+  % todo: hm, this Z is poorly named. it's really natpair_integers.Z
+  % todo: say a few words about how we handle division by zero?
+
+In our development we prove thas the standard library's default rationals do indeed implement \lstinline|Rationals|, as do implementations of the \lstinline|QType| module interface.
+While the latter is rather ad-hoc from a theoretical perspective, it is nevertheless of great practical interest because it is used for the very efficient \lstinline|BigQ| rationals based on machine integers~\cite{something}.
+Hence, theory and programs developed on our \lstinline|Rationals| interface applies and can make immediate use of these efficient rationals.
+As stated earlier, we plan to rebase the computable real number implementation in CoRN on this interface, precisely so that it may be instantiated with efficient implementations like these.
+
+We also plan to do the exact same thing for polynomials, which can be abstractly specified as a free commutative algebra, and for which diverse implementations exist (coefficient lists, Bernstein polynomials, etc.).
 
@@
 sharpen our understanding of the relation with Canonical Structures and with Unification Hints.
+
+This work has been partially funded by the FORMATH project, nr.\ 243847, of the FET program within the 7th Framework program of the European Commission.
 %Jeremy, Thierry

### 2010-06-22 jar.tex
Improve presentation of Coq definitions of numerical interfaces.

@@
 \begin{lstlisting}
-  Class Naturals A {e plus mult zero one} `{!NaturalsToSemiRing A} :=
-    { naturals_ring:> @SemiRing A e plus mult zero one
-    ; naturals_to_semiring_mor:> Π `{SemiRing B}, SemiRing_Morphism (naturals_to_semiring A B)
+  Context `{SemiRing A} `{!NaturalsToSemiRing A}.
+  Class Naturals: Prop :=
+    { naturals_ring:> SemiRing A
+    ; naturals_to_semiring_mor:> Π `{SemiRing B},
+        SemiRing_Morphism (naturals_to_semiring A B)
     ; naturals_initial:> Initial (bundle_semiring A) }.
@@
 \begin{lstlisting}
-  Class Integers A {e plus mult zero one opp} `{IntegersToRing A} :=
-    { integers_ring:> @Ring A e plus mult zero one opp
-    ; integers_to_ring_mor:> Π `{Ring B}, Ring_Morphism (integers_to_ring A B)
+  Context `{Ring A} `{IntegersToRing A}.
+  Class Integers: Prop :=
+    { integers_ring:> Ring A
+    ; integers_to_ring_mor:> Π `{Ring B},
+        Ring_Morphism (integers_to_ring A B)
     ; integers_initial:> Initial (ring.object A) }.
@@
 \begin{lstlisting}
-  Class Rationals A {e plus mult opp zero one mult_inv } `{Π x y: A, Decision (x = y)} {inj_inv}: Prop :=
-    { rationals_field:> @Field A e plus mult opp zero one mult_inv
-    ; rationals_frac: Surjective (λ p => integers_to_ring (Z nat) A (fst p) * / integers_to_ring (Z nat) A (snd p)) (inv:=inj_inv)
+  Context `{Field A} `{Π x y: A, Decision (x = y)} {inj_inv}.
+  Class Rationals: Prop :=
+    { rationals_field:> Field A
+    ; rationals_frac: Surjective
+        (λ p => integers_to_ring (Z nat) A (fst p) *
+          / integers_to_ring (Z nat) A (snd p)) (inv:=inj_inv)
     ; rationals_embed_ints: Injective (integers_to_ring (Z nat) A) }.
 \end{lstlisting}
-Here, \lstinline|Z| is an \lstinling|Integers| implementation paramerized by a \lstinline|Naturals| implementation, for which we just take \lstinline|nat|.
-The choice of \lstinline|Z nat| here is immaterial; we could have picked any other (or even an explicitly arbitrary) implementation of \lstinilne|Integers|, but doing so would provide no benefit.
+Here, \lstinline|Z| is an \lstinline|Integers| implementation paramerized by a \lstinline|Naturals| implementation, for which we just take \lstinline|nat|.
+The choice of \lstinline|Z nat| here is immaterial; we could have picked any other (or even an explicitly arbitrary) implementation of \lstinline|Integers|, but doing so would provide no benefit.
   % todo: hm, this Z is poorly named. it's really natpair_integers.Z
   % todo: say a few words about how we handle division by zero?
+
+% todo: say a word about section variable/context discharging, because we use it everywhere.
 

### 2010-06-23 jar.tex
Work on conclusion; reintegrate several commented out parts (with heavy massaging).

@@
 
-% outline:
-% 
-% - preliminaries
-%     -> coq (couple of sentences about programming in a combined framework with logic)
-%     -> type classes
-% - on bundling
-%     -> introduce problem
-%     -> naive solution
-%     -> etc
-%     -> "we first begin by showing how unbundling is in some sense the "right" way to do things"
-%     in doing this, however, we will find that doing things this way is somewhat painful
-%     -> derived principles
-% - Managing unbundled structure with type classes
-%     -> painful if flexible, which is why not done before
-%     -> with type classes, managable.
-% 
-% - toward the algebraic hierarchy
-% 
-%   leaving things unbundled is one thing, but we need more for the alg. hierarchy. suppose we define a class for semigroups based on yada.
-%   -> operational type classes (parameterized over a single operation = constant), structural type classes (parameterized over all the components that make up the structure).
-% 
-% - advantages
-% 
-% - numeric classes
-% 
-% - cat.theory
-% 
-% - univ. alg.
-% 
-% - quoting
-
-%todo: have a word somewhere about "super classes"
-
 In this paper we focus on the Coq proof assistant. We conjecture that the methods can be transferred
@@
 
-Propositions in Coq are types, which themselves have types called \emph{sorts}.
-Coq features a distinguished sort called \lstinline|Prop| that one may voluntarily choose to use as the sort of types representing propositions.
+Propositions in Coq are types~\cite{ITT,CMCP}, which themselves have types called \emph{sorts}.
+Coq features a distinguished sort called \lstinline|Prop| that one may voluntarily choose to use as the sort for types representing propositions.
 The distinguishing feature of the \lstinline|Prop| sort is that Coq will not permit terms of non-\lstinline|Prop| types to depend on the values of inhabitants of \lstinline|Prop| types (i.e.
@@
 
+% As originally conceived~\cite{ITT,CMCP}, Propositions are Types and we can extract information from
+% them.
+Often we do \emph{not} want functions to depend on proofs.
+In Coq, this is the paradigmatic way of working.
+We are forbidden to extract information from proofs of \lstinline|Prop|s.
+Especially with the aid of the Program machinery, this allows us to flexibly work with simply typed program and their Hoare style correctness proofs.
+
 Adopting this regime of discrimination establishes a weak form of proof irrelevance, in that changing a proof can never affect the result of value computations.
@@
 A stronger notion of proof irrelevance in which any two inhabitants of a \lstinline|Prop| type are made \emph{convertible} is planned for a future Coq version.
+% todo: process?:
+%One may be tempted to propose Propositions as [ ]-types~\cite{awodey2004propositions}. Here the
+%construction $[P]:=\{\top\mid p:P\}$ squashes all the information about how we proved $P$. In Coq,
+%the [ ]-types are isomorphic to the type of sort \lstinline|Prop|. However,
+%there are places where we needed proof \emph{relevance}; e.g.\ in the proof of the first isomorphism
+%theorem above.In practice, manual dead-code analysis seems useful. Making Harrop formulas proof
+%irrelevant seems to be a good first approximation~\cite{lcf:spi:03}. However, a further refinement
+%of Propositions as types seems necessary.
+% Making Harrop formulas proof irrelevant seems to be a good first approximation~\cite{lcf:spi:03}.
+An orthogonal and further reaching proposal is to replace the Coq type theory with the Calculus of implicit constructions with implicit Σ-types~\cite{miquel2001implicit,barras2008implicit,Bernardo}.
+
+Occasionally there is some ambiguity as to whether a certain piece of information (such as a witness to an existential statement) is strictly ``proof matter'' (and thus belongs in the \lstinline|Prop| sort) or actually of further computational interest (and thus does \emph{not} belong to the \lstinline|Prop| sort).
+We will see one such case when we discuss the first homomorphism theorem in section \ref{homothm}.
+Coq provides a modest level of \emph{universe-polymorphism} so that we may avoid duplication when trying to support \lstinline|Prop|-sorted and non-\lstinline|Prop|-sorted content with a single set of definitions.
+
+% However, we would also like to use Coq as a dependently typed programming language.
+This requires a hybrid approach, were we put informative propositions in \lstinline|Type| and non-informative ones in \lstinline|Prop|.
+% Eelis: I don't see the logic here.
 
@@
 
-Effectively keeping track of, resolving, and combining proofs of equivalence-ness and propriety when the user attempts to substitute a given (sub)term using a given equality, requires nontrivial infrastructure and support from the system.
+Because the Coq type theory lacks ``real'' quotient types (as it would make type checking undecidable), the conventional way of working with abstract structures in Coq is to always start with \emph{setoids} (aka Bishop sets): types equipped with an equivalence relation~\cite{Bishop67,Hofmann,Capretta}.
+As we will see in section ~\ref{univ}, this way of working by nature lends itself well to particularly simple definitions of notions such as quotient algebras.
+
+Effectively keeping track of, resolving, and combining proofs of equivalence-ness and propriety when the user attempts to substitute a given (sub)term using a given equality, is known as ``setoid rewriting'', and requires nontrivial infrastructure and support from the system.
 The Coq implementation of these mechanisms was largely rewritten in 200?
-by Matthieu Sozeau~\ref{rewriting}.
-(todo: say that it's now much better than before)
+by Matthieu Sozeau~\cite{Setoid-rewrite} to make it more flexible and to replace the old special-purpose setoid/morphism registration command with a clean type class based interface.
 
@@
 Thus, type classes in Coq are realized by relatively minor syntactic aids that bring together existing facilities of the theory and the system into a coherent idiom, rather than by introduction of a new category of qualitatively different definitions with their own dedicated semantics.
-
-% BS: put back
-% The Coq type theory lacks quotient
-% types, as it would make type checking undecidable. % todo: reference
-% Instead it supports setoids (aka Bishop sets), a pair of a type and an equivalence relation~\cite{Bishop67,Hofmann,Capretta}.
-Recently, improved support for setoids has been added to the system~\cite{Setoid-rewrite}.
-Functions respecting these equivalence relations are called \lstinline|Proper|.
-% As we will see in Section~\ref{univ} below this functionality can also be used to work with
-% quotient algebras.
 
@@
 \end{quote}
+% todo: better example
+
 Without further support from the system, this would have to be written as
@@
 
+% note: we don't mention "super classes" because by Matthieu's terminology only parameters-of-class-type are "super classes", and for those that means only the operational classes, which we certainly don't think of as being part of the hierarchy.
+
 However, we can easily add additional inheritance relations by declaring corresponding class instances.
@@
 What we really want these fields to express is ``\emph{if} in a certain context we know something to be a \lstinline|Monoid_Morphism|, \emph{then} realize that the source and target are \lstinline|Monoid|s''.
-However, the current instance resolution implementation does not support this style of \emph{forward} reasoning, and only supports \emph{backward} reasoning: had we registered \lstinline|monmor_to| and \lstinline|monmor_from| as instance resolution hints, we would in fact be saying ``\emph{if} trying to establish that something is a \lstinline|Monoid|, \emph{then} try finding a \lstinline|Monoid_Morphism| to or from it'', which quickly degenerates into a wild goose chase.
+However, the current instance resolution implementation has little support for this style of \emph{forward} reasoning, and is really primarily oriented on \emph{backward} reasoning: had we registered \lstinline|monmor_to| and \lstinline|monmor_from| as instance resolution hints, we would in fact be saying ``\emph{if} trying to establish that something is a \lstinline|Monoid|, \emph{then} try finding a \lstinline|Monoid_Morphism| to or from it'', which quickly degenerates into a wild goose chase.
 We will return to this point in section~\ref{conclusion}.
@@
 % First, multiple inheritance becomes trivial: \lstinline|SemiRing| inherits two \lstinline|Monoid| structures on the same carrier and setoid relation, using ordinary named arguments to achieve ``manifest fields''; see section~\ref{manifest}.
-% 
-% Second, because our terms are small and independent and never refer to proofs, we are invulnerable to concerns about efficiency and ambiguity of projection paths that plague existing solutions, obviating the need for extensions like the proposed coercion pullbacks~\cite{Hints}.
-% 
+
 % Third, since our structural type classes are mere predicates, overlap between their instances is a non-issue.
@@
 \subsection{The first homomorphism theorem}
+\label{homothm}
 
@@
 
-% todo: decision
 The generic \lstinline|Decision| instance for \lstinline|Naturals| equality implemented by mapping to \lstinline|nat| will typically be far less efficient than a specialized implementation for a particular representation of the natural numbers.
-Fortunately, with Coq's type classes it is no problem for instances overlapping in this way to co-exist, and we can even deprioritize the generic instance so that instance resolution will always pick the specialization when the representation is known.
+Fortunately, with Coq's type classes it is no problem for instances overlapping in this way to co-exist, and we can even deprioritize the generic instance (using a numeric priority annotation) so that instance resolution will always pick the specialization when the representation is known.
 
@@
 
-%integers all the same:
-
-% At the time of writing, our development includes a fully integrated formalization of a nontrivial portion of category theory and multisorted universal algebra, including various categories (e.g.\ the category \lstinline|Cat| of categories, and finitary algebraic categories  defined by a theory which we instantiate to obtain the categories of monoids, semirings, and rings), functors (including automatically generated forgetful functors), natural transformations, adjunctions, initial models of equational theories constructed from term algebras, transference of proofs between isomorphic models of equational theories, subalgebras, congruences, quotients, products, and the first homomorphism theorem~\cite{meinke1993universal}.
-
-EELIS UP TO HERE
-
-\subsection{Categorical universal algebra}
-Having discussed category theory, we now turn to its application to universal algebra.
-
-%varieties/closed_terms.v
-We define the forgetful functor $U$ from the category of τ-algebras to the category of sets and its left-adjoint $F$: the term algebra consisting of closed terms, i.e.\ terms with the empty type as set of variables.
+% Eelis: and when we do, we can write something like:
+% Given a ring $R$, the $R$-algebra $R[X]$ of polynomials, is the free $R$-algebra on a set $X$.
+% We provide two implementations of polynomials: the
+% standard representation using lists of coefficients and the Bernstein representation. A Bernstein
+% basis polynomial is one of the form:
+% \[b_{\nu,n}(x) = {n \choose \nu} x^{\nu} \left( 1 - x \right)^{n - \nu}, \quad \nu = 0, \ldots, n.\]
+% A Bernstein polynomial is a linear combination of these basic polynomials. Bernstein polynomials
+% have been used for efficient computations inside the Coq system~\cite{ZumkellerPhD}.
+
+% Eelis: TODO: process: At the time of writing, our development includes a fully integrated formalization of a nontrivial portion of category theory and multisorted universal algebra, including various categories (e.g.\ the category \lstinline|Cat| of categories, and finitary algebraic categories  defined by a theory which we instantiate to obtain the categories of monoids, semirings, and rings), functors (including automatically generated forgetful functors), natural transformations, adjunctions, initial models of equational theories constructed from term algebras, transference of proofs between isomorphic models of equational theories, subalgebras, congruences, quotients, products, and the first homomorphism theorem~\cite{meinke1993universal}.
+
+%\subsection{Categorical universal algebra}
+
+% We define the forgetful functor $U$ from the category of τ-algebras to the category of sets and its left-adjoint $F$: the term algebra consisting of closed terms, i.e.\ terms with the empty type as set of variables.
 This term algebra is the initial τ-algebra.
@@
 We will use the resulting expression monad in Section~\ref{quote} below.
+% Eelis: oh?
 
 %The product of two $\tau$-algebras is their categorical product.
-
-\begin{comment}
-Given a sub\emph{set} of an algebra, we define the sub\emph{algebra} generated by it.
-An interesting fact to mention is the use of the following heterogeneous equality between
-elements of the algebra and of the subalgebra, i.e.\ terms of different types may be equal.
-\begin{lstlisting}
-  Fixpoint heq {o}: op_type carrier o -> op_type v o -> Prop :=
-    match o with
-    | constant x => fun a b => `a == b
-    | function x y => fun a b => forall u, heq (a u) (b (`u))
-    end.
-\end{lstlisting}
-\end{comment}
-
-We connect the abstract theory to the concrete theory. Concretely, let $\tau_m$ be the theory
-of monoids. Given a (concrete) monoid, we can construct the corresponding $\tau_m$-algebra. 
-Conversely, given a $\tau_m$-algebra, we construct an instance of the \lstinline|Monoid| type class.
-%varieties/monoid.v
-\begin{lstlisting}
-Variable o: variety.Object theory.
-Global Instance: SemiGroupOp (o tt) := algebra_op theory mult.
-Global Instance: MonoidUnit (o tt) := algebra_op theory one.
-Global Instance from_object: Monoid (o tt).
-\end{lstlisting}
-
-In the last line, instance resolution `automatically' finds the operation and the unit specified in
-the lines before.
-
-This interplay between concrete algebraic structure and their expressions on the one hand, and models of equational theories on the other is occasionally a source of tension (in that translation in either direction is not yet fully automatic).
+% 
+% \begin{comment}
+% Given a sub\emph{set} of an algebra, we define the sub\emph{algebra} generated by it.
+% An interesting fact to mention is the use of the following heterogeneous equality between
+% elements of the algebra and of the subalgebra, i.e.\ terms of different types may be equal.
+% \begin{lstlisting}
+%   Fixpoint heq {o}: op_type carrier o -> op_type v o -> Prop :=
+%     match o with
+%     | constant x => fun a b => `a == b
+%     | function x y => fun a b => forall u, heq (a u) (b (`u))
+%     end.
+% \end{lstlisting}
+% \end{comment}
+% 
+% We connect the abstract theory to the concrete theory. Concretely, let $\tau_m$ be the theory
+% of monoids. Given a (concrete) monoid, we can construct the corresponding $\tau_m$-algebra. 
+% Conversely, given a $\tau_m$-algebra, we construct an instance of the \lstinline|Monoid| type class.
+% %varieties/monoid.v
+% \begin{lstlisting}
+% Variable o: variety.Object theory.
+% Global Instance: SemiGroupOp (o tt) := algebra_op theory mult.
+% Global Instance: MonoidUnit (o tt) := algebra_op theory one.
+% Global Instance from_object: Monoid (o tt).
+% \end{lstlisting}
+% 
+% In the last line, instance resolution `automatically' finds the operation and the unit specified in
+% the lines before.
+% 
+% This interplay between concrete algebraic structure and their expressions on the one hand, and models of equational theories on the other is occasionally a source of tension (in that translation in either direction is not yet fully automatic).
 However, it opens the door to the possibility of fully internalized implementations of generic tactics for algebraic manipulation, no longer requiring plugins.
 We come back to this when we describe automatic quotation of concrete expressions into universal algebra expressions; see Section~\ref{quote}.
-
-\section{Interfaces using category theory}\label{interfaces}\label{modul}
-We have characterized the naturals as the initial object in the category of semirings and derived
-many of their properties from this interface. Unary, binary and machine numbers are
-instances of this interface, so we can directly apply these results to all these instances.
-Similarly, the integers are the initial ring. The rationals are the field of fractions of the
-integers. 
-Given a ring $R$, the $R$-algebra $R[X]$ of polynomials, is the free $R$-algebra on a set $X$.
-We provide two implementations of polynomials: the
-standard representation using lists of coefficients and the Bernstein representation. A Bernstein
-basis polynomial is one of the form:
-\[b_{\nu,n}(x) = {n \choose \nu} x^{\nu} \left( 1 - x \right)^{n - \nu}, \quad \nu = 0, \ldots, n.\]
-A Bernstein polynomial is a linear combination of these basic polynomials. Bernstein polynomials
-have been used for efficient computations inside the Coq system~\cite{ZumkellerPhD}.
-
-We encourage more efficient implementations by assigning the default implementation a
-low priority. For example, the distance function on the natural numbers, which is derived from its
-semiring structure, is assigned priority 10.
-\begin{lstlisting}
-  Global Program Instance: NatDistance N | 10 := ...
-\end{lstlisting}
-
-In future work we aim to base our development of the reals
-on an abstract dense set, allowing us to use the efficient dyadic
-rationals~\cite{boldo2009combining} as a base for exact real number computation in
-Coq~\cite{Riemann,Oconnor:real}. The use of category theory has been important in these developments.
+% 
+%
 
@@
 %which moreover works on setoid equalities.
-
-\section{Canonical structures}\label{canonical}
-Packed classes use canonical structures for the algebraic hierarchy. Both canonical structures and
-type classes may be seen as instances of hints in unification~\cite{Hints}. Some uses of canonical
-structures can be replaced by type class instances. The user manual (2.7.15) uses canonical
-structures to derive the setoid equality on the natural
-numbers in the following example \lstinline|(S x)==(S y)|. In this case type classes provide similar proof
-terms. Canonical structures give
-\begin{lstlisting}
-@equiv(Build_Setoid nat (@eq nat)(@eq_equivalence nat))(S x)(S y)
-\end{lstlisting}
-
-which includes an explicit proof that \lstinline|(@eq nat)| is an equivalence,
-whereas we obtain \lstinline|@equiv nat (@eq nat) (S x) (S y)|.
 
@@
 
-\subsection{Propositions as Types}
-As originally conceived~\cite{ITT,CMCP}, Propositions are Types and we can extract information from
-them.
-Often we do \emph{not} want functions to depend on proofs.
-In Coq, this is the paradigmatic way of working.
-We are forbidden to extract information from proofs of \lstinline|Prop|s.
-Especially with the aid of the Program machinery, this allows us to flexibly work with simply typed program and their Hoare style correctness proofs.
-However, we would also like to use Coq as a dependently typed programming language.
-This requires a hybrid approach, were we put informative propositions in \lstinline|Type| and non-informative ones in \lstinline|Prop|.
-Making Harrop formulas proof irrelevant seems to be a good first approximation~\cite{lcf:spi:03}.
-An orthogonal and further reaching proposal is to replace the Coq type theory with the Calculus of implicit constructions with implicit
-Σ-types~\cite{miquel2001implicit,barras2008implicit,Bernardo}.
-
-%One may be tempted to propose Propositions as [ ]-types~\cite{awodey2004propositions}. Here the
-%construction $[P]:=\{\top\mid p:P\}$ squashes all the information about how we proved $P$. In Coq,
-%the [ ]-types are isomorphic to the type of sort \lstinline|Prop|. However,
-%there are places where we needed proof \emph{relevance}; e.g.\ in the proof of the first isomorphism
-%theorem above.In practice, manual dead-code analysis seems useful. Making Harrop formulas proof
-%irrelevant seems to be a good first approximation~\cite{lcf:spi:03}. However, a further refinement
-%of Propositions as types seems necessary.
-
-\section{Coq wishlist}
-% todo: maybe merge with conclusions and add "and future work"
-
-While we have already found the predicate-class-based style of formalization to work very well for our development thus far, and believe it has the potential to be \emph{the} modern solution to this problem, in our work we have encountered a number of limitations in the Coq implementation of key features of the tool, that keep us from making an unconditional recommendation for universal adoption of these techniques.
-
-\paragraph{Universe-polymorphic definitions}
-Coq has supported universe-polymorphism for parameterized inductive definitions (different instantiations of which are automatically of the lowest possible sort taking into account the universe level constraints on the arguments) for some time.
+% 
+% \section{Coq wishlist}
+% % todo: maybe merge with conclusions and add "and future work"
+% 
+% \paragraph{Universe-polymorphic definitions}
+% Coq has supported universe-polymorphism for parameterized inductive definitions (different instantiations of which are automatically of the lowest possible sort taking into account the universe level constraints on the arguments) for some time.
 However, the same functionality has not been available for \emph{definitions}, primarily because of efficiency concerns.
@@
 This should solve our problems.
-
-\paragraph{Implicit generalization}
-
-The current syntax for implicit generalization 
-%current syntax very subtle.
+% 
+% \paragraph{Implicit generalization}
+% 
+% The current syntax for implicit generalization 
+% %current syntax very subtle.
 various combinations of `, class-ness, {} vs (), !
@@
 Just a remaining source of the occasional uglyness in declarations.
-
-\paragraph{Instance resolution efficiency} ...
-
-The current implementation is in some sense only a prototype.
-
-Backward
-
-Forward
-
-It seems likely that known techniques from established resolution engines can be reused to significantly increase performance.
-
-In reply to our inquiries about these issues the Coq development team has agreed that these are legitimate concerns and wishes, and indicated that they are committed to addressing them.
+%
 
 \section{Conclusions}
-Telescopes have been criticized~\cite{Packed} for the lack of multiple inheritance and
-the efficiency penalty of a long chain of coercion projections. Packed classes~\cite{Packed} provide
-a solution to these problems. We have provided an alternative solution based on type classes.
-
-We conclude that unification hints, canonical structures and type classes all appear to have
-similar expressive power. Canonical structures, being tied to the unifier, are more robust, but
-seem to require more ingenuity. Type classes are easier to use. Unification hints, may be a
-good generalization, but are absent from Coq. We encourage their inclusion to the
-Coq system.
+
+While bundling operational and propositional components of abstract structures into records may seem natural at first, doing so actually introduces many serious problems.
+Keeping things unbundled avoids these problems and is more flexible in theory, and now with type classes finally also in practice, or so we hope to have demonstrated.
+
+%Telescopes have been criticized~\cite{Packed} for the lack of multiple inheritance and
+%the efficiency penalty of a long chain of coercion projections. Packed classes~\cite{Packed} provide
+%a solution to these problems. We have provided an alternative solution based on type classes.
+
+It has been suggested that canonical structures are more robust (in the sense of being much more restricted and internally closed compared to the comparatively wild and open-ended proof search used for instance resolution).
+However, in addition to the unfortunate fact that the restrictions force one into bundled representations, a second consequence is that using them for anything beyond basic use cases requires significant ingenuity, whereas type classes are fairly easy and straightforward to use.
+
+% \section{Canonical structures}\label{canonical}
+% Packed classes use canonical structures for the algebraic hierarchy. Both canonical structures and
+% type classes may be seen as instances of hints in unification~\cite{Hints}. Some uses of canonical
+% structures can be replaced by type class instances. The user manual (2.7.15) uses canonical
+% structures to derive the setoid equality on the natural
+% numbers in the following example \lstinline|(S x)==(S y)|. In this case type classes provide similar proof
+% terms. Canonical structures give
+% \begin{lstlisting}
+% @equiv(Build_Setoid nat (@eq nat)(@eq_equivalence nat))(S x)(S y)
+% \end{lstlisting}
+% which includes an explicit proof that \lstinline|(@eq nat)| is an equivalence,
+% whereas we obtain \lstinline|@equiv nat (@eq nat) (S x) (S y)|.
+
+Unification hints are a more general mechanism than type classes, and could provide a more precise account of the interaction between implicit argument inference and proof search.
+Perhaps, then, a fruitful approach could be to use unification hints as the underlying mechanism with type classes as an end-user interface encapsulating a particularly convenient idiom for using them.
+% We encourage their inclusion to the Coq system. -- Eelis: not quite ready to go there yet
+
+The only real pending concern that keeps us from making an unequivocal endorsement of type classes as a versatile, expressive, and elegant means of organizing proof developments, is instance resolution efficiency.
+In more complex parts of our development we are now experiencing increasingly serious efficiency problems, despite having already made sacrifices by artifically inhibiting many natural class instances in order not to further strain instance resolution.
+
+%While we have already found the predicate-class-based style of formalization to work very well for our development thus far, and believe it has the potential to be \emph{the} modern solution to this problem, in our work we have encountered a number of limitations in the Coq implementation of key features of the tool, that keep us from making an unconditional recommendation for universal adoption of these techniques.
+
+Fortunately, the Coq development team recognizes that this is a real problem, and furthermore agrees that there is plenty of potential for improvement of the current implementation, potentially borrowing techniques from the vast literature on efficient implementation of Prolog-style resolution (which the hint-based proof search used for instance resolution greatly resembles).
+
+We are currently in the process of retrofitting the rationals interface into CoRN.
+In future work we aim to base our development of its reals on an abstract dense set, allowing us to use the efficient dyadic rationals~\cite{boldo2009combining} as a base for exact real number computation in Coq~\cite{Riemann,Oconnor:real}.
+The use of category theory has been important in these developments.
 

### 2010-06-23 jar.tex
Reintegrate small note about ssreflect.
Started work on quotation section.

@@
   % maybe have a chapter called "interfaces are primary, not implementations" (also involve polynomials)
-  % say this somewhere: \footnote{For an interesting (if fairly restrictive) alternative design in which unicity of term representations \emph{is} required for all objects, see the algebraic hierarchy in the ssreflect\cite{ssreflect} standard library.}
 
@@
 
+For an interesting (if fairly restrictive) alternative approach based on the idea of requiring unicity of representation for all objects so that Leibniz equality is always applicable, see the algebraic hierarchy part of the ssreflect\cite{ssreflect} libraries.
+
 \paragraph{Type classes}
@@
 
-% note: we don't mention "super classes" because by Matthieu's terminology only parameters-of-class-type are "super classes", and for those that means only the operational classes, which we certainly don't think of as being part of the hierarchy.
+% note: we don't mention "super classes" because by Matthieu's terminology only parameters-of-class-type are "super classes", and for us that means only the operational classes, which we certainly don't think of as being part of the hierarchy.
+Matthieu calls fields of class type declared with ":>" substructures.
 
@@
 Following our idiom, we introduce operational type classes for the \emph{components} of a category:
-
 \begin{lstlisting}
@@
 category~\cite{saibi1995constructive}:
-
 \begin{lstlisting}
@@
 
-To give a further taste of what universal algebra in our development looks like, we consider the definitions involved in the first homomorphism theorem in more detail:
+To give a further taste of what universal algebra in our development looks like, we consider the definitions involved in the first homomorphism theorem~\cite{meinke1993universal} in more detail:
 \begin{theorem}[First homomorphism theorem]
@@
 
-% Eelis: TODO: process: At the time of writing, our development includes a fully integrated formalization of a nontrivial portion of category theory and multisorted universal algebra, including various categories (e.g.\ the category \lstinline|Cat| of categories, and finitary algebraic categories  defined by a theory which we instantiate to obtain the categories of monoids, semirings, and rings), functors (including automatically generated forgetful functors), natural transformations, adjunctions, initial models of equational theories constructed from term algebras, transference of proofs between isomorphic models of equational theories, subalgebras, congruences, quotients, products, and the first homomorphism theorem~\cite{meinke1993universal}.
+% Eelis: TODO: process: At the time of writing, our development includes a fully integrated formalization of a nontrivial portion of category theory and multisorted universal algebra, including various categories (e.g.\ the category \lstinline|Cat| of categories, and finitary algebraic categories  defined by a theory which we instantiate to obtain the categories of monoids, semirings, and rings), products
 
@@
 We come back to this when we describe automatic quotation of concrete expressions into universal algebra expressions; see Section~\ref{quote}.
-% 
-%
-
-\section{Type class quoting}\label{quote}
-Unification hints~\cite{Hints} allow one to semi-automatically construct a quote
-function.\footnote{Gonthier provides similar functionality using canonical structures.} This feature is absent from Coq. Fortunately, type classes provide similar functionality as we will now show.
-
-We define a term language for monoids
-\begin{lstlisting}
-Inductive Expr (V: Type) := Mult (a b: Expr V) | Zero | Var (v: V).
-\end{lstlisting}
-
-The expression type is parametrized over the set of variable indices. Hence, we diverge from~\cite{Hints}, 
-which uses \lstinline|nat| for variables thereby introducing bound problems and dummy variables.
-
- An expression is only meaningful in the context of a variable
-assignment:\footnote{Categorically: \lstinline|nat| is
-an \lstinline|Expr|-algebra and \lstinline|eval| is \lstinline|map vars| composed with the algebra map.}
+
+\section{Quoting with type classes}\label{quote}
+
+A common need when interfacing generic theory and utilities developed for algebraic structures (such as normalization procedures) with concrete instances of these structures is to take a concrete expression or statement in a model of a particular algebraic structure, and translate it to a symbolic expression or statement in the language of the algebra's signature, so that its structure can be inspected.
+
+Traditionally, proof assistants such as Coq have provided sophisticated tactics or built-in commands to support this ``quoting''.
+In~\cite{Hints}, TODO et.
+al.
+show that unification hints, a very general way of facilitating user-defined extensions to term and type inference, can be used to semi-automatically build quote functions without dropping to a meta-level.\footnote{Gonthier provides similar functionality using ingenious use of canonical structures.} This feature is absent from Coq, but fortunately type classes let us do very much the same thing, as we will now show.
+
+For ease of presentation we show only a proof of concept for a very concrete language.
+We are currently working to integrate this technique with our existing universal algebra infrastructure, in particular its data types for algebraic terms.%  Preliminary results in this area (such as a simple monoid normalization tactic) suggest yada.
+
+For the present example, we define a term language for monoids
+\begin{lstlisting}
+Inductive Expr (V: Type) := Mult (a b: Expr V) | One | Var (v: V).
+\end{lstlisting}
+
+The expression type is parametrized over the set of variable indices.
+Hence, we diverge from~\cite{Hints}, which uses \lstinline|nat| for variables, thereby introducing bound problems and a need for dummy variables.
+
+An expression is only meaningful in the context of a variable assignment:\footnote{Categorically: \lstinline|nat| is an \lstinline|Expr|-algebra and \lstinline|eval| is \lstinline|map vars| composed with the algebra map.}
 \begin{lstlisting}
@@
   match e with
-  | Zero => 0
+  | One => 1
   | Mult a b => eval vs a * eval vs b
@@
 
-We have used this technique to implement a tactic to normalize monoid expressions.
+We have used this technique to implement a tactic to normalize monoid expressions, and plan to extend yada yada.
 %It would be interesting to use this technique to implement the \lstinline|congruence| tactic, which implements

### 2010-06-24 jar.tex
Added MSCS style files. Gives strange errors.  Reverted changes partially. Better wait for the final version. Note: MSCS says use footnotes scarsely.
Revise quoting section. Merge and fix natural-transformation and adjunction sections.

@@
 \usepackage{amsmath}
+\usepackage{amssymb}
 \usepackage{upgreek}
@@
 % todo:
-  % what to say about terms referring on proofs?
   % cite that other type class based development in the context of the alternative bundling schemes.
@@
   {λ}{{$\uplambda\ $}}1
-  {σ}{{$\upsigma\ $}}1
+  {σ}{{$\upsigma$}}1
+  {η}{{$\upeta$}}1
+  {φ}{{$\upphi$}}1
   {∃}{{$\exists$}}1
@@
   {⟶}{{$\longrightarrow\ $}}1
+  {⇛}{{$\Rrightarrow\ $}}1
   {∧}{{$\land$}}1
   {∀}{{$\forall$}}1
-  {Π}{{$\Uppi\ $}}1
+  {Π}{{$\Uppi$}}1
   {η}{{$\upeta$}}1
@@
   {∘}{{$\circ\ $}}1
-  {◎}{{$\odot\ $}}1
+  {◎}{{$\odot$}}1
   { ≡ }{{$\equiv\ $}}1
@@
 
-\subsection{Natural transformations}
-Natural transformations are formalized as:
+\subsection{Natural transformations and adjunctions}
+
+We introduce a convenient notation for the type of the computational content of a natural transformation between two functors:
+\begin{lstlisting}
+  Notation "F ⇛ G" := (Π x, F x ⟶ G x).
+\end{lstlisting}
+Now assume the following context:
 \begin{lstlisting}
   Context `{Category C} `{Category D}
-   `{!Functor (F: C → D) Fa} `{!Functor (G: C → D) Ga}.
-   Class NaturalTransformation (η: Π c, F c ⟶ G c): Prop :=
-     natural: Π (x y: C) (f: x ⟶ y), η$$ y ∘ fmap F f = fmap G f ∘ η$$ x.
+    `{Functor (F: C → D)} `{Functor (G: D → C)}.
+\end{lstlisting}
+The naturality property is easy to write:
+\begin{lstlisting}
+  Class NaturalTransformation (η: F ⇛ G): Prop :=
+    { naturaltrans_from: Functor F
+    ; naturaltrans_to: Functor G
+    ; natural: Π `(f: x ⟶ y), η$$ y ◎$$ fmap F f = fmap G f ◎$$ η$$ x }.
 \end{lstlisting}
@@
 
-We have defined the category of setoids, the dual of a category, products, and hence co-products. 
-We prove that two definitions of adjunction are equivalent. The proofs are concise and follow the textbooks.
-
-The usual size problems in the definition of the category of categories can be avoided by using universe polymorphism. 
-However, we need to avoid making \lstinline|Relation| a \lstinline|Definition|, since \lstinline|Definitions| are not (yet?) universe polymorphic.
-
-\subsection{Adjunctions}
+% We have defined the category of setoids, the dual of a category, products, and hence co-products.
+
+% The usual size problems in the definition of the category of categories can be avoided by using universe polymorphism. 
+% However, we need to avoid making \lstinline|Relation| a \lstinline|Definition|, since \lstinline|Definitions| are not (yet?) universe polymorphic.
 
 Adjunctions can be defined in different ways. A nice symmetric definition is the following:
-
-\begin{lstlisting}
-  Context `{Category A} `{Category X}
-    F `{!Functor (F: X → A) F'}
-    G `{!Functor (G: A → X) G'}
-    (φ: Π {x a}, (F x ⟶ a) → (x ⟶ G a))
-    `{Π x a, Inverse (φ x a)}
-    `{Π x a, Bijective (φ x a)}.
-
-  Class Adjunction: Prop :=
+\begin{lstlisting}
+  Class Adjunction (φ: Π `(F c ⟶ d), (c ⟶ G d)): Prop :=
     { adjunction_left_functor: Functor F _
     ; adjunction_right_functor: Functor G _
-   ; natural_left `(k: a ⟶ a'): `((fmap G k ◎) ∘ φ = @φ x _ ∘ (k ◎))
-   ; natural_right `(h: x' ⟶ x):`((◎ h) ∘ @φ _ a = φ ∘ (◎ fmap F h)) }.
-\end{lstlisting}
-
-blabla, alternative definition, equivalence theorem nice stress test, preliminary results promising except for rewriting/instance-resolution efficiency.
+    ; natural_left `(f: d ⟶ d') c: (fmap G f ◎) ∘ φ$$ = φ (c:=c) ∘ (f ◎)
+    ; natural_right `(f: c' ⟶ c) d: (◎ f) ∘ φ (d:=d) = φ$$ ∘ (◎ fmap F f) }.
+\end{lstlisting}
+An alternative definition is the following:
+\begin{lstlisting}
+  Class AltAdjunction (η: id ⇛ G ∘ F) (φ: Π `(f: c ⟶ G d), F c ⟶ d): Prop :=
+    { alt_adjunction_natural_unit: NaturalTransformation η
+    ; alt_adjunction_factor: Π `(f: c ⟶ G d),
+        is_sole ((f =) ∘ (◎ η$$ c) ∘ fmap G) (φ f) }.
+\end{lstlisting}
+Formalizing the (nontrivial) proof that these two definitions are equivalent provides a nice test for our definitions.
+Preliminary results indicate that we can very concisely and closely follow the textbook proof.
+% todo: ref?
 
@@
 \begin{lstlisting}
-  Class Algebra `{Π a, Equiv (carriers a)} `{AlgebraOps σ carriers}: Prop :=
+  Class Algebra `{Π a, Equiv (carriers a)} `{AlgebraOps σ$$ carriers}: Prop :=
       { algebra_setoids:> Π a, Setoid (carriers a)
@@
 \begin{lstlisting}
-Context `{Algebra σ A}.
+Context `{Algebra σ$$ A}.
 Class Congruence (e: Π s: sorts σ, relation (v s)): Prop :=
@@
 \begin{lstlisting}
-  Context `{Algebra σ A} `{Algebra σ B} `{!HomoMorphism σ A B f}.
+  Context `{Algebra σ$$ A} `{Algebra σ$$ B} `{!HomoMorphism σ$$ A B f}.
 \end{lstlisting}
@@
 \begin{lstlisting}
-   Instance co: Congruence σ (~).
+   Instance co: Congruence σ$$ (~).
 \end{lstlisting}
@@
 \begin{lstlisting}
-  Definition image s (b: B s): Type := sigT (λ a => f s a = b).
+  Definition image s (b: B s): Type := sigT (λa => f s a = b).
 
@@
 \begin{lstlisting}
-  Context `{Algebra σ A} (P: Π s, A s → Type).
+  Context `{Algebra σ$$ A} (P: Π s, A s → Type).
   Class ClosedSubset: Type :=
@@
     algebra.object σ A (algebra_equiv:=(~)).
-  Definition subobject: algebra.Object σ :=
+  Definition subobject: algebra.Object σ$$ :=
     algebra.object σ (ua_subalgebraT.carrier image).
@@
 
-For the present example, we define a term language for monoids
+For the present example, we define an ad-hoc term language for monoids
 \begin{lstlisting}
@@
 \end{lstlisting}
-
 The expression type is parametrized over the set of variable indices.
@@
 
-An expression is only meaningful in the context of a variable assignment:\footnote{Categorically: \lstinline|nat| is an \lstinline|Expr|-algebra and \lstinline|eval| is \lstinline|map vars| composed with the algebra map.}
+Suppose now that we want to quote \lstinline|nat| expressions built from 1 and multiplication.
+To describe the relation we want the symbolic expression to have to the original expression, we first define how symbolic expressions evaluate to values (given a variable assignment):
+%An expression is only meaningful in the context of a variable assignment:\footnote{Categorically: \lstinline|nat| is an \lstinline|Expr|-algebra and \lstinline|eval| is \lstinline|map vars| composed with the algebra map.}
 \begin{lstlisting}
@@
 %$TX:=1+T^2X$, may be seen as the prototypical monad. 
-The \lstinline|Expr| monad, i.e.\ the free \lstinline|Expr|-algebra monad on the category of \lstinline|Type|s with extensional functions%\marginpar{Proof in classquote}
-, provides a bind operation which describes the arrows in the Kleisli category. Concretely,
-\begin{lstlisting}
-Fixpoint bind {V W} (f: V → Expr W) (e: Expr V): Expr W :=
-  match e with
-  | Zero => Zero
-  | Mult a b => Mult (bind f a) (bind f b)
-  | Var v => vs v
-  end.
-\end{lstlisting}
+% The \lstinline|Expr| monad, i.e.\ the free \lstinline|Expr|-algebra monad on the category of \lstinline|Type|s with extensional functions%\marginpar{Proof in classquote}
+% , provides a bind operation which describes the arrows in the Kleisli category. Concretely,
+% \begin{lstlisting}
+% Fixpoint bind {V W} (f: V → Expr W) (e: Expr V): Expr W :=
+%   match e with
+%   | Zero => Zero
+%   | Mult a b => Mult (bind f a) (bind f b)
+%   | Var v => vs v
+%   end.
+% \end{lstlisting}
+% Eelis: above commented out because we never actually use it, and it's really not pertinent to the discussion.
 
@@
 
-\noindent We define some simple combinators for variable packs:
-%
+We can now state our goal: given an expression of type \lstinline|nat|, we seek to construct an \lstinline|Expr V| for some appropriate \lstinline|V| along with a variable assignment, such that evaluation of the latter yields the former.
+Because we will be doing this incrementally, we introduce a few simple variable ``heap combinators'':
 \begin{lstlisting}
@@
 \end{lstlisting}
-
-\noindent These last two combinators are the `constructors' of an implicitly defined subset of
- Galina terms, representing heaps, for which we implement syntactic lookup with type classes.
+These last two combinators are the `constructors' of an implicitly defined subset of Gallina terms, representing heaps, for which we will implement syntactic lookup with type classes in a moment.
 The heap can also be defined explicitly, with no essential change in the code.
-Given a heap and value, \lstinline|Lookup| instances give the value's index in the heap:
-% \marginpar{A
-% canonical structures approach would allow us to do this in Coq?}
-\begin{lstlisting}
-  Class Lookup {A} (x: Value) (f: Vars A) :=
-    { lookup: A; lookup_correct: f lookup = x }.
-\end{lstlisting}
-
-% Context (x: Value) {A B} (va: Vars A) (vb: Vars B).
-
-If the heap is a merge of two heaps and we can find the value's index in the left heap, we can
-access it by indexing the merged heap (and vice versa)
-\begin{lstlisting}
-  Global Instance lookup_left `{!Lookup x va}: Lookup x (merge va vb)
-    := { lookup := inl (lookup x va) }.
-\end{lstlisting}
-
-If the heap is just a singlevar, we can easily index it.
-\begin{lstlisting}
-Global Program Instance: Lookup x (singlevar x) := { lookup := tt }.
-\end{lstlisting}
-
-%One useful operation we need before we get to Quote relates to variables and expression evaluation. 
- As its name suggests, \lstinline|map_var| maps an expression's variable indices.
-This reindexing function is the the \lstinline|map| of the \lstinline|Expr|-monad.
-
-\begin{lstlisting}
-Definition map_var {V W: Type} (f: V → W):
-    Expr V → Expr W :=
-  fix F (e: Expr V): Expr W :=
-    match e with
-    | Mult a b => Mult (F a) (F b)
-    | Zero => Zero
-    | Var v => Var (f v)
-    end.
-\end{lstlisting}
-
-In \lstinline|Quote| below, the idea is that \lstinline|V, l|, and \lstinline|n| are all input variables, while \lstinline|V'| and \lstinline|r| are
-output variables (in the sense that we will rely on unification to generate them). \lstinline|V| and \lstinline|l|
-represent the current heap, \lstinline|n| represents the value we want to quote, and \lstinline|V'| and \lstinline|r'| represent
-the heap of newly encountered variables during the quotation.
-  This explains the type of quote: it is an expression that refers either to variables from
-the old heap, or to newly encountered variables. Finally, \lstinline|eval_quote| is the usual correctness property, which now merges the two heaps.
-
-\begin{lstlisting}
-  Class Quote {V} (l: Vars V) (n: Value) {V'} (r: Vars V') :=
+
+With these, we can define the primary ingredient, the \lstinline|Quote| class:
+\begin{lstlisting}
+  Class Quote {V} (l: Vars V) (n: Value) {V'} (r: Vars V'): Type :=
     { quote: Expr (V + V')
@@
 \end{lstlisting}
-
-Our first instance for \lstinline|Zero| is easy. The `novars' in the result type reflects the fact that no
-new variables are encountered.
-\begin{lstlisting}
-  Global Program Instance quote_zero V (v: Vars V):
-    Quote v 0 novars := { quote := Zero }.
-\end{lstlisting}
-
-The instance for multiplication is a bit more complex. The first line consists of
- variable declarations. The second line is important. `Quote x y z' must be read as
- `quoting y with existing heap x generates new heap z', so the second line basically just
-shuffles heaps around.
- The third line contains some \lstinline|map_var|'s because the heap shuffling must be
-reflected in the variable indices, but apart from that it is just constructing a \lstinline|Mult| term with
-quoted subterms.
-
-\begin{lstlisting}
-Global Program Instance quote_mult V (v: Vars V)
-  n V' (v': Vars V') m V'' (v'': Vars V'')
-  `{!Quote v n v'} `{!Quote (merge v v') m v''}:
-  Quote v (n * m) (merge v' v'') :=
-  { quote := Mult (map_var bla (quote n)) (map_var sum_assoc (quote m)) }.
-\end{lstlisting}
-
-The instance where we recognize values that are already in the heap is expressed by the Lookup requirement, which will only be fulfilled if the Lookup instances defined above can find the value in the heap.
-The novars in the \lstinline|Quote v x novars| result
-   reflects that this quotation does not generate new variables.
-
-\begin{lstlisting}
-Global Program Instance quote_old_var
-  V (v: Vars V) x {i: Lookup x v}:
-  Quote v x novars | 8 := { quote := Var (inl (lookup x v)) }.
-\end{lstlisting}
-
-\noindent Finally, the instance for new variables. We give this a lower priority so that it is only
-used if Lookup fails. The \lstinline|8| in the previous example is a random number less than \lstinline|9| below. 
+We can think of \lstinline|Quote| as the type for a family of Prolog-like syntax-directed resolution functions, which will take as input \lstinline|V| and \lstinline|l| representing previously encountered holes (opaque subexpressions that could not be destructured further) and their values, along with a concrete term \lstinline|n| to be quoted.
+Their ``output'' will consist not only of the fields in the class, but also of \lstinline|V'| and \lstinline|r| representing additional holes and their values.
+Hence, a type class constraint of the form \lstinline|Quote x y z| should be read as ``quoting y with existing heap x generates new heap z''.
+
+The \lstinline|Quote| instance for 1 illustrates the basic idea:
+\begin{lstlisting}
+  Instance quote_one V (v: Vars V): Quote v 1 novars := { quote := One }.
+\end{lstlisting}
+The expression ``1'' can be quoted in any context \lstinline|(V, v)|, introduces no new variables, and the symbolic term representing it is just \lstinline|One|.
+(The \lstinline|eval_quote| field is turned into a trivial proof obligation.)
+
+The \lstinline|Quote| instance for multiplication is a little more subtle, but really only does a bit of heap juggling:
+\begin{lstlisting}
+  Instance quote_mult V (v: Vars V) n V' (v': Vars V') m V'' (v'': Vars V'')
+    `{!Quote v n v'} `{!Quote (merge v v') m v''}: Quote v (n * m) (merge v' v'') :=
+      { quote := Mult (map_var shift (quote n)) (map_var sum_assoc (quote m)) }.
+\end{lstlisting}
+
+These two instances specify how 1 and multiplications are to be quoted, but what about other expressions?
+For these, we want to distinguish two kinds: expressions we have seen before, and those we have not.
+To make this distinction, we need to be able to look up expressions in variable heaps to see if they're already there.
+Importantly, we must do this not by really comparing the values they evaluate to, but by actually browsing the term denoting the variable heap (that is, a composition from \lstinline|novars|, \lstinline|singlevar|, and \lstinline|merge|).
+This, too, is a job for a type class:
+\begin{lstlisting}
+  Class Lookup {A} (x: Value) (v: Vars A) := { key: A; key_correct: v key = x }.
+\end{lstlisting}
+Our first \lstinline|Lookup| instance states that \lstinline|x| can be looked up in \lstinline|singlevar x|:
+\begin{lstlisting}
+  Instance singlevar_lookup (x: Value): Lookup x (singlevar x) := { key := tt }.
+\end{lstlisting}
+Next (and last), if an expression can be looked up in a pack, then it can also be looked up when that pack is merged with another pack:
+\begin{lstlisting}
+  Context (x: Value) {A B} (va: Vars A) (vb: Vars B).
+
+  Instance lookup_left `{!Lookup x va}: Lookup x (merge va vb)
+    := { key := inl (key x va) }.
+
+  Instance lookup_right `{!Lookup x vb}: Lookup x (merge va vb)
+    := { key := inr (key x vb) }.
+\end{lstlisting}
+
+With \lstinline|Lookup|, we can now define a \lstinline|Quote| instance for previously encountered expressions:
+\begin{lstlisting}
+  Instance quote_old_var V (v: Vars V) x {Lookup x v}:
+    Quote v x novars | 8 := { quote := Var (inl (key x v)) }.
+\end{lstlisting}
+
+If none of the \lstinline|Quote| instances defined so far apply, the term in question is a newly encountered hole.
+For this case we define a catch-all instance with a low priority, which yields a singleton heap containing the expression:
+\begin{lstlisting}
+  Instance quote_new_var V (v: Vars V) x: Quote v x (singlevar x) | 9
+    := { quote := Var (inr tt) }.
+\end{lstlisting}
+And with that, we can start quoting:
+\begin{lstlisting}
+Goal Π x y (P: Value → Prop), P ((x * y) * (x * 1)).
+  intros.
+  rewrite <- eval_quote.
+\end{lstlisting}
+The \lstinline|rewrite| rewrites the goal to (something that reduces to):
+\begin{lstlisting}
+  P (eval
+     (merge novars
+        (merge (merge (singlevar x) (singlevar y)) (merge novars novars)))
+     (Mult (Mult (Var (inr (inl (inl ())))) (Var (inr (inl (inr ())))))
+        (Mult (Var (inr (inl (inl ())))) One)))
+\end{lstlisting}
+
+Using a small additional utility lemma we can also easily quote equalities into \lstinline|eval p e = eval p' e'| form with a shared variable pack (so that identical expressions on both sides of the equation don't end up as different variables).
+
+Notice that we have not made any use of \lstinline|Ltac|, Coq's tactic language.
+Instead, we have used instance resolution as a unification-based programming language to steer the unifier into inferring the symbolic quotation.
+
 %A similar method is used in Coq's notation mechanism.
@@
 
-\begin{lstlisting}
-Global Program Instance quote_new_var V (v: Vars V) x:
-  Quote v x (singlevar x) | 9 := { quote := Var (inr tt) }.
-\end{lstlisting}
-
-We have defined a light-weight quoting function. The following code turns the goal into \lstinline|(eval variable_pack quote)|.
-Here \lstinline|eval_quote'| is a modification of \lstinline|eval_quote| which works in the empty context:
-\begin{lstlisting}
-Goal Π x y (P: Value $\to$ Prop), P ((x * y) * (x * 0)).
-  intros. rewrite <- (eval_quote' _).
-\end{lstlisting}
-
 We have used this technique to implement a tactic to normalize monoid expressions, and plan to extend yada yada.
@@
 
-\subsection{Big operators}
+\section{Big operators}
 Canonical structures have been used to provide a uniform treatment~\cite{bertot2008canonical} of big

### 2010-06-24 jar.tex
Minor improvements.

@@
+% \documentclass{mscs}
 \documentclass[a4paper,10pt,runningheads]{llncs}
@@
   % maybe have a chapter called "interfaces are primary, not implementations" (also involve polynomials)
+  % mention that we use same RingPlus/RingMult operational type classes (with notation) in SemiRing/Ring.
 
@@
 \title{The algebraic hierarchy in type theory using type classes}
-\author{Bas Spitters \and Eelis van der Weegen}
+\author{Bas Spitters \and Eelis van der Weegen} % For MSCS \\ Radboud University Nijmegen}
 \institute{Radboud University Nijmegen}
@@
 
-\paragraph{Equality, setoids, and rewriting}
+% \paragraph{Equality, setoids, and rewriting}
 
@@
 \end{quote}
-% todo: better example
+(Please forgive the admittedly silly statement; we are presently concerned only with the syntax.)
 
@@
 Hence, if we want notations, what we really need are canonical names for these components.
-This is easily accomplished with single-field type classes containing one component each, which we will call \emph{operational type classes}\footnote{These single-field type classes are used in the same way in the \lstinline|Clean| standard library \cite{something}.}:
+This is easily accomplished with single-field type classes containing one component each, which we will call \emph{operational type classes}\footnote{These single-field type classes are used in the same way in the \lstinline|Clean| standard library~\cite{clean}.}:
 \begin{lstlisting}
@@
 \begin{lstlisting}
-  Instance decide_conj `{Decision P} `{Decision Q}: Decision (P ∧ Q).
-\end{lstlisting}
-With these in place, we can just say \lstinline|decide (x = y ∧ p = q)| and let instance resolution automatically compose a decision procedure that can decide the specified proposition.
+  Instance decide_conj `{Decision P} `{Decision Q}: Decision (P ∧$$ Q).
+\end{lstlisting}
+With these in place, we can just say \lstinline|decide (x = y ∧$ $ p = q)| and let instance resolution automatically compose a decision procedure that can decide the specified proposition.
 This style of syntax-directed implicit composition of algorithms is very convenient and highly expressive.
@@
 Thus, axiomatic structural properties and inheritance have precisely the same status as separately proved structural properties and inheritance, reflecting natural mathematical ideology.
-Again, contrast this with bundled approaches, where axiomatic inheritance relations determine projection paths, and where additional inheritance relations lead to additional and ambiguous projection paths for the same operations.
+Again, contrast this with bundled approaches, where axiomatic inheritance relations determine projection paths, and where additional inheritance relations require rebundling and lead to additional and ambiguous projection paths for the same operations.
 
@@
 Definition merge {A B} (a: Vars A) (b: Vars B): Vars (A+B) :=
-  λ i => if i then a j else b j.
+  λ i => match i with inl j => a j | inr j => b j end.
 \end{lstlisting}
@@
     { quote: Expr (V + V')
-    ; eval_quote: @eval (V+V') (merge l r) quote = n }.
+    ; eval_quote: eval (merge l r) quote = n }.
 \end{lstlisting}
@@
 \section{Conclusions}
+\label{conclusion}
 
 While bundling operational and propositional components of abstract structures into records may seem natural at first, doing so actually introduces many serious problems.
-Keeping things unbundled avoids these problems and is more flexible in theory, and now with type classes finally also in practice, or so we hope to have demonstrated.
+Keeping things unbundled avoids these problems and is more flexible in theory, and now with type classes finally also in practice---or so we hope to have demonstrated.
 
@@
 Unification hints are a more general mechanism than type classes, and could provide a more precise account of the interaction between implicit argument inference and proof search.
-Perhaps, then, a fruitful approach could be to use unification hints as the underlying mechanism with type classes as an end-user interface encapsulating a particularly convenient idiom for using them.
+Perhaps a fruitful approach could be to use unification hints as the underlying mechanism with type classes as an end-user interface encapsulating a particularly convenient idiom for using them.
 % We encourage their inclusion to the Coq system. -- Eelis: not quite ready to go there yet
@@
 %Jeremy, Thierry
+% MSCS recommends harvard, but does not specify which bst file to use.
 \bibliographystyle{plain}

### 2010-06-25 jar.tex
Fix figure reference.
Citations, references, wording, ...

@@
 \begin{document}
-\title{The algebraic hierarchy in type theory using type classes}
+\title{Type Classes for Mathematics in Type Theory}
 \author{Bas Spitters \and Eelis van der Weegen} % For MSCS \\ Radboud University Nijmegen}
@@
 To support mathematically sound abstract interfaces for $\N$, $\Z$, and $\Q$, our formalization includes portions of category theory and multisorted universal algebra.
-Algebra flourishes by the interplay between syntax and semantics. The Prolog-like
-abilities of type class unification allow us to conveniently define a quote function thus facilitating the use of reflective techniques.
+Algebra flourishes by the interplay between syntax and semantics.
+The Prolog-like abilities of type class unification allow us to conveniently define a quote function thus facilitating the use of reflective techniques.\marginpar{Do TODOs}
   % todo: handle.
@@
 In section~\ref{predicateclasses} we show how type classes can make the use of ``unbundled'' purely predicate based interfaces for abstract structures practical.
-Next, in section~\ref{hierarchy}, we discuss our algebraic hierarchy implemented using such predicate classes.
+Next, in section~\ref{classes}, we discuss our algebraic hierarchy implemented using such predicate classes.
 
@@
 
-For an interesting (if fairly restrictive) alternative approach based on the idea of requiring unicity of representation for all objects so that Leibniz equality is always applicable, see the algebraic hierarchy part of the ssreflect\cite{ssreflect} libraries.
+For an interesting (if fairly restrictive) alternative approach based on the idea of requiring unicity of representation for all objects so that Leibniz equality is always applicable, see the algebraic hierarchy part of the ssreflect~\cite{Packed} libraries.
 
@@
 
-% \subsection{Type classes}
-% Implementation in Ltac, Notation etc\\
-% Coercions\\
-% Unification hints
-
 \section{Bundling is bad}\label{bundling}
@@
 Higher structures, too, can all be expressed as predicates (expressing laws) over a number of carriers, relations, and operations.
-While optimally flexible in principle, in practice \emph{naive} adoption of this approach (i.e.
-without using type classes) leads to substantial inconveniences in actual use: when \emph{stating} theorems about abstract instances of such structures, one must enumerate all components along with the structure (predicate) of interest.
+While optimally flexible in principle, in practice \emph{naive} adoption of this approach (i.e.\ without using type classes) leads to substantial inconveniences in actual use: when \emph{stating} theorems about abstract instances of such structures, one must enumerate all components along with the structure (predicate) of interest.
 And when \emph{applying} such theorems, one must either enumerate any non-inferrable components, or let the system spawn awkward metavariables to be resolved at a later time.
 Importantly, this also hinders proof search for proofs of the structure predicates, making any nontrivial use of theorems a laborious experience.
-Finally, the lack of \emph{canonical names} for particular components of abstract structures makes it impossible to associate idiomatic notations with them.
+Finally, the lack of \emph{canonical names} for particular components of abstract structures makes it impossible to associate idiomatic notations with them.\marginpar{something about opacity?}
 
@@
 
-% hm, say something about opacity
-
-In the absence of type classes, these are all very real problems, and for this reason the two biggest formalizations of abstract algebraic structures in Coq today (namely CoRN\ref{corn} and Ssreflect\ref{ssreflect}), both use \emph{bundled} representation schemes, using records with one or more of the components as fields instead of parameters.
+In the absence of type classes, these are all very real problems, and for this reason the two biggest formalizations of abstract algebraic structures in Coq today (namely CoRN\ref{C-corn} and Ssreflect\ref{Packed}), both use \emph{bundled} representation schemes, using records with one or more of the components as fields instead of parameters.
 For reflexive relations, the following is a fully bundled representation---the other end of the spectrum:
@@
 
-Manifest fields~\ref{manifestrecords} have been proposed and implemented in the Matita system to address exactly this problem.
+Manifest fields~\ref{DBLP:conf/types/Luo08} have been proposed and implemented in the Matita system to address exactly this problem.
 We hope to convince the reader that type system extensions of this nature, designed to mitigate particular symptoms of the bundled approach, are less elegant than a solution that avoids the problem altogether by using predicate-like type classes in place of bundled records.
@@
 
-\section{Predicate classes and operational classes}
+\section{Predicate classes and operational classes}\label{predicateclasses}
 
@@
 \begin{figure}
-\label{inheritancediagram}
 \centering
@@
 \caption{Inheritance relations established by \lstinline|:>| fields.}
+\label{inheritancediagram}
 \end{figure}
@@
 
-% todo: problem with ambiguous projection paths: quoting will not notice convertability.
+% todo: problem with ambiguous projection paths: quoting will not notice convertibility.
 
@@
 \end{lstlisting}
-In order to use this with \lstinline|Initial|, we define an additional conversion instance that takes a \lstinline|NaturalsToSemiRing| along with a proof showing that it yields \lstinline|SemiRing_Morphism|s, and builds an \lstinline|InitialArrow| instance out of it invoking another conversion function that translates concrete \lstinline|SemiRing_Morphism| proofs into univeral algebra \lstinline|Homomorphism|s instantiated with the semiring signature, which make up the arrows in the category.
+In order to use this with \lstinline|Initial|, we define an additional conversion instance that takes a \lstinline|NaturalsToSemiRing| along with a proof showing that it yields \lstinline|SemiRing_Morphism|s, and builds an \lstinline|InitialArrow| instance out of it invoking another conversion function that translates concrete \lstinline|SemiRing_Morphism| proofs into univeral algebra \lstinline|Homomorphism|s instantiated with the semiring signature, which make up the arrows in the category.\marginpar{Zin te lang}
 
@@
 Basing theory and programs on this abstract interface instead of on specific implementation (such as the ubiquitous Peano naturals \lstinline|nat| in the Coq standard library) is not only cleaner mathematically, but also facilitates easy swapping between implementations.
-And this benefit is far from theoretical, as diverse and non-contrived representations of the natural numbers are abound (unary, binary, factor multisets, arrays of native machine words, etc.).
-
-Some theory and programs about naturals are more easily (or can only be) proved (resp.
-defined) for concrete implementations and then ``lifted'' to the abstract interface.
+And this benefit is far from theoretical, as diverse representations of the natural numbers are abound; for instance, unary, binary, factor multisets, arrays of native machine words.
+
+Some theory and programs about naturals are more easily proved, respectively defined, for concrete implementations and then ``lifted'' to the abstract interface.
 For relatively simple things like decidability of equality, all that is needed to make this lifting easy is a handful of simple lemmas translating the initiality property back into statements saying that \lstinline|naturals_to_semiring| is extensionally equal to other semiring morphisms and involutive between \lstinline|Naturals| implementations.
-For the example of decidability, these let us define decidability for an arbitrary \lstinline|Naturals| implementation simply by mapping to the \lstinline|nat| implementation using \lstinline|naturals_to_semiring|, deciding equality there, and reflecting the equality or inequality proof back to the abstract implementation using involutivity of \lstinline|naturals_to_semiring|.
+For the example of decidability, these let us define decidability for an arbitrary \lstinline|Naturals| implementation simply by mapping to the \lstinline|nat| implementation using \lstinline|naturals_to_semiring|, deciding equality there, and reflecting the equality or inequality proof back to the abstract implementation using involutivity of \lstinline|naturals_to_semiring|.\marginpar{Word: Involutive?}
 
@@
 Combined with the trivially proved property that any two initial objects in a category are isomorphic, this lets us tranfer proofs of such statements between implementations of \lstinline|Naturals|, requiring only that we reflect the concrete statement (expressed in terms of the operational type classes) to a symbolic statement in the language of semirings.
-We intend to eventually make this reflection completely automatic using type class based quotation techniques along the lines of those described in~\ref{quoting}.
+We intend to eventually make this reflection completely automatic using type class based quotation techniques along the lines of those described in Section~\ref{quote}.
 
 Thanks to our close integration of universal algebra we can actually obtain a \lstinline|Naturals| implementation completely automatically, by invoking a generic construction of initial models built from the closed term algebra for the signature along with a setoid equality expressing the congruence closure of the identities in the equational theory.
-However, this implementation is very useful neither in terms of efficiency, nor as a canonical implementation (to be used as the basis for theory and programs that are then subsequently lifted), because for example deciding equality for the aforementioned setoid equality is far harder than for, say, \lstinline|nat| (because it requires a normalization procedure).
+However, this implementation is not very useful, neither in terms of efficiency, nor as a canonical implementation (to be used as the basis for theory and programs that are then subsequently lifted).
+For example, defining a normalization procedure to decide the aforementioned setoid equality is far harder than deciding equality for, say, \lstinline|nat|.
 
@@
 The generic \lstinline|Decision| instance for \lstinline|Naturals| equality implemented by mapping to \lstinline|nat| will typically be far less efficient than a specialized implementation for a particular representation of the natural numbers.
-Fortunately, with Coq's type classes it is no problem for instances overlapping in this way to co-exist, and we can even deprioritize the generic instance (using a numeric priority annotation) so that instance resolution will always pick the specialization when the representation is known.
+Fortunately, with Coq's type classes it is no problem for instances overlapping in this way to co-exist.
+We can even deprioritize the generic instance so that instance resolution will always pick the specialization when the representation is known.
 
@@
 
-Other operations on natural numbers that are sensible choices for specialization are subtraction, distance, division/multiplication by 2, etc.
+Other examples of operations on natural numbers that are sensible choices for specialization include: subtraction, distance, binary division and multiplication.
 
@@
 
-The rationals are characterized as a decidable field with an injective ring morphism from a canonical implementation of the integers and for which division of such integers is surjective:
+The rationals are characterized as a decidable field with an injective ring morphism from a canonical implementation of the integers and a surjection of fractions of such integers:
 \begin{lstlisting}
@@
 Here, \lstinline|Z| is an \lstinline|Integers| implementation paramerized by a \lstinline|Naturals| implementation, for which we just take \lstinline|nat|.
-The choice of \lstinline|Z nat| here is immaterial; we could have picked any other (or even an explicitly arbitrary) implementation of \lstinline|Integers|, but doing so would provide no benefit.
+The choice of \lstinline|Z nat| here is immaterial; we could have picked another, or even a generic, implementation of \lstinline|Integers|, but doing so would provide no benefit.
   % todo: hm, this Z is poorly named. it's really natpair_integers.Z
@@
 
-In our development we prove thas the standard library's default rationals do indeed implement \lstinline|Rationals|, as do implementations of the \lstinline|QType| module interface.
-While the latter is rather ad-hoc from a theoretical perspective, it is nevertheless of great practical interest because it is used for the very efficient \lstinline|BigQ| rationals based on machine integers~\cite{something}.
+In our development we prove that the standard library's default rationals do indeed implement \lstinline|Rationals|, as do implementations of the \lstinline|QType| module interface.
+While the latter is rather ad-hoc from a theoretical perspective, it is nevertheless of great practical interest because it is used for the very efficient \lstinline|BigQ| rationals based on machine integers~\cite{machineintegers}.
 Hence, theory and programs developed on our \lstinline|Rationals| interface applies and can make immediate use of these efficient rationals.
@@
 
-We also plan to do the exact same thing for polynomials, which can be abstractly specified as a free commutative algebra, and for which diverse implementations exist (coefficient lists, Bernstein polynomials, etc.).
+We plan to provide an abstract interface for polynomials as a free commutative algebra. This would unify existing implementations such as coefficient lists and Bernstein polynomials.
 
@@
 al.
-show that unification hints, a very general way of facilitating user-defined extensions to term and type inference, can be used to semi-automatically build quote functions without dropping to a meta-level.\footnote{Gonthier provides similar functionality using ingenious use of canonical structures.} This feature is absent from Coq, but fortunately type classes let us do very much the same thing, as we will now show.
+show that unification hints, a very general way of facilitating user-defined extensions to term and type inference, can be used to semi-automatically build quote functions without dropping to a meta-level.\footnote{Gonthier provides similar functionality by ingeniously using canonical structures.} This feature is absent from Coq, but fortunately type classes also allow us to do this, as we will now show.
 
@@
 \end{lstlisting}
-The expression ``1'' can be quoted in any context \lstinline|(V, v)|, introduces no new variables, and the symbolic term representing it is just \lstinline|One|.
-(The \lstinline|eval_quote| field is turned into a trivial proof obligation.)
+The expression `1' can be quoted in any context \lstinline|(V, v)|, introduces no new variables, and the symbolic term representing it is just \lstinline|One|.
+The \lstinline|eval_quote| field is turned into a trivial proof obligation.
 
@@
 To make this distinction, we need to be able to look up expressions in variable heaps to see if they're already there.
-Importantly, we must do this not by really comparing the values they evaluate to, but by actually browsing the term denoting the variable heap (that is, a composition from \lstinline|novars|, \lstinline|singlevar|, and \lstinline|merge|).
+Importantly, we must do this not by comparing the values they evaluate to, but by actually browsing the term denoting the variable heap --- that is, a composition from \lstinline|novars|, \lstinline|singlevar|, and \lstinline|merge|.
 This, too, is a job for a type class:
@@
 \end{lstlisting}
-Next (and last), if an expression can be looked up in a pack, then it can also be looked up when that pack is merged with another pack:
+Finally, if an expression can be looked up in a pack, then it can also be looked up when that pack is merged with another pack:
 \begin{lstlisting}
@@
 
-We have used this technique to implement a tactic to normalize monoid expressions, and plan to extend yada yada.
+% We have used this technique to implement a tactic to normalize monoid expressions, and plan to extend yada yada.
 %It would be interesting to use this technique to implement the \lstinline|congruence| tactic, which implements
@@
 Canonical structures have been used to provide a uniform treatment~\cite{bertot2008canonical} of big
-operators (like $\Pi,\sum, \max$). These operators extend a pair of a binary and a 0-ary operation
+operators, like $\Pi,\Sigma, \max$. These operators extend a pair of a binary and a 0-ary operation
 to an $n$-ary operation for any $n$. Categorically, one considers the algebra maps from non-empty
@@
 
+\begin{comment}
 \section{Subset types and proof irrelevance}\label{PI}
@@
 to pack the proofs together with the operations, but it is too early to tell.
-
-% 
+\end{comment}
+
 % \section{Coq wishlist}
@@
 Just a remaining source of the occasional uglyness in declarations.
-%
-
-\section{Conclusions}
-\label{conclusion}
+
+\section{Conclusions}\label{conclusion}
 
 While bundling operational and propositional components of abstract structures into records may seem natural at first, doing so actually introduces many serious problems.
-Keeping things unbundled avoids these problems and is more flexible in theory, and now with type classes finally also in practice---or so we hope to have demonstrated.
+With type classes we avoid these problems by avoiding bundling altogether.
 
@@
 
-It has been suggested that canonical structures are more robust (in the sense of being much more restricted and internally closed compared to the comparatively wild and open-ended proof search used for instance resolution).
-However, in addition to the unfortunate fact that the restrictions force one into bundled representations, a second consequence is that using them for anything beyond basic use cases requires significant ingenuity, whereas type classes are fairly easy and straightforward to use.
+It has been suggested that canonical structures are more robust by being more restricted compared to the wild and open-ended proof search of instance resolution.
+However, these restrictions force one into bundled representations, and, moreover, their more advanced usage requires significant ingenuity, whereas type classes usage is straightforward.
 
@@
 In more complex parts of our development we are now experiencing increasingly serious efficiency problems, despite having already made sacrifices by artifically inhibiting many natural class instances in order not to further strain instance resolution.
-
 %While we have already found the predicate-class-based style of formalization to work very well for our development thus far, and believe it has the potential to be \emph{the} modern solution to this problem, in our work we have encountered a number of limitations in the Coq implementation of key features of the tool, that keep us from making an unconditional recommendation for universal adoption of these techniques.
-
-Fortunately, the Coq development team recognizes that this is a real problem, and furthermore agrees that there is plenty of potential for improvement of the current implementation, potentially borrowing techniques from the vast literature on efficient implementation of Prolog-style resolution (which the hint-based proof search used for instance resolution greatly resembles).
+Fortunately, there is plenty of potential for improvement of the current implementation.
+One source is the vast literature on efficient implementation of Prolog-style resolution, which the hint-based proof search used for instance resolution greatly resembles.
 
@@
 
-An obvious topic for future research is the extension from equational logic to partial Horn
-logic~\cite{palmgren2007partial}. Another topic would be to fully, but practically, embrace the
+An obvious topic for future research is the extension from equational logic with dependent types~\cite{Cartmell,palmgren2007partial}. Another topic would be to fully, but practically, embrace the
 categorical approach to universal algebra~\cite{pitts2001categorical}.
 
-According to \lstinline|coqwc|, our development consists of 4895 lines of
-specifications and 769 lines of proofs.\marginpar{update,targz}
+According to \lstinline|coqwc|, our development consists of 5660 lines of specifications and 937 lines of proofs.
 
@@
 % I understand that this is the algorithm underlying congruence and perhaps first-order.
-% 
-% Type classes vs soft typing (Mizar types).
-% Type classes and types in homalg?
 
@@
 This work has been partially funded by the FORMATH project, nr.\ 243847, of the FET program within the 7th Framework program of the European Commission.
-%Jeremy, Thierry
 % MSCS recommends harvard, but does not specify which bst file to use.

### 2010-06-25 jar.tex
fixing references

@@
 
-In the absence of type classes, these are all very real problems, and for this reason the two biggest formalizations of abstract algebraic structures in Coq today (namely CoRN\ref{C-corn} and Ssreflect\ref{Packed}), both use \emph{bundled} representation schemes, using records with one or more of the components as fields instead of parameters.
+In the absence of type classes, these are all very real problems, and for this reason the two biggest formalizations of abstract algebraic structures in Coq today, CoRN~\cite{C-corn} and Ssreflect~\cite{Packed}, both use \emph{bundled} representation schemes, using records with one or more of the components as fields instead of parameters.
 For reflexive relations, the following is a fully bundled representation---the other end of the spectrum:

### 2010-06-28 jar.tex
Revise abstract to make it more in line with title. Fix typo.

--- 2010-06-25 jar.tex
@@
 \begin{abstract}
-We present a new formalization of the algebraic hierarchy in Coq, exploiting its new type class mechanism to make practical a solution formerly thought infeasible.
-Our approach addresses both traditional challenges as well as new ones resulting from our ambition to build upon this development a library of constructive analysis in which abstraction penalties inhibiting efficient computation are reduced to a minimum.
-To support mathematically sound abstract interfaces for $\N$, $\Z$, and $\Q$, our formalization includes portions of category theory and multisorted universal algebra.
+The introduction of first-class type classes in the Coq system calls for re-examination of the basic interfaces we use for mathematical formalization in type theory.
+% especially since existing solutions fall short from a proof engineering perspective.
+We present a new set of foundational type classes for mathematics that take full advantage of their unique features to make practical a particularly flexible approach formerly thought infeasible.
+Thus, we address both traditional proof engineering challenges as well as new ones resulting from our ambition to build upon this development a library of constructive analysis in which abstraction penalties inhibiting efficient computation are reduced to a minimum.
+
+The base of our development consists of type classes representing a standard algebraic hierarchy, as well as portions of category theory and universal algebra.
+We build upon this foundation a set of mathematically sound abstract interfaces for different kinds of numbers, succinctly expressed using categorical language and universal algebra constructions.
+Strategic use of type classes lets us support these high-level theory-friendly definitions while still enabling efficient implementations unhindered by gratuitous indirection or conversion or projection.
+
 Algebra flourishes by the interplay between syntax and semantics.
-The Prolog-like abilities of type class unification allow us to conveniently define a quote function thus facilitating the use of reflective techniques.\marginpar{Do TODOs}
-  % todo: handle.
+The Prolog-like abilities of type class instance resolution allow us to conveniently define a quote function, thus facilitating the use of reflective techniques.
 \end{abstract}
@@
 Some theory and programs about naturals are more easily proved, respectively defined, for concrete implementations and then ``lifted'' to the abstract interface.
-For relatively simple things like decidability of equality, all that is needed to make this lifting easy is a handful of simple lemmas translating the initiality property back into statements saying that \lstinline|naturals_to_semiring| is extensionally equal to other semiring morphisms and involutive between \lstinline|Naturals| implementations.
+For relatively simple things like decidability of equality, all that is needed to make this lifting easy is a handful of simple lemmas translating the initiality property back into statements saying that \lstinline|naturals_to_semiring| is extensionally equal to other semiring morphisms and involutary between \lstinline|Naturals| implementations.
 For the example of decidability, these let us define decidability for an arbitrary \lstinline|Naturals| implementation simply by mapping to the \lstinline|nat| implementation using \lstinline|naturals_to_semiring|, deciding equality there, and reflecting the equality or inequality proof back to the abstract implementation using involutivity of \lstinline|naturals_to_semiring|.\marginpar{Word: Involutive?}

### 2010-06-28 jar.tex
Break up overly long sentence.

@@
 \end{lstlisting}
-In order to use this with \lstinline|Initial|, we define an additional conversion instance that takes a \lstinline|NaturalsToSemiRing| along with a proof showing that it yields \lstinline|SemiRing_Morphism|s, and builds an \lstinline|InitialArrow| instance out of it invoking another conversion function that translates concrete \lstinline|SemiRing_Morphism| proofs into univeral algebra \lstinline|Homomorphism|s instantiated with the semiring signature, which make up the arrows in the category.\marginpar{Zin te lang}
+In order to use this with \lstinline|Initial|, we define an additional conversion instance that takes a \lstinline|NaturalsToSemiRing| along with a proof showing that it yields \lstinline|SemiRing_Morphism|s, and builds an \lstinline|InitialArrow| instance out of it.
+This conversion instance in turn invokes another conversion function that translates concrete \lstinline|SemiRing_Morphism| proofs into univeral algebra \lstinline|Homomorphism|s instantiated with the semiring signature, which make up the arrows in the category.
 

### 2010-06-28 jar.tex
Replace some "i.e."'s. I'm less eager to replace "e.g."'s.

@@
 
-For the algebraic hierarchy---a critical component in any library of formalized mathematics---these challenges include structure inference, handling of multiple inheritance, idiomatic use of notations, and convenient algebraic manipulation (e.g.
-rewriting).
+For the algebraic hierarchy---a critical component in any library of formalized mathematics---these challenges include structure inference, handling of multiple inheritance, idiomatic use of notations, and convenient algebraic manipulation (e.g.\ rewriting).
 
@@
 Coq features a distinguished sort called \lstinline|Prop| that one may voluntarily choose to use as the sort for types representing propositions.
-The distinguishing feature of the \lstinline|Prop| sort is that Coq will not permit terms of non-\lstinline|Prop| types to depend on the values of inhabitants of \lstinline|Prop| types (i.e.
-proof terms).
+The distinguishing feature of the \lstinline|Prop| sort is that Coq will not permit terms of non-\lstinline|Prop| types to depend on the values of inhabitants of \lstinline|Prop| types (that is, proof terms).
 
@@
 Higher structures, too, can all be expressed as predicates (expressing laws) over a number of carriers, relations, and operations.
-While optimally flexible in principle, in practice \emph{naive} adoption of this approach (i.e.\ without using type classes) leads to substantial inconveniences in actual use: when \emph{stating} theorems about abstract instances of such structures, one must enumerate all components along with the structure (predicate) of interest.
+While optimally flexible in principle, in practice \emph{naive} adoption of this approach (that is, without using type classes) leads to substantial inconveniences in actual use: when \emph{stating} theorems about abstract instances of such structures, one must enumerate all components along with the structure (predicate) of interest.
 And when \emph{applying} such theorems, one must either enumerate any non-inferrable components, or let the system spawn awkward metavariables to be resolved at a later time.
@@
 A second problem with bundling is that as the bundled records are stacked to represent higher and higher structures, the projection paths for their components grow longer and longer, resulting in ever more unwieldy terms (coercions and notations can make this less painful).
-Further, if one tries to implement some semblance of sharing in a bundled representation, these projection paths additionally become non-canonical, and still more extensions have been proposed to address this symptom (e.g.
-coercion pullbacks~\cite{Hints}).
+Further, if one tries to implement some semblance of sharing in a bundled representation, these projection paths additionally become non-canonical, and still more extensions have been proposed to address this symptom (e.g.\ coercion pullbacks~\cite{Hints}).
 
@@
 We ought to say a few words about our use of \lstinline|fmap|.
-The usual mathematical notational convention for functor application is to use the name of the functor to refer to both its object map and its arrow map, relying on additional conventions regarding object/arrow names for disambiguation (i.e.
-``\lstinline|F x|'' and ``\lstinline|F f|'' map an object and an arrow, respectively, because ``\lstinline|x|'' and ``\lstinline|f|'' are conventional names for objects and arrows, respectively.
+The usual mathematical notational convention for functor application is to use the name of the functor to refer to both its object map and its arrow map, relying on additional conventions regarding object/arrow names for disambiguation: ``\lstinline|F x|'' and ``\lstinline|F f|'' map an object and an arrow, respectively, because ``\lstinline|x|'' and ``\lstinline|f|'' are conventional names for objects and arrows, respectively.
 

### 2010-06-28 jar.tex
Add paragraph about opacity and unpredictability of instance resolution.

@@
 
-\lstinline|Decision| is a very general purpose type class that also works for predicates.
+\lstinline|Decision| is a very general-purpose type class that also works for predicates.
 For instance, to declare a parameter expressing decidability of, say, (setoid) equality on a type \lstinline|X|, we write: \lstinline|`{Π a b: X, Decision (a = b)}|.
@@
 There, even \emph{strong} proof irrelevance (which would make terms convertible that differ only in what proofs they refer to) would not make these difficulties go away entirely, because high-level tactics that rely on quotation of terms require syntactic identity (rather than ``mere'' convertibility) to recognize identical subterms.
+
+Because predicate classes only provide contextual information and are insulated from the actual algebraic expressions, their instances can always be kept entirely opaque---only their existence matters.
+Together, these properties largely defuse an argument occasionally voiced against type classes concerning perceived unpredictability of instance resolution.
+While it is certainly true that in contexts with redundant information it can become hard to predict which instance of a predicate class will be found by proof search, this is simply a non-issue because it \emph{does not matter} which one is found.
 

### 2010-06-28 jar.tex
In same paragraph, add remark about operational type classes.
Add paragraph about re-use of operational type classes. Fix some code whitespace formatting.

@@
   % maybe have a chapter called "interfaces are primary, not implementations" (also involve polynomials)
-  % mention that we use same RingPlus/RingMult operational type classes (with notation) in SemiRing/Ring.
 
@@
 Importantly, this also hinders proof search for proofs of the structure predicates, making any nontrivial use of theorems a laborious experience.
-Finally, the lack of \emph{canonical names} for particular components of abstract structures makes it impossible to associate idiomatic notations with them.\marginpar{something about opacity?}
+Finally, the lack of \emph{canonical names} for particular components of abstract structures makes it impossible to associate idiomatic notations with them.
 
@@
 While it is certainly true that in contexts with redundant information it can become hard to predict which instance of a predicate class will be found by proof search, this is simply a non-issue because it \emph{does not matter} which one is found.
+And for operational type classes the issue rarely arises because their instances are not nearly as abundant.
 
@@
 Much more so than the existing canonical structures facility, because canonical structures can only be used with bundled representations, which we argued against in section~\ref{bundling}.
+
+The operational type classes used in \lstinline|SemiRing| for zero, one, multiplication and addition, are the same ones used by \lstinline|Ring| and \lstinline|Field| (not shown).
+Thus, the realization that a particular semiring is in fact a ring or field (etc.) has no bearing on how one refers to the operations in question, which is as it should be.
+The realization that a particular semigroup is part of a semiring \emph{does} call for a new (canonical) name, simply because of the need for disambiguation.
+The introduction of these additional names for the same operation is quite harmless in practice, because canonical names established by operational type class fields are just trivial identity functions anyway, so that in most contexts the distinction reduces away instantly.
 
@@
   ; comp_assoc w x y z (a: w ⟶ x) (b: x ⟶ y) (c: y ⟶ z):
-      c ◎ (b ◎ a) = (c ◎ b) ◎ a
-  ; id_l `(a: x ⟶ y): cat_id ◎ a = a
-  ; id_r `(a: x ⟶ y): a ◎ cat_id = a }.
+      c ◎$$ (b ◎$$ a) = (c ◎$$ b) ◎$$ a
+  ; id_l `(a: x ⟶ y): cat_id ◎$$ a = a
+  ; id_r `(a: x ⟶ y): a ◎$$ cat_id = a }.
 \end{lstlisting}
@@
     ; preserves_comp `(f: y ⟶ z) `(g: x ⟶ y):
-      fmap (f ◎ g) = fmap f ◎ fmap g }.
+      fmap (f ◎$$ g) = fmap f ◎$$ fmap g }.
 \end{lstlisting}

### 2010-06-28 jar.tex
Add paragraph about implicit generalization and the functionality we would like to see.

@@
   % cite that other type class based development in the context of the alternative bundling schemes.
-  % maybe have a chapter called "interfaces are primary, not implementations" (also involve polynomials)
 
@@
 
-% todo: mention proof irrelevance here somewhere, and terms that depend on proofs.
-
 The observant reader may wonder whether it might be beneficial to go one step further and unbundle proof of laws and inherited substructures as well.
@@
 Using implicit generalization, we can write:
-
 \begin{lstlisting}
@@
 \end{lstlisting}
-% TODO: describe implicit generalization in two words
-% TODO: at some point down the line in our coq-inadequacies section, describe the implicit generalization improvements we need.
+The backtick tells Coq to insert implicit declarations of further parameters to \lstinline|SemiGroup G|, namely those declared as \lstinline|e| and \lstinline|op| above.
+It further lets us omit a name for the \lstinline|SemiGroup G| parameter itself as well.
+All of these will be given automatically generated names that we will never refer to.
+
 Thus, we have reached the mathematical ideal we aimed for.
+
+While we are on the topic of implicit generalization, we mention one inadequacy concerning their current implementation that we feel should be addressed for the facility to be a completely satisfying solution.
+While the syntax already supports variants (not shown above) that differ in how exactly different kinds of arguments are inferred and/or generalized, there is no support to have an argument be ``inferred if possible, generalized otherwise".
+We have found that the need for such a policy arises naturally when declaring a parameter of class type in a contexts where \emph{some} of its components are already available, while others are to be newly introduced.
+The current workaround in these cases involves providing names for components that are then never referred to, which is a bit awkward.
 

### 2010-06-28 jar.tex
Fix awkward style in abstract.

@@
 The base of our development consists of type classes representing a standard algebraic hierarchy, as well as portions of category theory and universal algebra.
-We build upon this foundation a set of mathematically sound abstract interfaces for different kinds of numbers, succinctly expressed using categorical language and universal algebra constructions.
+On this foundation we build a set of mathematically sound abstract interfaces for different kinds of numbers, succinctly expressed using categorical language and universal algebra constructions.
 Strategic use of type classes lets us support these high-level theory-friendly definitions while still enabling efficient implementations unhindered by gratuitous indirection or conversion or projection.
@@
 In our hierarchy, these ``substructures'' by themselves establish the inheritance diagram in Figure~\ref{inheritancediagram}.
-% todo: for some reason the figure number reference is wrong here..
 

### 2010-06-28 jar.tex
Fix some citations; other fixes.

@@
 Effectively keeping track of, resolving, and combining proofs of equivalence-ness and propriety when the user attempts to substitute a given (sub)term using a given equality, is known as ``setoid rewriting'', and requires nontrivial infrastructure and support from the system.
-The Coq implementation of these mechanisms was largely rewritten in 200?
-by Matthieu Sozeau~\cite{Setoid-rewrite} to make it more flexible and to replace the old special-purpose setoid/morphism registration command with a clean type class based interface.
+The Coq implementation of these mechanisms was largely rewritten by Matthieu Sozeau~\cite{Setoid-rewrite} to make it more flexible and to replace the old special-purpose setoid/morphism registration command with a clean type class based interface.
 
@@
 
-In the absence of type classes, these are all very real problems, and for this reason the two biggest formalizations of abstract algebraic structures in Coq today (namely CoRN\ref{C-corn} and Ssreflect\ref{Packed}), both use \emph{bundled} representation schemes, using records with one or more of the components as fields instead of parameters.
+In the absence of type classes, these are all very real problems, and for this reason the two biggest formalizations of abstract algebraic structures in Coq today (namely CoRN\cite{C-corn} and Ssreflect\cite{Packed}), both use \emph{bundled} representation schemes, using records with one or more of the components as fields instead of parameters.
 For reflexive relations, the following is a fully bundled representation---the other end of the spectrum:
@@
 
-Manifest fields~\ref{DBLP:conf/types/Luo08} have been proposed and implemented in the Matita system to address exactly this problem.
+Manifest fields~\cite{DBLP:conf/types/Luo08} have been proposed and implemented in the Matita system to address exactly this problem.
 We hope to convince the reader that type system extensions of this nature, designed to mitigate particular symptoms of the bundled approach, are less elegant than a solution that avoids the problem altogether by using predicate-like type classes in place of bundled records.
@@
 For relatively simple things like decidability of equality, all that is needed to make this lifting easy is a handful of simple lemmas translating the initiality property back into statements saying that \lstinline|naturals_to_semiring| is extensionally equal to other semiring morphisms and involutary between \lstinline|Naturals| implementations.
-For the example of decidability, these let us define decidability for an arbitrary \lstinline|Naturals| implementation simply by mapping to the \lstinline|nat| implementation using \lstinline|naturals_to_semiring|, deciding equality there, and reflecting the equality or inequality proof back to the abstract implementation using involutivity of \lstinline|naturals_to_semiring|.\marginpar{Word: Involutive?}
+For the example of decidability, these let us define decidability for an arbitrary \lstinline|Naturals| implementation simply by mapping to the \lstinline|nat| implementation using \lstinline|naturals_to_semiring|, deciding equality there, and reflecting the equality or inequality proof back to the abstract implementation using the fact that \lstinline|naturals_to_semiring| is an involution.
 

### 2010-06-28 jar.tex

@@
 
-In the absence of type classes, these are all very real problems, and for this reason the two biggest formalizations of abstract algebraic structures in Coq today (namely CoRN\cite{C-corn} and Ssreflect\cite{Packed}), both use \emph{bundled} representation schemes, using records with one or more of the components as fields instead of parameters.
+In the absence of type classes, these are all very real problems, and for this reason the two biggest formalizations of abstract algebraic structures in Coq today, CoRN~\cite{C-corn} and Ssreflect~\cite{Packed}, both use \emph{bundled} representation schemes, using records with one or more of the components as fields instead of parameters.
 For reflexive relations, the following is a fully bundled representation---the other end of the spectrum:

### 2010-06-29 jar.tex
Stylistic adjustments in conclusion.

@@
 
-It has been suggested that canonical structures are more robust by being more restricted compared to the wild and open-ended proof search of instance resolution.
-However, these restrictions force one into bundled representations, and, moreover, their more advanced usage requires significant ingenuity, whereas type classes usage is straightforward.
+It has been suggested that canonical structures are more robust because of their more restricted nature compared to the wild and open-ended proof search of instance resolution.
+However, these restrictions force one into bundled representations, and moreover, their more advanced usage requires significant ingenuity, whereas type class usage is straightforward.
+Furthermore, wild and open-ended proof search is harmless for predicate classes for which only existence---not identity---matters.
 
@@
 Unification hints are a more general mechanism than type classes, and could provide a more precise account of the interaction between implicit argument inference and proof search.
-Perhaps a fruitful approach could be to use unification hints as the underlying mechanism with type classes as an end-user interface encapsulating a particularly convenient idiom for using them.
+It is not a great stretch to conjecture that a fruitful approach might be to use unification hints as the underlying mechanism, with type classes as an end-user interface encapsulating a particularly convenient idiom for using them.
 % We encourage their inclusion to the Coq system. -- Eelis: not quite ready to go there yet
@@
 %While we have already found the predicate-class-based style of formalization to work very well for our development thus far, and believe it has the potential to be \emph{the} modern solution to this problem, in our work we have encountered a number of limitations in the Coq implementation of key features of the tool, that keep us from making an unconditional recommendation for universal adoption of these techniques.
-Fortunately, there is plenty of potential for improvement of the current implementation.
+Fortunately, there is plenty of potential for improvement of the current instance resolution implementation.
 One source is the vast literature on efficient implementation of Prolog-style resolution, which the hint-based proof search used for instance resolution greatly resembles.
@@
 The use of category theory has been important in these developments.
+% Eelis: not sure i understand.
+"has been"?
+it's not done.
 

### 2010-06-29 jar.tex
Replace bigops section with 'Sequences and universes' section.

@@
 
-\section{Big operators}
+\section{Sequences and universes}
+
+Finite sequences are another example of a concept that can be represented in many different ways: as cons-lists, maps from bounded naturals, array-queues, etc.
+Here, too, the introduction of an abstract interface facilitates implementation independence.
+
+Mathematically, finite sequences can be characterized as free monoids over sets.
+A categorical way of expressing this is in terms of adjunctions.
+As before, we \emph{could} fully embrace this perspective, paying no heed to practicality of implementation and usage, and define a relatively succinct type class for sequences as follows:
+\begin{lstlisting}
+Class PoshSequence
+   (free: setoid.Object → monoid.Object) `{Fmap free}
+   (singleton: id ⇛ monoid.forget ∘ free)
+   (extend: `((x ⟶ monoid.forget y) → (free x ⟶ y))): Prop :=
+   { sequence_adjunction: AltAdjunction singleton extend
+   ; extend_morphism: `(Setoid_Morphism (extend x y)) }.
+\end{lstlisting}
+Here, \lstinline|monoid.forget| is the forgetful functor from monoids to sets.
+
+However, we \emph{do} care about practicality, and so we will again take a more concrete perspective, starting with operational type classes for the characteristic operations:
+\begin{lstlisting}
+  Context `{Functor (seq: Type → Type)}.
+
+  Class Extend := extend: Π {x y} `{SemiGroupOp y} `{MonoidUnit y},
+    (x → y) → (seq x → y).
+  Class Singleton := singleton: Π x, x → seq x.
+\end{lstlisting}
+With these, we can define the predicate class for sequences:
+\begin{lstlisting}
+  Class Sequence
+   `{Π a, MonoidUnit (seq a)} `{Π a, SemiGroupOp (seq a)}
+   `{Π a, Equiv a → Equiv (seq a)} `{Singleton} `{Extend}: Prop := ...
+\end{lstlisting}
+%    { ... structural and propriety proofs sufficient to bundle up the categories and arrows ...
+%    ; ... AltAdjunction stated using the bundled ... }.
+On top of this interface we can build theory about typical list operations such as maps, folds, their relation to \lstinline|singleton| and \lstinline|extend|, et cetera.
+We can also generically define ``bigops'' for summation and products, and easily show properties like distributivity, all without ever mentioning cons lists.
+
+Unfortunately, tragedy strikes when, after having defined this theory, we try to show that regular cons lists implement the abstract \lstinline|Sequence| interface.
+When we get to the point where we want to define the \lstinline|Singleton| operation, Coq emits a universe inconsistency error.
+The problem is that because of the categorical constructions involved, the theory forces \lstinline|Singleton| to inhabit a relatively high universe level, making it incompatible with lowly \lstinline|list|.
+
+Universe polymorphism could in principle most likely solve this problem, but its current implementation in Coq only supports universe polymorphic inductive definitions, while \lstinline|Singleton| is a regular definition.
+Universe polymorphic \emph{definitions} have historically not been supported in Coq, primarily because of efficiency concerns.
+However, we have taken up the issue with the Coq development team, and they have agreed to introduce the means to let one turn on universe polymorphism for definitions voluntarily on a per-definition basis.
+With such functionality we could make \lstinline|Singleton| universe polymorphic, and hopefully resolve these problems.
+
+In other places in our development, too, we have encountered universe inconsistencies that could be traced back to universe monomorphic definitions being forced into disparate universes (\lstinline|Equiv| being a typical example).
+Hence, we consider the support for universe polymorphic definitions that is currently being implemented to be of great importance to the general applicability and scalability of our approach.
+
+% todo: not sure why we didn't have this problem with, say, nat/Naturals.
+
+\begin{comment}
 Canonical structures have been used to provide a uniform treatment~\cite{bertot2008canonical} of big
@@
 \end{lstlisting}
-\marginpar{Prove distributivity?}
+\end{comment}
 
@@
 
-% \section{Coq wishlist}
-% % todo: maybe merge with conclusions and add "and future work"
-% 
-% \paragraph{Universe-polymorphic definitions}
-% Coq has supported universe-polymorphism for parameterized inductive definitions (different instantiations of which are automatically of the lowest possible sort taking into account the universe level constraints on the arguments) for some time.
-However, the same functionality has not been available for \emph{definitions}, primarily because of efficiency concerns.
-In our development, bla bla misery and drama when doing this and that.
-In response to our inquiries about this issue, the Coq development team has agreed to implement ``opt-in'' universe polymorphism, where only expressly annotated definitions are made universe-polymorphic.
-This should solve our problems.
-% 
-% \paragraph{Implicit generalization}
-% 
-% The current syntax for implicit generalization 
-% %current syntax very subtle.
-various combinations of `, class-ness, {} vs (), !
-or no !, make _ mean different things.
-Worse, one strategy of ``infer if possible, generalize otherwise'' is not supported while it seems like a very commonly needed thing.
-The Coq development team recognizes this as a legitimate problem, but fears the implementation might not be straightforward, and as of this writing we are not aware of plans to address this.
-That said, a new proof engine has recently, whcih might make this more attainable.
-Fortunately, not that big a problem.
-Just a remaining source of the occasional uglyness in declarations.
-
 \section{Conclusions}\label{conclusion}
@@
 
+% \paragraph{Universe-polymorphic definitions}
+% Coq has supported universe-polymorphism for parameterized inductive definitions (different instantiations of which are automatically of the lowest possible sort taking into account the universe level constraints on the arguments) for some time.
+However, the same functionality has not been available for \emph{definitions}, primarily because of efficiency concerns.
+In our development, bla bla misery and drama when doing this and that.
+In response to our inquiries about this issue, the Coq development team has agreed to implement ``opt-in'' universe polymorphism, where only expressly annotated definitions are made universe-polymorphic.
+This should solve our problems.
+
 The only real pending concern that keeps us from making an unequivocal endorsement of type classes as a versatile, expressive, and elegant means of organizing proof developments, is instance resolution efficiency.

### 2010-06-29 jar.tex
Note universe polymorphism as a problem in conclusion and refer to relevant section.

@@
 
-The only real pending concern that keeps us from making an unequivocal endorsement of type classes as a versatile, expressive, and elegant means of organizing proof developments, is instance resolution efficiency.
+There are really only two pending concerns that keeps us from making an unequivocal endorsement of type classes as a versatile, expressive, and elegant means of organizing proof developments.
+The first and lesser of the two is universe polymorphism for definitions as described in the previous section.
+The second is instance resolution efficiency.
 In more complex parts of our development we are now experiencing increasingly serious efficiency problems, despite having already made sacrifices by artifically inhibiting many natural class instances in order not to further strain instance resolution.

### 2010-06-30 jar.tex
Complete outline in introduction.

@@
 \begin{abstract}
-The introduction of first-class type classes in the Coq system calls for re-examination of the basic interfaces we use for mathematical formalization in type theory.
+The introduction of first-class type classes in the Coq system calls for re-examination of the basic interfaces used for mathematical formalization in type theory.
 % especially since existing solutions fall short from a proof engineering perspective.
-We present a new set of foundational type classes for mathematics that take full advantage of their unique features to make practical a particularly flexible approach formerly thought infeasible.
+We present a new set of foundational type classes for mathematics and take full advantage of their unique features to make practical a particularly flexible approach formerly thought infeasible.
 Thus, we address both traditional proof engineering challenges as well as new ones resulting from our ambition to build upon this development a library of constructive analysis in which abstraction penalties inhibiting efficient computation are reduced to a minimum.
@@
 % ideally should be more or less readable by Haskell person. that would be neat. so that's what the preliminaries should prepare for.
+In this paper we focus on the Coq proof assistant. We conjecture that the methods can be transferred
+to any type theory based proof assistant supporting type classes such as
+Matita~\cite{asperti2007user}.
 
@@
 Next, in section~\ref{classes}, we discuss our algebraic hierarchy implemented using such predicate classes.
+In sections~\ref{cat} and~\ref{univ} we give a taste of what category theory and universal algebra look like in our development, and in section~\ref{numbers} we use these facilities to build abstract interfaces for numbers.
+In order to illustrate a very different use of type classes, we discuss the implementation of a quoting function for algebraic terms in terms of type classes, in section~\ref{quoting}.
+In section~\ref{sequences} we hint at an interface for sequences but describe how a limitation in the Coq system makes its use problematic.
+We end with conclusions in section~\ref{conclusions}.
 
@@
 With this in place, we continue with a discussion of
-
-In this paper we focus on the Coq proof assistant. We conjecture that the methods can be transferred
-to any type theory based proof assistant supporting type classes such as
-Matita~\cite{asperti2007user}.
 
@@
 However, the current instance resolution implementation has little support for this style of \emph{forward} reasoning, and is really primarily oriented on \emph{backward} reasoning: had we registered \lstinline|monmor_to| and \lstinline|monmor_from| as instance resolution hints, we would in fact be saying ``\emph{if} trying to establish that something is a \lstinline|Monoid|, \emph{then} try finding a \lstinline|Monoid_Morphism| to or from it'', which quickly degenerates into a wild goose chase.
-We will return to this point in section~\ref{conclusion}.
-% todo: do this
+We will return to this point in section~\ref{conclusions}.
 
@@
 
-\section{Numerical interfaces}
+\section{Numerical interfaces}\label{numbers}
 
@@
 Combined with the trivially proved property that any two initial objects in a category are isomorphic, this lets us tranfer proofs of such statements between implementations of \lstinline|Naturals|, requiring only that we reflect the concrete statement (expressed in terms of the operational type classes) to a symbolic statement in the language of semirings.
-We intend to eventually make this reflection completely automatic using type class based quotation techniques along the lines of those described in Section~\ref{quote}.
+We intend to eventually make this reflection completely automatic using type class based quotation techniques along the lines of those described in Section~\ref{quoting}.
 
@@
 
-\section{Quoting with type classes}\label{quote}
+\section{Quoting with type classes}\label{quoting}
 
@@
 \end{lstlisting}
-The expression type is parametrized over the set of variable indices.
+The expression type is parameterized over the set of variable indices.
 Hence, we diverge from~\cite{Hints}, which uses \lstinline|nat| for variables, thereby introducing bound problems and a need for dummy variables.
@@
 
-\section{Sequences and universes}
+\section{Sequences and universes}\label{sequences}
 
@@
 
-\section{Conclusions}\label{conclusion}
+\section{Conclusions}\label{conclusions}
 

### 2010-06-30 jar.tex
Fix TODO naming author of unification-hints paper.
Polishing.

@@
   {∀}{{$\forall$}}1
-  {Π}{{$\Uppi$}}1
+  {Π}{{$\Uppi$ }}1
   {η}{{$\upeta$}}1
@@
   {<-}{{$\leftarrow\ $}}1
-  {~}{{$\sim\ $}}1
+  {~}{{$\sim$}}1
 }
@@
 
-In the absence of type classes, these are all very real problems, and for this reason the two biggest formalizations of abstract algebraic structures in Coq today, CoRN~\cite{C-corn} and Ssreflect~\cite{Packed}, both use \emph{bundled} representation schemes, using records with one or more of the components as fields instead of parameters.
+In the absence of type classes, these are all very real problems, and for this reason the two largest formalizations of abstract algebraic structures in Coq today, CoRN~\cite{C-corn} and Ssreflect~\cite{Packed}, both use \emph{bundled} representation schemes, using records with one or more of the components as fields instead of parameters.
 For reflexive relations, the following is a fully bundled representation---the other end of the spectrum:
@@
 Manifest fields~\cite{DBLP:conf/types/Luo08} have been proposed and implemented in the Matita system to address exactly this problem.
-We hope to convince the reader that type system extensions of this nature, designed to mitigate particular symptoms of the bundled approach, are less elegant than a solution that avoids the problem altogether by using predicate-like type classes in place of bundled records.
+We hope to convince the reader that type system extensions of this nature, designed to mitigate particular symptoms of the bundled approach, are less elegant than a solution (described in the next section) that avoids the problem altogether by using predicate-like type classes in place of bundled records.
 
@@
 
-The observant reader may wonder whether it might be beneficial to go one step further and unbundle proof of laws and inherited substructures as well.
+The observant reader may wonder whether it might be beneficial to go one step further and unbundle proofs of laws and inherited substructures as well.
 This is not the case, because there is no point in sharing them.
@@
 \end{lstlisting}
-(The curly brackets mark \lstinline|P| as an implicit argument.) Because \lstinline|e| and \lstinline{op} are freshly introduced local names, we cannot bind notations to them prior to this theorem.
+Because \lstinline|e| and \lstinline{op} are freshly introduced local names, we cannot bind notations to them prior to this theorem.
 Hence, if we want notations, what we really need are canonical names for these components.
@@
   Theorem L G e op {P: SemiGroup G e op}: Π x y z,
-    equiv e (sg_op op (sg_op op x y) z) (sg_op op (sg_op op x y) z).
+    equiv e (sg_op op (sg_op op x y) z) (sg_op op (sg_op op z y) x).
 \end{lstlisting}
@@
 While the syntax already supports variants (not shown above) that differ in how exactly different kinds of arguments are inferred and/or generalized, there is no support to have an argument be ``inferred if possible, generalized otherwise".
-We have found that the need for such a policy arises naturally when declaring a parameter of class type in a contexts where \emph{some} of its components are already available, while others are to be newly introduced.
+We have found that the need for such a policy arises naturally when declaring a parameter of class type in a context where \emph{some} of its components are already available, while others are to be newly introduced.
 The current workaround in these cases involves providing names for components that are then never referred to, which is a bit awkward.
@@
 
-Before we proceed with a discussion of the algebraic hierarchy based on type classes, we briefly highlight one specific operational type class because we will use it later and because it is a particularly nice illustration of another neat application of operational type classes.
+Before we proceed with a discussion of the algebraic hierarchy based on predicate classes and operational classes, we briefly highlight one specific operational type class because we will use it later and because it is a particularly nice illustration of another neat application of operational type classes.
 The operation in question is that of deciding a proposition:
@@
 Fields declared with \lstinline|:>| are registered as hints for instance resolution, so that in any context where \lstinline|(A, =, +, 0, *, 1)| is known to be a \lstinline|SemiRing|, \lstinline|(A, =, +, 0)| and \lstinline|(A, =, *, 1)| are automatically known to be \lstinline|CommutativeMonoid|s (and so on, transitively, because instance resolution is recursive).
-In our hierarchy, these ``substructures'' by themselves establish the inheritance diagram in Figure~\ref{inheritancediagram}.
-
-\begin{figure}
-\centering
+In our hierarchy, these ``substructures'' by themselves establish the following inheritance diagram:
+
+\vspace{5mm}
+\begin{center}
 \includegraphics{hierarchy.pdf}
-\caption{Inheritance relations established by \lstinline|:>| fields.}
-\label{inheritancediagram}
-\end{figure}
+\end{center}
+\vspace{5mm}
 
@@
 
-Having described the basic principles of our approach, in the remainder of this paper we tour other parts of our development, further illustrating what a state of the art formal development of foundational mathematical structures can look like with a modern proof assistant.
+Having described the basic principles of our approach, in the remainder of this paper we tour other parts of our development, further illustrating what a state of the art formal development of foundational mathematical structures can look like with a modern proof assistant based on type theory.
 
@@
 The \lstinline|Initial| class itself further requires these ``initial arrows'' to be unique.
-Having \lstinline|InitialArrows| as a type class means that we can always simply say \lstinline|initial_arrow y| whenever \lstinline|y| is known to be an object in a category known to have an initial object (where ``known to'' should be read as ``can be determined by instance resolution'').
+Having \lstinline|InitialArrows| as an operational type class means that we can always simply say \lstinline|initial_arrow y| whenever \lstinline|y| is known to be an object in a category known to have an initial object (where ``known'' should be read as ``can be determined by instance resolution'').
 
@@
 
-  Class Functor `(Fmap): Prop :=
+  Class Functor `{Fmap}: Prop :=
     { functor_from: Category C
@@
 In addition to not being supported by Coq, option (1) would violate our policy of leaving components unbundled.
-For (2), if it could be made to work at all (which is not clear at all), F would need a pretty egregious type considering that arrow types are indexed by objects, and that the type of the arrow map (namely ``\lstinline|Π x y, (x ⟶ y) → (F x ⟶ F y)|'') must refer to the object map.
-
-We feel that these issues are not limitations of the Coq system, but merely reflect the fact that notationally identifying these two distinct and interdependent maps is an abuse of notation of sufficient severety to make it ill-suited to a formal development where software engineering concerns apply.
+For (2), if it could be made to work at all (which is not clear at all), F would need a pretty egregious type considering that arrow types are indexed by objects, and that the type of the arrow map (namely \lstinline|Π x y, (x ⟶ y) → (F x ⟶ F y)|) must refer to the object map.
+
+We feel that these issues are not limitations of the Coq system, but merely reflect the fact that notationally identifying these two distinct and interdependent maps is an abuse of notation of sufficient severity to make it ill-suited to a formal development where software engineering concerns apply.
 Hence, we do not adopt this practice, and use ``\lstinline|fmap F|'' (name taken from the Haskell standard library) to refer to the arrow map of a functor \lstinline|F|.
@@
 To specify the natural numbers and the integers as initial objects in the categories of semirings and rings, respectively, definitions of these categories are needed.
-While one could define both of them ``manually'', greater economy can be achieved by recognizing that both semirings and rings can be defined by equational theories, \emph{varieties} (that is, categories consisting of models for a fixed theory, with homomorphisms between them) of which can be defined generically.
+While one could define both of them ``manually'', greater economy can be achieved by recognizing that both semirings and rings can be defined by equational theories, for which \emph{varieties} can be defined generically (varieties are categories consisting of models for a fixed theory with homomorphisms between them).
 
@@
 
-  Class AlgebraOps (σ: Signature) (carriers: sorts σ → Type)
+  Class AlgebraOps (σ: Signature) (carriers: sorts σ$$ → Type)
     := algebra_op: Π o: operation σ, op_type carriers (σ o).
@@
 An \lstinline|EqEntailment| consists of premises and a conclusion represented by an inductively defined statement grammar, which in turn uses an inductively defined term grammar.
-A detailed discussion of these definitions and the theory developed for them is beyond the scope of this paper,
+A detailed discussion of these definitions and the theory developed for them is beyond the scope of this paper.
 
@@
     { variety_algebra:> Algebra et carriers
-  ; variety_laws: Π s, et_laws et s → (Π vars, eval_stmt et vars s) }.
-\end{lstlisting}
+    ; variety_laws: Π s, et_laws et s → (Π $\hspace{-1.5mm}$vars, eval_stmt et vars s) }.
+\end{lstlisting}
+  % todo: why do i need the manual \hspace!?
 
@@
 
-In addition to this ``variety'' category, we also have categories of lawless algebras, as well as forgetful functors from the former to the latter, and from the latter to the category of setoids.
+In addition to this variety category, we also have categories of lawless algebras, as well as forgetful functors from the former to the latter, and from the latter to the category of setoids.
 
@@
 \begin{theorem}[First homomorphism theorem]
-If $A$ and $B$ are algebras, and $f$ is a homomorphism from $A$ to $B$, then the equivalence relation defined by $a\sim b$ if and only if $f(a)=f(b)$ is a congruence on $A$, and the algebra $A/\sim$ is isomorphic to the image of $f$, which is a subalgebra of B.
+If $A$ and $B$ are algebras, and $f$ is a homomorphism from $A$ to $B$, then the equivalence relation defined by $a\sim b$ if and only if $f(a)=f(b)$ is a congruence on $A$, and the algebra $A/\hspace{-1mm}\sim$ is isomorphic to the image of $f$, which is a subalgebra of B.
 \end{theorem}
@@
   { congruence_proper:> Π s, Proper (equiv ==> equiv ==> iff) (e s)
-    ; congruence_quotient:> @Algebra et v e _ }.
+  ; congruence_quotient:> Algebra σ$$ v (e:=e) }.
 \end{lstlisting}
@@
 Basing theory and programs on this abstract interface instead of on specific implementation (such as the ubiquitous Peano naturals \lstinline|nat| in the Coq standard library) is not only cleaner mathematically, but also facilitates easy swapping between implementations.
-And this benefit is far from theoretical, as diverse representations of the natural numbers are abound; for instance, unary, binary, factor multisets, arrays of native machine words.
+And this benefit is far from theoretical, as diverse representations of the natural numbers are abound; for instance, unary, binary, factor multisets, and arrays of native machine words.
 
@@
 \begin{lstlisting}
-Definition calculate_things `{Naturals N} `{!Π n m: N, Decision (n = m)}
+Definition calculate_things `{Naturals N} `{Π$$n m: N, Decision (n = m)}
   (a b: nat): ... := ... decide (a = b) ... .
@@
 Hence, theory and programs developed on our \lstinline|Rationals| interface applies and can make immediate use of these efficient rationals.
-As stated earlier, we plan to rebase the computable real number implementation in CoRN on this interface, precisely so that it may be instantiated with efficient implementations like these.
-
-We plan to provide an abstract interface for polynomials as a free commutative algebra. This would unify existing implementations such as coefficient lists and Bernstein polynomials.
+We plan to rebase the computable real number implementation in CoRN on this interface, precisely so that it may be instantiated with efficient implementations like these.
+
+We also plan to provide an abstract interface for polynomials as a free commutative algebra. This would unify existing implementations such as coefficient lists and Bernstein polynomials.
 
@@
 Traditionally, proof assistants such as Coq have provided sophisticated tactics or built-in commands to support this ``quoting''.
-In~\cite{Hints}, TODO et.
+In~\cite{Hints}, Asperti et.
 al.
@@
 Their ``output'' will consist not only of the fields in the class, but also of \lstinline|V'| and \lstinline|r| representing additional holes and their values.
-Hence, a type class constraint of the form \lstinline|Quote x y z| should be read as ``quoting y with existing heap x generates new heap z''.
+Hence, a type class constraint of the form \lstinline|Quote x y z| should be read as ``quoting \lstinline|y| with existing heap \lstinline|x| generates new heap \lstinline|z|''.
 
@@
   Instance quote_mult V (v: Vars V) n V' (v': Vars V') m V'' (v'': Vars V'')
-    `{!Quote v n v'} `{!Quote (merge v v') m v''}: Quote v (n * m) (merge v' v'') :=
-      { quote := Mult (map_var shift (quote n)) (map_var sum_assoc (quote m)) }.
+    `{!Quote v n v'} `{!Quote (merge v v') m v''}:
+      Quote v (n * m) (merge v' v'') :=
+      { quote :=
+        Mult (map_var shift (quote n)) (map_var sum_assoc (quote m)) }.
 \end{lstlisting}
@@
 
-Using a small additional utility lemma we can also easily quote equalities into \lstinline|eval p e = eval p' e'| form with a shared variable pack (so that identical expressions on both sides of the equation don't end up as different variables).
+Using a small additional utility lemma we can also easily quote equalities into \lstinline|eval p e = eval p e'| form with a shared variable pack (so that identical expressions on both sides of the equation don't end up as different variables).
 
@@
 A categorical way of expressing this is in terms of adjunctions.
-As before, we \emph{could} fully embrace this perspective, paying no heed to practicality of implementation and usage, and define a relatively succinct type class for sequences as follows:
+As with the numeric interfaces, we \emph{could} fully embrace this perspective, paying no heed to practicality of implementation and usage, and define a relatively succinct type class for sequences as follows:
 \begin{lstlisting}
@@
 
+We emphasize that these efficiency problems only affect type checking; efficiency of computation using type-checked terms is not affected.
+
 We are currently in the process of retrofitting the rationals interface into CoRN.

### 2010-06-30 jar.tex
Back to 'multiplication/division by 2.

@@
 
-Other examples of operations on natural numbers that are sensible choices for specialization include: subtraction, distance, binary division and multiplication.
+Other examples of operations on natural numbers that are sensible choices for specialization include: subtraction, distance, and division and multiplication by 2.
 

### 2010-06-30 jar.tex
Fixed scare quotes

--- 2010-06-30 jar.tex
@@
 telescopes), packed classes~\cite{Packed}, and occasionally modules.
-We present a new approach based entirely on Coq's new type class mechanism, and show how it can be used to make fully ``unbundled'' predicate-representations of algebraic structures practical to work with.
+We present a new approach based entirely on Coq's new type class mechanism, and show how it can be used to make fully `unbundled' predicate-representations of algebraic structures practical to work with.
 
@@
 Then, in section~\ref{bundling}, we give a very concrete introduction to the issue of \emph{bundling}, arguably the biggest design dimension when building interfaces for abstract structures.
-In section~\ref{predicateclasses} we show how type classes can make the use of ``unbundled'' purely predicate based interfaces for abstract structures practical.
+In section~\ref{predicateclasses} we show how type classes can make the use of `unbundled' purely predicate based interfaces for abstract structures practical.
 Next, in section~\ref{classes}, we discuss our algebraic hierarchy implemented using such predicate classes.
@@
 
-Occasionally there is some ambiguity as to whether a certain piece of information (such as a witness to an existential statement) is strictly ``proof matter'' (and thus belongs in the \lstinline|Prop| sort) or actually of further computational interest (and thus does \emph{not} belong to the \lstinline|Prop| sort).
+Occasionally there is some ambiguity as to whether a certain piece of information (such as a witness to an existential statement) is strictly `proof matter' (and thus belongs in the \lstinline|Prop| sort) or actually of further computational interest (and thus does \emph{not} belong to the \lstinline|Prop| sort).
 We will see one such case when we discuss the first homomorphism theorem in section \ref{homothm}.
@@
 
-The ``native'' notion of equality in Coq is that of terms being convertible, which in this context constitutes Leibniz equality, naturally propositionalized by an inductive type family \lstinline|eq| with a single constructor \lstinline|eq_refl| of type \lstinline|Π (T: Type) (x: T), x ≡ x|, where ``\lstinline|a ≡ b|'' is notation for \lstinline|eq T a b|.
-Here we diverge from Coq tradition and reserve ``\lstinline|=|'' for setoid equality, as this is the equality we will be working with most of the time.
+The `native' notion of equality in Coq is that of terms being convertible, which in this context constitutes Leibniz equality, naturally propositionalized by an inductive type family \lstinline|eq| with a single constructor \lstinline|eq_refl| of type \lstinline|Π (T: Type) (x: T), x ≡ x|, where `\lstinline|a ≡ b|' is notation for \lstinline|eq T a b|.
+Here we diverge from Coq tradition and reserve \lstinline|=| for setoid equality, as this is the equality we will be working with most of the time.
 Importantly, since convertibility is a congruence, a proof of \lstinline|a ≡ b| lets us substitute \lstinline|b| for \lstinline|a| anywhere inside a term without further conditions.
@@
 
-Because the Coq type theory lacks ``real'' quotient types (as it would make type checking undecidable), the conventional way of working with abstract structures in Coq is to always start with \emph{setoids} (aka Bishop sets): types equipped with an equivalence relation~\cite{Bishop67,Hofmann,Capretta}.
+Because the Coq type theory lacks quotient types (as it would make type checking undecidable), the conventional way of working with abstract structures in Coq is to always start with \emph{setoids} (aka Bishop sets): types equipped with an equivalence relation~\cite{Bishop67,Hofmann,Capretta}.
 As we will see in section ~\ref{univ}, this way of working by nature lends itself well to particularly simple definitions of notions such as quotient algebras.
@@
 In Haskell and Isabelle, type classes and their instances are second class, in that they are handled as specialized syntactic constructs whose semantics are given specifically by the type class apparatus.
-By contrast, the expressivity of dependent types and inductive families as supported in Coq, combined with the use of pre-existing technology in the system (namely proof search and implicit arguments) enable a \emph{first class} type class implementation~\cite{DBLP:conf/tphol/SozeauO08}: classes are ordinary record types (``dictionaries''), instances are ordinary constants of these record types (registered as ``hints'' with the proof search machinery), class constraints are ordinary implicit parameters, and instance resolution is achieved by augmenting the unification algorithm to invoke ordinary proof search for implicit arguments of class type.
+By contrast, the expressivity of dependent types and inductive families as supported in Coq, combined with the use of pre-existing technology in the system (namely proof search and implicit arguments) enable a \emph{first class} type class implementation~\cite{DBLP:conf/tphol/SozeauO08}: classes are ordinary record types (`dictionaries'), instances are ordinary constants of these record types (registered as `hints' with the proof search machinery), class constraints are ordinary implicit parameters, and instance resolution is achieved by augmenting the unification algorithm to invoke ordinary proof search for implicit arguments of class type.
 
@@
 Superficially, this instantly solves the problems described above: reflexive relations can now be declared and passed as self-contained packages, and the \lstinline|rr_rel| projection now constitutes a ``canonical name'' for relations that are known to be reflexive, and we could bind a notation to it.
-While there is no conventional notation for reflexive relations, the situation is the same in the context of, say, a semiring, where we would bind ``$+$'' and ``$*$'' notations to the record field projections for the addition and multiplication operations, respectively.
+While there is no conventional notation for reflexive relations, the situation is the same in the context of, say, a semiring, where we would bind $+$ and $*$ notations to the record field projections for the addition and multiplication operations, respectively.
 
@@
 In our example, lack of sharing support rears its head as soon as we try to define \lstinline|EquivalenceRelation| in terms of \lstinline|ReflexiveRelation| and its hypothetical siblings bundling symmetric and transitive relations.
-There, we would need some way to make sure that when we ``inherit'' \lstinline|ReflexiveRelation|, \lstinline|SymmetricRelation|, and \lstinline|TransitiveRelation| by adding them as fields in our bundled record, they all refer to the same carrier and relation.
+There, we would need some way to make sure that when we `inherit' \lstinline|ReflexiveRelation|, \lstinline|SymmetricRelation|, and \lstinline|TransitiveRelation| by adding them as fields in our bundled record, they all refer to the same carrier and relation.
 Adding additional fields stating equalities between the three bundled carriers and relations is neither easily accomplished (because one would need to work with heterogenous equality) nor would it permit natural use of the resulting structure (because one would constantly have to rewrite things back and forth).
@@
 This is not the case, because there is no point in sharing them.
-After all, by (weak) proof irrelevance, the ``value'' of such proofs can be of no consequence anyway.
+After all, by (weak) proof irrelevance, the `value' of such proofs can be of no consequence anyway.
 Indeed, parameterizing on proofs would be actively harmful because instantiations differing only in the proof argument would express the same thing yet be nonconvertible, requiring awkward conversions and hindering automation.
@@
 As an aside, we note that the distinction between the class field name and the infix operator notation bound to it is really just a mildly awkward Coq artifact.
-In Haskell, where operators can themselves be used as names, there would be no need to have the \lstinline|equiv| and \lstinline|sg_op| names in addition to the operator ``names''.
+In Haskell, where operators can themselves be used as names, there would be no need to have the \lstinline|equiv| and \lstinline|sg_op| names in addition to the operator `names'.
 
@@
 \end{lstlisting}
-We are not particularly worried about overly large contexts, especially because most of the ``extra'' hypotheses we have compared to bundled approaches are declarations of relations, operators, and constants, which are all in some sense ``inert'' with respect to proof search.
+We are not particularly worried about overly large contexts, especially because most of the `extra' hypotheses we have compared to bundled approaches are declarations of relations, operators, and constants, which are all in some sense `inert' with respect to proof search.
 Hence, we do not foresee problems with large contexts for any but the most complex formalizations.
@@
 Fields declared with \lstinline|:>| are registered as hints for instance resolution, so that in any context where \lstinline|(A, =, +, 0, *, 1)| is known to be a \lstinline|SemiRing|, \lstinline|(A, =, +, 0)| and \lstinline|(A, =, *, 1)| are automatically known to be \lstinline|CommutativeMonoid|s (and so on, transitively, because instance resolution is recursive).
-In our hierarchy, these ``substructures'' by themselves establish the following inheritance diagram:
+In our hierarchy, these `substructures' by themselves establish the following inheritance diagram:
 
@@
 This kind of redundant knowledge (which arises naturally in mathematics) is never a problem in our setup, because the use of operational type classes ensures that terms composed of algebraic operations and relations never refer to structure proofs.
-We find that this is a tremendous relief compared to approaches that \emph{do} intermix the two and where one must be careful to ensure that such terms refer to the ``right'' proofs of properties.
-There, even \emph{strong} proof irrelevance (which would make terms convertible that differ only in what proofs they refer to) would not make these difficulties go away entirely, because high-level tactics that rely on quotation of terms require syntactic identity (rather than ``mere'' convertibility) to recognize identical subterms.
+We find that this is a tremendous relief compared to approaches that \emph{do} intermix the two and where one must be careful to ensure that such terms refer to the `right' proofs of properties.
+There, even \emph{strong} proof irrelevance (which would make terms convertible that differ only in what proofs they refer to) would not make these difficulties go away entirely, because high-level tactics that rely on quotation of terms require syntactic identity (rather than `mere' convertibility) to recognize identical subterms.
 
@@
 Doing this consistently minimizes administrative noise in the code, bringing us closer to ordinary mathematical vernacular.
-Indeed, we believe that type classes are an elegant and apt formalization of the seemingly casual manner in which ordinary mathematical presentation assumes implicit administration and use of a ``database'' of properties previously proved.
+Indeed, we believe that type classes are an elegant and apt formalization of the seemingly casual manner in which ordinary mathematical presentation assumes implicit administration and use of a `database' of properties previously proved.
 Much more so than the existing canonical structures facility, because canonical structures can only be used with bundled representations, which we argued against in section~\ref{bundling}.
@@
 The \lstinline|Initial| class itself further requires these ``initial arrows'' to be unique.
-Having \lstinline|InitialArrows| as an operational type class means that we can always simply say \lstinline|initial_arrow y| whenever \lstinline|y| is known to be an object in a category known to have an initial object (where ``known'' should be read as ``can be determined by instance resolution'').
+Having \lstinline|InitialArrows| as an operational type class means that we can always simply say \lstinline|initial_arrow y| whenever \lstinline|y| is known to be an object in a category known to have an initial object (where `known' should be read as ``can be determined by instance resolution'').
 
@@
 We ought to say a few words about our use of \lstinline|fmap|.
-The usual mathematical notational convention for functor application is to use the name of the functor to refer to both its object map and its arrow map, relying on additional conventions regarding object/arrow names for disambiguation: ``\lstinline|F x|'' and ``\lstinline|F f|'' map an object and an arrow, respectively, because ``\lstinline|x|'' and ``\lstinline|f|'' are conventional names for objects and arrows, respectively.
+The usual mathematical notational convention for functor application is to use the name of the functor to refer to both its object map and its arrow map, relying on additional conventions regarding object/arrow names for disambiguation: \lstinline|F x| and \lstinline|F f| map an object and an arrow, respectively, because \lstinline|x| and \lstinline|f| are conventional names for objects and arrows, respectively.
 
@@
 We feel that these issues are not limitations of the Coq system, but merely reflect the fact that notationally identifying these two distinct and interdependent maps is an abuse of notation of sufficient severity to make it ill-suited to a formal development where software engineering concerns apply.
-Hence, we do not adopt this practice, and use ``\lstinline|fmap F|'' (name taken from the Haskell standard library) to refer to the arrow map of a functor \lstinline|F|.
+Hence, we do not adopt this practice, and use \lstinline|fmap F| (name taken from the Haskell standard library) to refer to the arrow map of a functor \lstinline|F|.
 
@@
 To specify the natural numbers and the integers as initial objects in the categories of semirings and rings, respectively, definitions of these categories are needed.
-While one could define both of them ``manually'', greater economy can be achieved by recognizing that both semirings and rings can be defined by equational theories, for which \emph{varieties} can be defined generically (varieties are categories consisting of models for a fixed theory with homomorphisms between them).
+While one could define both of them manually, greater economy can be achieved by recognizing that both semirings and rings can be defined by equational theories, for which \emph{varieties} can be defined generically (varieties are categories consisting of models for a fixed theory with homomorphisms between them).
 
@@
 
-A multisorted signature enumerates sorts, operations, and specifies the ``types'' of the operations as non-empty lists of sorts, where the final element denotes the result type:
+A multisorted signature enumerates sorts, operations, and specifies the `types' of the operations as non-empty lists of sorts, where the final element denotes the result type:
 \begin{lstlisting}
@@
 
-What remains is to show that carriers+\lstinline|Equivs|+\lstinline|AlgebraOps| satisfying \lstinline|InVariety| for a given \lstinline|EquationalTheory| do indeed form a \lstinline|Category| (the ``variety'').
+What remains is to show that carriers+\lstinline|Equivs|+\lstinline|AlgebraOps| satisfying \lstinline|InVariety| for a given \lstinline|EquationalTheory| do indeed form a \lstinline|Category| (the `variety').
 Since we need a type for the objects in the \lstinline|Category|, at this point we have no choice but to bundle components and proof together in a record:
@@
 \end{lstlisting}
-While succinct, this definition is not a satisfactory abstraction because the use of \lstinline|ObjectInVariety| for the type of the \lstinline|A| component ``leaks'' the fact that we used this one particular universal algebraic construction of the category, which is just an implementation choice.
+While succinct, this definition is not a satisfactory abstraction because the use of \lstinline|ObjectInVariety| for the type of the \lstinline|A| component `leaks' the fact that we used this one particular universal algebraic construction of the category, which is just an implementation choice.
 Furthermore, this definition needs an additional layer of class instances to relate it to the \lstinline|SemiRing| class from our algebraic hierarchy.
@@
 
-Some theory and programs about naturals are more easily proved, respectively defined, for concrete implementations and then ``lifted'' to the abstract interface.
+Some theory and programs about naturals are more easily proved, respectively defined, for concrete implementations and then lifted to the abstract interface.
 For relatively simple things like decidability of equality, all that is needed to make this lifting easy is a handful of simple lemmas translating the initiality property back into statements saying that \lstinline|naturals_to_semiring| is extensionally equal to other semiring morphisms and involutary between \lstinline|Naturals| implementations.
@@
 
-Traditionally, proof assistants such as Coq have provided sophisticated tactics or built-in commands to support this ``quoting''.
+Traditionally, proof assistants such as Coq have provided sophisticated tactics or built-in commands to support this quoting.
 In~\cite{Hints}, Asperti et.
@@
 We can think of \lstinline|Quote| as the type for a family of Prolog-like syntax-directed resolution functions, which will take as input \lstinline|V| and \lstinline|l| representing previously encountered holes (opaque subexpressions that could not be destructured further) and their values, along with a concrete term \lstinline|n| to be quoted.
-Their ``output'' will consist not only of the fields in the class, but also of \lstinline|V'| and \lstinline|r| representing additional holes and their values.
+Their `output' will consist not only of the fields in the class, but also of \lstinline|V'| and \lstinline|r| representing additional holes and their values.
 Hence, a type class constraint of the form \lstinline|Quote x y z| should be read as ``quoting \lstinline|y| with existing heap \lstinline|x| generates new heap \lstinline|z|''.
@@
 On top of this interface we can build theory about typical list operations such as maps, folds, their relation to \lstinline|singleton| and \lstinline|extend|, et cetera.
-We can also generically define ``bigops'' for summation and products, and easily show properties like distributivity, all without ever mentioning cons lists.
+We can also generically define `bigops' for summation and products, and easily show properties like distributivity, all without ever mentioning cons lists.
 

### 2010-06-30 jar.tex
Fix two typos.

--- 2010-06-30 jar.tex
@@
 As part of our universal algebra theory, we have proved that proofs of statements in the language of an equational theory can be transferred between isomorphic implementations.
-Combined with the trivially proved property that any two initial objects in a category are isomorphic, this lets us tranfer proofs of such statements between implementations of \lstinline|Naturals|, requiring only that we reflect the concrete statement (expressed in terms of the operational type classes) to a symbolic statement in the language of semirings.
+Combined with the trivially proved property that any two initial objects in a category are isomorphic, this lets us transfer proofs of such statements between implementations of \lstinline|Naturals|, requiring only that we reflect the concrete statement (expressed in terms of the operational type classes) to a symbolic statement in the language of semirings.
 We intend to eventually make this reflection completely automatic using type class based quotation techniques along the lines of those described in Section~\ref{quoting}.
@@
 The second is instance resolution efficiency.
-In more complex parts of our development we are now experiencing increasingly serious efficiency problems, despite having already made sacrifices by artifically inhibiting many natural class instances in order not to further strain instance resolution.
+In more complex parts of our development we are now experiencing increasingly serious efficiency problems, despite having already made sacrifices by artificially inhibiting many natural class instances in order not to further strain instance resolution.
 %While we have already found the predicate-class-based style of formalization to work very well for our development thus far, and believe it has the potential to be \emph{the} modern solution to this problem, in our work we have encountered a number of limitations in the Coq implementation of key features of the tool, that keep us from making an unconditional recommendation for universal adoption of these techniques.

### 2010-06-30 jar.tex
In section 4, use [Theorem] consistently and rename L to 'example'.

@@
 \begin{quote}
-Theorem L: For $x, y, z$ in a semigroup $G$, $x * y * z = z * y * x$.
+Theorem: For $x, y, z$ in a semigroup $G$, $x * y * z = z * y * x$.
 \end{quote}
@@
 \begin{lstlisting}
-  Theorem L G e op {P: SemiGroup G e op}:
+  Theorem example G e op {P: SemiGroup G e op}:
     Π x y z, e (op (op x y) z) (op (op z y) x).
@@
 \begin{lstlisting}
-  Theorem L G e op {P: SemiGroup G e op}: Π x y z, x & y & z = z & y & x.
+  Theorem example G e op {P: SemiGroup G e op}:
+    Π x y z, x & y & z = z & y & x.
 \end{lstlisting}
@@
 \begin{lstlisting}
-  Theorem L G e op {P: SemiGroup G e op}: Π x y z,
-    equiv e (sg_op op (sg_op op x y) z) (sg_op op (sg_op op z y) x).
+  Theorem example G e op {P: SemiGroup G e op}:
+    Π x y z, equiv e (sg_op op (sg_op op x y) z) (sg_op op (sg_op op z y) x).
 \end{lstlisting}
@@
 \begin{lstlisting}
-  Lemma L `{SemiGroup G}: Π x y z: G, x & y & z = z & y & x.
+  Theorem example `{SemiGroup G}: Π x y z: G, x & y & z = z & y & x.
 \end{lstlisting}

### 2010-06-30 jar.tex
Many lay-out, text changes

--- 2010-06-30 jar.tex
@@
 % especially since existing solutions fall short from a proof engineering perspective.
-We present a new set of foundational type classes for mathematics and take full advantage of their unique features to make practical a particularly flexible approach formerly thought infeasible.
+We present a new set of type classes for mathematics and take full advantage of their unique features to make practical a particularly flexible approach formerly thought infeasible.
 Thus, we address both traditional proof engineering challenges as well as new ones resulting from our ambition to build upon this development a library of constructive analysis in which abstraction penalties inhibiting efficient computation are reduced to a minimum.
@@
 For the algebraic hierarchy---a critical component in any library of formalized mathematics---these challenges include structure inference, handling of multiple inheritance, idiomatic use of notations, and convenient algebraic manipulation (e.g.\ rewriting).
-
 Several solutions have been proposed for the Coq theorem prover: dependent records~\cite{DBLP:journals/jsc/GeuversPWZ02} (a.k.a.
@@
 Doing this requires that we have clean abstract interfaces, and mathematics tells us what these should look like: we represent $\N$, $\Z$, and $\Q$ as \emph{interfaces} specifying an initial semiring, an initial ring, and a field of integral fractions, respectively.
-
-To express these elegantly and without duplication, our development\footnote{The sources are available
+To express these interfaces elegantly and without duplication, our development\footnote{The sources are available
 at:~\url{http://www.eelis.net/research/math-classes/}} includes an integrated formalization of parts of category theory and multi-sorted universal algebra, all expressed using type classes for optimum effect.
@@
 
-\paragraph{Outline}
-In section~\ref{preliminaries} we briefly describe the Coq system and some of its facilities that are of particular interest to us (most notably its implementation of type classes).
+\paragraph{Outline.}
+In section~\ref{preliminaries} we briefly describe the Coq system and its implementation of type classes.
 Then, in section~\ref{bundling}, we give a very concrete introduction to the issue of \emph{bundling}, arguably the biggest design dimension when building interfaces for abstract structures.
@@
 In order to illustrate a very different use of type classes, we discuss the implementation of a quoting function for algebraic terms in terms of type classes, in section~\ref{quoting}.
-In section~\ref{sequences} we hint at an interface for sequences but describe how a limitation in the Coq system makes its use problematic.
+In section~\ref{sequences} we hint at an interface for sequences, but describe how a limitation in the Coq system makes its use problematic.
 We end with conclusions in section~\ref{conclusions}.
@@
 constructions~\cite{CoquandHuet,CoquandPaulin}, a dependent type theory with (co)inductive types; see~\cite{Coq,BC04}.
-In true Curry-Howard fashion, it can be thought of either as an excessively pure (if somewhat pedantic) functional programming language with an extremely expressive type system, or as a language for mathematical statements and proofs.
+In true Curry-Howard fashion, it is both an excessively pure, if somewhat pedantic, functional programming language with an extremely expressive type system, and a language for mathematical statements and proofs.
 We highlight some aspect of Coq that are of particular relevance to our development.
 
-\paragraph{Types and propositions}
+\paragraph{Types and propositions.}
 
 Propositions in Coq are types~\cite{ITT,CMCP}, which themselves have types called \emph{sorts}.
-Coq features a distinguished sort called \lstinline|Prop| that one may voluntarily choose to use as the sort for types representing propositions.
-The distinguishing feature of the \lstinline|Prop| sort is that Coq will not permit terms of non-\lstinline|Prop| types to depend on the values of inhabitants of \lstinline|Prop| types (that is, proof terms).
-
+Coq features a distinguished sort called \lstinline|Prop| that one may choose to use as the sort for types representing propositions.
+The distinguishing feature of the \lstinline|Prop| sort is that terms of non-\lstinline|Prop| type may not depend on the values of inhabitants of \lstinline|Prop| types (that is, proof terms).
 % As originally conceived~\cite{ITT,CMCP}, Propositions are Types and we can extract information from
@@
 Especially with the aid of the Program machinery, this allows us to flexibly work with simply typed program and their Hoare style correctness proofs.
-
-Adopting this regime of discrimination establishes a weak form of proof irrelevance, in that changing a proof can never affect the result of value computations.
+This regime of discrimination establishes a weak form of proof irrelevance, in that changing a proof can never affect the result of value computations.
 On a very practical level, this lets Coq safely erase all \lstinline|Prop| components when extracting certified programs to OCaml or Haskell.
-
 A stronger notion of proof irrelevance in which any two inhabitants of a \lstinline|Prop| type are made \emph{convertible} is planned for a future Coq version.
@@
 
-The `native' notion of equality in Coq is that of terms being convertible, which in this context constitutes Leibniz equality, naturally propositionalized by an inductive type family \lstinline|eq| with a single constructor \lstinline|eq_refl| of type \lstinline|Π (T: Type) (x: T), x ≡ x|, where `\lstinline|a ≡ b|' is notation for \lstinline|eq T a b|.
+The `native' notion of equality in Coq is that of terms being convertible, which in this context constitutes Leibniz equality, naturally propositionalized by an inductive type family \lstinline|eq| with a single constructor
+\begin{lstlisting}
+ eq_refl : Π (T: Type) (x: T), x ≡ x,
+\end{lstlisting}
+where `\lstinline|a ≡ b|' is notation for \lstinline|eq T a b|.
 Here we diverge from Coq tradition and reserve \lstinline|=| for setoid equality, as this is the equality we will be working with most of the time.
 Importantly, since convertibility is a congruence, a proof of \lstinline|a ≡ b| lets us substitute \lstinline|b| for \lstinline|a| anywhere inside a term without further conditions.
-
-We mention this explicitly only because rewriting \emph{does} give rise to conditions when we depart from Leibniz equality and introduce equivalence relations representing abstractions over common representations of the same conceptual object (for instance when identifying formal fractions of integers representing the same ratio).
+We mention this explicitly only because rewriting \emph{does} give rise to conditions when we depart from Leibniz equality and introduce equivalence relations representing abstractions over common representations of the same conceptual object.
+For instance, consider formal fractions of integers as a representation of rationals.
 Rewriting a subterm using such an equality is permitted only if the subterm is an argument of a function that has been proven to \emph{respect} the equality.
@@
 
-Because the Coq type theory lacks quotient types (as it would make type checking undecidable), the conventional way of working with abstract structures in Coq is to always start with \emph{setoids} (aka Bishop sets): types equipped with an equivalence relation~\cite{Bishop67,Hofmann,Capretta}.
-As we will see in section ~\ref{univ}, this way of working by nature lends itself well to particularly simple definitions of notions such as quotient algebras.
+Because the Coq type theory lacks quotient types (as it would make type checking undecidable), one usually bases abstract structures on a \emph{setoid} (`Bishop set'): a type equipped with an equivalence relation~\cite{Bishop67,Hofmann,Capretta}.
+As we will see in section~\ref{univ}, this way of working pays off when working with notions such as quotient algebras.
 
@@
 
-For an interesting (if fairly restrictive) alternative approach based on the idea of requiring unicity of representation for all objects so that Leibniz equality is always applicable, see the algebraic hierarchy part of the ssreflect~\cite{Packed} libraries.
-
-\paragraph{Type classes}
-
-Type classes have been a great success story in the Haskell functional programming language, as a means for organizing interfaces of abstract structures.
-The type class implementation that is now part of Coq provides a superset of their functionality, but implemented in a different way.
-
-In Haskell and Isabelle, type classes and their instances are second class, in that they are handled as specialized syntactic constructs whose semantics are given specifically by the type class apparatus.
+The algebraic hierarchy of the ssreflect~\cite{Packed} libraries uses an interesting (if fairly restrictive) alternative approach.
+It requires unicity of representation for all objects so that Leibniz equality is always applicable.
+
+\paragraph{Type classes.}
+
+Type classes have been a great success story in the Haskell functional programming language, as a means of organizing interfaces of abstract structures.
+Coq's type classes provide a superset of their functionality, but implemented in a different way.
+
+In Haskell and Isabelle, type classes and their instances are second class.
+They are handled as specialized syntactic constructs whose semantics are given specifically by the type class apparatus.
 By contrast, the expressivity of dependent types and inductive families as supported in Coq, combined with the use of pre-existing technology in the system (namely proof search and implicit arguments) enable a \emph{first class} type class implementation~\cite{DBLP:conf/tphol/SozeauO08}: classes are ordinary record types (`dictionaries'), instances are ordinary constants of these record types (registered as `hints' with the proof search machinery), class constraints are ordinary implicit parameters, and instance resolution is achieved by augmenting the unification algorithm to invoke ordinary proof search for implicit arguments of class type.
-
 Thus, type classes in Coq are realized by relatively minor syntactic aids that bring together existing facilities of the theory and the system into a coherent idiom, rather than by introduction of a new category of qualitatively different definitions with their own dedicated semantics.
@@
 
-Higher structures, too, can all be expressed as predicates (expressing laws) over a number of carriers, relations, and operations.
+More elaborate structures, too, can be expressed as predicates (expressing laws) over a number of carriers, relations, and operations.
 While optimally flexible in principle, in practice \emph{naive} adoption of this approach (that is, without using type classes) leads to substantial inconveniences in actual use: when \emph{stating} theorems about abstract instances of such structures, one must enumerate all components along with the structure (predicate) of interest.
@@
 
-Manifest fields~\cite{DBLP:conf/types/Luo08} have been proposed and implemented in the Matita system to address exactly this problem.
+Manifest fields~\cite{DBLP:conf/types/Luo08} have been proposed, and implemented in the Matita system, to address exactly this problem.
 We hope to convince the reader that type system extensions of this nature, designed to mitigate particular symptoms of the bundled approach, are less elegant than a solution (described in the next section) that avoids the problem altogether by using predicate-like type classes in place of bundled records.
@@
 \end{lstlisting}
-However, the same argument as before leads to the conclusion that \mbox{\lstinline|EquivalenceRelation|,} too, should be a predicate.
+However, as before we conclude that \mbox{\lstinline|EquivalenceRelation|,} too, should be a predicate.
 Indeed, it would be rather strange for the interface of equivalence relations to differ qualitatively from the interface of reflexive relations.
 
-Another attempt to recover some grouping might be to bundle the carrier with the relation into a (lawless) record, but this too hinders sharing: as soon as we try to define an algebraic structure with two reflexive relations on the same carrier, we need awkward hacks to establish equality between the carrier projections of two different (carrier, relation) bundles.
-
-Even bundling just the operations of an algebraic structure together in a record (with the carrier as a parameter) does not work, as it again leads to the same problem when, for example, one attempts to define a hypothetical algebraic structure with two binary relations and a constant such that both binary relations form a monoid with the constant.
+Another attempt to recover some grouping might be to bundle the carrier with the relation into a (lawless) record, but this too hinders sharing.
+As soon as we try to define an algebraic structure with two reflexive relations on the same carrier, we need awkward hacks to establish equality between the carrier projections of two different (carrier, relation) bundles.
+
+Even bundling just the operations of an algebraic structure together in a record (with the carrier as a parameter) leads to the same problem when, for example, one attempts to define a hypothetical algebraic structure with two binary relations and a constant such that both binary relations form a monoid with the constant.
 
 A second problem with bundling is that as the bundled records are stacked to represent higher and higher structures, the projection paths for their components grow longer and longer, resulting in ever more unwieldy terms (coercions and notations can make this less painful).
-Further, if one tries to implement some semblance of sharing in a bundled representation, these projection paths additionally become non-canonical, and still more extensions have been proposed to address this symptom (e.g.\ coercion pullbacks~\cite{Hints}).
+Further, if one tries to implement some semblance of sharing in a bundled representation, these projection paths additionally become non-canonical, and still more extensions have been proposed to address this symptom, e.g.\ coercion pullbacks~\cite{Hints}.
 
@@
 \end{quote}
-(Please forgive the admittedly silly statement; we are presently concerned only with the syntax.)
+(This silly statement allows us to clearly present the syntax.)
 
@@
 \end{lstlisting}
-where e and the op's are filled in by instance resolution.
+where \lstinline|e| and the \lstinline|op|s are filled in by instance resolution.
 
@@
 Since \lstinline|CommutativeMonoid| indirectly includes a \lstinline|SemiGroup| field, which in turn includes a \lstinline|Equivalence| field, having a \lstinline|SemiRing| proof means having two distinct proofs that the equality relation is an equivalence.
-This kind of redundant knowledge (which arises naturally in mathematics) is never a problem in our setup, because the use of operational type classes ensures that terms composed of algebraic operations and relations never refer to structure proofs.
+This kind of redundant knowledge ()which arises naturally) is never a problem in our setup, because the use of operational type classes ensures that terms composed of algebraic operations and relations never refer to structure proofs.
 We find that this is a tremendous relief compared to approaches that \emph{do} intermix the two and where one must be careful to ensure that such terms refer to the `right' proofs of properties.
@@
 Together, these properties largely defuse an argument occasionally voiced against type classes concerning perceived unpredictability of instance resolution.
-While it is certainly true that in contexts with redundant information it can become hard to predict which instance of a predicate class will be found by proof search, this is simply a non-issue because it \emph{does not matter} which one is found.
-And for operational type classes the issue rarely arises because their instances are not nearly as abundant.
+While it is certainly true that in contexts with redundant information it can become hard to predict which instance of a predicate class will be found by proof search, it simply \emph{does not matter} which one is found.
+Moreover, for operational type classes the issue rarely arises because their instances are not nearly as abundant.
 
@@
 
-The reader will note that we use names for properties like distributivity and absorption.
-Indeed, these are type classes as well (which is why we declare them with \lstinline|:>|).
+We use names for properties like distributivity and absorption, because these are type classes as well (which is why we declare them with \lstinline|:>|).
 It has been our experience that almost any generic predicate worth naming is worth representing as a predicate type class, so that its proofs will be resolved as instances behind the scenes whenever possible.
@@
 The operational type classes used in \lstinline|SemiRing| for zero, one, multiplication and addition, are the same ones used by \lstinline|Ring| and \lstinline|Field| (not shown).
-Thus, the realization that a particular semiring is in fact a ring or field (etc.) has no bearing on how one refers to the operations in question, which is as it should be.
+Thus, the realization that a particular semiring is in fact a ring or field has no bearing on how one refers to the operations in question, which is as it should be.
 The realization that a particular semigroup is part of a semiring \emph{does} call for a new (canonical) name, simply because of the need for disambiguation.
-The introduction of these additional names for the same operation is quite harmless in practice, because canonical names established by operational type class fields are just trivial identity functions anyway, so that in most contexts the distinction reduces away instantly.
+The introduction of these additional names for the same operation is quite harmless in practice, because canonical names established by operational type class fields are identity functions, so that in most contexts the distinction reduces away instantly.
 
@@
 Some clarification is in order to explain the role of the \lstinline|Context| declaration of the two monoids.
-While it may look we are parameterizing \lstinline|Monoid_Morphism| on monoid-ness proofs (which would be a gross violation of our idiom), in fact it is only parameterized on the monoid \emph{components} declared through implicit generalization of the \lstinline|Monoid| declarations, because it only refers to those.
+While \lstinline|Monoid_Morphism| seemingly depends on monoid-ness proofs (which would be a gross violation of our idiom), in fact it is only parameterized on the monoid \emph{components} declared through implicit generalization of the \lstinline|Monoid| declarations, because it only refers to those.
 Here, we use declarations of predicate class parameters merely as convenient shorthands to declare their components.
@@
 
-Unfortunately, there is actually one annoying wrinkle here, which will also explain why, as the observant reader will have noticed, we do not register these two fields as instance resolution hints (by declaring them with \lstinline|:>|).
+Unfortunately, there is actually one annoying wrinkle here, which will also explain why we do not register these two fields as instance resolution hints (by declaring them with \lstinline|:>|).
 What we really want these fields to express is ``\emph{if} in a certain context we know something to be a \lstinline|Monoid_Morphism|, \emph{then} realize that the source and target are \lstinline|Monoid|s''.
@@
 In addition to not being supported by Coq, option (1) would violate our policy of leaving components unbundled.
-For (2), if it could be made to work at all (which is not clear at all), F would need a pretty egregious type considering that arrow types are indexed by objects, and that the type of the arrow map (namely \lstinline|Π x y, (x ⟶ y) → (F x ⟶ F y)|) must refer to the object map.
+For (2), if it could be made to work at all, F would need a pretty egregious type considering that arrow types are indexed by objects, and that the type of the arrow map
+\begin{lstlisting}
+ Π x y, (x ⟶ y) → (F x ⟶ F y)
+\end{lstlisting}
+must refer to the object map.
 
@@
 To specify the natural numbers and the integers as initial objects in the categories of semirings and rings, respectively, definitions of these categories are needed.
-While one could define both of them manually, greater economy can be achieved by recognizing that both semirings and rings can be defined by equational theories, for which \emph{varieties} can be defined generically (varieties are categories consisting of models for a fixed theory with homomorphisms between them).
+While one could define both of them manually, greater economy can be achieved by recognizing that both semirings and rings can be defined by equational theories, for which \emph{varieties} can be defined generically.
+Varieties are categories consisting of models for a fixed theory with homomorphisms between them.
 
 To this end, we have formalized some of the theory of multisorted universal algebra and equational theories.
-We chose not to revive existing old (and somewhat stale) formalizations~\cite{DBLP:conf/tphol/Capretta99,dominguez2008formalizing} of universal algebra, because an important aim for us has been to find out what level of elegance, convenience, and integration can be achieved by leveraging the state of the art in Coq facilities (of which type classes are the most important example).
+We chose not to revive existing formalizations~\cite{DBLP:conf/tphol/Capretta99,dominguez2008formalizing} of universal algebra, because an important aim for us has been to find out what level of elegance, convenience, and integration can be achieved by leveraging the state of the art in Coq facilities (of which type classes are the most important example).
 
@@
 \end{lstlisting}
-In order to use this with \lstinline|Initial|, we define an additional conversion instance that takes a \lstinline|NaturalsToSemiRing| along with a proof showing that it yields \lstinline|SemiRing_Morphism|s, and builds an \lstinline|InitialArrow| instance out of it.
+To use this with \lstinline|Initial|, we define an additional conversion instance that takes a \lstinline|NaturalsToSemiRing| along with a proof showing that it yields \mbox{\lstinline|SemiRing_Morphism|s}, and builds an \lstinline|InitialArrow| instance out of it.
 This conversion instance in turn invokes another conversion function that translates concrete \lstinline|SemiRing_Morphism| proofs into univeral algebra \lstinline|Homomorphism|s instantiated with the semiring signature, which make up the arrows in the category.
@@
 Hence, theory and programs developed on our \lstinline|Rationals| interface applies and can make immediate use of these efficient rationals.
-We plan to rebase the computable real number implementation in CoRN on this interface, precisely so that it may be instantiated with efficient implementations like these.
+We plan to rebase the computable real number implementation~\cite{Oconnor:real} on this interface, precisely so that it may be instantiated with efficient implementations like these.
 
@@
 Traditionally, proof assistants such as Coq have provided sophisticated tactics or built-in commands to support this quoting.
-In~\cite{Hints}, Asperti et.
-al.
-show that unification hints, a very general way of facilitating user-defined extensions to term and type inference, can be used to semi-automatically build quote functions without dropping to a meta-level.\footnote{Gonthier provides similar functionality by ingeniously using canonical structures.} This feature is absent from Coq, but fortunately type classes also allow us to do this, as we will now show.
+Unification hints~\cite{Hints}, a very general way of facilitating user-defined extensions to term and type inference, can be used to semi-automatically build quote functions without dropping to a meta-level.\footnote{Gonthier provides similar functionality by ingeniously using canonical structures.} This feature is absent from Coq, but fortunately type classes also allow us to do this, as we will now show.
 
@@
 One source is the vast literature on efficient implementation of Prolog-style resolution, which the hint-based proof search used for instance resolution greatly resembles.
-
 We emphasize that these efficiency problems only affect type checking; efficiency of computation using type-checked terms is not affected.

### 2010-07-01 jar.tex
References for manifest fields, more precise about adjunctions

@@
 
-Manifest fields~\cite{DBLP:conf/types/Luo08} have been proposed, and implemented in the Matita system, to address exactly this problem.
+Manifest fields~\cite{Pollack:2002} have been proposed to address exactly this problem.
+In fact, a semblance has been implemented in the Matita system~\cite{sacerdoti2008working}.
 We hope to convince the reader that type system extensions of this nature, designed to mitigate particular symptoms of the bundled approach, are less elegant than a solution (described in the next section) that avoids the problem altogether by using predicate-like type classes in place of bundled records.
@@
 Formalizing the (nontrivial) proof that these two definitions are equivalent provides a nice test for our definitions.
-Preliminary results indicate that we can very concisely and closely follow the textbook proof.
+As a first step, we have constructed the unit and co-unit of the adjunction, thus proving MacLane's Theorem 1.
+We have concisely and closely followed his proof~\cite{CatWork}.
 % todo: ref?
@@
 The expression type is parameterized over the set of variable indices.
+Below we use an implicitly defined heap of such variables.
 Hence, we diverge from~\cite{Hints}, which uses \lstinline|nat| for variables, thereby introducing bound problems and a need for dummy variables.

### 2010-07-01 jar.tex
Clarify use of naturals_to_semiring, bigops.

@@
   {⟶}{{$\longrightarrow\ $}}1
-  {⇛}{{$\Rrightarrow\ $}}1
+  {⇛}{{$\Longrightarrow\ $}}1
   {∧}{{$\land$}}1
@@
 
-Some theory and programs about naturals are more easily proved, respectively defined, for concrete implementations and then lifted to the abstract interface.
-For relatively simple things like decidability of equality, all that is needed to make this lifting easy is a handful of simple lemmas translating the initiality property back into statements saying that \lstinline|naturals_to_semiring| is extensionally equal to other semiring morphisms and involutary between \lstinline|Naturals| implementations.
-For the example of decidability, these let us define decidability for an arbitrary \lstinline|Naturals| implementation simply by mapping to the \lstinline|nat| implementation using \lstinline|naturals_to_semiring|, deciding equality there, and reflecting the equality or inequality proof back to the abstract implementation using the fact that \lstinline|naturals_to_semiring| is an involution.
-
-For more complex properties such as injectivity of partially applied addition and multiplication, we take a longer detour.
+Since initial objects in categories are isomorphic, we can easily derive that \lstinline|naturals_to_semiring| gives isomorphisms between different \lstinline|Naturals| implementations:
+\begin{lstlisting}
+  Lemma iso_naturals `{Naturals A} `{Naturals B}:
+    Π a: A, naturals_to_semiring B A (naturals_to_semiring A B a) = a.
+\end{lstlisting}
+This is very useful, because some properties of and operations on naturals are more easily proved, respectively defined, for concrete implementations (such as \lstinline|nat|) and then `lifted' to the abstract \lstinline|Naturals| interface so that they work for arbitrary implementations.
+For example, while showing decidability for an arbitrary \lstinline|Naturals| implementation directly is tricky, it is very easy to show decidability for \lstinline|nat|.
+Using \lstinline|iso_naturals|, the latter can be very straightforwardly used to implement the former.
+
+To lift properties such as injectivity of partially applied addition and multiplication from \lstinline|nat| to arbitrary \lstinline|Naturals| implementations, we take a longer detour.
 As part of our universal algebra theory, we have proved that proofs of statements in the language of an equational theory can be transferred between isomorphic implementations.
-Combined with the trivially proved property that any two initial objects in a category are isomorphic, this lets us transfer proofs of such statements between implementations of \lstinline|Naturals|, requiring only that we reflect the concrete statement (expressed in terms of the operational type classes) to a symbolic statement in the language of semirings.
+Hence, we can transfer proofs of such statements between implementations of \lstinline|Naturals|, requiring only that we reflect the concrete statement (expressed in terms of the operational type classes) to a symbolic statement in the language of semirings.
 We intend to eventually make this reflection completely automatic using type class based quotation techniques along the lines of those described in Section~\ref{quoting}.
@@
 %    ; ... AltAdjunction stated using the bundled ... }.
-On top of this interface we can build theory about typical list operations such as maps, folds, their relation to \lstinline|singleton| and \lstinline|extend|, et cetera.
-We can also generically define `bigops' for summation and products, and easily show properties like distributivity, all without ever mentioning cons lists.
+On top of this interface we can build theory about typical sequence operations such as maps, folds, their relation to \lstinline|singleton| and \lstinline|extend|, et cetera.
+We can also generically define `big operators' for sums ($\sum$) and products ($\prod$) of sequences, and easily show properties like distributivity, all without ever mentioning cons lists.
 

### 2010-07-01 jar.tex
Improvements/corrections suggested by James.

@@
 On this foundation we build a set of mathematically sound abstract interfaces for different kinds of numbers, succinctly expressed using categorical language and universal algebra constructions.
-Strategic use of type classes lets us support these high-level theory-friendly definitions while still enabling efficient implementations unhindered by gratuitous indirection or conversion or projection.
-
-Algebra flourishes by the interplay between syntax and semantics.
+Strategic use of type classes lets us support these high-level theory-friendly definitions while still enabling efficient implementations unhindered by gratuitous indirection, conversion or projection.
+
+Algebra thrives on the interplay between syntax and semantics.
 The Prolog-like abilities of type class instance resolution allow us to conveniently define a quote function, thus facilitating the use of reflective techniques.
@@
 On a very practical level, this lets Coq safely erase all \lstinline|Prop| components when extracting certified programs to OCaml or Haskell.
-A stronger notion of proof irrelevance in which any two inhabitants of a \lstinline|Prop| type are made \emph{convertible} is planned for a future Coq version.
+
+% A stronger notion of proof irrelevance in which any two inhabitants of a \lstinline|Prop| type are made \emph{convertible} is planned for a future Coq version.
+  % commented out because not really relevant. only re-add with explanation of relevance.
+
 % todo: process?:
@@
 
-The `native' notion of equality in Coq is that of terms being convertible, which in this context constitutes Leibniz equality, naturally propositionalized by an inductive type family \lstinline|eq| with a single constructor
+The `native' notion of equality in Coq is that of terms being convertible, naturally reified as a proposition by the inductive type family \lstinline|eq| with single constructor \lstinline|eq_refl|:
 \begin{lstlisting}
@@
 
-The algebraic hierarchy of the ssreflect~\cite{Packed} libraries uses an interesting (if fairly restrictive) alternative approach.
+The algebraic hierarchy of the \textsc{Ssreflect} libraries~\cite{Packed} uses an interesting (if fairly restrictive) alternative approach.
 It requires unicity of representation for all objects so that Leibniz equality is always applicable.
@@
 
-Algebraic structures are expressed in terms of a number of carriers, a number of operations and relations on these carriers, and a number of laws that the operations and relations satisfy.
+Algebraic structures are expressed in terms of a number of carrier sets, a number of operations and relations on these carriers, and a number of laws that the operations and relations satisfy.
 In a system like Coq, we have different options when it comes to representing the grouping of these components.
@@
 More elaborate structures, too, can be expressed as predicates (expressing laws) over a number of carriers, relations, and operations.
-While optimally flexible in principle, in practice \emph{naive} adoption of this approach (that is, without using type classes) leads to substantial inconveniences in actual use: when \emph{stating} theorems about abstract instances of such structures, one must enumerate all components along with the structure (predicate) of interest.
+While optimally flexible in principle, in practice \emph{na\"ive} adoption of this approach (that is, without using type classes) leads to substantial inconveniences in actual use: when \emph{stating} theorems about abstract instances of such structures, one must enumerate all components along with the structure (predicate) of interest.
 And when \emph{applying} such theorems, one must either enumerate any non-inferrable components, or let the system spawn awkward metavariables to be resolved at a later time.
@@
 
-In the absence of type classes, these are all very real problems, and for this reason the two largest formalizations of abstract algebraic structures in Coq today, CoRN~\cite{C-corn} and Ssreflect~\cite{Packed}, both use \emph{bundled} representation schemes, using records with one or more of the components as fields instead of parameters.
+In the absence of type classes, these are all very real problems, and for this reason the two largest formalizations of abstract algebraic structures in Coq today, CoRN~\cite{C-corn} and \textsc{Ssreflect}~\cite{Packed}, both use \emph{bundled} representation schemes, using records with one or more of the components as fields instead of parameters.
 For reflexive relations, the following is a fully bundled representation---the other end of the spectrum:
@@
 
-We will tackle the problems associated with the structure predicates approach one by one, starting with those encountered during theorem \emph{application}.
+We will tackle the problems associated with the use of structure predicates one by one, starting with those encountered during theorem \emph{application}.
 Suppose we have defined \lstinline|SemiGroup| as a structure predicate as follows\footnote{Note that defining \lstinline|SemiGroup| as a record instead of a straight conjunction does not make it any less of a predicate.
@@
 
-The declarations of the two inherited \lstinline|CommutativeMonoid| structures nicely illustrate how predicate classes naturally support not just multiple inheritance, but \emph{overlapping} multiple inheritance, where the inherited structures may share components (in this case carrier and equivalence relation).
+The declarations of the two inherited \lstinline|CommutativeMonoid| structures in \lstinline|SemiRing| nicely illustrate how predicate classes naturally support not just multiple inheritance, but \emph{overlapping} multiple inheritance, where the inherited structures may share components (in this case carrier and equivalence relation).
 The carrier \lstinline|A|, being an explicit argument, is specified as normal.
-The equivalence relation, being an implicit argument of class type, is resolved automatically to \lstinline|e| and passed.
-The binary operation and constant would normally be automatically resolved and passed as well, but we override the inference mechanism locally using Coq's existing named argument facility (which is only syntactic sugar of the most superficial kind) in order to specify the correct pairings.
+The equivalence relation, being an implicit argument of class type, is resolved automatically to \lstinline|e|.
+The binary operation and constant would normally be automatically resolved as well, but we override the inference mechanism locally using Coq's existing named argument facility (which is only syntactic sugar of the most superficial kind) in order to explicitly pair multiplication with 1 for the first \lstinline|CommutativeMonoid| substructure, and addition with 0 for the second \lstinline|CommutativeMonoid| substructure.
 Again, contrast this with type system extensions such as Matita's manifest records, which are required to make this work when the records bundle components such as \lstinline|op| and \lstinline{unit} as \emph{fields} instead of parameters.
@@
   Record EquationalTheory :=
-    { et_sig:> Signature
-    ; et_laws:> EqEntailment et_sig → Prop }.
+    { eqt_sig:> Signature
+    ; eqt_laws:> EqEntailment eqt_sig → Prop }.
 \end{lstlisting}
@@
     { variety_algebra:> Algebra et carriers
-    ; variety_laws: Π s, et_laws et s → (Π $\hspace{-1.5mm}$vars, eval_stmt et vars s) }.
+    ; variety_laws: Π s, eqt_laws et s → (Π $\hspace{-1.5mm}$vars, eval_stmt et vars s) }.
 \end{lstlisting}
@@
 
-What remains is to show that carriers+\lstinline|Equivs|+\lstinline|AlgebraOps| satisfying \lstinline|InVariety| for a given \lstinline|EquationalTheory| do indeed form a \lstinline|Category| (the `variety').
+What remains is to show that carrier sets together with \lstinline|Equiv|s and \lstinline|AlgebraOps| satisfying \lstinline|InVariety| for a given \lstinline|EquationalTheory| do indeed form a \lstinline|Category| (the `variety').
 Since we need a type for the objects in the \lstinline|Category|, at this point we have no choice but to bundle components and proof together in a record:
@@
 \end{lstlisting}
-\lstinline|sigT| represents a \lstinline|Type|-sorted existential. \lstinline|ClosedSubset| is defined elsewhere as:
+The \lstinline|sigT| type constructor is a \lstinline|Type|-sorted existential quantifier. \lstinline|ClosedSubset| is defined elsewhere as:
 \begin{lstlisting}
@@
 \end{lstlisting}
-\lstinline|op_closed| is defined by recursion over the symbolic operation types.
-
-The reason we define \lstinline|image| and \lstinline|ClosedSubset| in \lstinline|Type| rather than \lstinline|Prop| is that since the final goal of the proof is to establish an isomorphism in the category of σ-algebras (where arrows are algebra homomorphisms), we will eventually need to map elements in the subalgebra defined by \lstinline|image| back to their pre-image in \lstinline|A|.
+Here, \lstinline|op_closed| is defined by recursion over the symbolic operation types.
+
+The reason we define \lstinline|image| and \lstinline|ClosedSubset| in \lstinline|Type| rather than in \lstinline|Prop| is that since the final goal of the proof is to establish an isomorphism in the category of σ-algebras (where arrows are algebra homomorphisms), we will eventually need to map elements in the subalgebra defined by \lstinline|image| back to their pre-image in \lstinline|A|.
 
@@
 \end{lstlisting}
-To use this with \lstinline|Initial|, we define an additional conversion instance that takes a \lstinline|NaturalsToSemiRing| along with a proof showing that it yields \mbox{\lstinline|SemiRing_Morphism|s}, and builds an \lstinline|InitialArrow| instance out of it.
+The instance for \lstinline|nat| is defined as follows:
+\begin{lstlisting}
+Instance nat_to_semiring: NaturalsToSemiRing nat :=
+  λ _ _ _ _ _ => fix f (n: nat) := match n with 0 => 0 | S m => f m + 1 end.
+\end{lstlisting}
+
+To use \lstinline|NaturalsToSemiRing| with \lstinline|Initial|, we define an additional conversion instance that takes a \lstinline|NaturalsToSemiRing| along with a proof showing that it yields \mbox{\lstinline|SemiRing_Morphism|s}, and builds an \lstinline|InitialArrow| instance out of it.
 This conversion instance in turn invokes another conversion function that translates concrete \lstinline|SemiRing_Morphism| proofs into univeral algebra \lstinline|Homomorphism|s instantiated with the semiring signature, which make up the arrows in the category.
@@
 
-Traditionally, proof assistants such as Coq have provided sophisticated tactics or built-in commands to support this quoting.
+Traditionally, proof assistants such as Coq have provided sophisticated tactics or built-in commands to support such \emph{quoting}.
 Unification hints~\cite{Hints}, a very general way of facilitating user-defined extensions to term and type inference, can be used to semi-automatically build quote functions without dropping to a meta-level.\footnote{Gonthier provides similar functionality by ingeniously using canonical structures.} This feature is absent from Coq, but fortunately type classes also allow us to do this, as we will now show.
@@
 Below we use an implicitly defined heap of such variables.
-Hence, we diverge from~\cite{Hints}, which uses \lstinline|nat| for variables, thereby introducing bound problems and a need for dummy variables.
+Hence, we diverge from~\cite{Hints}, which uses \lstinline|nat| for variable indices, thereby introducing a need for dummy variables for out-of-bounds indices.
 
@@
   Definition Value := nat.
-  Definition Vars V := V → Value.
-
-  Fixpoint eval {V} (vs: Vars V) (e: Expr V): Value :=
+  Definition Env V := V → Value.
+
+  Fixpoint eval {V} (vs: Env V) (e: Expr V): Value :=
     match e with
@@
 \begin{lstlisting}
-  Definition novars: Vars False := False_rect _.
-  Definition singlevar (x: Value): Vars unit := λ _ => x.
-  Definition merge {A B} (a: Vars A) (b: Vars B): Vars (A+B) :=
+  Definition novars: Env False := False_rect _.
+  Definition singlevar (x: Value): Env unit := λ _ => x.
+  Definition merge {A B} (a: Env A) (b: Env B): Env (A+B) :=
     λ i => match i with inl j => a j | inr j => b j end.
@@
 \begin{lstlisting}
-  Class Quote {V} (l: Vars V) (n: Value) {V'} (r: Vars V'): Type :=
+  Class Quote {V} (l: Env V) (n: Value) {V'} (r: Env V'): Type :=
     { quote: Expr (V + V')
@@
 \begin{lstlisting}
-  Instance quote_one V (v: Vars V): Quote v 1 novars := { quote := One }.
+  Instance quote_one V (v: Env V): Quote v 1 novars := { quote := One }.
 \end{lstlisting}
@@
 \begin{lstlisting}
-  Instance quote_mult V (v: Vars V) n V' (v': Vars V') m V'' (v'': Vars V'')
+  Instance quote_mult V (v: Env V) n V' (v': Env V') m V'' (v'': Env V'')
     `{!Quote v n v'} `{!Quote (merge v v') m v''}:
@@
 \begin{lstlisting}
-  Class Lookup {A} (x: Value) (v: Vars A) := { key: A; key_correct: v key = x }.
+  Class Lookup {A} (x: Value) (v: Env A) := { key: A; key_correct: v key = x }.
 \end{lstlisting}
@@
 \begin{lstlisting}
-  Context (x: Value) {A B} (va: Vars A) (vb: Vars B).
+  Context (x: Value) {A B} (va: Env A) (vb: Env B).
 
@@
 \begin{lstlisting}
-  Instance quote_old_var V (v: Vars V) x {Lookup x v}:
+  Instance quote_old_var V (v: Env V) x {Lookup x v}:
     Quote v x novars | 8 := { quote := Var (inl (key x v)) }.
@@
 \begin{lstlisting}
-  Instance quote_new_var V (v: Vars V) x: Quote v x (singlevar x) | 9
+  Instance quote_new_var V (v: Env V) x: Quote v x (singlevar x) | 9
     := { quote := Var (inr tt) }.
@@
 
-Using a small additional utility lemma we can also easily quote equalities into \lstinline|eval p e = eval p e'| form with a shared variable pack (so that identical expressions on both sides of the equation don't end up as different variables).
+The following additional utility lemma lets us quote equalities with a shared heap (so that an opaque expression that occurs on both sides of the equation is not represented by two distinct variables):
+\begin{lstlisting}
+  Lemma quote_equality {V} {v: Env V} {V'} {v': Env V'} (l r: Value)
+    `{!Quote novars l v} `{!Quote v r v'}:
+      let heap := merge v v' in
+      eval heap (map_var shift quote) = eval heap quote → l = r.
+\end{lstlisting}
 
@@
 On top of this interface we can build theory about typical sequence operations such as maps, folds, their relation to \lstinline|singleton| and \lstinline|extend|, et cetera.
-We can also generically define `big operators' for sums ($\sum$) and products ($\prod$) of sequences, and easily show properties like distributivity, all without ever mentioning cons lists.
-
-Unfortunately, tragedy strikes when, after having defined this theory, we try to show that regular cons lists implement the abstract \lstinline|Sequence| interface.
+We can also generically define `big operators' for sums ($\sum$) and products ($\prod$) of sequences, and easily show properties like distributivity, all without ever mentioning cons-lists.
+
+Unfortunately, disaster strikes when, after having defined this theory, we try to show that regular cons-lists implement the abstract \lstinline|Sequence| interface.
 When we get to the point where we want to define the \lstinline|Singleton| operation, Coq emits a universe inconsistency error.
@@
 Georges Gonthier and Claudio Sacerdoti Coen have helped to
-sharpen our understanding of the relation with Canonical Structures and with Unification Hints.
+sharpen our understanding of the relation with Canonical Structures and with Unification Hints. Finally, James McKinna suggested many corrections and improvements.
 

### 2010-07-01 jar.tex
Use fewer scare quotes.

@@
 They are handled as specialized syntactic constructs whose semantics are given specifically by the type class apparatus.
-By contrast, the expressivity of dependent types and inductive families as supported in Coq, combined with the use of pre-existing technology in the system (namely proof search and implicit arguments) enable a \emph{first class} type class implementation~\cite{DBLP:conf/tphol/SozeauO08}: classes are ordinary record types (`dictionaries'), instances are ordinary constants of these record types (registered as `hints' with the proof search machinery), class constraints are ordinary implicit parameters, and instance resolution is achieved by augmenting the unification algorithm to invoke ordinary proof search for implicit arguments of class type.
+By contrast, the expressivity of dependent types and inductive families as supported in Coq, combined with the use of pre-existing technology in the system (namely proof search and implicit arguments) enable a \emph{first class} type class implementation~\cite{DBLP:conf/tphol/SozeauO08}: classes are ordinary record types (`dictionaries'), instances are ordinary constants of these record types (registered as \emph{hints} with the proof search machinery), class constraints are ordinary implicit parameters, and instance resolution is achieved by augmenting the unification algorithm to invoke ordinary proof search for implicit arguments of class type.
 Thus, type classes in Coq are realized by relatively minor syntactic aids that bring together existing facilities of the theory and the system into a coherent idiom, rather than by introduction of a new category of qualitatively different definitions with their own dedicated semantics.
@@
 \end{lstlisting}
-Superficially, this instantly solves the problems described above: reflexive relations can now be declared and passed as self-contained packages, and the \lstinline|rr_rel| projection now constitutes a ``canonical name'' for relations that are known to be reflexive, and we could bind a notation to it.
+Superficially, this instantly solves the problems described above: reflexive relations can now be declared and passed as self-contained packages, and the \lstinline|rr_rel| projection now constitutes a canonical name for relations that are known to be reflexive, and we could bind a notation to it.
 While there is no conventional notation for reflexive relations, the situation is the same in the context of, say, a semiring, where we would bind $+$ and $*$ notations to the record field projections for the addition and multiplication operations, respectively.
@@
 \end{lstlisting}
-We are not particularly worried about overly large contexts, especially because most of the `extra' hypotheses we have compared to bundled approaches are declarations of relations, operators, and constants, which are all in some sense `inert' with respect to proof search.
+We are not particularly worried about overly large contexts, especially because most of the `extra' hypotheses we have compared to bundled approaches are declarations of relations, operators, and constants, which are all in some sense inert with respect to proof search.
 Hence, we do not foresee problems with large contexts for any but the most complex formalizations.
@@
 Fields declared with \lstinline|:>| are registered as hints for instance resolution, so that in any context where \lstinline|(A, =, +, 0, *, 1)| is known to be a \lstinline|SemiRing|, \lstinline|(A, =, +, 0)| and \lstinline|(A, =, *, 1)| are automatically known to be \lstinline|CommutativeMonoid|s (and so on, transitively, because instance resolution is recursive).
-In our hierarchy, these `substructures' by themselves establish the following inheritance diagram:
+In our hierarchy, these substructures by themselves establish the following inheritance diagram:
 
@@
 This kind of redundant knowledge ()which arises naturally) is never a problem in our setup, because the use of operational type classes ensures that terms composed of algebraic operations and relations never refer to structure proofs.
-We find that this is a tremendous relief compared to approaches that \emph{do} intermix the two and where one must be careful to ensure that such terms refer to the `right' proofs of properties.
+We find that this is a tremendous relief compared to approaches that \emph{do} intermix the two and where one must be careful to ensure that such terms refer to the \emph{right} proofs of properties.
 There, even \emph{strong} proof irrelevance (which would make terms convertible that differ only in what proofs they refer to) would not make these difficulties go away entirely, because high-level tactics that rely on quotation of terms require syntactic identity (rather than `mere' convertibility) to recognize identical subterms.
@@
 
-Taking these conversion functions for granted, we will also need a ``down-to-earth'' representation of the initiality arrows if we are to give a \lstinline|SemiRing|-based definition of the interface for natural numbers.
+Taking these conversion functions for granted, we will also need a ``down to earth'' representation of the initiality arrows if we are to give a \lstinline|SemiRing|-based definition of the interface for natural numbers.
 Once again, we introduce an operational type class to represent this particular component:
@@
 \end{lstlisting}
-This is very useful, because some properties of and operations on naturals are more easily proved, respectively defined, for concrete implementations (such as \lstinline|nat|) and then `lifted' to the abstract \lstinline|Naturals| interface so that they work for arbitrary implementations.
+This is very useful, because some properties of and operations on naturals are more easily proved, respectively defined, for concrete implementations (such as \lstinline|nat|) and then \emph{lifted} to the abstract \lstinline|Naturals| interface so that they work for arbitrary implementations.
 For example, while showing decidability for an arbitrary \lstinline|Naturals| implementation directly is tricky, it is very easy to show decidability for \lstinline|nat|.

### 2010-07-01 jar.tex
Sacrifice !'s in parameter declarations to improve presentation.

@@
 \begin{lstlisting}
-  Class Category (O: Type) `{!Arrows O} `{Π x y: O, Equiv (x ⟶ y)}
-      `{!CatId O} `{!CatComp O}: Prop :=
+  Class Category (O: Type) `{Arrows O} `{Π x y: O, Equiv (x ⟶ y)}
+      `{CatId O} `{CatComp O}: Prop :=
     { arrow_equiv:> Π x y, Setoid (x ⟶ y)
@@
     (et: EquationalTheory) (carriers: sorts et → Type)
-    {e: Π a, Equiv (carriers a)} `{!AlgebraOps et carriers}: Prop :=
+    {e: Π a, Equiv (carriers a)} `{AlgebraOps et carriers}: Prop :=
     { variety_algebra:> Algebra et carriers
@@
 \begin{lstlisting}
-  Context `{Algebra σ$$ A} `{Algebra σ$$ B} `{!HomoMorphism σ$$ A B f}.
+  Context `{Algebra σ$$ A} `{Algebra σ$$ B} `{HomoMorphism σ$$ A B f}.
 \end{lstlisting}
@@
 \begin{lstlisting}
-  Class Naturals (A: ObjectInVariety semiring_theory) `{!InitialArrow A}: Prop :=
+  Class Naturals (A: ObjectInVariety semiring_theory) `{InitialArrow A}: Prop :=
     { naturals_initial:> Initial A }.
@@
 \begin{lstlisting}
-  Context `{SemiRing A} `{!NaturalsToSemiRing A}.
+  Context `{SemiRing A} `{NaturalsToSemiRing A}.
   Class Naturals: Prop :=
@@
   Instance quote_mult V (v: Env V) n V' (v': Env V') m V'' (v'': Env V'')
-    `{!Quote v n v'} `{!Quote (merge v v') m v''}:
+    `{Quote v n v'} `{Quote (merge v v') m v''}:
       Quote v (n * m) (merge v' v'') :=
@@
 
-  Instance lookup_left `{!Lookup x va}: Lookup x (merge va vb)
+  Instance lookup_left `{Lookup x va}: Lookup x (merge va vb)
     := { key := inl (key x va) }.
 
-  Instance lookup_right `{!Lookup x vb}: Lookup x (merge va vb)
+  Instance lookup_right `{Lookup x vb}: Lookup x (merge va vb)
     := { key := inr (key x vb) }.
@@
   Lemma quote_equality {V} {v: Env V} {V'} {v': Env V'} (l r: Value)
-    `{!Quote novars l v} `{!Quote v r v'}:
+    `{Quote novars l v} `{Quote v r v'}:
       let heap := merge v v' in

### 2010-07-02 compile.sh
rename

@@ -1,6 +1,6 @@
 #!/bin/sh
 
-cp jar.tex tmp.tex
+cp mscs.tex tmp.tex
 dot -Tpdf hierarchy.dot > hierarchy.pdf
 
 set -e
@@ -9,5 +9,5 @@ bibtex tmp
 pdflatex tmp.tex
 bibtex tmp
 pdflatex tmp.tex
-cat tmp.pdf > jar.pdf
+cat tmp.pdf > mscs.pdf
 rm tmp.*

### 2010-07-07 papers/mscs/mscs2egu.tex
Reorganize files.

% This is file MSCS2egu.tex
% v1.2 released 3rd October 2001
%   (based on MSCSguid.tex for LaTeX 2.09)
% Copyright (C) 1997-2001 Cambridge University Press

\NeedsTeXFormat{LaTeX2e}

\documentclass{mscs}

\title[Mathematical Structures in Computer Science]
  {\LaTeXe\ guide for MSCS authors}
\author[\LaTeXe\ guide]
  {C\ls A\ls M\ls B\ls R\ls I\ls D\ls G\ls E\ns
   \TeX\ls -\ls T\ls O\ls -\ls T\ls Y\ls P\ls E}
\date{October 2001}

\pubyear{2001}
\pagerange{\pageref{firstpage}--\pageref{lastpage}}
\volume{00}
\doi{S0960129501003425}

\newcommand\eg{\textit{e.g.\ }}
\newcommand\etc{\textit{etc}}

\newtheorem{lemma}{Lemma}[section]

\begin{document}

\label{firstpage}
\maketitle

\begin{abstract}
This guide is for authors who are preparing papers for the
\emph{Journal of Mathematical Structures in Computer Science}
using the \LaTeX\ document preparation system and the CUP MSCS class file.
\end{abstract}

\tableofcontents

\ifprodtf \newpage \else \vspace*{-1\baselineskip}\fi

\section{Introduction}

The layout design for the \emph{Journal of Mathematical Structures
in Computer Science} has been implemented as a \LaTeX\ class file.
The MSCS class is based on the \verb"ARTICLE" class as discussed in
the \LaTeX\ manual.
Commands which differ from the standard \LaTeX\ interface, or which are
provided in addition to the standard interface, are explained in this
guide. This guide is not a substitute for the \LaTeX\ manual itself.

\ifprodtf \else \newpage\fi

\subsection{Introduction to \LaTeX}

The \LaTeX\ document preparation system is a special version of the
\TeX\ typesetting program.
\LaTeX\ adds to \TeX\ a collection of commands which simplify typesetting for
the author by allowing him/her to concentrate on the logical structure of the
document rather than its visual layout.

\LaTeX\ provides a consistent and comprehensive document preparation
interface. There are simple-to-use commands for generating a table of
contents, lists of figures and/or tables, and indexes.
\LaTeX\ can automatically number list entries, equations, figures,
tables, and footnotes, as well as parts, sections and subsections.
Using this numbering system, bibliographic citations, page references and
cross references to any other numbered entity (\eg sections, equations,
figures, list entries) are quite straightforward.

\subsection{The MSCS document class}

The use of document classes allows a simple change of style (or
style option) to transform the appearance of your document.
The CUP MSCS class file preserves the standard \LaTeX\ interface
such that any document which can be  produced using the standard
\LaTeX\ \verb"ARTICLE" class, can also be produced with the MSCS class.
However, the measure (or width of text) is wider than for \verb"ARTICLE",
therefore line breaks will change and long equations may need re-setting.
Authors should not introduce hard line and page breaks into the
text. The finished pages will be typeset in the Times typeface, while
most authors use Computer Modern. This change of typeface will alter both
line and page breaks.

It is essential that you do not modify \verb"mscs.cls". When
additional macros are required, place them in a separate file with
the filename extension \verb".sty" (e.g. \verb"mymacros.sty").
Then load these macros with \verb"\usepackage":
\begin{verbatim}
  \documentclass{mscs}
\end{verbatim}
\begin{verbatim}
  \usepackage{mymacros}
\end{verbatim}
In this way you will distinguished clearly between the main class file
and your own macros, which can then be found easily by the journal editor.
Do not forget to submit your optional style file for publication along
with your input file.

\section{Using the MSCS class}

First, copy the file \verb"mscs.cls" into the correct subdirectory on your
system. The MSCS document class is implemented as a complete document
class \emph{not} a document class option.
In order to use the MSCS class, replace \verb"article" by \verb"mscs" in
the \verb"\documentclass" command at the beginning of your document:
\begin{verbatim}
  \documentclass{article}
\end{verbatim}
is replaced by,
\begin{verbatim}
  \documentclass{mscs}
\end{verbatim}

\subsection{Document class options}

In general, the standard document class options should \emph{not} be
used with the MSCS class:
\begin{itemize}
  \item \texttt{10pt}, \texttt{11pt}, \texttt{12pt} -- unavailable.
  \item \texttt{draft}, \texttt{twoside} (no associated style file) --
        \texttt{twoside} is the default.
  \item \texttt{fleqn}, \texttt{leqno}, \texttt{titlepage}, \texttt{twocolumn} --
        unavailable.
  \item \texttt{proc}, \texttt{ifthen} -- these `packages' can be used if necessary.
\end{itemize}
%
\ifprodtf
The following new class options are provided:
\begin{itemize}
  \item \texttt{prodtf} -- tells the class file that we want to use the
    production typeface. This option also resets the odd, even and top
    margins.
\end{itemize}
\fi

\section{Additional facilities}

In addition to all the standard \LaTeX\ design elements, the MSCS class
includes the following features:
\begin{itemize}
  \item Additional commands for typesetting the title page. Extended
        commands for specifying a short version of the title and author(s)
        for the running headlines.
  \item Full width and narrow figures and tables.
  \item Unnumbered Theorem-like environments (produced with \verb"\newtheorem").
  \item A \verb"proof" environment.
  \item Control of enumerated lists.
\end{itemize}
Once you have used any of these additional facilities in your document,
you should not process it with a standard \LaTeX\ class file.

\subsection{Titles and author's names}

In the MSCS class, the title of the article and the author's name (or
authors' names) are used both at the beginning of the article for the
main title and throughout the article as running headlines at the top of
every page. The title is used on odd-numbered pages (rectos) and the
author's name appears on even-numbered pages (versos).
Although the main heading can run to several lines of text, the running
head line must be a single line.
Moreover, the main heading can also incorporate new line commands
(e.g.\ \verb"\\") but these are not acceptable in a running headline.
To enable you to specify an alternative short title and author's name, the
standard \verb"\title" and \verb"\author" commands have been extended to
take an optional argument to be used as the running headline:
\begin{verbatim}
  \title[A short title]{The full title which can be as long
                        as necessary}
  \author[Author's name]{The full author's name \\
                         with affiliation if necessary}
\end{verbatim}

The following example, taken from Danvy and Filinski (1992), shows how
this is done with more than one author, with different affiliations
and a grant acknowledgement.
\begin{verbatim}
  \title[Representing control]
    {Representing control:\\ a study of the CPS transformation}
  \author[O. Danvy and A. Filinski]
    {O\ls L\ls I\ls V\ls I\ls E\ls R\ns D\ls A\ls N\ls V\ls Y$^1$%
     \thanks{This work was partly supported by NSF under grant
      CCR-9102625.} \ns and\ns A\ls N\ls D\ls R\ls Z\ls E\ls J\ns
      F\ls I\ls L\ls I\ls N\ls S\ls K\ls I$^2$\\
      $^1$ Department of Computing and Information
      Sciences, Kansas State University, Manhattan,\addressbreak
      Kansas 66506, USA.
      \addressbreak $^2$ School of Computer Science, Carnegie Mellon
      University, Pittsburgh, Pensylvania 15213, USA.}
\end{verbatim}
Note the use of \verb"\addressbreak". It is
needed when an affiliation extends into 2 or more lines and is inserted
at the beginning of each affiliation line except the last. Note also the
use of \verb"\ls" and \verb"\ns" respectively to letterspace the authors'
names and insert an appropriate space between the names.

Once you have used these extended versions of \verb"\author" and
\verb"\title", do not use them with a standard \LaTeX\ class file.

\subsection{Figures and Tables}

The \texttt{figure} and \texttt{table} environments are implemented as described
in the \LaTeX\ Manual to provide consecutively numbered floating inserts
for illustrations and tables respectively.
The standard inserts and their captions are formatted unjustified on a
restricted (30~pica) measure.

Wide figures and tables which require the full measure can
be produced using the \texttt{figure*} and \texttt{table*} environments which
are normally used to provide double-column inserts in two-columned
documents. For example,
\begin{verbatim}
  \begin{figure*}
    \vspace{12cm}
    \caption{Test of the Gibbs adsorption isotherm for
         oxygen on copper (Bauer, Speiser \& Hirth, 1976).}
  \end{figure*}
\end{verbatim}
These wide inserts and their captions are formatted unjustified over the
full text width.

\subsection{Theorems}

Any theorem-like environments which are required should be defined
as usual in the document preamble using the \verb"\newtheorem" command.
The default for theorem numbering is by section (\eg 3.1) and is
reset at the beginning of each new section.

The \verb"\newtheorem" command also generates an unnumbered version of
any theorem-like environments defined. These can be accessed using the
\verb"*"-form of the environment. e.g.
\begin{verbatim}
  \newtheorem{lemma}{Lemma}[section]
  \begin{lemma}
    If $\mathbf{PP}_0 \vdash a\colon A$, then there exists a
    closed $\lambda$-term.
  \end{lemma}

  \begin{lemma*}
    Conversely, if $\mathbf{PN}_0 \vdash a\colon A$, then
    there exists a closed SK-term $a^\circ$.
  \end{lemma*}
\end{verbatim}
Which produces:
  \begin{lemma}
    If $\mathbf{PP}_0 \vdash a\colon A$, then there exists a
    closed $\lambda$-term.
  \end{lemma}

  \begin{lemma*}
    Conversely, if $\mathbf{PN}_0 \vdash a\colon A$, then
    there exists a closed SK-term $a^\circ$.
  \end{lemma*}
The standard optional argument [\ ] may also be used with any
unnumbered environments, created with \verb"\newtheorem". A new
command (\verb"\removebrackets") is also provided which allows you
to remove the automatic parenthesis from around the optional
argument in the output, thus alowing further flexability.

\subsection{Proofs}

The \verb"proof" environment has been added to provide a consistent
format for proofs. For example,
\begin{verbatim}
  \begin{proof}
    Use $K_\lambda$ and $S_\lambda$ to translate combinators
    into $\lambda$-terms. For the converse, translate
    $\lambda x$ \ldots by [$x$] \ldots and use induction
    and the lemma.
  \end{proof}
\end{verbatim}
produces the following text:
\begin{proof}
  Use $K_\lambda$ and $S_\lambda$ to translate combinators
  into $\lambda$-terms. For the converse, translate
  $\lambda x$ \ldots by [$x$] \ldots and use induction
  and the lemma.
\end{proof}

The final \usebox{\proofbox} will not be included if the \verb"proof*"
environment is used. Note that the proof-box is drawn by a new macro
\verb"\proofbox" and can be placed in text by typing \linebreak
\verb"\usebox{\proofbox}".

The proof environment can also take an optional argument which allows you
to produce `special' proofs, e.g.
%
  \begin{proof}[Proof of Theorem~1.6.]
  Use $K_\lambda$ and $S_\lambda$ to translate combinators
  into $\lambda$-terms. For the converse, translate
  $\lambda x$ \ldots by [$x$] \ldots and use induction
  and the lemma.
  \end{proof}
%
Which was produced like this:
%
\begin{verbatim}
  \begin{proof}[Proof of Theorem~1.6.]
  Use $K_\lambda$ and $S_\lambda$ to translate combinators
  into $\lambda$-terms. For the converse, translate
  $\lambda x$ \ldots by [$x$] \ldots and use induction
  and the lemma.
  \end{proof}
\end{verbatim}
%
Notice that once the optional argument is used, you have to type all of
the text which is to appear as the heading (including the final full-stop).

\subsection{Lists}

The MSCS class provides the three standard list environments plus an
additional unnumbered list:
\begin{itemize}
  \item Numbered lists, created using the \verb"enumerate" environment.
  \item Bulleted lists, created using the \verb"itemize" environment.
  \item Labelled lists, created using the \verb"description" environment.
\end{itemize}
The enumerated list numbers each list item with an arabic numeral;
alternative styles can be achieved by inserting a redefinition of the
number labelling command after the \verb"\begin{enumerate}".
For example, a list numbered with roman numerals inside parentheses can be
produced by the following commands:
\begin{verbatim}
    \begin{enumerate}
     \renewcommand{\theenumi}{(\roman{enumi})}
     \item first item
           :
    \end{enumerate}
\end{verbatim}
This produces the following list:
\begin{enumerate}
  \renewcommand{\theenumi}{(\roman{enumi})}
  \item first item
  \item second item
  \item etc...
\end{enumerate}
In the last example, the label for item 3 (iii) clashed with text
because the standard list indentation is designed to be sufficient for
arabic numerals rather than the wider roman numerals.
In order to enable different labels to be used more easily, the
\verb"enumerate" environment in the MSCS class can be given an optional
argument which (like a standard \verb"bibliography" environment)
specifies the \emph{widest label}.
For example,
\begin{enumerate}[(iii)]
\renewcommand{\theenumi}{(\roman{enumi})}
  \item first item
  \item second item
  \item etc...
\end{enumerate}
was produced by the following input:
\begin{verbatim}
    \begin{enumerate}[(iii)]
     \renewcommand{\theenumi}{(\roman{enumi})}
     \item first item
           :
    \end{enumerate}
\end{verbatim}
Remember, once you have used the optional argument on the \verb"enumerate"
environment, do not process your document with a standard \LaTeX\ class
file.

\section{Some guidelines for using standard facilities}

The following notes may help you achieve the best effects with the MSCS
class file.

\subsection{Sections}

\LaTeX\ provides five levels of section headings and they are all
defined in the MSCS class file:
\begin{itemize}
  \item Heading A -- \verb"\section".
  \item Heading B -- \verb"\subsection".
  \item Heading C -- \verb"\subsubsection".
  \item Heading D -- \verb"\paragraph".
  \item Heading E -- \verb"\subparagraph".
\end{itemize}
Section numbers are given for sections, subsection and subsubsection
headings.

One additional heading is provided: \verb"\xhead". This
heading is provided to obtain unnumbered `A' headings inside appendices
(normally \verb"\section"(\verb"*") commands inside appendices have the word
`Appendix' pre-appended).

\subsection{Running headlines}

As described above, the title of the article and the author's name (or
authors' names) are used as running headlines at the top of every page.
The title is used on odd-numbered pages (rectos) and the author's name
appears on even-numbered pages (versos).

The \verb"\pagestyle" and \verb"\thispagestyle" commands should
\emph{not} be used.
Similarly, the commands \verb"\markright" and \verb"\markboth" should not
be necessary.

\subsection{Illustrations (or figures)}

The MSCS class will cope with most positioning of your illustrations and
tables and you should not normally use the optional positional qualifiers
on the \verb"figure" environment which would override these decisions.
Figure captions should be below the figure itself therefore the
\verb"\caption" command should appear after the figure or space left for
an illustration. For example, Figure~\ref{sample-figure} is produced
using the following commands:
\begin{figure}
  \vspace{3cm}
  \caption{An example figure with space for artwork.}
  \label{sample-figure}
\end{figure}
\begin{verbatim}
  \begin{figure}
    \vspace{3cm}
    \caption{An example figure with space for artwork.}
    \label{sample-figure}
  \end{figure}
\end{verbatim}

\subsection{Tables}

The MSCS class will cope with most positioning of your illustrations and
tables and you should not normally use the optional positional qualifiers
on the \verb"table" environment which would override these decisions.
Table captions should be at the top therefore the \verb"\caption" command
should appear before the body of the table.
For example, Table~\ref{sample-table} is produced using the following
commands:
%
\begin{table}
  \caption{An example table}
    \begin{tabular}{cccc}
     \hline
     \hline
     Figure & $hA$ & $hB$ & $hC$\\
     \hline
     2 & $\exp\left(\pi i\frac58\right)$
       & $\exp\left(\pi i\frac18\right)$ & $0$\\
     3 & $-1$    & $\exp\left(\pi i\frac34\right)$ & $1$\\
     4 & $-4+3i$ & $-4+3i$ & $\frac74$\\
     5 & $-2$    & $-2$    & $\frac54 i$ \\
     \hline
     \hline
    \end{tabular}
  \label{sample-table}
\end{table}
\begin{verbatim}
  \begin{table}
    \caption{An example table}
    \begin{tabular}{cccc}
     \hline \hline
     Figure & $hA$ & $hB$ & $hC$\\
     \hline
     2 & $\exp\left(\pi i\frac58\right)$
       & $\exp\left(\pi i\frac18\right)$ & $0$\\
     3 & $-1$    & $\exp\left(\pi i\frac34\right)$ & $1$\\
     4 & $-4+3i$ & $-4+3i$ & $\frac74$\\
     5 & $-2$    & $-2$    & $\frac54 i$ \\
     \hline \hline
    \end{tabular}
    \label{sample-table}
  \end{table}
\end{verbatim}
%
The \verb"tabular" environment should be used to produce ruled tables;
it has been modified for the MSCS class so that additional vertical space
is inserted on either side of a rule (produced by \verb"\hline"). Commands
to redefine quantities such as \verb"\arraystretch" should be omitted.

\subsubsection{Continued captions.}

If a table or figure will not fit onto a single page and has to broken into
more than one part, the subsequent parts should be should have a caption like:
\hbox{Fig. 1 \textit{continued}.} or \hbox{Table 1 \textit{continued}.} To achieve
this, use \verb"\contcaption" instead of \verb"\caption", in the subsequent
figure or table environments.  The \verb"\contcaption" uses exactly the same
syntax as the normal \verb"\caption" command, except it does not step
the counter.

\subsection{Appendices}

Use the standard \LaTeX\ \verb"\appendix" command. After this any
\verb"\section" commands will start a new appendix.

Appendices should normally be placed after the main sections
and before an acknowledgement and the list of references. This ensures
that the references can be found easily at the end of the article.
Occasionally, appendices are so different from the main text, as in
this document, that they should be placed at the end.

\subsection{Bibliography}

As with standard \LaTeX, there are two ways of producing
a bibliography; either by compiling a list of references
by hand (using a \verb"thebibliography" environment), or
by using Bib\TeX\ with a suitable bibliographic database.

\subsubsection{Citations in the text.}

MSCS journal class uses the author-date system for
references. \LaTeX\ does not handle this system
effectively and many authors simply key the reference
citations by hand. This method is
accepted by the journal, though authors are encouraged to
use \verb"\cite" and the \verb"\bibitem" label because these
help with consistent citations and efficient editing.

In standard \LaTeX\ you make a citation such as
\cite{JCR} by typing \verb"\cite{JCR}". The key \verb"JCR"
establishes the cross-reference to the bibliography (see below)
and \verb"\cite" can extract the \verb"\bibitem" label to create
the in-text citation. Citations such as Toyn \linebreak \emph{et al.}
(1987) cannot be produced in this way and must be typed in the text.

In MSCS the \verb"\cite" command has been extended (using code from
\verb"apalike.sty") so that a list of references can be generated.
For instance, \cite{DEK,AJ1,TDR} was produced by typing
\verb"\cite{DEK,AJ1,TDR}".

\subsubsection{The list of references.}

The following listing shows examples of references
prepared in the style of the journal and they are
typeset below. Note that the \verb"\bibitem" arguments
\verb"[<label>]" and \verb"{<key>}" are only needed if
you are using \verb"\cite".
%
\begin{verbatim}
\begin{thebibliography}{}
 \bibitem[Augustsson and Johnsson 1987]{AJ1}
   Augustsson,~L. and Johnsson,~T. (1987) LML users' manual. PMG
   Report, Department of Computer Science, Chalmers University of
   Technology, Goteborg, Sweden.
 \bibitem[Conklin 1987]{JC}
   Conklin,~J. (1987) Hypertext: an introduction and survey.
   \emph{IEEE Computer} \textbf{20}~(9), 17--41.
\bibitem[Dijkstra 1976]{EWD}
   Dijkstra,~E.\,W. (1976) \emph{A Discipline of Programming}.
   Prentice-Hall.
\bibitem[Knuth 1984]{DEK}
   Knuth,~D.\,E. (1984) Literate programming. \emph{BCS Comput.\ J.}
   \textbf{27}~(2), 97--111 (May).
\bibitem[Danvy and Filinski 1992]{DO}
   Danvy, O. and Filinski, A. (1992) Representing control: a study
   of the CPS transformation \emph{Math. Struct. in Comp. Science}
   \textbf{2}~(2), 361--391.
\bibitem[Reynolds 1968]{JCR}
   Reynolds,~J.\,C. (1969) Transformation systems and the
   algebraic structure of atomic formulas. In B. Meltzer and
   D. Michie (editors), \emph{Machine Intelligence} \textbf{5}, 135--151.
   Edinburgh University Press.
\bibitem[Toyn \emph{et al.} 1987]{TDR}
   Toyn,~I., Dix,~A. and Runciman,~C. (1987) Performance
   polymorphism. In \emph{Functional Programming Languages and
   Computer Architecture, Lecture Notes in Computer Science}
   \textbf{274}, 325--346. Springer-Verlag.
\end{thebibliography}
\end{verbatim}
%
\begin{thebibliography}{}
 \bibitem[Augustsson and Johnsson 1987]{AJ1}
   Augustsson,~L. and Johnsson,~T. (1987) LML users' manual. PMG
   Report, Department of Computer Science, Chalmers University of
   Technology, Goteborg, Sweden.
 \bibitem[Conklin 1987]{JC}
   Conklin,~J. (1987) Hypertext: an introduction and survey.
   \emph{IEEE Computer} \textbf{20}~(9), 17--41.
\bibitem[Dijkstra 1976]{EWD}
   Dijkstra,~E.\,W. (1976) \emph{A Discipline of Programming}.
   Prentice-Hall.
\bibitem[Knuth 1984]{DEK}
   Knuth,~D.\,E. (1984) Literate programming. \emph{BCS Comput.\ J.}
   \textbf{27}~(2), 97--111 (May).
\bibitem[Danvy and Filinski 1992]{DO}
   Danvy, O. and Filinski, A. (1992) Representing control: a study
   of the CPS transformation \emph{Math. Struct. in Comp. Science}
   \textbf{2}~(2), 361--391.
\bibitem[Reynolds 1968]{JCR}
   Reynolds,~J.\,C. (1969) Transformation systems and the
   algebraic structure of atomic formulas. In B. Meltzer and
   D. Michie (editors), \emph{Machine Intelligence} \textbf{5}, 135--151.
   Edinburgh University Press.
\bibitem[Toyn \emph{et al.} 1987]{TDR}
   Toyn,~I., Dix,~A. and Runciman,~C. (1987) Performance
   polymorphism. In \emph{Functional Programming Languages and
   Computer Architecture, Lecture Notes in Computer Science}
   \textbf{274}, 325--346. Springer-Verlag.
\end{thebibliography}

\newpage
\appendix

\section{Special commands in \textmd{\texttt{mscs.cls}}}

The following is a summary of the new commands, optional
arguments and environments which have been added to the
standard \LaTeX\ user-interface in creating the MSCS class file.

\vspace{6pt}

\begin{tabular}{lp{8cm}}
\emph{New commands}     & \\
\verb"\addressbreak"    & allows you to break long affiliations into shorter lines.\\
\verb"\contcaption"     & for continued captions.\\
\verb"\ls"              & to letter space the author's name. \\
\verb"\ns"              & to insert space between an author's names. \\
\verb"\proofbox"        & used with \verb"\usebox" to place a proofbox
                          in text. \\
\verb"\removebrackets"  & removes the automatic parenthesis from around
                          the optional argument in environments created
                          by \verb"\newtheorem".\\
\verb"\xhead"           & used in Appendices to produce unnumbered
                          \verb"\section"s (without the appended `Appendix').\\[6.5pt]
\emph{New optional arguments} & \\
\verb"[<short title>]"  & in the \verb"\title" command: to define a right running
                          headline that is different from the article title. \\
\verb"[<short author>]" & in the \verb"\author" command: to define a left running
                          headline with text that is different from the
                          authors' names as typeset at the article opening. \\
\verb"[<widest label>]" & in \verb"\begin{enumerate}": to ensure the correct alignment
                          of numbered lists.\\[6.5pt]
\emph{New environments}  & \\
\verb"figure*"          & this environment is redefined in \verb"mscs.cls" to typeset
                          a long figure legend to the full text-width. \\
\verb"table*"           & this environment is redefined in \verb"mscs.cls" to typeset
                          wide tables, so that the table caption and rules are the
                          full text-width. \\
\verb"proof"            & to typeset mathematical proofs. \\
\verb"proof*"           & to typeset mathematical proofs without the
                          terminating proofbox.
\end{tabular}

\ifprodtf \else \newpage\fi
\section{Notes for editors}

\setcounter{subsubsection}{0}

This appendix contains additional information which may be useful to
those who are involved with the final production stages of an article.
Authors, who are generally not typesetting the final pages in the
journal's typeface, do not need this information.

\ifprodtf
\subsubsection{Setting the production typeface.}

The global \verb"\documentclass" option `\verb"prodtf"' sets up
\verb"mscs.cls" to typeset in the production typeface -- Monotype Times.
e.g.
%
\begin{verbatim}
  \documentclass[prodtf]{mscs}
\end{verbatim}
\fi

\subsubsection{Catchline commands.}

To be placed in the preamble:
\begin{itemize}
  \item \verb"\pubyear{}"
  \item \verb"\volume{}"
  \item \verb"\pagerange{}"
  \item \verb"\doi{}"
\end{itemize}

\subsubsection{Footnotes.}

If a footnote falls at the bottom of a page, it is possible for the
footnote to appear on the following page (a feature of \TeX ).
Check for this.

\subsubsection{Table footnotes and long captions.}

Set the table in a minipage that has the same width as the table body.
The caption will be set to the same width as the table and the footnotes
will fall at the bottom of the minipage.

\subsubsection{Font sizes.}

The MSCS class file defines all the standard \LaTeX\ font sizes:
\begin{itemize}
  \item \verb"\indexsize" -- {\indexsize This is indexsize text.}
  \item \verb"\footnotesize" -- {\footnotesize This is footnotesize text.}
  \item \verb"\small" -- {\small This is small text.}
  \item \verb"\normalsize" -- This is normalsize text (default).
  \item \verb"\LARGE" -- {\LARGE This is LARGE text.}
\end{itemize}
%
All these sizes are summarized in Table~\ref{tab:fontsizes}.
%
\begin{table*}[hbt]
\caption{Type sizes for \LaTeX\ size-changing commands.}
\label{tab:fontsizes}
 \begin{tabular}{lcl}
 \hline\hline
 \textit{Size}        &  \textit{size/baseline} & \textit{Usage}\\
 \hline
 \verb"\tiny"         &  5pt/6pt  & -- \\
 \verb"\scriptsize"   &  7pt/8pt  & -- \\
 \verb"\indexsize"    &  8pt/9pt  & index entries. \\
 \verb"\footnotesize" &  8pt/10pt & footnotes, table body. \\
 \verb"\small"        &  9pt/12pt & abstract, references, quote, \\
                      &           & quotations, received date, \\
                      &           & affiliation, figure captions. \\
 \verb"\normalsize"   & 10pt/13pt & main text size, author,\\
                      &           & section headings, etc. \\
 \verb"\large"        & 12pt/14pt & -- \\
 \verb"\Large"        & 14pt/18pt & -- \\
 \verb"\LARGE"        & 17pt/21pt & article title, part title. \\
 \verb"\huge"         & 20pt/25pt & -- \\
 \verb"\Huge"         & 25pt/30pt & -- \\
 \hline\hline
 \end{tabular}
\end{table*}

\ifprodtf
%
\newcommand\lra{\quad\longrightarrow\quad}

\section{Macros provided by {\tt mscs2esy.sty}}

\subsection{Automatic font/character changes}

\begin{itemize}\itemsep=6pt
\item The \verb|\le|, \verb|\leq|, \verb|\ge|, \verb|\geq| commands
use the equivalent AMS slanted symbols:
\[
\oldle \oldleq \oldge \oldgeq
 \lra
\le \leq \ge \geq
\]
The normal characters can be obtained by using the \verb|\old| form
(\eg \verb|\oldge|).
\end{itemize}

\subsection{Additional fonts}

\begin{itemize}\itemsep=6pt
\item The complete (v1) AMS symbols are available using the normal names:
\[
\hbox{\verb"\boxdot \boxplus \boxtimes"} \lra
  \boxdot \boxplus \boxtimes
\]
The \verb"\emptyset" ($\emptyset$) symbol is substituted from the
CUP Pi fonts.  The normal empty set symbol ($\oldemptyset$) can be obtained by
using \verb"\oldemptyset".

\item Blackboard bold:
\[
\hbox{\verb"$\mathbb{ABC}$"} \lra \mathbb{ABC}
\]

\item Fraktur/Gothic:
\[
   \hbox{\verb"$\mathfrak{ABC}$"} \lra \mathfrak{ABC}
\]

\item Monotype Script and bold Script:\\[6pt]
\verb"  $\mathscr{ABCabc}$ " $\lra \mathscr{ABCabc}$\\
\verb"  $\mathbscr{ABCabc}$" $\lra \mathbscr{ABCabc}$

\item Bold math italic/symbols are provided by the \verb"\boldsymbol" command
(\verb"\bmath" is provided as an alias). \verb"mscs2esy" also defines most of the
symbols from Appendix F of the \TeX book. These can be obtained by using
their normal (unbold) symbol name prefixed with a `b'. \eg \verb|\nabla|
becomes \verb|\bnabla|. The only exception to this rule is \verb|\eta|,
which whould lead to a clash with \verb|\beta|. In this case use
\verb|\boldeta| for bold eta.

\item Upright Greek: The \verb"\mathup" and \verb"\mathbup" macros are provided
(by \verb"upmath.sty") to obtain upright (and bold upright) lower-case Greek characters.
All of the lower-case Greek characters are pre-defined.\\[6pt]
%
\verb"  $\ualpha$" $\lra \ualpha$ \qquad
\verb"$\ubalpha$"  $\lra \ubalpha$

\item Sans serif symbols:
The following commands provide easy access to the various sans serif faces, in text
and math mode.\\[6pt]
%
\verb"  \textsf{text}  " $\lra$ \textsf{text}
  \qquad \verb"\mathsf{math}  " $\lra \mathsf{math}$\\
\verb"  \textsfi{text} " $\lra$ \textsfi{text}
  \qquad \verb"\mathsfi{math} " $\lra \mathsfi{math}$\\
\verb"  \textsfb{text} " $\lra$ \textsfb{text}
  \qquad \verb"\mathsfb{math} " $\lra \mathsfb{math}$\\
\verb"  \textsfbi{text}" $\lra$ \textsfbi{text}
  \qquad \verb"\mathsfbi{math}" $\lra \mathsfbi{math}$\\[6pt]

\end{itemize}
%
\fi

\label{lastpage}

\end{document}

### 2010-07-07 papers/itp10-roughdiamond/roughdiamond.tex

--- 2010-03-26 roughdiamond.tex
@@
 
-Conventional wisdom warns that while this approach is theoretically very flexible (as we will see momentarily), one risks extreme inconvenience both in having to declare and pass around all these structural components all the time, as well as in losing notations (because we no longer project named operations out of records).
+Conventional wisdom warns that while this approach is theoretically very flexible, one risks extreme inconvenience both in having to declare and pass around all these structural components all the time, as well as in losing notations (because we no longer project named operations out of records).
 
@@
 
-First, multiple inheritance (a major complication in current solutions) becomes trivial: |SemiRing| inherits two |Monoid| structures on the same carrier and setoid relation, using ordinary named arguments (rather than dedicated extensions~\cite{DBLP:conf/types/Luo08}) to achieve ``manifest fields''.
+First, multiple inheritance becomes trivial: |SemiRing| inherits two |Monoid| structures on the same carrier and setoid relation, using ordinary named arguments (rather than dedicated extensions~\cite{DBLP:conf/types/Luo08}) to achieve ``manifest fields''.
 
@@
 
-Third, since our structural type classes are mere predicates, overlap between their instances---another worry with existing solutions---is a non-issue.
+Third, since our structural type classes are mere predicates, overlap between their instances is a non-issue.
 Together with the previous point, this gives us tremendous freedom to posit multiple structures on the same operations and relations, including ones derived implicitly via subclasses: by simply declaring a |SemiRing| class instance showing that a ring is a semiring, results about semirings immediately apply implicitly to any known ring, without us having to explicitly encode this relation in the hierarchy definition itself, and without needing any projection or translation of carriers or operations.
@@
 Realizing that much code duplication for the various algebraic structures in the hierarchy could be avoided by employing universal algebra constructions, we then proceeded to formalize some of the theory of multisorted universal algebra and equational theories, using it to automatically construct varieties of algebras.
-We purposely did not try to use existing Coq formalizations~\cite{DBLP:conf/tphol/Capretta99,dominguez2008formalizing} of universal algebra for this, because an important goal for us is to find out exactly what level of elegance, convenience, and integration can be achieved using the state of the art technology offered by the Coq system (of which type classes are the most important instance).
+We avoided existing formalizations~\cite{DBLP:conf/tphol/Capretta99,dominguez2008formalizing} of universal algebra, because we aimed to find out what level of elegance, convenience, and integration can be achieved using the state of the art technology (of which type classes are the most important instance).
 
-At the time of writing, our development includes a fully integrated formalization of a nontrivial portion of category theory and multisorted universal algebra, including various categories (e.g.
-the category |Cat| of categories, and generic variety categories which we instantiate to obtain the categories of monoids, semirings, and rings), functors (including automatically generated forgetful functors), natural transformations, adjunctions, initial models of equational theories constructed from term algebras, transference of proofs between isomorphic models of equational theories, subalgebras, congruences, quotients, products, and the first homomorphism theorem.
+At the time of writing, our development includes a fully integrated formalization of a nontrivial portion of category theory and multisorted universal algebra, including various categories (e.g.\ the category |Cat| of categories, and generic variety categories which we instantiate to obtain the categories of monoids, semirings, and rings), functors (including automatically generated forgetful functors), natural transformations, adjunctions, initial models of equational theories constructed from term algebras, transference of proofs between isomorphic models of equational theories, subalgebras, congruences, quotients, products, and the first homomorphism theorem.
 

### 2010-07-07 papers/mscs/defManSSR.tex

--- 2010-05-28 defManSSR.tex
@@
 	    destruct, induction, using, refine, inversion, injection,
-         rewrite, congr, unlock, compute, ring, field, fourier, 
+         rewrite, congr, unlock, compute, field, fourier,
             replace, fold, unfold, change, cutrewrite, simpl,
@@
 morekeywords=[5]{
-         by, done, exact, reflexivity, tauto, romega, omega,
+         by, done, exact, tauto, romega, omega,
          assumption, solve, contradiction, discriminate},

### 2010-11-23 papers/mscs/mscs.tex
Replace Π's with ∀'s in MSCS paper.

--- 2010-07-01 jar.tex
@@
 \newcommand{\Q}{\ensuremath{\mathbbm{Q}}}
+
 \usepackage{color}
@@
   {∧}{{$\land$}}1
-  {∀}{{$\forall$}}1
-  {Π}{{$\Uppi$ }}1
+  {∀}{{$\forall\ $}}1
   {η}{{$\upeta$}}1
@@
 \begin{lstlisting}
-  eq_refl : Π (T: Type) (x: T), x ≡ x,
+  eq_refl : ∀ (T: Type) (x: T), x ≡ x,
 \end{lstlisting}
@@
 \begin{lstlisting}
-  Definition reflexive {A: Type} (R: relation A): Prop := Π a, R a a.
+  Definition reflexive {A: Type} (R: relation A): Prop := ∀ a, R a a.
 \end{lstlisting}
@@
     ; rr_rel: relation rr_carrier
-    ; rr_proof: Π x, rr_rel x x }.
+    ; rr_proof: ∀ x, rr_rel x x }.
 \end{lstlisting}
@@
   Theorem example G e op {P: SemiGroup G e op}:
-    Π x y z, e (op (op x y) z) (op (op z y) x).
+    ∀ x y z, e (op (op x y) z) (op (op z y) x).
 \end{lstlisting}
@@
 \begin{lstlisting}
-  Π (G: Type) (e: Equiv G) (op: SemiGroupOp G), Prop
+  ∀ (G: Type) (e: Equiv G) (op: SemiGroupOp G), Prop
 \end{lstlisting}
@@
   Theorem example G e op {P: SemiGroup G e op}:
-    Π x y z, x & y & z = z & y & x.
+    ∀ x y z, x & y & z = z & y & x.
 \end{lstlisting}
@@
   Theorem example G e op {P: SemiGroup G e op}:
-    Π x y z, equiv e (sg_op op (sg_op op x y) z) (sg_op op (sg_op op z y) x).
+    ∀ x y z, equiv e (sg_op op (sg_op op x y) z) (sg_op op (sg_op op z y) x).
 \end{lstlisting}
@@
 \begin{lstlisting}
-  Theorem example `{SemiGroup G}: Π x y z: G, x & y & z = z & y & x.
+  Theorem example `{SemiGroup G}: ∀ x y z: G, x & y & z = z & y & x.
 \end{lstlisting}
@@
 \lstinline|Decision| is a very general-purpose type class that also works for predicates.
-For instance, to declare a parameter expressing decidability of, say, (setoid) equality on a type \lstinline|X|, we write: \lstinline|`{Π a b: X, Decision (a = b)}|.
+For instance, to declare a parameter expressing decidability of, say, (setoid) equality on a type \lstinline|X|, we write: \lstinline|`{∀ a b: X, Decision (a = b)}|.
 To then use this (unnamed) decider to decide a particular equality, we simply say \lstinline|decide (x = y)|, and instance resolution will resolve the decider we declared.
@@
   Class CatComp O `{Arrows O} :=
-    comp: Π {x y z}, (y ⟶ z) → (x ⟶ y) → (x ⟶ z).
+    comp: ∀ {x y z}, (y ⟶ z) → (x ⟶ y) → (x ⟶ z).
 
@@
 \begin{lstlisting}
-  Class Category (O: Type) `{Arrows O} `{Π x y: O, Equiv (x ⟶ y)}
+  Class Category (O: Type) `{Arrows O} `{∀$$x y: O, Equiv (x ⟶ y)}
       `{CatId O} `{CatComp O}: Prop :=
-    { arrow_equiv:> Π x y, Setoid (x ⟶ y)
-    ; comp_proper:> Π x y z, Proper (equiv ==> equiv ==> equiv) comp
+    { arrow_equiv:> ∀ x y, Setoid (x ⟶ y)
+    ; comp_proper:> ∀ x y z, Proper (equiv ==> equiv ==> equiv) comp
     ; comp_assoc w x y z (a: w ⟶ x) (b: x ⟶ y) (c: y ⟶ z):
@@
   Context `{Category X}.
-  Class InitialArrows (x: X): Type := initial_arrow: Π y, x ⟶ y.
+  Class InitialArrows (x: X): Type := initial_arrow: ∀ y, x ⟶ y.
   Class Initial (x: X) `{InitialArrows x}: Prop :=
-    initial_arrow_unique: Π y (a: x ⟶ y), a = initial_arrow y.
+    initial_arrow_unique: ∀ y (a: x ⟶ y), a = initial_arrow y.
 \end{lstlisting}
@@
   Class Fmap: Type :=
-    fmap: Π {v w: C}, (v ⟶ w) → (map_obj v ⟶ map_obj w).
+    fmap: ∀ {v w: C}, (v ⟶ w) → (map_obj v ⟶ map_obj w).
 
@@
     ; functor_to: Category D
-    ; functor_morphism:> Π a b: C, Setoid_Morphism (@fmap _ a b)
+    ; functor_morphism:> ∀ a b: C, Setoid_Morphism (@fmap _ a b)
     ; preserves_id: `(fmap (cat_id: a ⟶ a) = cat_id)
@@
 \begin{lstlisting}
- Π x y, (x ⟶ y) → (F x ⟶ F y)
+ ∀ x y, (x ⟶ y) → (F x ⟶ F y)
 \end{lstlisting}
@@
 \begin{lstlisting}
-  Notation "F ⇛ G" := (Π x, F x ⟶ G x).
+  Notation "F ⇛ G" := (∀ x, F x ⟶ G x).
 \end{lstlisting}
@@
     ; naturaltrans_to: Functor G
-    ; natural: Π `(f: x ⟶ y), η$$ y ◎$$ fmap F f = fmap G f ◎$$ η$$ x }.
+    ; natural: ∀ `(f: x ⟶ y), η$$ y ◎$$ fmap F f = fmap G f ◎$$ η$$ x }.
 \end{lstlisting}
@@
 \begin{lstlisting}
-  Class Adjunction (φ: Π `(F c ⟶ d), (c ⟶ G d)): Prop :=
+  Class Adjunction (φ: ∀ `(F c ⟶ d), (c ⟶ G d)): Prop :=
     { adjunction_left_functor: Functor F _
@@
 \begin{lstlisting}
-  Class AltAdjunction (η: id ⇛ G ∘ F) (φ: Π `(f: c ⟶ G d), F c ⟶ d): Prop :=
+  Class AltAdjunction (η: id ⇛ G ∘ F) (φ: ∀ `(f: c ⟶ G d), F c ⟶ d): Prop :=
     { alt_adjunction_natural_unit: NaturalTransformation η
-    ; alt_adjunction_factor: Π `(f: c ⟶ G d),
+    ; alt_adjunction_factor: ∀ `(f: c ⟶ G d),
         is_sole ((f =) ∘ (◎ η$$ c) ∘ fmap G) (φ f) }.
@@
   Class AlgebraOps (σ: Signature) (carriers: sorts σ$$ → Type)
-    := algebra_op: Π o: operation σ, op_type carriers (σ o).
+    := algebra_op: ∀ o: operation σ, op_type carriers (σ o).
 \end{lstlisting}
@@
 \begin{lstlisting}
-  Class Algebra `{Π a, Equiv (carriers a)} `{AlgebraOps σ$$ carriers}: Prop :=
-      { algebra_setoids:> Π a, Setoid (carriers a)
-      ; algebra_propers:> Π o: σ, Proper (=) (algebra_op o) }.
+  Class Algebra `{∀$$a, Equiv (carriers a)} `{AlgebraOps σ$$ carriers}: Prop :=
+      { algebra_setoids:> ∀ a, Setoid (carriers a)
+      ; algebra_propers:> ∀ o: σ, Proper (=) (algebra_op o) }.
 \end{lstlisting}
@@
     (et: EquationalTheory) (carriers: sorts et → Type)
-    {e: Π a, Equiv (carriers a)} `{AlgebraOps et carriers}: Prop :=
+    {e: ∀ a, Equiv (carriers a)} `{AlgebraOps et carriers}: Prop :=
     { variety_algebra:> Algebra et carriers
-    ; variety_laws: Π s, eqt_laws et s → (Π $\hspace{-1.5mm}$vars, eval_stmt et vars s) }.
-\end{lstlisting}
-  % todo: why do i need the manual \hspace!?
+    ; variety_laws: ∀ s, eqt_laws et s → (∀$$vars, eval_stmt et vars s) }.
+\end{lstlisting}
 
@@
     { variety_carriers:> sorts et → Type
-    ; variety_equiv: Π a, Equiv (variety_carriers a)
+    ; variety_equiv: ∀ a, Equiv (variety_carriers a)
     ; variety_op: AlgebraOps et variety_carriers
@@
   Context `{Algebra σ$$ A}.
-  Class Congruence (e: Π s: sorts σ, relation (v s)): Prop :=
-    { congruence_proper:> Π s, Proper (equiv ==> equiv ==> iff) (e s)
+  Class Congruence (e: ∀ s: sorts σ, relation (v s)): Prop :=
+    { congruence_proper:> ∀ s, Proper (equiv ==> equiv ==> iff) (e s)
     ; congruence_quotient:> Algebra σ$$ v (e:=e) }.
@@
 \begin{lstlisting}
-  Context `{Algebra σ$$ A} (P: Π s, A s → Type).
+  Context `{Algebra σ$$ A} (P: ∀ s, A s → Type).
   Class ClosedSubset: Type :=
-    { subset_proper: Π s x x', x = x' → iffT (P s x) (P s x')
-    ; subset_closed: Π o, op_closed (algebra_op o) }.
+    { subset_proper: ∀ s x x', x = x' → iffT (P s x) (P s x')
+    ; subset_closed: ∀ o, op_closed (algebra_op o) }.
 \end{lstlisting}
@@
   Class NaturalsToSemiRing (A: Type) :=
-    naturals_to_semiring: Π B `{RingMult B} `{RingPlus B} `{RingOne B}
+    naturals_to_semiring: ∀ B `{RingMult B} `{RingPlus B} `{RingOne B}
       `{RingZero B}, A → B.
@@
     { naturals_ring:> SemiRing A
-    ; naturals_to_semiring_mor:> Π `{SemiRing B},
+    ; naturals_to_semiring_mor:> ∀ `{SemiRing B},
         SemiRing_Morphism (naturals_to_semiring A B)
@@
   Lemma iso_naturals `{Naturals A} `{Naturals B}:
-    Π a: A, naturals_to_semiring B A (naturals_to_semiring A B a) = a.
+    ∀ a: A, naturals_to_semiring B A (naturals_to_semiring A B a) = a.
 \end{lstlisting}
@@
 \begin{lstlisting}
-Definition calculate_things `{Naturals N} `{Π$$n m: N, Decision (n = m)}
+Definition calculate_things `{Naturals N} `{∀$$n m: N, Decision (n = m)}
   (a b: nat): ... := ... decide (a = b) ... .
@@
     { integers_ring:> Ring A
-    ; integers_to_ring_mor:> Π `{Ring B},
+    ; integers_to_ring_mor:> ∀ `{Ring B},
         Ring_Morphism (integers_to_ring A B)
@@
 \begin{lstlisting}
-  Context `{Field A} `{Π x y: A, Decision (x = y)} {inj_inv}.
+  Context `{Field A} `{∀ x y: A, Decision (x = y)} {inj_inv}.
   Class Rationals: Prop :=
@@
 
-We also plan to provide an abstract interface for polynomials as a free commutative algebra. This would unify existing implementations such as coefficient lists and Bernstein polynomials.
+We also plan to provide an abstract interface for polynomials as a free commutative algebra.
+This would unify existing implementations such as coefficient lists and Bernstein polynomials; see~\cite{ZumkellerPhD} for the latter.
 
@@
 \begin{lstlisting}
-  Goal Π x y (P: Value → Prop), P ((x * y) * (x * 1)).
+  Goal ∀ x y (P: Value → Prop), P ((x * y) * (x * 1)).
     intros.
@@
 
-  Class Extend := extend: Π {x y} `{SemiGroupOp y} `{MonoidUnit y},
+  Class Extend := extend: ∀ {x y} `{SemiGroupOp y} `{MonoidUnit y},
     (x → y) → (seq x → y).
-  Class Singleton := singleton: Π x, x → seq x.
+  Class Singleton := singleton: ∀ x, x → seq x.
 \end{lstlisting}
@@
   Class Sequence
-   `{Π a, MonoidUnit (seq a)} `{Π a, SemiGroupOp (seq a)}
-   `{Π a, Equiv a → Equiv (seq a)} `{Singleton} `{Extend}: Prop := ...
+   `{∀ a, MonoidUnit (seq a)} `{∀ a, SemiGroupOp (seq a)}
+   `{∀ a, Equiv a → Equiv (seq a)} `{Singleton} `{Extend}: Prop := ...
 \end{lstlisting}

### 2010-11-25 papers/mscs/mscs.tex
Fix typo.

@@
 Since \lstinline|CommutativeMonoid| indirectly includes a \lstinline|SemiGroup| field, which in turn includes a \lstinline|Equivalence| field, having a \lstinline|SemiRing| proof means having two distinct proofs that the equality relation is an equivalence.
-This kind of redundant knowledge ()which arises naturally) is never a problem in our setup, because the use of operational type classes ensures that terms composed of algebraic operations and relations never refer to structure proofs.
+This kind of redundant knowledge (which arises naturally) is never a problem in our setup, because the use of operational type classes ensures that terms composed of algebraic operations and relations never refer to structure proofs.
 We find that this is a tremendous relief compared to approaches that \emph{do} intermix the two and where one must be careful to ensure that such terms refer to the \emph{right} proofs of properties.

### 2010-11-26 papers/mscs/mscs.tex
First batch of MSCS paper modifications addressing concerns brought up by referees.

@@
 More elaborate structures, too, can be expressed as predicates (expressing laws) over a number of carriers, relations, and operations.
-While optimally flexible in principle, in practice \emph{na\"ive} adoption of this approach (that is, without using type classes) leads to substantial inconveniences in actual use: when \emph{stating} theorems about abstract instances of such structures, one must enumerate all components along with the structure (predicate) of interest.
+While optimally flexible in principle, in practice \emph{naive} adoption of this approach (that is, without using type classes) leads to substantial inconveniences in actual use: when \emph{stating} theorems about abstract instances of such structures, one must enumerate all components along with the structure (predicate) of interest.
 And when \emph{applying} such theorems, one must either enumerate any non-inferrable components, or let the system spawn awkward metavariables to be resolved at a later time.
@@
 For reflexive relations, the following is a fully bundled representation---the other end of the spectrum:
+\newpage
 \begin{lstlisting}
@@
 
-We will tackle the problems associated with the use of structure predicates one by one, starting with those encountered during theorem \emph{application}.
+To show that the fully unbundled approach with structures represented by predicates can be made feasible using type classes, we will tackle one by one the problems traditionally associated with their use, starting with those encountered during theorem \emph{application}.
 Suppose we have defined \lstinline|SemiGroup| as a structure predicate as follows\footnote{Note that defining \lstinline|SemiGroup| as a record instead of a straight conjunction does not make it any less of a predicate.
@@
 While it is certainly true that in contexts with redundant information it can become hard to predict which instance of a predicate class will be found by proof search, it simply \emph{does not matter} which one is found.
-Moreover, for operational type classes the issue rarely arises because their instances are not nearly as abundant.
+Moreover, for operational type classes the issue rarely arises because their instances are not nearly as abundant, and are systematically shared.
 
@@
 
-We use names for properties like distributivity and absorption, because these are type classes as well (which is why we declare them with \lstinline|:>|).
+We use names for properties like distributivity and absorption, because these are type classes as well (which is why we declare their instances with \lstinline|:>|).
 It has been our experience that almost any generic predicate worth naming is worth representing as a predicate type class, so that its proofs will be resolved as instances behind the scenes whenever possible.
 Doing this consistently minimizes administrative noise in the code, bringing us closer to ordinary mathematical vernacular.
-Indeed, we believe that type classes are an elegant and apt formalization of the seemingly casual manner in which ordinary mathematical presentation assumes implicit administration and use of a `database' of properties previously proved.
-Much more so than the existing canonical structures facility, because canonical structures can only be used with bundled representations, which we argued against in section~\ref{bundling}.
+Indeed, we believe that type classes provide an elegant and apt formalization of the seemingly casual manner in which ordinary mathematical presentation assumes implicit administration and use of a `database' of properties previously proved.
 
@@
 
-The \lstinline|monmor_to| and \lstinline|monmor_from| fields are not an absolute necessity, but eliminate the need for theorems to declare \lstinline|Monoid| parameters when they already declare \lstinline|Monoid_Morphism| parameters.
-This is an instance where we can freely posit structural properties without worrying about potential problems when such information turns out to be redundant in contexts where the source and target of the morphism are already known to be \lstinline|Monoid|s.
+We include the \lstinline|monmor_to| and \lstinline|monmor_from| fields because it does not make much sense to talk about monoid morphisms between non-monoids, and having these fields removes the need for \lstinline|Monoid| class constraints when we are already parameterizing definitions or theory on a \lstinline|Monoid_Morphism|.
+On the other hand, we will also wish to talk about monoid morphisms between \emph{known} monoids, and in these cases the fields will be strictly redundant.
+As mentioned before, it is a strength of our approach that such redundant knowledge is entirely harmless, so that we may freely posit these structural properties whenever they make sense and provide convenience, without risking rebundling tar-pits or projection path ambiguities down the line.
 
@@
 \end{lstlisting}
-Given an interpretation of the sorts (mapping each symbolic sort to a carrier), an interpretation for the operations is easily written as an operational type class:
-\begin{lstlisting}
-  Variables (σ: Signature) (carriers: sorts σ → Type).
-
-  Definition op_type: ne_list (sorts σ) → Type := fold (→) ∘ map carrier.
-
-  Class AlgebraOps (σ: Signature) (carriers: sorts σ$$ → Type)
-    := algebra_op: ∀ o: operation σ, op_type carriers (σ o).
+Given an interpretation of the sorts (mapping each symbolic sort to a carrier type), interpretations of the operations are easily represented by an operational type class:
+\begin{lstlisting}
+  Variables (σ: Signature) (carriers: sorts σ$$ → Type).
+
+  Class AlgebraOps :=
+    algebra_op: ∀ o: operation σ, fold (→$\hspace{-1mm}$) (map carriers (operation_type σ$$ o)).
 \end{lstlisting}
@@
 \begin{lstlisting}
-  Class Algebra `{∀$$a, Equiv (carriers a)} `{AlgebraOps σ$$ carriers}: Prop :=
+  Class Algebra `{∀$$a, Equiv (carriers a)} `{AlgebraOps}: Prop :=
       { algebra_setoids:> ∀ a, Setoid (carriers a)
@@
 \end{lstlisting}
-The \lstinline|(=)| referred to in \lstinline|algebra_propers| is an \lstinline|Equiv| instance expressing setoid-respecting extensionality for the function types generated by \lstinline|op_type|.
+The \lstinline|(=)| referred to in \lstinline|algebra_propers| is an automatically derived \lstinline|Equiv| instance expressing setoid-respecting extensionality for the function types produced by the \lstinline|fold| in \lstinline|AlgebraOps|.
 
@@
 
-To characterize structures such as semirings and rings, we need equational theories, consisting of a signature together with laws (represented by a predicate over equality entailments):
+To adequately characterize structures such as semirings and rings, we need not just a signature that enumerates and gives the types of their operations, but also a specification of the axioms (laws) that these operations must satisfy.
+For this, we define \lstinline|EquationalTheory| as a signature together with a set of laws, the latter represented by a predicate over equality entailments:
 \begin{lstlisting}
@@
 \begin{theorem}[First homomorphism theorem]
-If $A$ and $B$ are algebras, and $f$ is a homomorphism from $A$ to $B$, then the equivalence relation defined by $a\sim b$ if and only if $f(a)=f(b)$ is a congruence on $A$, and the algebra $A/\hspace{-1mm}\sim$ is isomorphic to the image of $f$, which is a subalgebra of B.
+If $A$ and $B$ are algebras, and $f$ is a homomorphism from $A$ to $B$, then the equivalence relation $\sim$ defined by ``$a\sim b \leftrightarrow f(a)=f(b)$'' is a congruence on $A$, and the quotient algebra $A/\hspace{-1mm}\sim$ is isomorphic to the image of $f$, which is a subalgebra of B.
 \end{theorem}
@@
 \begin{lstlisting}
-  Context `{Algebra σ$$ A} `{Algebra σ$$ B} `{HomoMorphism σ$$ A B f}.
-\end{lstlisting}
-With \lstinline|~| defined as described, the first part of the proof is simply the definition of the following instance:
+  Context `{HomoMorphism σ$$ A B f}.
+\end{lstlisting}
+With \lstinline|~| defined as indicated, the first part of the proof is simply the definition of the following instance:
 \begin{lstlisting}
@@
 Unfortunately, Coq's universe polymorphism is not yet up to the task of letting us use a single set of definitions to handle both cases.
-In particular, there is no universe polymorphism for definitions (as opposed to inductive definitions) yet.
+In particular, there is no universe polymorphism for ordinary definitions (as opposed to inductive definitions) yet.
 We will return to this point later.

### 2010-12-08 papers/mscs/mscs.tex
Couple more referee-guided improvements in MSCS paper.

@@
 \end{lstlisting}
-Then by (1) making \lstinline|SemiGroup| a \emph{class} (by replacing the \lstinline|Record| keyword with the \lstinline|Class| keyword), (2) marking its proofs as \emph{instances} (again by replacing a single keyword), and (3) marking the \lstinline|SemiGroup| parameter of semigroup theorems as implicit (by using curly instead of round brackets), we no longer have to pass \lstinline|SemiGroup| proofs around manually ourselves, letting instance resolution do it for us instead.
+Then by (1) making \lstinline|SemiGroup| a \emph{class} (by replacing the \lstinline|Record| keyword with the \lstinline|Class| keyword), (2) marking its proofs as \emph{instances} (by replacing the \lstinline|Lemma| keyword with the \lstinline|Instance| keyword), and (3) marking the \lstinline|SemiGroup| parameter of semigroup theorems as implicit (by using curly instead of round brackets), we no longer have to pass \lstinline|SemiGroup| proofs around manually ourselves, letting instance resolution do it for us instead.
 Because instance resolution is part of the unifier, this also works when the statement of the theorem we wish to apply only mentions some of the components (which admittedly doesn't make much sense for semigroups).
@@
 For ease of presentation we show only a proof of concept for a very concrete language.
-We are currently working to integrate this technique with our existing universal algebra infrastructure, in particular its data types for algebraic terms.%  Preliminary results in this area (such as a simple monoid normalization tactic) suggest yada.
+We are currently working to integrate this technique with our existing universal algebra infrastructure.
+In particular, the latter's term data type should be ideally suited to serve as a generic symbolic representation of terms in a wide class of algebras.
+This should let us implement the basic setup of the technique once and for all, so that quotation for new algebraic structures can be enabled with minimal effort.
+  %  Preliminary results in this area (such as a simple monoid normalization tactic) suggest yada.
 

### 2010-12-09 papers/mscs/mscs.tex
Still more referee-guided improvements in MSCS paper.

@@
 In section~\ref{predicateclasses} we show how type classes can make the use of `unbundled' purely predicate based interfaces for abstract structures practical.
-Next, in section~\ref{classes}, we discuss our algebraic hierarchy implemented using such predicate classes.
+
+In the rest of the paper, we tour the key components in our development, leading up to the numerical interfaces.
+This will not only show the pleasant style of formalization that rigorous use of type classes enables, but will also show that eager adoption and incorporation of more abstract mathematical perspectives (which are traditionally often ignored when doing dependently typed programming on concrete data structures in type theory), is not only feasible but actually practical and beneficial.
+
+ In section~\ref{classes}, we discuss our algebraic hierarchy implemented using type classes.
 In sections~\ref{cat} and~\ref{univ} we give a taste of what category theory and universal algebra look like in our development, and in section~\ref{numbers} we use these facilities to build abstract interfaces for numbers.
 In order to illustrate a very different use of type classes, we discuss the implementation of a quoting function for algebraic terms in terms of type classes, in section~\ref{quoting}.
-In section~\ref{sequences} we hint at an interface for sequences, but describe how a limitation in the Coq system makes its use problematic.
+In section~\ref{sequences} we hint at an interface for sequences, but describe how a limitation in the current implementation of Coq makes its use problematic.
 We end with conclusions in section~\ref{conclusions}.
@@
 
-% \paragraph{Equality, setoids, and rewriting}
+\paragraph{Equality, setoids, and rewriting}
 
@@
 where `\lstinline|a ≡ b|' is notation for \lstinline|eq T a b|.
-Here we diverge from Coq tradition and reserve \lstinline|=| for setoid equality, as this is the equality we will be working with most of the time.
+Here we diverge from Coq tradition and reserve the \lstinline|=| symbol for \emph{setoid} equality (to be discussed momentarily), as this is the equality we will be working with most of the time.
 Importantly, since convertibility is a congruence, a proof of \lstinline|a ≡ b| lets us substitute \lstinline|b| for \lstinline|a| anywhere inside a term without further conditions.
-We mention this explicitly only because rewriting \emph{does} give rise to conditions when we depart from Leibniz equality and introduce equivalence relations representing abstractions over common representations of the same conceptual object.
-For instance, consider formal fractions of integers as a representation of rationals.
-Rewriting a subterm using such an equality is permitted only if the subterm is an argument of a function that has been proven to \emph{respect} the equality.
-Such a function is called \emph{proper}, and propriety must be proved for each function in whose arguments we wish to enable rewriting.
+We mention this explicitly only because such rewriting \emph{does} give rise to conditions when we depart from raw convertibility and introduce equivalence relations
+that express how possibly distinct (unconvertible) terms may represent the same conceptual object.
+A typical example would be different formal integer fractions representing the same rational number.
+Rewriting a subterm using a proof of such an equality is permitted only if the subterm is argument to a function that has been proven to \emph{respect} the equality.
+Such a function is called \emph{proper} with respect to the equality in question, and propriety must be proved for each function in whose arguments we wish to enable rewriting.
 
@@
 \begin{lstlisting}
-  Context `{Field A} `{∀ x y: A, Decision (x = y)} {inj_inv}.
+  Context `{Field A} `{∀$$x y: A, Decision (x = y)} {inj_inv}.
   Class Rationals: Prop :=
@@
 
-Universe polymorphism could in principle most likely solve this problem, but its current implementation in Coq only supports universe polymorphic inductive definitions, while \lstinline|Singleton| is a regular definition.
-Universe polymorphic \emph{definitions} have historically not been supported in Coq, primarily because of efficiency concerns.
+Universe polymorphism could in principle most likely solve this problem, but its current implementation in Coq only supports universe polymorphic \emph{inductive} definitions, while \lstinline|Singleton| is a regular definition.
+Universe polymorphic regular definitions have historically not been supported in Coq, primarily because of efficiency concerns.
 However, we have taken up the issue with the Coq development team, and they have agreed to introduce the means to let one turn on universe polymorphism for definitions voluntarily on a per-definition basis.

### 2010-12-09 papers/mscs/mscs.tex
More referee-guided MSCS paper improvements.

@@
 
-For the algebraic hierarchy---a critical component in any library of formalized mathematics---these challenges include structure inference, handling of multiple inheritance, idiomatic use of notations, and convenient algebraic manipulation (e.g.\ rewriting).
+For the algebraic hierarchy---a critical component in any library of formalized mathematics---these challenges include structure inference, handling of multiple inheritance, equality of axiomatically posited and derived structure, idiomatic use of notations, support for models based on quotient representations, and convenient algebraic manipulation (e.g.\ rewriting).
 Several solutions have been proposed for the Coq theorem prover: dependent records~\cite{DBLP:journals/jsc/GeuversPWZ02} (a.k.a.
 telescopes), packed classes~\cite{Packed}, and occasionally modules.
-We present a new approach based entirely on Coq's new type class mechanism, and show how it can be used to make fully `unbundled' predicate-representations of algebraic structures practical to work with.
-
-Since we intend to use this development as a basis for constructive analysis with practical certified exact real arithmetic, an additional objective and motivation in our design is to facilitate \emph{efficient} computation.
-In particular, we want to be able to effortlessly swap implementations of number representations.
+We present a new solution based entirely on the use of Coq's new type class facility to make fully `unbundled' predicate-representations of algebraic structures practical to work with.
+
+Our development is not merely aimed at formalization of theory, and our choice of a system based on type theory is no accident.
+It is our explicit ambition for the interfaces and theory we develop to be employed directly for the specification and parameterization of efficiently executable implementations of structures and procedures, expressed using type theory's native term reduction as a programming language.
+Thus, our work very much belongs in the long tradition of trying to realize to the greatest possible extent the promise of type theory to truly unite mathematics and certified (functional) programming, without making painful sacrifices on either side.
+
+Because our ``ultimate'' goal is to use this development as a basis for constructive analysis with practical certified exact real arithmetic, and because numerical structures are ideal test subjects for our algebraic hierarchy, we shall use these to motivate and demonstrate the key parts of our development.
+Since we are concerned with \emph{efficient} computation, we want to be able to effortlessly swap implementations of number representations.
 Doing this requires that we have clean abstract interfaces, and mathematics tells us what these should look like: we represent $\N$, $\Z$, and $\Q$ as \emph{interfaces} specifying an initial semiring, an initial ring, and a field of integral fractions, respectively.
-To express these interfaces elegantly and without duplication, our development\footnote{The sources are available
-at:~\url{http://www.eelis.net/research/math-classes/}} includes an integrated formalization of parts of category theory and multi-sorted universal algebra, all expressed using type classes for optimum effect.
+To express these interfaces elegantly and without duplication, our development\footnote{The sources are available at:~\url{http://www.eelis.net/research/math-classes/}} includes an integrated formalization of parts of category theory and multi-sorted universal algebra, all expressed using type classes for optimum effect.
 
@@
 
-The `native' notion of equality in Coq is that of terms being convertible, naturally reified as a proposition by the inductive type family \lstinline|eq| with single constructor \lstinline|eq_refl|:
+The `native' notion of equality in Coq is that of term convertibility, naturally reified as a proposition by the inductive type family \lstinline|eq: ∀ (T: Type), T → T → Prop| with single constructor \lstinline|eq_refl|:
 \begin{lstlisting}
@@
 where `\lstinline|a ≡ b|' is notation for \lstinline|eq T a b|.
-Here we diverge from Coq tradition and reserve the \lstinline|=| symbol for \emph{setoid} equality (to be discussed momentarily), as this is the equality we will be working with most of the time.
+Here we diverge from Coq tradition and reserve the `\lstinline|a = b|' notation for \emph{setoid} equality (to be discussed momentarily), as this is the equality we will be working with most of the time.
+
 Importantly, since convertibility is a congruence, a proof of \lstinline|a ≡ b| lets us substitute \lstinline|b| for \lstinline|a| anywhere inside a term without further conditions.
-We mention this explicitly only because such rewriting \emph{does} give rise to conditions when we depart from raw convertibility and introduce equivalence relations
-that express how possibly distinct (unconvertible) terms may represent the same conceptual object.
-A typical example would be different formal integer fractions representing the same rational number.
+We mention this explicitly only because such rewriting \emph{does} give rise to conditions when we depart from raw convertibility and introduce equivalence relations that express how possibly distinct (unconvertible) terms may represent the same conceptual object.
+Rational numbers represented by (non-reduced) formal integer fractions are a typical example.
 Rewriting a subterm using a proof of such an equality is permitted only if the subterm is argument to a function that has been proven to \emph{respect} the equality.
@@
 
-The algebraic hierarchy of the \textsc{Ssreflect} libraries~\cite{Packed} uses an interesting (if fairly restrictive) alternative approach.
-It requires unicity of representation for all objects so that Leibniz equality is always applicable.
+The algebraic hierarchy of the \textsc{Ssreflect} libraries~\cite{Packed} uses an alternative approach.
+It simply requires canonical representation of all objects, so that setoid equality is not needed.
+Of course, this policy severely restricts the freedom one has when implementing models of abstract structures.
+Indeed, for some sets, canonical representation schemes do not exist.
+The constructive reals, which are of particular interest to us, are an example of such a set.
 

### 2010-12-10 papers/mscs/mscs.tex
Change citation and bibliography style in MSCS paper.

@@
 \usepackage{amsmath}
+\usepackage{harvard}
 \usepackage{amssymb}
@@
 The Coq proof assistant is based on the calculus of inductive
-constructions~\cite{CoquandHuet,CoquandPaulin}, a dependent type theory with (co)inductive types; see~\cite{Coq,BC04}.
+constructions~\cite{CoquandHuet,CoquandPaulin}, a dependent type theory with (co)inductive types; see~\cite{BC04,Coq}.
 In true Curry-Howard fashion, it is both an excessively pure, if somewhat pedantic, functional programming language with an extremely expressive type system, and a language for mathematical statements and proofs.
@@
 
-Propositions in Coq are types~\cite{ITT,CMCP}, which themselves have types called \emph{sorts}.
+Propositions in Coq are types~\cite{CMCP,ITT}, which themselves have types called \emph{sorts}.
 Coq features a distinguished sort called \lstinline|Prop| that one may choose to use as the sort for types representing propositions.
@@
 We are currently in the process of retrofitting the rationals interface into CoRN.
-In future work we aim to base our development of its reals on an abstract dense set, allowing us to use the efficient dyadic rationals~\cite{boldo2009combining} as a base for exact real number computation in Coq~\cite{Riemann,Oconnor:real}.
+In future work we aim to base our development of its reals on an abstract dense set, allowing us to use the efficient dyadic rationals~\cite{boldo2009combining} as a base for exact real number computation in Coq~\cite{Oconnor:real,Riemann}.
 The use of category theory has been important in these developments.
@@
 % MSCS recommends harvard, but does not specify which bst file to use.
-\bibliographystyle{plain}
+\bibliographystyle{agsm}
 \bibliography{alg}

### 2010-12-10 papers/mscs/mscs.tex

@@
 Because the Coq type theory lacks quotient types (as it would make type checking undecidable), one usually bases abstract structures on a \emph{setoid} (`Bishop set'): a type equipped with an equivalence relation~\cite{Bishop67,Hofmann,Capretta}.
-As we will see in section~\ref{univ}, this way of working pays off when working with notions such as quotient algebras.
+Palmgren~\cite{palmgren2009constructivist} shows that Bishop sets have pleasant categorical properties, which translate to a powerful implicit type structure.
+It would be of interest to actually provide machine support for this type structure.
+As we will see in section~\ref{univ}, working with setoids pays off when working with notions such as quotient algebras.
 
@@
 
-Type classes have been a great success story in the Haskell functional programming language, as a means of organizing interfaces of abstract structures.
+Type classes \cite{wadler1989adhoc} have been a great success story in the Haskell functional programming language, as a means of organizing interfaces of abstract structures.
 Coq's type classes provide a superset of their functionality, but implemented in a different way.
@@
 Thus, type classes in Coq are realized by relatively minor syntactic aids that bring together existing facilities of the theory and the system into a coherent idiom, rather than by introduction of a new category of qualitatively different definitions with their own dedicated semantics.
+
+The basic idea of using type-class-like facilities for structuring computerized mathematics dates back to the \textsc{AXIOM} computer algebra system~\cite{jenks1992axiom}.
+Weber~\cite{weber1993type} pursues the analogy between \textsc{AXIOM}'s so-called categories and type classes in Haskell.
+Santas~\cite{santas1995type} pursues analogies between type classes, \textsc{AXIOM} categories and existential types.
+Existential types are present in Haskell, but absent from Coq.
 

### 2010-12-13 papers/mscs/mscs.tex
Adding references.

@@
 \usepackage{amsmath}
-\usepackage{harvard}
+% \usepackage{harvard}
 \usepackage{amssymb}
@@
 Our development is not merely aimed at formalization of theory, and our choice of a system based on type theory is no accident.
-It is our explicit ambition for the interfaces and theory we develop to be employed directly for the specification and parameterization of efficiently executable implementations of structures and procedures, expressed using type theory's native term reduction as a programming language.
-Thus, our work very much belongs in the long tradition of trying to realize to the greatest possible extent the promise of type theory to truly unite mathematics and certified (functional) programming, without making painful sacrifices on either side.
+It is our explicit ambition for the interfaces and theory we develop to be employed directly for the specification and parameterization of efficiently executable procedures and data structures, implemented using type theory's native term reduction as a programming language.
+Thus, our work very much belongs in the long tradition of trying to realize the promise of type theory to truly unite mathematical formalization and certified (functional) programming, without making painful sacrifices on either side.
 
@@
 
- In section~\ref{classes}, we discuss our algebraic hierarchy implemented using type classes.
+ In section~\ref{classes}, we discuss our algebraic hierarchy implemented with type classes.
 In sections~\ref{cat} and~\ref{univ} we give a taste of what category theory and universal algebra look like in our development, and in section~\ref{numbers} we use these facilities to build abstract interfaces for numbers.
@@
 Historically, using bundled representations has nevertheless been an acceptable trade-off, because (1) the unbundled alternative was such a pain, and (2) the standard algebraic hierarchy (up to, say, fields and modules) is not all that wild.
+Further instances of the cost of sharing can be found in Section~\ref{classes}.
 
@@
 With these in place, our type class for categories follows the usual type-theoretical definition of a
-category~\cite{saibi1995constructive}:
+category~\cite{saibi1995constructive,PhdSAIBI}:
 \begin{lstlisting}
@@
 % MSCS recommends harvard, but does not specify which bst file to use.
-\bibliographystyle{agsm}
+\bibliographystyle{plain}
 \bibliography{alg}

### 2010-12-14 papers/mscs/mscs.tex
Undo accidental disabling of Harvard style bibliography and citation.
Revert "Adding references."
This reverts commit 11741048f231681f0d212b34e54060cc40271f2f.

@@
 \usepackage{amsmath}
-% \usepackage{harvard}
+\usepackage{harvard}
 \usepackage{amssymb}
@@
 Historically, using bundled representations has nevertheless been an acceptable trade-off, because (1) the unbundled alternative was such a pain, and (2) the standard algebraic hierarchy (up to, say, fields and modules) is not all that wild.
-Further instances of the cost of sharing can be found in Section~\ref{classes}.
 
@@
 With these in place, our type class for categories follows the usual type-theoretical definition of a
-category~\cite{saibi1995constructive,PhdSAIBI}:
+category~\cite{saibi1995constructive}:
 \begin{lstlisting}

### 2010-12-14 papers/mscs/mscs.tex
Fitting into 30pp

@@
 \newcommand{\Q}{\ensuremath{\mathbbm{Q}}}
-
+\usepackage{wrapfig}
 \usepackage{color}
@@
 It is our explicit ambition for the interfaces and theory we develop to be employed directly for the specification and parameterization of efficiently executable procedures and data structures, implemented using type theory's native term reduction as a programming language.
-Thus, our work very much belongs in the long tradition of trying to realize the promise of type theory to truly unite mathematical formalization and certified (functional) programming, without making painful sacrifices on either side.
+Thus, our work belongs in the long tradition of realize the promise of type theory to truly unite mathematical formalization and certified (functional) programming, without making painful sacrifices on either side.
 
@@
 
-\paragraph{Outline.}
+\emph{Outline.}
 In section~\ref{preliminaries} we briefly describe the Coq system and its implementation of type classes.
@@
 For reflexive relations, the following is a fully bundled representation---the other end of the spectrum:
-\newpage
+
 \begin{lstlisting}
@@
 
-In the next section, we show that type-classification of structure predicates and their component parameters has the potential to remedy the problems associated with the naive unbundled predicate approach, giving us the best of both worlds.
+In the next section, we show that type-classification of structure predicates and their component parameters has the potential to remedy the problems associated with the naive unbundled predicate approach.
 
@@
 
-Next, we turn to problems concerning theorem \emph{declaration}. As a reference point, our ideal will be the common mathematical vernacular, where one simply says:
+Next, we turn to problems concerning theorem \emph{declaration}. Our ideal will be the common mathematical vernacular, where one simply says:
 \begin{quote}
@@
 While the syntax already supports variants (not shown above) that differ in how exactly different kinds of arguments are inferred and/or generalized, there is no support to have an argument be ``inferred if possible, generalized otherwise".
-We have found that the need for such a policy arises naturally when declaring a parameter of class type in a context where \emph{some} of its components are already available, while others are to be newly introduced.
+The need for such a policy arises naturally when declaring a parameter of class type in a context where \emph{some} of its components are already available, while others are to be newly introduced.
 The current workaround in these cases involves providing names for components that are then never referred to, which is a bit awkward.
@@
 One aspect of the predicate approach we have not mentioned thus far is that in proofs parameterized by abstract structures, all components become hypotheses in the context.
-For the example theorem above, the context looks like:
+For the theorem above, the context looks like:
 \begin{lstlisting}
@@
 \end{lstlisting}
+
 All of \lstinline|Equiv|, \lstinline|RingPlus|, \lstinline|RingMult|, \lstinline|RingZero|, and \lstinline|RingOne| are operational (single-field) classes, with bound notations \lstinline|=|, \lstinline|+|, \lstinline|*|, \lstinline|0|, and \lstinline|1|, respectively.
-
+\begin{wrapfigure}{r}{.4\textwidth}
+    \includegraphics[scale=.8]{hierarchy.pdf}
+%   \caption{}
+\end{wrapfigure}
 Let us briefly highlight some additional aspects of this style of structure definition in more detail.
@@
 Fields declared with \lstinline|:>| are registered as hints for instance resolution, so that in any context where \lstinline|(A, =, +, 0, *, 1)| is known to be a \lstinline|SemiRing|, \lstinline|(A, =, +, 0)| and \lstinline|(A, =, *, 1)| are automatically known to be \lstinline|CommutativeMonoid|s (and so on, transitively, because instance resolution is recursive).
-In our hierarchy, these substructures by themselves establish the following inheritance diagram:
-
-\vspace{5mm}
-\begin{center}
-\includegraphics{hierarchy.pdf}
-\end{center}
-\vspace{5mm}
+In our hierarchy, these substructures by themselves establish the inheritance diagram.
 
@@
 
-\paragraph{Acknowledgements.}
+\emph{Acknowledgements.}
 This research was stimulated by the new possibilities provided by the introduction of type classes
@@
 Georges Gonthier and Claudio Sacerdoti Coen have helped to
-sharpen our understanding of the relation with Canonical Structures and with Unification Hints. Finally, James McKinna suggested many corrections and improvements.
+sharpen our understanding of the relation with Canonical Structures and with Unification Hints. Finally, James McKinna suggested many improvements.
 

### 2010-12-15 papers/mscs/mscs.tex

@@
 \definecolor{dkviolet}{rgb}{0.6,0,0.8}
-
-% preliminary choices:
-%   don't discuss difference between maximally-inserted and non-maximally-inserted implicit arguments.
-
-% todo:
-  % cite that other type class based development in the context of the alternative bundling schemes.
 
@@
 The introduction of first-class type classes in the Coq system calls for re-examination of the basic interfaces used for mathematical formalization in type theory.
-% especially since existing solutions fall short from a proof engineering perspective.
 We present a new set of type classes for mathematics and take full advantage of their unique features to make practical a particularly flexible approach formerly thought infeasible.
@@
 
-% ideally should be more or less readable by Haskell person. that would be neat. so that's what the preliminaries should prepare for.
 In this paper we focus on the Coq proof assistant. We conjecture that the methods can be transferred
@@
 
- %Having derived some principles concerning bundling, we then proceed to show (in section \ref{managing}) how type classes may be employed to adopt these principles without compromising ease of use.
-Next, in section \ref{hierarchy}, we introduce a number of additional conventions that let us complete a consistent idiom for interfaces to algebraic structures.
-With this in place, we continue with a discussion of
-
 \section{Preliminaries}
 \label{preliminaries}
-
-% ideally, this should make the paper readable for Haskell people
 
@@
 The distinguishing feature of the \lstinline|Prop| sort is that terms of non-\lstinline|Prop| type may not depend on the values of inhabitants of \lstinline|Prop| types (that is, proof terms).
-% As originally conceived~\cite{ITT,CMCP}, Propositions are Types and we can extract information from
-% them.
-Often we do \emph{not} want functions to depend on proofs.
-In Coq, this is the paradigmatic way of working.
-We are forbidden to extract information from proofs of \lstinline|Prop|s.
-Especially with the aid of the Program machinery, this allows us to flexibly work with simply typed program and their Hoare style correctness proofs.
 This regime of discrimination establishes a weak form of proof irrelevance, in that changing a proof can never affect the result of value computations.
 On a very practical level, this lets Coq safely erase all \lstinline|Prop| components when extracting certified programs to OCaml or Haskell.
-
-% A stronger notion of proof irrelevance in which any two inhabitants of a \lstinline|Prop| type are made \emph{convertible} is planned for a future Coq version.
-  % commented out because not really relevant. only re-add with explanation of relevance.
-
-% todo: process?:
-%One may be tempted to propose Propositions as [ ]-types~\cite{awodey2004propositions}. Here the
-%construction $[P]:=\{\top\mid p:P\}$ squashes all the information about how we proved $P$. In Coq,
-%the [ ]-types are isomorphic to the type of sort \lstinline|Prop|. However,
-%there are places where we needed proof \emph{relevance}; e.g.\ in the proof of the first isomorphism
-%theorem above.In practice, manual dead-code analysis seems useful. Making Harrop formulas proof
-%irrelevant seems to be a good first approximation~\cite{lcf:spi:03}. However, a further refinement
-%of Propositions as types seems necessary.
-% Making Harrop formulas proof irrelevant seems to be a good first approximation~\cite{lcf:spi:03}.
-An orthogonal and further reaching proposal is to replace the Coq type theory with the Calculus of implicit constructions with implicit Σ-types~\cite{miquel2001implicit,barras2008implicit,Bernardo}.
 
@@
 Coq provides a modest level of \emph{universe-polymorphism} so that we may avoid duplication when trying to support \lstinline|Prop|-sorted and non-\lstinline|Prop|-sorted content with a single set of definitions.
-
-% However, we would also like to use Coq as a dependently typed programming language.
-This requires a hybrid approach, were we put informative propositions in \lstinline|Type| and non-informative ones in \lstinline|Prop|.
-% Eelis: I don't see the logic here.
 
@@
 
-% \section{The Type-Classified Algebraic Hierarchy}\label{classes}
-% We represent each structure in the algebraic hierarchy as a type class.
-This immediately leads to the familiar question of which components of the structure should become parameters of the class, and which should become fields.
-By far the most important design choice in our development is the decision to turn all \emph{structural} components (i.e.
-carriers, relations, and operations) into parameters, keeping only \emph{properties} as fields.
-Type classes defined in this way are essentially predicates with automatically resolved proofs.
-% 
-% Conventional wisdom warns that while this approach is theoretically very flexible, one risks extreme inconvenience both in having to declare and pass around all these structural components all the time, as well as in losing notations (because we no longer project named operations out of records).
-These legitimate concerns will be addressed in Sections~\ref{bundle}-\ref{classes}.
-Section~\ref{univ} abstracts to universal
-% algebra, we develop the basic theory to the point of the first homomorphism theorem.
-% Section~\ref{interfaces} provides interfaces to usual data structures by providing their universal
-% properties. To allow symbolic manipulation of semantic objects one needs a quote function. Usually,
-% this function is implemented in Coq's tactics language. A more convenient way is to use type class
-% unification; see section~\ref{quote}. In section~\ref{canonical}, we end with a short comparison
-% between type classes and canonical structures.
-
 \section{Bundling is bad}\label{bundling}
@@
 
-% hm, i used to have paragraphs here about how it would be unclear how this approach could support structure inference and structure inheritance, but those are actually pretty straightforward using ordinary proof search.
-
 In the absence of type classes, these are all very real problems, and for this reason the two largest formalizations of abstract algebraic structures in Coq today, CoRN~\cite{C-corn} and \textsc{Ssreflect}~\cite{Packed}, both use \emph{bundled} representation schemes, using records with one or more of the components as fields instead of parameters.
@@
 While there is no conventional notation for reflexive relations, the situation is the same in the context of, say, a semiring, where we would bind $+$ and $*$ notations to the record field projections for the addition and multiplication operations, respectively.
-
-% While not observable in this example, when combined with other specialized facilities in Coq (namely coercions and/or canonical structures), this bundled representation also lets one address the problems of structure inference inheritance to some extent.
 
@@
 
-%For completeness, we mention an alternative record-based approach known as ``packed classes'', as used in the ssreflect library.
-In the packed classes idiom, each algebraic structure is represented by not one but a trinity of interdependent records along with packing/unpacking functions.
-These are then stacked in an intricate manner designed to reach a particular hybrid packing balance that, combined with the use of canonical structures and coercions, accomplishes some of the goals above.
-
-%We have found it hard to assess the general merits of this approach, partly because of its complexity, and partly because the only actual usage example (the hierarchy in the ssreflect library) is not only wedded to a very specialized view on equality and decidability that does not support setoid relations from the outset, but also diverges significantly from standard mathematical dichotomy, making it hard to recognize familiar structures and presentation.
-The reasons for this (as we understand them) are pragmatic and twofold.
-
-%First, the ssreflect library was originally developed strictly to support very practical work in a specialized domain of mathematics involving decidable and mostly finite structures.
-Second, at the time of ssreflect's initial development, support for setoid rewriting in Coq was not nearly as stable and refined as it is today, which made ``Leibniz equality at all costs'' a more reasonable strategy at the time than it might be considered today.
-
-%Work to extend the ssreflect framework with techniques to allow representation of a less restricted class of mathematical structures has only recently begun and is still ongoing, though there are currently no plans to support undecidable structures.
-
-%Because in our development we will be working with computable reals, equality on which can never be either Leibniz or decidable, we conclude that the ssreflect library is not usable for us.
-
 The observant reader may wonder whether it might be beneficial to go one step further and unbundle proofs of laws and inherited substructures as well.
@@
 Indeed, parameterizing on proofs would be actively harmful because instantiations differing only in the proof argument would express the same thing yet be nonconvertible, requiring awkward conversions and hindering automation.
-
-% Another benefit of our approach is that it works very smoothly when these records are in fact type
-% classes. When relations and operations are bundled and do \emph{not} appear as parameters of the
-% structure record/class, to express that some collection of relations and
-% operations possess the structure as a type class constraint seems to require manifest fields; see
-% Section~\ref{manifest}. With
-% our unbundled approach, on the other hand, such type class constraints are easily expressed without
-% any extensions as type inhabitation queries
-% % (e.g. ``hey Coq, find me a proof of [Monoid some_rel some_op some_unit]'')
-% of types living in Prop, so that we don't even care what instances are found (thus embracing proof
-% irrelevance).
-
-% Leaving the equivalence proofs bundled means that when one has a proof that a structure
-% is a semiring, one has two proofs (via the two monoids) that the equivalence relation is an
-% equivalence. Fortunately, this redundancy is entirely harmless since our terms
-% never refer (directly or indirectly) to proofs, which is the case when the relations and operations
-% they are composed from are left unbundled. Here, too, we enjoy the warm embrace of proof
-% irrelevance.
-% 
-% From the discussion so far, we derive some principles:
-% \begin{enumerate}
-%  \item structural concepts (like Reflexive, Setoid, Monoid) should have their structural components
-% (i.e. carriers, relations, and operations) as parameters, and nothing else;
-%  \item (as a consequence) these concepts should be defined as predicates (which includes records and
-% classes living in Prop), and their proofs may be kept opaque;
-%  \item (as a second consequence) algebraic expressions need/should never refer (directly or
-% indirectly) to proofs about the algebraic structure.\footnote{The reciprocal in a field
-% famously does refer to a proof that the element is non-zero, but not to a proof that the operations
-% indeed form a field.}
-% \end{enumerate}
-% % \setcounter{enumi}{3}
-% 
-% The principles listed so far are not particularly controversial, in fact this
-% approach ensures maximum flexibility~\cite{Hints}. However,
-% there is a common perception that unbundling carriers, relations, and operations this way invariably
-% leads either to unmanageably big algebraic expressions (involving endless projections and/or implicit
-% arguments), or to unmanageably long and tedious enumeration of carriers/relations/operations, or to
-% problems establishing canonical names for operations (since we no longer project them out of
-% structure instances all the time).
-% 
-% In the next sections, we show how systematic use of type classes (and their support infrastructure
-% in Coq), combined with the use of a special form of type class we call an `operational type class'
-% (as opposed to a `structural' type class we might use in place of the records shown thus far), lets
-% us avoid these problems.
 
@@
 The current workaround in these cases involves providing names for components that are then never referred to, which is a bit awkward.
-
-% We note that operational type classes allow us to avoid Coq's notation \lstinline|scope| mechanism.
-% hm, i don't think this is true.
-we don't need scopes for different representations' use of operators because we properly recognize their commonality, but that can be done without operational type classes as well.
-
-% TODO: get this in:
-% Although the expression now \emph{looks} like it is bound to some
-% specific semigroup structure (which with traditional approaches would imply projections from bundles
-% and/or reference to proofs), it is really only referring near-directly to the actual operations
-% involved, with the semigroup proof existing only as opaque `knowledge about the operations' which we
-% may use in the proof. This lack of projections keeps our terms small and independent, and keeps
-% rewriting simple and sane; see Section~\ref{canonical}.
 
@@
     \includegraphics[scale=.8]{hierarchy.pdf}
-%   \caption{}
 \end{wrapfigure}
@@
 
-% note: we don't mention "super classes" because by Matthieu's terminology only parameters-of-class-type are "super classes", and for us that means only the operational classes, which we certainly don't think of as being part of the hierarchy.
-Matthieu calls fields of class type declared with ":>" substructures.
-
 However, we can easily add additional inheritance relations by declaring corresponding class instances.
@@
 Again, contrast this with bundled approaches, where axiomatic inheritance relations determine projection paths, and where additional inheritance relations require rebundling and lead to additional and ambiguous projection paths for the same operations.
-
-% todo: problem with ambiguous projection paths: quoting will not notice convertibility.
 
@@
 
-% rather strict separation of operations and proofs of properties about them (resolved by instance resolution) provides a very pleasing environment to work in, without any need for hacks.
-% terms composed of algebraic operations and relations 
-% separation of , this kind of redundance is never a problem for us. 
-% it is /never/ a problem to know a fact twice.
-% multiple proofs of equality, and even of semigroup-ness (since both Monoid), but that's simply not a problem, since none of our terms ever refers to such proofs.
-We have found that this rather strict separation of operations and proofs of properties about them (resolved by instance resolution) provides a very pleasing environment to work in, without any need for hacks.
-
 We use names for properties like distributivity and absorption, because these are type classes as well (which is why we declare their instances with \lstinline|:>|).
@@
 We will return to this point in section~\ref{conclusions}.
-
-% : with predicate classes we can essentially pose queries of the form ``do I already know that proposition P holds?'', whereas canonical structures
-% ``do I already have a composite structure X of which Y is a specific component?
-% 
-% arbitrary predicates, based on associating particular
-
-% Having argued that the \emph{all-structure-as-parameters} approach \emph{can} be made practical, we enumerate some of the benefits that make it worthwhile.
-% 
-% First, multiple inheritance becomes trivial: \lstinline|SemiRing| inherits two \lstinline|Monoid| structures on the same carrier and setoid relation, using ordinary named arguments to achieve ``manifest fields''; see section~\ref{manifest}.
-
-% Third, since our structural type classes are mere predicates, overlap between their instances is a non-issue.
-Together with the previous point, this gives us tremendous freedom to posit multiple structures on the same operations and relations, including ones derived implicitly via subclasses: by simply declaring a \lstinline|SemiRing| class instance showing that a ring is a semiring, results about semirings immediately apply implicitly to any known ring, without us having to explicitly encode this relation in the hierarchy definition itself, and without needing any projection or translation of carriers or operations.
-% 
-% Fourth, \emph{Structure-as-parameters} helps setoid-rewriting: type class resolution
-% can find the equivalence relation in the context.\footnote{
-% A similar style should be possible for, say, the \lstinline|Ring| tactic, instead of
-% declaring the ring structure by a separate command, we would rely on type class resolution to find it.}
-% We note that \lstinline|op| does not depend on the proof that \lstinline|e| is an equivalence.
-As explained above we use Coq's implicit quantification (\lstinline|`{}|) to avoid having to write all the parameters when \emph{stating} a theorem and Coq's maximally inserted implicit arguments to find the parameters when \emph{applying} a theorem.
-Both features are new in Coq and stem from the type class implementation.
-% 
-% We mention the trade-off between bigger contexts versus bigger terms. Our contexts are bigger than
-% those of telescopes or packed classes. In our experience, this has been relatively
-% harmless\footnote{However, Coq's data structure for contexts is not very efficient. Gonthier fears that this
-% may be a bottleneck for huge developments. It seems that the data structure chosen
-% in~\cite{asperti2009compact} will behave better.}%
-% : most terms in the context are there to support canonical names. Bigger terms
-% \emph{do} cause problems: 1. when proofs are part of mathematical objects we need to share these
-% proofs to allow rewriting. Moreover, it prohibits Opaque proofs and `proof irrelevance'. 2. The
-% projection paths may not be canonical.
-% 
-% Coercion pullbacks~\cite{Hints} were introduced to address problems with multiple coercions paths,
-% as in the definition of a semiring: a type with two monoid structures on it. We avoid some
-% of these problems by explicitly specifying the fields. We emphasize that the semiring properties are
-% automatically derived from the ring properties, although the properties of a semiring are not
-% structurally included in the ring properties.
-% 
-% Let us now stop and think to what extent this approach suffers from all the problems commonly
-% associated with it. In particular, let us imagine what happens to our terms and contexts when we
-% want to talk about nested structures such as polynomials over the ring. For a concrete
-% representation of the polynomials the context will just contain the context for an abstract ring.
-% For abstract reasoning about polynomials the context will grow with abstract operations on
-% polynomials. Consequently, the context will grow linearly, as opposed to exponentially.
 
@@
 This definition is based on the 2-categorical idea of having equality only on arrows, not on objects.
-%Similarly, we will have equality on natural transformations, but not on functors. % todo: ORLY?
 
@@
 \end{lstlisting}
-  % Todo: Having to add "$$" to get whitespace is awful. we can't add "\ " in η's literate-replacement though, because then we also get space before the colon in "η: t".
-
-% We have defined the category of setoids, the dual of a category, products, and hence co-products.
-
-% The usual size problems in the definition of the category of categories can be avoided by using universe polymorphism. 
-% However, we need to avoid making \lstinline|Relation| a \lstinline|Definition|, since \lstinline|Definitions| are not (yet?) universe polymorphic.
 
@@
 The \lstinline|Program| command generates proof obligations (not shown) expressing that these two arrows are indeed homomorphisms. The proof of the theorem itself is trivial.
-
-% \footnote{In Bishop's words~\cite[p.12]{Bishop/Bridges:1985}: The axiom of choice is used to extract
-%elements from equivalence classes where they should never have been put in the first place.}
 
@@
 The choice of \lstinline|Z nat| here is immaterial; we could have picked another, or even a generic, implementation of \lstinline|Integers|, but doing so would provide no benefit.
-  % todo: hm, this Z is poorly named. it's really natpair_integers.Z
-  % todo: say a few words about how we handle division by zero?
-
-% todo: say a word about section variable/context discharging, because we use it everywhere.
 
@@
 
-% Eelis: and when we do, we can write something like:
-% Given a ring $R$, the $R$-algebra $R[X]$ of polynomials, is the free $R$-algebra on a set $X$.
-% We provide two implementations of polynomials: the
-% standard representation using lists of coefficients and the Bernstein representation. A Bernstein
-% basis polynomial is one of the form:
-% \[b_{\nu,n}(x) = {n \choose \nu} x^{\nu} \left( 1 - x \right)^{n - \nu}, \quad \nu = 0, \ldots, n.\]
-% A Bernstein polynomial is a linear combination of these basic polynomials. Bernstein polynomials
-% have been used for efficient computations inside the Coq system~\cite{ZumkellerPhD}.
-
-% Eelis: TODO: process: At the time of writing, our development includes a fully integrated formalization of a nontrivial portion of category theory and multisorted universal algebra, including various categories (e.g.\ the category \lstinline|Cat| of categories, and finitary algebraic categories  defined by a theory which we instantiate to obtain the categories of monoids, semirings, and rings), products
-
-%\subsection{Categorical universal algebra}
-
-% We define the forgetful functor $U$ from the category of τ-algebras to the category of sets and its left-adjoint $F$: the term algebra consisting of closed terms, i.e.\ terms with the empty type as set of variables.
-This term algebra is the initial τ-algebra.
-Every adjunction defines a monad by composition of the functors $U$ and $F$.
-We will use the resulting expression monad in Section~\ref{quote} below.
-% Eelis: oh?
-
-%The product of two $\tau$-algebras is their categorical product.
-% 
-% \begin{comment}
-% Given a sub\emph{set} of an algebra, we define the sub\emph{algebra} generated by it.
-% An interesting fact to mention is the use of the following heterogeneous equality between
-% elements of the algebra and of the subalgebra, i.e.\ terms of different types may be equal.
-% \begin{lstlisting}
-%   Fixpoint heq {o}: op_type carrier o -> op_type v o -> Prop :=
-%     match o with
-%     | constant x => fun a b => `a == b
-%     | function x y => fun a b => forall u, heq (a u) (b (`u))
-%     end.
-% \end{lstlisting}
-% \end{comment}
-% 
-% We connect the abstract theory to the concrete theory. Concretely, let $\tau_m$ be the theory
-% of monoids. Given a (concrete) monoid, we can construct the corresponding $\tau_m$-algebra. 
-% Conversely, given a $\tau_m$-algebra, we construct an instance of the \lstinline|Monoid| type class.
-% %varieties/monoid.v
-% \begin{lstlisting}
-% Variable o: variety.Object theory.
-% Global Instance: SemiGroupOp (o tt) := algebra_op theory mult.
-% Global Instance: MonoidUnit (o tt) := algebra_op theory one.
-% Global Instance from_object: Monoid (o tt).
-% \end{lstlisting}
-% 
-% In the last line, instance resolution `automatically' finds the operation and the unit specified in
-% the lines before.
-% 
-% This interplay between concrete algebraic structure and their expressions on the one hand, and models of equational theories on the other is occasionally a source of tension (in that translation in either direction is not yet fully automatic).
-However, it opens the door to the possibility of fully internalized implementations of generic tactics for algebraic manipulation, no longer requiring plugins.
-We come back to this when we describe automatic quotation of concrete expressions into universal algebra expressions; see Section~\ref{quote}.
-
 \section{Quoting with type classes}\label{quoting}
@@
 This should let us implement the basic setup of the technique once and for all, so that quotation for new algebraic structures can be enabled with minimal effort.
-  %  Preliminary results in this area (such as a simple monoid normalization tactic) suggest yada.
 
@@
 To describe the relation we want the symbolic expression to have to the original expression, we first define how symbolic expressions evaluate to values (given a variable assignment):
-%An expression is only meaningful in the context of a variable assignment:\footnote{Categorically: \lstinline|nat| is an \lstinline|Expr|-algebra and \lstinline|eval| is \lstinline|map vars| composed with the algebra map.}
 \begin{lstlisting}
@@
 
-%Monads are trees with grafting~\cite{MonadsGrafting}: the tree monad
-%$TX:=1+T^2X$, may be seen as the prototypical monad. 
-% The \lstinline|Expr| monad, i.e.\ the free \lstinline|Expr|-algebra monad on the category of \lstinline|Type|s with extensional functions%\marginpar{Proof in classquote}
-% , provides a bind operation which describes the arrows in the Kleisli category. Concretely,
-% \begin{lstlisting}
-% Fixpoint bind {V W} (f: V → Expr W) (e: Expr V): Expr W :=
-%   match e with
-%   | Zero => Zero
-%   | Mult a b => Mult (bind f a) (bind f b)
-%   | Var v => vs v
-%   end.
-% \end{lstlisting}
-% Eelis: above commented out because we never actually use it, and it's really not pertinent to the discussion.
-
-% We have shown that |Expr|, i.e.\ |bind| together with |Var|, is a monad on the category of Types
-% with extensional functions between them~\marginpar{Really?}.
-
 We can now state our goal: given an expression of type \lstinline|nat|, we seek to construct an \lstinline|Expr V| for some appropriate \lstinline|V| along with a variable assignment, such that evaluation of the latter yields the former.
@@
 Instead, we have used instance resolution as a unification-based programming language to steer the unifier into inferring the symbolic quotation.
-
-%A similar method is used in Coq's notation mechanism.
-% We recommend a change in
-% implementation supporting an agda style precedence relation which is only restricted to be a
-% directed acyclic graph~\cite{danielsson2009parsing}.
-
-% We have used this technique to implement a tactic to normalize monoid expressions, and plan to extend yada yada.
-%It would be interesting to use this technique to implement the \lstinline|congruence| tactic, which implements
-%the congruence closure algorithm~\cite{corbineau2007deciding}. One would naturally obtain a tactic
-%which moreover works on setoid equalities.
 
@@
 \end{lstlisting}
-%    { ... structural and propriety proofs sufficient to bundle up the categories and arrows ...
-%    ; ... AltAdjunction stated using the bundled ... }.
 On top of this interface we can build theory about typical sequence operations such as maps, folds, their relation to \lstinline|singleton| and \lstinline|extend|, et cetera.
@@
 
-% todo: not sure why we didn't have this problem with, say, nat/Naturals.
-
-\begin{comment}
-Canonical structures have been used to provide a uniform treatment~\cite{bertot2008canonical} of big
-operators, like $\Pi,\Sigma, \max$. These operators extend a pair of a binary and a 0-ary operation
-to an $n$-ary operation for any $n$. Categorically, one considers the algebra maps from non-empty
-lists, lists, inhabited finite sets and finite sets to the carrier of a semigroup, monoid,
-commutative semigroup, commutative monoid. Hence we want to reuse the libraries for lists etc.\ as much as possible.
-
-Again, we can use type classes, instead of canonical structures, to deduce the relevant monoid operation:
-\begin{lstlisting}
-  Definition seq_sum
-    `{Sequence A T} `{RingPlus A} `{z: RingZero A}: T $\to$ A
-    := @seq_to_monoid A T _ A ring_plus z id.
-  Eval compute in seq_sum [3; 2].
-\end{lstlisting}
-\end{comment}
-
-\begin{comment}
-\section{Subset types and proof irrelevance}\label{PI}
-We have mentioned proof irrelevance a number of times before. It is an important theme in our
-approach. The \lstinline|Program| machinery allows one to conveniently write programs with Hoare-style
-pre- and post-conditions. I.e.\ functions $f: \{ A \mid P \} \to \{ B \mid Q \}$. Both side
-conditions
-are intended to be proof irrelevant, they are in \lstinline|Prop|. Presently, Coq does not support such subset
-types, thus forcing the user to manually prove that \lstinline|f| does not depend on the proof of \lstinline|P|.
-The addition of proof irrelevance and subset types to Coq is pursued by
-Barras~\cite{Barras:subset,Werner08}. 
-% Squash is a monad and allows many constructions: subsets, unions, images, heterogeneous equality,
-% \ldots.
-
-Our approach may be intuitively explained by `telescopic' subset types. We recall the use of telescopes in record types~\cite{pollack2000dependently}. Given a telescope\[
-T=[x_1:A_1][x_2:A_2(x_1)]\cdots[x_k:A_k(x_1,\ldots,x_k)].
-\]
-There is an inductively defined type $\Sigma T$, with a single constructor, given by the following 
-formation and introduction rule in pseudo-code for $k=2$:
-\begin{align*}
- A_1 :& Type\\
- A_2 :& \Pi A_1,Type\\
-\cline{1-2}
-\Sigma T :& Type
-\end{align*}
-
-\begin{align*}
- \Sigma T:Type\quad a_1 &:A_1 \quad a_2 : A_2(a_1)\\
-\cline{1-3}
-(a_1,a_2) &: \Sigma T
-\end{align*}
-
-For our methodology to work we have to be able to transform the occurring record into a telescopic
-subset $\{ A \mid  B\}$. We moreover conjecture that the terms in \lstinline|B| only depend on the
-variables in \lstinline|A|, but not in \lstinline|B| itself. However, we do not need this.
-
-We are naturally lead to the following methodology: Let $\phi$ be a statement in type
-theory, i.e.\ an alternation of $\Pi$s and $\Sigma$s.
-By the type theoretic axiom of choice, we first Skolemize to $\Sigma \overrightarrow{h}.
-\Pi \overrightarrow{a}.
-P$ and then transfer the variable $\overrightarrow{h}$ to the parameters.
-Concretely, the statement of surjectivity is skolemized to a surjection with an explicit right inverse, i.e.\ a split epi.
-We move the right inverse to the parameters.
-As a side effect, the distinction between classical and constructive mathematics virtually disappears for such explicit statements.
-
-The addition of implicit Σ-types to Coq would change many things. Perhaps it would even allow us
-to pack the proofs together with the operations, but it is too early to tell.
-\end{comment}
-
 \section{Conclusions}\label{conclusions}
@@
 With type classes we avoid these problems by avoiding bundling altogether.
-
-%Telescopes have been criticized~\cite{Packed} for the lack of multiple inheritance and
-%the efficiency penalty of a long chain of coercion projections. Packed classes~\cite{Packed} provide
-%a solution to these problems. We have provided an alternative solution based on type classes.
 
@@
 
-% \section{Canonical structures}\label{canonical}
-% Packed classes use canonical structures for the algebraic hierarchy. Both canonical structures and
-% type classes may be seen as instances of hints in unification~\cite{Hints}. Some uses of canonical
-% structures can be replaced by type class instances. The user manual (2.7.15) uses canonical
-% structures to derive the setoid equality on the natural
-% numbers in the following example \lstinline|(S x)==(S y)|. In this case type classes provide similar proof
-% terms. Canonical structures give
-% \begin{lstlisting}
-% @equiv(Build_Setoid nat (@eq nat)(@eq_equivalence nat))(S x)(S y)
-% \end{lstlisting}
-% which includes an explicit proof that \lstinline|(@eq nat)| is an equivalence,
-% whereas we obtain \lstinline|@equiv nat (@eq nat) (S x) (S y)|.
-
 Unification hints are a more general mechanism than type classes, and could provide a more precise account of the interaction between implicit argument inference and proof search.
 It is not a great stretch to conjecture that a fruitful approach might be to use unification hints as the underlying mechanism, with type classes as an end-user interface encapsulating a particularly convenient idiom for using them.
-% We encourage their inclusion to the Coq system. -- Eelis: not quite ready to go there yet
-
-% \paragraph{Universe-polymorphic definitions}
-% Coq has supported universe-polymorphism for parameterized inductive definitions (different instantiations of which are automatically of the lowest possible sort taking into account the universe level constraints on the arguments) for some time.
-However, the same functionality has not been available for \emph{definitions}, primarily because of efficiency concerns.
-In our development, bla bla misery and drama when doing this and that.
-In response to our inquiries about this issue, the Coq development team has agreed to implement ``opt-in'' universe polymorphism, where only expressly annotated definitions are made universe-polymorphic.
-This should solve our problems.
 
@@
 In more complex parts of our development we are now experiencing increasingly serious efficiency problems, despite having already made sacrifices by artificially inhibiting many natural class instances in order not to further strain instance resolution.
-%While we have already found the predicate-class-based style of formalization to work very well for our development thus far, and believe it has the potential to be \emph{the} modern solution to this problem, in our work we have encountered a number of limitations in the Coq implementation of key features of the tool, that keep us from making an unconditional recommendation for universal adoption of these techniques.
 Fortunately, there is plenty of potential for improvement of the current instance resolution implementation.
@@
 The use of category theory has been important in these developments.
-% Eelis: not sure i understand.
-"has been"?
-it's not done.
 
@@
 According to \lstinline|coqwc|, our development consists of 5660 lines of specifications and 937 lines of proofs.
-
-% Alt-Ergo congruence closure parametrized by an equational theory.
-% As usual either implement in Coq (verify the algorithm), or Check traces, or Check Alt-Ergo (Work
-% in progress).
-
-% http://lara.epfl.ch/dokuwiki/sav09:congruence_closure_algorithm_and_correctness
-% I understand that this is the algorithm underlying congruence and perhaps first-order.
 
